---
title: "Calculate Isoform Proportions - NMD Lung Cells"
author: "Analysis Script"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(data.table)
library(DT)
library(rtracklayer)
```

# Analysis Overview

**Objective:** Calculate per-sample isoform proportions for all genes across DMSO and Smg1i treatments.

**Approach:**
- Read raw transcript counts
- Filter out zero-count isoforms across all samples
- Map transcripts to genes using SQANTI (PB.x) and GTF (ENST)
- Calculate isoform proportions per sample, per gene
- Generate separate output files per cell type and treatment

# Source Files

**Input Data:**
- Raw counts: `data/qdf_raw_counts_2026.1.3.csv`
- Phenotype: `pheno/nmd_pheno_longreadbamids_2025.1.1.csv`
- SQANTI classification: `results/sqanti_runs/merged_collapsed/isoforms_classification_TPM.txt`
- GTF annotations: `reference_files/gencode.v49.primary_assembly.annotation.chrnamesedited.gtf`

**Output Directory:** `tmp/`

# Load Data

## Raw Counts Matrix

```{r load-counts}
cat("Loading raw counts matrix...\n")
counts <- fread("../data/qdf_raw_counts_2026.1.3.csv", data.table = FALSE)
cat("Dimensions:", nrow(counts), "transcripts x", ncol(counts), "columns\n")
head(counts[, 1:5])
```

## Phenotype Data

```{r load-pheno}
cat("Loading phenotype data...\n")
pheno <- read_csv("../pheno/nmd_pheno_longreadbamids_2025.1.1.csv")
cat("Dimensions:", nrow(pheno), "samples x", ncol(pheno), "columns\n")
head(pheno)
```

## Transcript-to-Gene Mappings

### SQANTI Classification (PB.x isoforms)

```{r load-sqanti}
cat("Loading SQANTI classification...\n")
sqanti <- fread("../results/sqanti_runs/merged_collapsed/isoforms_classification_TPM.txt",
                data.table = FALSE,
                select = c("isoform", "associated_gene"))
cat("Dimensions:", nrow(sqanti), "isoforms x", ncol(sqanti), "columns\n")

# Filter to only ENSG-mapped genes (exclude novelGenes)
sqanti_filtered <- sqanti %>%
  filter(str_starts(associated_gene, "ENSG"))

cat("After filtering to ENSG genes:", nrow(sqanti_filtered), "isoforms\n")
cat("Removed", nrow(sqanti) - nrow(sqanti_filtered), "novel gene isoforms\n")

# Keep novelGene isoforms for separate analysis
sqanti_novel <- sqanti %>%
  filter(!str_starts(associated_gene, "ENSG"))

cat("\nNovel gene isoforms:", nrow(sqanti_novel), "\n")
head(sqanti_filtered)
```

### GTF Annotations (ENST transcripts)

```{r load-gtf}
cat("Loading GTF file...\n")
gtf <- import("../reference_files/gencode.v49.primary_assembly.annotation.chrnamesedited.gtf")
gtf_df <- as.data.frame(gtf)

# Extract transcript to gene mapping
gtf_mapping <- gtf_df %>%
  filter(type == "transcript") %>%
  select(transcript_id, gene_id) %>%
  distinct()

cat("Dimensions:", nrow(gtf_mapping), "transcripts x", ncol(gtf_mapping), "columns\n")
head(gtf_mapping)
```

# Data Processing

## Filter Zero-Count Isoforms

Remove isoforms with zero counts across all 38 samples.

```{r filter-zeros}
# Get sample columns (skip first two: tname and length)
sample_cols <- 3:ncol(counts)

# Calculate row sums across samples
row_sums <- rowSums(counts[, sample_cols])

# Filter to non-zero isoforms
counts_filtered <- counts[row_sums > 0, ]

cat("Original transcripts:", nrow(counts), "\n")
cat("After filtering zeros:", nrow(counts_filtered), "\n")
cat("Removed:", nrow(counts) - nrow(counts_filtered), "zero-count transcripts\n")
```

## Prepare Counts Matrix

```{r prepare-counts}
# Extract transcript IDs and sample counts
transcript_ids <- counts_filtered[[1]]  # First column is transcript name
count_matrix <- counts_filtered[, sample_cols]
rownames(count_matrix) <- transcript_ids

cat("Count matrix dimensions:", nrow(count_matrix), "transcripts x", ncol(count_matrix), "samples\n")
```

## Map BAM Filenames to Sample Metadata

```{r map-samples}
# Extract BAM filename from full path
pheno <- pheno %>%
  mutate(bam_name = basename(bam),
         bam_name = str_remove(bam_name, "\\.bam$"))

# Get sample column names from count matrix
count_colnames <- colnames(count_matrix)

# Map count columns to phenotype
sample_map <- tibble(count_col = count_colnames) %>%
  left_join(pheno, by = c("count_col" = "bam_name"))

cat("Sample mapping:\n")
head(sample_map)

cat("\nSamples by treatment:\n")
table(sample_map$treatment)

cat("\nSamples by cell type:\n")
table(sample_map$ct)
```

## Analyze Novel Isoform Expression

```{r analyze-novel}
# Get expression data for novel isoforms
novel_transcripts <- sqanti_novel$isoform

# Filter count matrix to novel transcripts
novel_counts <- count_matrix[rownames(count_matrix) %in% novel_transcripts, ]

cat("Novel isoforms with expression data:", nrow(novel_counts), "\n")

# Calculate statistics per transcript
novel_expr <- novel_counts %>%
  as.data.frame() %>%
  rownames_to_column("transcript_id") %>%
  pivot_longer(cols = -transcript_id, names_to = "count_col", values_to = "raw_count") %>%
  left_join(sample_map %>% select(count_col, treatment, ct), by = "count_col") %>%
  group_by(transcript_id, treatment) %>%
  summarize(
    n_samples = n(),
    mean_count = mean(raw_count, na.rm = TRUE),
    median_count = median(raw_count, na.rm = TRUE),
    max_count = max(raw_count, na.rm = TRUE),
    n_nonzero = sum(raw_count > 0),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = treatment,
    values_from = c(mean_count, median_count, max_count, n_nonzero),
    names_glue = "{.value}_{treatment}"
  ) %>%
  left_join(sqanti_novel %>% select(isoform, associated_gene),
            by = c("transcript_id" = "isoform"))

# Add fold change
novel_expr <- novel_expr %>%
  mutate(
    fc_mean = ifelse(mean_count_DMSO > 0, mean_count_Smg1i / mean_count_DMSO, NA),
    log2fc_mean = log2(fc_mean)
  ) %>%
  arrange(desc(mean_count_DMSO + mean_count_Smg1i))

cat("\nTop novel isoforms by expression:\n")
if (nrow(novel_expr) > 1000) {
  cat("Showing top 1000 of", nrow(novel_expr), "novel isoforms\n")
  datatable(head(novel_expr, 1000), options = list(pageLength = 25, scrollX = TRUE), filter = 'top')
} else {
  datatable(novel_expr, options = list(pageLength = 25, scrollX = TRUE), filter = 'top')
}

# Write to file
novel_output_file <- "../tmp/novel_isoforms_expression_2026.1.3.tsv"
header_lines <- c(
  "# Novel Isoform Expression Summary",
  sprintf("# Generated: %s", Sys.time()),
  "# Script: code/calculate_isoform_proportions_2026.1.3.Rmd",
  "# ",
  "# Description: Expression levels of novel isoforms (not mapped to ENSG genes)",
  sprintf("# Novel isoforms: %d", nrow(novel_expr)),
  "# ",
  "# Columns:",
  "#   - transcript_id: PB.x.x isoform ID",
  "#   - mean_count_DMSO/Smg1i: Mean raw counts across all samples",
  "#   - median_count_DMSO/Smg1i: Median raw counts",
  "#   - max_count_DMSO/Smg1i: Maximum raw counts",
  "#   - n_nonzero_DMSO/Smg1i: Number of samples with non-zero counts",
  "#   - associated_gene: novelGene ID from SQANTI",
  "#   - fc_mean: Fold change (Smg1i/DMSO)",
  "#   - log2fc_mean: Log2 fold change",
  "#"
)

writeLines(header_lines, novel_output_file)
write_tsv(novel_expr, novel_output_file, append = TRUE)

cat(sprintf("\nNovel isoform data written to: %s\n", basename(novel_output_file)))
```

## Create Combined Transcript-to-Gene Mapping

```{r create-mapping}
# Combine SQANTI and GTF mappings
# SQANTI uses "isoform" and "associated_gene"
# GTF uses "transcript_id" and "gene_id"

sqanti_map <- sqanti_filtered %>%
  dplyr::rename(transcript_id = isoform, gene_id = associated_gene)

combined_mapping <- bind_rows(
  sqanti_map,
  gtf_mapping
) %>%
  distinct(transcript_id, .keep_all = TRUE)  # Keep first occurrence

cat("Combined mapping dimensions:", nrow(combined_mapping), "transcripts\n")
cat("Transcripts in counts matrix:", nrow(count_matrix), "\n")

# Check mapping coverage
mapped_transcripts <- sum(transcript_ids %in% combined_mapping$transcript_id)
cat("Transcripts successfully mapped:", mapped_transcripts,
    "(", round(100 * mapped_transcripts / length(transcript_ids), 1), "%)\n")
```

# Calculate Isoform Proportions

## Function to Calculate Proportions

```{r function-calc-props}
calculate_isoform_proportions <- function(counts_mat, transcript_to_gene, sample_metadata) {
  # Convert to long format
  counts_long <- counts_mat %>%
    as.data.frame() %>%
    rownames_to_column("transcript_id") %>%
    pivot_longer(cols = -transcript_id,
                 names_to = "count_col",
                 values_to = "raw_count")

  # Add gene IDs
  counts_long <- counts_long %>%
    left_join(transcript_to_gene, by = "transcript_id")

  # Add sample metadata
  counts_long <- counts_long %>%
    left_join(sample_metadata, by = "count_col")

  # Calculate gene totals per sample
  gene_totals <- counts_long %>%
    group_by(sample, ct, treatment, gene_id) %>%
    summarize(gene_total = sum(raw_count, na.rm = TRUE),
              n_isoforms = n_distinct(transcript_id),
              .groups = "drop")

  # Calculate isoform proportions
  isoform_props <- counts_long %>%
    left_join(gene_totals, by = c("sample", "ct", "treatment", "gene_id")) %>%
    mutate(isoform_proportion = ifelse(gene_total > 0, raw_count / gene_total, NA)) %>%
    select(cell_type = ct, treatment, sample, gene_id, transcript_id,
           raw_count, gene_total, isoform_proportion, n_isoforms) %>%
    filter(!is.na(gene_id))  # Remove unmapped transcripts

  return(isoform_props)
}
```

## Calculate for All Samples

```{r calc-all-props}
cat("Calculating isoform proportions for all samples...\n")
all_proportions <- calculate_isoform_proportions(count_matrix, combined_mapping, sample_map)

cat("Results dimensions:", nrow(all_proportions), "rows x", ncol(all_proportions), "columns\n")
head(all_proportions, 20)
```

## Filter Transcripts by Expression

Apply filtering criteria to remove low-expression, non-reproducible transcripts.

**Filter Criteria:**
- Total counts > 10 per treatment/cell type
- At least 2 samples with ≥2 counts

```{r filter-expression}
cat("Calculating filter statistics per transcript...\n")

# Calculate filter metrics per transcript per treatment per cell type
filter_stats <- all_proportions %>%
  group_by(cell_type, treatment, transcript_id, gene_id) %>%
  summarize(
    total_counts = sum(raw_count),
    n_samples_with_2plus = sum(raw_count >= 2),
    n_samples_total = n(),
    max_count = max(raw_count),
    .groups = "drop"
  )

cat("Initial transcripts:", n_distinct(all_proportions$transcript_id), "\n\n")

# Define filter criteria
# Keep if: total_counts > 10 AND n_samples_with_2plus >= 2
transcripts_pass <- filter_stats %>%
  filter(total_counts > 10 & n_samples_with_2plus >= 2) %>%
  distinct(transcript_id) %>%
  pull(transcript_id)

cat("Transcripts passing filter (in at least one treatment/cell type):",
    length(transcripts_pass), "\n")

# Identify high-count single-sample transcripts (interesting outliers)
high_count_single_sample <- filter_stats %>%
  filter(total_counts > 10 & n_samples_with_2plus == 1) %>%
  arrange(desc(total_counts))

cat("High-count single-sample transcripts (>10 counts, only 1 sample ≥2):",
    nrow(high_count_single_sample), "\n\n")

# Save high-count single-sample transcripts
if (nrow(high_count_single_sample) > 0) {
  high_count_file <- "../tmp/high_count_singlesample_transcripts_2026.1.3.tsv"
  header_lines <- c(
    "# High-Count Single-Sample Transcripts",
    sprintf("# Generated: %s", Sys.time()),
    "# Script: code/calculate_isoform_proportions_2026.1.3.Rmd",
    "# ",
    "# Description: Transcripts with >10 total counts but only 1 sample with ≥2 counts",
    "# These may represent sample-specific expression, artifacts, or biological outliers",
    sprintf("# Total transcripts: %d", nrow(high_count_single_sample)),
    "# ",
    "# Columns:",
    "#   - cell_type, treatment: Experimental condition",
    "#   - transcript_id, gene_id: Identifiers",
    "#   - total_counts: Sum of raw counts across all samples in this treatment/cell type",
    "#   - n_samples_with_2plus: Number of samples with ≥2 counts (always 1 for this file)",
    "#   - n_samples_total: Total samples in this treatment/cell type",
    "#   - max_count: Maximum count in any single sample",
    "#"
  )

  writeLines(header_lines, high_count_file)
  write_tsv(high_count_single_sample, high_count_file, append = TRUE)
  cat(sprintf("Saved to: %s\n\n", basename(high_count_file)))

  # Show examples in table
  cat("Examples of high-count single-sample transcripts:\n")
  if (nrow(high_count_single_sample) > 1000) {
    cat("Showing top 1000 of", nrow(high_count_single_sample), "transcripts\n")
    datatable(head(high_count_single_sample, 1000),
              options = list(pageLength = 25, scrollX = TRUE), filter = 'top')
  } else {
    datatable(high_count_single_sample,
              options = list(pageLength = 25, scrollX = TRUE), filter = 'top')
  }
}

# Filter main dataset
cat("\nApplying filter to main dataset...\n")
all_proportions_filtered <- all_proportions %>%
  filter(transcript_id %in% transcripts_pass)

cat("Rows before filter:", nrow(all_proportions), "\n")
cat("Rows after filter:", nrow(all_proportions_filtered), "\n")
cat("Reduction:", round(100 * (1 - nrow(all_proportions_filtered)/nrow(all_proportions)), 1), "%\n\n")

# Update variable name for downstream use
all_proportions <- all_proportions_filtered
```

## Summary Statistics

```{r summary-stats}
cat("\nSummary by cell type and treatment:\n")
summary_stats <- all_proportions %>%
  filter(raw_count > 0) %>%  # Only count transcripts with actual expression in this group
  group_by(cell_type, treatment) %>%
  summarize(
    n_samples = n_distinct(sample),
    n_genes = n_distinct(gene_id),
    n_transcripts = n_distinct(transcript_id),
    mean_isoforms_per_gene = n_distinct(transcript_id) / n_distinct(gene_id),
    .groups = "drop"
  )

datatable(summary_stats, options = list(pageLength = 20))
```

# Export Results

Generate separate files for each cell type and treatment combination.

```{r export-files}
# Cell type mapping
celltype_map <- c(
  "AT" = "at2",
  "DD" = "dd",
  "DD_ALI" = "ddali",
  "DO" = "do",
  "FB" = "fb",
  "MV" = "mv"
)

# Treatment mapping
treatment_map <- c(
  "DMSO" = "dmso",
  "Smg1i" = "smg1i"
)

# Get unique combinations
combinations <- all_proportions %>%
  distinct(cell_type, treatment) %>%
  arrange(cell_type, treatment)

cat("Generating", nrow(combinations), "output files:\n\n")

for (i in 1:nrow(combinations)) {
  ct <- combinations$cell_type[i]
  trt <- combinations$treatment[i]

  # Create filename
  ct_code <- celltype_map[ct]
  trt_code <- treatment_map[trt]
  filename <- sprintf("../tmp/isoform_proportions_%s_%s_2026.1.3.tsv", ct_code, trt_code)

  # Filter data
  subset_data <- all_proportions %>%
    filter(cell_type == ct, treatment == trt)

  # Create header
  header_lines <- c(
    sprintf("# Isoform Proportions: %s - %s", ct, trt),
    sprintf("# Generated: %s", Sys.time()),
    sprintf("# Script: code/calculate_isoform_proportions_2026.1.3.Rmd"),
    "# ",
    "# Source Data:",
    "#   - Raw counts: data/qdf_raw_counts_2026.1.3.csv",
    "#   - Phenotype: pheno/nmd_pheno_longreadbamids_2025.1.1.csv",
    "#   - SQANTI: results/sqanti_runs/merged_collapsed/isoforms_classification_TPM.txt",
    "#   - GTF: reference_files/gencode.v49.primary_assembly.annotation.chrnamesedited.gtf",
    "# ",
    sprintf("# Samples: %d", n_distinct(subset_data$sample)),
    sprintf("# Genes: %d", n_distinct(subset_data$gene_id)),
    sprintf("# Transcripts: %d", n_distinct(subset_data$transcript_id)),
    "# ",
    "# Filters applied:",
    "#   - Removed isoforms with zero counts across all 38 samples",
    "#   - Filtered to ENSG-mapped genes only (excluded novelGenes)",
    "#   - Expression filter: total_counts > 10 AND ≥2 samples with ≥2 counts per treatment/cell type",
    "#   - High-count single-sample transcripts saved separately",
    sprintf("#   - cell_type == '%s'", ct),
    sprintf("#   - treatment == '%s'", trt),
    "#"
  )

  # Write file with column headers
  writeLines(header_lines, filename)
  write_tsv(subset_data, filename, append = TRUE, col_names = TRUE)

  cat(sprintf("%d. %s (%d rows)\n", i, basename(filename), nrow(subset_data)))
}

cat("\nAll files written to tmp/ directory\n")
```

# Preview Results

Show a sample of results from DD + DMSO (largest sample size).

```{r preview-results}
dd_dmso <- all_proportions %>%
  filter(cell_type == "DD", treatment == "DMSO")

cat("DD + DMSO results:", nrow(dd_dmso), "rows\n")
cat("Samples:", unique(dd_dmso$sample), "\n\n")

# Show genes with multiple isoforms
multi_isoform <- dd_dmso %>%
  filter(n_isoforms > 1) %>%
  arrange(desc(n_isoforms), gene_id, sample, desc(isoform_proportion))

if (nrow(multi_isoform) > 1000) {
  cat("Showing top 1000 of", nrow(multi_isoform), "rows\n")
  datatable(head(multi_isoform, 1000),
            options = list(pageLength = 25, scrollX = TRUE),
            filter = 'top')
} else {
  datatable(multi_isoform,
            options = list(pageLength = 25, scrollX = TRUE),
            filter = 'top')
}
```

# Session Info

```{r session-info}
sessionInfo()
```
