---
title: "Gene-Level Expression Comparison: Isoform-Aggregated vs Direct Quantification"
author: "NMD Study"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# Overview

This analysis compares gene-level expression quantifications derived from two different sources:

1. **Isoform-aggregated counts**: Transcript-level counts (long-read PacBio data) summed to gene level
2. **Direct gene-level counts**: Gene-level quantifications from independent pipeline

The comparison evaluates technical concordance between quantification methods and identifies genes with discrepant expression estimates.

## Key Considerations

- Only transcripts with Ensembl gene IDs are included in aggregation
- Datasets are matched by common Ensembl gene IDs and samples
- TMM normalization is applied separately to each matched dataset
- Comparisons are performed at multiple levels: overall, per-sample, and per-cell-type

---

# Load Libraries

```{r load-libraries}
library(edgeR)
library(tidyverse)
library(knitr)
library(SummarizedExperiment)
```

---

# Comparison Function

Define a modular function to compare gene-level expression between isoform-aggregated and direct gene-level quantifications.

```{r comparison-function}
compare_gene_expression <- function(dge_isoform,
                                     gene_data,
                                     gene_id_col = "ensembl_gene_id",
                                     output_prefix = "gene_expression_comparison") {

  # Prepare gene-level data ------------------------------------------------------
  # Handle DGEList, SummarizedExperiment, or counts matrix input

  if (is(gene_data, "DGEList")) {
    dge_gene <- gene_data
    cat("Gene-level input: DGEList with", nrow(dge_gene), "genes\n")
  } else if (is(gene_data, "SummarizedExperiment")) {
    # Extract counts from SummarizedExperiment
    cat("Gene-level input: SummarizedExperiment with", nrow(gene_data), "genes\n")

    # Get counts matrix
    gene_counts <- assay(gene_data, "counts")

    # Parse sample metadata from column names (format: CELLTYPE###_TREATMENT)
    sample_names <- colnames(gene_data)

    # Extract donor ID, cell type, and treatment
    parsed_metadata <- data.frame(
      sample_name = sample_names,
      stringsAsFactors = FALSE
    )

    # Parse format like "AT2001V_DMSO" -> ct="AT2", id="001V", treatment="DMSO"
    # Split by underscore
    split_names <- strsplit(sample_names, "_")
    parsed_metadata$treatment <- sapply(split_names, function(x) x[2])

    # Extract cell type and ID from first part (e.g., "AT2001V")
    # Donor ID is always 3 digits + 1 letter at the end (e.g., "001V")
    # Cell type is everything before that
    first_part <- sapply(split_names, function(x) x[1])
    parsed_metadata$id <- substr(first_part, nchar(first_part) - 3, nchar(first_part))
    parsed_metadata$ct <- substr(first_part, 1, nchar(first_part) - 4)

    cat("  Parsed sample metadata from", ncol(gene_counts), "samples\n")

    # Create DGEList
    dge_gene <- DGEList(counts = gene_counts,
                        samples = parsed_metadata)
  } else if (is.matrix(gene_data) || is.data.frame(gene_data)) {
    # Create DGEList from counts matrix using isoform sample info
    cat("Gene-level input: Counts matrix with", nrow(gene_data), "genes\n")
    cat("Creating DGEList using sample metadata from isoform data\n")
    dge_gene <- DGEList(counts = as.matrix(gene_data),
                        samples = dge_isoform$samples)
  } else {
    stop("gene_data must be a DGEList, SummarizedExperiment, or counts matrix")
  }

  # Report input dimensions
  cat("\nInput dimensions:\n")
  cat("  Isoform-level:", nrow(dge_isoform$counts), "transcripts x",
      ncol(dge_isoform$counts), "samples\n")
  cat("  Gene-level:", nrow(dge_gene$counts), "genes x",
      ncol(dge_gene$counts), "samples\n")

  # Filter and aggregate isoforms to gene level ---------------------------------
  # Only include transcripts with valid Ensembl gene IDs

  if (!gene_id_col %in% colnames(dge_isoform$genes)) {
    stop(paste("Column", gene_id_col, "not found in dge_isoform$genes"))
  }

  gene_ids <- dge_isoform$genes[[gene_id_col]]

  # Filter to transcripts with valid Ensembl gene IDs
  has_gene_id <- !is.na(gene_ids) & gene_ids != "" & gene_ids != "NA"

  # Exclude fusion transcripts (multi-gene IDs like "ENSG00000000003_ENSG00000000419")
  # These represent fusion transcripts that map to multiple genes
  is_fusion <- grepl("_ENSG", gene_ids, fixed = TRUE)

  cat("\nFiltering transcripts:\n")
  cat("  Total transcripts:", length(gene_ids), "\n")
  cat("  With valid gene IDs:", sum(has_gene_id),
      "(", round(100 * sum(has_gene_id) / length(gene_ids), 1), "%)\n")
  cat("  Fusion transcripts (excluded):", sum(is_fusion & has_gene_id),
      "(", round(100 * sum(is_fusion & has_gene_id) / sum(has_gene_id), 1), "%)\n")

  # Keep only non-fusion transcripts with valid gene IDs
  keep_transcripts <- has_gene_id & !is_fusion
  cat("  Kept for aggregation:", sum(keep_transcripts),
      "(", round(100 * sum(keep_transcripts) / length(gene_ids), 1), "%)\n")

  # Subset and aggregate
  isoform_counts_filtered <- dge_isoform$counts[keep_transcripts, ]
  gene_ids_filtered <- gene_ids[keep_transcripts]

  # Strip version numbers from gene IDs (e.g., "ENSG00000000003.16" -> "ENSG00000000003")
  gene_ids_filtered <- sub("\\.\\d+$", "", gene_ids_filtered)

  # Sum transcript counts by gene
  gene_counts_from_isoforms <- rowsum(isoform_counts_filtered,
                                      group = gene_ids_filtered)
  cat("Aggregated to", nrow(gene_counts_from_isoforms), "genes\n")

  # Match datasets by Ensembl gene ID --------------------------------------------
  # Find common genes and samples between datasets

  gene_ids_gene <- rownames(dge_gene$counts)

  # Strip version numbers from gene-level IDs (e.g., "ENSG00000000003.16" -> "ENSG00000000003")
  gene_ids_gene_clean <- sub("\\.\\d+$", "", gene_ids_gene)

  # Update rownames in gene counts
  rownames(dge_gene$counts) <- gene_ids_gene_clean

  common_genes <- intersect(rownames(gene_counts_from_isoforms), gene_ids_gene_clean)

  # Match samples by metadata (donor ID, cell type, treatment)
  # Create matching keys for both datasets

  # For isoform data
  iso_donor_id <- if ("id" %in% colnames(dge_isoform$samples)) {
    dge_isoform$samples$id
  } else if ("sample_id" %in% colnames(dge_isoform$samples)) {
    dge_isoform$samples$sample_id
  } else {
    stop("Could not find donor ID column in isoform samples")
  }

  iso_key <- paste(iso_donor_id,
                   dge_isoform$samples$ct,
                   dge_isoform$samples$treatment,
                   sep = "_")
  names(iso_key) <- colnames(gene_counts_from_isoforms)

  # For gene data
  gene_donor_id <- if ("sample_id" %in% colnames(dge_gene$samples)) {
    dge_gene$samples$sample_id
  } else if ("id" %in% colnames(dge_gene$samples)) {
    dge_gene$samples$id
  } else {
    stop("Could not find donor ID column in gene samples")
  }

  gene_key <- paste(gene_donor_id,
                    dge_gene$samples$ct,
                    dge_gene$samples$treatment,
                    sep = "_")
  names(gene_key) <- colnames(dge_gene$counts)

  # Find common sample keys
  common_keys <- intersect(iso_key, gene_key)

  cat("\nMatching datasets:\n")
  cat("  Common genes:", length(common_genes), "\n")
  cat("  Common samples (by metadata):", length(common_keys), "\n")

  if (length(common_genes) == 0) stop("No common genes found!")
  if (length(common_keys) == 0) stop("No common samples found!")

  # Get sample names for each dataset
  iso_samples <- names(iso_key)[iso_key %in% common_keys]
  gene_samples <- names(gene_key)[gene_key %in% common_keys]

  # Reorder gene samples to match isoform sample order
  iso_keys_ordered <- iso_key[iso_samples]
  gene_samples_ordered <- gene_samples[match(iso_keys_ordered, gene_key[gene_samples])]

  # Subset both datasets to common genes and matched samples
  counts_iso_matched <- gene_counts_from_isoforms[common_genes, iso_samples]
  counts_gene_matched <- dge_gene$counts[common_genes, gene_samples_ordered]

  # Rename gene columns to match isoform column names for consistency
  colnames(counts_gene_matched) <- colnames(counts_iso_matched)

  # Create matched DGEList objects -----------------------------------------------
  # Use sample metadata from isoform DGEList

  samples_matched <- dge_isoform$samples[iso_samples, ]

  dge_iso_matched <- DGEList(counts = counts_iso_matched,
                             samples = samples_matched)
  dge_gene_matched <- DGEList(counts = counts_gene_matched,
                              samples = samples_matched)

  # TMM normalize separately -----------------------------------------------------
  # Apply TMM normalization independently to each matched dataset

  dge_iso_matched <- calcNormFactors(dge_iso_matched, method = "TMM")
  dge_gene_matched <- calcNormFactors(dge_gene_matched, method = "TMM")

  cat("\nTMM normalization factors:\n")
  cat("  Isoform range: [",
      round(min(dge_iso_matched$samples$norm.factors), 3), ", ",
      round(max(dge_iso_matched$samples$norm.factors), 3), "]\n", sep = "")
  cat("  Gene range: [",
      round(min(dge_gene_matched$samples$norm.factors), 3), ", ",
      round(max(dge_gene_matched$samples$norm.factors), 3), "]\n", sep = "")

  # Calculate CPM
  cpm_iso <- cpm(dge_iso_matched, log = FALSE)
  cpm_gene <- cpm(dge_gene_matched, log = FALSE)

  # Calculate log-transformed CPM for correlations
  log_cpm_iso <- log2(cpm_iso + 1)
  log_cpm_gene <- log2(cpm_gene + 1)

  # Overall correlations ---------------------------------------------------------
  # Correlation across all genes and samples

  # Untransformed CPM
  overall_cor_spearman <- cor(as.vector(cpm_iso), as.vector(cpm_gene),
                              method = "spearman")
  overall_cor_pearson <- cor(as.vector(cpm_iso), as.vector(cpm_gene),
                             method = "pearson")

  # Log-transformed CPM
  overall_cor_pearson_log <- cor(as.vector(log_cpm_iso), as.vector(log_cpm_gene),
                                 method = "pearson")

  # Per-sample correlations ------------------------------------------------------
  # Correlation within each sample

  sample_cors <- data.frame(
    sample = iso_samples,
    spearman = sapply(iso_samples, function(s) {
      cor(cpm_iso[, s], cpm_gene[, s], method = "spearman")
    }),
    pearson = sapply(iso_samples, function(s) {
      cor(cpm_iso[, s], cpm_gene[, s], method = "pearson")
    }),
    pearson_log = sapply(iso_samples, function(s) {
      cor(log_cpm_iso[, s], log_cpm_gene[, s], method = "pearson")
    })
  )

  # Add sample metadata (exclude 'sample' column if it exists to avoid duplicates)
  metadata_cols <- samples_matched[iso_samples, ]
  if ("sample" %in% colnames(metadata_cols)) {
    metadata_cols$sample <- NULL
  }
  sample_cors <- cbind(sample_cors, metadata_cols)

  # Per-cell-type correlations ---------------------------------------------------
  # Correlation within each cell type (if cell type info available)

  ct_cors <- NULL
  if ("ct" %in% colnames(samples_matched)) {
    cell_types <- unique(samples_matched$ct)
    ct_cors <- lapply(cell_types, function(ct) {
      ct_samples <- iso_samples[samples_matched$ct == ct]
      ct_cpm_iso <- as.vector(cpm_iso[, ct_samples])
      ct_cpm_gene <- as.vector(cpm_gene[, ct_samples])
      ct_log_cpm_iso <- as.vector(log_cpm_iso[, ct_samples])
      ct_log_cpm_gene <- as.vector(log_cpm_gene[, ct_samples])

      data.frame(
        cell_type = ct,
        n_samples = length(ct_samples),
        spearman = cor(ct_cpm_iso, ct_cpm_gene, method = "spearman"),
        pearson = cor(ct_cpm_iso, ct_cpm_gene, method = "pearson"),
        pearson_log = cor(ct_log_cpm_iso, ct_log_cpm_gene, method = "pearson")
      )
    })
    ct_cors <- do.call(rbind, ct_cors)
  }

  # Gene-level statistics --------------------------------------------------------
  # Calculate mean expression and fold differences

  mean_cpm_iso <- rowMeans(cpm_iso)
  mean_cpm_gene <- rowMeans(cpm_gene)

  # Log2 fold difference (isoform vs gene)
  fold_diff <- log2((mean_cpm_iso + 1) / (mean_cpm_gene + 1))

  # Define expressed genes (mean CPM > 1)
  expressed_genes <- (mean_cpm_iso > 1) | (mean_cpm_gene > 1)

  # Define discrepant genes (>2-fold difference and expressed)
  discrepant_genes <- abs(fold_diff) > 1 & expressed_genes

  # Create comparison data frame
  comparison_df <- data.frame(
    gene = common_genes,
    mean_cpm_isoform = mean_cpm_iso,
    mean_cpm_gene = mean_cpm_gene,
    log2_fold_diff = fold_diff,
    abs_log2_fold_diff = abs(fold_diff),
    expressed = expressed_genes,
    discrepant = discrepant_genes,
    stringsAsFactors = FALSE
  )

  # Return results ---------------------------------------------------------------

  results <- list(
    comparison_df = comparison_df,
    sample_correlations = sample_cors,
    cell_type_correlations = ct_cors,
    overall_correlations = c(spearman = overall_cor_spearman,
                            pearson = overall_cor_pearson,
                            pearson_log = overall_cor_pearson_log),
    cpm_isoform = cpm_iso,
    cpm_gene = cpm_gene,
    log_cpm_isoform = log_cpm_iso,
    log_cpm_gene = log_cpm_gene,
    dge_isoform_matched = dge_iso_matched,
    dge_gene_matched = dge_gene_matched
  )

  # Save results -----------------------------------------------------------------

  saveRDS(results, file = paste0("../results/rds/", output_prefix, "_2026.2.7.rds"))
  write.csv(comparison_df,
            file = paste0("../results/", output_prefix, "_2026.2.7.csv"),
            row.names = FALSE)

  cat("\nResults saved to:\n")
  cat("  -", paste0("results/rds/", output_prefix, "_2026.2.7.rds\n"))
  cat("  -", paste0("results/", output_prefix, "_2026.2.7.csv\n"))

  return(results)
}
```

---

# Analysis 1: DGEList vs DGEList

Compare isoform-aggregated counts with direct gene-level quantifications (both in DGEList format).

## Load Data

```{r load-data}
# Load isoform-level DGEList (nofilter version)
dge_isoform <- readRDS("../rds/dge_isoform_nofilter_2026.2.7.rds")

# Load gene-level DGEList (nofilter version)
dge_gene <- readRDS("../rds/dge_gene_nofilter_2026.1.3.rds")

cat("Loaded DGEList objects:\n")
cat("  Isoform:", nrow(dge_isoform), "transcripts x", ncol(dge_isoform), "samples\n")
cat("  Gene:", nrow(dge_gene), "genes x", ncol(dge_gene), "samples\n")
```

## Identify Gene ID Column

```{r identify-gene-id-col}
# Check available columns in isoform annotation
cat("Available columns in dge_isoform$genes:\n")
print(colnames(dge_isoform$genes))

# Determine which column contains Ensembl gene IDs
gene_id_col <- if ("ensembl_gene_id" %in% colnames(dge_isoform$genes)) {
  "ensembl_gene_id"
} else if ("gene_id" %in% colnames(dge_isoform$genes)) {
  "gene_id"
} else if ("gene_id_ens115_sqanti" %in% colnames(dge_isoform$genes)) {
  "gene_id_ens115_sqanti"
} else if ("hgnc_id" %in% colnames(dge_isoform$genes)) {
  "hgnc_id"
} else {
  stop("Could not find Ensembl gene ID column")
}

cat("\nUsing gene ID column:", gene_id_col, "\n")
```

## Run Comparison

```{r run-comparison}
results1 <- compare_gene_expression(
  dge_isoform = dge_isoform,
  gene_data = dge_gene,
  gene_id_col = gene_id_col,
  output_prefix = "comparison_nofilter_dgelist"
)
```

---

# Results

## Overall Correlation

```{r overall-correlation}
cat("Overall correlation across all genes and samples:\n")
cat("  Spearman:", round(results1$overall_correlations["spearman"], 4), "\n")
cat("  Pearson [log2(CPM+1)]:", round(results1$overall_correlations["pearson_log"], 4), "\n")
```

**Overall correlations:**

- **Spearman** (rank-based): `r round(results1$overall_correlations["spearman"], 3)`
- **Pearson on log2(CPM+1)**: `r round(results1$overall_correlations["pearson_log"], 3)`

The Spearman correlation of `r round(results1$overall_correlations["spearman"], 3)` indicates `r if(results1$overall_correlations["spearman"] > 0.7) "good" else "moderate"` rank concordance, while the log-transformed Pearson of `r round(results1$overall_correlations["pearson_log"], 3)` shows `r if(results1$overall_correlations["pearson_log"] > 0.7) "strong" else "moderate"` linear concordance.

## Per-Sample Correlations

```{r per-sample-correlations}
# Summary statistics
cat("Per-sample Spearman correlation:\n")
cat("  Mean:", round(mean(results1$sample_correlations$spearman), 4), "\n")
cat("  Range: [", round(min(results1$sample_correlations$spearman), 4), ", ",
    round(max(results1$sample_correlations$spearman), 4), "]\n", sep = "")

# Display table with sample metadata
results1$sample_correlations %>%
  select(sample, ct, treatment, spearman, pearson_log) %>%
  arrange(desc(spearman)) %>%
  kable(digits = 4,
        caption = "Table 1: Per-sample correlations between isoform-aggregated and gene-level quantifications (Analysis 1)",
        col.names = c("Sample", "Cell Type", "Treatment", "Spearman", "Pearson (log)"))
```

Per-sample correlations range from **`r round(min(results1$sample_correlations$spearman), 3)`** to **`r round(max(results1$sample_correlations$spearman), 3)`**, with a mean of **`r round(mean(results1$sample_correlations$spearman), 3)`**.

## Gene-Level Summary

```{r gene-summary}
comp_df <- results1$comparison_df

cat("Gene-level statistics:\n")
cat("  Total common genes:", nrow(comp_df), "\n")
cat("  Expressed genes (mean CPM > 1):", sum(comp_df$expressed), "\n")
cat("  Discrepant genes (>2-fold diff, expressed):", sum(comp_df$discrepant), "\n")
cat("\nSummary statistics for expressed genes:\n")
cat("  Median absolute CPM difference:",
    round(median(abs(comp_df$mean_cpm_isoform - comp_df$mean_cpm_gene)[comp_df$expressed]), 2), "\n")
cat("  Mean absolute CPM difference:",
    round(mean(abs(comp_df$mean_cpm_isoform - comp_df$mean_cpm_gene)[comp_df$expressed]), 2), "\n")
cat("  Median |log2 fold difference|:",
    round(median(comp_df$abs_log2_fold_diff[comp_df$expressed]), 3), "\n")
```

Of **`r format(nrow(comp_df), big.mark=",")`** common genes, **`r format(sum(comp_df$expressed), big.mark=",")`** are expressed (mean CPM > 1) and **`r sum(comp_df$discrepant)`** show discrepant quantifications (>2-fold difference).

## Top Discrepant Genes

```{r top-discrepant}
comp_df %>%
  filter(expressed) %>%
  arrange(desc(abs_log2_fold_diff)) %>%
  head(20) %>%
  select(gene, log2_fold_diff, mean_cpm_isoform, mean_cpm_gene) %>%
  kable(digits = 3,
        caption = "Table 2: Top 20 genes with largest expression differences (Analysis 1)",
        col.names = c("Gene", "log2(Isoform/Gene)", "Mean CPM (Isoform)", "Mean CPM (Gene)"))
```

These genes show the largest discrepancies between quantification methods. Positive log2 fold differences indicate higher expression in isoform-aggregated data; negative values indicate higher expression in direct gene-level data.

## Gene Length Effects

```{r gene-length-analysis}
# Identify dominant isoform for each gene
# Use mean expression across all samples to determine dominant isoform

# Get isoform-level expression data
iso_cpm <- cpm(dge_isoform, log = FALSE)

# Filter to isoforms with valid gene IDs and not fusion transcripts
gene_ids <- dge_isoform$genes$gene_id_ens115_sqanti
has_gene_id <- !is.na(gene_ids) & gene_ids != "" & gene_ids != "NA"
is_fusion <- grepl("_ENSG", gene_ids, fixed = TRUE)
keep_tx <- has_gene_id & !is_fusion

# Strip version numbers and get filtered data
gene_ids_clean <- sub("\\.\\d+$", "", gene_ids[keep_tx])
iso_cpm_filtered <- iso_cpm[keep_tx, ]
iso_lengths <- dge_isoform$genes$length[keep_tx]

# Calculate mean expression per isoform across samples
mean_iso_expr <- rowMeans(iso_cpm_filtered)

# Create data frame with isoform info
iso_df <- data.frame(
  txid = rownames(iso_cpm_filtered),
  gene_id = gene_ids_clean,
  length = iso_lengths,
  mean_expr = mean_iso_expr,
  stringsAsFactors = FALSE
)

# Remove isoforms without length information
iso_df <- iso_df[!is.na(iso_df$length), ]

# Identify dominant isoform per gene (highest mean expression)
dominant_isoforms <- iso_df %>%
  group_by(gene_id) %>%
  arrange(desc(mean_expr)) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  select(gene_id, dominant_length = length, dominant_expr = mean_expr)

cat("Identified dominant isoforms for", nrow(dominant_isoforms), "genes\n")
cat("Gene length range:", min(dominant_isoforms$dominant_length), "-",
    max(dominant_isoforms$dominant_length), "bp\n")
cat("Median gene length:", median(dominant_isoforms$dominant_length), "bp\n")

# Merge with comparison data
comp_df_length <- comp_df %>%
  left_join(dominant_isoforms, by = c("gene" = "gene_id")) %>%
  filter(!is.na(dominant_length), expressed)

cat("\nGenes with length info and expressed:", nrow(comp_df_length), "\n")

# Analyze relationship between length and discrepancy
cor_length_discrepancy <- cor(comp_df_length$dominant_length,
                               comp_df_length$abs_log2_fold_diff,
                               method = "spearman")
cat("Spearman correlation between gene length and absolute fold difference:",
    round(cor_length_discrepancy, 4), "\n")

# Bin genes by length quartiles
comp_df_length <- comp_df_length %>%
  mutate(length_bin = cut(dominant_length,
                          breaks = quantile(dominant_length, probs = c(0, 0.25, 0.5, 0.75, 1)),
                          labels = c("Q1 (shortest)", "Q2", "Q3", "Q4 (longest)"),
                          include.lowest = TRUE))

# Calculate statistics by length bin
length_stats <- comp_df_length %>%
  group_by(length_bin) %>%
  summarise(
    n_genes = n(),
    median_length = median(dominant_length),
    median_abs_fold_diff = median(abs_log2_fold_diff),
    pct_discrepant = 100 * sum(discrepant) / n(),
    .groups = "drop"
  )

cat("\n")
print(kable(length_stats, digits = 2,
            caption = "Table 4: Gene length effects on quantification concordance",
            col.names = c("Length Quartile", "N Genes", "Median Length (bp)",
                         "Median |log2FC|", "% Discrepant")))
```

**Gene length analysis:**

Dominant isoform lengths were identified for **`r nrow(comp_df_length)`** expressed genes. The Spearman correlation between gene length and absolute fold difference is **`r round(cor_length_discrepancy, 3)`**, suggesting `r if(abs(cor_length_discrepancy) > 0.1) "a relationship" else "minimal relationship"` between gene length and quantification discrepancy.

```{r length-effect-scatter, fig.width=8, fig.height=6}
# Scatter plot: gene length vs discrepancy
ggplot(comp_df_length, aes(x = dominant_length, y = abs_log2_fold_diff)) +
  geom_point(alpha = 0.3, size = 1, color = "steelblue") +
  geom_smooth(method = "loess", color = "coral2", se = TRUE) +
  scale_x_log10() +
  labs(x = "Gene Length (bp, log scale)",
       y = "Absolute log2 Fold Difference",
       title = "Figure 5A: Gene Length vs Quantification Discrepancy (Analysis 1)") +
  theme_bw(base_size = 12)
```

```{r length-effect-boxplot, fig.width=8, fig.height=6}
# Boxplot: discrepancy by length quartile
ggplot(comp_df_length, aes(x = length_bin, y = abs_log2_fold_diff)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6, outlier.alpha = 0.3) +
  labs(x = "Gene Length Quartile",
       y = "Absolute log2 Fold Difference",
       title = "Figure 5B: Quantification Discrepancy by Length Quartile (Analysis 1)") +
  theme_bw(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The analysis shows `r if(abs(cor_length_discrepancy) > 0.1) paste0("that ", if(cor_length_discrepancy > 0) "longer" else "shorter", " genes tend to have larger discrepancies") else "minimal systematic effect of gene length on quantification concordance"`.

---

# Scatter Plot: Mean Expression

```{r scatter-plot, fig.width=8, fig.height=8}
# Prepare data for plotting
plot_df <- comp_df %>%
  mutate(category = case_when(
    discrepant ~ "Discrepant (>2-fold)",
    TRUE ~ "Concordant"
  ))

# Create scatter plot
ggplot(plot_df, aes(x = mean_cpm_gene, y = mean_cpm_isoform, color = category)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  scale_color_manual(values = c("Concordant" = "steelblue",
                                 "Discrepant (>2-fold)" = "coral2")) +
  labs(x = "Mean CPM (Direct Gene-Level)",
       y = "Mean CPM (Isoform-Aggregated)",
       title = "Figure 1: Comparison of Gene-Level Expression Quantifications (Analysis 1)",
       subtitle = paste0("Spearman r = ", round(results1$overall_correlations["spearman"], 3)),
       color = "Category") +
  theme_bw(base_size = 12) +
  theme(legend.position = "top",
        aspect.ratio = 1)
```

The scatter plot shows overall concordance between quantification methods on the original CPM scale. Discrepant genes (coral points) deviate from the diagonal unity line.

## Scatter Plot: Log2-Transformed Expression

```{r scatter-plot-log2, fig.width=8, fig.height=8}
# Prepare data with log2-transformed values
# These are the same values used for Pearson correlation calculation
plot_df_log2 <- comp_df %>%
  mutate(
    log2_cpm_gene = log2(mean_cpm_gene + 1),
    log2_cpm_isoform = log2(mean_cpm_isoform + 1),
    category = case_when(
      discrepant ~ "Discrepant (>2-fold)",
      TRUE ~ "Concordant"
    )
  )

# Create scatter plot with log2-transformed data
ggplot(plot_df_log2, aes(x = log2_cpm_gene, y = log2_cpm_isoform, color = category)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  scale_color_manual(values = c("Concordant" = "steelblue",
                                 "Discrepant (>2-fold)" = "coral2")) +
  labs(x = "log2(CPM + 1) - Direct Gene-Level",
       y = "log2(CPM + 1) - Isoform-Aggregated",
       title = "Figure 2: Log2-Transformed Expression Comparison (Analysis 1)",
       subtitle = paste0("Pearson r = ", round(results1$overall_correlations["pearson_log"], 3),
                        " (same data used for correlation)"),
       color = "Category") +
  theme_bw(base_size = 12) +
  theme(legend.position = "top",
        aspect.ratio = 1)
```

This plot shows the same comparison using log2-transformed CPM values, which is the data used for calculating the Pearson correlation (r = `r round(results1$overall_correlations["pearson_log"], 3)`). The linear scale reveals the correlation structure more clearly than the log-log plot.

---

# MA Plot: Difference vs Average

```{r ma-plot, fig.width=10, fig.height=6}
# Prepare data for MA plot
ma_df <- comp_df %>%
  mutate(
    A = 0.5 * (log2(mean_cpm_isoform + 1) + log2(mean_cpm_gene + 1)),
    M = log2_fold_diff,
    category = case_when(
      discrepant ~ "Discrepant (>2-fold)",
      TRUE ~ "Concordant"
    )
  )

# Create MA plot
ggplot(ma_df, aes(x = A, y = M, color = category)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_hline(yintercept = c(-1, 1), linetype = "dotted", color = "red") +
  scale_color_manual(values = c("Concordant" = "steelblue",
                                 "Discrepant (>2-fold)" = "coral2")) +
  labs(x = "Average log2(CPM + 1)",
       y = "log2(Isoform / Gene)",
       title = "Figure 3: MA Plot - Difference vs Average Expression (Analysis 1)",
       subtitle = "Dotted lines indicate ±2-fold difference",
       color = "Category") +
  theme_bw(base_size = 12) +
  theme(legend.position = "top")
```

The MA plot visualizes fold differences as a function of average expression level. Most genes cluster around M=0 (no difference), with discrepant genes showing systematic deviations.

---

# Sample Correlation Heatmap

```{r correlation-heatmap, fig.width=10, fig.height=4}
# Extract sample correlations by cell type and treatment
cor_plot_df <- results1$sample_correlations %>%
  select(sample, ct, treatment, spearman) %>%
  arrange(ct, treatment)

# Create bar plot
ggplot(cor_plot_df, aes(x = reorder(sample, spearman), y = spearman, fill = ct)) +
  geom_col() +
  geom_hline(yintercept = mean(cor_plot_df$spearman),
             linetype = "dashed", color = "red") +
  facet_wrap(~treatment, scales = "free_x") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Sample",
       y = "Spearman Correlation",
       title = "Figure 4: Per-Sample Correlations by Cell Type and Treatment (Analysis 1)",
       subtitle = paste0("Red line indicates mean correlation (",
                        round(mean(cor_plot_df$spearman), 3), ")"),
       fill = "Cell Type") +
  theme_bw(base_size = 11) +
  theme(legend.position = "right")
```

Per-sample correlations show `r if(max(cor_plot_df$spearman) - min(cor_plot_df$spearman) < 0.05) "consistent" else "variable"` concordance across cell types and treatments.

---

# Session Info

```{r session-info}
sessionInfo()
```

---

# Analysis 2: Length-Corrected Gene Counts (Salmon)

Compare isoform-aggregated counts with length-corrected gene-level quantifications from Salmon. Salmon quantifications account for effective gene length, which may reduce discrepancies caused by isoform length differences.

## Load Salmon Data

```{r load-salmon-data}
# Load Salmon SummarizedExperiment
se_salmon <- readRDS("../data/salmon.merged.gene_counts_scaled.rds")

cat("Loaded Salmon SummarizedExperiment:\n")
cat("  Dimensions:", nrow(se_salmon), "genes x", ncol(se_salmon), "samples\n")
cat("  Assays:", paste(assayNames(se_salmon), collapse = ", "), "\n")
```

## Run Comparison with Salmon Data

```{r run-comparison-salmon}
results2 <- compare_gene_expression(
  dge_isoform = dge_isoform,
  gene_data = se_salmon,
  gene_id_col = gene_id_col,
  output_prefix = "comparison_nofilter_salmon"
)
```

## Salmon Results

### Overall Correlation

```{r overall-correlation-salmon}
cat("Overall correlation (Salmon length-corrected):\n")
cat("  Spearman:", round(results2$overall_correlations["spearman"], 4), "\n")
cat("  Pearson [log2(CPM+1)]:", round(results2$overall_correlations["pearson_log"], 4), "\n")
```

**Salmon length-corrected correlations:**

- **Spearman**: `r round(results2$overall_correlations["spearman"], 3)`
- **Pearson on log2(CPM+1)**: `r round(results2$overall_correlations["pearson_log"], 3)`

### Comparison with Analysis 1

```{r comparison-analysis1-vs-2}
comp_table <- data.frame(
  Method = c("Standard gene counts", "Salmon length-corrected"),
  Spearman = c(results1$overall_correlations["spearman"],
               results2$overall_correlations["spearman"]),
  Pearson_log = c(results1$overall_correlations["pearson_log"],
                  results2$overall_correlations["pearson_log"])
)

kable(comp_table, digits = 4,
      caption = "Table 3: Comparison of correlation metrics between standard and length-corrected quantifications",
      col.names = c("Method", "Spearman", "Pearson [log2(CPM+1)]"))
```

Length correction `r if(results2$overall_correlations["pearson_log"] > results1$overall_correlations["pearson_log"]) "improves" else "does not substantially improve"` the log-transformed Pearson correlation by `r round(abs(results2$overall_correlations["pearson_log"] - results1$overall_correlations["pearson_log"]), 3)`.

### Scatter Plot: Salmon vs Isoform-Aggregated

```{r scatter-plot-salmon, fig.width=8, fig.height=8}
# Prepare data for plotting
comp_df2 <- results2$comparison_df
plot_df2 <- comp_df2 %>%
  mutate(category = case_when(
    discrepant ~ "Discrepant (>2-fold)",
    TRUE ~ "Concordant"
  ))

# Create scatter plot
ggplot(plot_df2, aes(x = mean_cpm_gene, y = mean_cpm_isoform, color = category)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  scale_color_manual(values = c("Concordant" = "steelblue",
                                 "Discrepant (>2-fold)" = "coral2")) +
  labs(x = "Mean CPM (Salmon Length-Corrected)",
       y = "Mean CPM (Isoform-Aggregated)",
       title = "Figure 6: Comparison with Salmon Length-Corrected Quantifications (Analysis 2)",
       subtitle = paste0("Spearman r = ", round(results2$overall_correlations["spearman"], 3),
                        ", Pearson [log2] r = ", round(results2$overall_correlations["pearson_log"], 3)),
       color = "Category") +
  theme_bw(base_size = 12) +
  theme(legend.position = "top",
        aspect.ratio = 1)
```

The scatter plot shows the comparison between Salmon length-corrected quantifications and isoform-aggregated counts. Salmon quantifications account for effective gene length but show slightly lower correlation compared to standard gene counts (Analysis 1).

### Scatter Plot: Log2-Transformed Expression (Salmon)

```{r scatter-plot-salmon-log2, fig.width=8, fig.height=8}
# Prepare data with log2-transformed values
# These are the same values used for Pearson correlation calculation
plot_df2_log2 <- comp_df2 %>%
  mutate(
    log2_cpm_gene = log2(mean_cpm_gene + 1),
    log2_cpm_isoform = log2(mean_cpm_isoform + 1),
    category = case_when(
      discrepant ~ "Discrepant (>2-fold)",
      TRUE ~ "Concordant"
    )
  )

# Create scatter plot with log2-transformed data
ggplot(plot_df2_log2, aes(x = log2_cpm_gene, y = log2_cpm_isoform, color = category)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  scale_color_manual(values = c("Concordant" = "steelblue",
                                 "Discrepant (>2-fold)" = "coral2")) +
  labs(x = "log2(CPM + 1) - Salmon Length-Corrected",
       y = "log2(CPM + 1) - Isoform-Aggregated",
       title = "Figure 7: Log2-Transformed Expression Comparison - Salmon (Analysis 2)",
       subtitle = paste0("Pearson r = ", round(results2$overall_correlations["pearson_log"], 3),
                        " (same data used for correlation)"),
       color = "Category") +
  theme_bw(base_size = 12) +
  theme(legend.position = "top",
        aspect.ratio = 1)
```

This plot shows the Salmon comparison using log2-transformed CPM values, which is the data used for calculating the Pearson correlation (r = `r round(results2$overall_correlations["pearson_log"], 3)`). The linear scale reveals the correlation structure more clearly than the log-log plot.
