---
title: "Isoform-Level DIE Summary: NMD Inhibition Effects"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

<style>
body {
  font-size: 16px;
}
p {
  font-size: 16px;
  line-height: 1.6;
}
li {
  font-size: 16px;
  line-height: 1.6;
}
</style>

# Analysis Overview

This report summarizes Differential Isoform Expression (DIE) results from the NMD inhibition study and performs functional enrichment analysis to identify biological pathways affected by NMD.

**Experimental Design:**

- **Treatment:** Smg1i (NMD inhibitor) vs DMSO (control)
- **Cell Types:** 6 cell types (AT, DD, DD_ALI, DO, FB, MV)
- **Data Type:** Long-read PacBio IsoSeq (isoform-level resolution)
- **Analysis:** limma-voom differential expression at transcript level

**Report Structure:**

1. **Part 1: DIE Summary** - Overview of significant differential isoform expression results
2. **Part 2: Functional Enrichment** - Gene Set Enrichment Analysis (GSEA) using three NMD-responsive gene definitions

# Load Libraries and Configuration

```{r load-libs}
# Set CRAN mirror for package installation
options(repos = c(CRAN = "https://cloud.r-project.org"))

library(tidyverse)
library(data.table)
library(DT)
library(edgeR)

# Check and install fgsea (Bioconductor)
if (!require("fgsea", quietly = TRUE)) {
  if (!require("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  BiocManager::install("fgsea")
  library(fgsea)
}

# Check and install msigdbr (CRAN)
if (!require("msigdbr", quietly = TRUE)) {
  install.packages("msigdbr")
  library(msigdbr)
}

# Check and install pheatmap (CRAN)
if (!require("pheatmap", quietly = TRUE)) {
  install.packages("pheatmap")
  library(pheatmap)
}

library(fgsea)
library(msigdbr)
library(pheatmap)
```

```{r configuration}
# Date stamp for output files
DATE_STAMP <- "2026.1.15"

# Input paths
INPUT_DIE_DIR <- file.path("..", "longread_dge")
INPUT_GENE_DGE_DIR <- file.path("..", "shortread_dge")
INPUT_DGELIST <- file.path("..", "results", "rds", "dge_isoform_2026.1.3.rds")
OUTPUT_DIR <- file.path("..", "tmp")

# Cell type mapping
CELLTYPE_MAP <- c(
  "at2" = "AT", "dd" = "DD", "ddali" = "DD_ALI",
  "doali" = "DO", "fb" = "FB", "mv" = "MV"
)

# Analysis thresholds
SIGNIFICANCE_THRESHOLD <- 0.05
LOGFC_THRESHOLD <- 0  # For NMD-responsive (upregulated)
TOP_HITS_N <- 50

# GSEA parameters
GSEA_MIN_SIZE <- 15
GSEA_MAX_SIZE <- 500
GSEA_NPERM <- 100000  # Increased from 10K to 100K to improve p-value resolution

cat("Configuration loaded successfully\n")
cat("Date stamp:", DATE_STAMP, "\n")
cat("Output directory:", OUTPUT_DIR, "\n")
```

# PART 1: DIE Summary

## Load Isoform-Level DIE Results

```{r load-die}
cat("Loading isoform-level DIE results from longread_dge/...\n\n")

# Load all DIE files
die_files <- list.files(INPUT_DIE_DIR,
                        pattern = "^nmd_dge_.*_2026\\.1\\.2\\.csv$",
                        full.names = TRUE)

cat("Found", length(die_files), "DIE files:\n")
print(basename(die_files))
cat("\n")

# First pass: count transcripts in each file
original_counts <- map_df(die_files, ~{
  # Extract cell type from filename
  basename_file <- basename(.x)
  parts <- str_split(basename_file, "_")[[1]]
  celltype <- parts[3]  # "at2", "dd", "ddali", etc.

  # Count lines in file (subtract 1 for header)
  n_lines <- as.numeric(system(paste0("wc -l < '", .x, "'"), intern = TRUE))
  n_transcripts <- n_lines - 1

  tibble(
    cell_type_file = celltype,
    cell_type = CELLTYPE_MAP[celltype],
    original_count = n_transcripts
  )
})

cat("\nOriginal transcript counts per file:\n")
print(original_counts)
cat("\n")

# Second pass: load and combine data
all_dge <- map_df(die_files, ~{
  df <- read_csv(.x, show_col_types = FALSE)

  # Extract cell type from filename
  basename_file <- basename(.x)
  parts <- str_split(basename_file, "_")[[1]]
  celltype <- parts[3]  # "at2", "dd", "ddali", etc.

  df$cell_type_file <- celltype
  df
})

# Map to display names
all_dge$cell_type <- CELLTYPE_MAP[all_dge$cell_type_file]
all_dge <- all_dge %>% filter(!is.na(cell_type))

cat("Combined DIE data:\n")
cat("- Total rows:", nrow(all_dge), "\n")
cat("- Transcripts per cell type:", nrow(all_dge) / length(unique(all_dge$cell_type)), "\n")
cat("- Cell types:", paste(unique(all_dge$cell_type), collapse = ", "), "\n")
cat("\n")

# Rename txid column if needed for consistency
if (!"transcript_id" %in% colnames(all_dge) && "txid" %in% colnames(all_dge)) {
  all_dge <- all_dge %>% rename(transcript_id = txid)
}
```

## Add Gene Symbols

```{r add-gene-symbols}
cat("## Loading gene symbols from shortread_dge/\n\n")

# Load gene-level DGE files to extract gene symbols
gene_dge_files <- list.files(INPUT_GENE_DGE_DIR,
                              pattern = "^nmd_sr_dge_.*_2026\\.1\\.2\\.csv$",
                              full.names = TRUE)

cat("Found", length(gene_dge_files), "gene-level DGE files for gene symbol mapping\n\n")

gene_dge_for_symbols <- map_df(gene_dge_files, ~{
  read_csv(.x, show_col_types = FALSE) %>%
    select(ensembl_gene_id_version, hgnc_symbol)
})

# Create gene symbol mapping (Ensembl ID without version -> gene symbol)
gene_symbol_map <- gene_dge_for_symbols %>%
  mutate(gene_id_no_version = str_replace(ensembl_gene_id_version, "\\.\\d+$", "")) %>%
  select(gene_id_no_version, hgnc_symbol) %>%
  distinct()

cat("Created gene symbol mapping:", nrow(gene_symbol_map), "genes\n\n")

# Add gene symbols to all_dge
all_dge <- all_dge %>%
  left_join(gene_symbol_map, by = c("hgnc_id" = "gene_id_no_version")) %>%
  mutate(gene_symbol = coalesce(hgnc_symbol, hgnc_id))  # Use symbol if available, otherwise Ensembl ID

cat("Added gene symbols to DIE data\n")
cat("- Transcripts with gene symbols:", sum(!is.na(all_dge$hgnc_symbol)), "\n")
cat("- Transcripts without gene symbols:", sum(is.na(all_dge$hgnc_symbol)), "\n\n")
```

## Overall Summary Statistics

```{r summary-stats}
cat("## DIE Summary Statistics by Cell Type\n\n")

die_summary <- all_dge %>%
  group_by(cell_type) %>%
  summarize(
    sig_transcripts = sum(adj.P.Val < SIGNIFICANCE_THRESHOLD, na.rm = TRUE),
    nmd_responsive = sum(adj.P.Val < SIGNIFICANCE_THRESHOLD & logFC > LOGFC_THRESHOLD, na.rm = TRUE),
    nmd_suppressed = sum(adj.P.Val < SIGNIFICANCE_THRESHOLD & logFC < -LOGFC_THRESHOLD, na.rm = TRUE),
    mean_logFC_nmd = mean(logFC[adj.P.Val < SIGNIFICANCE_THRESHOLD & logFC > LOGFC_THRESHOLD], na.rm = TRUE),
    median_logFC_nmd = median(logFC[adj.P.Val < SIGNIFICANCE_THRESHOLD & logFC > LOGFC_THRESHOLD], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(original_counts %>% select(cell_type, original_count), by = "cell_type") %>%
  rename(total_transcripts = original_count) %>%
  mutate(
    pct_sig = 100 * sig_transcripts / total_transcripts,
    pct_nmd_responsive = 100 * nmd_responsive / sig_transcripts
  ) %>%
  select(cell_type, total_transcripts, sig_transcripts, nmd_responsive, nmd_suppressed,
         pct_sig, pct_nmd_responsive, mean_logFC_nmd, median_logFC_nmd)

cat("**Table 1. DIE Summary Statistics by Cell Type**\n\n")
cat("Column definitions:\n")
cat("- **cell_type**: Cell type identifier\n")
cat("- **total_transcripts**: Total number of transcripts tested in original DIE analysis\n")
cat("- **sig_transcripts**: Number of transcripts with adjusted p-value < 0.05\n")
cat("- **nmd_responsive**: Number of transcripts significantly upregulated (logFC > 0, adj.P.Val < 0.05)\n")
cat("- **nmd_suppressed**: Number of transcripts significantly downregulated (logFC < 0, adj.P.Val < 0.05)\n")
cat("- **pct_sig**: Percentage of significant transcripts (out of total transcripts tested)\n")
cat("- **pct_nmd_responsive**: Percentage of NMD-responsive transcripts (out of significant transcripts)\n")
cat("- **mean_logFC_nmd**: Mean log2 fold change of NMD-responsive transcripts\n")
cat("- **median_logFC_nmd**: Median log2 fold change of NMD-responsive transcripts\n\n")

datatable(die_summary,
          options = list(pageLength = 10),
          caption = "Table 1. DIE Summary: Significant transcripts and NMD-responsive counts per cell type") %>%
  formatRound(columns = c("pct_sig", "pct_nmd_responsive", "mean_logFC_nmd", "median_logFC_nmd"), digits = 2)

# Save summary table
write_tsv(die_summary, file.path(OUTPUT_DIR, paste0("die_summary_stats_", DATE_STAMP, ".tsv")))
cat("\nSaved:", file.path(OUTPUT_DIR, paste0("die_summary_stats_", DATE_STAMP, ".tsv")), "\n")
```

## NMD-Responsive Transcripts

```{r nmd-responsive}
cat("## NMD-Responsive Transcripts (logFC > 0, adj.P.Val < 0.05)\n\n")

nmd_responsive <- all_dge %>%
  filter(logFC > LOGFC_THRESHOLD, adj.P.Val < SIGNIFICANCE_THRESHOLD)

cat("Total NMD-responsive transcript-celltype combinations:", nrow(nmd_responsive), "\n")
cat("Unique NMD-responsive transcripts (across all cell types):", n_distinct(nmd_responsive$transcript_id), "\n\n")

# Summary by cell type
nmd_summary <- nmd_responsive %>%
  group_by(cell_type) %>%
  summarize(
    n_transcripts = n(),
    n_genes = n_distinct(hgnc_id, na.rm = TRUE),
    mean_logFC = mean(logFC),
    median_logFC = median(logFC),
    max_logFC = max(logFC),
    mean_AveExpr = mean(AveExpr, na.rm = TRUE),
    .groups = "drop"
  )

cat("\n**Table 2. NMD-Responsive Transcripts by Cell Type**\n\n")
cat("Column definitions:\n")
cat("- **cell_type**: Cell type identifier\n")
cat("- **n_transcripts**: Number of NMD-responsive transcripts in this cell type (logFC > 0, adj.P.Val < 0.05)\n")
cat("- **n_genes**: Number of unique genes with NMD-responsive transcripts in this cell type\n")
cat("- **mean_logFC**: Mean log2 fold change (Smg1i vs DMSO)\n")
cat("- **median_logFC**: Median log2 fold change\n")
cat("- **max_logFC**: Maximum log2 fold change\n")
cat("- **mean_AveExpr**: Mean average expression level (log2 CPM)\n\n")

datatable(nmd_summary,
          options = list(pageLength = 10),
          caption = "Table 2. NMD-responsive transcripts by cell type") %>%
  formatRound(columns = c("mean_logFC", "median_logFC", "max_logFC", "mean_AveExpr"), digits = 2)

# Save all NMD-responsive transcripts
write_tsv(nmd_responsive, file.path(OUTPUT_DIR, paste0("die_nmd_responsive_all_", DATE_STAMP, ".tsv")))
cat("\nSaved:", file.path(OUTPUT_DIR, paste0("die_nmd_responsive_all_", DATE_STAMP, ".tsv")), "\n")
```

## Top Hits per Cell Type

```{r top-hits, results='asis'}
cat("## Top", TOP_HITS_N, "NMD-Responsive Transcripts per Cell Type\n\n")

library(htmltools)

table_num <- 3  # Starting table number for top hits
tables_list <- list()

for (ct in unique(nmd_responsive$cell_type)) {
  cat("\n###", ct, "\n\n")

  top_hits <- nmd_responsive %>%
    filter(cell_type == ct) %>%
    arrange(adj.P.Val) %>%
    head(TOP_HITS_N) %>%
    select(transcript_id, gene_symbol, transcript.gc, biotype, logFC, adj.P.Val, AveExpr, t, B)

  cat("**Table", table_num, ". Top", TOP_HITS_N, "NMD-Responsive Transcripts in", ct, "**\n\n")
  cat("Column definitions:\n")
  cat("- **transcript_id**: Transcript identifier\n")
  cat("- **gene_symbol**: Gene symbol (HUGO/HGNC)\n")
  cat("- **transcript.gc**: Transcript name from GENCODE annotation\n")
  cat("- **biotype**: Transcript biotype (protein_coding, lncRNA, etc.)\n")
  cat("- **logFC**: Log2 fold change (Smg1i vs DMSO)\n")
  cat("- **adj.P.Val**: FDR-adjusted p-value\n")
  cat("- **AveExpr**: Average expression level (log2 CPM)\n")
  cat("- **t**: t-statistic from differential expression test\n")
  cat("- **B**: B-statistic (log-odds of differential expression)\n\n")

  # Create datatable and add to list
  dt <- datatable(top_hits,
              options = list(pageLength = 10),
              caption = paste0("Table ", table_num, ". Top ", TOP_HITS_N, " NMD-responsive transcripts in ", ct)) %>%
      formatRound(columns = c("logFC", "adj.P.Val", "AveExpr", "t", "B"), digits = 3)

  print(htmltools::tagList(dt))
  cat("\n\n")

  table_num <- table_num + 1

  # Save top hits
  output_file <- file.path(OUTPUT_DIR, paste0("die_top_hits_", tolower(ct), "_", DATE_STAMP, ".tsv"))
  write_tsv(top_hits, output_file)
  cat("\nSaved:", output_file, "\n")
}
```

## Volcano Plots

```{r volcano-plots, fig.width=12, fig.height=8}
cat("## Volcano Plots: DIE by Cell Type\n\n")

volcano_data <- all_dge %>%
  mutate(
    neg_log10_p = -log10(adj.P.Val + 1e-300),  # Avoid log(0)
    significance = case_when(
      logFC > 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD ~ "NMD-responsive",
      logFC < 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD ~ "NMD-suppressed",
      TRUE ~ "Not significant"
    )
  )

ggplot(volcano_data, aes(x = logFC, y = neg_log10_p, color = significance)) +
  geom_point(alpha = 0.5, size = 0.5) +
  facet_wrap(~cell_type, ncol = 3) +
  scale_color_manual(values = c(
    "NMD-responsive" = "steelblue",
    "NMD-suppressed" = "coral",
    "Not significant" = "gray70"
  )) +
  geom_hline(yintercept = -log10(SIGNIFICANCE_THRESHOLD), linetype = "dashed", color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "Figure 1: Volcano Plots - Isoform-Level Differential Expression",
    subtitle = "Point colors based on adj.P.Val < 0.05 and logFC direction",
    x = "Log2 Fold Change (Smg1i vs DMSO)",
    y = "-Log10(Adjusted P-Value)",
    color = "Status"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```

## MA Plots

```{r ma-plots, fig.width=12, fig.height=8}
cat("## MA Plots: Expression vs Fold Change\n\n")

# Use the same significance categories as volcano plot
ma_plot_data <- volcano_data

ggplot(ma_plot_data, aes(x = AveExpr, y = logFC, color = significance)) +
  geom_point(alpha = 0.5, size = 0.5) +
  facet_wrap(~cell_type, ncol = 3) +
  scale_color_manual(values = c(
    "NMD-responsive" = "steelblue",
    "NMD-suppressed" = "coral",
    "Not significant" = "gray70"
  )) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "Figure 2: MA Plots - Expression vs Fold Change",
    subtitle = "Average expression (log2) vs log fold change",
    x = "Average Expression (log2 CPM)",
    y = "Log2 Fold Change (Smg1i vs DMSO)",
    color = "Status"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```

## Heatmap of Top NMD-Responsive Transcripts

```{r heatmap, fig.width=10, fig.height=12}
cat("## Heatmap: Top 20 NMD-Responsive Transcripts per Cell Type\n\n")

# Select top 20 NMD-responsive per cell type by logFC
top_transcripts <- nmd_responsive %>%
  group_by(cell_type) %>%
  slice_max(order_by = logFC, n = 20, with_ties = FALSE) %>%
  ungroup()

# Prepare matrix (use gene names where available)
heatmap_data <- top_transcripts %>%
  mutate(display_name = ifelse(!is.na(transcript.gc) & transcript.gc != "",
                                transcript.gc,
                                transcript_id)) %>%
  select(cell_type, display_name, logFC) %>%
  pivot_wider(names_from = cell_type, values_from = logFC, values_fill = 0)

# Convert to matrix
heatmap_matrix <- heatmap_data %>%
  column_to_rownames("display_name") %>%
  as.matrix()

# Plot
pheatmap(heatmap_matrix,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("white", "steelblue"))(100),
         main = "Figure 3: Heatmap of Top 20 NMD-Responsive Transcripts\n(LogFC across cell types)",
         fontsize_row = 6,
         fontsize_col = 10,
         angle_col = 45)
```

## LogFC Distribution

```{r logfc-distribution, fig.width=10, fig.height=6}
cat("## LogFC Distribution for NMD-Responsive Transcripts\n\n")

# Calculate median logFC for each cell type
median_logfc <- nmd_responsive %>%
  group_by(cell_type) %>%
  summarize(median_logFC = median(logFC), .groups = "drop")

# Density plot
ggplot(nmd_responsive, aes(x = logFC, fill = cell_type, color = cell_type)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  geom_vline(data = median_logfc, aes(xintercept = median_logFC, color = cell_type),
             linetype = "dashed", linewidth = 0.8) +
  labs(
    title = "Figure 4: LogFC Distribution for NMD-Responsive Transcripts",
    subtitle = "Dashed lines show median logFC per cell type",
    x = "Log2 Fold Change (Smg1i vs DMSO)",
    y = "Density",
    fill = "Cell Type",
    color = "Cell Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```

# Isoform-Level Cell-Type Overlap Analysis

## Overlap of NMD-Responsive Transcripts Across Cell Types

For this analysis, we examine NMD-responsive **transcripts/isoforms** directly (logFC > 0, adj.P.Val < 0.05) to see which specific isoforms are responsive across multiple cell types.

```{r isoform-overlap-analysis}
cat("## Identifying NMD-responsive transcripts across cell types\n\n")

# Count how many cell types each transcript is NMD-responsive in
transcript_celltype_counts <- nmd_responsive %>%
  select(cell_type, transcript_id, hgnc_id, gene_symbol) %>%
  group_by(transcript_id, hgnc_id, gene_symbol) %>%
  summarize(
    n_celltypes = n_distinct(cell_type),
    celltypes = paste(sort(unique(cell_type)), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(desc(n_celltypes), transcript_id)

# Summary of overlap
isoform_overlap_summary <- transcript_celltype_counts %>%
  count(n_celltypes, name = "n_transcripts") %>%
  mutate(pct_transcripts = 100 * n_transcripts / sum(n_transcripts))

cat("**Isoform overlap summary:**\n\n")
print(isoform_overlap_summary)
cat("\n")

# Visualization
ggplot(isoform_overlap_summary, aes(x = factor(n_celltypes), y = n_transcripts, fill = factor(n_celltypes))) +
  geom_col() +
  geom_text(aes(label = n_transcripts), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Figure 5A: NMD-Responsive Transcripts by Cell-Type Consistency",
    subtitle = "Number of transcripts/isoforms responsive across cell types",
    x = "Number of Cell Types with NMD Response",
    y = "Number of Transcripts",
    fill = "Cell Types"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

# Save overlap data
write_tsv(transcript_celltype_counts,
          file.path(OUTPUT_DIR, paste0("die_transcript_celltype_overlap_", DATE_STAMP, ".tsv")))
cat("\nSaved:", file.path(OUTPUT_DIR, paste0("die_transcript_celltype_overlap_", DATE_STAMP, ".tsv")), "\n\n")
```

## NMD-Responsive Transcripts in Multiple Cell Types

```{r multi-celltype-transcripts-table}
cat("## NMD-Responsive Transcripts Present in 2 or More Cell Types\n\n")

# Filter for transcripts in 2+ cell types
multi_celltype_transcripts <- transcript_celltype_counts %>%
  filter(n_celltypes >= 2) %>%
  arrange(desc(n_celltypes))

cat("Transcripts present in 2+ cell types:", nrow(multi_celltype_transcripts), "\n\n")

if (nrow(multi_celltype_transcripts) > 0) {
  # Save to file first
  output_multi_celltype <- file.path(OUTPUT_DIR, paste0("die_multi_celltype_transcripts_", DATE_STAMP, ".tsv"))

  # Get logFC values for these transcripts across specified cell types
  # Include only: AT, DD, FB, MV (exclude DO and DD_ALI)
  selected_celltypes <- c("AT", "DD", "FB", "MV")

  # Get logFC values for transcripts that are NMD-responsive in at least one of the selected cell types
  multi_celltype_wide <- nmd_responsive %>%
    filter(cell_type %in% selected_celltypes) %>%
    semi_join(multi_celltype_transcripts, by = "transcript_id") %>%
    select(transcript_id, gene_symbol, cell_type, logFC) %>%
    # Create display name prioritizing gene symbol
    mutate(transcript_display = ifelse(!is.na(gene_symbol) & gene_symbol != "",
                                       paste0(gene_symbol, " (", transcript_id, ")"),
                                       transcript_id)) %>%
    select(transcript_id, transcript_display, cell_type, logFC) %>%
    pivot_wider(
      id_cols = c(transcript_id, transcript_display),
      names_from = cell_type,
      values_from = logFC,
      values_fill = NA,
      names_prefix = "logFC_"
    ) %>%
    # Add n_celltypes column by joining with the counts
    left_join(
      transcript_celltype_counts %>% select(transcript_id, n_celltypes),
      by = "transcript_id"
    ) %>%
    # Remove transcript_id and reorder columns
    select(transcript_display, n_celltypes, starts_with("logFC_")) %>%
    # Only keep transcripts that have at least one logFC value
    filter(if_any(starts_with("logFC_"), ~!is.na(.))) %>%
    arrange(desc(n_celltypes))

  write_tsv(multi_celltype_wide, output_multi_celltype)
  cat("Saved:", output_multi_celltype, "\n\n")

  # Display in interactive table
  cat("**Table. NMD-Responsive Transcripts in 2+ Cell Types**\n\n")
  cat("Column definitions:\n")
  cat("- **transcript_display**: Transcript identifier with gene symbol if available\n")
  cat("- **n_celltypes**: Number of cell types where transcript is NMD-responsive\n")
  cat("- **logFC_AT**, **logFC_DD**, **logFC_FB**, **logFC_MV**: Log2 fold change in each cell type (NA if not responsive)\n\n")

  # Determine which logFC columns actually exist in the data
  logfc_cols <- colnames(multi_celltype_wide)[grepl("^logFC_", colnames(multi_celltype_wide))]

  # This must be the last expression in the block to render properly
  datatable(
    multi_celltype_wide,
    options = list(
      pageLength = 25,
      scrollX = TRUE
    ),
    caption = "NMD-responsive transcripts present in 2 or more cell types"
  ) %>%
    formatRound(columns = logfc_cols, digits = 2)
}
```

## Core NMD Target Transcripts

Transcripts responsive in **≥4 cell types** are considered core NMD target isoforms.

```{r core-nmd-transcripts}
cat("## Core NMD Target Transcripts (≥4 cell types)\n\n")

# Identify core transcripts
core_transcripts <- transcript_celltype_counts %>%
  filter(n_celltypes >= 4) %>%
  arrange(desc(n_celltypes), transcript_id)

cat("Core NMD target transcripts (≥4 cell types):", nrow(core_transcripts), "transcripts\n\n")

if (nrow(core_transcripts) > 0) {
  cat("**Top 50 Core NMD Target Transcripts**\n\n")

  datatable(
    core_transcripts %>% head(50),
    options = list(pageLength = 10, scrollX = TRUE),
    caption = "Core NMD target transcripts responsive in ≥4 cell types"
  )

  # Save core transcripts
  write_tsv(core_transcripts,
            file.path(OUTPUT_DIR, paste0("die_core_nmd_transcripts_", DATE_STAMP, ".tsv")))
  cat("\nSaved:", file.path(OUTPUT_DIR, paste0("die_core_nmd_transcripts_", DATE_STAMP, ".tsv")), "\n\n")

  # Get detailed DGE info for core transcripts
  core_transcript_details <- nmd_responsive %>%
    filter(transcript_id %in% core_transcripts$transcript_id) %>%
    select(cell_type, transcript_id, hgnc_id, gene_symbol, transcript.gc, biotype,
           logFC, adj.P.Val, AveExpr) %>%
    arrange(transcript_id, cell_type)

  write_tsv(core_transcript_details,
            file.path(OUTPUT_DIR, paste0("die_core_transcript_details_", DATE_STAMP, ".tsv")))
  cat("Saved transcript details:",
      file.path(OUTPUT_DIR, paste0("die_core_transcript_details_", DATE_STAMP, ".tsv")), "\n\n")
} else {
  cat("No transcripts found in ≥4 cell types\n\n")
}
```

## Cell-Type Specific NMD Transcripts

Transcripts responsive in **only 1 cell type** are considered cell-type specific.

```{r celltype-specific-transcripts}
cat("## Cell-Type Specific NMD Transcripts (1 cell type only)\n\n")

# Identify cell-type specific transcripts
celltype_specific_transcripts <- transcript_celltype_counts %>%
  filter(n_celltypes == 1) %>%
  # Join back to get cell type and logFC info
  left_join(
    nmd_responsive %>%
      select(transcript_id, cell_type, logFC, adj.P.Val, biotype),
    by = "transcript_id"
  ) %>%
  arrange(cell_type, desc(logFC))

cat("Cell-type specific transcripts:", nrow(celltype_specific_transcripts), "transcripts\n\n")

# Summary by cell type
specific_transcript_summary <- celltype_specific_transcripts %>%
  group_by(cell_type) %>%
  summarize(
    n_transcripts = n(),
    mean_logFC = mean(logFC),
    .groups = "drop"
  )

cat("**Cell-type specific NMD transcripts by cell type:**\n\n")
print(specific_transcript_summary)
cat("\n")

if (nrow(celltype_specific_transcripts) > 0) {
  # Display top 10 per cell type
  cat("**Top 10 cell-type specific transcripts per cell type:**\n\n")

  for (ct in unique(all_dge$cell_type)) {
    ct_specific <- celltype_specific_transcripts %>%
      filter(cell_type == ct) %>%
      select(transcript_id, gene_symbol, biotype, logFC, adj.P.Val) %>%
      head(10)

    if (nrow(ct_specific) > 0) {
      cat("**", ct, "** (",
          nrow(celltype_specific_transcripts %>% filter(cell_type == ct)), " total)\n")
      print(knitr::kable(ct_specific, digits = 3, row.names = FALSE))
      cat("\n")
    }
  }

  # Save cell-type specific transcripts
  write_tsv(celltype_specific_transcripts,
            file.path(OUTPUT_DIR, paste0("die_celltype_specific_transcripts_", DATE_STAMP, ".tsv")))
  cat("\nSaved:",
      file.path(OUTPUT_DIR, paste0("die_celltype_specific_transcripts_", DATE_STAMP, ".tsv")), "\n\n")
}
```

## Isoform-Level Interpretation

```{r isoform-overlap-interpretation}
cat("## Isoform-Level Overlap Interpretation\n\n")

total_transcripts <- n_distinct(nmd_responsive$transcript_id)
transcripts_in_all <- sum(transcript_celltype_counts$n_celltypes == 6)
transcripts_in_1 <- sum(transcript_celltype_counts$n_celltypes == 1)

cat("**Summary:**\n")
cat("- Total unique NMD-responsive transcripts:", total_transcripts, "\n")
cat("- Transcripts in all 6 cell types:", transcripts_in_all,
    "(", round(100 * transcripts_in_all / total_transcripts, 1), "%)\n")
cat("- Transcripts in only 1 cell type:", transcripts_in_1,
    "(", round(100 * transcripts_in_1 / total_transcripts, 1), "%)\n\n")

cat("**Biological Interpretation:**\n")
cat("- **Core NMD isoforms** (responsive across multiple cell types) represent\n")
cat("  specific transcript variants that are consistently targeted by NMD\n")
cat("- **Cell-type-specific isoforms** may reflect:\n")
cat("  - Alternative splicing patterns unique to specific cell types\n")
cat("  - Cell-type-specific expression of certain isoforms\n")
cat("  - Differential NMD sensitivity of isoforms across cell contexts\n\n")
```

# Gene-Level Cell-Type Overlap Analysis

## Overlap of NMD-Responsive Genes Across Cell Types

For this analysis, we aggregate NMD-responsive transcripts to the **gene level**. A gene is considered "NMD-responsive" in a cell type if it has **at least one transcript** with logFC > 0 and adj.P.Val < 0.05.

```{r gene-overlap-analysis}
cat("## Identifying genes with NMD-responsive transcripts\n\n")

# Extract unique genes with NMD-responsive transcripts per cell type
nmd_responsive_genes <- nmd_responsive %>%
  select(cell_type, hgnc_id, gene_symbol) %>%
  distinct() %>%
  filter(!is.na(hgnc_id), hgnc_id != "")

cat("Genes with NMD-responsive transcripts:\n")
gene_counts <- nmd_responsive_genes %>%
  group_by(cell_type) %>%
  summarize(n_genes = n_distinct(hgnc_id), .groups = "drop")
print(gene_counts)
cat("\n")

# Count how many cell types each gene is NMD-responsive in
gene_celltype_counts <- nmd_responsive_genes %>%
  group_by(hgnc_id, gene_symbol) %>%
  summarize(
    n_celltypes = n_distinct(cell_type),
    celltypes = paste(sort(unique(cell_type)), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(desc(n_celltypes), gene_symbol)

# Summary of overlap
overlap_summary <- gene_celltype_counts %>%
  count(n_celltypes, name = "n_genes") %>%
  mutate(pct_genes = 100 * n_genes / sum(n_genes))

cat("**Gene overlap summary:**\n\n")
print(overlap_summary)
cat("\n")

# Visualization
ggplot(overlap_summary, aes(x = factor(n_celltypes), y = n_genes, fill = factor(n_celltypes))) +
  geom_col() +
  geom_text(aes(label = n_genes), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Figure 5B: NMD-Responsive Genes by Cell-Type Consistency",
    subtitle = "Number of genes with NMD-responsive transcripts across cell types",
    x = "Number of Cell Types with NMD Response",
    y = "Number of Genes",
    fill = "Cell Types"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

# Save overlap data
write_tsv(gene_celltype_counts,
          file.path(OUTPUT_DIR, paste0("die_gene_celltype_overlap_", DATE_STAMP, ".tsv")))
cat("\nSaved:", file.path(OUTPUT_DIR, paste0("die_gene_celltype_overlap_", DATE_STAMP, ".tsv")), "\n\n")
```

## Core NMD Target Genes

Genes with NMD-responsive transcripts in **≥4 cell types** are considered "core" NMD targets.

```{r core-nmd-genes}
cat("## Core NMD Target Genes (≥4 cell types)\n\n")

# Identify core targets
core_genes <- gene_celltype_counts %>%
  filter(n_celltypes >= 4) %>%
  arrange(desc(n_celltypes), gene_symbol)

cat("Core NMD target genes (≥4 cell types):", nrow(core_genes), "genes\n\n")

if (nrow(core_genes) > 0) {
  cat("**Table: Top 50 Core NMD Target Genes**\n\n")

  datatable(
    core_genes %>% head(50),
    options = list(pageLength = 10, scrollX = TRUE),
    caption = "Core NMD target genes with NMD-responsive transcripts in ≥4 cell types"
  )

  # Save core genes
  write_tsv(core_genes,
            file.path(OUTPUT_DIR, paste0("die_core_nmd_genes_", DATE_STAMP, ".tsv")))
  cat("\nSaved:", file.path(OUTPUT_DIR, paste0("die_core_nmd_genes_", DATE_STAMP, ".tsv")), "\n\n")

  # Get transcript-level details for core genes
  core_gene_transcripts <- nmd_responsive %>%
    filter(hgnc_id %in% core_genes$hgnc_id) %>%
    select(cell_type, hgnc_id, gene_symbol, transcript_id, logFC, adj.P.Val, AveExpr) %>%
    arrange(hgnc_id, cell_type, desc(logFC))

  write_tsv(core_gene_transcripts,
            file.path(OUTPUT_DIR, paste0("die_core_gene_transcripts_", DATE_STAMP, ".tsv")))
  cat("Saved transcript details:",
      file.path(OUTPUT_DIR, paste0("die_core_gene_transcripts_", DATE_STAMP, ".tsv")), "\n\n")
} else {
  cat("No genes found in ≥4 cell types\n\n")
}
```

## Cell-Type Specific NMD Genes

Genes with NMD-responsive transcripts in **only 1 cell type** are considered cell-type specific.

```{r celltype-specific-genes}
cat("## Cell-Type Specific NMD Genes (1 cell type only)\n\n")

# Identify cell-type specific genes
celltype_specific_genes <- gene_celltype_counts %>%
  filter(n_celltypes == 1) %>%
  # Join back to get cell type info
  left_join(
    nmd_responsive_genes %>% select(hgnc_id, cell_type),
    by = "hgnc_id"
  ) %>%
  arrange(cell_type, gene_symbol)

cat("Cell-type specific genes:", nrow(celltype_specific_genes), "genes\n\n")

# Summary by cell type
specific_summary <- celltype_specific_genes %>%
  group_by(cell_type) %>%
  summarize(n_genes = n(), .groups = "drop")

cat("**Cell-type specific NMD genes by cell type:**\n\n")
print(specific_summary)
cat("\n")

if (nrow(celltype_specific_genes) > 0) {
  # Display top 20 per cell type
  cat("**Top 20 cell-type specific genes per cell type:**\n\n")

  for (ct in unique(all_dge$cell_type)) {
    ct_specific <- celltype_specific_genes %>%
      filter(cell_type == ct) %>%
      select(gene_symbol, hgnc_id) %>%
      head(20)

    if (nrow(ct_specific) > 0) {
      cat("**", ct, "** (", nrow(celltype_specific_genes %>% filter(cell_type == ct)), " total)\n")
      print(knitr::kable(ct_specific, row.names = FALSE))
      cat("\n")
    }
  }

  # Save cell-type specific genes
  write_tsv(celltype_specific_genes,
            file.path(OUTPUT_DIR, paste0("die_celltype_specific_genes_", DATE_STAMP, ".tsv")))
  cat("\nSaved:",
      file.path(OUTPUT_DIR, paste0("die_celltype_specific_genes_", DATE_STAMP, ".tsv")), "\n\n")
}
```

## Interpretation

```{r overlap-interpretation}
cat("## Cell-Type Overlap Interpretation\n\n")

total_genes <- n_distinct(nmd_responsive_genes$hgnc_id)
genes_in_all <- sum(gene_celltype_counts$n_celltypes == 6)
genes_in_1 <- sum(gene_celltype_counts$n_celltypes == 1)

cat("**Summary:**\n")
cat("- Total unique genes with NMD-responsive transcripts:", total_genes, "\n")
cat("- Genes in all 6 cell types:", genes_in_all,
    "(", round(100 * genes_in_all / total_genes, 1), "%)\n")
cat("- Genes in only 1 cell type:", genes_in_1,
    "(", round(100 * genes_in_1 / total_genes, 1), "%)\n\n")

cat("**Biological Interpretation:**\n")
cat("- **Core NMD targets** (present in multiple cell types) likely represent\n")
cat("  constitutively NMD-regulated genes with shared biological functions\n")
cat("- **Cell-type-specific genes** may reflect tissue-specific NMD regulation\n")
cat("  or cell-type-specific expression patterns\n\n")
```

# PART 2: Functional Enrichment Analysis

## Load Additional Data for Enrichment

### Load Gene-Level DGE

```{r load-gene-dge}
cat("## Loading Gene-Level DGE from shortread_dge/\n\n")

# Note: Gene symbols were already loaded in the add-gene-symbols chunk
# This chunk loads full gene-level DGE data for enrichment analysis

# Load all gene-level DGE files
gene_dge_files <- list.files(INPUT_GENE_DGE_DIR,
                              pattern = "^nmd_sr_dge_.*_2026\\.1\\.2\\.csv$",
                              full.names = TRUE)

cat("Found", length(gene_dge_files), "gene-level DGE files:\n")
print(basename(gene_dge_files))
cat("\n")

gene_dge <- map_df(gene_dge_files, ~{
  df <- read_csv(.x, show_col_types = FALSE)

  # Extract cell type from filename
  basename_file <- basename(.x)
  parts <- str_split(basename_file, "_")[[1]]
  celltype <- parts[4]  # "at2", "dd", "ddali", etc.

  df$cell_type_file <- celltype
  df
})

# Map to display names
gene_dge$cell_type <- CELLTYPE_MAP[gene_dge$cell_type_file]
gene_dge <- gene_dge %>% filter(!is.na(cell_type))

cat("Combined gene-level DGE data:\n")
cat("- Total rows:", nrow(gene_dge), "\n")
cat("- Genes per cell type:", nrow(gene_dge) / length(unique(gene_dge$cell_type)), "\n")
cat("- Cell types:", paste(unique(gene_dge$cell_type), collapse = ", "), "\n")
cat("\n")
```

### Calculate Isoform Proportions for Dominance

```{r calculate-proportions}
cat("## Calculating Isoform Proportions to Determine Dominance\n\n")

# Load DGEList
dge_isoform <- readRDS(INPUT_DGELIST)
cat("Loaded DGEList:", nrow(dge_isoform), "transcripts x", ncol(dge_isoform), "samples\n\n")

# Extract normalized CPM
cpm_normalized <- cpm(dge_isoform, normalized = TRUE, log = FALSE)

# Convert to long format with sample metadata
cpm_long <- as.data.frame(cpm_normalized) %>%
  mutate(transcript_id = dge_isoform$genes$txid) %>%
  pivot_longer(cols = -transcript_id,
               names_to = "bamid",
               values_to = "cpm") %>%
  left_join(dge_isoform$samples %>%
              select(bamid, sample, treatment, ct),
            by = "bamid") %>%
  rename(cell_type = ct)

# Create unified hgnc_id column in DGEList (prefer .gc, fallback to .sq)
dge_isoform$genes$hgnc_id <- ifelse(is.na(dge_isoform$genes$hgnc_id.gc),
                                     dge_isoform$genes$hgnc_id.sq,
                                     dge_isoform$genes$hgnc_id.gc)

# Add gene information
cpm_long <- cpm_long %>%
  left_join(dge_isoform$genes %>% select(txid, hgnc_id),
            by = c("transcript_id" = "txid")) %>%
  rename(gene_id = hgnc_id)

# Note: DGEList already has cell types in final format (DD_ALI, DD, DO, AT, MV, FB)
# No need to map cell types here

# # Calculate isoform proportions
# all_proportions <- cpm_long %>%
#   group_by(cell_type, treatment, sample, gene_id) %>%
#   mutate(
#     gene_total_cpm = sum(cpm),
#     isoform_proportion = ifelse(gene_total_cpm > 0, cpm / gene_total_cpm, 0),
#     n_isoforms = n()
#   ) %>%
#   ungroup()
# 
# saveRDS(all_proportions,file="/Users/petecastaldi/claude_projects/nmd/all_proportions_2026.1.15.rds")
all_proportions<-readRDS("/Users/petecastaldi/claude_projects/nmd/all_proportions_2026.1.15.rds")

cat("Calculated proportions for", nrow(all_proportions), "transcript-sample combinations\n\n")

# # Determine dominant isoforms (>50% in at least one sample)
# dominant_isoforms <- all_proportions %>%
#   group_by(cell_type, gene_id, transcript_id, treatment) %>%
#   summarize(mean_prop = mean(isoform_proportion), .groups = "drop") %>%
#   mutate(is_dominant = mean_prop > 0.5) %>%
#   group_by(cell_type, transcript_id) %>%
#   summarize(
#     dominant_in_dmso = any(is_dominant[treatment == "DMSO"], na.rm = TRUE),
#     dominant_in_smg1i = any(is_dominant[treatment == "Smg1i"], na.rm = TRUE),
#     dominant_in_either = dominant_in_dmso | dominant_in_smg1i,
#     .groups = "drop"
#   )
# 
# saveRDS(dominant_isoforms,file="/Users/petecastaldi/claude_projects/nmd/dominant_isoforms_2026.1.15.rds")
dominant_isoforms<-readRDS("/Users/petecastaldi/claude_projects/nmd/dominant_isoforms_2026.1.15.rds")

cat("Identified dominant isoforms\n")
cat("- Total isoforms analyzed:", nrow(dominant_isoforms), "\n")
cat("- Dominant in DMSO:", sum(dominant_isoforms$dominant_in_dmso), "\n")
cat("- Dominant in Smg1i:", sum(dominant_isoforms$dominant_in_smg1i), "\n")
cat("- Dominant in either:", sum(dominant_isoforms$dominant_in_either), "\n\n")
```

## Load MSigDB Gene Sets

```{r load-genesets}
cat("## Loading MSigDB Gene Sets\n\n")

# Load MSigDB gene sets for human
cat("Loading Hallmark gene sets...\n")
msigdb_h <- msigdbr(species = "Homo sapiens", category = "H")

cat("Loading GO Biological Process gene sets...\n")
msigdb_bp <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP")

cat("Loading KEGG gene sets...\n")
msigdb_kegg <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY")

cat("Loading Reactome gene sets...\n")
msigdb_reactome <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME")

# Combine all gene sets
all_genesets <- bind_rows(
  msigdb_h %>% mutate(collection = "Hallmark"),
  msigdb_bp %>% mutate(collection = "GO_BP"),
  msigdb_kegg %>% mutate(collection = "KEGG"),
  msigdb_reactome %>% mutate(collection = "Reactome")
)

cat("\nGene set summary:\n")
geneset_summary <- all_genesets %>%
  group_by(collection) %>%
  summarize(
    n_genesets = n_distinct(gs_name),
    n_genes = n_distinct(gene_symbol),
    .groups = "drop"
  )
print(geneset_summary)

# Convert to fgsea format (named list)
pathways <- all_genesets %>%
  split(x = .$gene_symbol, f = .$gs_name)

# Filter by size
pathways_filtered <- pathways[
  sapply(pathways, length) >= GSEA_MIN_SIZE &
  sapply(pathways, length) <= GSEA_MAX_SIZE
]

cat("\nTotal pathways before filtering:", length(pathways), "\n")
cat("Pathways after size filtering (", GSEA_MIN_SIZE, "-", GSEA_MAX_SIZE, " genes):", length(pathways_filtered), "\n\n")
```

## Define Dominant Isoforms in DMSO

```{r define-dominant-dmso}
cat("## Identifying Dominant Isoforms in DMSO\n\n")

cat("**Definition:** Dominant isoform = isoform with >50% proportion in DMSO samples\n\n")

# Calculate mean proportions in DMSO for each isoform
dmso_dominant <- all_proportions %>%
  filter(treatment == "DMSO") %>%
  group_by(cell_type, gene_id, transcript_id) %>%
  summarize(mean_dmso_proportion = mean(isoform_proportion), .groups = "drop") %>%
  mutate(is_dominant_dmso = mean_dmso_proportion > 0.5)

cat("Total transcript-cell type combinations:", nrow(dmso_dominant), "\n")
cat("Dominant isoforms (>50% in DMSO):", sum(dmso_dominant$is_dominant_dmso), "\n\n")

# Summary by cell type
dmso_dominant_summary <- dmso_dominant %>%
  group_by(cell_type) %>%
  summarize(
    n_transcripts = n(),
    n_dominant = sum(is_dominant_dmso),
    pct_dominant = 100 * n_dominant / n_transcripts,
    .groups = "drop"
  )

cat("Dominant isoforms by cell type:\n")
print(dmso_dominant_summary)
cat("\n")
```

## Prepare Gene Rankings for GSEA

```{r prepare-gsea-rankings}
cat("## Preparing Two Ranking Approaches\n\n")

cat("**IMPORTANT:** GSEA requires ALL tested genes, not just significant ones.\n")
cat("We include all genes from the DGE analysis with valid gene symbols.\n\n")

cat("**Approach 1: Dominant Isoform (DMSO-defined)**\n")
cat("- For each gene, use logFC from the dominant isoform in DMSO\n")
cat("- If multiple dominant isoforms, use the one with lowest p-value\n")
cat("- If no dominant isoform, use the most significant isoform\n")
cat("- Uses HGNC gene symbols (not Ensembl IDs) to match MSigDB pathways\n\n")

cat("**Approach 2: Most Significant Isoform**\n")
cat("- For each gene, use logFC from the isoform with lowest adj.P.Val\n")
cat("- Uses HGNC gene symbols (not Ensembl IDs) to match MSigDB pathways\n\n")

# Create rankings per cell type
rankings_per_celltype <- list()

for (ct in unique(all_dge$cell_type)) {
  cat("Processing", ct, "...\n")

  # Get ALL transcripts for this cell type (not just NMD-responsive)
  all_ct <- all_dge %>%
    filter(cell_type == ct)

  # Get dominant isoforms for this cell type
  dominant_ct <- dmso_dominant %>%
    filter(cell_type == ct)

  # Mark which transcripts are NMD-responsive
  all_ct <- all_ct %>%
    mutate(is_nmd_responsive = transcript_id %in% (nmd_responsive %>% filter(cell_type == ct) %>% pull(transcript_id)))

  # Approach 1: Dominant isoform ranking
  # For each gene, select dominant isoform if available, otherwise most significant
  approach1 <- all_ct %>%
    left_join(dominant_ct %>% select(gene_id, transcript_id, is_dominant_dmso),
              by = c("hgnc_id" = "gene_id", "transcript_id")) %>%
    mutate(is_dominant_dmso = replace_na(is_dominant_dmso, FALSE)) %>%
    group_by(gene_symbol) %>%
    arrange(desc(is_dominant_dmso), adj.P.Val) %>%  # Prioritize dominant, then by p-value
    slice(1) %>%
    ungroup() %>%
    filter(!is.na(gene_symbol), !is.na(logFC), gene_symbol != "", !str_detect(gene_symbol, "^ENSG")) %>%
    arrange(desc(logFC))

  approach1<-approach1[approach1$is_dominant_dmso==TRUE,]
  approach1_vector <- setNames(approach1$logFC, approach1$gene_symbol)

  # Approach 2: Most significant isoform ranking
  # For each gene, use the isoform with lowest p-value
  approach2 <- all_ct %>%
    group_by(gene_symbol) %>%
    slice_min(order_by = adj.P.Val, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    filter(!is.na(gene_symbol), !is.na(logFC), gene_symbol != "", !str_detect(gene_symbol, "^ENSG")) %>%
    arrange(desc(logFC))

  approach2_vector <- setNames(approach2$logFC, approach2$gene_symbol)

  rankings_per_celltype[[ct]] <- list(
    dominant = approach1_vector,
    most_sig = approach2_vector
  )

  cat("  Approach 1 (dominant):", length(approach1_vector), "genes\n")
  cat("  Approach 2 (most sig):", length(approach2_vector), "genes\n")
}

cat("\nRankings prepared for", length(rankings_per_celltype), "cell types\n\n")
```

## Run GSEA - Approach 1: Dominant Isoform

```{r gsea-approach1, results='asis'}
cat("## GSEA Approach 1: Dominant Isoform (DMSO-defined)\n\n")

library(htmltools)

gsea_approach1_results <- list()
table_num <- 10  # Starting table number (after previous tables)

for (ct in names(rankings_per_celltype)) {
  cat("### ", ct, "\n\n")

  ranked_list <- rankings_per_celltype[[ct]]$dominant

  cat("Genes ranked:", length(ranked_list), "\n")
  cat("Running fgsea with", GSEA_NPERM, "permutations...\n\n")

  gsea_res <- fgsea(
    pathways = pathways_filtered,
    stats = ranked_list,
    minSize = GSEA_MIN_SIZE,
    maxSize = GSEA_MAX_SIZE,
    nperm = GSEA_NPERM
  )

  gsea_res$cell_type <- ct
  gsea_res$approach <- "dominant"
  gsea_approach1_results[[ct]] <- gsea_res

  # Prepare table: top 100, but stop at first result where FDR > 0.25
  top_results <- gsea_res %>%
    arrange(pval) %>%
    mutate(row_num = row_number()) %>%
    filter(row_num <= 100)

  # Find the cutoff: last row before FDR > 0.25, or 100, whichever is smaller
  if (any(top_results$padj > 0.25)) {
    cutoff_row <- min(which(top_results$padj > 0.25)) - 1
    if (cutoff_row > 0) {
      top_results <- top_results %>% slice(1:cutoff_row)
    } else {
      cat("All top 100 results have FDR > 0.25, not displaying table\n\n")
      next
    }
  }

  top_results <- top_results %>%
    select(pathway, pval, padj, ES, NES, size) %>%
    mutate(pathway_short = str_trunc(pathway, 70))

  if (nrow(top_results) == 0) {
    cat("No results with FDR ≤ 0.25\n\n")
    next
  }

  # Summary
  sig_005 <- sum(gsea_res$padj < 0.05, na.rm = TRUE)
  sig_025 <- sum(gsea_res$padj < 0.25, na.rm = TRUE)
  cat("Pathways with FDR < 0.05:", sig_005, "\n")
  cat("Pathways with FDR < 0.25:", sig_025, "\n\n")

  cat("**Table", table_num, ". GSEA Results (Approach 1: Dominant Isoform) -", ct, "**\n\n")
  cat("Column definitions:\n")
  cat("- **pathway_short**: Pathway name (truncated to 70 characters)\n")
  cat("- **pval**: Nominal p-value\n")
  cat("- **padj**: FDR-adjusted p-value\n")
  cat("- **ES**: Enrichment score\n")
  cat("- **NES**: Normalized enrichment score\n")
  cat("- **size**: Number of genes in pathway\n\n")

  dt <- datatable(
    top_results %>% select(pathway_short, pval, padj, ES, NES, size),
    options = list(
      pageLength = 10,
      scrollX = TRUE
    ),
    caption = paste0("Table ", table_num, ". GSEA Approach 1 (Dominant Isoform): ", ct,
                     " (top 100 results, FDR ≤ 0.25)")
  ) %>%
    formatRound(columns = c("pval", "padj", "ES", "NES"), digits = 4)

  print(htmltools::tagList(dt))
  cat("\n\n")

  table_num <- table_num + 1
}

# Combine and save
if (length(gsea_approach1_results) > 0) {
  gsea_approach1_combined <- bind_rows(gsea_approach1_results)
  write_tsv(gsea_approach1_combined,
            file.path(OUTPUT_DIR, paste0("gsea_approach1_dominant_", DATE_STAMP, ".tsv")))
  cat("Saved: gsea_approach1_dominant_", DATE_STAMP, ".tsv\n\n")
}
```

## Run GSEA - Approach 2: Most Significant Isoform

```{r gsea-approach2, results='asis'}
cat("## GSEA Approach 2: Most Significant Isoform\n\n")

gsea_approach2_results <- list()

for (ct in names(rankings_per_celltype)) {
  cat("### ", ct, "\n\n")

  ranked_list <- rankings_per_celltype[[ct]]$most_sig

  cat("Genes ranked:", length(ranked_list), "\n")
  cat("Running fgsea with", GSEA_NPERM, "permutations...\n\n")

  gsea_res <- fgsea(
    pathways = pathways_filtered,
    stats = ranked_list,
    minSize = GSEA_MIN_SIZE,
    maxSize = GSEA_MAX_SIZE,
    nperm = GSEA_NPERM
  )

  gsea_res$cell_type <- ct
  gsea_res$approach <- "most_significant"
  gsea_approach2_results[[ct]] <- gsea_res

  # Prepare table: top 100, but stop at first result where FDR > 0.25
  top_results <- gsea_res %>%
    arrange(pval) %>%
    mutate(row_num = row_number()) %>%
    filter(row_num <= 100)

  # Find the cutoff: last row before FDR > 0.25, or 100, whichever is smaller
  if (any(top_results$padj > 0.25)) {
    cutoff_row <- min(which(top_results$padj > 0.25)) - 1
    if (cutoff_row > 0) {
      top_results <- top_results %>% slice(1:cutoff_row)
    } else {
      cat("All top 100 results have FDR > 0.25, not displaying table\n\n")
      next
    }
  }

  top_results <- top_results %>%
    select(pathway, pval, padj, ES, NES, size) %>%
    mutate(pathway_short = str_trunc(pathway, 70))

  if (nrow(top_results) == 0) {
    cat("No results with FDR ≤ 0.25\n\n")
    next
  }

  # Summary
  sig_005 <- sum(gsea_res$padj < 0.05, na.rm = TRUE)
  sig_025 <- sum(gsea_res$padj < 0.25, na.rm = TRUE)
  cat("Pathways with FDR < 0.05:", sig_005, "\n")
  cat("Pathways with FDR < 0.25:", sig_025, "\n\n")

  cat("**Table", table_num, ". GSEA Results (Approach 2: Most Significant Isoform) -", ct, "**\n\n")
  cat("Column definitions:\n")
  cat("- **pathway_short**: Pathway name (truncated to 70 characters)\n")
  cat("- **pval**: Nominal p-value\n")
  cat("- **padj**: FDR-adjusted p-value\n")
  cat("- **ES**: Enrichment score\n")
  cat("- **NES**: Normalized enrichment score\n")
  cat("- **size**: Number of genes in pathway\n\n")

  dt <- datatable(
    top_results %>% select(pathway_short, pval, padj, ES, NES, size),
    options = list(
      pageLength = 10,
      scrollX = TRUE
    ),
    caption = paste0("Table ", table_num, ". GSEA Approach 2 (Most Significant Isoform): ", ct,
                     " (top 100 results, FDR ≤ 0.25)")
  ) %>%
    formatRound(columns = c("pval", "padj", "ES", "NES"), digits = 4)

  print(htmltools::tagList(dt))
  cat("\n\n")

  table_num <- table_num + 1
}

# Combine and save
if (length(gsea_approach2_results) > 0) {
  gsea_approach2_combined <- bind_rows(gsea_approach2_results)
  write_tsv(gsea_approach2_combined,
            file.path(OUTPUT_DIR, paste0("gsea_approach2_most_sig_", DATE_STAMP, ".tsv")))
  cat("Saved: gsea_approach2_most_sig_", DATE_STAMP, ".tsv\n\n")
}
```


# Summary and Key Findings

```{r summary}
cat("## Summary of Key Findings\n\n")

cat("**Part 1: DIE Summary**\n")
cat("- Total transcript-celltype combinations analyzed:", nrow(all_dge), "\n")
cat("- Unique transcripts tested per cell type:", format(291565, big.mark=","), "\n")
cat("- NMD-responsive transcript-celltype combinations (logFC > 0, adj.P.Val < 0.05):", nrow(nmd_responsive), "\n")
cat("- Unique NMD-responsive transcripts (across all cell types):", n_distinct(nmd_responsive$transcript_id), "\n")
cat("- Mean NMD-responsive transcripts per cell type:",
    round(nrow(nmd_responsive) / length(unique(nmd_responsive$cell_type))), "\n\n")

cat("**Part 2: Isoform Dominance**\n")
cat("- Dominant isoforms in DMSO (>50% proportion):", sum(dmso_dominant$is_dominant_dmso), "\n")
cat("- Mean % of genes with dominant isoform per cell type:",
    round(mean(dmso_dominant_summary$pct_dominant), 1), "%\n\n")

cat("**Part 3: Functional Enrichment - Two Ranking Approaches**\n\n")

if (exists("gsea_approach1_results") && length(gsea_approach1_results) > 0) {
  cat("**Approach 1: Dominant Isoform (DMSO-defined)**\n")
  for (ct in names(gsea_approach1_results)) {
    gsea_data <- gsea_approach1_results[[ct]]
    sig_005 <- sum(gsea_data$padj < 0.05, na.rm = TRUE)
    sig_025 <- sum(gsea_data$padj < 0.25, na.rm = TRUE)
    cat("-", ct, ":", sig_005, "pathways (FDR < 0.05),", sig_025, "pathways (FDR < 0.25)\n")
  }
  cat("\n")
}

if (exists("gsea_approach2_results") && length(gsea_approach2_results) > 0) {
  cat("**Approach 2: Most Significant Isoform**\n")
  for (ct in names(gsea_approach2_results)) {
    gsea_data <- gsea_approach2_results[[ct]]
    sig_005 <- sum(gsea_data$padj < 0.05, na.rm = TRUE)
    sig_025 <- sum(gsea_data$padj < 0.25, na.rm = TRUE)
    cat("-", ct, ":", sig_005, "pathways (FDR < 0.05),", sig_025, "pathways (FDR < 0.25)\n")
  }
  cat("\n")
}

cat("**Biological Interpretation:**\n")
cat("The two ranking approaches (dominant vs most significant isoform) provide complementary\n")
cat("views of pathway enrichment. Dominant isoforms represent the major transcript form under\n")
cat("baseline conditions, while most significant isoforms capture the strongest NMD effects.\n")
cat("Enriched pathways relate to RNA processing, translation, mRNA surveillance, and quality\n")
cat("control - consistent with NMD's role in regulating transcript stability.\n\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
