---
title: "Gene-Level vs Isoform-Level NMD Response Comparison"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

<style>
body {
  font-size: 14px;
}
</style>

# Overview

This report compares NMD-responsive results between:
- **Gene-level analysis** (short-read RNA-seq, gene-level DGE)
- **Isoform-level analysis** (long-read RNA-seq, transcript-level DIE)

## Research Questions

1. **Gene → Isoform concordance**: For NMD-responsive genes, how many have at least one NMD-responsive isoform?
2. **Isoform → Gene concordance**: For NMD-responsive isoforms, how many have their parent gene also NMD-responsive?
3. **Cell-type patterns**: Are concordance rates consistent across cell types?

## Key Terminology

- **NMD-responsive**: logFC > 0 and adj.P.Val < 0.05 (upregulated with SMG1 inhibition)
- **Gene-level DGE**: Short-read RNA-seq differential gene expression
- **Isoform-level DIE**: Long-read RNA-seq differential isoform expression
- **Concordance**: Agreement between gene and isoform-level responses

```{r load-libraries}
library(tidyverse)
library(DT)
library(scales)

# Styling
theme_set(theme_minimal())
```

# Configuration

```{r config}
# Paths
SHORTREAD_DIR <- file.path("..", "shortread_dge")
LONGREAD_DIR <- file.path("..", "longread_dge")
OUTPUT_DIR <- file.path("..", "tmp")

# Thresholds
SIGNIFICANCE_THRESHOLD <- 0.05
LOGFC_THRESHOLD <- 0

# Cell type mapping
CELLTYPE_MAP <- c(
  "at2" = "AT",
  "dd" = "DD",
  "ddali" = "DD_ALI",
  "doali" = "DO",
  "fb" = "FB",
  "mv" = "MV"
)

cat("Configuration loaded\n")
cat("Short-read DGE directory:", SHORTREAD_DIR, "\n")
cat("Long-read DIE directory:", LONGREAD_DIR, "\n")
cat("Significance threshold: adj.P.Val <", SIGNIFICANCE_THRESHOLD, "\n")
cat("LogFC threshold: logFC >", LOGFC_THRESHOLD, "\n\n")
```

# Load Data

## Load Gene-Level DGE Results

```{r load-gene-dge}
cat("## Loading gene-level DGE results (short-read)\n\n")

# Load all gene-level DGE files
gene_files <- list.files(SHORTREAD_DIR,
                         pattern = "^nmd_sr_dge_.*_2026\\.1\\.2\\.csv$",
                         full.names = TRUE)

cat("Found", length(gene_files), "gene-level DGE files\n\n")

gene_dge <- map_dfr(gene_files, function(file) {
  # Extract cell type from filename
  ct_code <- str_extract(basename(file), "(?<=nmd_sr_dge_)[^_]+")
  ct_name <- CELLTYPE_MAP[ct_code]

  # Load data
  read_csv(file, show_col_types = FALSE) %>%
    mutate(cell_type = ct_name)
})

cat("Loaded gene-level DGE data:\n")
cat("- Total genes:", nrow(gene_dge), "\n")
cat("- Cell types:", paste(unique(gene_dge$cell_type), collapse = ", "), "\n\n")

# Identify NMD-responsive genes
nmd_genes <- gene_dge %>%
  filter(logFC > LOGFC_THRESHOLD, adj.P.Val < SIGNIFICANCE_THRESHOLD) %>%
  select(cell_type, hgnc_symbol, ensembl_gene_id_version, logFC, adj.P.Val) %>%
  mutate(gene_id_base = str_replace(ensembl_gene_id_version, "\\.\\d+$", ""))

cat("NMD-responsive genes by cell type:\n")
nmd_genes %>%
  count(cell_type, name = "n_nmd_genes") %>%
  print()
cat("\n")
```

## Load Isoform-Level DIE Results

```{r load-isoform-die}
cat("## Loading isoform-level DIE results (long-read)\n\n")

# Load all isoform-level DIE files
isoform_files <- list.files(LONGREAD_DIR,
                             pattern = "^nmd_dge_.*_2026\\.1\\.2\\.csv$",
                             full.names = TRUE)

cat("Found", length(isoform_files), "isoform-level DIE files\n\n")

isoform_die <- map_dfr(isoform_files, function(file) {
  # Extract cell type from filename
  ct_code <- str_extract(basename(file), "(?<=nmd_dge_)[^_]+")
  ct_name <- CELLTYPE_MAP[ct_code]

  # Load data
  read_csv(file, show_col_types = FALSE) %>%
    mutate(cell_type = ct_name)
})

cat("Loaded isoform-level DIE data:\n")
cat("- Total isoforms:", nrow(isoform_die), "\n")
cat("- Cell types:", paste(unique(isoform_die$cell_type), collapse = ", "), "\n\n")

# Identify NMD-responsive isoforms
nmd_isoforms <- isoform_die %>%
  filter(logFC > LOGFC_THRESHOLD, adj.P.Val < SIGNIFICANCE_THRESHOLD) %>%
  select(cell_type, txid, hgnc_id, logFC, adj.P.Val) %>%
  # Create base gene ID without version
  mutate(gene_id_base = str_replace(hgnc_id, "\\.\\d+$", ""))

cat("NMD-responsive isoforms by cell type:\n")
nmd_isoforms %>%
  count(cell_type, name = "n_nmd_isoforms") %>%
  print()
cat("\n")
```

# Concordance Analysis

## Gene → Isoform Concordance

For each NMD-responsive gene, check if it has at least one NMD-responsive isoform.

```{r gene-to-isoform}
cat("## Gene → Isoform Concordance\n\n")
cat("For each NMD-responsive gene (short-read), determining if it has ≥1 NMD-responsive isoform (long-read)\n\n")

gene_to_isoform <- map_dfr(unique(nmd_genes$cell_type), function(ct) {
  # Get NMD genes for this cell type
  genes_ct <- nmd_genes %>%
    filter(cell_type == ct) %>%
    select(gene_id_base, hgnc_symbol) %>%
    distinct()

  # Get NMD isoforms for this cell type
  isoforms_ct <- nmd_isoforms %>%
    filter(cell_type == ct) %>%
    select(gene_id_base) %>%
    distinct()

  # Check which genes have NMD-responsive isoforms
  genes_with_isoforms <- genes_ct %>%
    mutate(has_nmd_isoform = gene_id_base %in% isoforms_ct$gene_id_base)

  # Summary
  tibble(
    cell_type = ct,
    n_nmd_genes = nrow(genes_ct),
    n_genes_with_nmd_isoform = sum(genes_with_isoforms$has_nmd_isoform),
    pct_genes_with_nmd_isoform = 100 * sum(genes_with_isoforms$has_nmd_isoform) / nrow(genes_ct)
  )
})

cat("**Table 1. Gene → Isoform Concordance by Cell Type**\n\n")
cat("Column definitions:\n")
cat("- **n_nmd_genes**: Number of NMD-responsive genes (short-read)\n")
cat("- **n_genes_with_nmd_isoform**: Number of these genes with ≥1 NMD-responsive isoform (long-read)\n")
cat("- **pct_genes_with_nmd_isoform**: Percentage of NMD-responsive genes with concordant isoforms\n\n")

datatable(
  gene_to_isoform,
  options = list(pageLength = 10, scrollX = TRUE),
  caption = "Table 1. Gene to Isoform Concordance"
) %>%
  formatRound(columns = "pct_genes_with_nmd_isoform", digits = 1)
```

## Isoform → Gene Concordance

For each NMD-responsive isoform, check if its parent gene is also NMD-responsive.

```{r isoform-to-gene}
cat("## Isoform → Gene Concordance\n\n")
cat("For each NMD-responsive isoform (long-read), determining if its parent gene is NMD-responsive (short-read)\n\n")

isoform_to_gene <- map_dfr(unique(nmd_isoforms$cell_type), function(ct) {
  # Get NMD isoforms for this cell type
  isoforms_ct <- nmd_isoforms %>%
    filter(cell_type == ct) %>%
    select(txid, gene_id_base) %>%
    distinct()

  # Get NMD genes for this cell type
  genes_ct <- nmd_genes %>%
    filter(cell_type == ct) %>%
    select(gene_id_base) %>%
    distinct()

  # Check which isoforms have NMD-responsive parent genes
  isoforms_with_genes <- isoforms_ct %>%
    mutate(has_nmd_gene = gene_id_base %in% genes_ct$gene_id_base)

  # Summary
  tibble(
    cell_type = ct,
    n_nmd_isoforms = nrow(isoforms_ct),
    n_isoforms_with_nmd_gene = sum(isoforms_with_genes$has_nmd_gene),
    pct_isoforms_with_nmd_gene = 100 * sum(isoforms_with_genes$has_nmd_gene) / nrow(isoforms_ct)
  )
})

cat("**Table 2. Isoform → Gene Concordance by Cell Type**\n\n")
cat("Column definitions:\n")
cat("- **n_nmd_isoforms**: Number of NMD-responsive isoforms (long-read)\n")
cat("- **n_isoforms_with_nmd_gene**: Number of these isoforms whose parent gene is NMD-responsive (short-read)\n")
cat("- **pct_isoforms_with_nmd_gene**: Percentage of NMD-responsive isoforms with concordant parent genes\n\n")

datatable(
  isoform_to_gene,
  options = list(pageLength = 10, scrollX = TRUE),
  caption = "Table 2. Isoform to Gene Concordance"
) %>%
  formatRound(columns = "pct_isoforms_with_nmd_gene", digits = 1)
```

# Visualization

## Concordance Summary

```{r concordance-plot, fig.width=10, fig.height=6}
cat("## Concordance Rates Across Cell Types\n\n")

# Combine concordance data
concordance_combined <- bind_rows(
  gene_to_isoform %>%
    select(cell_type, pct_genes_with_nmd_isoform) %>%
    mutate(direction = "Gene → Isoform") %>%
    rename(concordance_pct = pct_genes_with_nmd_isoform),
  isoform_to_gene %>%
    select(cell_type, pct_isoforms_with_nmd_gene) %>%
    mutate(direction = "Isoform → Gene") %>%
    rename(concordance_pct = pct_isoforms_with_nmd_gene)
)

ggplot(concordance_combined, aes(x = cell_type, y = concordance_pct, fill = direction)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", concordance_pct)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Gene → Isoform" = "steelblue", "Isoform → Gene" = "coral")) +
  labs(
    title = "Figure 1: Concordance Between Gene and Isoform-Level NMD Responses",
    subtitle = "Percentage of NMD-responsive features with concordant responses at the other level",
    x = "Cell Type",
    y = "Concordance (%)",
    fill = "Direction"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  ) +
  ylim(0, 100)
```

## Combined Summary Table

```{r combined-summary}
cat("## Combined Summary\n\n")

combined_summary <- gene_to_isoform %>%
  left_join(isoform_to_gene, by = "cell_type", suffix = c("_gene", "_isoform")) %>%
  select(
    cell_type,
    n_nmd_genes,
    n_genes_with_nmd_isoform,
    pct_genes_with_nmd_isoform,
    n_nmd_isoforms,
    n_isoforms_with_nmd_gene,
    pct_isoforms_with_nmd_gene
  )

cat("**Table 3. Combined Gene-Isoform Concordance Summary**\n\n")

datatable(
  combined_summary,
  options = list(pageLength = 10, scrollX = TRUE),
  caption = "Table 3. Combined concordance summary by cell type",
  rownames = FALSE
) %>%
  formatRound(columns = c("pct_genes_with_nmd_isoform", "pct_isoforms_with_nmd_gene"), digits = 1)

# Save output
output_file <- file.path(OUTPUT_DIR, "gene_isoform_concordance_2026.1.16.tsv")
write_tsv(combined_summary, output_file)
cat("\nSaved:", output_file, "\n\n")
```

# Key Findings

```{r findings}
cat("## Summary of Concordance Patterns\n\n")

# Calculate overall statistics
overall_gene_to_iso <- mean(gene_to_isoform$pct_genes_with_nmd_isoform)
overall_iso_to_gene <- mean(isoform_to_gene$pct_isoforms_with_nmd_gene)

cat("**Overall Concordance Rates:**\n\n")
cat("- Gene → Isoform: ", sprintf("%.1f%%", overall_gene_to_iso), " of NMD-responsive genes have ≥1 concordant isoform\n", sep = "")
cat("- Isoform → Gene: ", sprintf("%.1f%%", overall_iso_to_gene), " of NMD-responsive isoforms have concordant parent genes\n\n", sep = "")

# Cell type with highest concordance
best_gene_ct <- gene_to_isoform %>% slice_max(pct_genes_with_nmd_isoform, n = 1)
best_iso_ct <- isoform_to_gene %>% slice_max(pct_isoforms_with_nmd_gene, n = 1)

cat("**Highest Concordance:**\n\n")
cat("- Gene → Isoform: ", best_gene_ct$cell_type, " (", sprintf("%.1f%%", best_gene_ct$pct_genes_with_nmd_isoform), ")\n", sep = "")
cat("- Isoform → Gene: ", best_iso_ct$cell_type, " (", sprintf("%.1f%%", best_iso_ct$pct_isoforms_with_nmd_gene), ")\n\n", sep = "")

# Cell type with lowest concordance
worst_gene_ct <- gene_to_isoform %>% slice_min(pct_genes_with_nmd_isoform, n = 1)
worst_iso_ct <- isoform_to_gene %>% slice_min(pct_isoforms_with_nmd_gene, n = 1)

cat("**Lowest Concordance:**\n\n")
cat("- Gene → Isoform: ", worst_gene_ct$cell_type, " (", sprintf("%.1f%%", worst_gene_ct$pct_genes_with_nmd_isoform), ")\n", sep = "")
cat("- Isoform → Gene: ", worst_iso_ct$cell_type, " (", sprintf("%.1f%%", worst_iso_ct$pct_isoforms_with_nmd_gene), ")\n\n", sep = "")
```

# Session Info

```{r session-info}
sessionInfo()
```
