---
title: "Gene-Level DGE Summary: NMD Inhibition Effects"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r config, include=FALSE}
# =============================================================================
# CONFIGURATION: Edit these paths to match your system
# =============================================================================

# Date stamp for output files (YYYY.M.D format)
DATE_STAMP <- "2026.1.15"

# Input file paths
INPUT_DGE_DIR <- file.path("..", "shortread_dge")

# Output directory
OUTPUT_DIR <- file.path("..", "tmp")

# Output file paths
OUTPUT_SUMMARY <- file.path(OUTPUT_DIR, paste0("gene_summary_statistics_", DATE_STAMP, ".tsv"))
OUTPUT_TOP_HITS <- file.path(OUTPUT_DIR, paste0("gene_top_nmd_responsive_", DATE_STAMP, ".tsv"))
OUTPUT_CELLTYPE_OVERLAP <- file.path(OUTPUT_DIR, paste0("gene_celltype_overlap_", DATE_STAMP, ".tsv"))
OUTPUT_CORE_TARGETS <- file.path(OUTPUT_DIR, paste0("gene_core_nmd_targets_", DATE_STAMP, ".tsv"))
OUTPUT_CELLTYPE_SPECIFIC <- file.path(OUTPUT_DIR, paste0("gene_celltype_specific_", DATE_STAMP, ".tsv"))
OUTPUT_CHARACTERISTICS <- file.path(OUTPUT_DIR, paste0("gene_characteristics_", DATE_STAMP, ".tsv"))
OUTPUT_GSEA_ALL <- file.path(OUTPUT_DIR, paste0("gene_gsea_results_", DATE_STAMP, ".tsv"))
OUTPUT_GSEA_SIG <- file.path(OUTPUT_DIR, paste0("gene_gsea_significant_", DATE_STAMP, ".tsv"))

# Analysis parameters
DGE_FILE_PATTERN <- "^nmd_sr_dge_.*_2026\\.1\\.2\\.csv$"
SIGNIFICANCE_THRESHOLD <- 0.05
TOP_HITS_N <- 500

# Cell type mapping (file codes to display names)
CELLTYPE_MAP <- c(
  "at2" = "AT",
  "dd" = "DD",
  "ddali" = "DD_ALI",
  "doali" = "DO",
  "fb" = "FB",
  "mv" = "MV"
)

# Check that input directory exists
if (!dir.exists(INPUT_DGE_DIR)) {
  stop("DGE directory not found: ", INPUT_DGE_DIR)
}
if (!dir.exists(OUTPUT_DIR)) {
  dir.create(OUTPUT_DIR, recursive = TRUE)
  cat("Created output directory:", OUTPUT_DIR, "\n")
}

cat("Configuration loaded successfully\n")
cat("Input DGE directory:", INPUT_DGE_DIR, "\n")
cat("Output directory:", OUTPUT_DIR, "\n\n")
```

<style>
body {
  font-size: 16px;
}
p {
  font-size: 16px;
  line-height: 1.6;
}
li {
  font-size: 16px;
  line-height: 1.6;
}
</style>

# Analysis Overview

This report analyzes **gene-level differential expression** between Smg1i (NMD inhibitor) and DMSO control across 6 lung cell types using short-read RNA sequencing data.

**Research Questions:**

1. How many genes are responsive to NMD inhibition in each cell type?
2. Which genes are the most significant NMD targets?
3. Are NMD-responsive genes consistent across cell types or cell-type-specific?
4. Do NMD-responsive genes have distinct expression characteristics?
5. What biological pathways are enriched among NMD-responsive genes?

**Key Terminology:**

- **NMD-responsive genes:** Genes with positive log fold change (logFC > 0) and adjusted p-value < 0.05, indicating upregulation when NMD is inhibited by Smg1i treatment
- **Gene-level analysis:** Analysis at the gene level (not isoform/transcript level), aggregating all isoforms per gene

# Load Libraries

```{r load-libs}
library(tidyverse)
library(data.table)
library(DT)
library(edgeR)
library(tibble)
library(rlang)
```

# Load Gene-Level DGE Data

```{r load-dge}
# Read gene-level DGE results for each cell type
dge_files <- list.files(INPUT_DGE_DIR, pattern = DGE_FILE_PATTERN, full.names = TRUE)

cat("Loading", length(dge_files), "gene-level DGE result files...\n")

all_gene_dge <- map_df(dge_files, ~{
  df <- read_csv(.x, show_col_types = FALSE)

  # Extract cell type from filename
  basename <- basename(.x)
  parts <- str_split(basename, "_")[[1]]
  celltype <- parts[4]  # nmd_sr_dge_CELLTYPE_date.csv

  df$cell_type_file <- celltype
  df
})

# Map cell type codes to display names
all_gene_dge$cell_type <- CELLTYPE_MAP[all_gene_dge$cell_type_file]

# Filter out any unmapped cell types
all_gene_dge <- all_gene_dge %>% filter(!is.na(cell_type))

# Validate and clean gene symbols
all_gene_dge <- all_gene_dge %>%
  mutate(
    # Mark invalid gene symbols
    valid_symbol = !is.na(hgnc_symbol) & hgnc_symbol != "" & !str_detect(hgnc_symbol, "^ENSG"),
    # For display, use hgnc_symbol if valid, otherwise ensembl_gene_id_version
    gene_display = ifelse(valid_symbol, hgnc_symbol, ensembl_gene_id_version)
  )

cat("Loaded", nrow(all_gene_dge), "gene-level DGE results\n")
cat("Cell types:", paste(unique(all_gene_dge$cell_type), collapse = ", "), "\n")
cat("Genes per cell type:", n_distinct(all_gene_dge$ensembl_gene_id_version) / n_distinct(all_gene_dge$cell_type), "\n")
cat("Genes with valid HGNC symbols:", sum(all_gene_dge$valid_symbol),
    "(", round(100 * mean(all_gene_dge$valid_symbol), 1), "%)\n\n")
```

# Summary Statistics

```{r summary-stats}
cat("## Overall Summary by Cell Type\n\n")

# Calculate summary statistics per cell type
summary_stats <- all_gene_dge %>%
  group_by(cell_type) %>%
  summarize(
    n_valid_symbols = sum(valid_symbol),
    n_sig = sum(adj.P.Val < SIGNIFICANCE_THRESHOLD),
    pct_sig = 100 * n_sig / n(),
    n_nmd_responsive = sum(logFC > 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD),
    pct_nmd_responsive = 100 * n_nmd_responsive / n_sig,  # As % of significant genes
    n_nmd_suppressed = sum(logFC < 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD),
    pct_nmd_suppressed = 100 * n_nmd_suppressed / n_sig,  # As % of significant genes
    mean_logFC_responsive = mean(logFC[logFC > 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD]),
    median_logFC_responsive = median(logFC[logFC > 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD]),
    min_logFC = min(logFC),
    max_logFC = max(logFC),
    .groups = "drop"
  )

cat("**Table 1. Gene-level DGE summary statistics by cell type**\n\n")
cat("Column definitions:\n")
cat("- **n_valid_symbols**: Genes with valid HGNC symbols\n")
cat("- **n_sig**: Significant genes (adj.P.Val < 0.05)\n")
cat("- **pct_sig**: Percent of tested genes that are significant\n")
cat("- **n_nmd_responsive**: Genes upregulated by Smg1i (logFC > 0, adj.P.Val < 0.05)\n")
cat("- **pct_nmd_responsive**: Percent of significant genes that are NMD-responsive\n")
cat("- **n_nmd_suppressed**: Genes downregulated by Smg1i (logFC < 0, adj.P.Val < 0.05)\n")
cat("- **pct_nmd_suppressed**: Percent of significant genes that are NMD-suppressed\n")
cat("- **mean/median_logFC_responsive**: Average fold change for NMD-responsive genes\n\n")

datatable(summary_stats,
          options = list(pageLength = 10, scrollX = TRUE),
          caption = "Table 1. Gene-level DGE summary statistics by cell type") %>%
  formatRound(columns = c("pct_sig", "pct_nmd_responsive", "pct_nmd_suppressed",
                          "mean_logFC_responsive", "median_logFC_responsive",
                          "min_logFC", "max_logFC"),
              digits = 2)

# Save summary statistics
write_tsv(summary_stats, OUTPUT_SUMMARY)
cat("\nSaved:", OUTPUT_SUMMARY, "\n\n")
```

## Identify NMD-Responsive Genes

```{r identify-nmd-responsive}
# Define NMD-responsive genes
nmd_responsive_genes <- all_gene_dge %>%
  filter(logFC > 0, adj.P.Val < SIGNIFICANCE_THRESHOLD) %>%
  select(cell_type, ensembl_gene_id_version, hgnc_symbol, gene_display,
         logFC, adj.P.Val, AveExpr, t, B)

cat("Total NMD-responsive genes (across all cell types):", nrow(nmd_responsive_genes), "\n")
cat("Unique NMD-responsive genes:", n_distinct(nmd_responsive_genes$ensembl_gene_id_version), "\n\n")

# Summary by cell type
nmd_summary <- nmd_responsive_genes %>%
  group_by(cell_type) %>%
  summarize(
    n_nmd_genes = n(),
    mean_logFC = mean(logFC),
    median_logFC = median(logFC),
    .groups = "drop"
  )

print(nmd_summary)
cat("\n")
```

# Top NMD-Responsive Genes

```{r top-hits, results='asis'}
cat("## Top", TOP_HITS_N, "NMD-Responsive Genes per Cell Type\n\n")

# Create top hits tables for each cell type
table_num <- 2  # Start at 2 since Table 1 is summary statistics

for (ct in unique(all_gene_dge$cell_type)) {
  cat("### ", ct, "\n\n")

  top_hits <- nmd_responsive_genes %>%
    filter(cell_type == ct) %>%
    arrange(adj.P.Val) %>%
    head(TOP_HITS_N) %>%
    select(hgnc_symbol, gene_display, logFC, adj.P.Val, AveExpr)

  cat("**Table", table_num, ". Top", TOP_HITS_N, "NMD-Responsive Genes in", ct, "**\n\n")
  cat("Column definitions:\n")
  cat("- **hgnc_symbol**: HGNC gene symbol\n")
  cat("- **gene_display**: Gene name for display (symbol if available, otherwise Ensembl ID)\n")
  cat("- **logFC**: Log2 fold change (Smg1i vs DMSO)\n")
  cat("- **adj.P.Val**: FDR-adjusted p-value\n")
  cat("- **AveExpr**: Average log2 expression\n\n")

  dt <- datatable(
    top_hits,
    options = list(
      pageLength = 10,
      scrollX = TRUE
    ),
    caption = paste0("Table ", table_num, ". Top ", TOP_HITS_N, " NMD-responsive genes in ", ct)
  ) %>%
    formatRound(columns = c("logFC", "AveExpr"), digits = 3)

  print(htmltools::tagList(dt))
  cat("\n\n")

  table_num <- table_num + 1
}

# Save combined top hits
write_tsv(nmd_responsive_genes, OUTPUT_TOP_HITS)
cat("\nSaved all NMD-responsive genes:", OUTPUT_TOP_HITS, "\n\n")
```

# Visualizations

## Volcano Plots

```{r volcano-plots, fig.width=12, fig.height=8}
# Prepare data for volcano plot
volcano_data <- all_gene_dge %>%
  mutate(
    neg_log10_pval = -log10(P.Value),
    is_nmd_responsive = logFC > 0 & adj.P.Val < SIGNIFICANCE_THRESHOLD,
    significance = case_when(
      is_nmd_responsive ~ "NMD-responsive",
      adj.P.Val < SIGNIFICANCE_THRESHOLD ~ "Significant (other)",
      TRUE ~ "Not significant"
    )
  )

# Volcano plot
ggplot(volcano_data, aes(x = logFC, y = neg_log10_pval, color = significance)) +
  geom_point(alpha = 0.5, size = 0.5) +
  facet_wrap(~cell_type, ncol = 3) +
  scale_color_manual(values = c(
    "NMD-responsive" = "steelblue",
    "Significant (other)" = "coral",
    "Not significant" = "gray70"
  )) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "Figure 1: Volcano Plots - Gene-Level Differential Expression",
    subtitle = "Point colors based on adj.P.Val < 0.05 and logFC > 0",
    x = "Log2 Fold Change (Smg1i vs DMSO)",
    y = "-Log10(P-Value)",
    color = "Status"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```

## MA Plots

```{r ma-plots, fig.width=12, fig.height=8}
# MA plot
ggplot(volcano_data, aes(x = AveExpr, y = logFC, color = significance)) +
  geom_point(alpha = 0.5, size = 0.5) +
  facet_wrap(~cell_type, ncol = 3) +
  scale_color_manual(values = c(
    "NMD-responsive" = "steelblue",
    "Significant (other)" = "coral",
    "Not significant" = "gray70"
  )) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "Figure 2: MA Plots - Expression vs Fold Change",
    subtitle = "Average expression (log2) vs log fold change",
    x = "Average Expression (log2)",
    y = "Log2 Fold Change (Smg1i vs DMSO)",
    color = "Status"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```

## LogFC Distribution

```{r logfc-distribution, fig.width=10, fig.height=6}
# Calculate median logFC for each cell type
median_logfc <- nmd_responsive_genes %>%
  group_by(cell_type) %>%
  summarize(median_logFC = median(logFC), .groups = "drop")

# Density plot
ggplot(nmd_responsive_genes, aes(x = logFC, fill = cell_type, color = cell_type)) +
  geom_density(alpha = 0.3, linewidth = 1) +
  geom_vline(data = median_logfc, aes(xintercept = median_logFC, color = cell_type),
             linetype = "dashed", linewidth = 0.8) +
  labs(
    title = "Figure 3: LogFC Distribution for NMD-Responsive Genes",
    subtitle = "Dashed lines show median logFC per cell type",
    x = "Log2 Fold Change (Smg1i vs DMSO)",
    y = "Density",
    fill = "Cell Type",
    color = "Cell Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )
```

# Cell-Type Comparison Analysis

## Overlap Analysis

```{r celltype-overlap}
cat("## Identifying genes responsive across multiple cell types\n\n")

# Count how many cell types each gene is NMD-responsive in
gene_celltype_counts <- nmd_responsive_genes %>%
  group_by(ensembl_gene_id_version, hgnc_symbol, gene_display) %>%
  summarize(
    n_celltypes = n_distinct(cell_type),
    celltypes = paste(sort(unique(cell_type)), collapse = ", "),
    mean_logFC = mean(logFC),
    .groups = "drop"
  ) %>%
  arrange(desc(n_celltypes), desc(mean_logFC))

# Summary of overlap
overlap_summary <- gene_celltype_counts %>%
  count(n_celltypes, name = "n_genes") %>%
  mutate(pct_genes = 100 * n_genes / sum(n_genes))

cat("**Table 8. Genes by number of cell types with NMD response**\n\n")
print(overlap_summary)
cat("\n")

# Visualization
ggplot(overlap_summary, aes(x = factor(n_celltypes), y = n_genes, fill = factor(n_celltypes))) +
  geom_col() +
  geom_text(aes(label = n_genes), vjust = -0.5, size = 4) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Figure 4: NMD-Responsive Genes by Cell-Type Consistency",
    x = "Number of Cell Types with NMD Response",
    y = "Number of Genes",
    fill = "Cell Types"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

# Save overlap data
write_tsv(gene_celltype_counts, OUTPUT_CELLTYPE_OVERLAP)
cat("\nSaved:", OUTPUT_CELLTYPE_OVERLAP, "\n\n")
```

## NMD-Responsive Genes in Multiple Cell Types

```{r multi-celltype-table}
cat("## NMD-Responsive Genes Present in 2 or More Cell Types\n\n")

# Filter for genes in 2+ cell types
multi_celltype_genes <- gene_celltype_counts %>%
  filter(n_celltypes >= 2) %>%
  arrange(desc(n_celltypes), desc(mean_logFC))

cat("Genes present in 2+ cell types:", nrow(multi_celltype_genes), "\n\n")

if (nrow(multi_celltype_genes) > 0) {
  # Get logFC values for these genes across specified cell types
  # Include only: AT, DD, FB, MV (exclude DO and DD_ALI)
  selected_celltypes <- c("AT", "DD", "FB", "MV")

  # Get logFC values for genes that are NMD-responsive in at least one of the selected cell types
  multi_celltype_wide <- nmd_responsive_genes %>%
    filter(cell_type %in% selected_celltypes) %>%
    semi_join(multi_celltype_genes, by = "ensembl_gene_id_version") %>%
    select(ensembl_gene_id_version, gene_display, cell_type, logFC) %>%
    pivot_wider(
      id_cols = c(ensembl_gene_id_version, gene_display),
      names_from = cell_type,
      values_from = logFC,
      values_fill = NA,
      names_prefix = "logFC_"
    ) %>%
    # Add n_celltypes column by joining with the counts
    left_join(
      gene_celltype_counts %>% select(ensembl_gene_id_version, n_celltypes),
      by = "ensembl_gene_id_version"
    ) %>%
    # Remove ensembl ID and reorder columns: gene_display, n_celltypes, then logFC columns
    select(gene_display, n_celltypes, starts_with("logFC_")) %>%
    # Only keep genes that have at least one logFC value (are responsive in at least one selected cell type)
    filter(if_any(starts_with("logFC_"), ~!is.na(.))) %>%
    arrange(desc(n_celltypes))

  # Save to file first
  output_multi_celltype <- file.path(OUTPUT_DIR, "gene_multi_celltype_nmd_2026.1.15.tsv")
  write_tsv(multi_celltype_wide, output_multi_celltype)
  cat("Saved:", output_multi_celltype, "\n\n")

  # Display in interactive table
  cat("**Table 8A. NMD-Responsive Genes in 2+ Cell Types**\n\n")
  cat("Column definitions:\n")
  cat("- **gene_display**: Gene symbol (HGNC) or Ensembl ID\n")
  cat("- **n_celltypes**: Number of cell types where gene is NMD-responsive\n")
  cat("- **logFC_AT**, **logFC_DD**, **logFC_FB**, **logFC_MV**: Log2 fold change in each cell type (NA if not responsive)\n\n")

  # Determine which logFC columns actually exist in the data
  logfc_cols <- colnames(multi_celltype_wide)[grepl("^logFC_", colnames(multi_celltype_wide))]

  # This must be the last expression in the block to render properly
  datatable(
    multi_celltype_wide,
    options = list(
      pageLength = 25,
      scrollX = TRUE
    ),
    caption = "Table 8A. NMD-responsive genes present in 2 or more cell types"
  ) %>%
    formatRound(columns = logfc_cols, digits = 2)
}
```

## Core NMD Targets

```{r core-targets}
cat("## Genes responsive in >= 4 cell types (Core NMD targets)\n\n")

# Identify core targets (responsive in at least 4 cell types)
core_targets <- gene_celltype_counts %>%
  filter(n_celltypes >= 4) %>%
  arrange(desc(n_celltypes), desc(mean_logFC))

cat("Core NMD targets (≥4 cell types):", nrow(core_targets), "genes\n\n")

if (nrow(core_targets) > 0) {
  # Get logFC values for core targets across all cell types
  core_targets_wide <- nmd_responsive_genes %>%
    semi_join(core_targets, by = "ensembl_gene_id_version") %>%
    select(gene_display, cell_type, logFC) %>%
    pivot_wider(names_from = cell_type, values_from = logFC, values_fill = NA)

  # Display top 50
  cat("**Table 9. Top 50 Core NMD Targets**\n\n")
  datatable(
    core_targets_wide %>% head(50),
    options = list(pageLength = 10, scrollX = TRUE),
    caption = "Table 9. Core NMD targets with logFC values across cell types"
  ) %>%
    formatRound(columns = 2:ncol(core_targets_wide), digits = 2)

  # Heatmap of top 30 core targets
  if (nrow(core_targets) >= 30) {
    top_core <- core_targets %>% head(30) %>% pull(ensembl_gene_id_version)

    heatmap_data <- nmd_responsive_genes %>%
      filter(ensembl_gene_id_version %in% top_core) %>%
      # Prioritize HGNC symbol, fall back to gene_display if not available
      mutate(display_name = ifelse(!is.na(hgnc_symbol) & hgnc_symbol != "", hgnc_symbol, gene_display)) %>%
      select(display_name, cell_type, logFC) %>%
      pivot_wider(names_from = cell_type, values_from = logFC, values_fill = 0) %>%
      column_to_rownames("display_name") %>%
      as.matrix()

    # Simple heatmap using base graphics
    heatmap(heatmap_data,
            Colv = NA,  # Don't cluster columns
            scale = "none",
            col = colorRampPalette(c("white", "steelblue"))(50),
            margins = c(8, 12),
            main = "Figure 5: Heatmap of Top 30 Core NMD Targets\n(LogFC across cell types)")
  }

  # Save core targets
  write_tsv(core_targets_wide, OUTPUT_CORE_TARGETS)
  cat("\nSaved:", OUTPUT_CORE_TARGETS, "\n\n")
} else {
  cat("No genes found in ≥4 cell types\n\n")
}
```

## Cell-Type Specific Genes

```{r celltype-specific}
cat("## Genes responsive in only 1 cell type (Cell-type specific)\n\n")

# Identify cell-type specific genes
celltype_specific <- gene_celltype_counts %>%
  filter(n_celltypes == 1) %>%
  # Join back to get cell type info
  left_join(
    nmd_responsive_genes %>% select(ensembl_gene_id_version, cell_type, logFC, adj.P.Val),
    by = "ensembl_gene_id_version"
  ) %>%
  arrange(cell_type, desc(logFC))

cat("Cell-type specific genes:", nrow(celltype_specific), "genes\n\n")

# Summary by cell type
specific_summary <- celltype_specific %>%
  group_by(cell_type) %>%
  summarize(
    n_genes = n(),
    mean_logFC_specific = mean(logFC),
    median_logFC_specific = median(logFC),
    .groups = "drop"
  )

cat("**Table 10. Cell-type specific NMD-responsive genes**\n\n")
print(specific_summary)
cat("\n")

if (nrow(celltype_specific) > 0) {
  # Display top 10 per cell type
  cat("**Top 10 cell-type specific genes per cell type:**\n\n")

  for (ct in unique(all_gene_dge$cell_type)) {
    ct_specific <- celltype_specific %>%
      filter(cell_type == ct) %>%
      select(gene_display, logFC, adj.P.Val) %>%
      head(10)

    if (nrow(ct_specific) > 0) {
      cat("**", ct, "**\n")
      print(knitr::kable(ct_specific, digits = 3))
      cat("\n")
    }
  }

  # Save cell-type specific genes
  write_tsv(celltype_specific, OUTPUT_CELLTYPE_SPECIFIC)
  cat("\nSaved:", OUTPUT_CELLTYPE_SPECIFIC, "\n\n")
}
```

# Gene Characteristics Analysis

## Expression Level Distribution

We need condition-specific expression data. Let's load the DGEList to get per-sample expression values.

```{r load-dgelist-for-expression}
cat("## Loading DGEList for condition-specific expression\n\n")

# Load the gene-level DGEList (if available)
DGELIST_PATH <- file.path("..", "results", "rds", "dge_gene_2026.1.3.rds")

if (file.exists(DGELIST_PATH)) {
  dge_gene <- readRDS(DGELIST_PATH)

  # Extract normalized CPM
  cpm_normalized <- cpm(dge_gene, normalized = TRUE, log = FALSE)

  # Check what columns exist in samples metadata
  cat("DGEList samples columns:", paste(colnames(dge_gene$samples), collapse = ", "), "\n")

  # Get sample metadata - adapt to actual column names
  sample_metadata <- dge_gene$samples

  # The CPM matrix column names match the rownames of dge_gene$samples
  # We need to use rownames as the join key, not the sample_id column (which contains donor IDs)
  sample_metadata <- sample_metadata %>%
    rownames_to_column("sample_name")  # Use sample_name for the long sample identifier

  # Check if we have treatment/condition info
  if ("treatment" %in% colnames(sample_metadata) && "ct" %in% colnames(sample_metadata)) {
    # Use the ct column and map to match DGE results
    sample_metadata <- sample_metadata %>%
      select(sample_name, treatment, ct) %>%
      mutate(
        # Map DGEList cell types to match the CELLTYPE_MAP used in DGE loading
        cell_type = case_when(
          ct == "AT2" ~ "AT",        # AT2 in DGEList -> AT in DGE results
          ct == "DO_ALI" ~ "DO",     # DO_ALI in DGEList -> DO in DGE results
          TRUE ~ ct                   # DD, DD_ALI, FB, MV match directly
        )
      ) %>%
      select(-ct)
  } else if ("group" %in% colnames(sample_metadata)) {
    # Try to extract from group column (might be formatted as celltype_treatment)
    cat("Extracting treatment and cell type from 'group' column\n")
    sample_metadata <- sample_metadata %>%
      select(sample_name, group) %>%
      mutate(
        # Try to parse group name (e.g., "AT_DMSO", "DD_Smg1i")
        cell_type = str_extract(group, "^[^_]+"),
        treatment = str_extract(group, "[^_]+$")
      ) %>%
      select(sample_name, treatment, cell_type)
  } else {
    cat("Could not find treatment/cell type information in DGEList\n")
    has_condition_data <- FALSE
  }

  if (exists("sample_metadata")) {
    # Convert to long format with sample metadata
    # The colnames of the CPM matrix match the rownames of dge_gene$samples
    cpm_long <- as.data.frame(cpm_normalized) %>%
      mutate(dgelist_gene_id = rownames(.)) %>%
      pivot_longer(cols = -dgelist_gene_id, names_to = "sample_name", values_to = "cpm") %>%
      left_join(sample_metadata, by = "sample_name")

    # Add gene information from DGEList
    # Check what columns exist in dge_gene$genes
    cat("DGEList genes columns:", paste(colnames(dge_gene$genes), collapse = ", "), "\n")

    gene_info <- dge_gene$genes %>%
      rownames_to_column("dgelist_gene_id")

    # Try to find the Ensembl ID column
    ensembl_col <- NULL
    if ("ensembl_gene_id_version" %in% colnames(gene_info)) {
      ensembl_col <- "ensembl_gene_id_version"
    } else if ("ensembl_gene_id" %in% colnames(gene_info)) {
      ensembl_col <- "ensembl_gene_id"
    } else if ("gene_id" %in% colnames(gene_info)) {
      ensembl_col <- "gene_id"
    }

    if (!is.null(ensembl_col)) {
      cat("Using", ensembl_col, "as the gene identifier\n")
      gene_info <- gene_info %>%
        select(dgelist_gene_id, !!sym(ensembl_col), hgnc_symbol)
    } else {
      cat("WARNING: Could not find Ensembl gene ID column in DGEList\n")
      gene_info <- gene_info %>%
        select(dgelist_gene_id, hgnc_symbol)
    }

    cpm_long <- cpm_long %>%
      left_join(gene_info, by = "dgelist_gene_id")

    cat("Loaded per-sample expression data\n")
    cat("Samples:", n_distinct(cpm_long$sample_name), "\n")

    # Check what gene ID column we have
    if (!is.null(ensembl_col)) {
      cat("Genes:", n_distinct(cpm_long[[ensembl_col]]), "\n")
      cat("Gene identifier column:", ensembl_col, "\n")
    } else {
      cat("Using hgnc_symbol as gene identifier\n")
    }

    cat("Treatments:", paste(unique(cpm_long$treatment), collapse = ", "), "\n")
    cat("Cell types:", paste(unique(cpm_long$cell_type), collapse = ", "), "\n\n")

    has_condition_data <- TRUE
    # Store ensembl_col for later use
    gene_id_column <- ensembl_col
  }
} else {
  cat("DGEList file not found, using AveExpr from DGE results\n\n")
  has_condition_data <- FALSE
  gene_id_column <- NULL
}
```

### Expression Distribution in DMSO

```{r expression-dmso, fig.width=10, fig.height=6}
if (has_condition_data) {
  # Calculate mean expression per gene per treatment per cell type
  if (!is.null(gene_id_column)) {
    gene_expr_by_condition <- cpm_long %>%
      filter(!is.na(treatment), !is.na(cell_type)) %>%
      group_by(cell_type, treatment, !!sym(gene_id_column), hgnc_symbol) %>%
      summarize(mean_cpm = mean(cpm), .groups = "drop") %>%
      rename(gene_id = !!sym(gene_id_column))
  } else {
    # If no Ensembl ID, just use hgnc_symbol
    gene_expr_by_condition <- cpm_long %>%
      filter(!is.na(treatment), !is.na(cell_type)) %>%
      group_by(cell_type, treatment, hgnc_symbol) %>%
      summarize(mean_cpm = mean(cpm), .groups = "drop") %>%
      mutate(gene_id = hgnc_symbol)
  }

  cat("Gene expression by condition:\n")
  cat("- Cell types:", paste(unique(gene_expr_by_condition$cell_type), collapse = ", "), "\n")
  cat("- Treatments:", paste(unique(gene_expr_by_condition$treatment), collapse = ", "), "\n\n")

  # Join with NMD-responsive classification
  nmd_gene_ids <- nmd_responsive_genes %>%
    select(cell_type, ensembl_gene_id_version) %>%
    distinct()

  cat("NMD-responsive genes cell types:", paste(unique(nmd_gene_ids$cell_type), collapse = ", "), "\n\n")

  # Map cell types if they don't match
  # Check if cell_type values match between datasets
  ct_match <- any(gene_expr_by_condition$cell_type %in% nmd_gene_ids$cell_type)

  if (!ct_match) {
    cat("WARNING: Cell type names don't match between DGEList and DGE results\n")
    cat("Attempting to map cell types...\n\n")

    # Try to map using CELLTYPE_MAP or reverse mapping
    # Get unique cell types from both
    dgelist_cts <- unique(gene_expr_by_condition$cell_type)
    dge_cts <- unique(nmd_gene_ids$cell_type)

    cat("DGEList cell types:", paste(dgelist_cts, collapse = ", "), "\n")
    cat("DGE result cell types:", paste(dge_cts, collapse = ", "), "\n\n")

    # Create reverse mapping (display names back to codes if needed)
    reverse_map <- setNames(names(CELLTYPE_MAP), CELLTYPE_MAP)

    # Try mapping DGEList cell types to match DGE results
    gene_expr_by_condition <- gene_expr_by_condition %>%
      mutate(cell_type_mapped = case_when(
        cell_type %in% names(reverse_map) ~ reverse_map[cell_type],
        cell_type %in% names(CELLTYPE_MAP) ~ CELLTYPE_MAP[cell_type],
        TRUE ~ cell_type
      ))

    # Use mapped cell type - ensure it's character
    gene_expr_by_condition <- gene_expr_by_condition %>%
      mutate(cell_type = as.character(cell_type_mapped)) %>%
      select(-cell_type_mapped)

    cat("After mapping, DGEList cell types:", paste(unique(gene_expr_by_condition$cell_type), collapse = ", "), "\n\n")
  }

  gene_expr_dmso <- gene_expr_by_condition %>%
    filter(treatment == "DMSO", mean_cpm > 0) %>%
    mutate(
      log2_cpm = log2(mean_cpm + 1),
      is_nmd_responsive = paste0(cell_type, gene_id) %in%
        paste0(nmd_gene_ids$cell_type, nmd_gene_ids$ensembl_gene_id_version)
    ) %>%
    mutate(
      classification = ifelse(is_nmd_responsive, "NMD-responsive", "Other genes")
    )

  cat("DMSO expression data rows:", nrow(gene_expr_dmso), "\n")
  cat("NMD-responsive genes in DMSO:", sum(gene_expr_dmso$is_nmd_responsive), "\n\n")

  # Only plot if we have data
  if (nrow(gene_expr_dmso) > 0 && n_distinct(gene_expr_dmso$cell_type) > 0) {
    # Violin plot for DMSO
    ggplot(gene_expr_dmso, aes(x = classification, y = log2_cpm, fill = classification)) +
    geom_violin(alpha = 0.6) +
    geom_boxplot(width = 0.2, alpha = 0.8, outlier.shape = NA) +
    facet_wrap(~cell_type, ncol = 3) +
    scale_fill_manual(values = c(
      "NMD-responsive" = "steelblue",
      "Other genes" = "gray70"
    )) +
    labs(
      title = "Figure 6A: Expression Level Distribution in DMSO",
      subtitle = "NMD-responsive genes vs all other genes",
      x = "Gene Classification",
      y = "Log2(CPM + 1)",
      fill = "Classification"
    ) +
    theme_minimal() +
    theme(
      legend.position = "top",
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12),
      plot.title = element_text(size = 14, face = "bold")
    )
  } else {
    cat("No data available for plotting after filtering\n")
  }
} else {
  cat("Condition-specific data not available\n")
}
```

### Expression Distribution in Smg1i

```{r expression-smg1i, fig.width=10, fig.height=6}
if (has_condition_data) {
  gene_expr_smg1i <- gene_expr_by_condition %>%
    filter(treatment == "Smg1i", mean_cpm > 0) %>%
    mutate(
      log2_cpm = log2(mean_cpm + 1),
      is_nmd_responsive = paste0(cell_type, gene_id) %in%
        paste0(nmd_gene_ids$cell_type, nmd_gene_ids$ensembl_gene_id_version)
    ) %>%
    mutate(
      classification = ifelse(is_nmd_responsive, "NMD-responsive", "Other genes")
    )

  if (nrow(gene_expr_smg1i) > 0 && n_distinct(gene_expr_smg1i$cell_type) > 0) {
    # Violin plot for Smg1i
    ggplot(gene_expr_smg1i, aes(x = classification, y = log2_cpm, fill = classification)) +
    geom_violin(alpha = 0.6) +
    geom_boxplot(width = 0.2, alpha = 0.8, outlier.shape = NA) +
    facet_wrap(~cell_type, ncol = 3) +
    scale_fill_manual(values = c(
      "NMD-responsive" = "steelblue",
      "Other genes" = "gray70"
    )) +
    labs(
      title = "Figure 6B: Expression Level Distribution in Smg1i",
      subtitle = "NMD-responsive genes vs all other genes",
      x = "Gene Classification",
      y = "Log2(CPM + 1)",
      fill = "Classification"
    ) +
    theme_minimal() +
    theme(
      legend.position = "top",
      axis.text.x = element_text(angle = 45, hjust = 1),
      axis.text = element_text(size = 10),
      axis.title = element_text(size = 12),
      plot.title = element_text(size = 14, face = "bold")
    )

    # Statistical summary
    expr_summary <- bind_rows(
      gene_expr_dmso %>%
        group_by(cell_type, classification) %>%
        summarize(
          treatment = "DMSO",
          mean_log2cpm = mean(log2_cpm),
          median_log2cpm = median(log2_cpm),
          .groups = "drop"
        ),
      gene_expr_smg1i %>%
        group_by(cell_type, classification) %>%
        summarize(
          treatment = "Smg1i",
          mean_log2cpm = mean(log2_cpm),
          median_log2cpm = median(log2_cpm),
          .groups = "drop"
        )
    )

    cat("\n**Table 11. Expression level summary by classification and treatment:**\n\n")
    datatable(expr_summary,
              options = list(pageLength = 10, scrollX = TRUE),
              caption = "Table 11. Expression level summary by classification and treatment") %>%
      formatRound(columns = c("mean_log2cpm", "median_log2cpm"), digits = 2)
  } else {
    cat("No data available for plotting after filtering\n")
  }
} else {
  cat("Condition-specific data not available\n")
}
```

### All Genes: Expression in DMSO

```{r all-genes-expression-dmso, fig.width=10, fig.height=6}
if (has_condition_data) {
  # Calculate mean expression across all cell types for all genes in DMSO
  all_genes_dmso <- gene_expr_by_condition %>%
    filter(treatment == "DMSO", mean_cpm > 0) %>%
    group_by(gene_id, hgnc_symbol) %>%
    summarize(
      mean_cpm = mean(mean_cpm),
      .groups = "drop"
    )

  if (nrow(all_genes_dmso) > 0) {
    ggplot(all_genes_dmso, aes(x = mean_cpm)) +
      geom_density(fill = "gray70", alpha = 0.6) +
      scale_x_log10() +
      labs(
        title = "Figure 6C: All Genes - Expression Distribution in DMSO",
        subtitle = "Mean expression across all cell types",
        x = "Mean CPM in DMSO (log10)",
        y = "Density"
      ) +
      theme_minimal() +
      theme(
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 13),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11)
      )
  } else {
    cat("No data available for Figure 6C\n")
  }
} else {
  cat("Condition-specific data not available\n")
}
```

### NMD-Responsive Genes: Expression in DMSO

```{r nmd-genes-expression-dmso, fig.width=10, fig.height=6}
if (has_condition_data) {
  # Get NMD-responsive genes with their DMSO expression
  nmd_genes_dmso <- gene_expr_by_condition %>%
    filter(treatment == "DMSO") %>%
    inner_join(
      nmd_responsive_genes %>%
        select(cell_type, ensembl_gene_id_version) %>%
        distinct(),
      by = c("cell_type", "gene_id" = "ensembl_gene_id_version")
    ) %>%
    group_by(gene_id, hgnc_symbol) %>%
    summarize(
      mean_cpm = mean(mean_cpm),
      .groups = "drop"
    ) %>%
    filter(mean_cpm > 0)

  if (nrow(nmd_genes_dmso) > 0) {
    ggplot(nmd_genes_dmso, aes(x = mean_cpm)) +
      geom_density(fill = "steelblue", alpha = 0.6) +
      scale_x_log10() +
      labs(
        title = "Figure 6D: NMD-Responsive Genes - Expression in DMSO",
        subtitle = "Mean expression across cell types for NMD-responsive genes",
        x = "Mean CPM in DMSO (log10)",
        y = "Density"
      ) +
      theme_minimal() +
      theme(
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 13),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11)
      )
  } else {
    cat("No data available for Figure 6D\n")
  }
} else {
  cat("Condition-specific data not available\n")
}
```

### NMD-Responsive Genes: DMSO vs Smg1i Expression

```{r nmd-genes-dmso-vs-smg1i, fig.width=10, fig.height=6}
if (has_condition_data) {
  # Get expression in both conditions for NMD-responsive genes
  nmd_genes_both <- gene_expr_by_condition %>%
    inner_join(
      nmd_responsive_genes %>%
        select(cell_type, ensembl_gene_id_version) %>%
        distinct(),
      by = c("cell_type", "gene_id" = "ensembl_gene_id_version")
    ) %>%
    group_by(gene_id, hgnc_symbol, treatment) %>%
    summarize(
      mean_cpm = mean(mean_cpm),
      .groups = "drop"
    ) %>%
    filter(mean_cpm > 0)

  if (nrow(nmd_genes_both) > 0) {
    ggplot(nmd_genes_both, aes(x = mean_cpm, fill = treatment, color = treatment)) +
      geom_density(alpha = 0.4, linewidth = 1) +
      scale_x_log10() +
      scale_fill_manual(values = c("DMSO" = "gray70", "Smg1i" = "steelblue")) +
      scale_color_manual(values = c("DMSO" = "gray70", "Smg1i" = "steelblue")) +
      labs(
        title = "Figure 6E: NMD-Responsive Genes - Expression: DMSO vs Smg1i",
        subtitle = "Mean expression across cell types",
        x = "Mean CPM (log10)",
        y = "Density",
        fill = "Treatment",
        color = "Treatment"
      ) +
      theme_minimal() +
      theme(
        legend.position = "top",
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 13),
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11)
      )
  } else {
    cat("No data available for Figure 6E\n")
  }
} else {
  cat("Condition-specific data not available\n")
}
```

## Effect Size Analysis

```{r effect-size, fig.width=10, fig.height=6}
cat("\n## Distribution of fold changes for NMD-responsive genes\n\n")

# Boxplot of logFC
ggplot(nmd_responsive_genes, aes(x = cell_type, y = logFC, fill = cell_type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  geom_hline(yintercept = 2, linetype = "dashed", color = "red", linewidth = 0.8) +
  labs(
    title = "Figure 7: LogFC Distribution for NMD-Responsive Genes",
    subtitle = "Red line: logFC = 2 (4-fold change threshold)",
    x = "Cell Type",
    y = "Log2 Fold Change",
    fill = "Cell Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 11),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

# Identify genes with extreme fold changes
extreme_fc_genes <- nmd_responsive_genes %>%
  filter(abs(logFC) > 2) %>%
  arrange(desc(logFC))

cat("\n**Table 12. Genes with extreme fold changes (|logFC| > 2):**", nrow(extreme_fc_genes), "genes\n\n")

extreme_fc_summary <- extreme_fc_genes %>%
  group_by(cell_type) %>%
  summarize(n_extreme = n(), .groups = "drop")

print(extreme_fc_summary)
cat("\n")
```

## Expression vs Fold Change Relationship

```{r expr-vs-fc, fig.width=10, fig.height=6}
cat("## Relationship between baseline expression and fold change\n\n")

# Scatter plot with trend line
ggplot(nmd_responsive_genes, aes(x = AveExpr, y = logFC)) +
  geom_point(alpha = 0.4, size = 1, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red", linewidth = 1) +
  facet_wrap(~cell_type, ncol = 3) +
  labs(
    title = "Figure 8: Expression Level vs Fold Change",
    subtitle = "For NMD-responsive genes. Red line: linear fit",
    x = "Average Expression (log2)",
    y = "Log2 Fold Change"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

# Calculate correlation
correlation_results <- nmd_responsive_genes %>%
  group_by(cell_type) %>%
  summarize(
    n_genes = n(),
    correlation = ifelse(n() >= 3, cor(AveExpr, logFC, method = "pearson", use = "complete.obs"), NA_real_),
    p_value = ifelse(n() >= 3,
                     tryCatch(cor.test(AveExpr, logFC)$p.value, error = function(e) NA_real_),
                     NA_real_),
    .groups = "drop"
  )

cat("\n**Table 13. Correlation between expression level and fold change:**\n\n")
print(correlation_results)
cat("\n")

# Save characteristics data
characteristics_data <- nmd_responsive_genes %>%
  select(cell_type, ensembl_gene_id_version, hgnc_symbol, gene_display,
         logFC, adj.P.Val, AveExpr) %>%
  mutate(extreme_fc = abs(logFC) > 2)

write_tsv(characteristics_data, OUTPUT_CHARACTERISTICS)
cat("Saved:", OUTPUT_CHARACTERISTICS, "\n\n")
```

# GSEA: Pathway Enrichment Analysis

## Load Gene Sets

```{r load-genesets}
cat("## Loading MSigDB gene sets for GSEA\n\n")

# Set CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Load GSEA libraries
if (!require("fgsea", quietly = TRUE)) {
  if (!require("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  BiocManager::install("fgsea")
  library(fgsea)
}

if (!require("msigdbr", quietly = TRUE)) {
  install.packages("msigdbr")
  library(msigdbr)
}

library(fgsea)
library(msigdbr)

# GSEA parameters
GSEA_MIN_SIZE <- 15
GSEA_MAX_SIZE <- 500
GSEA_NPERM <- 10000
GSEA_FDR_THRESHOLD <- 0.05

cat("GSEA Parameters:\n")
cat("- Pathway size:", GSEA_MIN_SIZE, "-", GSEA_MAX_SIZE, "genes\n")
cat("- Permutations:", GSEA_NPERM, "\n")
cat("- Significance threshold: FDR <", GSEA_FDR_THRESHOLD, "\n\n")

# Load MSigDB gene sets
cat("Loading Hallmark gene sets...\n")
msigdb_h <- msigdbr(species = "Homo sapiens", category = "H")

cat("Loading GO Biological Process gene sets...\n")
msigdb_bp <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP")

cat("Loading KEGG gene sets...\n")
msigdb_kegg <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:KEGG_LEGACY")

cat("Loading Reactome gene sets...\n")
msigdb_reactome <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "CP:REACTOME")

# Combine all gene sets
all_genesets <- bind_rows(
  msigdb_h %>% mutate(collection = "Hallmark"),
  msigdb_bp %>% mutate(collection = "GO_BP"),
  msigdb_kegg %>% mutate(collection = "KEGG"),
  msigdb_reactome %>% mutate(collection = "Reactome")
)

cat("\nGene set summary:\n")
geneset_summary <- all_genesets %>%
  group_by(collection) %>%
  summarize(
    n_genesets = n_distinct(gs_name),
    n_genes = n_distinct(gene_symbol),
    .groups = "drop"
  )
print(geneset_summary)

# Convert to fgsea format
pathways <- all_genesets %>%
  split(x = .$gene_symbol, f = .$gs_name)

# Filter by size
pathways_filtered <- pathways[
  sapply(pathways, length) >= GSEA_MIN_SIZE &
  sapply(pathways, length) <= GSEA_MAX_SIZE
]

cat("\nTotal pathways before filtering:", length(pathways), "\n")
cat("Pathways after size filtering:", length(pathways_filtered), "\n\n")
```

## Prepare Gene Rankings

```{r prepare-rankings}
cat("## Preparing gene-level rankings for GSEA\n\n")

cat("**Ranking strategy:**\n")
cat("- Rank ALL genes by logFC (not just significant ones)\n")
cat("- Use valid HGNC gene symbols for matching with MSigDB pathways\n")
cat("- This captures the full distribution of NMD effects\n\n")

# Create rankings per cell type
rankings_per_celltype <- list()

for (ct in unique(all_gene_dge$cell_type)) {
  cat("Preparing ranked list for", ct, "...\n")

  # Get all genes with valid symbols for this cell type
  # Handle duplicates by keeping the gene with the best (lowest) P.Value
  # Use P.Value instead of adj.P.Val to minimize ties
  ranked_data <- all_gene_dge %>%
    filter(cell_type == ct, valid_symbol) %>%
    group_by(hgnc_symbol) %>%
    slice_min(order_by = P.Value, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(desc(logFC))

  # Create named vector for GSEA
  ranked_vector <- setNames(ranked_data$logFC, ranked_data$hgnc_symbol)

  rankings_per_celltype[[ct]] <- ranked_vector

  cat("  Genes in ranked list:", length(ranked_vector), "\n")
  cat("  Unique gene symbols:", n_distinct(ranked_data$hgnc_symbol), "\n")
  cat("  Genes with logFC > 0:", sum(ranked_data$logFC > 0), "\n")
  cat("  Range:", round(min(ranked_data$logFC), 2), "to", round(max(ranked_data$logFC), 2), "\n\n")
}
```

## Run GSEA

```{r run-gsea, results='asis'}
cat("## Running GSEA Per Cell Type\n\n")

library(htmltools)

gsea_results_all <- list()
table_num <- 14  # Continue numbering after Table 13

for (ct in names(rankings_per_celltype)) {
  cat("### ", ct, "\n\n")

  ranked_list <- rankings_per_celltype[[ct]]

  cat("Running fgsea with", GSEA_NPERM, "permutations...\n\n")

  gsea_res <- fgsea(
    pathways = pathways_filtered,
    stats = ranked_list,
    minSize = GSEA_MIN_SIZE,
    maxSize = GSEA_MAX_SIZE,
    nperm = GSEA_NPERM
  )

  gsea_res$cell_type <- ct
  gsea_results_all[[ct]] <- gsea_res

  # Summary
  sig_pathways <- sum(gsea_res$padj < GSEA_FDR_THRESHOLD, na.rm = TRUE)
  cat("Pathways with FDR <", GSEA_FDR_THRESHOLD, ":", sig_pathways, "\n\n")

  # Display top 100 results
  top_100 <- gsea_res %>%
    arrange(pval) %>%
    head(100) %>%
    select(pathway, pval, padj, ES, NES, size, leadingEdge) %>%
    mutate(
      pathway_short = str_trunc(pathway, 60),
      leadingEdge = sapply(leadingEdge, function(x) paste(head(x, 5), collapse = ", "))
    ) %>%
    select(pathway_short, pval, padj, ES, NES, size, leadingEdge)

  cat("**Table", table_num, ". Top 100 GSEA Results for", ct, "**\n\n")

  dt <- datatable(
    top_100,
    options = list(
      pageLength = 10,
      scrollX = TRUE
    ),
    caption = paste0("Top 100 GSEA results for ", ct, " (ranked by p-value)")
  ) %>%
    formatRound(columns = c("pval", "padj", "ES", "NES"), digits = 4)

  print(htmltools::tagList(dt))
  cat("\n\n")

  table_num <- table_num + 1
}

# Combine all results
gsea_combined <- bind_rows(gsea_results_all)

cat("\n## Overall GSEA Results\n\n")
cat("- Cell types analyzed:", paste(unique(gsea_combined$cell_type), collapse = ", "), "\n")
cat("- Total pathways tested:", n_distinct(gsea_combined$pathway), "\n")
cat("- Total significant results (FDR <", GSEA_FDR_THRESHOLD, "):",
    sum(gsea_combined$padj < GSEA_FDR_THRESHOLD, na.rm = TRUE), "\n\n")

# Save results
write_tsv(gsea_combined, OUTPUT_GSEA_ALL)
cat("Saved all GSEA results:", OUTPUT_GSEA_ALL, "\n")

# Save significant results
gsea_significant <- gsea_combined %>%
  filter(padj < GSEA_FDR_THRESHOLD) %>%
  arrange(padj)

if (nrow(gsea_significant) > 0) {
  write_tsv(gsea_significant, OUTPUT_GSEA_SIG)
  cat("Saved significant GSEA results:", OUTPUT_GSEA_SIG, "\n\n")
}
```

## Cross-Cell-Type Pathway Analysis

```{r pathway-consistency}
if (nrow(gsea_significant) > 0) {
  cat("## Pathways Consistent Across Cell Types\n\n")

  # Find pathways significant in multiple cell types
  pathway_counts <- gsea_significant %>%
    count(pathway) %>%
    arrange(desc(n)) %>%
    filter(n > 1)

  cat("Pathways significant in multiple cell types:", nrow(pathway_counts), "\n\n")

  if (nrow(pathway_counts) > 0) {
    cat("**Table 20. Top pathways by consistency:**\n\n")
    datatable(
      pathway_counts %>% head(20),
      options = list(pageLength = 10, scrollX = TRUE),
      caption = "Table 20. Pathways significant in multiple cell types"
    )

    # Show details for most consistent pathways
    cat("\n**Details for top consistent pathways:**\n\n")
    top_consistent <- pathway_counts %>% head(5) %>% pull(pathway)

    for (pw in top_consistent) {
      cat("\n**", pw, "**\n\n")
      pw_details <- gsea_significant %>%
        filter(pathway == pw) %>%
        select(cell_type, NES, padj) %>%
        arrange(cell_type)
      print(knitr::kable(pw_details, digits = 4))
    }
  }
} else {
  cat("## Cross-Cell-Type Pathway Analysis\n\n")
  cat("No significant pathways found (FDR < ", GSEA_FDR_THRESHOLD, ")\n\n")
}
```

# Key Findings

```{r key-findings}
cat("\n## Summary of Key Findings\n\n")

cat("**Gene-Level NMD Response:**\n")
cat("- Average", round(mean(summary_stats$n_nmd_responsive)),
    "genes per cell type are NMD-responsive (",
    round(mean(summary_stats$pct_nmd_responsive), 1), "% of tested genes)\n")
cat("- Mean logFC for NMD-responsive genes:",
    round(mean(summary_stats$mean_logFC_responsive), 2), "\n")
cat("- Median logFC:", round(mean(summary_stats$median_logFC_responsive), 2), "\n\n")

cat("**Cell-Type Consistency:**\n")
cat("- Unique NMD-responsive genes across all cell types:",
    n_distinct(nmd_responsive_genes$ensembl_gene_id_version), "\n")
if (exists("core_targets") && nrow(core_targets) > 0) {
  cat("- Core NMD targets (≥4 cell types):", nrow(core_targets), "genes\n")
}
if (exists("celltype_specific") && nrow(celltype_specific) > 0) {
  cat("- Cell-type specific genes (1 cell type only):", nrow(celltype_specific), "genes\n")
}
cat("\n")

cat("**Gene Characteristics:**\n")
if (exists("extreme_fc_genes")) {
  cat("- Genes with extreme fold changes (|logFC| > 2):", nrow(extreme_fc_genes), "\n")
}
cat("- Expression-fold change correlation: varies by cell type\n")
cat("  (see correlation table above for details)\n\n")

if (exists("gsea_significant") && nrow(gsea_significant) > 0) {
  cat("**GSEA - Pathway Enrichment:**\n")
  cat("- Total significant pathway enrichments:", nrow(gsea_significant), "\n")
  cat("- Pathways with high NMD effect identified across cell types\n")
  cat("- See GSEA section for detailed pathway analysis\n\n")
} else {
  cat("**GSEA - Pathway Enrichment:**\n")
  cat("- No pathways reached significance threshold\n")
  cat("- NMD-responsive genes may be broadly distributed across pathways\n\n")
}
```

# Session Info

```{r session-info}
sessionInfo()
```
