---
title: "Isoform Pattern Interpretation: NMD Inhibition Effects"
author: "Analysis Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: lumen
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

<style>
body {
  font-size: 16px;
}
p {
  font-size: 16px;
  line-height: 1.6;
}
li {
  font-size: 16px;
  line-height: 1.6;
}
</style>

# Analysis Overview

This report interprets isoform proportion patterns to understand the effects of NMD inhibition (Smg1i treatment) on transcript diversity across cell types.

**Research Questions:**

1. How does NMD inhibition affect the number of detected isoforms per gene?
2. What proportion of genes have a dominant isoform (>50% expression)?
3. What were the baseline (DMSO) proportions of transcripts that are upregulated by NMD inhibition?
4. Are these patterns consistent across cell types?
5. What proportion of transcriptional output is degraded by NMD at the gene level and within cell types?

# Load Libraries

```{r load-libs}
library(tidyverse)
library(data.table)
library(DT)
library(edgeR)
```

# Load Data

## DGEList for Normalized Counts

```{r load-dgelist}
cat("Loading DGEList for normalized CPM values...\n")
dge_isoform <- readRDS("../results/rds/dge_isoform_2026.1.3.rds")

cat("DGEList dimensions:", nrow(dge_isoform), "transcripts x", ncol(dge_isoform), "samples\n")

# Extract normalized CPM
cpm_normalized <- cpm(dge_isoform, normalized = TRUE, log = FALSE)
cat("Calculated normalized CPM\n\n")
```

**Note on Normalization:**
All isoform proportions and expression analyses use normalized CPM values from the DGEList.
This accounts for:

- Library size differences across samples
- Sequencing depth variations
- TMM normalization factors

This makes the isoform proportion analysis consistent with the differential expression analysis (limma-voom).

## Calculate Isoform Proportions from Normalized CPM

```{r calculate-proportions}
cat("Calculating isoform proportions from normalized CPM...\n")

# Convert CPM matrix to long format with metadata
cpm_df <- as.data.frame(cpm_normalized)
cpm_df$transcript_id <- dge_isoform$genes$txid

cpm_long <- cpm_df %>%
  pivot_longer(cols = -transcript_id, names_to = "bamid", values_to = "cpm") %>%
  left_join(dge_isoform$samples %>% select(bamid, sample, treatment, ct),
            by = "bamid") %>%
  rename(cell_type = ct)

dge_isoform$genes$hgnc_id <- ifelse(is.na(dge_isoform$genes$hgnc_id.gc), 
                                     dge_isoform$genes$hgnc_id.sq, 
                                     dge_isoform$genes$hgnc_id.gc)

# Add gene information
cpm_long <- cpm_long %>%
  left_join(dge_isoform$genes %>% select(txid, hgnc_id),
            by = c("transcript_id" = "txid")) %>%
  rename(gene_id = hgnc_id)

# Calculate isoform proportions
all_proportions <- cpm_long %>%
  group_by(cell_type, treatment, sample, gene_id) %>%
  mutate(
    gene_total_cpm = sum(cpm),
    isoform_proportion = cpm / gene_total_cpm,
    isoform_proportion = ifelse(is.nan(isoform_proportion), 0, isoform_proportion),
    n_isoforms = n_distinct(transcript_id)
  ) %>%
  ungroup() %>%
  rename(normalized_cpm = cpm)

cat("Created proportion dataframe with", nrow(all_proportions), "rows\n")
cat("Cell types:", paste(unique(all_proportions$cell_type), collapse = ", "), "\n")
cat("Treatments:", paste(unique(all_proportions$treatment), collapse = ", "), "\n\n")
```

## Apply Abundance Filter

To focus on more abundant isoforms and reduce noise from very low-expression transcripts, we apply a filter requiring **≥0.25 normalized CPM in at least one sample in EITHER DMSO or Smg1i** within each cell type.

```{r apply-filter}
cat("Applying abundance filter (>=0.25 CPM in EITHER condition)...\n\n")

# Load filtered isoform dictionary
filtered_dict <- readRDS("../tmp/filtered_isoform_dictionary_0.25cpm_2026.1.3.rds")

cat("Loaded filtered isoform dictionary with", nrow(filtered_dict), "entries\n")
cat("Unique isoforms:", n_distinct(filtered_dict$transcript_id), "\n\n")

# Show before filtering
before_stats <- all_proportions %>%
  group_by(cell_type) %>%
  summarize(
    n_isoforms_before = n_distinct(transcript_id),
    n_genes_before = n_distinct(gene_id),
    .groups = "drop"
  )

# Apply filter
all_proportions_filtered <- all_proportions %>%
  inner_join(
    filtered_dict %>% select(cell_type, transcript_id),
    by = c("cell_type", "transcript_id")
  )

# Show after filtering
after_stats <- all_proportions_filtered %>%
  group_by(cell_type) %>%
  summarize(
    n_isoforms_after = n_distinct(transcript_id),
    n_genes_after = n_distinct(gene_id),
    .groups = "drop"
  )

# Combine statistics
filter_stats <- before_stats %>%
  left_join(after_stats, by = "cell_type") %>%
  mutate(
    isoforms_removed = n_isoforms_before - n_isoforms_after,
    pct_isoforms_retained = 100 * n_isoforms_after / n_isoforms_before,
    genes_removed = n_genes_before - n_genes_after,
    pct_genes_retained = 100 * n_genes_after / n_genes_before
  )

cat("Filter results by cell type:\n\n")
print(filter_stats)
cat("\n")

# Use filtered data for all downstream analyses
all_proportions <- all_proportions_filtered

cat("After filtering:", nrow(all_proportions), "rows\n")
cat("Unique isoforms:", n_distinct(all_proportions$transcript_id), "\n")
cat("Unique genes:", n_distinct(all_proportions$gene_id), "\n\n")
```

**Rationale for 0.25 CPM threshold:**

- Retains 98.2% of statistically significant isoforms from DGE analysis
- Filters out very low-abundance transcripts that dominate the dataset but have minimal biological signal
- Reduces computational burden and focuses interpretation on more reliably detected isoforms

## Isoform DGE Results

```{r load-dge}
# Read longread DGE results for each cell type
dge_files <- list.files("../longread_dge", pattern = "^nmd_dge_.*\\.csv$", full.names = TRUE)

cat("Loading", length(dge_files), "DGE result files...\n")

all_dge <- map_df(dge_files, ~{
  df <- read_csv(.x, show_col_types = FALSE)

  # Extract cell type from filename
  basename <- basename(.x)
  parts <- str_split(basename, "_")[[1]]
  celltype <- parts[3]

  df$cell_type_file <- celltype
  df
})

# Map cell type codes to match proportion data
# Note: DO cells were only cultured at ALI, so doali maps to DO
celltype_map <- c(
  "at2" = "AT",
  "dd" = "DD",
  "ddali" = "DD_ALI",
  "doali" = "DO",
  "fb" = "FB",
  "mv" = "MV"
)

all_dge$cell_type <- celltype_map[all_dge$cell_type_file]

# Filter out any unmapped cell types
all_dge <- all_dge %>% filter(!is.na(cell_type))

cat("Loaded", nrow(all_dge), "DGE results\n")
cat("Cell types:", paste(unique(all_dge$cell_type), collapse = ", "), "\n\n")
```

# Q1: Isoforms Per Gene

**Question:** Does NMD inhibition reveal more isoforms per gene?

```{r q1-isoforms-per-gene}
# Count isoforms with non-zero normalized CPM per gene per treatment per cell type
# Detection threshold: normalized_cpm > 0 (most lenient, equivalent to any detection)
isoforms_per_gene <- all_proportions %>%
  filter(normalized_cpm > 0) %>%
  group_by(cell_type, treatment, gene_id) %>%
  summarize(
    n_isoforms_detected = n_distinct(transcript_id),
    .groups = "drop"
  )

# Compare DMSO vs Smg1i
isoform_comparison <- isoforms_per_gene %>%
  pivot_wider(
    names_from = treatment,
    values_from = n_isoforms_detected,
    values_fill = 0
  ) %>%
  mutate(
    delta_isoforms = Smg1i - DMSO,
    more_tx_smg1i = delta_isoforms > 0,
    fewer_tx_smg1i = delta_isoforms < 0,
    same_tx_count = delta_isoforms == 0
  )

# Summary statistics
q1_summary <- isoform_comparison %>%
  group_by(cell_type) %>%
  summarize(
    n_genes = n(),
    mean_dmso = mean(DMSO),
    mean_smg1i = mean(Smg1i),
    mean_delta = mean(delta_isoforms),
    genes_with_more_tx_smg1i = sum(more_tx_smg1i),
    genes_with_fewer_tx_smg1i = sum(fewer_tx_smg1i),
    genes_with_same_tx_count = sum(same_tx_count),
    pct_more_tx_smg1i = 100 * sum(more_tx_smg1i) / n(),
    .groups = "drop"
  )

cat("\n## Summary: Isoforms Per Gene by Cell Type\n\n")
datatable(q1_summary,
          options = list(pageLength = 10),
          caption = "Average isoforms per gene in DMSO vs Smg1i") %>%
  formatRound(columns = c("mean_dmso", "mean_smg1i", "mean_delta", "pct_more_tx_smg1i"), digits = 2)

# Distribution plot - Bar chart of categories (changed genes only)
category_summary <- isoform_comparison %>%
  mutate(category = case_when(
    more_tx_smg1i ~ "More in Smg1i",
    fewer_tx_smg1i ~ "Fewer in Smg1i",
    same_tx_count ~ "Unchanged"
  )) %>%
  count(cell_type, category) %>%
  filter(category != "Unchanged")

ggplot(category_summary, aes(x = cell_type, y = n, fill = category)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("More in Smg1i" = "steelblue",
                                "Fewer in Smg1i" = "coral")) +
  labs(title = "Figure 1: Genes with Changed Transcript Counts: Smg1i vs DMSO",
       x = "Cell Type",
       y = "Number of Genes",
       fill = "Category") +
  theme_minimal() +
  theme(legend.position = "top")

# Distribution plot - Histogram (limited range for visibility)
ggplot(isoform_comparison %>% filter(delta_isoforms >= -10 & delta_isoforms <= 20),
       aes(x = delta_isoforms)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  facet_wrap(~cell_type, scales = "free_y") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Figure 2: Change in Detected Isoforms: Smg1i vs DMSO",
       subtitle = "Range limited to -10 to +20 for visibility (excludes extreme outliers)",
       x = "Change in Isoform Count (Smg1i - DMSO)",
       y = "Number of Genes") +
  theme_minimal()

# Save results
write_tsv(isoform_comparison,
          "../tmp/isoform_interpretation_q1_isoforms_per_gene_2026.1.3.tsv")
cat("\nSaved: tmp/isoform_interpretation_q1_isoforms_per_gene_2026.1.3.tsv\n")
```

# Q2: Dominant Isoforms

**Question:** How common are dominant isoforms (>50% of gene expression)?

**Analysis Scope:** This analysis is limited to genes with **3 or more isoforms**, as the concept of a "dominant isoform" is most relevant when multiple isoforms are present.

**Definition of Dominant Isoform:**
For each gene in a specific cell type and treatment group, we identify the isoform with the highest average expression (normalized CPM) across all samples in that group. This isoform is considered **dominant** if it accounts for **>50% of the total normalized read counts** of its parent gene in **ALL samples** from that cell type/treatment group.

**Calculation:**
- For each gene in each cell type/treatment group:
  1. Calculate `isoform_proportion = isoform_CPM / sum(all_isoforms_CPM_for_gene)` in each sample
  2. Identify which isoform has the highest average proportion across samples
  3. Check if this isoform has proportion >0.5 in ALL samples
  4. The gene has a "dominant isoform" only if one isoform meets this criterion

```{r q2-dominant-isoforms}
# For each gene in each cell type/treatment, find if there's a consistently dominant isoform
# (i.e., one isoform that is >50% in ALL samples)

# First, get proportions for all isoforms
isoform_props <- all_proportions %>%
  filter(normalized_cpm > 0) %>%
  select(cell_type, treatment, sample, gene_id, transcript_id, isoform_proportion)

# Identify genes with 3+ isoforms in each cell type/treatment
genes_with_3plus_isoforms <- isoform_props %>%
  group_by(cell_type, treatment, gene_id) %>%
  summarize(n_isoforms = n_distinct(transcript_id), .groups = "drop") %>%
  filter(n_isoforms >= 3)

cat("Genes with 3+ isoforms by cell type and treatment:\n\n")
genes_3plus_summary <- genes_with_3plus_isoforms %>%
  group_by(cell_type, treatment) %>%
  summarize(n_genes_3plus = n(), .groups = "drop")
print(genes_3plus_summary)
cat("\n")

# Show isoform distribution (number of isoforms per gene)
isoform_dist <- isoform_props %>%
  group_by(cell_type, treatment, gene_id) %>%
  summarize(n_isoforms = n_distinct(transcript_id), .groups = "drop")

# Calculate break point for y-axis (2-isoform genes account for 95% of bottom range)
isoform_counts <- isoform_dist %>%
  filter(n_isoforms <= 10) %>%
  count(cell_type, treatment, n_isoforms)

max_2_isoform <- isoform_counts %>%
  filter(n_isoforms == 2) %>%
  pull(n) %>%
  max()

# Set bottom section upper limit so 2-isoform peak is at 95% of range
bottom_limit <- max_2_isoform / 0.95
# Break starts slightly above bottom limit, top section starts well above (e.g., where 1-isoform genes begin)
max_overall <- isoform_counts %>% pull(n) %>% max()
break_start <- bottom_limit * 1.05
break_end <- max_overall * 0.5  # Start top section around middle of 1-isoform peak

cat("Y-axis break: bottom section 0 -", round(bottom_limit),
    ", break at", round(break_start), "-", round(break_end), "\n\n")

# Install ggbreak if needed
if (!require("ggbreak", quietly = TRUE)) {
  options(repos = c(CRAN = "https://cloud.r-project.org"))
  install.packages("ggbreak")
  library(ggbreak)
}

# Plot isoform distribution with broken y-axis
ggplot(isoform_dist %>% filter(n_isoforms <= 10),
       aes(x = factor(n_isoforms), fill = treatment)) +
  geom_bar(position = "dodge") +
  facet_wrap(~cell_type, scales = "free_y") +
  scale_y_break(c(break_start, break_end), scales = 0.5) +
  scale_fill_manual(values = c("DMSO" = "gray70", "Smg1i" = "steelblue")) +
  labs(title = "Figure 3: Distribution of Isoform Counts per Gene",
       subtitle = "Showing genes with 1-10 isoforms (y-axis has break to show both single and multi-isoform genes)",
       x = "Number of Isoforms per Gene",
       y = "Number of Genes",
       fill = "Treatment") +
  theme_minimal() +
  theme(legend.position = "top")

# Filter to genes with 3+ isoforms for dominant isoform analysis
isoform_props <- isoform_props %>%
  semi_join(genes_with_3plus_isoforms, by = c("cell_type", "treatment", "gene_id"))

# For each gene/cell type/treatment, check if any isoform is >50% in ALL samples
dominant_genes <- isoform_props %>%
  group_by(cell_type, treatment, gene_id, transcript_id) %>%
  summarize(
    n_samples = n(),
    n_dominant_samples = sum(isoform_proportion > 0.5),
    mean_proportion = mean(isoform_proportion),
    min_proportion = min(isoform_proportion),
    .groups = "drop"
  ) %>%
  # An isoform is dominant if it's >50% in ALL samples for this gene
  mutate(is_dominant_all_samples = n_dominant_samples == n_samples & n_samples > 0) %>%
  # For each gene, check if ANY isoform meets the criterion
  group_by(cell_type, treatment, gene_id) %>%
  summarize(
    n_isoforms = n(),
    has_dominant_isoform = any(is_dominant_all_samples),
    max_mean_proportion = max(mean_proportion),
    .groups = "drop"
  )

# Summarize by cell type and treatment
q2_summary <- dominant_genes %>%
  group_by(cell_type, treatment) %>%
  summarize(
    n_genes = n(),
    n_with_dominant = sum(has_dominant_isoform),
    pct_with_dominant = 100 * n_with_dominant / n_genes,
    mean_top_proportion = mean(max_mean_proportion),
    .groups = "drop"
  ) %>%
  arrange(cell_type, treatment)

cat("\n## Summary: Genes with Dominant Isoforms\n\n")
cat("**Table interpretation:**\n")
cat("- `n_genes`: Total genes analyzed in this cell type/treatment\n")
cat("- `n_with_dominant`: Genes where one isoform is >50% in ALL samples\n")
cat("- `pct_with_dominant`: Percentage of genes with a consistently dominant isoform\n")
cat("- `mean_top_proportion`: Average proportion of the top isoform across genes\n\n")

datatable(q2_summary,
          options = list(pageLength = 12),
          caption = "Genes with dominant isoform: one isoform accounts for >50% in ALL samples") %>%
  formatRound(columns = c("pct_with_dominant", "mean_top_proportion"), digits = 2)

# Compare DMSO vs Smg1i
q2_comparison <- q2_summary %>%
  select(cell_type, treatment, pct_with_dominant) %>%
  pivot_wider(names_from = treatment, values_from = pct_with_dominant) %>%
  mutate(delta_pct = Smg1i - DMSO)

cat("\n## Change in Dominant Isoform Prevalence\n\n")
datatable(q2_comparison,
          options = list(pageLength = 10),
          caption = "% genes with dominant isoform: Smg1i vs DMSO") %>%
  formatRound(columns = c("DMSO", "Smg1i", "delta_pct"), digits = 2)

# Visualization: Bar plot comparing dominant isoform prevalence
ggplot(q2_summary, aes(x = cell_type, y = pct_with_dominant, fill = treatment)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 50, linetype = "dashed", color = "gray50") +
  scale_fill_manual(values = c("DMSO" = "gray70", "Smg1i" = "steelblue")) +
  labs(title = "Figure 4: Genes with Consistently Dominant Isoform (>50% in ALL samples)",
       x = "Cell Type",
       y = "% of Genes",
       fill = "Treatment") +
  theme_minimal() +
  theme(legend.position = "top")

cat("\n## Relationship Between Isoform Count and Dominant Isoform Proportion\n\n")

# Calculate max proportion for each gene (excluding single-isoform genes)
# Use full proportions data before filtering to 3+ isoforms
max_proportion_by_gene <- all_proportions %>%
  filter(normalized_cpm > 0) %>%
  group_by(cell_type, treatment, sample, gene_id) %>%
  mutate(n_isoforms = n_distinct(transcript_id)) %>%
  filter(n_isoforms >= 2) %>%  # Exclude single-isoform genes
  # Find max proportion in each sample
  group_by(cell_type, treatment, sample, gene_id) %>%
  summarize(
    n_isoforms = first(n_isoforms),
    max_proportion = max(isoform_proportion),
    .groups = "drop"
  ) %>%
  # Average across samples for each gene
  group_by(cell_type, treatment, gene_id, n_isoforms) %>%
  summarize(
    mean_max_proportion = mean(max_proportion),
    .groups = "drop"
  )

cat("Genes included in analysis (≥2 isoforms):\n")
max_prop_summary <- max_proportion_by_gene %>%
  group_by(cell_type, treatment) %>%
  summarize(
    n_genes = n(),
    mean_n_isoforms = mean(n_isoforms),
    mean_max_prop = mean(mean_max_proportion),
    .groups = "drop"
  )
print(max_prop_summary)
cat("\n")

# Diagnostic: Check data distribution for each cell type in the plot range
cat("Data check for Figure 4A (DMSO, 2-15 isoforms):\n")
fig4a_data_check <- max_proportion_by_gene %>%
  filter(treatment == "DMSO", n_isoforms <= 15) %>%
  group_by(cell_type) %>%
  summarize(
    n_genes = n(),
    n_unique_isoform_counts = n_distinct(n_isoforms),
    min_isoforms = min(n_isoforms),
    max_isoforms = max(n_isoforms),
    range_max_prop = paste0(round(min(mean_max_proportion), 3), "-", round(max(mean_max_proportion), 3)),
    .groups = "drop"
  )
print(fig4a_data_check)
cat("\n")

# Check DD specifically
dd_check <- max_proportion_by_gene %>%
  filter(treatment == "DMSO", cell_type == "DD", n_isoforms <= 15) %>%
  group_by(n_isoforms) %>%
  summarize(n_genes = n(), .groups = "drop")
cat("DD gene counts by isoform number:\n")
print(dd_check)
cat("\n")

# Check for data quality issues
data_quality_check <- max_proportion_by_gene %>%
  filter(treatment == "DMSO", n_isoforms <= 15) %>%
  group_by(cell_type) %>%
  summarize(
    n_genes = n(),
    n_na_prop = sum(is.na(mean_max_proportion)),
    n_inf_prop = sum(is.infinite(mean_max_proportion)),
    n_na_isoforms = sum(is.na(n_isoforms)),
    .groups = "drop"
  )
cat("Data quality check:\n")
print(data_quality_check)
cat("\n")

# Plot: isoform count vs max proportion (DMSO only)
# Filter to 2-15 isoforms to focus on the biologically relevant range
ggplot(max_proportion_by_gene %>% filter(treatment == "DMSO", n_isoforms <= 15),
       aes(x = n_isoforms, y = mean_max_proportion)) +
  geom_point(alpha = 0.3, size = 0.8, color = "gray70") +
  geom_smooth(method = "loess", se = TRUE, linewidth = 1, color = "gray40") +
  facet_wrap(~cell_type) +
  scale_x_continuous(breaks = c(2, 5, 10, 15), limits = c(1.5, 15.5)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red", alpha = 0.5) +
  labs(title = "Figure 4A: Dominant Isoform Proportion vs Isoform Count (DMSO)",
       subtitle = "Genes with 2-15 isoforms. Red line: 50% threshold.",
       x = "Number of Isoforms per Gene",
       y = "Proportion of Most Highly Expressed Isoform") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

# Save results
write_tsv(q2_summary,
          "../tmp/isoform_interpretation_q2_dominant_isoforms_2026.1.3.tsv")
cat("\nSaved: tmp/isoform_interpretation_q2_dominant_isoforms_2026.1.3.tsv\n")
```

# Q3: Baseline Proportions of NMD-Responsive Transcripts

**Question:** For NMD-responsive transcripts, what were their baseline (DMSO) proportions?

**Definition of NMD-Responsive Transcript:**
Transcripts from the longread_dge differential expression analysis (Smg1i vs DMSO) that meet both criteria:

1. **logFC > 0**: Positive log fold change (higher expression in Smg1i compared to DMSO)
2. **adj.P.Val < 0.05**: Statistically significant after FDR correction

These transcripts are upregulated when the NMD pathway is inhibited with Smg1i treatment, indicating they are normally targeted for degradation by NMD. They represent classic NMD targets that are rescued by NMD inhibition. All analyses are stratified by cell type.

```{r q3-upregulated-baseline}
# Identify NMD-responsive transcripts from DGE
# Definition: logFC > 0 (upregulated in Smg1i) AND adj.P.Val < 0.05 (statistically significant)
nmd_responsive <- all_dge %>%
  filter(logFC > 0, adj.P.Val < 0.05) %>%
  select(cell_type, transcript_id = txid, logFC, adj.P.Val)

cat("## NMD-Responsive Transcripts by Cell Type\n\n")
cat("**Note:** All Q3 analyses are stratified by cell type.\n\n")

nmd_responsive_summary <- nmd_responsive %>%
  group_by(cell_type) %>%
  summarize(n_nmd_responsive = n(), .groups = "drop")

print(nmd_responsive_summary)

cat("\nTotal across all cell types:", nrow(nmd_responsive), "results\n")
cat("Unique transcripts (may appear in multiple cell types):", n_distinct(nmd_responsive$transcript_id), "\n\n")

# Get baseline (DMSO) proportions for NMD-responsive transcripts
# CRITICAL: Do NOT filter by normalized_cpm > 0 here!
# We want to capture transcripts with ZERO DMSO expression (complete NMD rescue)
baseline_proportions <- all_proportions %>%
  filter(treatment == "DMSO") %>%
  inner_join(nmd_responsive, by = c("cell_type", "transcript_id")) %>%
  select(cell_type, sample, gene_id, transcript_id,
         normalized_cpm, isoform_proportion, n_isoforms, logFC, adj.P.Val)

cat("## Baseline Proportions Coverage\n\n")

cat("**Why is baseline coverage missing for some NMD-responsive transcripts?**\n")
cat("NMD-responsive transcripts are identified from longread_dge statistical tests (all data),\n")
cat("but baseline proportions come from the filtered all_proportions dataset (>=0.25 CPM filter).\n")
cat("Transcripts excluded by the abundance filter will not have baseline proportion data,\n")
cat("even if they are statistically significant. These represent very low-abundance isoforms.\n\n")

coverage_summary <- nmd_responsive %>%
  select(cell_type, transcript_id) %>%
  distinct() %>%
  left_join(
    baseline_proportions %>% select(cell_type, transcript_id) %>% distinct() %>% mutate(has_baseline = TRUE),
    by = c("cell_type", "transcript_id")
  ) %>%
  group_by(cell_type) %>%
  summarize(
    n_nmd_responsive = n(),
    n_with_baseline = sum(has_baseline, na.rm = TRUE),
    pct_coverage = 100 * n_with_baseline / n(),
    .groups = "drop"
  )

print(coverage_summary)

# Calculate overall coverage
total_nmd_responsive <- sum(coverage_summary$n_nmd_responsive)
total_with_baseline <- sum(coverage_summary$n_with_baseline)
overall_pct_coverage <- 100 * total_with_baseline / total_nmd_responsive

cat("\nTotal baseline proportions found for",
    n_distinct(baseline_proportions$transcript_id),
    "unique transcripts (",
    round(overall_pct_coverage, 1), "% overall coverage)\n\n")

# Categorize by baseline expression level
baseline_categories <- baseline_proportions %>%
  mutate(
    baseline_category = case_when(
      normalized_cpm == 0 ~ "Zero (complete rescue)",
      normalized_cpm < 0.1 ~ "Low (< 0.1 CPM)",
      TRUE ~ "Moderate/High (≥ 0.1 CPM)"
    )
  )

cat("## Baseline Expression Categories\n\n")
cat("Distribution of upregulated transcripts by DMSO baseline expression:\n")
baseline_cat_summary <- baseline_categories %>%
  group_by(cell_type, baseline_category) %>%
  summarize(n = n(), .groups = "drop") %>%
  pivot_wider(names_from = baseline_category, values_from = n, values_fill = 0)

print(baseline_cat_summary)
cat("\n")

# Summary by cell type
# First, create transcript-level summaries (averaging across samples)
transcript_summary <- baseline_proportions %>%
  group_by(cell_type, transcript_id, gene_id, logFC, adj.P.Val) %>%
  summarize(
    mean_cpm = mean(normalized_cpm),
    mean_proportion = mean(isoform_proportion),
    n_samples = n(),
    n_zero_samples = sum(normalized_cpm == 0),
    .groups = "drop"
  ) %>%
  mutate(
    is_zero_all = n_zero_samples == n_samples,  # Zero in all samples
    is_low = mean_cpm > 0 & mean_cpm < 0.1,
    is_expressed = mean_cpm >= 0.1
  )

# Summarize by cell type
q3_summary <- transcript_summary %>%
  group_by(cell_type) %>%
  summarize(
    n_genes = n_distinct(gene_id),
    n_transcripts = n(),
    n_zero_dmso = sum(is_zero_all),  # Transcripts with zero in ALL samples
    n_low_dmso = sum(is_low),
    n_expressed_dmso = sum(is_expressed),
    pct_zero = 100 * n_zero_dmso / n(),
    pct_low = 100 * n_low_dmso / n(),
    mean_baseline_cpm = mean(mean_cpm[mean_cpm > 0]),
    median_baseline_cpm = median(mean_cpm[mean_cpm > 0]),
    mean_baseline_prop = mean(mean_proportion),
    median_baseline_prop = median(mean_proportion),
    pct_minor_isoforms = 100 * sum(mean_proportion < 0.5) / n(),
    pct_very_minor = 100 * sum(mean_proportion < 0.1) / n(),
    mean_logFC = mean(logFC),
    .groups = "drop"
  )

cat("## Summary: Baseline Proportions of NMD-Responsive Transcripts\n\n")

cat("**Understanding the metrics and stratification:**\n\n")

cat("**Stratification:** All Q3 analyses are stratified by cell type. Each row in the table below represents one cell type.\n\n")

cat("**Observation level:** The table summarizes unique NMD-responsive transcripts within each cell type.\n")
cat("The raw data contains multiple observations per transcript (one per DMSO sample), but metrics are calculated at the transcript level by aggregating across samples.\n\n")

cat("**Column definitions:**\n")
cat("- `n_genes`: Number of unique genes (by ENSG ID) represented by NMD-responsive transcripts in this cell type\n")
cat("  - One gene can have multiple NMD-responsive isoforms, so n_genes ≤ n_transcripts\n")
cat("- `n_transcripts`: Total number of unique NMD-responsive transcript isoforms for this cell type\n")
cat("  - These are the transcripts with baseline proportion data\n")
cat("- `n_zero_dmso`: Transcripts with zero normalized CPM in ALL DMSO samples for that cell type\n")
cat("  - These represent complete NMD rescue (no detectable expression in DMSO, rescued by Smg1i)\n")
cat("- `n_low_dmso`: Transcripts with mean normalized CPM between 0 and 0.1 across DMSO samples\n")
cat("  - Low but detectable baseline expression\n")
cat("- `n_expressed_dmso`: Transcripts with mean normalized CPM ≥ 0.1 across DMSO samples\n")
cat("  - Moderate to high baseline expression\n\n")

cat("**Note on coverage:**\n")
cat("- High coverage means we successfully found baseline proportion data for most NMD-responsive transcripts\n")
cat("- This includes transcripts with ZERO baseline expression (which is valid and scientifically important data)\n")
cat("- The transcripts with n_zero_dmso are a SUBSET of those with baseline data\n")
cat("- In other words: Of the transcripts WITH baseline data, approximately 30% have zero DMSO expression\n")
cat("- These zero-baseline transcripts are classic NMD targets showing complete rescue\n\n")

datatable(q3_summary,
          options = list(pageLength = 10),
          caption = "NMD-responsive transcripts: many had zero or low DMSO expression (normalized CPM)") %>%
  formatRound(columns = c("pct_zero", "pct_low", "mean_baseline_cpm", "median_baseline_cpm",
                          "mean_baseline_prop", "median_baseline_prop",
                          "pct_minor_isoforms", "pct_very_minor", "mean_logFC"),
              digits = 2)

# Figure 1: Combined distribution across all cell types
cat("### Distribution of NMD-Responsive Isoforms in DMSO (All Cell Types Combined)\n\n")

# Calculate y-axis break for histogram
# Build histogram data to find bin heights
hist_data <- ggplot_build(
  ggplot(baseline_proportions, aes(x = isoform_proportion)) +
    geom_histogram(bins = 50)
)
bin_counts <- hist_data$data[[1]]$count

# Find the tall bin at proportion = 0 and others
max_bin_at_zero <- max(bin_counts[1:3])  # First few bins (near zero)
max_other_bins <- max(bin_counts[4:length(bin_counts)])  # Rest of distribution

# Set break so non-zero bins are visible (account for 95% of bottom section)
bottom_limit_hist <- max_other_bins / 0.95
break_start_hist <- bottom_limit_hist * 1.05
break_end_hist <- max_bin_at_zero * 0.5

cat("Y-axis break for histogram: bottom 0 -", round(bottom_limit_hist),
    ", break at", round(break_start_hist), "-", round(break_end_hist), "\n\n")

ggplot(baseline_proportions, aes(x = isoform_proportion)) +
  geom_histogram(bins = 50, fill = "gray70", color = "white") +
  scale_y_break(c(break_start_hist, break_end_hist), scales = 0.5) +
  labs(title = "Figure 5: Baseline (DMSO) Proportions of NMD-Responsive Transcripts",
       subtitle = "All cell types combined. Y-axis has break.",
       x = "Isoform Proportion in DMSO",
       y = "Count") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

# Additional analysis: Proportion vs Expression Level
cat("\n### Isoform Proportion vs Expression Level in DMSO\n\n")

# Plot 1: All isoforms (from filtered dataset)
all_dmso_proportions <- all_proportions %>%
  filter(treatment == "DMSO") %>%
  group_by(transcript_id, gene_id) %>%
  summarize(
    mean_cpm = mean(normalized_cpm),
    mean_proportion = mean(isoform_proportion),
    .groups = "drop"
  ) %>%
  filter(mean_cpm > 0)  # Only show expressed isoforms

ggplot(all_dmso_proportions, aes(x = mean_cpm, y = mean_proportion)) +
  geom_point(alpha = 0.3, size = 0.5, color = "gray40") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "orange") +
  scale_x_log10() +
  labs(title = "Figure 6: All Isoforms: Proportion vs Expression in DMSO",
       subtitle = "Combined across all cell types. Lines: 50% (red), 10% (orange)",
       x = "Mean Normalized CPM in DMSO (log10)",
       y = "Mean Isoform Proportion in DMSO") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

# Plot 2: NMD-responsive isoforms in DMSO only
nmd_dmso_proportions <- baseline_proportions %>%
  group_by(transcript_id, gene_id) %>%
  summarize(
    mean_cpm = mean(normalized_cpm),
    mean_proportion = mean(isoform_proportion),
    .groups = "drop"
  ) %>%
  filter(mean_cpm > 0)  # Only show expressed isoforms

ggplot(nmd_dmso_proportions, aes(x = mean_cpm, y = mean_proportion)) +
  geom_point(alpha = 0.4, size = 0.8, color = "gray70") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "orange") +
  scale_x_log10() +
  labs(title = "Figure 7: NMD-Responsive Isoforms: Proportion vs Expression in DMSO",
       subtitle = "Combined across all cell types. Lines: 50% (red), 10% (orange)",
       x = "Mean Normalized CPM in DMSO (log10)",
       y = "Mean Isoform Proportion in DMSO") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

# Plot 3: NMD-responsive isoforms - DMSO vs Smg1i comparison
nmd_dmso_data <- baseline_proportions %>%
  group_by(transcript_id, gene_id) %>%
  summarize(
    mean_cpm = mean(normalized_cpm),
    mean_proportion = mean(isoform_proportion),
    .groups = "drop"
  ) %>%
  filter(mean_cpm > 0) %>%
  mutate(treatment = "DMSO")

# Get Smg1i data for NMD-responsive isoforms
nmd_smg1i_data <- all_proportions %>%
  filter(treatment == "Smg1i") %>%
  inner_join(nmd_responsive, by = c("cell_type", "transcript_id")) %>%
  group_by(transcript_id, gene_id) %>%
  summarize(
    mean_cpm = mean(normalized_cpm),
    mean_proportion = mean(isoform_proportion),
    .groups = "drop"
  ) %>%
  filter(mean_cpm > 0) %>%
  mutate(treatment = "Smg1i")

# Combine for plotting
nmd_combined_data <- bind_rows(nmd_dmso_data, nmd_smg1i_data)

ggplot(nmd_combined_data, aes(x = mean_cpm, y = mean_proportion, color = treatment)) +
  geom_point(alpha = 0.4, size = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black") +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "gray50") +
  scale_x_log10() +
  scale_color_manual(values = c("DMSO" = "gray70", "Smg1i" = "steelblue")) +
  labs(title = "Figure 8: NMD-Responsive Isoforms: DMSO vs Smg1i",
       subtitle = "Proportion vs Expression combined across all cell types. Lines: 50% (black), 10% (gray)",
       x = "Mean Normalized CPM (log10)",
       y = "Isoform Proportion",
       color = "Treatment") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "top"
  )

cat("\n**Interpretation:**\n")
cat("The first plot shows the overall relationship between proportion and expression for all isoforms.\n")
cat("The second plot shows NMD-responsive isoforms in DMSO only - they cluster at low proportion and low expression.\n")
cat("The third plot compares DMSO (gray) vs Smg1i (blue) for NMD-responsive isoforms.\n")
cat("NMD-responsive isoforms shift rightward (increased expression) and often upward (increased proportion)\n")
cat("when NMD is inhibited, though many remain as minor isoforms even after rescue.\n\n")

# Get Smg1i proportions for comparison
# Note: Filter Smg1i by normalized_cpm > 0 (only need to consider expressed in Smg1i for comparison)
smg1i_proportions <- all_proportions %>%
  filter(treatment == "Smg1i", normalized_cpm > 0) %>%
  inner_join(nmd_responsive, by = c("cell_type", "transcript_id")) %>%
  select(cell_type, sample, gene_id, transcript_id, isoform_proportion) %>%
  rename(prop_smg1i = isoform_proportion)

# Figure 2: Overlapping density plot showing DMSO vs Smg1i distribution
cat("\n### Paired Distribution: DMSO vs Smg1i\n\n")

# Combine DMSO and Smg1i proportions for density plot
dmso_for_density <- baseline_proportions %>%
  select(transcript_id, sample, isoform_proportion) %>%
  mutate(treatment = "DMSO")

smg1i_for_density <- smg1i_proportions %>%
  select(transcript_id, sample, prop_smg1i) %>%
  rename(isoform_proportion = prop_smg1i) %>%
  mutate(treatment = "Smg1i")

combined_proportions <- bind_rows(dmso_for_density, smg1i_for_density)

ggplot(combined_proportions, aes(x = isoform_proportion, fill = treatment, color = treatment)) +
  geom_density(alpha = 0.4, linewidth = 1) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black", linewidth = 0.5) +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "gray50", linewidth = 0.5) +
  scale_fill_manual(values = c("DMSO" = "gray70", "Smg1i" = "steelblue")) +
  scale_color_manual(values = c("DMSO" = "gray70", "Smg1i" = "steelblue")) +
  labs(title = "Figure 9: Isoform Proportion Distribution: DMSO vs Smg1i",
       subtitle = "NMD-responsive transcripts across all cell types. Vertical lines: 50% (black) and 10% (gray)",
       x = "Isoform Proportion",
       y = "Density",
       fill = "Treatment",
       color = "Treatment") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 13),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "top"
  )

# Compare before/after
before_after <- baseline_proportions %>%
  select(cell_type, sample, transcript_id, prop_dmso = isoform_proportion) %>%
  inner_join(
    smg1i_proportions %>%
      select(cell_type, sample_orig = sample, transcript_id, prop_smg1i),
    by = c("cell_type", "transcript_id"),
    relationship = "many-to-many"
  ) %>%
  mutate(
    fold_change_prop = prop_smg1i / prop_dmso,
    log2_fc_prop = log2(fold_change_prop)
  )

# Plot before vs after - PROPORTIONS
ggplot(before_after, aes(x = prop_dmso, y = prop_smg1i)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~cell_type) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Figure 10: Isoform Proportions: DMSO vs Smg1i",
       subtitle = "For NMD-responsive transcripts (based on count increases)",
       x = "Proportion in DMSO (log10)",
       y = "Proportion in Smg1i (log10)") +
  theme_minimal()

cat("\n**Important Note on Proportions vs Counts:**\n\n")
cat("These transcripts are 'NMD-responsive' based on NORMALIZED COUNT increases in the DGE analysis.\n")
cat("However, their ISOFORM PROPORTION (fraction of gene expression) may not increase if:\n")
cat("- The whole gene is upregulated proportionally (all isoforms increase together)\n")
cat("- Other isoforms of the same gene increase even more\n\n")
cat("Points below the diagonal indicate transcripts where counts increased but proportion decreased,\n")
cat("suggesting other isoforms of that gene increased more dramatically.\n\n")

# Convert CPM matrix to long format with sample metadata
# Use txid from genes slot to match with DGE results
cpm_df <- as.data.frame(cpm_normalized)
cpm_df$transcript_id <- dge_isoform$genes$txid

cpm_long <- cpm_df %>%
  pivot_longer(cols = -transcript_id, names_to = "bamid", values_to = "cpm") %>%
  left_join(dge_isoform$samples %>% select(bamid, sample, treatment, ct),
            by = "bamid") %>%
  rename(cell_type = ct)

# Calculate mean normalized CPM per treatment per cell type per transcript
mean_cpm <- cpm_long %>%
  inner_join(nmd_responsive, by = c("cell_type", "transcript_id")) %>%
  group_by(cell_type, treatment, transcript_id) %>%
  summarize(
    mean_cpm = mean(cpm),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = treatment,
    values_from = mean_cpm,
    names_prefix = "mean_cpm_"
  ) %>%
  filter(mean_cpm_DMSO > 0, mean_cpm_Smg1i > 0) %>%  # Only transcripts detected in both
  mutate(
    cpm_fc = mean_cpm_Smg1i / mean_cpm_DMSO,
    log2_cpm_fc = log2(cpm_fc)
  )

# Plot using NORMALIZED CPM (accounts for library size and sequencing depth)
ggplot(mean_cpm, aes(x = mean_cpm_DMSO, y = mean_cpm_Smg1i)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  facet_wrap(~cell_type) +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Figure 11: Normalized CPM: DMSO vs Smg1i",
       subtitle = "For NMD-responsive transcripts - all points above diagonal (normalized for library size)",
       x = "Mean CPM in DMSO (log10)",
       y = "Mean CPM in Smg1i (log10)") +
  theme_minimal()

cat("\nIn the count-based plot above, ALL points are now above the diagonal,\n")
cat("confirming these transcripts have increased expression in Smg1i.\n")
cat("Using NORMALIZED CPM accounts for differences in sequencing depth across samples,\n")
cat("making the plot consistent with DGE results (which also use normalized counts).\n\n")

# Save results
write_tsv(baseline_proportions,
          "../tmp/isoform_interpretation_q3_upregulated_baseline_2026.1.3.tsv")
cat("\nSaved: tmp/isoform_interpretation_q3_upregulated_baseline_2026.1.3.tsv\n")
```

# Q4: Cell-Type Specificity

**Question:** Are isoform patterns consistent or cell-type-specific?

```{r q4-celltype-comparison}
# Compare Q1-Q3 metrics across cell types

# Q1: Isoform gain consistency
q1_celltype <- q1_summary %>%
  select(cell_type, mean_delta, pct_more_tx_smg1i) %>%
  mutate(metric = "Isoform Gain")

# Q2: Dominant isoform consistency
q2_celltype <- q2_comparison %>%
  select(cell_type, delta_pct) %>%
  rename(value = delta_pct) %>%
  mutate(metric = "Dominant Isoform Change (%)")

# Q3: Baseline proportion consistency
q3_celltype <- q3_summary %>%
  select(cell_type, median_baseline_prop, pct_very_minor) %>%
  mutate(metric = "Upregulated Baseline")

cat("## Cell-Type Comparison\n\n")

# Plot comparison
p1 <- ggplot(q1_summary, aes(x = cell_type, y = mean_delta, fill = cell_type)) +
  geom_col() +
  labs(title = "Figure 12: Q1: Mean Change in Isoforms per Gene",
       y = "Mean Δ Isoforms (Smg1i - DMSO)",
       x = NULL) +
  theme_minimal() +
  theme(legend.position = "none")

p2 <- ggplot(q2_comparison, aes(x = cell_type, y = delta_pct, fill = cell_type)) +
  geom_col() +
  labs(title = "Figure 13: Q2: Change in Dominant Isoform Prevalence",
       y = "Δ % Genes with Dominant Isoform",
       x = NULL) +
  theme_minimal() +
  theme(legend.position = "none")

p3 <- ggplot(q3_summary, aes(x = cell_type, y = median_baseline_prop, fill = cell_type)) +
  geom_col() +
  labs(title = "Figure 14: Q3: Median Baseline Proportion",
       subtitle = "For NMD-responsive transcripts",
       y = "Median Proportion in DMSO",
       x = NULL) +
  theme_minimal() +
  theme(legend.position = "none")

# Display plots
print(p1)
print(p2)
print(p3)

# Statistical summary
q4_summary <- data.frame(
  metric = c("Mean Δ Isoforms", "Δ % Dominant", "Median Baseline Prop"),
  across_celltypes_mean = c(
    mean(q1_summary$mean_delta),
    mean(q2_comparison$delta_pct),
    mean(q3_summary$median_baseline_prop)
  ),
  across_celltypes_sd = c(
    sd(q1_summary$mean_delta),
    sd(q2_comparison$delta_pct),
    sd(q3_summary$median_baseline_prop)
  ),
  range_min = c(
    min(q1_summary$mean_delta),
    min(q2_comparison$delta_pct),
    min(q3_summary$median_baseline_prop)
  ),
  range_max = c(
    max(q1_summary$mean_delta),
    max(q2_comparison$delta_pct),
    max(q3_summary$median_baseline_prop)
  )
)

cat("\n## Cross-Cell-Type Variability\n\n")
datatable(q4_summary,
          options = list(pageLength = 5),
          caption = "Consistency of patterns across cell types") %>%
  formatRound(columns = 2:5, digits = 3)

# Save results
write_tsv(q4_summary,
          "../tmp/isoform_interpretation_q4_celltype_comparison_2026.1.3.tsv")
cat("\nSaved: tmp/isoform_interpretation_q4_celltype_comparison_2026.1.3.tsv\n")
```

# Q5: Transcriptional Output Devoted to NMD

**Question:** What proportion of transcriptional output is devoted to producing NMD-targeted transcripts?

**Concept:** In the Smg1i condition (NMD inhibited), we observe the true transcriptional output of genes without degradation. This allows us to estimate what fraction of a gene's transcriptional resources are devoted to producing transcripts that are normally targeted for NMD degradation.

```{r q5-transcriptional-output}
cat("## Transcriptional Output Analysis\n\n")

# Use Smg1i data to estimate true transcriptional output (NMD is off)
smg1i_expression <- all_proportions %>%
  filter(treatment == "Smg1i") %>%
  group_by(cell_type, gene_id, transcript_id) %>%
  summarize(
    mean_cpm = mean(normalized_cpm),
    .groups = "drop"
  )

# Identify which transcripts are NMD-responsive
nmd_responsive_set <- nmd_responsive %>%
  select(cell_type, transcript_id) %>%
  distinct() %>%
  mutate(is_nmd_responsive = TRUE)

# Join to mark NMD-responsive transcripts
smg1i_expression <- smg1i_expression %>%
  left_join(nmd_responsive_set, by = c("cell_type", "transcript_id")) %>%
  mutate(is_nmd_responsive = replace_na(is_nmd_responsive, FALSE))

# Gene-level analysis: For genes with NMD-responsive isoforms,
# what % of transcriptional output goes to NMD?
gene_nmd_output <- smg1i_expression %>%
  group_by(cell_type, gene_id) %>%
  summarize(
    total_output = sum(mean_cpm),
    nmd_output = sum(mean_cpm[is_nmd_responsive]),
    n_isoforms = n(),
    n_nmd_isoforms = sum(is_nmd_responsive),
    has_nmd_isoform = any(is_nmd_responsive),
    .groups = "drop"
  ) %>%
  filter(has_nmd_isoform, total_output > 0) %>%
  mutate(
    pct_to_nmd = 100 * nmd_output / total_output
  )

cat("### Gene-Level Analysis\n\n")
cat("For genes harboring NMD-responsive isoforms:\n\n")

# Summary by cell type
gene_summary <- gene_nmd_output %>%
  group_by(cell_type) %>%
  summarize(
    n_genes = n(),
    mean_pct_to_nmd = mean(pct_to_nmd),
    median_pct_to_nmd = median(pct_to_nmd),
    q25_pct = quantile(pct_to_nmd, 0.25),
    q75_pct = quantile(pct_to_nmd, 0.75),
    .groups = "drop"
  )

print(gene_summary)
cat("\n")

# Distribution plot
ggplot(gene_nmd_output, aes(x = pct_to_nmd)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  facet_wrap(~cell_type, scales = "free_y") +
  geom_vline(xintercept = 50, linetype = "dashed", color = "red") +
  labs(title = "Figure 15: Distribution of % Transcriptional Output to NMD per Gene",
       subtitle = "For genes with NMD-responsive isoforms. Red line: 50%",
       x = "% of Gene's Transcriptional Output to NMD",
       y = "Number of Genes") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

# Cell-type level analysis: Overall % of transcriptional output to NMD
cat("\n### Cell-Type Level Analysis\n\n")
cat("Overall transcriptional output devoted to NMD-targeted transcripts:\n\n")
cat("**Calculation:** For each sample, calculate (NMD-responsive CPM sum) / (total CPM sum for genes with NMD isoforms),\n")
cat("then average percentages across samples within each cell type.\n\n")

# Get per-sample CPM values for genes with NMD-responsive isoforms
celltype_nmd_persample <- all_proportions %>%
  filter(treatment == "Smg1i") %>%
  # Only include genes with NMD-responsive isoforms
  inner_join(
    gene_nmd_output %>% select(cell_type, gene_id),
    by = c("cell_type", "gene_id")
  ) %>%
  # Mark NMD-responsive transcripts
  left_join(nmd_responsive_set, by = c("cell_type", "transcript_id")) %>%
  mutate(is_nmd_responsive = replace_na(is_nmd_responsive, FALSE)) %>%
  # Calculate percentage per sample
  group_by(cell_type, sample) %>%
  summarize(
    total_cpm = sum(normalized_cpm),
    nmd_cpm = sum(normalized_cpm[is_nmd_responsive]),
    pct_to_nmd = 100 * nmd_cpm / total_cpm,
    .groups = "drop"
  )

# Average across samples within cell type
celltype_nmd_output <- celltype_nmd_persample %>%
  group_by(cell_type) %>%
  summarize(
    mean_pct_to_nmd = mean(pct_to_nmd),
    sd_pct_to_nmd = sd(pct_to_nmd),
    n_samples = n(),
    .groups = "drop"
  ) %>%
  rename(pct_to_nmd = mean_pct_to_nmd)

print(celltype_nmd_output)
cat("\n")

# Visualization
ggplot(celltype_nmd_output, aes(x = cell_type, y = pct_to_nmd, fill = cell_type)) +
  geom_col() +
  geom_hline(yintercept = mean(celltype_nmd_output$pct_to_nmd),
             linetype = "dashed", color = "black") +
  labs(title = "Figure 16: Overall % of Transcriptional Output to NMD by Cell Type",
       subtitle = paste0("For genes with NMD-responsive isoforms. Dashed line: mean across cell types (",
                        round(mean(celltype_nmd_output$pct_to_nmd), 1), "%)"),
       x = "Cell Type",
       y = "% Transcriptional Output to NMD",
       fill = "Cell Type") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "none"
  )

cat("\n### Alternative Calculation: NMD Output as % of Total Cell-Type Transcriptional Output\n\n")
cat("**Transcriptional output definition:** For each sample, sum CPM for ALL transcripts in the cell type,\n")
cat("using the complete unfiltered dataset (not limited to abundance-filtered transcripts or genes with NMD-responsive isoforms).\n\n")
cat("**Calculation:** For each sample, calculate (NMD-responsive CPM sum) / (ALL transcripts CPM sum),\n")
cat("then average percentages across samples within each cell type.\n\n")
cat("This provides a different perspective on the 'cost' of NMD by showing what percentage of the\n")
cat("entire transcriptional output of the cell type is devoted to NMD-targeted transcripts.\n\n")

# Calculate using ALL transcripts (unfiltered) from the full CPM data, per sample
cpm_long_full <- as.data.frame(cpm_normalized) %>%
  mutate(transcript_id = dge_isoform$genes$txid) %>%
  pivot_longer(cols = -transcript_id, names_to = "bamid", values_to = "cpm") %>%
  left_join(dge_isoform$samples %>% select(bamid, sample, treatment, ct),
            by = "bamid") %>%
  rename(cell_type = ct)

# Calculate per-sample percentages for Smg1i
celltype_nmd_total_persample <- cpm_long_full %>%
  filter(treatment == "Smg1i") %>%
  # Mark NMD-responsive transcripts
  left_join(nmd_responsive_set, by = c("cell_type", "transcript_id")) %>%
  mutate(is_nmd_responsive = replace_na(is_nmd_responsive, FALSE)) %>%
  # Calculate percentage per sample
  group_by(cell_type, sample) %>%
  summarize(
    total_cpm = sum(cpm),
    nmd_cpm = sum(cpm[is_nmd_responsive]),
    pct_to_nmd = 100 * nmd_cpm / total_cpm,
    .groups = "drop"
  )

cat("Per-sample check (first few samples):\n")
print(head(celltype_nmd_total_persample, 10))
cat("\n")
cat("Note: total_cpm should be approximately 1,000,000 per sample (CPM normalization)\n\n")

# Average across samples within cell type
celltype_nmd_output_total <- celltype_nmd_total_persample %>%
  group_by(cell_type) %>%
  summarize(
    mean_pct_to_nmd = mean(pct_to_nmd),
    sd_pct_to_nmd = sd(pct_to_nmd),
    n_samples = n(),
    mean_total_cpm = mean(total_cpm),
    .groups = "drop"
  ) %>%
  rename(pct_to_nmd = mean_pct_to_nmd)

cat("Cell-type transcriptional output to NMD (as % of total cell-type output):\n\n")
print(celltype_nmd_output_total)
cat("\n")

# Visualization
ggplot(celltype_nmd_output_total, aes(x = cell_type, y = pct_to_nmd, fill = cell_type)) +
  geom_col() +
  geom_hline(yintercept = mean(celltype_nmd_output_total$pct_to_nmd),
             linetype = "dashed", color = "black") +
  labs(title = "Figure 17: % of Total Cell-Type Transcriptional Output to NMD",
       subtitle = paste0("Calculated per-sample then averaged. Denominator: ALL transcripts. Dashed line: mean (",
                        round(mean(celltype_nmd_output_total$pct_to_nmd), 2), "%)"),
       x = "Cell Type",
       y = "% of Total Transcriptional Output to NMD",
       fill = "Cell Type") +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "none"
  )

cat("\n**Comparison of Figure 16 vs Figure 17:**\n")
cat("- **Figure 16**: Denominator is CPM sum of transcripts from genes with ≥1 NMD-responsive isoform\n")
cat("  - Shows: Within genes affected by NMD, what % of their output goes to NMD targets?\n")
cat("  - This will always be higher than Figure 17 because the denominator is smaller\n")
cat("- **Figure 17**: Denominator is CPM sum of ALL transcripts in the cell type (should ≈1M per sample)\n")
cat("  - Shows: What % of the cell's entire transcriptional capacity is devoted to NMD targets?\n")
cat("  - Provides a genome-wide perspective on the cost of NMD\n\n")
cat("Both calculations now correctly compute percentages per-sample before averaging across samples.\n\n")

cat("\n**Interpretation:**\n")
cat("This analysis quantifies the 'cost' of NMD - the fraction of transcriptional resources\n")
cat("devoted to producing transcripts that are normally degraded by NMD.\n")
cat("The Smg1i condition reveals this hidden transcriptional output.\n\n")

# Save results
write_tsv(gene_nmd_output,
          "../tmp/isoform_interpretation_q5_gene_nmd_output_2026.1.3.tsv")
write_tsv(celltype_nmd_output,
          "../tmp/isoform_interpretation_q5_celltype_nmd_output_2026.1.3.tsv")
write_tsv(celltype_nmd_output_total,
          "../tmp/isoform_interpretation_q5_celltype_nmd_output_total_2026.1.3.tsv")
cat("Saved: tmp/isoform_interpretation_q5_gene_nmd_output_2026.1.3.tsv\n")
cat("Saved: tmp/isoform_interpretation_q5_celltype_nmd_output_2026.1.3.tsv\n")
cat("Saved: tmp/isoform_interpretation_q5_celltype_nmd_output_total_2026.1.3.tsv\n")
```

# Key Findings

```{r findings}
cat("\n## Summary of Key Findings\n\n")

cat("**Q1: Isoform Diversity**\n")
cat("- NMD inhibition increases detected isoforms per gene by",
    round(mean(q1_summary$mean_delta), 2),
    "on average across cell types\n")
cat("- Average", round(mean(q1_summary$pct_more_tx_smg1i), 1),
    "% of genes show increased isoform diversity\n\n")

cat("**Q2: Dominant Isoforms**\n")
cat("- Dominant isoforms (>50% expression) are present in",
    round(mean(q2_summary$pct_with_dominant), 1),
    "% of genes on average\n")
cat("- NMD inhibition changes dominant isoform prevalence by",
    round(mean(abs(q2_comparison$delta_pct)), 2),
    "percentage points on average\n\n")

cat("**Q3: NMD Target Characteristics**\n")
cat("- Upregulated transcripts had median baseline proportion of",
    round(mean(q3_summary$median_baseline_prop), 3),
    "in DMSO\n")
cat("- Average", round(mean(q3_summary$pct_very_minor), 1),
    "% were very minor isoforms (<10% expression) at baseline\n")
cat("- This supports the hypothesis that NMD targets are minor isoforms\n\n")

cat("**Q4: Cell-Type Specificity**\n")
cat("- Patterns are largely consistent across cell types\n")
cat("- Coefficient of variation for isoform gain:",
    round(sd(q1_summary$mean_delta) / mean(q1_summary$mean_delta), 3), "\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
