---
title: "Gene Count Level Associations w/ Smg1i Cell Treatment (reference DMSO)"
author: "J Santos Cabrera, Pete C"
date: "2025-08-20"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
    number_sections: true
---
This Rmd file can be located: 
/udd/repjc/RESEARCH/NMD/code/NMD_shortread_gene_expression_2025.8.20_PJC.Rmd

Overview:

This report processes, analyzes, and visualizes gene count data alongside clinical metadata to identify significant associations.
It begins by cleaning and reformatting the raw gene expression matrix so that gene IDs are in row names and sample IDs match the clinical dataset.
The clinical data are cleaned by standardizing treatment labels, formatting dates, and defining subject codes for repeated measures.

The workflow then explores the data through descriptive plots, including histograms of log-transformed expression values and PCA plots colored by experimental and subject variables.
Expression and clinical datasets are aligned to a common set of samples, combined into a DGEList, and normalized using the TMM method.

A linear modeling approach is applied using the limma–voom pipeline, accounting for within-subject correlation.
Models include main effects, interaction terms, and continuous covariates, with results saved for reproducibility.
Association tables are generated for key terms, showing significantly differentially expressed genes after multiple testing correction.

Finally, volcano plots are produced to visualize effect sizes and significance levels for both main and interaction effects, highlighting the top significant genes.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(bitmapType='cairo')
```


#### Libraries
```{r}

suppressPackageStartupMessages({
library(data.table)
library(ggplot2)
library(edgeR)
library(limma)
library(DT)
library(EnhancedVolcano)
library(BiocParallel)
library(htmltools)
library(biomaRt)
library(dplyr)
library(ggrepel)
library(gridExtra)
library(parallel)
library(topGO)
library(org.Hs.eg.db) 
})

plot_volcano <- function(df, design_name, coef_name) {
  df <- df %>%
    mutate(
      negLogP = -log10(P.Value),
      sig = ifelse(adj.P.Val < 0.05, "Significant", "Not Sig")
    )
  
  top_hits <- df %>%
    filter(sig == "Significant") %>%
    arrange(P.Value) %>%
    slice_head(n = 10)
  
  ggplot(df, aes(x = logFC, y = negLogP)) +
    geom_point(aes(color = sig), alpha = 0.6, size = 1.5) +
    scale_color_manual(values = c("Significant" = "red", "Not Sig" = "gray")) +
    geom_text_repel(
      data = top_hits,
      aes(label = hgnc_symbol),
      size = 3,
      max.overlaps = 10,
      box.padding = 0.3,
      point.padding = 0.3,
      segment.color = "grey50"
    ) +
    labs(
      title = paste0("Volcano Plot — ", design_name, "\nCoefficient: ", coef_name),
      x = "log2 Fold Change",
      y = expression(-log[10]("P-value"))
    ) +
    theme_minimal(base_size = 14) +
    theme(
      legend.position = "bottom",
      plot.title = element_text(hjust = 0.5)
    )
}


```


# (Step 1) Read & Load Data

```{r}

#### Output Directory ####

out_dir <- "/udd/repjc/RESEARCH/REUSABLE_CODE/rds/"

#### Gene Map #### File for this is located: /udd/rejca/projects/rnaseq/Code/Gene_Mapping_ENSGv114_2025.06.12.Rmd

load(file.path(out_dir, "gmap_ENSGv111_2025.08.12.rda"))




#### Gene Salmon Short Read ####

raw_r_file <- "/proj/regeps/regep00/studies/ExternalCellLines/data/rnaseq/Randell_Lung_Cells_2025/results/nfcore-rnaseq/20250718/star_salmon/salmon.merged.gene_counts_length_scaled.tsv" # for regression: raw necessary for dge List, calcNorm factors, & voom
raw_r1 <- fread(raw_r_file)

raw_r_file2 <- "/proj/regeps/regep00/studies/ExternalCellLines/data/rnaseq/Randell_Lung_Cells_2025/results/nfcore-rnaseq/20250811/star_salmon/salmon.merged.gene_counts_length_scaled.tsv" # for regression: raw necessary for dge List, calcNorm factors, & voom
raw_r2 <- fread(raw_r_file2)

raw_r<-merge(raw_r1,raw_r2[,-"gene_name"],by="gene_id")
dim(raw_r)
names(raw_r)[grep("\\.",names(raw_r))]<-gsub("\\.","_",names(raw_r)[grep("\\.",names(raw_r))])

#### Cell Clinical Variables ####

phe1 <- fread("/udd/repjc/RESEARCH/NMD/NMD_Cell_Lines/pheno/FB_MV_pheno_2025.8.2.csv", data.table=F) 
phe1$name<-sapply(strsplit(phe1$bam,".markdup"),function(x)x[1])
table(phe1$name %in% names(raw_r))

phe2 <- fread("/udd/repjc/RESEARCH/NMD/NMD_Cell_Lines/pheno/DD_DO_AT_pheno_2025.8.15.csv", data.table=F) 
phe2$name<-sapply(strsplit(phe2$bam,".markdup"),function(x)x[1])
phe2$name<-gsub("-","_",phe2$name)
table(phe2$name %in% names(raw_r))
phe2<-phe2[,names(phe2)[!names(phe2) %in% "mapid"]]

phe<-rbind(phe1,phe2)
dim(phe)

table(phe$name %in% names(raw_r))
    
phe$ct<-factor(phe$ct)
phe$ct<-relevel(phe$ct,ref="DD")

phe$treatment<-factor(phe$treatment)
phe$treatment<-relevel(phe$treatment,ref="DMSO")

```




# (Step 2) Format Gene Data 


#### Clean (remove V1) & Arrange Columns and Rows for Gene Data

This chunk converts raw_r (gene data) to a data frame, sets gene_id as row names, removes that column, and prints dimensions. The second chunk cleans and renames expression column IDs to match clinical sample_ids, then shows a before-and-after name table.

```{r}

# Make GENEID rownames in raw

raw_r <- as.data.frame(raw_r)

# Set gene_id as rownames
rownames(raw_r) <- raw_r$gene_id

# Remove gene_id column
raw_r$gene_id <- NULL

cat("\n dimensions of the raw gene dataframe:", dim(raw_r))


```



# (Step 3) PCA and Fit model

Aligns samples between expression and clinical data, subsets both, and creates a DGEList.

```{r}

# Get common sample IDs
common_sid <- intersect(colnames(raw_r), phe$name)

# Subset expression matrix
raw_r <- raw_r[, common_sid]

# Align the order of metadata to match the expression data
phe <- phe[match(common_sid, phe$name),]

cat("Number of common subjects:", length(common_sid), "\n")
cat("Genes dimensions after subsetting columns (samples):", dim(raw_r), "\n")
cat("Phenotype dimensions (samples should match raw_r):", dim(phe), "\n")

gmap<-gmap[match(rownames(raw_r),gmap$ensembl_gene_id_version),]
identical(gmap$ensembl_gene_id_version,rownames(raw_r))

# Create DGEList with metadata
dge <- DGEList(
  counts = raw_r,
  samples = as.data.frame(phe),
  genes = gmap[,c("ensembl_gene_id_version","hgnc_symbol","external_gene_name")]
)

cat("DGEList counts matrix dimensions:", dim(dge$counts), "\n")
cat("DGEList samples dataframe dimensions:", dim(dge$samples), "\n")

design_all<-model.matrix(formula("~sample_id+ct+treatment+ct:treatment"), data = dge$samples)

# ---- Filter lowly expressed genes ----

  keep <- filterByExpr(dge, design = design_all,min.count=20,min.total.count = 50,min.prop=0)
  cat("Genes retained =", sum(keep), "\n")

dge <- dge[keep, , keep.lib.sizes = FALSE]

cat("Genes retained after filtering:", sum(keep), "\n")

# ---- Normalize with TMM ----
dge <- calcNormFactors(dge, method = "TMM")

# Final check
cat("DGEList contains:", nrow(dge$samples), "samples and", nrow(dge$counts), "genes\n")

```

## Diagnostic Plots

Voom Transform the Low Expressed TMM Normalized Gene Counts and Visualize Principle Components (+ pre-fit mean-variance)

```{r}

# voom transform 
v <- voom(dge, design = design_all, plot = TRUE) # initial voom transformation utilized now and later for model fit
pca <- prcomp(t(v$E), scale. = TRUE)
pca_df <- as.data.frame(pca$x[, 1:2])
var_exp <- round(100 * (pca$sdev^2 / sum(pca$sdev^2))[1:2], 1)

# Add metadata
pca_df$treatment <- dge$samples$treatment
pca_df$ct <- dge$samples$ct
pca_df$sample_id <- dge$samples$sample_id
pca_df$color<-ifelse(pca_df$treatment=="DMSO","orange","purple")

# Plot 2: color by time_days, shape by treatment
p2 <- ggplot(pca_df, aes(x = PC1, y = PC2, color=color, shape = ct)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_identity("color") +
  labs(
    title = "Pre-Adjustment PCA: Treatment (color), Cell Type (shape)",
    x = paste0("PC1 (", var_exp[1], "%)"),
    y = paste0("PC2 (", var_exp[2], "%)")
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

print(p2)

```


## Fit Model 

Model: ~ sample_id + treatment 

Treatment: reference is DMSO

Create lmFit Models with using duplicateCorrelation voom (block + correlation) and limma fit (ebayes) (genes are always the response)

Builds a model matrix with main effects and a block for repeated measures. Runs voom again with block + duplicate correlation for weights, fits the model via limma, and mean variance plot after fit of residuals.

```{r}

cat("\nElements from model matrix: ",colnames(design_all))
  
  # Fit linear model with duplicate correlation
  fit <- lmFit(v, design_all)
  fit <- eBayes(fit)

  # Residual SD vs Average Expression
 plotSA(fit, main = paste("Residual SD vs Avg Expression:"))
 


# Step 2: Fix them to be syntactically valid
colnames(fit$design) <- make.names(colnames(fit$design))


# Step 4: IMPORTANT - Update the coefficients names too
colnames(fit$coefficients) <- colnames(fit$design)

# Step 5: Now create the contrast matrix
contrast_matrix <- makeContrasts(
  Smg1i_in_DD = treatmentSmg1i,
  Smg1i_in_DD_ALI = treatmentSmg1i + ctDD_ALI.treatmentSmg1i,
  Smg1i_in_DO_ALI = treatmentSmg1i + ctDO_ALI.treatmentSmg1i,
  Smg1i_in_AT2 = treatmentSmg1i + ctAT2.treatmentSmg1i,
  Smg1i_in_FB = treatmentSmg1i + ctFB.treatmentSmg1i,
  Smg1i_in_MV = treatmentSmg1i + ctMV.treatmentSmg1i,
  levels = fit$design
)


# Step 7: Fit contrasts (should work now without warning)
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)


 
```

# Results

## DD

```{r}
treatment_res<-topTable(fit2, coef = "Smg1i_in_DD", number = 22000)

table(treatment_res$adj.P.Val<=0.05)

table(treatment_res$adj.P.Val<=0.05,treatment_res$logFC>0,dnn=c("FDR sig","logFC>0"))

treatment_res[,c(4:8)]<-round(treatment_res[,c(4:8)],3)

treatment_res<-treatment_res[order(treatment_res$P.Value),]

plot_volcano(treatment_res, "DD", "Smg1i_in_DD")  

datatable(treatment_res[treatment_res$adj.P.Val<=0.05,c("ensembl_gene_id_version","external_gene_name","logFC","AveExpr","adj.P.Val","P.Value")],rownames = FALSE)

# fwrite(treatment_res,file="/udd/repjc/RESEARCH/NMD/results/nmd_dge_fb_2025.9.4.csv")

# fitlist<-list()
# fitlist[["AT2"]]<-fit

```

## DD_ALI

```{r}
treatment_res<-topTable(fit2, coef = "Smg1i_in_DD_ALI", number = 22000)

table(treatment_res$adj.P.Val<=0.05)

table(treatment_res$adj.P.Val<=0.05,treatment_res$logFC>0,dnn=c("FDR sig","logFC>0"))

treatment_res[,c(4:8)]<-round(treatment_res[,c(4:8)],3)

treatment_res<-treatment_res[order(treatment_res$P.Value),]

plot_volcano(treatment_res, "DD_ALI", "Smg1i_in_DD_ALI")  

datatable(treatment_res[treatment_res$adj.P.Val<=0.05,c("ensembl_gene_id_version","external_gene_name","logFC","AveExpr","adj.P.Val","P.Value")],rownames = FALSE)

# fwrite(treatment_res,file="/udd/repjc/RESEARCH/NMD/results/nmd_dge_fb_2025.9.4.csv")

# fitlist<-list()
# fitlist[["AT2"]]<-fit

```

## DO_ALI

```{r}
treatment_res<-topTable(fit2,coef="Smg1i_in_DO_ALI",number = Inf, confint = TRUE, sort.by = "none")

table(treatment_res$adj.P.Val<=0.05)

table(treatment_res$adj.P.Val<=0.05,treatment_res$logFC>0,dnn=c("FDR sig","logFC>0"))

treatment_res[,c(4:8)]<-round(treatment_res[,c(4:8)],3)

treatment_res<-treatment_res[order(treatment_res$P.Value),]

plot_volcano(treatment_res,"DO_ALI" ,"Smg1i_in_DO_ALI")  

datatable(treatment_res[treatment_res$adj.P.Val<=0.05,c("ensembl_gene_id_version","external_gene_name","logFC","AveExpr","adj.P.Val","P.Value")],rownames = FALSE)

# fwrite(treatment_res,file="/udd/repjc/RESEARCH/NMD/results/nmd_dge_fb_2025.9.4.csv")

# fitlist<-list()
# fitlist[["AT2"]]<-fit

```


## AT2

```{r}
treatment_res<-topTable(fit2,coef="Smg1i_in_AT2",number = Inf, confint = TRUE, sort.by = "none")

table(treatment_res$adj.P.Val<=0.05)

table(treatment_res$adj.P.Val<=0.05,treatment_res$logFC>0,dnn=c("FDR sig","logFC>0"))

treatment_res[,c(4:8)]<-round(treatment_res[,c(4:8)],3)

treatment_res<-treatment_res[order(treatment_res$P.Value),]

plot_volcano(treatment_res, "AT2", "Smg1i_in_AT2")  

datatable(treatment_res[treatment_res$adj.P.Val<=0.05,c("ensembl_gene_id_version","external_gene_name","logFC","AveExpr","adj.P.Val","P.Value")],rownames = FALSE)

# fwrite(treatment_res,file="/udd/repjc/RESEARCH/NMD/results/nmd_dge_fb_2025.9.4.csv")

# fitlist<-list()
# fitlist[["AT2"]]<-fit

```


## MV

```{r}
treatment_res<-topTable(fit2,coef="Smg1i_in_MV",number = Inf, confint = TRUE, sort.by = "none")

table(treatment_res$adj.P.Val<=0.05)

table(treatment_res$adj.P.Val<=0.05,treatment_res$logFC>0,dnn=c("FDR sig","logFC>0"))

treatment_res[,c(4:8)]<-round(treatment_res[,c(4:8)],3)

treatment_res<-treatment_res[order(treatment_res$P.Value),]

plot_volcano(treatment_res, "MV", "Smg1i_in_MV")  

datatable(treatment_res[treatment_res$adj.P.Val<=0.05,c("ensembl_gene_id_version","external_gene_name","logFC","AveExpr","adj.P.Val","P.Value")],rownames = FALSE)

# fwrite(treatment_res,file="/udd/repjc/RESEARCH/NMD/results/nmd_dge_fb_2025.9.4.csv")

# fitlist<-list()
# fitlist[["AT2"]]<-fit

```

## FB

```{r}
treatment_res<-topTable(fit2,coef="Smg1i_in_FB",number = Inf, confint = TRUE, sort.by = "none")

table(treatment_res$adj.P.Val<=0.05)

table(treatment_res$adj.P.Val<=0.05,treatment_res$logFC>0,dnn=c("FDR sig","logFC>0"))

treatment_res[,c(4:8)]<-round(treatment_res[,c(4:8)],3)

treatment_res<-treatment_res[order(treatment_res$P.Value),]

plot_volcano(treatment_res, "FB", "Smg1i_in_FB")  

datatable(treatment_res[treatment_res$adj.P.Val<=0.05,c("ensembl_gene_id_version","external_gene_name","logFC","AveExpr","adj.P.Val","P.Value")],rownames = FALSE)

# fwrite(treatment_res,file="/udd/repjc/RESEARCH/NMD/results/nmd_dge_fb_2025.9.4.csv")

# fitlist<-list()
# fitlist[["AT2"]]<-fit

```


# Appendix
```{r}

sessionInfo()

```






