---
title: "Isoform Choice Analysis: Version 6.0"
author: "NMD Lung Cells Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: united
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(tidyverse)
library(knitr)
library(kableExtra)
library(patchwork)

base_dir <- "/Users/petecastaldi/claude_projects/nmd/results/isoform_transitions/Version_6.0"
```

# Overview

**Analysis Framework:** Dominant isoform-centric splicing choice characterization

**Research Question:** What splicing assembly choices do cells make when shifting from dominant to non-dominant isoforms?

---

# Executive Summary

```{r load-data}
# TODO: Load all result files
```

```{r summary-stats}
# TODO: Generate summary statistics
```

---

# Category 1: Co-occurrence Analysis

## Crude Co-occurrence Results

```{r cooccurrence-crude}
# TODO: Load and display crude results
```

## Structure-Controlled Co-occurrence

```{r cooccurrence-adjusted}
# TODO: Load and display adjusted results
```

## Comparison: Crude vs Adjusted

```{r cooccurrence-comparison}
# TODO: Compare crude vs adjusted ORs
```

---

# Category 2: Spatial Organization

## Positional Bias

```{r positional-bias}
# TODO: Load and visualize positional bias results
```

## Topology Enrichment

```{r topology-enrichment}
# TODO: Load and visualize topology results
```

## Proximity Analysis

```{r proximity}
# TODO: Load and visualize proximity results
```

---

# Category 3: Functional Context

## Regional Distribution

```{r regional-distribution}
# TODO: Load and visualize regional enrichment
```

## ORF Impact

```{r orf-impact}
# TODO: Summarize ORF-affecting changes
```

---

# Category 4: Pattern Classification

## Complexity-Event Relationship

```{r complexity-relationship}
# TODO: Load and visualize complexity analysis
```

## Pattern Frequencies

```{r pattern-frequencies}
# TODO: Display pattern frequencies by complexity
```

## Top Patterns

```{r top-patterns}
# TODO: Display most common patterns
```

---

# Conclusions

TODO: Synthesize findings across all categories

---

# Session Info

```{r session-info}
sessionInfo()
```
