---
title: "NMD Isoform Transitions Analysis (mashr Classification)"
subtitle: "Isopair Wrapper — Isocall Pipeline, mashr DIE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(Isopair)
library(dplyr)
library(tibble)
library(tidyr)
library(DT)
library(htmltools)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
if (has_ggplot2) library(ggplot2)
source("visualization_functions.R")

# DT rendering helper for results='asis' chunks
# Uses renderTags() + knit_meta_add() to register JS/CSS dependencies with knitr,
# so widgets render even when they are not the last expression in a chunk.
render_dt <- function(df, caption = NULL, ...) {
  widget <- datatable(df, caption = caption, rownames = FALSE,
                      class = "compact stripe",
                      options = list(pageLength = 25, dom = 'ft'), ...)
  rendered <- htmltools::renderTags(widget)
  knitr::knit_meta_add(rendered$dependencies)
  cat(rendered$html)
  cat("\n")
}

# Caching infrastructure: wrap expensive computations with cached_compute()
cached_compute <- function(key, expr, cache_dir = file.path(data_dir, "analysis_cache"),
                           force_recompute = FALSE) {
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
  cache_file <- file.path(cache_dir, paste0(key, ".rds"))
  if (!force_recompute && file.exists(cache_file)) {
    return(readRDS(cache_file))
  }
  result <- force(expr)
  saveRDS(result, cache_file)
  result
}

# Utility functions
ct_to_label <- function(ct) tolower(gsub("_", "", ct))

evt_types <- c("Alt_TSS", "Alt_TES", "A5SS", "A3SS", "SE",
               "Missing_Internal", "IR", "IR_diff",
               "Partial_IR_5", "Partial_IR_3")

compute_event_freq <- function(profiles) {
  n_pairs <- nrow(profiles)
  n_pir5 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_5")))
  n_pir3 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_3")))
  counts <- c(sum(profiles$tss_changed), sum(profiles$tes_changed),
              sum(profiles$n_a5ss > 0), sum(profiles$n_a3ss > 0),
              sum(profiles$n_se > 0), sum(profiles$n_missing_internal > 0),
              sum(profiles$n_ir > 0), sum(profiles$n_ir_diff > 0),
              n_pir5, n_pir3)
  data.frame(event_type = evt_types,
             n_pairs_with_event = counts,
             pct_of_pairs = round(100 * counts / n_pairs, 1))
}

# Display labels: internal c2/c4 → user-facing NMD/Control
comp_label <- function(comp) {
  switch(toupper(comp), C2 = "NMD", C4 = "Control", toupper(comp))
}

parse_key <- function(key) {
  parts <- strsplit(key, "_")[[1]]
  list(comp = comp_label(parts[1]),
       ct = paste(parts[-1], collapse = "_"))
}

# Count CDS bp in the comparator isoform upstream (5') of the first frameshift boundary.
# Used to determine how much of the protein is conserved vs novel after a frameshift.
count_conserved_cds_bp <- function(comp_id, fs_boundary, strand, cds_df, str_df) {
  cds_row <- cds_df[cds_df$isoform_id == comp_id, ]
  str_row <- str_df[str_df$isoform_id == comp_id, ]
  if (nrow(cds_row) == 0 || nrow(str_row) == 0) return(NA_integer_)
  cds_start <- cds_row$cds_start[1]
  cds_stop  <- cds_row$cds_stop[1]
  orf_len   <- cds_row$orf_length[1]
  e_starts  <- str_row$exon_starts[[1]]
  e_ends    <- str_row$exon_ends[[1]]

  # Order exons 5'→3'
  if (strand == "-") {
    ord <- order(e_starts, decreasing = TRUE)
  } else {
    ord <- order(e_starts)
  }
  e_starts <- e_starts[ord]
  e_ends   <- e_ends[ord]

  conserved <- 0L
  for (i in seq_along(e_starts)) {
    es <- e_starts[i]
    ee <- e_ends[i]
    # Intersect exon with CDS range
    cs <- max(es, cds_start)
    ce <- min(ee, cds_stop)
    if (cs > ce) next  # No CDS overlap

    if (strand == "+") {
      if (fs_boundary <= cs) break
      if (fs_boundary >= ce) {
        conserved <- conserved + (ce - cs + 1L)
      } else {
        conserved <- conserved + (fs_boundary - cs)
        break
      }
    } else {
      if (fs_boundary >= ce) break
      if (fs_boundary <= cs) {
        conserved <- conserved + (ce - cs + 1L)
      } else {
        conserved <- conserved + (ce - fs_boundary)
        break
      }
    }
  }
  conserved <- min(conserved, orf_len)
  as.integer(conserved)
}

# Build frame walk summary row for any profile key
# Restrict frame walk summary to pairs with coding references (frame_resolved not NA)
build_fw_row <- function(key) {
  fw <- set_results[[key]]$frame_walk
  if (is.null(fw) || is.null(fw$profile_summary) || nrow(fw$profile_summary) == 0) return(NULL)
  ps <- fw$profile_summary
  # Restrict to coding pairs: frame_resolved is NA when reference is non-coding
  ps <- ps[!is.na(ps$frame_resolved), ]
  if (nrow(ps) == 0) return(NULL)
  n <- nrow(ps)
  has_shift <- ps$n_frameshifts > 0
  n_shift <- sum(has_shift, na.rm = TRUE)
  n_unres <- sum(has_shift & ps$frame_resolved == FALSE, na.rm = TRUE)
  n_comp <- sum(ps$n_compensatory > 0, na.rm = TRUE)
  tibble::tibble(
    n_pairs = n,
    median_splice_groups = round(median(ps$n_splice_groups, na.rm = TRUE), 1),
    n_with_frameshift = n_shift,
    pct_with_frameshift = round(100 * n_shift / n, 1),
    n_unresolved = n_unres,
    pct_unresolved = round(100 * n_unres / n, 1),
    n_compensated = n_comp,
    pct_compensated = if (n_shift > 0) round(100 * n_comp / n_shift, 1) else 0,
    median_pct_orf_frameshifted = round(
      median(ps$pct_orf_frameshifted[ps$n_frameshifts > 0], na.rm = TRUE), 1)
  )
}

# ---------------------------------------------------------------------------
# Helper functions for rendering/computing per-cell-type results
# Used by both the all_samples main body and the per-CT tabbed section
# ---------------------------------------------------------------------------

# 1. Render event frequency table + summary stats
render_event_frequency <- function(ct) {
  cps <- cross_results[[ct]]
  if (is.null(cps)) return(invisible(NULL))
  ef <- cps$event_frequency
  if (is.null(ef) || isTRUE(ef$skipped)) return(invisible(NULL))

  ef_df <- do.call(rbind, lapply(ef$per_event, function(e) {
    dir_label <- gsub("enriched in A", "enriched in NMD",
                 gsub("enriched in B", "enriched in Control", e$direction))
    data.frame(
      event_type = e$event_type,
      NMD_n = e$count_a, NMD_pct = round(100 * e$count_a / e$n_a, 1),
      Ctrl_n = e$count_b, Ctrl_pct = round(100 * e$count_b / e$n_b, 1),
      OR = round(e$odds_ratio, 2),
      adj_p = e$adj_p_value,
      direction = dir_label
    )
  }))
  render_dt(ef_df, caption = sprintf("Event Type Prevalence: NMD vs Control — %s", ct))

  ec <- cps$event_count
  if (!is.null(ec) && !isTRUE(ec$skipped))
    cat(sprintf("Event count: Wilcoxon p = %s, d = %.3f (NMD median = %.0f, Control = %.0f)\n\n",
                format(ec$p_value, digits = 3), ec$cliffs_d, ec$median_a, ec$median_b))

  pc <- cps$profile_categories
  if (!is.null(pc) && !isTRUE(pc$skipped))
    cat(sprintf("Profile categories: Chi-square p = %s, V = %.3f\n\n",
                format(pc$p_value, digits = 3), pc$cramers_v))

  cooc_cps <- cps$cooccurrence
  if (!is.null(cooc_cps) && !isTRUE(cooc_cps$skipped))
    cat(sprintf("Co-occurrence structure is shared between NMD and Control (r = %.3f).\n\n",
                cooc_cps$correlation))

  pt <- cps$paired_tests
  if (!is.null(pt) && !isTRUE(pt$skipped)) {
    cat(sprintf("Paired analysis (%d gene-matched pairs): Wilcoxon p = %s; sign test: %.0f%% of genes have more events in NMD (p = %s)\n\n",
                pt$n_genes_overlap,
                format(pt$paired_wilcox$p_value, digits = 3),
                pt$sign_test$pct_a_more,
                format(pt$sign_test$p_value, digits = 3)))
  }
}

# 2. Compute structural divergence table
# Per-region length differences from direct isoform pair comparisons (ref - comp)
compute_divergence_row <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  div_c2 <- set_results[[c2_key]]$divergence
  div_c4 <- set_results[[c4_key]]$divergence
  if (is.null(div_c2) || nrow(div_c2) == 0 || is.null(div_c4) || nrow(div_c4) == 0) return(NULL)

  regions <- c("exonic_bp" = "Overall Exonic",
               "utr5_bp"   = "5' UTR",
               "cds_bp"    = "CDS",
               "utr3_bp"   = "3' UTR")
  diff_cols <- paste0(names(regions), "_diff")

  div_tbl <- do.call(rbind, lapply(seq_along(regions), function(i) {
    col <- diff_cols[i]
    ref_col <- paste0(names(regions)[i], "_ref")
    comp_col <- paste0(names(regions)[i], "_comp")

    # For overall exonic, use all pairs; for UTR/CDS, restrict to pairs
    # where both isoforms have >0 bp in that region
    if (col == "exonic_bp_diff") {
      c2_vals <- div_c2[[col]]
      c4_vals <- div_c4[[col]]
    } else {
      c2_mask <- div_c2[[ref_col]] > 0 & div_c2[[comp_col]] > 0
      c4_mask <- div_c4[[ref_col]] > 0 & div_c4[[comp_col]] > 0
      c2_vals <- div_c2[[col]][c2_mask]
      c4_vals <- div_c4[[col]][c4_mask]
    }

    wt_r <- if (length(c2_vals) > 1 && length(c4_vals) > 1)
      tryCatch(wilcox.test(c2_vals, c4_vals), error = function(e) NULL) else NULL

    data.frame(
      region = regions[i],
      NMD_n = length(c2_vals),
      NMD_median_diff = round(median(c2_vals, na.rm = TRUE), 0),
      NMD_mean_diff = round(mean(c2_vals, na.rm = TRUE), 0),
      NMD_pct_comp_shorter = round(100 * mean(c2_vals > 0, na.rm = TRUE), 1),
      Ctrl_n = length(c4_vals),
      Ctrl_median_diff = round(median(c4_vals, na.rm = TRUE), 0),
      Ctrl_mean_diff = round(mean(c4_vals, na.rm = TRUE), 0),
      Ctrl_pct_comp_shorter = round(100 * mean(c4_vals > 0, na.rm = TRUE), 1),
      wilcox_p = if (!is.null(wt_r)) wt_r$p.value else NA_real_
    )
  }))
  rownames(div_tbl) <- NULL
  div_tbl
}

# 3. Compute gain/loss balance row
compute_gain_loss_row <- function(ct) {
  cps <- cross_results[[ct]]
  if (is.null(cps)) return(NULL)
  gl <- cps$gain_loss
  if (is.null(gl) || isTRUE(gl$skipped)) return(NULL)
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  data.frame(
    cell_type = ct,
    NMD_med_net_bp = round(median(set_results[[c2_key]]$profiles$net_bp), 0),
    Ctrl_med_net_bp = round(median(set_results[[c4_key]]$profiles$net_bp), 0),
    wilcox_p = gl$p_value,
    cliffs_d = round(gl$cliffs_d, 3)
  )
}

# 4. Compute gain/loss direction by event type table
compute_gain_loss_direction <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)

  events_c2 <- dplyr::bind_rows(set_results[[c2_key]]$profiles$detailed_events)
  events_c4 <- dplyr::bind_rows(set_results[[c4_key]]$profiles$detailed_events)

  if (nrow(events_c2) == 0 || nrow(events_c4) == 0) return(NULL)
  dir_comp <- do.call(rbind, lapply(
    unique(c(events_c2$event_type, events_c4$event_type)), function(et) {
      e2 <- events_c2[events_c2$event_type == et, ]
      e4 <- events_c4[events_c4$event_type == et, ]
      g2 <- sum(e2$direction == "GAIN"); l2 <- sum(e2$direction == "LOSS")
      g4 <- sum(e4$direction == "GAIN"); l4 <- sum(e4$direction == "LOSS")
      ft <- if ((g2 + l2) > 0 && (g4 + l4) > 0)
        tryCatch(fisher.test(matrix(c(g2, l2, g4, l4), nrow = 2)), error = function(e) NULL) else NULL
      data.frame(
        event_type = et,
        NMD_gain = g2, NMD_loss = l2,
        NMD_pct_gain = round(100 * g2 / max(g2+l2, 1), 1),
        Ctrl_gain = g4, Ctrl_loss = l4,
        Ctrl_pct_gain = round(100 * g4 / max(g4+l4, 1), 1),
        fisher_p = if (!is.null(ft)) ft$p.value else NA_real_
      )
  }))
  dir_comp$adj_p <- p.adjust(dir_comp$fisher_p, "BH")
  dir_comp
}

# 5. Compute IR individual types table
compute_ir_individual <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  # IR_diff aggregates IR_diff_5, IR_diff_3, IR_diff_5_3
  ir_types <- c("IR", "IR_diff", "Partial_IR_5", "Partial_IR_3")

  events_c2 <- dplyr::bind_rows(set_results[[c2_key]]$profiles$detailed_events)
  events_c4 <- dplyr::bind_rows(set_results[[c4_key]]$profiles$detailed_events)

  ir_detail <- do.call(rbind, lapply(ir_types, function(et) {
    if (et == "IR_diff") {
      e2 <- events_c2[grepl("^IR_diff", events_c2$event_type), ]
      e4 <- events_c4[grepl("^IR_diff", events_c4$event_type), ]
    } else {
      e2 <- events_c2[events_c2$event_type == et, ]
      e4 <- events_c4[events_c4$event_type == et, ]
    }
    g2 <- sum(e2$direction == "GAIN"); l2 <- sum(e2$direction == "LOSS")
    g4 <- sum(e4$direction == "GAIN"); l4 <- sum(e4$direction == "LOSS")
    ft <- if ((g2 + l2) > 0 && (g4 + l4) > 0)
      tryCatch(fisher.test(matrix(c(g2, l2, g4, l4), nrow = 2)), error = function(e) NULL) else NULL
    data.frame(
      event_type = et,
      NMD_gain = g2, NMD_loss = l2, NMD_total = g2 + l2,
      NMD_pct_gain = round(100 * g2 / max(g2 + l2, 1), 1),
      Ctrl_gain = g4, Ctrl_loss = l4, Ctrl_total = g4 + l4,
      Ctrl_pct_gain = round(100 * g4 / max(g4 + l4, 1), 1),
      fisher_p = if (!is.null(ft)) ft$p.value else NA_real_
    )
  }))
  ir_detail$adj_p <- p.adjust(ir_detail$fisher_p, "BH")
  ir_detail
}

# 6. Compute cumulative IR burden row
compute_ir_cumulative_row <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  ir_types <- c("IR", "IR_diff", "Partial_IR_5", "Partial_IR_3")

  events_c2 <- dplyr::bind_rows(set_results[[c2_key]]$profiles$detailed_events)
  events_c4 <- dplyr::bind_rows(set_results[[c4_key]]$profiles$detailed_events)
  ir_c2 <- events_c2[events_c2$event_type %in% ir_types, ]
  ir_c4 <- events_c4[events_c4$event_type %in% ir_types, ]

  n_c2 <- nrow(set_results[[c2_key]]$profiles)
  n_c4 <- nrow(set_results[[c4_key]]$profiles)

  ir_per_pair_c2 <- sapply(set_results[[c2_key]]$profiles$detailed_events, function(de)
    sum(de$event_type %in% ir_types))
  ir_per_pair_c4 <- sapply(set_results[[c4_key]]$profiles$detailed_events, function(de)
    sum(de$event_type %in% ir_types))

  wt <- tryCatch(wilcox.test(ir_per_pair_c2, ir_per_pair_c4), error = function(e) NULL)

  g2 <- sum(ir_c2$direction == "GAIN"); l2 <- sum(ir_c2$direction == "LOSS")
  g4 <- sum(ir_c4$direction == "GAIN"); l4 <- sum(ir_c4$direction == "LOSS")
  ft <- if ((g2 + l2) > 0 && (g4 + l4) > 0)
    tryCatch(fisher.test(matrix(c(g2, l2, g4, l4), nrow = 2)), error = function(e) NULL) else NULL

  data.frame(
    cell_type = ct,
    NMD_total_ir = nrow(ir_c2),
    NMD_per_pair = round(nrow(ir_c2) / n_c2, 2),
    NMD_pct_gain = round(100 * g2 / max(g2 + l2, 1), 1),
    Ctrl_total_ir = nrow(ir_c4),
    Ctrl_per_pair = round(nrow(ir_c4) / n_c4, 2),
    Ctrl_pct_gain = round(100 * g4 / max(g4 + l4, 1), 1),
    count_wilcox_p = if (!is.null(wt)) wt$p.value else NA_real_,
    dir_fisher_p = if (!is.null(ft)) ft$p.value else NA_real_
  )
}

# 7. Compute PTC rates row
compute_ptc_rates_row <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)

  coding_ids <- cds$isoform_id[cds$coding_status == "coding"]

  p_c2 <- set_results[[c2_key]]$profiles
  cod_c2 <- p_c2[p_c2$reference_isoform_id %in% coding_ids &
                   p_c2$comparator_isoform_id %in% coding_ids, ]
  p_c4 <- set_results[[c4_key]]$profiles
  cod_c4 <- p_c4[p_c4$reference_isoform_id %in% coding_ids &
                   p_c4$comparator_isoform_id %in% coding_ids, ]
  if (nrow(cod_c2) == 0 || nrow(cod_c4) == 0) return(NULL)

  ptc_c2 <- sum(cod_c2$comparator_isoform_id %in% ptc$isoform_id[ptc$has_ptc == TRUE])
  ptc_c4 <- sum(cod_c4$comparator_isoform_id %in% ptc$isoform_id[ptc$has_ptc == TRUE])
  n_c2 <- nrow(cod_c2); n_c4 <- nrow(cod_c4)

  ft <- tryCatch({
    mat <- matrix(c(ptc_c2, n_c2 - ptc_c2, ptc_c4, n_c4 - ptc_c4), nrow = 2)
    fisher.test(mat)
  }, error = function(e) NULL)

  data.frame(
    cell_type = ct,
    n_coding_pairs = n_c2,
    NMD_comp_ptc_n = ptc_c2,
    NMD_comp_ptc_pct = round(100 * ptc_c2 / n_c2, 1),
    Ctrl_comp_ptc_n = ptc_c4,
    Ctrl_comp_ptc_pct = round(100 * ptc_c4 / n_c4, 1),
    fisher_OR = if (!is.null(ft)) round(ft$estimate, 2) else NA_real_,
    p_value = if (!is.null(ft)) ft$p.value else NA_real_
  )
}

# 8. Compute frame category count rows
compute_frame_category_rows <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  fc_c2 <- set_results[[c2_key]]$frame_comp
  fc_c4 <- set_results[[c4_key]]$frame_comp
  if (is.null(fc_c2) || is.null(fc_c4)) return(NULL)
  ps_c2 <- fc_c2$pair_summary
  ps_c4 <- fc_c4$pair_summary

  cats <- c("same_start_in_frame", "same_start_frameshift",
            "diff_start_same_frame", "diff_start_diff_frame",
            "diff_start_diff_frame_with_frameshift",
            "non_coding", "no_shared_cds")
  do.call(rbind, lapply(cats, function(cat) {
    data.frame(
      cell_type = ct,
      frame_category = cat,
      NMD_n = sum(ps_c2$frame_category == cat),
      NMD_pct = round(100 * sum(ps_c2$frame_category == cat) / nrow(ps_c2), 1),
      Ctrl_n = sum(ps_c4$frame_category == cat),
      Ctrl_pct = round(100 * sum(ps_c4$frame_category == cat) / nrow(ps_c4), 1)
    )
  }))
}

# 9. Compute PTC mechanism rows
compute_ptc_mechanism_rows <- function(ct) {
  coding_cats <- c("same_start_in_frame", "same_start_frameshift",
                   "diff_start_same_frame", "diff_start_diff_frame",
                   "diff_start_diff_frame_with_frameshift")
  ptc_pos_ids <- ptc$isoform_id[ptc$has_ptc == TRUE]
  ct_label <- ct_to_label(ct)

  mech <- do.call(rbind, lapply(c("c2", "c4"), function(comp) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) return(NULL)
    fc <- set_results[[key]]$frame_comp
    if (is.null(fc) || is.null(fc$pair_summary)) return(NULL)
    ps <- fc$pair_summary
    ps_coding <- ps[ps$frame_category %in% coding_cats, ]
    ps_ptc <- ps_coding[ps_coding$comparator_isoform_id %in% ptc_pos_ids, ]
    n_ptc_total <- nrow(ps_ptc)

    do.call(rbind, lapply(coding_cats, function(cat) {
      n_cat <- sum(ps_ptc$frame_category == cat)
      data.frame(
        cell_type = ct, comparison = comp_label(comp),
        frame_category = cat,
        n_ptc = n_cat, ptc_total = n_ptc_total,
        pct_of_ptc = round(100 * n_cat / max(n_ptc_total, 1), 1)
      )
    }))
  }))

  # Also compute frameshift enrichment test
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  fs_test <- NULL
  if (c2_key %in% names(set_results) && c4_key %in% names(set_results)) {
    fc_c2 <- set_results[[c2_key]]$frame_comp
    fc_c4 <- set_results[[c4_key]]$frame_comp
    if (!is.null(fc_c2) && !is.null(fc_c4)) {
      ps_c2 <- fc_c2$pair_summary; ps_c4 <- fc_c4$pair_summary
      n_ss_fs_c2 <- sum(ps_c2$frame_category == "same_start_frameshift")
      n_ss_if_c2 <- sum(ps_c2$frame_category == "same_start_in_frame")
      n_ss_fs_c4 <- sum(ps_c4$frame_category == "same_start_frameshift")
      n_ss_if_c4 <- sum(ps_c4$frame_category == "same_start_in_frame")
      fi <- tryCatch(
        fisher.test(matrix(c(n_ss_fs_c2, n_ss_if_c2, n_ss_fs_c4, n_ss_if_c4), nrow = 2)),
        error = function(e) NULL)
      fs_test <- data.frame(
        cell_type = ct,
        NMD_ss_total = n_ss_fs_c2 + n_ss_if_c2,
        NMD_fs_pct = round(100 * n_ss_fs_c2 / max(n_ss_fs_c2 + n_ss_if_c2, 1), 1),
        Ctrl_ss_total = n_ss_fs_c4 + n_ss_if_c4,
        Ctrl_fs_pct = round(100 * n_ss_fs_c4 / max(n_ss_fs_c4 + n_ss_if_c4, 1), 1),
        OR = if (!is.null(fi)) round(fi$estimate, 2) else NA_real_,
        p_value = if (!is.null(fi)) fi$p.value else NA_real_
      )
    }
  }
  list(mechanism = mech, fs_test = fs_test)
}

# 10. Compute frameshift outcome rows
compute_frameshift_outcome_rows <- function(ct) {
  ct_label <- ct_to_label(ct)
  rows <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) next
    fc <- set_results[[key]]$frame_comp
    if (is.null(fc) || is.null(fc$pair_summary)) next
    ps <- fc$pair_summary
    ss <- ps[ps$frame_category %in% c("same_start_in_frame", "same_start_frameshift"), ]
    n_same_start <- nrow(ss)
    n_in_frame <- sum(ss$frame_category == "same_start_in_frame")
    fs_pairs <- ss[ss$frame_category == "same_start_frameshift", ]
    n_fs_ptc <- 0L; n_fs_stop_last <- 0L; n_fs_other <- 0L
    if (nrow(fs_pairs) > 0) {
      fs_ptc <- merge(fs_pairs, ptc[, c("isoform_id", "has_ptc", "stop_in_last_exon")],
                      by.x = "comparator_isoform_id", by.y = "isoform_id", all.x = TRUE)
      n_fs_ptc       <- sum(fs_ptc$has_ptc == TRUE, na.rm = TRUE)
      no_ptc_mask    <- !is.na(fs_ptc$has_ptc) & fs_ptc$has_ptc == FALSE
      n_fs_stop_last <- sum(no_ptc_mask & fs_ptc$stop_in_last_exon == TRUE, na.rm = TRUE)
      n_fs_other     <- sum(no_ptc_mask & (is.na(fs_ptc$stop_in_last_exon) |
                                            fs_ptc$stop_in_last_exon == FALSE))
    }
    rows[[comp]] <- data.frame(
      cell_type = ct, comparison = comp_label(comp),
      n_same_start = n_same_start, n_in_frame = n_in_frame,
      n_fs_ptc = n_fs_ptc, n_fs_stop_last = n_fs_stop_last, n_fs_other = n_fs_other
    )
  }
  do.call(rbind, rows)
}

# 11. Compute frame-PTC cross-reference rows
compute_frame_ptc_xref_rows <- function(ct) {
  coding_cats <- c("same_start_in_frame", "same_start_frameshift",
                   "diff_start_same_frame", "diff_start_diff_frame",
                   "diff_start_diff_frame_with_frameshift")
  ptc_pos_ids <- ptc$isoform_id[ptc$has_ptc == TRUE]
  ct_label <- ct_to_label(ct)

  do.call(rbind, lapply(c("c2", "c4"), function(comp) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) return(NULL)
    fc <- set_results[[key]]$frame_comp
    if (is.null(fc) || is.null(fc$pair_summary)) return(NULL)
    ps <- fc$pair_summary

    do.call(rbind, lapply(coding_cats, function(cat) {
      cat_pairs <- ps[ps$frame_category == cat, ]
      n_cat <- nrow(cat_pairs)
      n_ptc <- sum(cat_pairs$comparator_isoform_id %in% ptc_pos_ids)
      data.frame(cell_type = ct, comparison = comp_label(comp),
                 frame_category = cat, n_pairs = n_cat, n_ptc = n_ptc,
                 ptc_pct = round(100 * n_ptc / max(n_cat, 1), 1))
    }))
  }))
}

# 12. Render frame walk by event type for one ct
render_frame_walk <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results)) return(invisible(NULL))

  fw_tables <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) next
    fw <- set_results[[key]]$frame_walk
    if (is.null(fw) || nrow(fw$events) == 0L) next
    fw_evts <- fw$events
    fw_tables[[comp]] <- do.call(rbind, lapply(unique(fw_evts$event_type), function(et) {
      e <- fw_evts[fw_evts$event_type == et, ]
      data.frame(event_type = et,
                 n_cds_events = nrow(e),
                 pct_frameshift = round(100 * sum(e$is_frameshift) / nrow(e), 1),
                 pct_compensatory = round(100 * sum(e$is_compensatory) / nrow(e), 1))
    }))
  }

  if (length(fw_tables) == 2) {
    fw_merged <- merge(fw_tables[["c2"]], fw_tables[["c4"]],
                       by = "event_type", suffixes = c("_NMD", "_Ctrl"), all = TRUE)
    render_dt(fw_merged, caption = sprintf("Frame Walk by Event Type: NMD vs Control — %s", ct))
  } else if ("c2" %in% names(fw_tables)) {
    render_dt(fw_tables[["c2"]], caption = sprintf("Frame Walk by Event Type: NMD — %s", ct))
  }
}

# 13. Compute novel protein metrics for one ct (C4 same_start_frameshift, no PTC)
compute_novel_protein <- function(ct) {
  ct_label <- ct_to_label(ct)
  c4_key <- paste0("c4_", ct_label)
  if (!c4_key %in% names(set_results)) return(NULL)
  fc <- set_results[[c4_key]]$frame_comp
  if (is.null(fc)) return(NULL)

  cached_compute(paste0("novel_protein_c4_v2_", ct_label), {
    ps <- fc$pair_summary
    rd <- fc$region_detail
    fs_pairs <- ps[ps$frame_category == "same_start_frameshift", ]
    if (nrow(fs_pairs) == 0) return(NULL)
    fs_merged <- merge(fs_pairs,
                       ptc[, c("isoform_id", "has_ptc", "orf_length")],
                       by.x = "comparator_isoform_id", by.y = "isoform_id", all.x = TRUE)
    fs_merged <- fs_merged[!is.na(fs_merged$has_ptc) & fs_merged$has_ptc == FALSE, ]
    if (nrow(fs_merged) == 0) return(NULL)
    strand_lookup <- setNames(cds$strand, cds$isoform_id)
    results <- lapply(seq_len(nrow(fs_merged)), function(i) {
      row <- fs_merged[i, ]
      comp_id <- row$comparator_isoform_id
      ref_id  <- row$reference_isoform_id
      s <- strand_lookup[comp_id]
      if (is.na(s)) return(NULL)
      pair_rd <- rd[rd$reference_isoform_id == ref_id &
                    rd$comparator_isoform_id == comp_id, ]
      if (nrow(pair_rd) == 0) return(NULL)
      div_regions <- pair_rd[pair_rd$ref_phase_at_start != pair_rd$comp_phase_at_start, ]
      if (nrow(div_regions) == 0) return(NULL)
      if (s == "+") { fs_boundary <- min(div_regions$region_start)
      } else { fs_boundary <- max(div_regions$region_end) }
      conserved_bp <- count_conserved_cds_bp(comp_id, fs_boundary, s, cds, structures)
      if (is.na(conserved_bp)) return(NULL)
      comp_orf <- row$orf_length
      novel_bp <- max(0L, comp_orf - conserved_bp)
      data.frame(
        gene_id = row$gene_id, comparator_isoform_id = comp_id,
        orf_length = comp_orf, conserved_bp = conserved_bp,
        novel_bp = novel_bp, novel_aa = novel_bp / 3,
        total_aa = comp_orf / 3, pct_novel = 100 * (novel_bp / 3) / (comp_orf / 3)
      )
    })
    do.call(rbind, results)
  })
}
```

# 1. Setup

This report analyzes structural differences between isoform pairs using the Isopair package,
with NMD classification derived from **mashr** (multivariate adaptive shrinkage) differential
isoform expression results. Mashr logFC values are shrunken relative to limma, but NMD
classification uses the pre-computed `nmd_responsive` flag (adj.P.Val < 0.05 & logFC > 0).
Main analyses cover all_samples (union NMD / intersection non-NMD) plus AT, DD, DO, FB, and MV.
DD_ALI is analyzed separately (Section 7) due to reversed biological patterns.

```{r load-data}
data_dir <- "data_mashr"

# Infrastructure
structures     <- readRDS(file.path(data_dir, "structures.rds"))
union_exons    <- readRDS(file.path(data_dir, "union_exons.rds"))
ue_mapping     <- readRDS(file.path(data_dir, "isoform_union_mapping.rds"))
cds            <- readRDS(file.path(data_dir, "cds.rds"))
cds_exons      <- readRDS(file.path(data_dir, "cds_exons.rds"))
annotated_ue   <- readRDS(file.path(data_dir, "annotated_ue.rds"))
ptc            <- readRDS(file.path(data_dir, "ptc.rds"))

# Study data
expr_mat       <- readRDS(file.path(data_dir, "expression_data.rds"))
sample_meta    <- readRDS(file.path(data_dir, "sample_metadata.rds"))
gene_map       <- readRDS(file.path(data_dir, "gene_map.rds"))
nmd_class      <- readRDS(file.path(data_dir, "nmd_classification.rds"))

# Build gene_id → gene_symbol lookup from mashr DE files
de_dir_mashr <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/mashr"
gene_symbol_lookup <- character(0)
for (de_f in list.files(de_dir_mashr, pattern = "^nmd_mashr_die_.*\\.csv$", full.names = TRUE)) {
  de_tmp <- read.csv(de_f, stringsAsFactors = FALSE)[, c("gene_id", "hgnc_symbol")]
  de_tmp <- de_tmp[!duplicated(de_tmp$gene_id) & !is.na(de_tmp$hgnc_symbol) &
                    de_tmp$hgnc_symbol != "", ]
  new_entries <- setdiff(de_tmp$gene_id, names(gene_symbol_lookup))
  gene_symbol_lookup[de_tmp$gene_id[de_tmp$gene_id %in% new_entries]] <-
    de_tmp$hgnc_symbol[de_tmp$gene_id %in% new_entries]
}

# Cell types with profiles (mashr: DO included, DD_ALI excluded from main)
cell_types <- c("all_samples", "AT", "DD", "DO", "FB", "MV")
comparisons <- c("c2", "c4")

# DD_ALI separation: main cell types for cross-CT analyses and meta-analysis
main_cts <- c("AT", "DD", "DO", "FB", "MV")
ali_ct <- "DD_ALI"
limma_data_dir <- "data"

# Load all available profiles
profiles_list <- list()
pairs_list    <- list()
for (ct in cell_types) {
  ct_label <- tolower(gsub("_", "", ct))
  for (comp in comparisons) {
    key <- paste(comp, ct_label, sep = "_")
    prof_file <- file.path(data_dir, sprintf("profiles_%s.rds", key))
    pair_file <- file.path(data_dir, sprintf("pairs_%s.rds", key))
    if (file.exists(prof_file)) {
      profiles_list[[key]] <- readRDS(prof_file)
      if (file.exists(pair_file)) pairs_list[[key]] <- readRDS(pair_file)
    }
  }
}

# Load DD_ALI profiles from original limma data directory for the DD_ALI section
for (comp in comparisons) {
  ali_label <- tolower(gsub("_", "", ali_ct))
  key <- paste(comp, ali_label, sep = "_")
  prof_file <- file.path(limma_data_dir, sprintf("profiles_%s.rds", key))
  pair_file <- file.path(limma_data_dir, sprintf("pairs_%s.rds", key))
  if (file.exists(prof_file)) {
    profiles_list[[key]] <- readRDS(prof_file)
    if (file.exists(pair_file)) pairs_list[[key]] <- readRDS(pair_file)
  }
}

cat(sprintf("Loaded %d profile sets across %d cell types (including DD_ALI from limma)\n",
            length(profiles_list), length(cell_types) + 1L))
```

```{r gene-match}
# Bidirectional gene-matching: restrict NMD and Control to the intersection of
# (gene_id, reference_isoform_id) pairs present in BOTH sets. After this step,
# every gene in NMD has a corresponding Control pair and vice versa.
gene_match_summary <- list()
# Include DD_ALI in gene-matching (loaded from limma) for the dedicated section
for (ct in unique(c(cell_types, ali_ct))) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(profiles_list) || !c4_key %in% names(profiles_list)) next

  nmd_before <- nrow(profiles_list[[c2_key]])
  control_before <- nrow(profiles_list[[c4_key]])

  # Intersection of (gene_id, reference_isoform_id)
  c2_gene_refs <- unique(profiles_list[[c2_key]][, c("gene_id", "reference_isoform_id")])
  c4_gene_refs <- unique(profiles_list[[c4_key]][, c("gene_id", "reference_isoform_id")])
  shared_refs <- inner_join(c2_gene_refs, c4_gene_refs, by = c("gene_id", "reference_isoform_id"))

  profiles_list[[c2_key]] <- semi_join(profiles_list[[c2_key]], shared_refs,
                                        by = c("gene_id", "reference_isoform_id"))
  profiles_list[[c4_key]] <- semi_join(profiles_list[[c4_key]], shared_refs,
                                        by = c("gene_id", "reference_isoform_id"))

  gene_match_summary[[ct]] <- data.frame(
    cell_type = ct,
    nmd_before = nmd_before,
    control_before = control_before,
    matched_genes = nrow(shared_refs),
    nmd_after = nrow(profiles_list[[c2_key]]),
    control_after = nrow(profiles_list[[c4_key]])
  )
}
gene_match_df <- do.call(rbind, gene_match_summary)
```

# 2. Data Flow

```{r data-flow-numbers}
# --- Gather numbers ---
n_isoforms <- nrow(expr_mat)
n_samples <- ncol(expr_mat)
n_dmso <- sum(sample_meta$treatment == "DMSO")
n_smg1i <- sum(sample_meta$treatment == "Smg1i")
n_genes_struct <- n_distinct(structures$gene_id)
coding_n <- sum(cds$coding_status == "coding")
unknown_n <- sum(cds$coding_status == "unknown")

nmd_summary <- do.call(rbind, lapply(names(nmd_class), function(ct) {
  data.frame(cell_type = ct,
             n_nmd = length(nmd_class[[ct]]$nmd),
             n_non_nmd = length(nmd_class[[ct]]$non_nmd))
}))
total_nmd <- length(unique(unlist(lapply(nmd_class, `[[`, "nmd"))))
total_non_nmd <- length(unique(unlist(lapply(nmd_class, `[[`, "non_nmd"))))

total_nmd_before <- sum(gene_match_df$nmd_before)
total_ctrl_before <- sum(gene_match_df$control_before)
total_nmd_after <- sum(gene_match_df$nmd_after)
total_ctrl_after <- sum(gene_match_df$control_after)
total_matched_genes <- sum(gene_match_df$matched_genes)
```

```{r data-flow-chart}
# --- Flowchart (Graphviz via DiagrammeR) ---
fmt <- function(x) format(x, big.mark = ",")

dot_spec <- sprintf('
digraph flowchart {
  graph [rankdir=TB, fontname="Helvetica", nodesep=1.6, ranksep=1.4, size="13,14"]
  node  [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=23, margin="0.65,0.5", width=6.5]
  edge  [color="#666666", penwidth=2.5]

  A [label=<<b>Isoform Expression</b><br/>%s isoforms × %d samples<br/>%d DMSO + %d Smg1i | 6 cell types>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  B [label=<<b>NMD Classification (mashr)</b><br/>NMD-sensitive: %s isoforms<br/>Non-NMD: %s isoforms>,
     fillcolor="#fff7bc", color="#d95f0e"]

  C [label=<<b>Pair Construction</b><br/>%s NMD pairs | %s Control pairs>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  D [label=<<b>Gene-Matching</b><br/>%s shared genes<br/>%s NMD | %s Control pairs>,
     fillcolor="#c7e9c0", color="#31a354"]

  A -> B -> C -> D
}',
  fmt(n_isoforms), n_samples, n_dmso, n_smg1i,
  fmt(total_nmd), fmt(total_non_nmd),
  fmt(total_nmd_before), fmt(total_ctrl_before),
  fmt(total_matched_genes), fmt(total_nmd_after), fmt(total_ctrl_after)
)

if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(dot_spec, width = 1000, height = 700)
} else {
  message("DiagrammeR not installed — flowchart not rendered")
}
```

## NMD Classification

Isoforms are classified per cell type using mashr (multivariate adaptive shrinkage)
differential isoform expression results. NMD-sensitive isoforms are identified by the
pre-computed `nmd_responsive` flag (adj.P.Val < 0.05 & logFC > 0, with mashr-shrunken
effect sizes). Non-NMD: adj.P.Val > 0.50 (no evidence of NMD regulation). Note: mashr
logFC values are shrunken relative to limma, but the NMD/non-NMD classification thresholds
are identical.

**Columns:** **cell_type** = cell type used for DE testing; **n_nmd** = number of isoforms classified as NMD-sensitive in that cell type; **n_non_nmd** = number of isoforms classified as non-NMD.

```{r nmd-classification-table, results='asis'}
render_dt(nmd_summary[nmd_summary$cell_type == "all_samples", ],
          caption = "NMD Classification Summary (all_samples)")
```

## Gene-Matching Detail

To enable paired NMD vs Control comparisons, we restrict both sets to genes where the same
reference isoform appears in both NMD and Control pairs. This ensures that any differences
between NMD and Control comparators are not confounded by different reference isoforms or gene sets.

**Columns:** **nmd_before/control_before** = pair counts before matching; **matched_genes** = genes with shared (gene_id, reference_isoform_id) in both sets; **nmd_after/control_after** = pair counts after restricting to matched genes.

```{r gene-matching-table, results='asis'}
render_dt(gene_match_df[gene_match_df$cell_type == "all_samples", ],
          caption = "Gene-Matching Summary (all_samples)")
```

## Profile Sets (after gene-matching)

```{r profile-sets-compute}
# Count splice events per pair: independent splicing decisions, computed by
# grouping non-terminal events that share a splice junction (union-find).
# A single splice event (one donor/acceptor choice) can produce multiple
# sequence events (e.g., skipping 3 exons = 1 splice event, 3 sequence events).
count_splice_events <- function(events) {
  if (is.null(events) || nrow(events) == 0) return(0L)
  # Exclude terminal events — these are transcription boundary differences, not splicing
  splice_types <- c("A5SS", "A3SS", "SE", "Missing_Internal", "IR",
                     "IR_diff", "IR_diff_5", "IR_diff_3", "IR_diff_5_3",
                     "Partial_IR_5", "Partial_IR_3")
  se <- events[events$event_type %in% splice_types, ]
  n <- nrow(se)
  if (n == 0L) return(0L)
  if (n == 1L) return(1L)

  # Union-find: group events sharing a splice junction
  parent <- seq_len(n)
  find <- function(x) {
    while (parent[x] != x) { parent[x] <<- parent[parent[x]]; x <- parent[x] }
    x
  }
  union_nodes <- function(a, b) {
    ra <- find(a); rb <- find(b)
    if (ra != rb) parent[ra] <<- rb
  }

  parse_jxns <- function(jxn_str) {
    if (is.na(jxn_str) || jxn_str == "") return(character(0))
    unlist(strsplit(jxn_str, ","))
  }

  jxn_map <- list()
  for (i in seq_len(n)) {
    all_jxns <- unique(c(
      parse_jxns(se$ref_junctions[i]),
      parse_jxns(se$comp_junctions[i])
    ))
    for (jxn in all_jxns) {
      if (jxn %in% names(jxn_map)) union_nodes(i, jxn_map[[jxn]])
      jxn_map[[jxn]] <- i
    }
  }
  # Events with no junctions (empty ref_junctions AND comp_junctions) each form their own group
  length(unique(vapply(seq_len(n), find, integer(1))))
}

prof_overview <- do.call(rbind, lapply(names(profiles_list), function(key) {
  p <- profiles_list[[key]]
  parts <- strsplit(key, "_")[[1]]

  # Per-pair splice event counts
  splice_events_per_pair <- vapply(p$detailed_events, count_splice_events, integer(1))

  data.frame(comparison = comp_label(parts[1]),
             cell_type = paste(parts[-1], collapse = "_"),
             n_profiles = nrow(p),
             n_genes = n_distinct(p$gene_id),
             total_sequence_events = sum(p$n_events),
             mean_sequence_events = round(mean(p$n_events), 2),
             total_splice_events = sum(splice_events_per_pair),
             mean_splice_events = round(mean(splice_events_per_pair), 2),
             pct_recon_pass = round(100 * mean(p$reconstruction_status == "PASS",
                                                na.rm = TRUE), 1))
}))
```

Each profile set represents isoform pairs analyzed for structural differences using
Isopair (`buildProfiles()` in `Isopair/R/profiles.R`; see METHODS.md §Event Detection
Algorithms). NMD pairs compare an NMD-sensitive comparator to a non-NMD reference;
Control pairs compare the next-best non-NMD isoform to the same reference.

The table reports two levels of event counting:

- A **sequence event** is an individual structural difference detected between two
  isoforms — e.g., a single skipped exon (SE), an alternative 5' splice site (A5SS),
  or a retained intron (IR). Each row in the `detailed_events` table is one sequence
  event. This is the finest-grained unit of structural difference.
- A **splice event** is an independent splicing decision — a single, different choice of
  donor and/or acceptor between the two isoforms. One splice event can produce multiple
  sequence events. For example, skipping three consecutive exons involves one splice
  decision (the donor of the upstream exon joins the acceptor of the downstream exon)
  but creates three Missing_Internal sequence events. Splice events are computed by
  grouping non-terminal sequence events that share a splice junction (union-find
  algorithm; see METHODS.md §Splice Group Assignment). Terminal events (Alt_TSS,
  Alt_TES) are excluded because they reflect transcription boundary differences, not
  splicing decisions.

**Columns:** **comparison** = NMD or Control; **cell_type** = cell type (or all_samples
for union/intersection); **n_profiles** = number of isoform pairs; **n_genes** = unique
genes represented; **total_sequence_events / mean_sequence_events** = sum and mean of
individual structural differences per pair; **total_splice_events / mean_splice_events** =
sum and mean of independent splicing decisions per pair; **pct_recon_pass** = % of pairs
where the comparator isoform can be perfectly reconstructed from the reference plus
detected events (quality control metric; 100% expected).

```{r profile-sets-table, results='asis'}
render_dt(prof_overview[prof_overview$cell_type == "allsamples", ],
          caption = "Profile Set Summary (all_samples, gene-matched)")
```

## Coding Status

NMD is a translation-dependent pathway: only protein-coding transcripts with an open reading
frame can harbour a premature termination codon. We therefore expect NMD-sensitive isoform pairs
to be overwhelmingly coding. This table characterizes the coding status of reference and
comparator isoforms in each profile set, based on SQANTI CDS annotations. A small fraction of
isoforms have unknown coding status (no CDS annotation in SQANTI); these are excluded from
downstream analyses that require reading-frame information (e.g., frameshift analysis).

```{r coding-status}
coding_tbl <- do.call(rbind, lapply(names(profiles_list), function(key) {
  p <- profiles_list[[key]]
  parts <- strsplit(key, "_")[[1]]
  comp <- comp_label(parts[1])
  ct <- paste(parts[-1], collapse = "_")

  ref_ids <- unique(p$reference_isoform_id)
  cmp_ids <- unique(p$comparator_isoform_id)

  ref_coding <- sum(ref_ids %in% cds$isoform_id[cds$coding_status == "coding"])
  cmp_coding <- sum(cmp_ids %in% cds$isoform_id[cds$coding_status == "coding"])

  # Pairs where both are coding
  ref_status <- cds$coding_status[match(p$reference_isoform_id, cds$isoform_id)]
  cmp_status <- cds$coding_status[match(p$comparator_isoform_id, cds$isoform_id)]
  both_coding <- sum(ref_status == "coding" & cmp_status == "coding", na.rm = TRUE)

  data.frame(
    Comparison = comp,
    Cell_Type = ct,
    N_Pairs = nrow(p),
    Ref_Coding = ref_coding,
    Ref_Pct = round(100 * ref_coding / length(ref_ids), 1),
    Comp_Coding = cmp_coding,
    Comp_Pct = round(100 * cmp_coding / length(cmp_ids), 1),
    Both_Coding = both_coding,
    Both_Pct = round(100 * both_coding / nrow(p), 1)
  )
}))
```

**Columns:** **Comparison** = NMD or Control; **Cell_Type** = cell type; **N_Pairs** = total
isoform pairs; **Ref_Coding / Ref_Pct** = number and percentage of unique reference isoforms
with a CDS annotation; **Comp_Coding / Comp_Pct** = same for comparator isoforms;
**Both_Coding / Both_Pct** = pairs where both reference and comparator are coding.

```{r coding-status-table, results='asis'}
render_dt(coding_tbl[coding_tbl$Cell_Type == "allsamples", ],
          caption = "Coding Status of Reference and Comparator Isoforms (all_samples)")
```

## Pair Subset Overview

The analyses in this report operate on different subsets of the isoform pair universe.
The flowcharts below show how NMD and Control pairs are progressively subdivided and
which report sections draw on each subset. Counts are for the all_samples set.

```{r pair-subset-flowchart, fig.width=16, fig.height=16}
# Combined NMD + Control pair subset flowchart
fmt <- function(x) format(x, big.mark = ",")

# Helper: compute subset counts for one profile set
compute_subset_counts <- function(profiles) {
  n_all <- nrow(profiles)

  ref_st <- cds$coding_status[match(profiles$reference_isoform_id, cds$isoform_id)]
  cmp_st <- cds$coding_status[match(profiles$comparator_isoform_id, cds$isoform_id)]
  n_both_coding <- sum(ref_st == "coding" & cmp_st == "coding", na.rm = TRUE)
  n_non_coding  <- n_all - n_both_coding

  fc <- tryCatch(
    compareIsoformFrames(profiles, cds_exons, cds),
    error = function(e) NULL
  )
  if (is.null(fc)) return(NULL)
  ps_ov <- fc$pair_summary

  ptc_pos_ids <- ptc$isoform_id[ptc$has_ptc == TRUE]
  count_ptc <- function(cat_name) {
    sum(ps_ov$comparator_isoform_id[ps_ov$frame_category == cat_name] %in% ptc_pos_ids)
  }

  n_ss_if  <- sum(ps_ov$frame_category == "same_start_in_frame")
  n_ss_fs  <- sum(ps_ov$frame_category == "same_start_frameshift")
  n_ds_sf  <- sum(ps_ov$frame_category == "diff_start_same_frame")
  n_ds_df  <- sum(ps_ov$frame_category == "diff_start_diff_frame")
  n_ds_dfs <- sum(ps_ov$frame_category == "diff_start_diff_frame_with_frameshift")
  n_nc     <- sum(ps_ov$frame_category == "non_coding")
  n_nsc    <- sum(ps_ov$frame_category == "no_shared_cds")

  list(
    all = n_all, coding = n_both_coding, non_coding = n_non_coding,
    same_start = n_ss_if + n_ss_fs, diff_start = n_ds_sf + n_ds_df + n_ds_dfs,
    no_shared_cds = n_nsc,
    ss_if = n_ss_if, ss_fs = n_ss_fs,
    ds_sf = n_ds_sf, ds_df = n_ds_df, ds_dfs = n_ds_dfs,
    fs_ptc = count_ptc("same_start_frameshift"),
    fs_no_ptc = n_ss_fs - count_ptc("same_start_frameshift"),
    if_ptc = count_ptc("same_start_in_frame"),
    if_no_ptc = n_ss_if - count_ptc("same_start_in_frame"),
    sf_ptc = count_ptc("diff_start_same_frame"),
    sf_no_ptc = n_ds_sf - count_ptc("diff_start_same_frame"),
    df_ptc = count_ptc("diff_start_diff_frame"),
    df_no_ptc = n_ds_df - count_ptc("diff_start_diff_frame"),
    dfs_ptc = count_ptc("diff_start_diff_frame_with_frameshift"),
    dfs_no_ptc = n_ds_dfs - count_ptc("diff_start_diff_frame_with_frameshift")
  )
}

# Helper: format a dual-count label line
dual <- function(n_nmd, n_ctrl) {
  sprintf('<font color="#d95f0e">NMD: %s</font>  |  <font color="#3182bd">Ctrl: %s</font>',
          fmt(n_nmd), fmt(n_ctrl))
}

as_c2_key <- "c2_allsamples"
as_c4_key <- "c4_allsamples"
dot_combined <- NULL

if (requireNamespace("DiagrammeR", quietly = TRUE) &&
    as_c2_key %in% names(profiles_list) &&
    as_c4_key %in% names(profiles_list)) {

  nmd  <- compute_subset_counts(profiles_list[[as_c2_key]])
  ctrl <- compute_subset_counts(profiles_list[[as_c4_key]])

  # After gene-matching, NMD and Control have the same number of pairs
  n_matched <- nmd$all
  stopifnot(nmd$all == ctrl$all)

  if (!is.null(nmd) && !is.null(ctrl)) {

    # Build PTC subdivision nodes and edges for terminal categories
    ptc_nodes <- ""
    ptc_edges <- ""
    ptc_rank_items <- c("FPTC", "FNPTC")

    # Same-start in-frame → PTC subdivision (if either has PTC > 0)
    if (nmd$if_ptc > 0 || ctrl$if_ptc > 0) {
      ptc_nodes <- paste0(ptc_nodes, sprintf('
  IFPTC  [label=<<b>Same Frame + PTC</b><br/>%s>, fillcolor="#fcbba1", color="#cb181d"]
  IFNPTC [label=<<b>Same Frame, No PTC</b><br/>%s>, fillcolor="#c6dbef", color="#3182bd"]
', dual(nmd$if_ptc, ctrl$if_ptc), dual(nmd$if_no_ptc, ctrl$if_no_ptc)))
      ptc_edges <- paste0(ptc_edges, "  IF -> IFPTC\n  IF -> IFNPTC\n")
      ptc_rank_items <- c(ptc_rank_items, "IFPTC", "IFNPTC")
    }
    # Diff-start same-frame → PTC subdivision
    if (nmd$sf_ptc > 0 || ctrl$sf_ptc > 0) {
      ptc_nodes <- paste0(ptc_nodes, sprintf('
  SFPTC  [label=<<b>Same Frame + PTC</b><br/>%s>, fillcolor="#fcbba1", color="#cb181d"]
  SFNPTC [label=<<b>Same Frame, No PTC</b><br/>%s>, fillcolor="#c6dbef", color="#3182bd"]
', dual(nmd$sf_ptc, ctrl$sf_ptc), dual(nmd$sf_no_ptc, ctrl$sf_no_ptc)))
      ptc_edges <- paste0(ptc_edges, "  FP -> SFPTC\n  FP -> SFNPTC\n")
      ptc_rank_items <- c(ptc_rank_items, "SFPTC", "SFNPTC")
    }
    # Diff-start diff-frame → PTC subdivision
    if (nmd$df_ptc > 0 || ctrl$df_ptc > 0) {
      ptc_nodes <- paste0(ptc_nodes, sprintf('
  DFPTC  [label=<<b>Diff Frame + PTC</b><br/>%s>, fillcolor="#fcbba1", color="#cb181d"]
  DFNPTC [label=<<b>Diff Frame, No PTC</b><br/>%s>, fillcolor="#c6dbef", color="#3182bd"]
', dual(nmd$df_ptc, ctrl$df_ptc), dual(nmd$df_no_ptc, ctrl$df_no_ptc)))
      ptc_edges <- paste0(ptc_edges, "  FD -> DFPTC\n  FD -> DFNPTC\n")
      ptc_rank_items <- c(ptc_rank_items, "DFPTC", "DFNPTC")
    }
    # Diff-start diff-frame + FS → PTC subdivision
    if (nmd$dfs_ptc > 0 || ctrl$dfs_ptc > 0) {
      ptc_nodes <- paste0(ptc_nodes, sprintf('
  DFSPTC  [label=<<b>Diff Frame + FS + PTC</b><br/>%s>, fillcolor="#fcbba1", color="#cb181d"]
  DFSNPTC [label=<<b>Diff Frame + FS, No PTC</b><br/>%s>, fillcolor="#c6dbef", color="#3182bd"]
', dual(nmd$dfs_ptc, ctrl$dfs_ptc), dual(nmd$dfs_no_ptc, ctrl$dfs_no_ptc)))
      ptc_edges <- paste0(ptc_edges, "  FDFS -> DFSPTC\n  FDFS -> DFSNPTC\n")
      ptc_rank_items <- c(ptc_rank_items, "DFSPTC", "DFSNPTC")
    }

    ptc_rank <- sprintf("  {rank=same; %s}", paste(ptc_rank_items, collapse = "; "))

    dot_combined <- sprintf('
digraph pair_subsets {
  graph [rankdir=TB, fontname="Helvetica", nodesep=0.6, ranksep=0.9,
         size="16,20", label=<<b>Pair Subset Overview (all_samples)</b><br/>
         <font color="#d95f0e">NMD (C2)</font> | <font color="#3182bd">Control (C4)</font>>,
         labelloc=t, labeljust=c, fontsize=24]
  node  [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=20,
         margin="0.35,0.25", width=3.5]
  edge  [color="#666666", penwidth=2.0, fontname="Helvetica", fontsize=14]

  ALL [label=<<b>Gene-Matched Pairs</b><br/>%s pairs<br/><font point-size="16">(same genes, same references)</font>>,
       fillcolor="#e8f4f8", color="#2c7fb8"]

  COD [label=<<b>Both-Coding Pairs</b><br/>%s>,
       fillcolor="#c7e9c0", color="#31a354"]
  NC  [label=<<b>Non-Coding</b><br/>%s<br/><font point-size="15">(one/both lack CDS)</font>>,
       fillcolor="#f0f0f0", color="#999999"]

  SS  [label=<<b>Same Start Site</b><br/>%s>,
       fillcolor="#fff7bc", color="#d95f0e"]
  DS  [label=<<b>Different Start Site</b><br/>%s>,
       fillcolor="#fff7bc", color="#d95f0e"]
  NSC [label=<<b>No Shared CDS</b><br/>%s>,
       fillcolor="#f0f0f0", color="#999999"]

  IF  [label=<<b>Same Frame</b><br/>%s>,
       fillcolor="#c6dbef", color="#3182bd"]
  FS  [label=<<b>Frameshift</b><br/>%s>,
       fillcolor="#fcbba1", color="#cb181d"]
  FP  [label=<<b>Same Frame</b><br/>%s>,
       fillcolor="#c6dbef", color="#3182bd"]
  FD  [label=<<b>Diff Frame</b><br/>%s>,
       fillcolor="#fff7bc", color="#d95f0e"]
  FDFS [label=<<b>Diff Frame + FS</b><br/>%s>,
       fillcolor="#fcbba1", color="#cb181d"]

  FPTC  [label=<<b>Frameshift + PTC</b><br/>%s>,
         fillcolor="#fcbba1", color="#cb181d"]
  FNPTC [label=<<b>Frameshift, No PTC</b><br/>%s<br/><font point-size="15">(novel proteins)</font>>,
         fillcolor="#dadaeb", color="#6a51a3"]

  %s

  node [shape=plaintext, style="", fontsize=20, fontcolor="#555555"]

  A_ALL [label=<<b>Used by:</b> Event Frequency, Gain/Loss,<br/>IR, Divergence, Co-occurrence,<br/>Heterogeneity, Meta-analysis<br/><font point-size="17">(Sections 4-6)</font>>]
  A_COD [label=<<b>Used by:</b> PTC Rates,<br/>Frame Comparison<br/><font point-size="17">(Section 3)</font>>]
  A_SS  [label=<<b>Used by:</b> Splicing-Induced<br/>Frameshift Enrichment test<br/><font point-size="17">(Section 3)</font>>]
  A_FS  [label=<<b>Used by:</b> Frameshift Outcome,<br/>Frame-PTC Cross-Reference<br/><font point-size="17">(Section 3)</font>>]
  A_PTC [label=<<b>Used by:</b> PTC Distance vs logFC,<br/>GAM models<br/><font point-size="17">(Section 3)</font>>]
  A_NP  [label=<<b>Used by:</b> Novel Protein<br/>Analysis<br/><font point-size="17">(Section 3)</font>>]

  ALL -> COD
  ALL -> NC

  COD -> SS
  COD -> DS
  COD -> NSC

  SS -> IF
  SS -> FS

  DS -> FP
  DS -> FD
  DS -> FDFS

  FS -> FPTC
  FS -> FNPTC

  %s

  edge [style=dashed, color="#aaaaaa", penwidth=1.0, arrowhead=none]
  ALL -> A_ALL
  COD -> A_COD
  SS  -> A_SS
  FS  -> A_FS
  FPTC -> A_PTC
  FNPTC -> A_NP

  {rank=same; ALL; A_ALL}
  {rank=same; COD; NC; A_COD}
  {rank=same; SS; DS; NSC; A_SS}
  {rank=same; IF; FS; FP; FD; FDFS; A_FS}
  %s
}',
    fmt(n_matched),
    dual(nmd$coding, ctrl$coding),
    dual(nmd$non_coding, ctrl$non_coding),
    dual(nmd$same_start, ctrl$same_start),
    dual(nmd$diff_start, ctrl$diff_start),
    dual(nmd$no_shared_cds, ctrl$no_shared_cds),
    dual(nmd$ss_if, ctrl$ss_if),
    dual(nmd$ss_fs, ctrl$ss_fs),
    dual(nmd$ds_sf, ctrl$ds_sf),
    dual(nmd$ds_df, ctrl$ds_df),
    dual(nmd$ds_dfs, ctrl$ds_dfs),
    dual(nmd$fs_ptc, ctrl$fs_ptc),
    dual(nmd$fs_no_ptc, ctrl$fs_no_ptc),
    ptc_nodes,
    ptc_edges,
    ptc_rank
    )
  }
}
```

```{r pair-subset-combined, fig.width=16, fig.height=16}
if (!is.null(dot_combined)) DiagrammeR::grViz(dot_combined, width = 1400, height = 1200)
```

```{r compute-all, include=FALSE}
# =============================================================================
# Silent compute chunk: all cached_compute() calls + comparePairSets
# No output — results stored in set_results / cross_results for display sections
# =============================================================================
set_results <- list()
cross_results <- list()

for (key in names(profiles_list)) {
  profiles <- profiles_list[[key]]
  message(sprintf("Computing analyses for %s...", key))

  prof_summary <- summarizeProfiles(profiles)
  divergence <- cached_compute(paste0("div_", key),
    tryCatch(quantifyPairDivergence(profiles, structures, cds), error = function(e) NULL))
  regional_impact <- cached_compute(paste0("ri_", key),
    tryCatch(quantifyRegionalImpact(profiles, annotated_ue, cds), error = function(e) NULL))
  cooc <- cached_compute(paste0("cooc_", key),
    tryCatch(testCooccurrence(profiles), error = function(e) NULL))
  event_regions <- cached_compute(paste0("er_", key),
    tryCatch(mapEventsToRegions(profiles, annotated_ue, cds), error = function(e) NULL))
  ptc_assoc <- cached_compute(paste0("ptc_", key),
    tryCatch(testPtcAssociation(ptc, profiles), error = function(e) NULL))
  frame_walk <- cached_compute(paste0("fw_", key),
    tryCatch(analyzeFrameWalk(profiles, cds, ptc), error = function(e) NULL))
  frame_comp <- cached_compute(paste0("fc_", key),
    tryCatch(compareIsoformFrames(profiles, cds_exons, cds), error = function(e) NULL))

  # Dependent computations (not cached separately — fast from cached event_regions)
  orf_bound <- NULL; orf_impact <- NULL
  if (!is.null(event_regions) && nrow(event_regions) > 0) {
    orf_bound <- tryCatch(testOrfBoundarySusceptibility(event_regions, annotated_ue),
                          error = function(e) NULL)
    orf_impact <- tryCatch(summarizeOrfImpact(event_regions, cds),
                           error = function(e) NULL)
  }

  set_results[[key]] <- list(
    profiles = profiles, prof_summary = prof_summary, divergence = divergence,
    regional_impact = regional_impact,
    cooc = cooc, event_regions = event_regions, orf_bound = orf_bound,
    orf_impact = orf_impact, ptc_assoc = ptc_assoc, frame_walk = frame_walk,
    frame_comp = frame_comp, evt_freq = compute_event_freq(profiles)
  )
}

# comparePairSets per cell type (cached, reused by display sections + meta-analysis)
# Include DD_ALI for the dedicated section (profiles loaded from limma data)
all_cts_for_cps <- unique(c(cell_types, ali_ct))
for (ct in all_cts_for_cps) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(profiles_list) || !c4_key %in% names(profiles_list)) next

  message(sprintf("Computing NMD vs Control comparison for %s...", ct))
  ptc_c2 <- ptc[ptc$isoform_id %in% profiles_list[[c2_key]]$comparator_isoform_id, ]
  ptc_c4 <- ptc[ptc$isoform_id %in% profiles_list[[c4_key]]$comparator_isoform_id, ]

  cross_results[[ct]] <- cached_compute(
    paste0("cps_", ct_label),
    tryCatch(
      comparePairSets(profiles_list[[c2_key]], profiles_list[[c4_key]],
                       ptc_a = ptc_c2, ptc_b = ptc_c4,
                       paired = TRUE, verbose = FALSE),
      error = function(e) {
        message(sprintf("comparePairSets failed for %s: %s", ct, e$message))
        NULL
      }
    )
  )
}

# Cell types available for display (DD_ALI excluded — dedicated section below)
all_cts_with_profiles <- unique(c(cell_types, ali_ct))
display_cts <- setdiff(
  all_cts_with_profiles[sapply(all_cts_with_profiles, function(ct) {
    ct_label <- ct_to_label(ct)
    any(paste0(c("c2_", "c4_"), ct_label) %in% names(set_results))
  })],
  ali_ct
)
per_ct <- setdiff(display_cts, "all_samples")
c2_per_ct <- per_ct[sapply(per_ct, function(ct)
  paste0("c2_", ct_to_label(ct)) %in% names(set_results))]

# Main cell types for cross-CT analyses (excluding DD_ALI outlier)
main_per_ct <- intersect(c2_per_ct, main_cts)
ali_in_c2 <- ali_ct %in% c2_per_ct

message("All computations complete.")
```


# 3. Event Type Glossary {.tabset}

Isopair detects 12 structural event types organized into 4 categories. Each tab below
shows a schematic diagram of the event alongside its definition. In all diagrams, the
**Reference** (top) is the dominant non-NMD isoform and the **Comparator** (bottom) is
the isoform being compared. Events are always described from the comparator's perspective:
a **GAIN** means the comparator has additional sequence relative to the reference, and a
**LOSS** means the comparator is missing sequence present in the reference.

```{r sec4-glossary, results='asis', fig.width=8, fig.height=3}
if (has_ggplot2) {
  # Synthetic schematic helper — creates a simple pair plot for glossary
  glossary_plot <- function(ref_s, ref_e, comp_s, comp_e, title, subtitle,
                            ref_cds = NULL, comp_cds = NULL, events = NULL) {
    plot_ucsc_pair(ref_s, ref_e, comp_s, comp_e, strand = "+",
                   ref_cds = ref_cds, comp_cds = comp_cds,
                   ref_label = "Reference", comp_label = "Comparator",
                   events = events, title = title, subtitle = subtitle)
  }

  # ---- Terminal Events ----

  cat("## Alt_TSS\n\n")
  cat("**Alternative Transcription Start Site.** The two isoforms begin at different\n")
  cat("genomic positions. Detected when the 5'-most exon of one isoform does not overlap\n")
  cat("the 5'-most exon of the other. Direction: GAIN if the comparator's TSS is further\n")
  cat("upstream (extending the transcript); LOSS if it is further downstream (truncating it).\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
    comp_s = c(250, 400, 700), comp_e = c(350, 550, 850),
    title = "Alt_TSS: Alternative Transcription Start Site",
    subtitle = "Comparator starts at a different position (here, downstream = LOSS)"
  )
  print(p); cat("\n\n")

  cat("## Alt_TES\n\n")
  cat("**Alternative Transcription End Site.** The two isoforms end at different\n")
  cat("genomic positions. Detected when the 3'-most exon of one isoform does not overlap\n")
  cat("the 3'-most exon of the other. Direction: GAIN if the comparator extends further;\n")
  cat("LOSS if it ends earlier.\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 900),
    comp_s = c(100, 400, 700), comp_e = c(200, 550, 800),
    title = "Alt_TES: Alternative Transcription End Site",
    subtitle = "Comparator ends earlier (LOSS)"
  )
  print(p); cat("\n\n")

  # ---- Boundary Events ----

  cat("## A5SS\n\n")
  cat("**Alternative 5' Splice Site.** An internal exon has a different 5' (donor)\n")
  cat("boundary between the two isoforms while sharing the same 3' boundary. The exon\n")
  cat("is extended or truncated at its 5' end. Direction: GAIN if the comparator's exon\n")
  cat("is longer; LOSS if shorter.\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
    comp_s = c(100, 350, 700), comp_e = c(200, 550, 850),
    title = "A5SS: Alternative 5' Splice Site",
    subtitle = "Comparator exon extends further at its 5' end (GAIN)"
  )
  print(p); cat("\n\n")

  cat("## A3SS\n\n")
  cat("**Alternative 3' Splice Site.** An internal exon has a different 3' (acceptor)\n")
  cat("boundary between the two isoforms while sharing the same 5' boundary. The exon\n")
  cat("is extended or truncated at its 3' end. Direction: GAIN if the comparator's exon\n")
  cat("is longer; LOSS if shorter.\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
    comp_s = c(100, 400, 700), comp_e = c(200, 600, 850),
    title = "A3SS: Alternative 3' Splice Site",
    subtitle = "Comparator exon extends further at its 3' end (GAIN)"
  )
  print(p); cat("\n\n")

  cat("## Partial_IR_5\n\n")
  cat("**Partial Intron Retention (5' end).** An intron is partially retained: one isoform's\n")
  cat("exon extends into the intron from its 5' side, covering what is intronic sequence in\n")
  cat("the other isoform. The 3' boundary of the retaining exon falls within the intron.\n")
  cat("Direction: GAIN if the comparator retains; LOSS if the reference retains.\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
    comp_s = c(100, 400, 700), comp_e = c(200, 620, 850),
    title = "Partial_IR_5: Partial Intron Retention (5' side)",
    subtitle = "Comparator exon extends into the intron from the 5' end (GAIN)"
  )
  print(p); cat("\n\n")

  cat("## Partial_IR_3\n\n")
  cat("**Partial Intron Retention (3' end).** An intron is partially retained: one isoform's\n")
  cat("exon extends into the intron from its 3' side, covering what is intronic sequence in\n")
  cat("the other isoform. The 5' boundary of the retaining exon falls within the intron.\n")
  cat("Direction: GAIN if the comparator retains; LOSS if the reference retains.\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
    comp_s = c(100, 400, 640), comp_e = c(200, 550, 850),
    title = "Partial_IR_3: Partial Intron Retention (3' side)",
    subtitle = "Comparator exon extends into the intron from the 3' end (GAIN)"
  )
  print(p); cat("\n\n")

  # ---- Exon-Level Events ----

  cat("## SE (Skipped Exon)\n\n")
  cat("**Skipped Exon.** An exon present in one isoform is completely absent from the\n")
  cat("other. The flanking exons splice directly to each other, skipping over the exon.\n")
  cat("Direction: GAIN if the comparator includes the exon (reference skips it);\n")
  cat("LOSS if the comparator skips it (reference includes it).\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 400, 600, 800), ref_e  = c(200, 500, 700, 900),
    comp_s = c(100, 400, 800),       comp_e = c(200, 500, 900),
    title = "SE: Skipped Exon",
    subtitle = "Comparator skips an internal exon present in the reference (LOSS)"
  )
  print(p); cat("\n\n")

  cat("## Missing_Internal\n\n")
  cat("**Missing Internal Exon.** Similar to SE but arising from a multi-exon skip:\n")
  cat("one isoform is missing one or more internal exons that are present in the other.\n")
  cat("Often co-occurs with other Missing_Internal events from the same splice decision.\n")
  cat("Direction: GAIN if the comparator has the exon; LOSS if the comparator lacks it.\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 350, 550, 750, 950), ref_e  = c(200, 450, 650, 850, 1050),
    comp_s = c(100, 950),                 comp_e = c(200, 1050),
    title = "Missing_Internal: Missing Internal Exon(s)",
    subtitle = "Comparator skips multiple internal exons (LOSS × 3)"
  )
  print(p); cat("\n\n")

  # ---- Intron Retention Events ----

  cat("## IR\n\n")
  cat("**Full Intron Retention.** One isoform has a single exon that spans the region\n")
  cat("covered by two or more exons (plus the intervening intron) in the other isoform.\n")
  cat("The retaining exon boundaries match exactly. This is a full retention where the\n")
  cat("entire intron is included in the retaining isoform's exon.\n")
  cat("Direction: GAIN if the comparator retains the intron; LOSS if the reference retains it.\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
    comp_s = c(100, 400),       comp_e = c(200, 850),
    title = "IR: Full Intron Retention",
    subtitle = "Comparator retains the intron — single exon spans what are two exons + intron in ref (GAIN)"
  )
  print(p); cat("\n\n")

  cat("## IR_diff_5\n\n")
  cat("**Intron Retention with 5' Boundary Difference.** Like IR, but the retaining\n")
  cat("exon's 5' boundary differs from the first overlapping exon in the other isoform.\n")
  cat("The intron is retained, but the exon boundary is shifted at the 5' end.\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
    comp_s = c(100, 370),       comp_e = c(200, 850),
    title = "IR_diff_5: Intron Retention with 5' Boundary Difference",
    subtitle = "Comparator retains the intron, but 5' boundary of retaining exon is shifted (GAIN)"
  )
  print(p); cat("\n\n")

  cat("## IR_diff_3\n\n")
  cat("**Intron Retention with 3' Boundary Difference.** Like IR, but the retaining\n")
  cat("exon's 3' boundary differs from the last overlapping exon in the other isoform.\n")
  cat("The intron is retained, but the exon boundary is shifted at the 3' end.\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
    comp_s = c(100, 400),       comp_e = c(200, 880),
    title = "IR_diff_3: Intron Retention with 3' Boundary Difference",
    subtitle = "Comparator retains the intron, but 3' boundary of retaining exon is shifted (GAIN)"
  )
  print(p); cat("\n\n")

  cat("## IR_diff_5_3\n\n")
  cat("**Intron Retention with Both Boundary Differences.** Like IR, but both the 5'\n")
  cat("and 3' boundaries of the retaining exon differ from the flanking exons in the\n")
  cat("other isoform. The intron is retained but the exon boundaries are shifted at\n")
  cat("both ends.\n\n")
  p <- glossary_plot(
    ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
    comp_s = c(100, 370),       comp_e = c(200, 880),
    title = "IR_diff_5_3: Intron Retention with Both Boundary Differences",
    subtitle = "Comparator retains the intron with shifted boundaries at both ends (GAIN)"
  )
  print(p); cat("\n\n")
}
```

In the IR section of this report, **IR_diff** refers to the aggregate of IR_diff_5,
IR_diff_3, and IR_diff_5_3 events — intron retentions where exon boundaries do not
match exactly.

# 4. Structural Differences

At baseline, non-NMD isoform pairs share ~50--58% of their exonic sequence and differ by a
median of ~700--800bp. IR and terminal events are the most prevalent event types, and the
next-best (comparator) isoform is typically shorter than the dominant (positive net_bp). This
represents the natural structural diversity between co-expressed isoforms from the same gene.

## Event Frequency

Prevalence of each splicing event type, measured as the percentage of isoform pairs containing
at least one instance of that event. NMD and Control are compared using Fisher's exact test per event type.

**Columns:** **event_type** = splicing event; **NMD_n/Ctrl_n** = number of pairs with the event; **NMD_pct/Ctrl_pct** = % of pairs; **OR** = odds ratio (>1 = more prevalent in NMD); **adj_p** = BH-adjusted p-value; **direction** = which group is enriched.

```{r sec4-event-freq, results='asis', fig.height=8}
render_event_frequency("all_samples")
```

## Structural Divergence

Length differences between reference and comparator isoforms, broken down by transcript
region. All differences are computed directly from the isoform pair as
`reference_bp − comparator_bp`, so **positive values** mean the comparator is shorter
(lost sequence) and **negative values** mean the comparator is longer (gained sequence).
For 5' UTR, CDS, and 3' UTR rows, only pairs where both isoforms have >0 bp in that
region are included (n column).

**Columns:** **region** = transcript region; **NMD_n/Ctrl_n** = number of pairs with
data for that region; **NMD_median_diff/Ctrl_median_diff** = median bp difference
(ref − comp); **NMD_mean_diff/Ctrl_mean_diff** = mean bp difference (ref − comp);
**NMD_pct_comp_shorter/Ctrl_pct_comp_shorter** = % of pairs where the
comparator is shorter; **wilcox_p** = Wilcoxon rank-sum test comparing NMD vs Control
differences.

**Note:** Per-region medians do not sum to the Overall Exonic median for two reasons:
(1) the median is not an additive statistic, and (2) the 5' UTR, CDS, and 3' UTR rows
are each computed on a filtered subset of pairs where both isoforms have >0 bp in that
region (see n columns), whereas Overall Exonic uses all pairs. The mean columns are
provided as a complementary summary since means are additive on the same set of pairs,
though they are more sensitive to outliers.

```{r sec4-divergence, results='asis'}
div_rows <- compute_divergence_row("all_samples")
if (!is.null(div_rows)) {
  knitr::kable(div_rows, row.names = FALSE,
               caption = "Structural Divergence by Region: NMD vs Control") |> print()
  cat("\n\n")
}
```

## Gain/Loss Balance

Net base pair change between reference and comparator isoforms. Computed as
`net_bp = bp_loss - bp_gain`, where LOSS and GAIN are from the comparator's perspective.
**Positive net_bp** means the comparator has lost more sequence than it gained (comparator is
shorter than the reference). **Negative net_bp** means the comparator has gained more sequence
(comparator is longer than the reference).

**Columns:** **NMD_med_net_bp/Ctrl_med_net_bp** = median net bp change; **wilcox_p** = Wilcoxon rank-sum test; **cliffs_d** = Cliff's delta effect size.

```{r sec4-gain-loss, results='asis'}
gl_rows <- compute_gain_loss_row("all_samples")
if (!is.null(gl_rows)) {
  knitr::kable(gl_rows, caption = "Net BP Change: NMD vs Control") |> print()
  cat("\n\n")
}
```

## Gain/Loss Direction by Event Type

For each event type, the direction indicates whether the comparator has GAINED or LOST that
feature relative to the reference. GAIN means the comparator has additional sequence (e.g., an
extra exon); LOSS means the comparator is missing sequence present in the reference.

**Columns:** **NMD_gain/NMD_loss** = event counts by direction in NMD pairs; **NMD_pct_gain** = % of events that are gains; **Ctrl_gain/Ctrl_loss/Ctrl_pct_gain** = same for Control; **fisher_p** = Fisher's exact test; **adj_p** = BH-adjusted p-value.

```{r sec4-gain-loss-direction, results='asis'}
dir_comp <- compute_gain_loss_direction("all_samples")
if (!is.null(dir_comp)) {
  render_dt(dir_comp, caption = "Gain/Loss Direction: NMD vs Control — all_samples")
}
```

## Intron Retention Analysis

Intron retention (IR) is the most prevalent event type at baseline, making it the dominant
structural difference between co-expressed isoforms from the same gene. IR events are detected
when one isoform has a single exon that spans the region covered by two or more exons (plus
the intervening intron) in the other isoform. Long-read sequencing may capture more IR than
short-read studies, since retained introns extend transcript length and are difficult to
assemble from short reads.

This section examines IR events (including IR, IR_diff, Partial_IR_5, Partial_IR_3) in
detail. **Direction reminder:** A GAIN means the comparator isoform has *retained* an intron
that is spliced out in the reference; a LOSS means the comparator has *lost* a retained intron
present in the reference (i.e., the comparator splices it out).

### IR Event Direction: Individual Types

GAIN/LOSS breakdown for each IR-family event type, stratified by NMD and Control.

```{r sec4-ir-individual, results='asis'}
ir_types <- c("IR", "IR_diff", "Partial_IR_5", "Partial_IR_3")

ir_detail <- compute_ir_individual("all_samples")
if (!is.null(ir_detail)) {
  knitr::kable(ir_detail,
               caption = "IR Event Direction: NMD vs Control — all_samples") |> print()
  cat("\n\n")
}
```

### Cumulative IR Burden

Total IR-family events (IR + IR_diff + Partial_IR_5 + Partial_IR_3) aggregated across all
subtypes, comparing NMD and Control. This measures the overall intron retention burden
per pair and whether the direction (GAIN vs LOSS) differs between comparison groups.

```{r sec4-ir-cumulative, results='asis', fig.height=8}
ir_cumul_rows <- compute_ir_cumulative_row("all_samples")

if (!is.null(ir_cumul_rows)) {
  cat("**Columns:** **NMD_total_ir/Ctrl_total_ir** = total IR-family events; ")
  cat("**NMD_per_pair/Ctrl_per_pair** = mean IR events per isoform pair; ")
  cat("**NMD_pct_gain/Ctrl_pct_gain** = % of IR events that are GAINs (comparator retains intron); ")
  cat("**count_wilcox_p** = Wilcoxon test on per-pair IR event counts (NMD vs Control); ")
  cat("**dir_fisher_p** = Fisher's exact test on GAIN vs LOSS proportions.\n\n")
  knitr::kable(ir_cumul_rows, caption = "Cumulative IR Burden: NMD vs Control") |> print()
  cat("\n\n")
}
```

## Distribution Plots

```{r sec4-dist-plots, results='asis', fig.height=6}
if (has_ggplot2) {
  # pct_shared from divergence — all_samples only
  pct_shared_data <- {
    rows <- list()
    for (comp in c("c2", "c4")) {
      key <- paste0(comp, "_allsamples")
      if (!key %in% names(set_results)) next
      div <- set_results[[key]]$divergence
      if (is.null(div) || nrow(div) == 0) next
      rows[[comp]] <- data.frame(comparison = comp_label(comp),
                                 pct_shared = div$pct_shared)
    }
    do.call(rbind, rows)
  }

  if (!is.null(pct_shared_data) && nrow(pct_shared_data) > 0) {
    p1 <- ggplot(pct_shared_data[!is.na(pct_shared_data$pct_shared), ],
                 aes(x = pct_shared, fill = comparison)) +
      geom_density(alpha = 0.5) +
      theme_minimal() +
      labs(title = "Sequence Shared (%): NMD vs Control (all_samples)",
           subtitle = "NMD shifted right = more similar (targeted changes)",
           x = "pct_shared", y = "Density", fill = "")
    print(p1)
    cat("\n\n")
  }

}
```

# 5. PTC & Reading Frame

```{r sec5-ptc-compute}
# Compute PTC rates for all_samples (used by inline references and display chunk below)
ptc_rows <- compute_ptc_rates_row("all_samples")
if (!is.null(ptc_rows)) rownames(ptc_rows) <- NULL
ptc_or_range <- if (!is.null(ptc_rows)) {
  or_vals <- ptc_rows$fisher_OR[!is.na(ptc_rows$fisher_OR)]
  sprintf("OR %s", round(or_vals[1]))
} else {
  "OR N/A"
}
ctrl_ptc_pcts <- if (!is.null(ptc_rows)) {
  round(ptc_rows$Ctrl_comp_ptc_pct[1])
} else {
  5
}
```

PTC enrichment (`r ptc_or_range`) is the single strongest signal distinguishing NMD-sensitive isoform
pairs from baseline structural variation. At baseline, ~`r ctrl_ptc_pcts`% of non-NMD comparator isoforms
contain a PTC. Reading frame is compared independently for each isoform at shared CDS regions,
distinguishing splicing-induced frameshifts (same start codon) from alternative initiation
(different start codons). NMD departs dramatically from Control in both PTC prevalence and
the consequences of frame disruption.

## PTC Rates

Premature termination codons (PTCs) are stop codons located >50 nucleotides upstream of the
last exon-exon junction, a position that triggers nonsense-mediated decay. This table compares
PTC prevalence in NMD comparator isoforms vs Control comparator isoforms, restricted to
**coding pairs** (both reference and comparator have a SQANTI CDS annotation). Non-coding
isoforms cannot harbour a PTC and are excluded from both the numerator and denominator.

PTC status is determined per-isoform using each isoform's own CDS (`computePtcStatus()` in
`Isopair/R/ptc.R`; see METHODS.md §Reading Frame and PTC Analysis). The 50-nt rule is
applied to each isoform's stop codon relative to its own last exon-exon junction.

**Columns:** **n_coding_pairs** = pairs where both reference and comparator are coding;
**NMD_comp_ptc_n / NMD_comp_ptc_pct** = number and % of NMD comparators with a PTC
(among coding pairs); **Ctrl_comp_ptc_n / Ctrl_comp_ptc_pct** = same for Control;
**fisher_OR** = odds ratio from Fisher's exact test (>1 = enriched in NMD);
**p_value** = Fisher's exact test p-value.

```{r sec5-ptc-rates, results='asis'}
if (!is.null(ptc_rows)) {
  knitr::kable(ptc_rows, row.names = FALSE,
               caption = "PTC Rates: NMD vs Control Comparators (coding pairs only)") |> print()
  cat("\n\n")
}
```

## Reading Frame Comparison

For each isoform pair, we compare the reading frame at shared CDS regions using each
isoform's own CDS annotation independently (`compareIsoformFrames()` in
`Isopair/R/ptc.R`; see METHODS.md §Reading Frame and PTC Analysis). CDS exon intervals
are extracted per isoform from the SQANTI GFF3 (`extractCdsExons()` in
`Isopair/R/annotate-regions.R`). For each pair, the function identifies genomic intervals
where both isoforms have CDS coverage, then computes the CDS phase (cumulative coding
bp mod 3) independently for each isoform at the 5' end of each shared interval.

The analysis first determines whether the two isoforms share the same start codon
(translation initiation site, with tolerance = 0), then evaluates whether shared CDS
regions are translated in the same reading frame. Every pair is assigned to exactly one
of six categories:

### Frame Category Counts

Each pair is classified into one of the following categories. The first four apply to
coding pairs with overlapping CDS; the last two are excluded from frame-specific analyses.

- **same_start_in_frame**: Both isoforms begin translation at the same position. All
  shared CDS regions are read in the same phase. Splicing differences between these
  isoforms affect sequence content but preserve the reading frame.
- **same_start_frameshift**: Both isoforms begin translation at the same position, but
  a splicing difference (e.g., a non-divisible-by-3 exon skip or boundary shift) causes
  downstream shared CDS to be read in a different phase. This is a splicing-induced
  frameshift.
- **diff_start_same_frame**: The two isoforms use different translation initiation
  sites, but shared CDS regions are read in the same phase (the start codon offset is a
  multiple of 3, or compensating splicing restores phase). No frameshift is implied.
- **diff_start_diff_frame**: The two isoforms use different translation initiation
  sites, and shared CDS regions are read in different phases with a constant phase offset.
  The frame difference is entirely attributable to the initiation site difference — all
  internal splicing differences are multiples of 3 (no frameshift).
- **diff_start_diff_frame_with_frameshift**: The two isoforms use different translation
  initiation sites, and the phase offset between them changes across shared CDS regions.
  This indicates a splicing-induced frameshift on top of the start codon offset.
- **non_coding**: One or both isoforms lack a CDS annotation (SQANTI coding_status ≠
  "coding"). These pairs cannot be evaluated for reading frame.
- **no_shared_cds**: Both isoforms are coding, but their CDS regions do not overlap
  genomically (e.g., entirely different exons are coding). No shared interval exists for
  phase comparison.

**Columns:** **cell_type** = cell type; **frame_category** = classification (see above);
**NMD_n / NMD_pct** = count and percentage of NMD (C2) pairs in this category;
**Ctrl_n / Ctrl_pct** = count and percentage of Control (C4) pairs. Percentages are
computed over all pairs in the respective set (including non_coding and no_shared_cds).

```{r sec5-frame-comp-detail, results='asis'}
# Detailed breakdown of frame categories — all_samples only
fc_detail <- compute_frame_category_rows("all_samples")
if (!is.null(fc_detail)) {
  widget <- datatable(fc_detail, caption = "Frame Category Counts: NMD vs Control (all_samples)",
                      rownames = FALSE, class = "compact stripe",
                      options = list(pageLength = nrow(fc_detail), dom = 'ft'))
  rendered <- htmltools::renderTags(widget)
  knitr::knit_meta_add(rendered$dependencies)
  cat(rendered$html)
  cat("\n")
}
```

```{r sec5-frame-comp-barchart, results='asis', fig.width=14, fig.height=6}
if (!is.null(fc_detail) && has_ggplot2) {
  coding_cats <- c("same_start_in_frame", "same_start_frameshift",
                   "diff_start_same_frame", "diff_start_diff_frame",
                   "diff_start_diff_frame_with_frameshift")
  cat_labels <- c(
    same_start_in_frame = "Same Start\nSame Frame",
    same_start_frameshift = "Same Start\nFrameshift",
    diff_start_same_frame = "Diff Start\nSame Frame",
    diff_start_diff_frame = "Diff Start\nDiff Frame",
    diff_start_diff_frame_with_frameshift = "Diff Start\nDiff Frame + FS"
  )

  ptc_pos_ids <- ptc$isoform_id[ptc$has_ptc == TRUE]

  # Build stacked data: PTC+ and No PTC per category × comparison (all_samples only)
  stacked_df <- do.call(rbind, lapply(c("c2", "c4"), function(comp) {
    key <- paste0(comp, "_allsamples")
    fc <- set_results[[key]]$frame_comp
    if (is.null(fc)) return(NULL)
    ps <- fc$pair_summary
    total_pairs <- nrow(set_results[[key]]$profiles)
    comp_lbl <- if (comp == "c2") "NMD" else "Control"

    do.call(rbind, lapply(coding_cats, function(cat) {
      cat_pairs <- ps[ps$frame_category == cat, ]
      n_cat <- nrow(cat_pairs)
      n_ptc <- sum(cat_pairs$comparator_isoform_id %in% ptc_pos_ids)
      n_no_ptc <- n_cat - n_ptc
      pct_ptc <- 100 * n_ptc / max(total_pairs, 1)
      pct_no_ptc <- 100 * n_no_ptc / max(total_pairs, 1)

      rbind(
        data.frame(frame_category = cat,
                   category_label = cat_labels[cat],
                   comparison = comp_lbl, ptc_status = "PTC+",
                   n = n_ptc, pct = pct_ptc),
        data.frame(frame_category = cat,
                   category_label = cat_labels[cat],
                   comparison = comp_lbl, ptc_status = "No PTC",
                   n = n_no_ptc, pct = pct_no_ptc)
      )
    }))
  }))

  stacked_df$category_label <- factor(stacked_df$category_label,
    levels = cat_labels[coding_cats])
  stacked_df$ptc_status <- factor(stacked_df$ptc_status, levels = c("PTC+", "No PTC"))

  stacked_df$fill_group <- paste(stacked_df$comparison, stacked_df$ptc_status, sep = " - ")
  stacked_df$fill_group <- factor(stacked_df$fill_group,
    levels = c("NMD - No PTC", "NMD - PTC+", "Control - No PTC", "Control - PTC+"))

  bar_totals <- aggregate(cbind(n, pct) ~ frame_category + category_label + comparison,
                          data = stacked_df, FUN = sum)

  # Fisher's test per category
  p_labels <- do.call(rbind, lapply(coding_cats, function(cat) {
    n_c2 <- fc_detail$NMD_n[fc_detail$frame_category == cat]
    n_c4 <- fc_detail$Ctrl_n[fc_detail$frame_category == cat]
    total_c2 <- sum(fc_detail$NMD_n)
    total_c4 <- sum(fc_detail$Ctrl_n)
    ft <- tryCatch({
      mat <- matrix(c(n_c2, total_c2 - n_c2, n_c4, total_c4 - n_c4), nrow = 2)
      fisher.test(mat)
    }, error = function(e) list(p.value = NA))
    p <- ft$p.value
    stars <- if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else "ns"
    max_pct <- max(bar_totals$pct[bar_totals$frame_category == cat], na.rm = TRUE)
    data.frame(category_label = cat_labels[cat],
               p_star = stars, max_pct = max_pct)
  }))
  p_labels$category_label <- factor(p_labels$category_label,
    levels = cat_labels[coding_cats])

  p_bar <- ggplot(stacked_df, aes(x = comparison, y = pct, fill = fill_group)) +
    geom_col(width = 0.7) +
    geom_text(data = bar_totals, aes(x = comparison, y = pct, label = n, fill = NULL),
              vjust = -0.3, size = 3) +
    geom_text(data = p_labels, aes(x = 1.5, y = max_pct + 4, label = p_star),
              inherit.aes = FALSE, size = 4, fontface = "bold") +
    facet_wrap(~ category_label, nrow = 1) +
    scale_fill_manual(
      values = c("NMD - No PTC" = "#ef8a62", "NMD - PTC+" = "#8c2d04",
                 "Control - No PTC" = "#67a9cf", "Control - PTC+" = "#2166ac"),
      name = NULL,
      labels = c("NMD - No PTC" = "NMD (No PTC)", "NMD - PTC+" = "NMD (PTC+)",
                 "Control - No PTC" = "Control (No PTC)", "Control - PTC+" = "Control (PTC+)")
    ) +
    theme_minimal(base_size = 12) +
    theme(strip.text = element_text(size = 10, face = "bold"),
          axis.text.x = element_text(angle = 0, hjust = 0.5),
          legend.position = "bottom",
          panel.grid.major.x = element_blank()) +
    labs(x = NULL, y = "% of all pairs",
         title = "Frame Category Distribution: NMD vs Control (all_samples)",
         subtitle = "Stacked by PTC status. Numbers above bars = total pair counts. *** p<0.001, ** p<0.01, * p<0.05, ns (Fisher's exact test)")
  print(p_bar)
  cat("\n\n")
}
```

### PTC Mechanism Breakdown

For PTC-containing comparator isoforms, what is the mechanism by which the PTC arises?
Each PTC+ comparator is classified by its frame category, revealing the route to premature
termination:

- **Same start, frameshift**: Splicing disrupts the reading frame, creating a premature stop
  downstream of the frameshift boundary.
- **Same start, in-frame**: PTC exists within the original reading frame (e.g., a retained
  intron containing a stop codon, or an exon skip that joins in-frame segments with a stop
  between them).
- **Diff start, same frame**: Alternative translation initiation, but shared CDS regions
  are read in the same phase — PTC arises from the shorter or different ORF.
- **Diff start, diff frame**: Alternative initiation reads shared CDS in a different phase.
- **Diff start, diff frame + FS**: Alternative initiation plus an additional splicing-induced
  frameshift.

The table and figure show these mechanism frequencies for both NMD and Control PTC+
comparators. A Fisher's exact test for splicing-induced frameshift enrichment among
same-start pairs is also reported, since this is the primary NMD-generating mechanism.

**PTC Mechanism table columns:** **frame_category** = reading frame classification of
the isoform pair (see §Frame Category Counts above); **NMD_n / NMD_pct** = number and
% of NMD PTC+ comparators in this category; **Ctrl_n / Ctrl_pct** = same for Control.
Percentages are of all PTC+ comparators in each set.

**Frameshift Enrichment table columns:** **cell_type** = cell type;
**NMD_ss_total / Ctrl_ss_total** = total same-start coding pairs (in-frame + frameshift)
in NMD / Control; **NMD_fs_pct / Ctrl_fs_pct** = % of same-start pairs that are
frameshifted; **OR** = odds ratio from Fisher's exact test (>1 = frameshift enriched
in NMD); **p_value** = Fisher's exact test p-value.

```{r sec5-ptc-mechanism, results='asis', fig.width=14, fig.height=6}
coding_cats <- c("same_start_in_frame", "same_start_frameshift",
                 "diff_start_same_frame", "diff_start_diff_frame",
                 "diff_start_diff_frame_with_frameshift")
mech_labels <- c(
  same_start_in_frame = "Same Start\nIn Frame",
  same_start_frameshift = "Same Start\nFrameshift",
  diff_start_same_frame = "Diff Start\nSame Frame",
  diff_start_diff_frame = "Diff Start\nDiff Frame",
  diff_start_diff_frame_with_frameshift = "Diff Start\nDiff Frame + FS"
)

mech_result <- compute_ptc_mechanism_rows("all_samples")
mech_rows <- mech_result$mechanism
fs_test_rows <- mech_result$fs_test

if (!is.null(mech_rows)) {
  nmd  <- mech_rows[mech_rows$comparison == "NMD", ]
  ctrl <- mech_rows[mech_rows$comparison == "Control", ]
  if (nrow(nmd) > 0 && nrow(ctrl) > 0) {
    mech_wide <- data.frame(
      frame_category = nmd$frame_category,
      NMD_n = nmd$n_ptc, NMD_pct = nmd$pct_of_ptc,
      Ctrl_n = ctrl$n_ptc, Ctrl_pct = ctrl$pct_of_ptc
    )
    rownames(mech_wide) <- NULL
    knitr::kable(mech_wide, row.names = FALSE,
      caption = "PTC Mechanism: Frame Category of PTC+ Comparators (NMD vs Control, all_samples)") |> print()
    cat("\n\n")
  }
}

if (!is.null(fs_test_rows)) {
  rownames(fs_test_rows) <- NULL
  knitr::kable(fs_test_rows, row.names = FALSE,
    caption = "Splicing-Induced Frameshift Enrichment (Same-Start Pairs): NMD vs Control (all_samples)") |> print()
  cat("\n\n")
}

# --- Bar chart: PTC mechanism distribution (all_samples only) ---
if (!is.null(mech_rows) && has_ggplot2) {
  mech_rows$category_label <- factor(mech_labels[mech_rows$frame_category],
                                     levels = mech_labels[coding_cats])

  # Fisher's test per category
  p_labels <- do.call(rbind, lapply(coding_cats, function(cat) {
    nmd_r  <- mech_rows[mech_rows$comparison == "NMD", ]
    ctrl_r <- mech_rows[mech_rows$comparison == "Control", ]
    n_nmd <- nmd_r$n_ptc[nmd_r$frame_category == cat]
    t_nmd <- nmd_r$ptc_total[nmd_r$frame_category == cat]
    n_ctrl <- ctrl_r$n_ptc[ctrl_r$frame_category == cat]
    t_ctrl <- ctrl_r$ptc_total[ctrl_r$frame_category == cat]
    if (length(n_nmd) == 0 || length(n_ctrl) == 0) return(NULL)

    ft <- tryCatch({
      mat <- matrix(c(n_nmd, t_nmd - n_nmd, n_ctrl, t_ctrl - n_ctrl), nrow = 2)
      fisher.test(mat)
    }, error = function(e) list(p.value = NA))
    p <- ft$p.value
    stars <- if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else "ns"
    max_pct <- max(mech_rows$pct_of_ptc[mech_rows$frame_category == cat], na.rm = TRUE)
    data.frame(category_label = mech_labels[cat],
               p_star = stars, max_pct = max_pct)
  }))
  p_labels$category_label <- factor(p_labels$category_label, levels = mech_labels[coding_cats])

  p_mech <- ggplot(mech_rows, aes(x = comparison, y = pct_of_ptc, fill = comparison)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = n_ptc), vjust = -0.3, size = 3) +
    geom_text(data = p_labels, aes(x = 1.5, y = max_pct + 5, label = p_star),
              inherit.aes = FALSE, size = 4, fontface = "bold") +
    facet_wrap(~ category_label, nrow = 1) +
    scale_fill_manual(values = c("NMD" = "#ef8a62", "Control" = "#67a9cf"), name = NULL) +
    theme_minimal(base_size = 12) +
    theme(strip.text = element_text(size = 10, face = "bold"),
          axis.text.x = element_text(angle = 0, hjust = 0.5),
          legend.position = "bottom",
          panel.grid.major.x = element_blank()) +
    labs(x = NULL, y = "% of PTC+ comparators",
         title = "PTC Mechanism: How Do PTCs Arise? (all_samples)",
         subtitle = "Distribution of PTC+ comparators across frame categories. Numbers = pair counts. *** p<0.001, ** p<0.01, * p<0.05 (Fisher's exact test)")
  print(p_mech)
  cat("\n\n")
}
```

### Frameshift-Causing Events by Frame Category

Which splicing event types cause the frameshifts in each frame category? Using the frame
walk analysis (`walkFrameAlongEvents()` in `Isopair/R/ptc.R`), each CDS-overlapping event
is flagged as `is_frameshift = TRUE` when its CDS overlap is not divisible by 3. This table
and chart show the distribution of frameshift-causing event types within each frame category,
for both NMD and Control.

Only pairs in frame categories that involve a frameshift are shown (`same_start_frameshift`,
`diff_start_diff_frame_with_frameshift`). For `same_start_in_frame` and
`diff_start_same_frame` pairs, no events cause frameshifts by definition.

**Columns:** **frame_category** = reading frame classification; **comparison** = NMD or
Control; **event_type** = the splicing event type causing the frameshift;
**n_fs_events** = number of frameshift-causing events of this type;
**pct_of_fs_events** = % of all frameshift-causing events in this frame category and
comparison group.

```{r sec5-fs-causing-events, results='asis', fig.width=12, fig.height=7}
# Frame categories with frameshifts
fs_frame_cats <- c("same_start_frameshift",
                   "diff_start_diff_frame_with_frameshift")

# Build table: for each frame category × comparison, which event types cause frameshifts?
fs_event_rows <- do.call(rbind, lapply(c("c2", "c4"), function(comp) {
  key <- paste0(comp, "_allsamples")
  if (!key %in% names(set_results)) return(NULL)
  fc <- set_results[[key]]$frame_comp
  fw <- set_results[[key]]$frame_walk
  if (is.null(fc) || is.null(fw)) return(NULL)
  ps <- fc$pair_summary
  fw_events <- fw$events
  comp_lbl <- comp_label(comp)

  do.call(rbind, lapply(fs_frame_cats, function(cat) {
    # Get pairs in this frame category
    cat_pairs <- ps[ps$frame_category == cat,
                    c("reference_isoform_id", "comparator_isoform_id")]
    if (nrow(cat_pairs) == 0) return(NULL)

    # Get frameshift-causing events for these pairs
    fs_evts <- fw_events[fw_events$is_frameshift == TRUE, ]
    fs_evts <- merge(fs_evts, cat_pairs,
                     by = c("reference_isoform_id", "comparator_isoform_id"))
    if (nrow(fs_evts) == 0) return(NULL)

    # Count by event type
    evt_counts <- as.data.frame(table(fs_evts$event_type), stringsAsFactors = FALSE)
    names(evt_counts) <- c("event_type", "n_fs_events")
    evt_counts$pct_of_fs_events <- round(100 * evt_counts$n_fs_events /
                                          sum(evt_counts$n_fs_events), 1)
    evt_counts$frame_category <- cat
    evt_counts$comparison <- comp_lbl
    evt_counts$n_pairs <- nrow(cat_pairs)
    evt_counts[order(-evt_counts$n_fs_events), ]
  }))
}))

if (!is.null(fs_event_rows) && nrow(fs_event_rows) > 0) {
  rownames(fs_event_rows) <- NULL
  knitr::kable(
    fs_event_rows[, c("frame_category", "comparison", "event_type",
                       "n_fs_events", "pct_of_fs_events")],
    row.names = FALSE,
    caption = "Frameshift-Causing Events by Frame Category (all_samples)") |> print()
  cat("\n\n")
}

# --- Stacked bar chart ---
if (!is.null(fs_event_rows) && nrow(fs_event_rows) > 0 && has_ggplot2) {
  fs_cat_labels <- c(
    same_start_frameshift = "Same Start\nFrameshift",
    diff_start_diff_frame_with_frameshift = "Diff Start\nDiff Frame + FS"
  )
  fs_event_rows$cat_label <- factor(
    fs_cat_labels[fs_event_rows$frame_category],
    levels = fs_cat_labels)

  p_fs_evt <- ggplot(fs_event_rows,
                     aes(x = comparison, y = pct_of_fs_events, fill = event_type)) +
    geom_col(width = 0.7) +
    facet_wrap(~ cat_label, nrow = 1) +
    theme_minimal(base_size = 12) +
    theme(strip.text = element_text(size = 11, face = "bold"),
          legend.position = "bottom",
          panel.grid.major.x = element_blank()) +
    labs(x = NULL, y = "% of frameshift-causing events",
         fill = "Event Type",
         title = "Which Splicing Events Cause Frameshifts? (all_samples)",
         subtitle = "Distribution of frameshift-causing event types within each frame category")
  print(p_fs_evt)
  cat("\n\n")
}
```

## Frameshift Consequences

Among pairs that share the same start codon, splicing-induced frameshifts are common in both
NMD and Control comparisons, but their downstream consequences differ sharply. In NMD pairs,
frameshifts overwhelmingly produce PTCs that trigger decay. In Control pairs, most frameshifts
place the new stop codon in the last exon (0 downstream EJCs), producing stable proteins with
rewritten C-terminal sequences — productive alternative splicing generating protein diversity.

### Frameshift Outcome Distribution

Same-start coding pairs are partitioned into four mutually exclusive outcome categories, each
expressed as a percentage of all same-start coding pairs (so bars across categories sum to
100% within each cell type). This uses the cross-isoform reading frame comparison
(`compareIsoformFrames()` in `Isopair/R/ptc.R`; see METHODS.md §Reading Frame and PTC
Analysis). Only pairs classified as `same_start_in_frame` or `same_start_frameshift` are
included, since these share translation initiation and any frame difference is attributable
to splicing.

The four categories are:

- **In Frame**: Same start codon, no frameshift — splicing differences preserve reading frame.
- **Frameshift → PTC**: Frameshift produces a premature stop codon (>50 nt upstream of last
  EJC), triggering NMD.
- **Frameshift → Stop in Last Exon**: Frameshift without PTC; stop codon falls in the last
  exon (0 downstream EJCs), producing a stable protein with novel C-terminal sequence.
- **Frameshift → Other**: Frameshift without PTC and stop not in last exon.

```{r sec5-frameshift-outcome, results='asis', fig.width=14, fig.height=6}
outcome_rows <- compute_frameshift_outcome_rows("all_samples")

if (!is.null(outcome_rows)) {
  outcome_tbl <- outcome_rows
  outcome_tbl$in_frame_pct <- round(100 * outcome_tbl$n_in_frame / outcome_tbl$n_same_start, 1)
  outcome_tbl$fs_ptc_pct <- round(100 * outcome_tbl$n_fs_ptc / outcome_tbl$n_same_start, 1)
  outcome_tbl$fs_stop_last_pct <- round(100 * outcome_tbl$n_fs_stop_last / outcome_tbl$n_same_start, 1)
  outcome_tbl$fs_other_pct <- round(100 * outcome_tbl$n_fs_other / outcome_tbl$n_same_start, 1)
  rownames(outcome_tbl) <- NULL
  knitr::kable(outcome_tbl[, c("cell_type", "comparison", "n_same_start",
                                "n_in_frame", "in_frame_pct",
                                "n_fs_ptc", "fs_ptc_pct",
                                "n_fs_stop_last", "fs_stop_last_pct",
                                "n_fs_other", "fs_other_pct")],
    row.names = FALSE,
    caption = "Frameshift Outcome Distribution: NMD vs Control (Same-Start Pairs, all_samples)") |> print()
  cat("\n\n")
}

# --- Stacked bar chart: outcomes as % of same-start pairs (all_samples only) ---
if (!is.null(outcome_rows) && has_ggplot2) {
  outcome_cats <- c("In Frame", "Frameshift\n\u2192 PTC",
                    "Frameshift\n\u2192 Stop in Last Exon", "Frameshift\n\u2192 Other")

  outcome_plot_df <- do.call(rbind, lapply(seq_len(nrow(outcome_rows)), function(i) {
    r <- outcome_rows[i, ]
    n <- r$n_same_start
    data.frame(
      comparison = r$comparison,
      outcome = outcome_cats,
      n = c(r$n_in_frame, r$n_fs_ptc, r$n_fs_stop_last, r$n_fs_other),
      pct = 100 * c(r$n_in_frame, r$n_fs_ptc, r$n_fs_stop_last, r$n_fs_other) / max(n, 1)
    )
  }))
  outcome_plot_df$outcome <- factor(outcome_plot_df$outcome, levels = outcome_cats)

  # Fisher's test per outcome category
  nmd_row <- outcome_rows[outcome_rows$comparison == "NMD", ]
  ctrl_row <- outcome_rows[outcome_rows$comparison == "Control", ]
  p_labels <- NULL
  if (nrow(nmd_row) > 0 && nrow(ctrl_row) > 0) {
    count_cols <- c("n_in_frame", "n_fs_ptc", "n_fs_stop_last", "n_fs_other")
    p_labels <- do.call(rbind, lapply(seq_along(outcome_cats), function(j) {
      ft <- tryCatch({
        mat <- matrix(c(nmd_row[[count_cols[j]]], nmd_row$n_same_start - nmd_row[[count_cols[j]]],
                        ctrl_row[[count_cols[j]]], ctrl_row$n_same_start - ctrl_row[[count_cols[j]]]), nrow = 2)
        fisher.test(mat)
      }, error = function(e) list(p.value = NA))
      p <- ft$p.value
      stars <- if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else "ns"
      max_pct <- max(outcome_plot_df$pct[outcome_plot_df$outcome == outcome_cats[j]], na.rm = TRUE)
      data.frame(outcome = outcome_cats[j], p_star = stars, max_pct = max_pct)
    }))
    p_labels$outcome <- factor(p_labels$outcome, levels = outcome_cats)
  }

  p_out <- ggplot(outcome_plot_df, aes(x = comparison, y = pct, fill = comparison)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = n), vjust = -0.3, size = 3) +
    { if (!is.null(p_labels)) geom_text(data = p_labels, aes(x = 1.5, y = max_pct + 4, label = p_star),
              inherit.aes = FALSE, size = 4, fontface = "bold") } +
    facet_wrap(~ outcome, nrow = 1) +
    scale_fill_manual(values = c("NMD" = "#ef8a62", "Control" = "#67a9cf"), name = NULL) +
    theme_minimal(base_size = 12) +
    theme(strip.text = element_text(size = 10, face = "bold"),
          axis.text.x = element_text(angle = 0, hjust = 0.5),
          legend.position = "bottom",
          panel.grid.major.x = element_blank()) +
    labs(x = NULL, y = "% of same-start pairs",
         title = "Frameshift Outcome Distribution: NMD vs Control (all_samples)",
         subtitle = "All categories sum to 100%. Numbers above bars = pair counts. *** p<0.001, ** p<0.01, * p<0.05 (Fisher's exact test)")
  print(p_out)
  cat("\n\n")
}
```

In NMD comparisons, splicing-induced frameshifts overwhelmingly produce PTCs that trigger decay.
In Control comparisons, most frameshifts place the new stop codon in the last exon
(0 downstream EJCs), producing a stable mRNA with a novel C-terminal protein sequence.

### Productive Frameshifts (Control)

Control (C4) pairs with same-start splicing-induced frameshifts and no PTC represent productive
alternative splicing: the frameshift rewrites the C-terminal protein sequence without triggering
NMD. These `r nrow(compute_novel_protein("all_samples"))` pairs produce stable mRNAs encoding
proteins with novel C-terminal sequences — a different protein, not a degraded one.

```{r sec5-novel-protein}
# Compute novel protein metrics for C4 same_start_frameshift pairs without PTC
# all_samples for main body; per-CT computed in Section 5
novel_protein_results <- list()
novel_protein_results[["all_samples"]] <- compute_novel_protein("all_samples")
```

**Columns:** **n_stable_novel** = C4 same-start frameshifts without PTC; **pct_of_same_start** = as % of all C4 same-start coding pairs; **median_novel_aa / iqr_novel_aa** = novel amino acids (translated in shifted frame); **median_total_aa** = total protein length; **median_pct_novel / iqr_pct_novel** = % of protein that is novel sequence.

```{r sec5-novel-protein-table, results='asis'}
np_as <- novel_protein_results[["all_samples"]]
novel_summary <- NULL
if (!is.null(np_as) && nrow(np_as) > 0) {
  fc <- set_results[["c4_allsamples"]]$frame_comp
  n_same_start <- sum(fc$pair_summary$frame_category %in%
    c("same_start_in_frame", "same_start_frameshift"))
  novel_summary <- data.frame(
    n_stable_novel = nrow(np_as),
    pct_of_same_start = round(100 * nrow(np_as) / n_same_start, 1),
    median_novel_aa = round(median(np_as$novel_aa), 0),
    iqr_novel_aa = paste0(round(quantile(np_as$novel_aa, 0.25), 0), "-",
                          round(quantile(np_as$novel_aa, 0.75), 0)),
    median_total_aa = round(median(np_as$total_aa), 0),
    median_pct_novel = round(median(np_as$pct_novel), 1),
    iqr_pct_novel = paste0(round(quantile(np_as$pct_novel, 0.25), 1), "-",
                           round(quantile(np_as$pct_novel, 0.75), 1))
  )
  knitr::kable(novel_summary, caption = "Stable Novel Proteins: Control (C4) Same-Start Frameshifts Without PTC (all_samples)") |> print()
  cat("\n\n")
}
```

```{r sec5-novel-protein-plot, fig.height=5}
if (has_ggplot2 && !is.null(np_as) && nrow(np_as) > 0) {
  p_novel <- ggplot(data.frame(pct_novel = np_as$pct_novel), aes(x = pct_novel)) +
    geom_density(fill = "steelblue", alpha = 0.5) +
    theme_minimal() +
    labs(title = "Novel Protein Sequence in Stable Control Frameshifts (all_samples)",
         subtitle = "C4 unresolved frameshifts without PTC — % of protein in shifted reading frame",
         x = "% of protein that is novel sequence", y = "Density")
  print(p_novel)
}
```

```{r sec5-pf-load-cache}
# Load pre-computed productive frameshift characterization
pf_cache_path <- file.path(data_dir, "analysis_cache", "productive_frameshift_allsamples.rds")
pf_data <- if (file.exists(pf_cache_path)) readRDS(pf_cache_path) else NULL
has_pf <- !is.null(pf_data)
if (has_pf) {
  pf_char <- pf_data$characterization
  pf_ms   <- pf_data$massspec
  pf_dom  <- pf_data$domains
  pf_cons <- pf_data$conservation
}
```

#### Frameshift Mechanisms

Splicing events that cause productive frameshifts — which event types shift reading frame
without introducing a PTC?

```{r sec5-pf-mechanisms, results='asis', fig.height=5}
if (has_pf && has_ggplot2) {
  # Expand comma-separated event types to one row per event type per pair
  event_types_raw <- pf_char$frameshift_event_types[!is.na(pf_char$frameshift_event_types)]
  all_event_types <- unlist(strsplit(event_types_raw, ","))
  event_freq <- sort(table(all_event_types), decreasing = TRUE)

  event_df <- data.frame(
    event_type = names(event_freq),
    count = as.integer(event_freq),
    stringsAsFactors = FALSE
  )
  event_df$pct <- round(100 * event_df$count / length(event_types_raw), 1)

  knitr::kable(event_df,
    col.names = c("Event Type", "Pairs", "% of Pairs"),
    caption = "Splicing events causing productive frameshifts (a pair may have >1 event type)") |> print()
  cat("\n\n")

  # Bar chart
  p_evt <- ggplot(event_df, aes(x = reorder(event_type, -count), y = count)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    geom_text(aes(label = count), vjust = -0.3, size = 3.5) +
    theme_minimal(base_size = 12) +
    labs(x = "Splicing Event Type", y = "Number of Pairs",
         title = "Frameshift-Causing Event Types in Productive Frameshifts")
  print(p_evt)

  # Scatter: pct_novel vs total protein length
  pf_plot <- pf_char[!is.na(pf_char$frameshift_event_types), ]
  # Use primary (first) event type for coloring
  pf_plot$primary_event <- sapply(strsplit(pf_plot$frameshift_event_types, ","), `[`, 1)

  p_scatter <- ggplot(pf_plot, aes(x = total_aa, y = pct_novel, color = primary_event)) +
    geom_point(alpha = 0.7, size = 2.5) +
    theme_minimal(base_size = 12) +
    labs(x = "Total Protein Length (aa)", y = "% Novel Sequence",
         color = "Primary Event",
         title = "Novel Protein Fraction vs Protein Size",
         subtitle = "Colored by primary frameshift-causing event type")
  print(p_scatter)
}
```

#### Functional Domain Impact

Domain annotations from Pfam/InterPro for ENST isoforms, classified relative to the frameshift
boundary: **conserved** (entirely upstream), **disrupted** (spanning the boundary), or
**novel_region** (entirely downstream, in the frameshifted sequence). For novel isoforms
without Ensembl IDs, conserved-region domains are proxied from the paired ENST isoform
(marked as "ref proxy" in the domain source column).

```{r sec5-pf-domains, results='asis', fig.height=5}
if (has_pf && nrow(pf_dom) > 0 && has_ggplot2) {
  # Per-pair summary: does this pair have any disrupted domains?
  pair_domain_summary <- pf_dom |>
    (\(d) {
      pairs_with_disrupted <- unique(d$gene_id[d$position_class == "disrupted"])
      pairs_with_any <- unique(d$gene_id)
      all_pairs <- unique(pf_char$gene_id)

      data.frame(
        category = c("At least one domain disrupted",
                      "Domains only in conserved region",
                      "No annotated domains"),
        n_pairs = c(
          length(pairs_with_disrupted),
          length(setdiff(pairs_with_any, pairs_with_disrupted)),
          length(setdiff(all_pairs, pairs_with_any))
        )
      )
    })()
  pair_domain_summary$pct <- round(100 * pair_domain_summary$n_pairs /
                                     nrow(pf_char), 1)

  knitr::kable(pair_domain_summary,
    col.names = c("Category", "N Pairs", "% of Total"),
    caption = "Domain disposition across productive frameshift pairs") |> print()
  cat("\n\n")

  # Stacked bar: position class distribution
  pos_freq <- as.data.frame(table(pf_dom$position_class))
  names(pos_freq) <- c("position", "count")
  pos_freq$position <- factor(pos_freq$position,
                                levels = c("conserved", "disrupted", "novel_region"))

  p_dom_bar <- ggplot(pos_freq, aes(x = position, y = count, fill = position)) +
    geom_col(alpha = 0.8) +
    geom_text(aes(label = count), vjust = -0.3, size = 3.5) +
    scale_fill_manual(values = c(conserved = "grey60", disrupted = "tomato",
                                  novel_region = "steelblue")) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none") +
    labs(x = "Domain Position Relative to Frameshift",
         y = "Number of Domain Annotations",
         title = "Functional Domain Disposition in Productive Frameshifts")
  print(p_dom_bar)

  # DT table: per-pair domain details
  dom_display <- pf_dom[, c("gene_id", "isoform_id", "isoform_role",
                              "domain_id", "domain_name", "domain_source",
                              "domain_start_aa", "domain_end_aa", "position_class")]
  dom_display$gene_symbol <- gene_symbol_lookup[dom_display$gene_id]
  dom_display <- dom_display[, c("gene_symbol", "gene_id", "isoform_id", "isoform_role",
                                   "domain_id", "domain_name", "domain_source",
                                   "domain_start_aa", "domain_end_aa", "position_class")]
  render_dt(dom_display,
    caption = "Domain annotations for productive frameshift isoforms (filterable)")
} else if (has_pf) {
  cat("No domain annotations were retrieved (biomaRt may have been unavailable).\n\n")
}
```

#### Proteomics Evidence

Mass-spectrometry evidence from the EBI Proteomics database for peptides unique to the
novel C-terminal region. Peptides are generated via in-silico tryptic digest (7-30 aa,
cleaving at K/R not followed by P) and filtered to those absent from the reference protein.

```{r sec5-pf-massspec, results='asis'}
if (has_pf && nrow(pf_ms) > 0) {
  n_queried_genes <- length(unique(pf_ms$comparator_isoform_id))
  n_hit_genes <- length(unique(pf_ms$comparator_isoform_id[pf_ms$ebi_hit]))
  n_gene_match <- length(unique(pf_ms$comparator_isoform_id[pf_ms$ebi_gene_match]))

  ms_summary <- data.frame(
    metric = c("Genes queried (with unique peptides)",
               "Genes with any EBI hit",
               "Genes with gene-matched hit",
               "Total unique peptides queried",
               "Peptides with EBI hits"),
    value = c(n_queried_genes, n_hit_genes, n_gene_match,
              nrow(pf_ms), sum(pf_ms$ebi_hit))
  )
  knitr::kable(ms_summary, col.names = c("Metric", "N"),
    caption = "Mass-spectrometry evidence for novel-region peptides") |> print()
  cat("\n\n")

  # Show hits if any
  ms_hits <- pf_ms[pf_ms$ebi_hit, ]
  if (nrow(ms_hits) > 0) {
    ms_hits$gene_symbol <- gene_symbol_lookup[ms_hits$gene_id]
    hit_display <- ms_hits[, c("gene_symbol", "gene_id", "comparator_isoform_id",
                                 "peptide", "peptide_length", "ebi_gene_match",
                                 "ebi_accession")]
    render_dt(hit_display, caption = "Peptides with EBI proteomics evidence")
  }
} else if (has_pf) {
  cat("No mass-spec queries were performed (no novel-region peptides met length criteria).\n\n")
}
```

Most productive frameshifts lack mass-spec evidence for their novel C-terminal peptides.
This is expected: (1) these are minor isoforms with lower expression; (2) the novel
regions are short (median ~40 aa), limiting the number of tryptic peptides; (3) proteomics
databases are biased toward principal isoforms. The few genes with hits typically correspond
to known alternative isoforms whose "novel" sequence is only novel relative to the dominant
isoform in the pair, not globally unknown.

#### Isoform Annotation Status

APPRIS, TSL (Transcript Support Level), and UniProt annotations for reference and comparator
isoforms. This reveals whether productive frameshifts involve well-characterized alternative
isoforms or poorly annotated transcripts.

```{r sec5-pf-conservation, results='asis'}
if (has_pf) {
  # APPRIS breakdown
  ref_appris <- table(pf_cons$ref_appris, useNA = "ifany")
  comp_appris <- table(pf_cons$comp_appris, useNA = "ifany")

  appris_df <- merge(
    data.frame(appris = names(ref_appris), reference = as.integer(ref_appris)),
    data.frame(appris = names(comp_appris), comparator = as.integer(comp_appris)),
    by = "appris", all = TRUE
  )
  appris_df$appris[is.na(appris_df$appris)] <- "(no annotation)"
  appris_df$reference[is.na(appris_df$reference)] <- 0
  appris_df$comparator[is.na(appris_df$comparator)] <- 0

  knitr::kable(appris_df, col.names = c("APPRIS", "Reference", "Comparator"),
    caption = "APPRIS classification: reference vs comparator isoforms") |> print()
  cat("\n\n")

  # TSL breakdown for comparators
  comp_tsl <- table(pf_cons$comp_tsl, useNA = "ifany")
  tsl_df <- data.frame(tsl = names(comp_tsl), count = as.integer(comp_tsl))
  tsl_df$tsl[is.na(tsl_df$tsl)] <- "(novel/no annotation)"
  knitr::kable(tsl_df, col.names = c("TSL", "Comparator Count"),
    caption = "Transcript Support Level for comparator (frameshifted) isoforms") |> print()
  cat("\n\n")

  # UniProt for ENST comparators
  enst_cons <- pf_cons[grepl("^ENST", pf_cons$comparator_isoform_id), ]
  if (any(!is.na(enst_cons$comp_uniprot_accession))) {
    uniprot_summary <- data.frame(
      category = c("Swiss-Prot (reviewed)", "TrEMBL (unreviewed)", "No UniProt entry"),
      count = c(
        sum(enst_cons$comp_uniprot_reviewed == "Swiss-Prot", na.rm = TRUE),
        sum(enst_cons$comp_uniprot_reviewed == "TrEMBL", na.rm = TRUE),
        sum(is.na(enst_cons$comp_uniprot_accession))
      )
    )
    knitr::kable(uniprot_summary, col.names = c("UniProt Status", "ENST Comparators"),
      caption = "UniProt protein existence for ENST comparator isoforms") |> print()
    cat("\n\n")

    # Show Swiss-Prot entries
    swissprot <- enst_cons[!is.na(enst_cons$comp_uniprot_reviewed) &
                             enst_cons$comp_uniprot_reviewed == "Swiss-Prot", ]
    if (nrow(swissprot) > 0) {
      swissprot$gene_symbol <- gene_symbol_lookup[swissprot$gene_id]
      sp_display <- swissprot[, c("gene_symbol", "comparator_isoform_id",
                                    "comp_uniprot_accession", "comp_protein_existence",
                                    "comp_appris", "comp_tsl")]
      render_dt(sp_display,
        caption = "ENST comparators with Swiss-Prot (reviewed) entries")
    }
  }
}
```

Productive frameshifts predominantly involve alternative (non-principal) isoforms as the
comparator. The reference (dominant) isoform is typically the APPRIS principal isoform,
while the frameshifted comparator is an alternative isoform — consistent with the
biological expectation that productive frameshifts represent regulated alternative splicing
producing functionally distinct protein variants.

### Frameshift Examples: Real Gene Cases

Four real gene examples illustrating the major categories of reading frame outcomes.
Panel A: a same-start splicing-induced frameshift that produces a PTC, triggering NMD.
Panel B: a different-start, different-frame isoform without PTC, producing a stable protein
with novel C-terminal sequence.
Panel C: a different-start, different-frame isoform with PTC, triggering NMD.
Panel D: a canonical poison exon — same start codon, in frame, but an included or retained
exonic region introduces a premature stop codon.

Each example is shown first as a full-gene overview, then zoomed to the region containing
the key splicing event.

```{r sec5-frameshift-examples, fig.width=12, fig.height=24}
if (has_ggplot2) {
  # Helper: resolve gene symbol from gene_id
  get_gene_sym <- function(gid) {
    sym <- gene_symbol_lookup[gid]
    if (is.na(sym) || sym == "") gid else sym
  }

  # Helper: zoom on a SPECIFIC subset of events (e.g., the frameshift-causing one)
  compute_event_zoom <- function(events, padding_frac = 0.3) {
    if (is.null(events) || nrow(events) == 0) return(NULL)
    ev_min <- min(events$five_prime, events$three_prime, na.rm = TRUE)
    ev_max <- max(events$five_prime, events$three_prime, na.rm = TRUE)
    span <- ev_max - ev_min
    pad <- max(span * padding_frac, 500)
    c(ev_min - pad, ev_max + pad)
  }

  # Helper: find frameshift-causing events using frame_walk
  # Returns data frame with five_prime/three_prime columns for compute_event_zoom()
  get_fs_events <- function(fw, ref_id, comp_id) {
    if (is.null(fw) || is.null(fw$events) || nrow(fw$events) == 0) return(NULL)
    fs <- fw$events[fw$events$reference_isoform_id == ref_id &
                    fw$events$comparator_isoform_id == comp_id &
                    fw$events$is_frameshift == TRUE, ]
    if (nrow(fs) == 0) return(NULL)
    # Return with five_prime/three_prime names for compute_event_zoom()
    data.frame(five_prime = fs$genomic_start, three_prime = fs$genomic_end)
  }

  # Helper: build a plot pair (full + zoom) and append to plot list
  build_panel <- function(pick, prof_key, ref_label, comp_label_text,
                          title_text, sub_text, annotations = list(),
                          zoom_events = NULL) {
    ref_ex <- get_exon_coords(pick$reference_isoform_id, structures)
    comp_ex <- get_exon_coords(pick$comparator_isoform_id, structures)
    ref_cds_r <- get_cds_range(pick$reference_isoform_id, cds)
    comp_cds_r <- get_cds_range(pick$comparator_isoform_id, cds)
    str_row <- structures[structures$isoform_id == pick$reference_isoform_id, ]
    gene_strand <- str_row$strand[1]

    profs <- set_results[[prof_key]]$profiles
    pair_prof <- profs[profs$reference_isoform_id == pick$reference_isoform_id &
                       profs$comparator_isoform_id == pick$comparator_isoform_id, ]
    pair_events <- if (nrow(pair_prof) > 0) pair_prof$detailed_events[[1]] else NULL

    plots <- list()
    plots[["full"]] <- plot_ucsc_pair(
      ref_ex$starts, ref_ex$ends, comp_ex$starts, comp_ex$ends, gene_strand,
      ref_cds = ref_cds_r, comp_cds = comp_cds_r,
      ref_label = ref_label, comp_label = comp_label_text,
      events = pair_events, annotations = annotations,
      title = title_text, subtitle = sub_text
    )

    # Zoom: use zoom_events if provided, otherwise all events
    zoom_target <- if (!is.null(zoom_events) && nrow(zoom_events) > 0) zoom_events else pair_events
    zoom_lim <- compute_event_zoom(zoom_target)
    if (!is.null(zoom_lim)) {
      plots[["zoom"]] <- plot_ucsc_pair(
        ref_ex$starts, ref_ex$ends, comp_ex$starts, comp_ex$ends, gene_strand,
        ref_cds = ref_cds_r, comp_cds = comp_cds_r,
        ref_label = ref_label, comp_label = comp_label_text,
        events = pair_events, annotations = annotations,
        title = paste0(title_text, " [ZOOM]"), subtitle = sub_text,
        xlim = zoom_lim
      )
    }
    list(plots = plots, strand = gene_strand, ref_cds = ref_cds_r, comp_cds = comp_cds_r)
  }

  # Use all_samples pooled data (matching main body)
  c2_key <- "c2_allsamples"
  c4_key <- "c4_allsamples"
  fc_c2 <- set_results[[c2_key]]$frame_comp
  fc_c4 <- set_results[[c4_key]]$frame_comp
  fw_c2 <- set_results[[c2_key]]$frame_walk
  fw_c4 <- set_results[[c4_key]]$frame_walk

  example_plots <- list()

  # --- Panel A: Same-start frameshift → PTC (NMD, C2) ---
  if (!is.null(fc_c2)) {
    ps_c2 <- fc_c2$pair_summary
    fs_c2 <- ps_c2[ps_c2$frame_category == "same_start_frameshift", ]
    fs_c2_ptc <- merge(fs_c2, ptc[ptc$has_ptc == TRUE, c("isoform_id", "has_ptc",
                       "ptc_distance", "n_downstream_ejcs")],
                       by.x = "comparator_isoform_id", by.y = "isoform_id")
    fs_c2_ptc <- merge(fs_c2_ptc,
                       structures[, c("isoform_id", "n_exons")],
                       by.x = "comparator_isoform_id", by.y = "isoform_id")
    names(fs_c2_ptc)[names(fs_c2_ptc) == "n_exons"] <- "comp_n_exons"
    fs_c2_ptc <- merge(fs_c2_ptc,
                       structures[, c("isoform_id", "n_exons")],
                       by.x = "reference_isoform_id", by.y = "isoform_id")
    names(fs_c2_ptc)[names(fs_c2_ptc) == "n_exons"] <- "ref_n_exons"
    fs_c2_ptc$exon_diff <- abs(fs_c2_ptc$ref_n_exons - fs_c2_ptc$comp_n_exons)
    candidates_a <- fs_c2_ptc[fs_c2_ptc$comp_n_exons >= 3 & fs_c2_ptc$comp_n_exons <= 10 &
                               fs_c2_ptc$exon_diff <= 1, ]
    if (nrow(candidates_a) > 0) {
      pick_a <- candidates_a[which.max(candidates_a$ptc_distance), ]

      # Find the frameshift-causing event for targeted zoom
      fs_evts_a <- get_fs_events(fw_c2, pick_a$reference_isoform_id,
                                  pick_a$comparator_isoform_id)

      # PTC annotation
      str_row_a <- structures[structures$isoform_id == pick_a$reference_isoform_id, ]
      strand_a <- str_row_a$strand[1]
      comp_cds_a <- get_cds_range(pick_a$comparator_isoform_id, cds)
      ptc_ann <- list()
      if (!is.null(comp_cds_a)) {
        ptc_pos <- if (strand_a == "+") comp_cds_a$cds_stop else comp_cds_a$cds_start
        ptc_ann <- list(
          list(type = "ptc_marker", x = ptc_pos, track = "comp",
               label = "PTC", color = "#E74C3C")
        )
      }

      gene_sym_a <- get_gene_sym(pick_a$gene_id)
      title_a <- sprintf("A) Same-Start Frameshift \u2192 PTC: %s (%s)",
                         gene_sym_a, pick_a$gene_id)
      sub_a <- sprintf("Ref: %s | Comp: %s\nPTC distance = %d nt, %d downstream EJC(s)",
                       pick_a$reference_isoform_id, pick_a$comparator_isoform_id,
                       round(pick_a$ptc_distance), pick_a$n_downstream_ejcs)

      # Zoom spans from frameshift-causing event to PTC position
      zoom_df_a <- fs_evts_a
      if (!is.null(zoom_df_a) && !is.null(comp_cds_a)) {
        ptc_pos_a <- if (strand_a == "+") comp_cds_a$cds_stop else comp_cds_a$cds_start
        zoom_df_a <- rbind(zoom_df_a,
                           data.frame(five_prime = ptc_pos_a, three_prime = ptc_pos_a))
      }

      panel_a <- build_panel(pick_a, c2_key, "Dominant\n(non-NMD)", "NMD-sensitive",
                             title_a, sub_a, annotations = ptc_ann,
                             zoom_events = zoom_df_a)
      example_plots <- c(example_plots, panel_a$plots)
    }
  }

  # --- Panel B: Diff-start diff-frame → Novel Protein (no PTC, C2) ---
  if (!is.null(fc_c2)) {
    ps_c2 <- fc_c2$pair_summary
    ds_c2 <- ps_c2[ps_c2$frame_category %in% c("diff_start_diff_frame",
                    "diff_start_diff_frame_with_frameshift"), ]
    # Exclude PTC+ comparators
    ds_c2_noptc <- merge(ds_c2, ptc[, c("isoform_id", "has_ptc")],
                         by.x = "comparator_isoform_id", by.y = "isoform_id", all.x = TRUE)
    ds_c2_noptc <- ds_c2_noptc[is.na(ds_c2_noptc$has_ptc) | ds_c2_noptc$has_ptc == FALSE, ]
    # Require coding comparator (has CDS)
    ds_c2_noptc <- ds_c2_noptc[ds_c2_noptc$comparator_isoform_id %in% cds$isoform_id, ]
    ds_c2_noptc <- merge(ds_c2_noptc,
                         structures[, c("isoform_id", "n_exons")],
                         by.x = "comparator_isoform_id", by.y = "isoform_id")
    candidates_b <- ds_c2_noptc[ds_c2_noptc$n_exons >= 3 & ds_c2_noptc$n_exons <= 10, ]
    # Prefer pairs with moderate frameshifted CDS for visual interest
    if (nrow(candidates_b) > 0) {
      candidates_b <- candidates_b[candidates_b$pct_shared_cds_frameshifted >= 30 &
                                    candidates_b$pct_shared_cds_frameshifted <= 80, ]
    }
    if (nrow(candidates_b) > 0) {
      pick_b <- candidates_b[1, ]

      # Annotation: different start codons
      ref_cds_b <- get_cds_range(pick_b$reference_isoform_id, cds)
      comp_cds_b <- get_cds_range(pick_b$comparator_isoform_id, cds)
      str_row_b <- structures[structures$isoform_id == pick_b$reference_isoform_id, ]
      strand_b <- str_row_b$strand[1]

      ds_ann <- list()
      if (!is.null(ref_cds_b) && !is.null(comp_cds_b)) {
        ref_start_pos <- if (strand_b == "+") ref_cds_b$cds_start else ref_cds_b$cds_stop
        comp_start_pos <- if (strand_b == "+") comp_cds_b$cds_start else comp_cds_b$cds_stop
        ds_ann <- list(
          list(type = "ptc_marker", x = ref_start_pos, track = "ref",
               label = "ATG", color = "#27AE60"),
          list(type = "ptc_marker", x = comp_start_pos, track = "comp",
               label = "ATG", color = "#27AE60")
        )
      }

      gene_sym_b <- get_gene_sym(pick_b$gene_id)
      title_b <- sprintf("B) Diff-Start Diff-Frame \u2192 Novel Protein: %s (%s)",
                         gene_sym_b, pick_b$gene_id)
      sub_b <- sprintf("Ref: %s | Comp: %s\nDifferent start codons, %.0f%% of shared CDS frameshifted, no PTC",
                       pick_b$reference_isoform_id, pick_b$comparator_isoform_id,
                       pick_b$pct_shared_cds_frameshifted)

      panel_b <- build_panel(pick_b, c2_key, "Dominant\n(non-NMD)", "NMD-sensitive",
                             title_b, sub_b, annotations = ds_ann,
                             zoom_events = NULL)
      example_plots <- c(example_plots, panel_b$plots)
    }
  }

  # --- Panel C: Diff-start diff-frame → PTC (NMD, C2) ---
  if (!is.null(fc_c2)) {
    ps_c2 <- fc_c2$pair_summary
    ds_c2 <- ps_c2[ps_c2$frame_category %in% c("diff_start_diff_frame",
                    "diff_start_diff_frame_with_frameshift"), ]
    ds_c2_ptc <- merge(ds_c2, ptc[ptc$has_ptc == TRUE, c("isoform_id", "has_ptc",
                       "ptc_distance", "n_downstream_ejcs")],
                       by.x = "comparator_isoform_id", by.y = "isoform_id")
    ds_c2_ptc <- merge(ds_c2_ptc,
                       structures[, c("isoform_id", "n_exons")],
                       by.x = "comparator_isoform_id", by.y = "isoform_id")
    names(ds_c2_ptc)[names(ds_c2_ptc) == "n_exons"] <- "comp_n_exons"
    candidates_c <- ds_c2_ptc[ds_c2_ptc$comp_n_exons >= 3 & ds_c2_ptc$comp_n_exons <= 10, ]
    if (nrow(candidates_c) > 0) {
      pick_c <- candidates_c[which.max(candidates_c$ptc_distance), ]

      str_row_c <- structures[structures$isoform_id == pick_c$reference_isoform_id, ]
      strand_c <- str_row_c$strand[1]
      comp_cds_c <- get_cds_range(pick_c$comparator_isoform_id, cds)
      ref_cds_c <- get_cds_range(pick_c$reference_isoform_id, cds)

      # Annotations: different start codons + PTC
      ptc_ds_ann <- list()
      if (!is.null(ref_cds_c) && !is.null(comp_cds_c)) {
        ref_start_pos <- if (strand_c == "+") ref_cds_c$cds_start else ref_cds_c$cds_stop
        comp_start_pos <- if (strand_c == "+") comp_cds_c$cds_start else comp_cds_c$cds_stop
        ptc_pos_c <- if (strand_c == "+") comp_cds_c$cds_stop else comp_cds_c$cds_start
        ptc_ds_ann <- list(
          list(type = "ptc_marker", x = ref_start_pos, track = "ref",
               label = "ATG", color = "#27AE60"),
          list(type = "ptc_marker", x = comp_start_pos, track = "comp",
               label = "ATG", color = "#27AE60"),
          list(type = "ptc_marker", x = ptc_pos_c, track = "comp",
               label = "PTC", color = "#E74C3C")
        )
      }

      gene_sym_c <- get_gene_sym(pick_c$gene_id)
      title_c <- sprintf("C) Diff-Start Diff-Frame \u2192 PTC: %s (%s)",
                         gene_sym_c, pick_c$gene_id)
      sub_c <- sprintf("Ref: %s | Comp: %s\nDifferent start codons, PTC distance = %d nt, %d downstream EJC(s)",
                       pick_c$reference_isoform_id, pick_c$comparator_isoform_id,
                       round(pick_c$ptc_distance), pick_c$n_downstream_ejcs)

      # Zoom on the PTC region
      zoom_df_c <- NULL
      if (!is.null(comp_cds_c)) {
        ptc_pos_c <- if (strand_c == "+") comp_cds_c$cds_stop else comp_cds_c$cds_start
        zoom_df_c <- data.frame(five_prime = ptc_pos_c, three_prime = ptc_pos_c)
      }

      panel_c <- build_panel(pick_c, c2_key, "Dominant\n(non-NMD)", "NMD-sensitive",
                             title_c, sub_c, annotations = ptc_ds_ann,
                             zoom_events = zoom_df_c)
      example_plots <- c(example_plots, panel_c$plots)
    }
  }

  # --- Panel D: Canonical Poison Exon (same start, in frame, PTC from included exon) ---
  if (!is.null(fc_c2)) {
    ps_c2 <- fc_c2$pair_summary
    sif_c2 <- ps_c2[ps_c2$frame_category == "same_start_in_frame", ]
    # Find in-frame pairs where the comparator has a PTC
    sif_c2_ptc <- merge(sif_c2, ptc[ptc$has_ptc == TRUE, c("isoform_id", "has_ptc",
                        "ptc_distance", "n_downstream_ejcs")],
                        by.x = "comparator_isoform_id", by.y = "isoform_id")
    # Merge exon counts
    sif_c2_ptc <- merge(sif_c2_ptc,
                        structures[, c("isoform_id", "n_exons")],
                        by.x = "comparator_isoform_id", by.y = "isoform_id")
    names(sif_c2_ptc)[names(sif_c2_ptc) == "n_exons"] <- "comp_n_exons"
    sif_c2_ptc <- merge(sif_c2_ptc,
                        structures[, c("isoform_id", "n_exons")],
                        by.x = "reference_isoform_id", by.y = "isoform_id")
    names(sif_c2_ptc)[names(sif_c2_ptc) == "n_exons"] <- "ref_n_exons"
    # Prefer cases where comparator has MORE exons (included exon) or same (retained intron)
    candidates_d <- sif_c2_ptc[sif_c2_ptc$comp_n_exons >= 3 &
                                sif_c2_ptc$comp_n_exons <= 12, ]
    if (nrow(candidates_d) > 0) {
      # Prefer high ptc_distance and multiple downstream EJCs for clear NMD
      candidates_d <- candidates_d[order(-candidates_d$n_downstream_ejcs,
                                          -candidates_d$ptc_distance), ]
      pick_d <- candidates_d[1, ]

      str_row_d <- structures[structures$isoform_id == pick_d$reference_isoform_id, ]
      strand_d <- str_row_d$strand[1]
      comp_cds_d <- get_cds_range(pick_d$comparator_isoform_id, cds)

      # PTC annotation
      ptc_pe_ann <- list()
      if (!is.null(comp_cds_d)) {
        ptc_pos_d <- if (strand_d == "+") comp_cds_d$cds_stop else comp_cds_d$cds_start
        ptc_pe_ann <- list(
          list(type = "ptc_marker", x = ptc_pos_d, track = "comp",
               label = "PTC", color = "#E74C3C")
        )
      }

      # Identify the event that introduces the PTC (SE, IR, etc.)
      profs_d <- set_results[[c2_key]]$profiles
      pair_prof_d <- profs_d[profs_d$reference_isoform_id == pick_d$reference_isoform_id &
                              profs_d$comparator_isoform_id == pick_d$comparator_isoform_id, ]
      pair_events_d <- if (nrow(pair_prof_d) > 0) pair_prof_d$detailed_events[[1]] else NULL

      # For zoom: find events closest to PTC position
      zoom_df_d <- NULL
      if (!is.null(pair_events_d) && nrow(pair_events_d) > 0 && !is.null(comp_cds_d)) {
        ptc_pos_d <- if (strand_d == "+") comp_cds_d$cds_stop else comp_cds_d$cds_start
        # Find events that overlap or are near the PTC
        ev_midpoints <- (pmin(pair_events_d$five_prime, pair_events_d$three_prime) +
                         pmax(pair_events_d$five_prime, pair_events_d$three_prime)) / 2
        nearest_idx <- which.min(abs(ev_midpoints - ptc_pos_d))
        nearest_ev <- pair_events_d[nearest_idx, , drop = FALSE]
        # Include PTC position in zoom range
        zoom_df_d <- data.frame(
          five_prime = c(pmin(nearest_ev$five_prime, nearest_ev$three_prime), ptc_pos_d),
          three_prime = c(pmax(nearest_ev$five_prime, nearest_ev$three_prime), ptc_pos_d)
        )
      }

      exon_diff_d <- pick_d$comp_n_exons - pick_d$ref_n_exons
      exon_note <- if (exon_diff_d > 0) {
        sprintf("comparator has %d extra exon(s)", exon_diff_d)
      } else if (exon_diff_d == 0) {
        "same exon count (retained intron or boundary change)"
      } else {
        sprintf("comparator has %d fewer exon(s)", abs(exon_diff_d))
      }

      gene_sym_d <- get_gene_sym(pick_d$gene_id)
      title_d <- sprintf("D) Poison Exon (In-Frame PTC): %s (%s)",
                         gene_sym_d, pick_d$gene_id)
      sub_d <- sprintf("Ref: %s | Comp: %s\nSame start, in frame, %s\nPTC distance = %d nt, %d downstream EJC(s)",
                       pick_d$reference_isoform_id, pick_d$comparator_isoform_id,
                       exon_note,
                       round(pick_d$ptc_distance), pick_d$n_downstream_ejcs)

      panel_d <- build_panel(pick_d, c2_key, "Dominant\n(non-NMD)", "NMD-sensitive",
                             title_d, sub_d, annotations = ptc_pe_ann,
                             zoom_events = zoom_df_d)
      example_plots <- c(example_plots, panel_d$plots)
    }
  }

  # Combine panels: full + zoom pairs
  if (length(example_plots) > 0) {
    if (requireNamespace("patchwork", quietly = TRUE)) {
      combined <- patchwork::wrap_plots(example_plots, ncol = 1) +
        patchwork::plot_annotation(
          title = "Reading Frame Consequences: Representative Gene Examples",
          theme = theme(plot.title = element_text(size = 14, face = "bold"))
        )
      print(combined)
    } else {
      for (pp in example_plots) print(pp)
    }
  }
}
```

## Stop Codon Distance to Last EJC

The distance from a transcript's stop codon to the last exon-exon junction (EJC) is a
key determinant of NMD susceptibility. When a stop codon sits >50 nt upstream of the last
EJC, the downstream junction(s) remain associated with the exon junction complex after
splicing, triggering NMD surveillance. Greater distance (more downstream EJCs) is generally
associated with more efficient NMD.

The distance is computed in mRNA coordinates as `last_EJC_position - stop_codon_position`.
Positive values indicate the stop codon is upstream of the last EJC (potential NMD trigger);
negative values indicate a stop codon downstream of the last EJC (normal termination).
Single-exon isoforms have no EJC and are excluded.

### Distribution of Stop-to-EJC Distance: All Comparators

Distribution of stop codon-to-last-EJC distances for **all** comparator isoforms (not just
PTC+), stratified by NMD and Control. This shows the full landscape of stop codon positioning
rather than conditioning on the PTC+ tail.

```{r sec5-stop-dist, results='asis', fig.height=6}
# Build stop-to-EJC distance data for all comparator isoforms — all_samples only
stop_dist_all <- {
  rows <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_allsamples")
    if (!key %in% names(set_results)) next
    comp_ids <- set_results[[key]]$profiles$comparator_isoform_id
    ptc_sub <- ptc[ptc$isoform_id %in% comp_ids & !is.na(ptc$ptc_distance), ]
    if (nrow(ptc_sub) == 0) next
    rows[[comp]] <- data.frame(
      comparison = comp_label(comp),
      ptc_distance = ptc_sub$ptc_distance,
      has_ptc = ptc_sub$has_ptc
    )
  }
  do.call(rbind, rows)
}

if (!is.null(stop_dist_all) && nrow(stop_dist_all) > 0 && has_ggplot2) {
  dist_summary <- stop_dist_all |>
    group_by(comparison) |>
    summarize(
      n_isoforms = n(),
      median_dist = round(median(ptc_distance), 0),
      pct_ptc_pos = round(100 * mean(has_ptc), 1),
      pct_negative = round(100 * mean(ptc_distance < 0), 1),
      .groups = "drop"
    )
  cat("**Columns:** **n_isoforms** = comparator isoforms with stop-to-EJC distance computed; ")
  cat("**median_dist** = median distance (nt); **pct_ptc_pos** = % with PTC (distance > 50 nt); ")
  cat("**pct_negative** = % with stop codon downstream of last EJC (normal termination).\n\n")
  knitr::kable(dist_summary, caption = "Stop Codon-to-Last EJC Distance: All Comparators (all_samples)") |> print()
  cat("\n\n")

  p_dist <- ggplot(stop_dist_all, aes(x = ptc_distance, fill = comparison)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
    annotate("text", x = 50, y = Inf, label = "PTC threshold (50 nt)",
             hjust = -0.05, vjust = 1.5, size = 3, color = "red") +
    scale_x_continuous(limits = c(
      quantile(stop_dist_all$ptc_distance, 0.01, na.rm = TRUE),
      quantile(stop_dist_all$ptc_distance, 0.99, na.rm = TRUE))) +
    theme_minimal() +
    labs(title = "Stop Codon Distance to Last EJC: NMD vs Control Comparators (all_samples)",
         subtitle = "All comparator isoforms (coding, multi-exon). Dashed red = 50 nt PTC threshold.",
         x = "Distance from stop codon to last EJC (nt)", y = "Density", fill = "")
  print(p_dist)
  cat("\n\n")
}
```

### NMD Response vs Stop Codon Distance (Meta-Analysis Average)

Among NMD-sensitive comparator isoforms, is there a relationship between the distance from
the stop codon to the last EJC and the magnitude of upregulation upon NMD inhibition? Rather
than showing per-cell-type scatter plots, this analysis uses the **mashr meta-analysis**
posterior mean averaged across 5 conditions (DD, AT, DO_ALI, FB, MV; DD_ALI excluded) to
produce a single consensus NMD response estimate per isoform. This leverages cross-condition
borrowing of information for more stable effect-size estimates.

```{r sec5-mashr-load, include=FALSE}
# Load mashr model for meta-analysis logFC
mashr_model_path <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/mashr/mashr_isoform_model_2026.3.10.rds"
mashr_model <- readRDS(mashr_model_path)

# Extract PosteriorMean matrix (isoforms x conditions)
post_mean <- ashr::get_pm(mashr_model)

# Average across 5 main conditions (exclude DD_ALI = Smg1i_in_DD_ALI)
meta_cols <- c("Smg1i_in_DD", "Smg1i_in_AT", "Smg1i_in_DO_ALI",
               "Smg1i_in_FB", "Smg1i_in_MV")
meta_logfc <- data.frame(
  txid = rownames(post_mean),
  mean_logFC = rowMeans(post_mean[, meta_cols, drop = FALSE])
)
```

```{r sec5-nmd-logfc-dist, results='asis', fig.height=6}
# Get all_samples NMD comparator isoform IDs
comp_ids_as <- set_results[["c2_allsamples"]]$profiles$comparator_isoform_id

# Merge meta-analysis logFC with PTC distance for all_samples comparators
ptc_sub <- ptc[ptc$isoform_id %in% comp_ids_as & !is.na(ptc$ptc_distance),
                c("isoform_id", "ptc_distance", "has_ptc")]
nmd_logfc_dist <- merge(meta_logfc[meta_logfc$txid %in% comp_ids_as, ],
                         ptc_sub, by.x = "txid", by.y = "isoform_id")

if (nrow(nmd_logfc_dist) > 0 && has_ggplot2) {
  p_logfc <- ggplot(nmd_logfc_dist, aes(x = ptc_distance, y = mean_logFC)) +
    geom_point(alpha = 0.15, size = 0.8) +
    geom_smooth(method = "loess", se = TRUE, color = "steelblue") +
    geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
    scale_x_continuous(limits = c(
      quantile(nmd_logfc_dist$ptc_distance, 0.01, na.rm = TRUE),
      quantile(nmd_logfc_dist$ptc_distance, 0.99, na.rm = TRUE))) +
    theme_minimal() +
    labs(title = sprintf("NMD Response vs Stop Codon Distance (all_samples, n = %d)",
                         nrow(nmd_logfc_dist)),
         subtitle = "Meta-analysis average (mashr PosteriorMean across DD, AT, DO_ALI, FB, MV). Blue = loess. Red = 50 nt PTC threshold.",
         x = "Distance from stop codon to last EJC (nt)",
         y = "Mean logFC (mashr, 5 conditions)")
  print(p_logfc)
  cat("\n\n")

  # Correlation summary
  cor_test <- cor.test(nmd_logfc_dist$ptc_distance, nmd_logfc_dist$mean_logFC,
                        method = "spearman")
  cor_summary <- data.frame(
    n = nrow(nmd_logfc_dist),
    spearman_rho = round(cor_test$estimate, 3),
    cor_p = cor_test$p.value
  )
  cat("**Columns:** **n** = NMD comparator isoforms (all_samples) with both meta-analysis logFC and stop-to-EJC distance; ")
  cat("**spearman_rho** = Spearman correlation between distance and mean logFC; ")
  cat("**cor_p** = p-value for the correlation.\n\n")
  knitr::kable(cor_summary,
               caption = "Correlation: Stop-to-EJC Distance vs Mean logFC (mashr meta-analysis, all_samples)") |> print()
  cat("\n\n")
}
```

### PTC Position Metrics: Which Best Predicts NMD Response?

The loess curve above suggests a non-linear relationship between stop codon distance and
logFC. To better understand what drives NMD efficiency, we compare three ways of measuring
PTC position using all_samples NMD comparators with mashr meta-analysis logFC:

- **PTC distance (nt)**: Raw nucleotide distance from stop codon to last EJC (truncated to 0–2000 nt to focus on the NMD-relevant range).
- **% CDS position**: Position of the PTC within the ORF as a percentage (`ptc_distance / orf_length * 100`), capped at 100%. Values that would exceed 100% arise when the distance from PTC to last EJC is longer than the ORF itself (i.e., many exons of 3' UTR lie downstream of an early PTC); these are set to 100%.
- **N downstream EJCs**: Number of exon-exon junctions downstream of the stop codon.

For each metric we fit a generalized additive model (GAM) to capture non-linear effects,
plus linear and quadratic models for comparison. All models are restricted to PTC+ isoforms
(distance > 50 nt) in the all_samples NMD comparison using mashr meta-analysis logFC.

```{r sec5-ptc-position-metrics, fig.height=5, fig.width=10}
# Merge PTC features with meta-analysis logFC for all_samples NMD comparators
as_comp_ids <- set_results[["c2_allsamples"]]$profiles$comparator_isoform_id

ptc_as <- ptc[ptc$isoform_id %in% as_comp_ids, ]
as_merged <- merge(meta_logfc[meta_logfc$txid %in% as_comp_ids, ], ptc_as,
                    by.x = "txid", by.y = "isoform_id")
as_merged <- as_merged[!is.na(as_merged$ptc_distance) & !is.na(as_merged$orf_length) &
                         as_merged$orf_length > 0, ]
as_merged$pct_cds_distance <- pmin(100, 100 * as_merged$ptc_distance / as_merged$orf_length)
as_merged$logFC <- as_merged$mean_logFC
as_ptc_pos <- as_merged[as_merged$has_ptc == TRUE, ]
```

```{r sec5-ptc-panel-plot, fig.height=5, fig.width=12}
if (has_ggplot2 && nrow(as_ptc_pos) > 0) {
  # PTC distance panel: restrict to 0-2000 nt
  dist_df <- data.frame(metric = "PTC distance (nt)", value = as_ptc_pos$ptc_distance,
                         logFC = as_ptc_pos$logFC)
  dist_df <- dist_df[dist_df$value <= 2000, ]

  ptc_long <- rbind(
    dist_df,
    data.frame(metric = "% CDS position", value = as_ptc_pos$pct_cds_distance,
               logFC = as_ptc_pos$logFC),
    data.frame(metric = "N downstream EJCs", value = as_ptc_pos$n_downstream_ejcs,
               logFC = as_ptc_pos$logFC)
  )
  ptc_long$metric <- factor(ptc_long$metric,
    levels = c("PTC distance (nt)", "% CDS position", "N downstream EJCs"))

  p_panel <- ggplot(ptc_long, aes(x = value, y = logFC)) +
    geom_point(alpha = 0.2, size = 0.8) +
    geom_smooth(method = "loess", span = 0.5, color = "steelblue", linewidth = 1.2) +
    facet_wrap(~metric, scales = "free_x") +
    theme_minimal() +
    labs(title = sprintf("all_samples NMD Comparators (PTC+, n = %d): Mean logFC vs PTC Position Metrics",
                         nrow(as_ptc_pos)),
         subtitle = "mashr meta-analysis mean logFC (DD, AT, DO_ALI, FB, MV)",
         x = "Metric value", y = "Mean logFC (mashr, 5 conditions)")
  print(p_panel)
}
```

```{r sec5-ejc-boxplot, fig.height=5, fig.width=9}
if (has_ggplot2 && nrow(as_ptc_pos) > 0) {
  ejc_counts <- as_ptc_pos |>
    count(n_downstream_ejcs) |>
    mutate(label = paste0("n=", n))

  p_box <- ggplot(as_ptc_pos, aes(x = factor(n_downstream_ejcs), y = logFC)) +
    geom_violin(fill = "lightblue", alpha = 0.5) +
    geom_boxplot(width = 0.15, outlier.alpha = 0.2) +
    geom_text(data = ejc_counts,
              aes(x = factor(n_downstream_ejcs),
                  y = max(as_ptc_pos$logFC) + 0.3, label = label),
              size = 2.8) +
    theme_minimal() +
    labs(title = "all_samples NMD Comparators: Mean logFC by Number of Downstream EJCs",
         subtitle = "mashr meta-analysis mean logFC rises sharply from 1-4 EJCs, then plateaus",
         x = "Number of downstream EJCs", y = "Mean logFC (mashr, 5 conditions)")
  print(p_box)
}
```

```{r sec5-gam-comparison, results='asis'}
if (requireNamespace("mgcv", quietly = TRUE) && nrow(as_ptc_pos) > 0) {
  n_ejc_unique <- length(unique(as_ptc_pos$n_downstream_ejcs))

  # GAM models
  gam_dist <- mgcv::gam(logFC ~ s(ptc_distance), data = as_ptc_pos)
  gam_pct  <- mgcv::gam(logFC ~ s(pct_cds_distance), data = as_ptc_pos)
  gam_ejc  <- mgcv::gam(logFC ~ s(n_downstream_ejcs, k = min(n_ejc_unique, 10)),
                          data = as_ptc_pos)
  gam_comb <- mgcv::gam(logFC ~ s(ptc_distance) +
                           s(n_downstream_ejcs, k = min(n_ejc_unique, 10)),
                          data = as_ptc_pos)

  # Linear and quadratic models
  lm_dist_lin <- lm(logFC ~ ptc_distance, data = as_ptc_pos)
  lm_pct_lin  <- lm(logFC ~ pct_cds_distance, data = as_ptc_pos)
  lm_ejc_lin  <- lm(logFC ~ n_downstream_ejcs, data = as_ptc_pos)
  lm_dist_q   <- lm(logFC ~ poly(ptc_distance, 2), data = as_ptc_pos)
  lm_pct_q    <- lm(logFC ~ poly(pct_cds_distance, 2), data = as_ptc_pos)
  lm_ejc_q    <- lm(logFC ~ poly(n_downstream_ejcs, 2), data = as_ptc_pos)

  model_comparison <- data.frame(
    metric = c(rep("ptc_distance (nt)", 3), rep("pct_cds_distance (%)", 3),
               rep("n_downstream_ejcs", 3), "combined (dist + ejcs)"),
    model = c(rep(c("Linear", "Quadratic", "GAM"), 3), "GAM"),
    R_squared = round(c(
      summary(lm_dist_lin)$r.squared, summary(lm_dist_q)$r.squared,
        summary(gam_dist)$r.sq,
      summary(lm_pct_lin)$r.squared, summary(lm_pct_q)$r.squared,
        summary(gam_pct)$r.sq,
      summary(lm_ejc_lin)$r.squared, summary(lm_ejc_q)$r.squared,
        summary(gam_ejc)$r.sq,
      summary(gam_comb)$r.sq
    ), 3),
    AIC = round(c(
      AIC(lm_dist_lin), AIC(lm_dist_q), AIC(gam_dist),
      AIC(lm_pct_lin), AIC(lm_pct_q), AIC(gam_pct),
      AIC(lm_ejc_lin), AIC(lm_ejc_q), AIC(gam_ejc),
      AIC(gam_comb)
    ), 0)
  )
  knitr::kable(model_comparison,
               caption = sprintf("Model Comparison: PTC Position Metrics vs Mean logFC (all_samples, n = %d PTC+ isoforms)",
                                  nrow(as_ptc_pos))) |> print()
}
```

The number of downstream exon-exon junctions (`n_downstream_ejcs`) is the strongest single
predictor of NMD response magnitude. Normalizing by CDS length
actually weakens the signal --- the absolute number of downstream junctions matters more than
the relative position within the transcript. The relationship is non-linear: logFC rises
sharply from 1 to ~4 downstream EJCs and then plateaus, consistent with EJC-based NMD
surveillance having diminishing returns beyond a threshold number of junctions. Quadratic
terms are highly significant for all three metrics, confirming the non-linear pattern.

## Frame-PTC Cross-Reference

PTC rates by frame category for both NMD and Control comparisons. This is the tabular form
of the PTC stacking in the Frame Category Distribution figure, showing how PTC prevalence
varies across frame categories and between comparison groups.

```{r sec5-frame-ptc-xref, results='asis'}
xref_rows <- compute_frame_ptc_xref_rows("all_samples")
if (!is.null(xref_rows)) {
  rownames(xref_rows) <- NULL
  knitr::kable(xref_rows, row.names = FALSE,
    caption = "PTC Rate by Frame Category: NMD vs Control (all_samples)") |> print()
  cat("\n\n")
}
```

## Frame Walk by Event Type

Breakdown of frameshift and compensatory rates by event type, comparing NMD and Control pairs.
Only events with CDS overlap are counted.

**Columns:** **event_type** = splicing event type; **n_cds_events** = number of events overlapping the CDS; **pct_frameshift** = % of those events that shift the reading frame (CDS overlap not divisible by 3); **pct_compensatory** = % that restore the frame after a prior shift. Suffixes **_NMD/_Ctrl** indicate the comparison group.

```{r sec5-frame-walk-by-event, results='asis'}
render_frame_walk("all_samples")
```

# 6. Cell-Type Results {.tabset}

Per-cell-type breakdowns of structural differences and PTC/reading frame analyses.
Each tab shows the same analyses as the all_samples sections above, restricted to a single
cell type.

```{r sec6-ct-results, results='asis', fig.width=10, fig.height=6}
for (ct in per_ct) {
  cat(sprintf("\n\n## %s {.tabset}\n\n", ct))

  # --- Structural Differences ---
  cat("\n\n### Structural Differences {.tabset}\n\n")

  cat("#### Event Frequency\n\n")
  render_event_frequency(ct)

  cat("\n\n#### Structural Divergence\n\n")
  div_row <- compute_divergence_row(ct)
  if (!is.null(div_row)) {
    knitr::kable(div_row, row.names = FALSE,
                 caption = sprintf("Structural Divergence by Region: NMD vs Control — %s", ct)) |> print()
    cat("\n\n")
  }

  cat("#### Gain/Loss Balance\n\n")
  gl_row <- compute_gain_loss_row(ct)
  if (!is.null(gl_row)) {
    knitr::kable(gl_row, caption = sprintf("Net BP Change: NMD vs Control — %s", ct)) |> print()
    cat("\n\n")
  }

  cat("#### Gain/Loss Direction\n\n")
  dc <- compute_gain_loss_direction(ct)
  if (!is.null(dc)) {
    render_dt(dc, caption = sprintf("Gain/Loss Direction: NMD vs Control — %s", ct))
  }
  cat("\n\n")

  cat("#### IR: Individual Types\n\n")
  ir_ind <- compute_ir_individual(ct)
  if (!is.null(ir_ind)) {
    knitr::kable(ir_ind, caption = sprintf("IR Event Direction: NMD vs Control — %s", ct)) |> print()
    cat("\n\n")
  }

  cat("#### IR: Cumulative Burden\n\n")
  ir_cum <- compute_ir_cumulative_row(ct)
  if (!is.null(ir_cum)) {
    knitr::kable(ir_cum, caption = sprintf("Cumulative IR Burden: NMD vs Control — %s", ct)) |> print()
    cat("\n\n")
  }

  cat("#### Distribution\n\n")
  if (has_ggplot2) {
    ct_label <- ct_to_label(ct)
    dist_rows <- list()
    for (comp in c("c2", "c4")) {
      key <- paste0(comp, "_", ct_label)
      if (!key %in% names(set_results)) next
      div <- set_results[[key]]$divergence
      if (is.null(div) || nrow(div) == 0) next
      dist_rows[[comp]] <- data.frame(comparison = comp_label(comp),
                                       pct_shared = div$pct_shared)
    }
    pct_shared_ct <- do.call(rbind, dist_rows)
    if (!is.null(pct_shared_ct) && nrow(pct_shared_ct) > 0) {
      p_dist <- ggplot(pct_shared_ct[!is.na(pct_shared_ct$pct_shared), ],
                       aes(x = pct_shared, fill = comparison)) +
        geom_density(alpha = 0.5) +
        theme_minimal() +
        labs(title = sprintf("Sequence Shared (%%): NMD vs Control — %s", ct),
             x = "pct_shared", y = "Density", fill = "")
      print(p_dist)
      cat("\n\n")
    }
  }

  # --- PTC & Reading Frame ---
  cat("\n\n### PTC & Reading Frame {.tabset}\n\n")

  cat("#### PTC Rates\n\n")
  ptc_row_ct <- compute_ptc_rates_row(ct)
  if (!is.null(ptc_row_ct)) {
    knitr::kable(ptc_row_ct, row.names = FALSE,
                 caption = sprintf("PTC Rates: NMD vs Control — %s", ct)) |> print()
    cat("\n\n")
  }

  cat("#### Frame Categories\n\n")
  fc_ct <- compute_frame_category_rows(ct)
  if (!is.null(fc_ct)) {
    knitr::kable(fc_ct, row.names = FALSE,
                 caption = sprintf("Frame Category Counts — %s", ct)) |> print()
    cat("\n\n")
  }

  cat("#### PTC Mechanism\n\n")
  mech_ct <- compute_ptc_mechanism_rows(ct)
  if (!is.null(mech_ct$mechanism)) {
    nmd_m <- mech_ct$mechanism[mech_ct$mechanism$comparison == "NMD", ]
    ctrl_m <- mech_ct$mechanism[mech_ct$mechanism$comparison == "Control", ]
    if (nrow(nmd_m) > 0 && nrow(ctrl_m) > 0) {
      mech_w <- data.frame(frame_category = nmd_m$frame_category,
                            NMD_n = nmd_m$n_ptc, NMD_pct = nmd_m$pct_of_ptc,
                            Ctrl_n = ctrl_m$n_ptc, Ctrl_pct = ctrl_m$pct_of_ptc)
      knitr::kable(mech_w, row.names = FALSE,
                   caption = sprintf("PTC Mechanism — %s", ct)) |> print()
      cat("\n\n")
    }
  }
  if (!is.null(mech_ct$fs_test)) {
    knitr::kable(mech_ct$fs_test, row.names = FALSE,
                 caption = sprintf("Frameshift Enrichment — %s", ct)) |> print()
    cat("\n\n")
  }

  cat("#### Frameshift Outcomes\n\n")
  fo_ct <- compute_frameshift_outcome_rows(ct)
  if (!is.null(fo_ct)) {
    fo_ct$in_frame_pct <- round(100 * fo_ct$n_in_frame / fo_ct$n_same_start, 1)
    fo_ct$fs_ptc_pct <- round(100 * fo_ct$n_fs_ptc / fo_ct$n_same_start, 1)
    fo_ct$fs_stop_last_pct <- round(100 * fo_ct$n_fs_stop_last / fo_ct$n_same_start, 1)
    fo_ct$fs_other_pct <- round(100 * fo_ct$n_fs_other / fo_ct$n_same_start, 1)
    knitr::kable(fo_ct[, c("comparison", "n_same_start",
                            "n_in_frame", "in_frame_pct",
                            "n_fs_ptc", "fs_ptc_pct",
                            "n_fs_stop_last", "fs_stop_last_pct",
                            "n_fs_other", "fs_other_pct")],
      row.names = FALSE,
      caption = sprintf("Frameshift Outcomes — %s", ct)) |> print()
    cat("\n\n")
  }

  cat("#### Novel Proteins\n\n")
  np_ct <- compute_novel_protein(ct)
  if (!is.null(np_ct) && nrow(np_ct) > 0) {
    c4_key_ct <- paste0("c4_", ct_to_label(ct))
    fc_ct_obj <- set_results[[c4_key_ct]]$frame_comp
    n_ss <- sum(fc_ct_obj$pair_summary$frame_category %in%
      c("same_start_in_frame", "same_start_frameshift"))
    np_sum <- data.frame(
      n_stable_novel = nrow(np_ct),
      pct_of_same_start = round(100 * nrow(np_ct) / n_ss, 1),
      median_novel_aa = round(median(np_ct$novel_aa), 0),
      median_pct_novel = round(median(np_ct$pct_novel), 1)
    )
    knitr::kable(np_sum, caption = sprintf("Novel Proteins — %s", ct)) |> print()
    cat("\n\n")
  }

  cat("#### Frame Walk\n\n")
  render_frame_walk(ct)
  cat("\n\n")

  # --- NMD Response vs Stop Distance (per-CT using mashr PosteriorMean) ---
  cat("### NMD Response vs Stop Distance\n\n")
  ct_to_mashr_col <- c(AT = "Smg1i_in_AT", DD = "Smg1i_in_DD", DD_ALI = "Smg1i_in_DD_ALI",
                         DO = "Smg1i_in_DO_ALI", FB = "Smg1i_in_FB", MV = "Smg1i_in_MV")
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  mashr_col <- ct_to_mashr_col[ct]
  if (!is.na(mashr_col) && c2_key %in% names(set_results) && has_ggplot2) {
    comp_ids_ct <- set_results[[c2_key]]$profiles$comparator_isoform_id
    ct_logfc <- data.frame(txid = rownames(post_mean),
                            logFC = post_mean[, mashr_col])
    ct_logfc <- ct_logfc[ct_logfc$txid %in% comp_ids_ct, ]
    ptc_sub_ct <- ptc[ptc$isoform_id %in% comp_ids_ct & !is.na(ptc$ptc_distance),
                        c("isoform_id", "ptc_distance", "has_ptc")]
    merged_ct <- merge(ct_logfc, ptc_sub_ct, by.x = "txid", by.y = "isoform_id")
    if (nrow(merged_ct) > 0) {
      p_ct <- ggplot(merged_ct, aes(x = ptc_distance, y = logFC)) +
        geom_point(alpha = 0.15, size = 0.8) +
        geom_smooth(method = "loess", se = TRUE, color = "steelblue") +
        geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
        scale_x_continuous(limits = c(
          quantile(merged_ct$ptc_distance, 0.01, na.rm = TRUE),
          quantile(merged_ct$ptc_distance, 0.99, na.rm = TRUE))) +
        theme_minimal() +
        labs(title = sprintf("NMD Response vs Stop Distance — %s (n = %d)", ct, nrow(merged_ct)),
             subtitle = sprintf("mashr PosteriorMean: %s", mashr_col),
             x = "Distance from stop codon to last EJC (nt)",
             y = sprintf("logFC (%s)", mashr_col))
      print(p_ct)
      cat("\n\n")
    }
  }
}
```

# 7. Cross-CT Consistency

## Event Frequency Heatmap

```{r sec7-heatmap, results='asis', fig.height=8}
if (length(main_per_ct) >= 2) {

  compute_freq_vec <- function(ct) {
    p <- profiles_list[[paste0("c2_", ct_to_label(ct))]]
    n <- nrow(p)
    n_pir5 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_5")))
    n_pir3 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_3")))
    data.frame(cell_type = ct, event_type = evt_types,
      pct = round(c(100*sum(p$tss_changed)/n, 100*sum(p$tes_changed)/n,
              100*sum(p$n_a5ss>0)/n, 100*sum(p$n_a3ss>0)/n,
              100*sum(p$n_se>0)/n, 100*sum(p$n_missing_internal>0)/n,
              100*sum(p$n_ir>0)/n, 100*sum(p$n_ir_diff>0)/n,
              100*n_pir5/n, 100*n_pir3/n), 1))
  }

  freq_matrix <- do.call(rbind, lapply(main_per_ct, compute_freq_vec))

  if (has_ggplot2) {
    p_heat <- ggplot(freq_matrix, aes(x = event_type, y = cell_type, fill = pct)) +
      geom_tile(color = "white") +
      geom_text(aes(label = sprintf("%.0f", pct)), size = 3) +
      scale_fill_gradient(low = "white", high = "steelblue", name = "% of pairs") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      labs(title = "NMD Event Frequency by Cell Type (pair-based %)",
           x = "Event Type", y = "Cell Type")
    print(p_heat)
    cat("\n\n")
  }

  freq_wide <- tidyr::pivot_wider(freq_matrix, names_from = event_type, values_from = pct)
  knitr::kable(freq_wide, caption = "NMD Event Frequency (% of pairs) by Cell Type") |> print()
  cat("\n\n")

} else {
  cat("Fewer than 2 main cell types with NMD profiles; skipping cross-cell-type analysis.\n\n")
}
```

## Heterogeneity Test

Tests whether the distribution of event types differs significantly across cell types.
A Pearson's chi-square test on the cell-type x event-type contingency table (raw pair
counts, not percentages). A significant result indicates that event type prevalence varies
across cell types beyond what would be expected by chance.

```{r sec7-heterogeneity, results='asis'}
if (length(main_per_ct) >= 2) {

  freq_count_list <- lapply(main_per_ct, function(ct) {
    p <- profiles_list[[paste0("c2_", ct_to_label(ct))]]
    n_pir5 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_5")))
    n_pir3 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_3")))
    c(sum(p$tss_changed), sum(p$tes_changed),
      sum(p$n_a5ss>0), sum(p$n_a3ss>0),
      sum(p$n_se>0), sum(p$n_missing_internal>0),
      sum(p$n_ir>0), sum(p$n_ir_diff>0),
      n_pir5, n_pir3)
  })
  freq_count_matrix <- do.call(rbind, freq_count_list)
  rownames(freq_count_matrix) <- main_per_ct
  colnames(freq_count_matrix) <- evt_types

  chi_main <- tryCatch(chisq.test(freq_count_matrix), error = function(e) NULL)
  if (!is.null(chi_main))
    cat(sprintf("**Main cell types** (AT, DD, DO, FB, MV): Chi-square = %.1f, df = %d, p = %s\n\n",
                chi_main$statistic, chi_main$parameter, format(chi_main$p.value, digits = 3)))

} else {
  cat("Fewer than 2 main cell types with NMD profiles; skipping heterogeneity test.\n\n")
}
```

# 8. Meta-Analysis

Random-effects meta-analysis pools NMD vs Control effect sizes across main cell types
(AT, DD, DO, FB, MV). DD_ALI is excluded due to reversed biological patterns that inflate
between-cell-type heterogeneity. all_samples is excluded as a superset.

## Pooled Effect Sizes Across Cell Types

Each metric from the NMD vs Control comparison is pooled across the five main cell types
using a random-effects model (metafor package, REML estimation). The table reports the
metrics discussed in earlier sections of this report, restricted to those with directional
interpretability.

**Effect size types and direction:**

- **Cliff's delta** (range -1 to +1): Nonparametric effect size for continuous comparisons.
  Positive values indicate NMD > Control; negative values indicate Control > NMD.
  - **event_count**: Positive = NMD pairs have more structural events per pair.
  - **gain_loss**: Positive = NMD pairs have higher net_bp (comparator shorter); negative = NMD comparator is longer.
- **Log odds ratio (log-OR)**: For binary outcomes. Positive = enriched in NMD; negative = enriched in Control.
  Exponentiate to get the odds ratio (e.g., log-OR of 2.3 = OR of ~10).
  - **ptc_rate**: Positive = PTC more frequent in NMD comparators.
  - **event_freq: {type}**: Positive = event type more prevalent in NMD pairs.

**Columns:** **metric** = analysis metric; **effect_size_type** = scale of the estimate;
**pooled_estimate** = meta-analytic effect size on the native scale (log-OR or Cliff's delta);
**OR** = exponentiated odds ratio for log-OR metrics (>1 = enriched in NMD; <1 = enriched in Control),
shown as "---" for Cliff's delta metrics;
**ci_lower/ci_upper** = 95% confidence interval on the native scale;
**p_value** = test of pooled effect = 0;
**I2** = heterogeneity (% of variance due to between-CT differences; >75% = high).

```{r sec8-meta, results='asis'}
if (requireNamespace("metafor", quietly = TRUE)) {
  # Use cached comparePairSets results — pool main CTs only (exclude DD_ALI + all_samples)
  cps_list <- cross_results[intersect(names(cross_results), main_cts)]
  cps_list <- cps_list[!sapply(cps_list, is.null)]

  if (length(cps_list) >= 2) {
    cat(sprintf("**Cell types pooled:** %s\n\n", paste(names(cps_list), collapse = ", ")))

    meta <- tryCatch(
      metaAnalyzePairSets(cps_list, verbose = FALSE),
      error = function(e) {
        cat(sprintf("Meta-analysis failed: %s\n\n", e$message))
        NULL
      }
    )

    if (!is.null(meta)) {
      # Only include metrics that are discussed in the report
      keep_metrics <- c("event_count", "gain_loss", "ptc_rate")
      scalar_sections <- intersect(
        names(meta)[vapply(meta, function(m) {
          !is.null(m$estimate) && !isTRUE(m$skipped)
        }, logical(1))],
        keep_metrics
      )

      meta_summary <- do.call(rbind, lapply(scalar_sections, function(mn) {
        m <- meta[[mn]]
        es_type <- switch(mn,
          event_count = "Cliff's delta",
          gain_loss = "Cliff's delta",
          ptc_rate = "log-OR",
          "---"
        )
        # For log-OR metrics, also show the exponentiated OR
        or_col <- if (es_type == "log-OR" && !is.null(m$pooled_or)) {
          sprintf("%.1f", m$pooled_or)
        } else { "---" }
        data.frame(
          metric = mn,
          effect_size_type = es_type,
          pooled_estimate = format(m$estimate, digits = 3),
          OR = or_col,
          ci_lower = format(m$ci_lower, digits = 3),
          ci_upper = format(m$ci_upper, digits = 3),
          p_value = format(m$p_value, digits = 3),
          I2 = sprintf("%.1f%%", m$I2)
        )
      }))

      # Add event_frequency per-event-type results (all log-OR)
      if (!is.null(meta$event_frequency) && is.list(meta$event_frequency)) {
        ef_rows <- do.call(rbind, lapply(names(meta$event_frequency), function(et) {
          m <- meta$event_frequency[[et]]
          if (isTRUE(m$skipped) || is.null(m$estimate)) return(NULL)
          or_col <- if (!is.null(m$pooled_or)) sprintf("%.2f", m$pooled_or) else
                      sprintf("%.2f", exp(m$estimate))
          data.frame(
            metric = paste0("event_freq: ", et),
            effect_size_type = "log-OR",
            pooled_estimate = format(m$estimate, digits = 3),
            OR = or_col,
            ci_lower = format(m$ci_lower, digits = 3),
            ci_upper = format(m$ci_upper, digits = 3),
            p_value = format(m$p_value, digits = 3),
            I2 = sprintf("%.1f%%", m$I2)
          )
        }))
        if (!is.null(ef_rows)) meta_summary <- rbind(meta_summary, ef_rows)
      }

      render_dt(meta_summary, caption = "Random-Effects Meta-Analysis (AT, DD, DO, FB, MV)")
    }
  } else {
    cat("Insufficient cell types for meta-analysis (need >= 2).\n\n")
  }
} else {
  cat("metafor package not available; skipping meta-analysis.\n\n")
}
```

# 9. DD_ALI

In AT, DD, DO, FB, and MV, NMD deviates from baseline in a consistent direction: more frameshifts,
more PTCs, more exon skipping, comparators longer, changes more targeted. DD_ALI reverses nearly
every one of these deviations --- its NMD transitions look more like an *amplification* of
baseline patterns than a departure from them.

Note: DD_ALI profiles shown here are from the **limma** classification (the original analysis),
since DD_ALI shows no meaningful PTC enrichment under any DE method and is excluded from all
main analyses. The comparison against main cell types uses mashr-classified results for the
main CTs.

```{r sec9-ddali-setup, include=FALSE}
ali_label <- ct_to_label(ali_ct)
c2_ali <- paste0("c2_", ali_label)
c4_ali <- paste0("c4_", ali_label)
has_ali <- c2_ali %in% names(set_results) && c4_ali %in% names(set_results)
```

## Key Metrics: DD_ALI vs Main Cell Types

```{r sec9-key-metrics, results='asis'}
if (has_ali) {

  range_str <- function(vals) {
    vals <- vals[!is.na(vals)]
    if (length(vals) == 0) return("\u2014")
    sprintf("%.0f\u2013%.0f", min(vals), max(vals))
  }

  ct_metrics <- lapply(main_per_ct, function(ct) {
    ct_label <- ct_to_label(ct)
    c2k <- paste0("c2_", ct_label); c4k <- paste0("c4_", ct_label)
    nmd_ptc <- set_results[[c2k]]$ptc_assoc
    ctrl_ptc <- set_results[[c4k]]$ptc_assoc
    nmd_div <- set_results[[c2k]]$divergence
    ctrl_div <- set_results[[c4k]]$divergence
    nmd_fw <- build_fw_row(c2k); ctrl_fw <- build_fw_row(c4k)
    list(
      nmd_ptc = if (!is.null(nmd_ptc)) 100 * nmd_ptc$comp_ptc_rate else NA,
      ctrl_ptc = if (!is.null(ctrl_ptc)) 100 * ctrl_ptc$comp_ptc_rate else NA,
      nmd_fw = if (!is.null(nmd_fw)) nmd_fw$pct_with_frameshift else NA,
      ctrl_fw = if (!is.null(ctrl_fw)) ctrl_fw$pct_with_frameshift else NA,
      nmd_net_bp = median(set_results[[c2k]]$profiles$net_bp),
      ctrl_net_bp = median(set_results[[c4k]]$profiles$net_bp),
      nmd_pct_shared = if (!is.null(nmd_div) && nrow(nmd_div) > 0)
        median(nmd_div$pct_shared, na.rm = TRUE) else NA,
      ctrl_pct_shared = if (!is.null(ctrl_div) && nrow(ctrl_div) > 0)
        median(ctrl_div$pct_shared, na.rm = TRUE) else NA
    )
  })

  ali_ptc <- set_results[[c2_ali]]$ptc_assoc
  ali_div <- set_results[[c2_ali]]$divergence
  ali_fw <- build_fw_row(c2_ali)
  ali_ptc_val <- if (!is.null(ali_ptc)) round(100 * ali_ptc$comp_ptc_rate, 1) else NA
  ali_fw_val <- if (!is.null(ali_fw)) ali_fw$pct_with_frameshift else NA
  ali_net_bp_val <- round(median(set_results[[c2_ali]]$profiles$net_bp), 0)
  ali_pct_shared_val <- if (!is.null(ali_div) && nrow(ali_div) > 0)
    round(median(ali_div$pct_shared, na.rm = TRUE), 1) else NA

  ddali_summary <- data.frame(
    metric = c("comp_ptc_%", "frameshift_%", "med_net_bp", "med_pct_shared"),
    Baseline_Ctrl = c(
      range_str(sapply(ct_metrics, `[[`, "ctrl_ptc")),
      range_str(sapply(ct_metrics, `[[`, "ctrl_fw")),
      range_str(sapply(ct_metrics, `[[`, "ctrl_net_bp")),
      range_str(sapply(ct_metrics, `[[`, "ctrl_pct_shared"))
    ),
    Main_CTs_NMD = c(
      range_str(sapply(ct_metrics, `[[`, "nmd_ptc")),
      range_str(sapply(ct_metrics, `[[`, "nmd_fw")),
      range_str(sapply(ct_metrics, `[[`, "nmd_net_bp")),
      range_str(sapply(ct_metrics, `[[`, "nmd_pct_shared"))
    ),
    DD_ALI_NMD = c(
      if (!is.na(ali_ptc_val)) sprintf("%.1f", ali_ptc_val) else "\u2014",
      if (!is.na(ali_fw_val)) sprintf("%.1f", ali_fw_val) else "\u2014",
      sprintf("%.0f", ali_net_bp_val),
      if (!is.na(ali_pct_shared_val)) sprintf("%.1f", ali_pct_shared_val) else "\u2014"
    ),
    DD_ALI_direction = c(
      "\u2248 baseline",
      "\u2248 baseline",
      "Amplified baseline",
      "Reversed"
    )
  )
  render_dt(ddali_summary, caption = "DD_ALI Reversal: Key Metrics")
  cat("\n\n")

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

## PTC & Frameshift

```{r sec9-ptc-frameshift, results='asis'}
if (has_ali) {

  ali_pa_c2 <- set_results[[c2_ali]]$ptc_assoc
  ali_pa_c4 <- set_results[[c4_ali]]$ptc_assoc
  ali_cps <- cross_results[[ali_ct]]
  ali_ptc_cps <- if (!is.null(ali_cps)) ali_cps$ptc_rate else NULL

  # Need ali_fw for this chunk too
  ali_fw <- build_fw_row(c2_ali)

  ptc_row <- data.frame(
    metric = c("Comparator PTC %", "Frameshift %", "Unresolved %"),
    NMD = c(
      round(100 * ali_pa_c2$comp_ptc_rate, 1),
      if (!is.null(ali_fw)) ali_fw$pct_with_frameshift else NA,
      if (!is.null(ali_fw)) ali_fw$pct_unresolved else NA
    ),
    Control = c(
      round(100 * ali_pa_c4$comp_ptc_rate, 1),
      {fw4 <- build_fw_row(c4_ali); if (!is.null(fw4)) fw4$pct_with_frameshift else NA},
      {if (!is.null(fw4)) fw4$pct_unresolved else NA}
    )
  )
  knitr::kable(ptc_row, caption = "DD_ALI: PTC & Frameshift Rates") |> print()
  cat("\n")

  if (!is.null(ali_ptc_cps) && !isTRUE(ali_ptc_cps$skipped))
    cat(sprintf("PTC Fisher's exact: OR = %.2f, p = %s\n\n",
                ali_ptc_cps$odds_ratio, format(ali_ptc_cps$p_value, digits = 3)))

  # Stop-to-EJC distance for DD_ALI comparators
  ali_comp_c2 <- set_results[[c2_ali]]$profiles$comparator_isoform_id
  ali_comp_c4 <- set_results[[c4_ali]]$profiles$comparator_isoform_id
  ali_ptc_c2 <- ptc[ptc$isoform_id %in% ali_comp_c2 & !is.na(ptc$ptc_distance), ]
  ali_ptc_c4 <- ptc[ptc$isoform_id %in% ali_comp_c4 & !is.na(ptc$ptc_distance), ]
  if (nrow(ali_ptc_c2) > 0 && nrow(ali_ptc_c4) > 0) {
    wt <- wilcox.test(ali_ptc_c2$ptc_distance, ali_ptc_c4$ptc_distance)
    cat(sprintf("Stop-to-EJC distance (all comparators): NMD median = %.0f nt, Control = %.0f nt, Wilcoxon p = %s\n\n",
                median(ali_ptc_c2$ptc_distance), median(ali_ptc_c4$ptc_distance),
                format(wt$p.value, digits = 3)))
  }

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

## Event Frequency

```{r sec9-event-freq, results='asis'}
if (has_ali) {

  ali_cps <- cross_results[[ali_ct]]
  ali_ef <- if (!is.null(ali_cps)) ali_cps$event_frequency else NULL
  if (!is.null(ali_ef) && !isTRUE(ali_ef$skipped)) {
    ef_df <- do.call(rbind, lapply(ali_ef$per_event, function(e) {
      dir_label <- gsub("enriched in A", "enriched in NMD",
                   gsub("enriched in B", "enriched in Control", e$direction))
      data.frame(
        event_type = e$event_type,
        NMD_n = e$count_a, NMD_pct = round(100 * e$count_a / e$n_a, 1),
        Ctrl_n = e$count_b, Ctrl_pct = round(100 * e$count_b / e$n_b, 1),
        OR = round(e$odds_ratio, 2),
        adj_p = e$adj_p_value,
        direction = dir_label
      )
    }))
    render_dt(ef_df, caption = "DD_ALI: Event Type Prevalence NMD vs Control")
    cat("\n")
  }

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

In main cell types, SE is enriched in NMD. In DD_ALI, SE is enriched in *Control*,
consistent with a fundamentally different NMD mechanism.

## Structural Divergence & Gain/Loss

```{r sec9-divergence, results='asis'}
if (has_ali) {

  ali_div_c2 <- set_results[[c2_ali]]$divergence
  ali_div_c4 <- set_results[[c4_ali]]$divergence
  if (!is.null(ali_div_c2) && nrow(ali_div_c2) > 0 &&
      !is.null(ali_div_c4) && nrow(ali_div_c4) > 0) {
    div_row <- data.frame(
      metric = c("med_pct_shared", "med_|exonic_diff|", "med_net_bp"),
      NMD = c(
        round(median(ali_div_c2$pct_shared, na.rm = TRUE), 1),
        round(median(abs(ali_div_c2$exonic_bp_diff)), 0),
        round(median(set_results[[c2_ali]]$profiles$net_bp), 0)
      ),
      Control = c(
        round(median(ali_div_c4$pct_shared, na.rm = TRUE), 1),
        round(median(abs(ali_div_c4$exonic_bp_diff)), 0),
        round(median(set_results[[c4_ali]]$profiles$net_bp), 0)
      )
    )
    knitr::kable(div_row, caption = "DD_ALI: Structural Divergence") |> print()
    cat("\n")
  }

  ali_cps <- cross_results[[ali_ct]]
  ali_gl <- if (!is.null(ali_cps)) ali_cps$gain_loss else NULL
  if (!is.null(ali_gl) && !isTRUE(ali_gl$skipped))
    cat(sprintf("Gain/loss: Wilcoxon p = %s, d = %.3f\n\n",
                format(ali_gl$p_value, digits = 3), ali_gl$cliffs_d))

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

In main cell types, NMD pairs share MORE sequence than baseline (~70% vs ~55%). DD_ALI
reverses this: NMD pairs share LESS sequence (~28%). DD_ALI NMD comparators are also
much longer than their reference (large positive net_bp), amplifying the baseline pattern
rather than reversing it.

## Gene Overlap with Main CTs

```{r sec9-gene-overlap, results='asis'}
if (has_ali) {

  ali_genes <- unique(profiles_list[[c2_ali]]$gene_id)
  main_gene_lists <- lapply(main_per_ct, function(ct)
    unique(profiles_list[[paste0("c2_", ct_to_label(ct))]]$gene_id))
  names(main_gene_lists) <- main_per_ct

  ali_overlap <- data.frame(
    cell_type = main_per_ct,
    ddali_genes = length(ali_genes),
    shared = sapply(main_per_ct, function(ct)
      length(intersect(ali_genes, main_gene_lists[[ct]]))),
    pct_of_ddali = sapply(main_per_ct, function(ct)
      round(100 * length(intersect(ali_genes, main_gene_lists[[ct]])) / length(ali_genes), 1))
  )
  render_dt(ali_overlap, caption = "DD_ALI Gene Overlap with Main CTs")
  cat("\n\n")

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

DD_ALI NMD-sensitive isoforms resemble highly divergent baseline isoforms (low sequence
sharing, high retention) rather than the targeted exon-skipping pattern seen in other cell
types. This suggests DD_ALI NMD targets arise through terminal/retention mechanisms rather
than frameshift-driven PTC introduction, possibly reflecting ALI culture conditions.

# 10. Key Figures


```{r sec10-key-figures, eval=has_ggplot2, fig.height=8}
if (has_ggplot2 && length(profiles_list) > 0) {

  # --- Figure 1: Event frequency by comparison type ---
  event_freq <- do.call(rbind, lapply(names(profiles_list), function(key) {
    p <- profiles_list[[key]]
    parts <- strsplit(key, "_")[[1]]
    s <- summarizeProfiles(p)
    data.frame(
      comparison = comp_label(parts[1]),
      cell_type = paste(parts[-1], collapse = "_"),
      event_type = names(s$event_counts),
      count = as.integer(s$event_counts),
      total = s$total_events
    )
  }))
  event_freq$pct <- 100 * event_freq$count / event_freq$total

  p1 <- ggplot(event_freq, aes(x = event_type, y = pct, fill = comparison)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~cell_type, scales = "free_y") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "Event Type Distribution by Comparison",
         subtitle = "Raw event counts (not pair-based percentages — see Section 4 tables)",
         x = "Event Type", y = "Percentage of Total Events",
         fill = "Comparison")
  print(p1)

}
```

# 11. Session Info

```{r session-info}
sessionInfo()
```
