---
title: "NMD Isoform Transitions Analysis (mashr Classification)"
subtitle: "Isopair Wrapper — Isocall Pipeline, mashr DIE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(Isopair)
library(dplyr)
library(tibble)
library(tidyr)
library(DT)
library(htmltools)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
if (has_ggplot2) library(ggplot2)
source("visualization_functions.R")

# DT rendering helper for results='asis' chunks
# Uses renderTags() + knit_meta_add() to register JS/CSS dependencies with knitr,
# so widgets render even when they are not the last expression in a chunk.
render_dt <- function(df, caption = NULL, ...) {
  widget <- datatable(df, caption = caption, rownames = FALSE,
                      class = "compact stripe",
                      options = list(pageLength = 25, dom = 'ft'), ...)
  rendered <- htmltools::renderTags(widget)
  knitr::knit_meta_add(rendered$dependencies)
  cat(rendered$html)
  cat("\n")
}

# Caching infrastructure: wrap expensive computations with cached_compute()
cached_compute <- function(key, expr, cache_dir = file.path(data_dir, "analysis_cache"),
                           force_recompute = FALSE) {
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
  cache_file <- file.path(cache_dir, paste0(key, ".rds"))
  if (!force_recompute && file.exists(cache_file)) {
    return(readRDS(cache_file))
  }
  result <- force(expr)
  saveRDS(result, cache_file)
  result
}

# Utility functions
ct_to_label <- function(ct) tolower(gsub("_", "", ct))

evt_types <- c("Alt_TSS", "Alt_TES", "A5SS", "A3SS", "SE",
               "Missing_Internal", "IR", "IR_diff",
               "Partial_IR_5", "Partial_IR_3")

compute_event_freq <- function(profiles) {
  n_pairs <- nrow(profiles)
  n_pir5 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_5")))
  n_pir3 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_3")))
  counts <- c(sum(profiles$tss_changed), sum(profiles$tes_changed),
              sum(profiles$n_a5ss > 0), sum(profiles$n_a3ss > 0),
              sum(profiles$n_se > 0), sum(profiles$n_missing_internal > 0),
              sum(profiles$n_ir > 0), sum(profiles$n_ir_diff > 0),
              n_pir5, n_pir3)
  data.frame(event_type = evt_types,
             n_pairs_with_event = counts,
             pct_of_pairs = round(100 * counts / n_pairs, 1))
}

# Display labels: internal c2/c4 → user-facing NMD/Control
comp_label <- function(comp) {
  switch(toupper(comp), C2 = "NMD", C4 = "Control", toupper(comp))
}

parse_key <- function(key) {
  parts <- strsplit(key, "_")[[1]]
  list(comp = comp_label(parts[1]),
       ct = paste(parts[-1], collapse = "_"))
}

# Count CDS bp in the comparator isoform upstream (5') of the first frameshift boundary.
# Used to determine how much of the protein is conserved vs novel after a frameshift.
count_conserved_cds_bp <- function(comp_id, fs_boundary, strand, cds_df, str_df) {
  cds_row <- cds_df[cds_df$isoform_id == comp_id, ]
  str_row <- str_df[str_df$isoform_id == comp_id, ]
  if (nrow(cds_row) == 0 || nrow(str_row) == 0) return(NA_integer_)
  cds_start <- cds_row$cds_start[1]
  cds_stop  <- cds_row$cds_stop[1]
  orf_len   <- cds_row$orf_length[1]
  e_starts  <- str_row$exon_starts[[1]]
  e_ends    <- str_row$exon_ends[[1]]

  # Order exons 5'→3'
  if (strand == "-") {
    ord <- order(e_starts, decreasing = TRUE)
  } else {
    ord <- order(e_starts)
  }
  e_starts <- e_starts[ord]
  e_ends   <- e_ends[ord]

  conserved <- 0L
  for (i in seq_along(e_starts)) {
    es <- e_starts[i]
    ee <- e_ends[i]
    # Intersect exon with CDS range
    cs <- max(es, cds_start)
    ce <- min(ee, cds_stop)
    if (cs > ce) next  # No CDS overlap

    if (strand == "+") {
      if (fs_boundary <= cs) break
      if (fs_boundary >= ce) {
        conserved <- conserved + (ce - cs + 1L)
      } else {
        conserved <- conserved + (fs_boundary - cs)
        break
      }
    } else {
      if (fs_boundary >= ce) break
      if (fs_boundary <= cs) {
        conserved <- conserved + (ce - cs + 1L)
      } else {
        conserved <- conserved + (ce - fs_boundary)
        break
      }
    }
  }
  conserved <- min(conserved, orf_len)
  as.integer(conserved)
}

# Build frame walk summary row for any profile key
# Restrict frame walk summary to pairs with coding references (frame_resolved not NA)
build_fw_row <- function(key) {
  fw <- set_results[[key]]$frame_walk
  if (is.null(fw) || is.null(fw$profile_summary) || nrow(fw$profile_summary) == 0) return(NULL)
  ps <- fw$profile_summary
  # Restrict to coding pairs: frame_resolved is NA when reference is non-coding
  ps <- ps[!is.na(ps$frame_resolved), ]
  if (nrow(ps) == 0) return(NULL)
  n <- nrow(ps)
  has_shift <- ps$n_frameshifts > 0
  n_shift <- sum(has_shift, na.rm = TRUE)
  n_unres <- sum(has_shift & ps$frame_resolved == FALSE, na.rm = TRUE)
  n_comp <- sum(ps$n_compensatory > 0, na.rm = TRUE)
  tibble::tibble(
    n_pairs = n,
    median_splice_groups = round(median(ps$n_splice_groups, na.rm = TRUE), 1),
    n_with_frameshift = n_shift,
    pct_with_frameshift = round(100 * n_shift / n, 1),
    n_unresolved = n_unres,
    pct_unresolved = round(100 * n_unres / n, 1),
    n_compensated = n_comp,
    pct_compensated = if (n_shift > 0) round(100 * n_comp / n_shift, 1) else 0,
    median_pct_orf_frameshifted = round(
      median(ps$pct_orf_frameshifted[ps$n_frameshifts > 0], na.rm = TRUE), 1)
  )
}
```

# 1. Setup

This report analyzes structural differences between isoform pairs using the Isopair package,
with NMD classification derived from **mashr** (multivariate adaptive shrinkage) differential
isoform expression results. Mashr logFC values are shrunken relative to limma, but NMD
classification uses the pre-computed `nmd_responsive` flag (adj.P.Val < 0.05 & logFC > 0).
Main analyses cover all_samples (union NMD / intersection non-NMD) plus AT, DD, DO, FB, and MV.
DD_ALI is analyzed separately (Section 7) due to reversed biological patterns.

```{r load-data}
data_dir <- "data_mashr"

# Infrastructure
structures     <- readRDS(file.path(data_dir, "structures.rds"))
union_exons    <- readRDS(file.path(data_dir, "union_exons.rds"))
ue_mapping     <- readRDS(file.path(data_dir, "isoform_union_mapping.rds"))
cds            <- readRDS(file.path(data_dir, "cds.rds"))
cds_exons      <- readRDS(file.path(data_dir, "cds_exons.rds"))
annotated_ue   <- readRDS(file.path(data_dir, "annotated_ue.rds"))
ptc            <- readRDS(file.path(data_dir, "ptc.rds"))

# Study data
expr_mat       <- readRDS(file.path(data_dir, "expression_data.rds"))
sample_meta    <- readRDS(file.path(data_dir, "sample_metadata.rds"))
gene_map       <- readRDS(file.path(data_dir, "gene_map.rds"))
nmd_class      <- readRDS(file.path(data_dir, "nmd_classification.rds"))

# Build gene_id → gene_symbol lookup from mashr DE files
de_dir_mashr <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/mashr"
gene_symbol_lookup <- character(0)
for (de_f in list.files(de_dir_mashr, pattern = "^nmd_mashr_die_.*\\.csv$", full.names = TRUE)) {
  de_tmp <- read.csv(de_f, stringsAsFactors = FALSE)[, c("gene_id", "hgnc_symbol")]
  de_tmp <- de_tmp[!duplicated(de_tmp$gene_id) & !is.na(de_tmp$hgnc_symbol) &
                    de_tmp$hgnc_symbol != "", ]
  new_entries <- setdiff(de_tmp$gene_id, names(gene_symbol_lookup))
  gene_symbol_lookup[de_tmp$gene_id[de_tmp$gene_id %in% new_entries]] <-
    de_tmp$hgnc_symbol[de_tmp$gene_id %in% new_entries]
}

# Cell types with profiles (mashr: DO included, DD_ALI excluded from main)
cell_types <- c("all_samples", "AT", "DD", "DO", "FB", "MV")
comparisons <- c("c2", "c4")

# DD_ALI separation: main cell types for cross-CT analyses and meta-analysis
main_cts <- c("AT", "DD", "DO", "FB", "MV")
ali_ct <- "DD_ALI"
limma_data_dir <- "data"

# Load all available profiles
profiles_list <- list()
pairs_list    <- list()
for (ct in cell_types) {
  ct_label <- tolower(gsub("_", "", ct))
  for (comp in comparisons) {
    key <- paste(comp, ct_label, sep = "_")
    prof_file <- file.path(data_dir, sprintf("profiles_%s.rds", key))
    pair_file <- file.path(data_dir, sprintf("pairs_%s.rds", key))
    if (file.exists(prof_file)) {
      profiles_list[[key]] <- readRDS(prof_file)
      if (file.exists(pair_file)) pairs_list[[key]] <- readRDS(pair_file)
    }
  }
}

# Load DD_ALI profiles from original limma data directory for the DD_ALI section
for (comp in comparisons) {
  ali_label <- tolower(gsub("_", "", ali_ct))
  key <- paste(comp, ali_label, sep = "_")
  prof_file <- file.path(limma_data_dir, sprintf("profiles_%s.rds", key))
  pair_file <- file.path(limma_data_dir, sprintf("pairs_%s.rds", key))
  if (file.exists(prof_file)) {
    profiles_list[[key]] <- readRDS(prof_file)
    if (file.exists(pair_file)) pairs_list[[key]] <- readRDS(pair_file)
  }
}

cat(sprintf("Loaded %d profile sets across %d cell types (including DD_ALI from limma)\n",
            length(profiles_list), length(cell_types) + 1L))
```

```{r gene-match}
# Bidirectional gene-matching: restrict NMD and Control to the intersection of
# (gene_id, reference_isoform_id) pairs present in BOTH sets. After this step,
# every gene in NMD has a corresponding Control pair and vice versa.
gene_match_summary <- list()
# Include DD_ALI in gene-matching (loaded from limma) for the dedicated section
for (ct in unique(c(cell_types, ali_ct))) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(profiles_list) || !c4_key %in% names(profiles_list)) next

  nmd_before <- nrow(profiles_list[[c2_key]])
  control_before <- nrow(profiles_list[[c4_key]])

  # Intersection of (gene_id, reference_isoform_id)
  c2_gene_refs <- unique(profiles_list[[c2_key]][, c("gene_id", "reference_isoform_id")])
  c4_gene_refs <- unique(profiles_list[[c4_key]][, c("gene_id", "reference_isoform_id")])
  shared_refs <- inner_join(c2_gene_refs, c4_gene_refs, by = c("gene_id", "reference_isoform_id"))

  profiles_list[[c2_key]] <- semi_join(profiles_list[[c2_key]], shared_refs,
                                        by = c("gene_id", "reference_isoform_id"))
  profiles_list[[c4_key]] <- semi_join(profiles_list[[c4_key]], shared_refs,
                                        by = c("gene_id", "reference_isoform_id"))

  gene_match_summary[[ct]] <- data.frame(
    cell_type = ct,
    nmd_before = nmd_before,
    control_before = control_before,
    matched_genes = nrow(shared_refs),
    nmd_after = nrow(profiles_list[[c2_key]]),
    control_after = nrow(profiles_list[[c4_key]])
  )
}
gene_match_df <- do.call(rbind, gene_match_summary)
```

# 2. Data Flow

```{r data-flow-numbers}
# --- Gather numbers ---
n_isoforms <- nrow(expr_mat)
n_samples <- ncol(expr_mat)
n_dmso <- sum(sample_meta$treatment == "DMSO")
n_smg1i <- sum(sample_meta$treatment == "Smg1i")
n_genes_struct <- n_distinct(structures$gene_id)
coding_n <- sum(cds$coding_status == "coding")
unknown_n <- sum(cds$coding_status == "unknown")

nmd_summary <- do.call(rbind, lapply(names(nmd_class), function(ct) {
  data.frame(cell_type = ct,
             n_nmd = length(nmd_class[[ct]]$nmd),
             n_non_nmd = length(nmd_class[[ct]]$non_nmd))
}))
total_nmd <- length(unique(unlist(lapply(nmd_class, `[[`, "nmd"))))
total_non_nmd <- length(unique(unlist(lapply(nmd_class, `[[`, "non_nmd"))))

total_nmd_before <- sum(gene_match_df$nmd_before)
total_ctrl_before <- sum(gene_match_df$control_before)
total_nmd_after <- sum(gene_match_df$nmd_after)
total_ctrl_after <- sum(gene_match_df$control_after)
total_matched_genes <- sum(gene_match_df$matched_genes)
```

```{r data-flow-chart}
# --- Flowchart (Graphviz via DiagrammeR) ---
fmt <- function(x) format(x, big.mark = ",")

dot_spec <- sprintf('
digraph flowchart {
  graph [rankdir=TB, fontname="Helvetica", nodesep=1.6, ranksep=1.4, size="13,14"]
  node  [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=23, margin="0.65,0.5", width=6.5]
  edge  [color="#666666", penwidth=2.5]

  A [label=<<b>Isoform Expression</b><br/>%s isoforms × %d samples<br/>%d DMSO + %d Smg1i | 6 cell types>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  B [label=<<b>NMD Classification (mashr)</b><br/>NMD-sensitive: %s isoforms<br/>Non-NMD: %s isoforms>,
     fillcolor="#fff7bc", color="#d95f0e"]

  C [label=<<b>Pair Construction</b><br/>%s NMD pairs | %s Control pairs>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  D [label=<<b>Gene-Matching</b><br/>%s shared genes<br/>%s NMD | %s Control pairs>,
     fillcolor="#c7e9c0", color="#31a354"]

  A -> B -> C -> D
}',
  fmt(n_isoforms), n_samples, n_dmso, n_smg1i,
  fmt(total_nmd), fmt(total_non_nmd),
  fmt(total_nmd_before), fmt(total_ctrl_before),
  fmt(total_matched_genes), fmt(total_nmd_after), fmt(total_ctrl_after)
)

if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(dot_spec, width = 1000, height = 700)
} else {
  message("DiagrammeR not installed — flowchart not rendered")
}
```

## NMD Classification

Isoforms are classified per cell type using mashr (multivariate adaptive shrinkage)
differential isoform expression results. NMD-sensitive isoforms are identified by the
pre-computed `nmd_responsive` flag (adj.P.Val < 0.05 & logFC > 0, with mashr-shrunken
effect sizes). Non-NMD: adj.P.Val > 0.50 (no evidence of NMD regulation). Note: mashr
logFC values are shrunken relative to limma, but the NMD/non-NMD classification thresholds
are identical.

**Columns:** **cell_type** = cell type used for DE testing; **n_nmd** = number of isoforms classified as NMD-sensitive in that cell type; **n_non_nmd** = number of isoforms classified as non-NMD.

```{r nmd-classification-table, results='asis'}
render_dt(nmd_summary, caption = "NMD Classification Summary")
```

## Gene-Matching Detail

To enable paired NMD vs Control comparisons, we restrict both sets to genes where the same
reference isoform appears in both NMD and Control pairs. This ensures that any differences
between NMD and Control comparators are not confounded by different reference isoforms or gene sets.

**Columns:** **nmd_before/control_before** = pair counts before matching; **matched_genes** = genes with shared (gene_id, reference_isoform_id) in both sets; **nmd_after/control_after** = pair counts after restricting to matched genes.

```{r gene-matching-table, results='asis'}
render_dt(gene_match_df, caption = "Gene-Matching Summary (bidirectional, per cell type)")
```

## Profile Sets (after gene-matching)

```{r profile-sets-compute}
# Count splice events per pair: independent splicing decisions, computed by
# grouping non-terminal events that share a splice junction (union-find).
# A single splice event (one donor/acceptor choice) can produce multiple
# sequence events (e.g., skipping 3 exons = 1 splice event, 3 sequence events).
count_splice_events <- function(events) {
  if (is.null(events) || nrow(events) == 0) return(0L)
  # Exclude terminal events — these are transcription boundary differences, not splicing
  splice_types <- c("A5SS", "A3SS", "SE", "Missing_Internal", "IR",
                     "IR_diff", "IR_diff_5", "IR_diff_3", "IR_diff_5_3",
                     "Partial_IR_5", "Partial_IR_3")
  se <- events[events$event_type %in% splice_types, ]
  n <- nrow(se)
  if (n == 0L) return(0L)
  if (n == 1L) return(1L)

  # Union-find: group events sharing a splice junction
  parent <- seq_len(n)
  find <- function(x) {
    while (parent[x] != x) { parent[x] <<- parent[parent[x]]; x <- parent[x] }
    x
  }
  union_nodes <- function(a, b) {
    ra <- find(a); rb <- find(b)
    if (ra != rb) parent[ra] <<- rb
  }

  parse_jxns <- function(jxn_str) {
    if (is.na(jxn_str) || jxn_str == "") return(character(0))
    unlist(strsplit(jxn_str, ","))
  }

  jxn_map <- list()
  for (i in seq_len(n)) {
    all_jxns <- unique(c(
      parse_jxns(se$ref_junctions[i]),
      parse_jxns(se$comp_junctions[i])
    ))
    for (jxn in all_jxns) {
      if (jxn %in% names(jxn_map)) union_nodes(i, jxn_map[[jxn]])
      jxn_map[[jxn]] <- i
    }
  }
  # Events with no junctions (empty ref_junctions AND comp_junctions) each form their own group
  length(unique(vapply(seq_len(n), find, integer(1))))
}

prof_overview <- do.call(rbind, lapply(names(profiles_list), function(key) {
  p <- profiles_list[[key]]
  parts <- strsplit(key, "_")[[1]]

  # Per-pair splice event counts
  splice_events_per_pair <- vapply(p$detailed_events, count_splice_events, integer(1))

  data.frame(comparison = comp_label(parts[1]),
             cell_type = paste(parts[-1], collapse = "_"),
             n_profiles = nrow(p),
             n_genes = n_distinct(p$gene_id),
             total_sequence_events = sum(p$n_events),
             mean_sequence_events = round(mean(p$n_events), 2),
             total_splice_events = sum(splice_events_per_pair),
             mean_splice_events = round(mean(splice_events_per_pair), 2),
             pct_recon_pass = round(100 * mean(p$reconstruction_status == "PASS",
                                                na.rm = TRUE), 1))
}))
```

Each profile set represents isoform pairs analyzed for structural differences using
Isopair (`buildProfiles()` in `Isopair/R/profiles.R`; see METHODS.md §Event Detection
Algorithms). NMD pairs compare an NMD-sensitive comparator to a non-NMD reference;
Control pairs compare the next-best non-NMD isoform to the same reference.

The table reports two levels of event counting:

- A **sequence event** is an individual structural difference detected between two
  isoforms — e.g., a single skipped exon (SE), an alternative 5' splice site (A5SS),
  or a retained intron (IR). Each row in the `detailed_events` table is one sequence
  event. This is the finest-grained unit of structural difference.
- A **splice event** is an independent splicing decision — a single, different choice of
  donor and/or acceptor between the two isoforms. One splice event can produce multiple
  sequence events. For example, skipping three consecutive exons involves one splice
  decision (the donor of the upstream exon joins the acceptor of the downstream exon)
  but creates three Missing_Internal sequence events. Splice events are computed by
  grouping non-terminal sequence events that share a splice junction (union-find
  algorithm; see METHODS.md §Splice Group Assignment). Terminal events (Alt_TSS,
  Alt_TES) are excluded because they reflect transcription boundary differences, not
  splicing decisions.

**Columns:** **comparison** = NMD or Control; **cell_type** = cell type (or all_samples
for union/intersection); **n_profiles** = number of isoform pairs; **n_genes** = unique
genes represented; **total_sequence_events / mean_sequence_events** = sum and mean of
individual structural differences per pair; **total_splice_events / mean_splice_events** =
sum and mean of independent splicing decisions per pair; **pct_recon_pass** = % of pairs
where the comparator isoform can be perfectly reconstructed from the reference plus
detected events (quality control metric; 100% expected).

```{r profile-sets-table, results='asis'}
render_dt(prof_overview, caption = "Profile Set Summary (gene-matched)")
```

## Coding Status

NMD is a translation-dependent pathway: only protein-coding transcripts with an open reading
frame can harbour a premature termination codon. We therefore expect NMD-sensitive isoform pairs
to be overwhelmingly coding. This table characterizes the coding status of reference and
comparator isoforms in each profile set, based on SQANTI CDS annotations. A small fraction of
isoforms have unknown coding status (no CDS annotation in SQANTI); these are excluded from
downstream analyses that require reading-frame information (e.g., frameshift analysis).

```{r coding-status}
coding_tbl <- do.call(rbind, lapply(names(profiles_list), function(key) {
  p <- profiles_list[[key]]
  parts <- strsplit(key, "_")[[1]]
  comp <- comp_label(parts[1])
  ct <- paste(parts[-1], collapse = "_")

  ref_ids <- unique(p$reference_isoform_id)
  cmp_ids <- unique(p$comparator_isoform_id)

  ref_coding <- sum(ref_ids %in% cds$isoform_id[cds$coding_status == "coding"])
  cmp_coding <- sum(cmp_ids %in% cds$isoform_id[cds$coding_status == "coding"])

  # Pairs where both are coding
  ref_status <- cds$coding_status[match(p$reference_isoform_id, cds$isoform_id)]
  cmp_status <- cds$coding_status[match(p$comparator_isoform_id, cds$isoform_id)]
  both_coding <- sum(ref_status == "coding" & cmp_status == "coding", na.rm = TRUE)

  data.frame(
    Comparison = comp,
    Cell_Type = ct,
    N_Pairs = nrow(p),
    Ref_Coding = ref_coding,
    Ref_Pct = round(100 * ref_coding / length(ref_ids), 1),
    Comp_Coding = cmp_coding,
    Comp_Pct = round(100 * cmp_coding / length(cmp_ids), 1),
    Both_Coding = both_coding,
    Both_Pct = round(100 * both_coding / nrow(p), 1)
  )
}))
```

**Columns:** **Comparison** = NMD or Control; **Cell_Type** = cell type; **N_Pairs** = total
isoform pairs; **Ref_Coding / Ref_Pct** = number and percentage of unique reference isoforms
with a CDS annotation; **Comp_Coding / Comp_Pct** = same for comparator isoforms;
**Both_Coding / Both_Pct** = pairs where both reference and comparator are coding.

```{r coding-status-table, results='asis'}
render_dt(coding_tbl, caption = "Coding Status of Reference and Comparator Isoforms")
```

## Pair Subset Overview

The analyses in this report operate on different subsets of the isoform pair universe.
The flowcharts below show how NMD and Control pairs are progressively subdivided and
which report sections draw on each subset. Counts are for the all_samples set.

```{r pair-subset-flowchart, fig.width=16, fig.height=16}
# Combined NMD + Control pair subset flowchart
fmt <- function(x) format(x, big.mark = ",")

# Helper: compute subset counts for one profile set
compute_subset_counts <- function(profiles) {
  n_all <- nrow(profiles)

  ref_st <- cds$coding_status[match(profiles$reference_isoform_id, cds$isoform_id)]
  cmp_st <- cds$coding_status[match(profiles$comparator_isoform_id, cds$isoform_id)]
  n_both_coding <- sum(ref_st == "coding" & cmp_st == "coding", na.rm = TRUE)
  n_non_coding  <- n_all - n_both_coding

  fc <- tryCatch(
    compareIsoformFrames(profiles, cds_exons, cds),
    error = function(e) NULL
  )
  if (is.null(fc)) return(NULL)
  ps_ov <- fc$pair_summary

  ptc_pos_ids <- ptc$isoform_id[ptc$has_ptc == TRUE]
  count_ptc <- function(cat_name) {
    sum(ps_ov$comparator_isoform_id[ps_ov$frame_category == cat_name] %in% ptc_pos_ids)
  }

  n_ss_if  <- sum(ps_ov$frame_category == "same_start_in_frame")
  n_ss_fs  <- sum(ps_ov$frame_category == "same_start_frameshift")
  n_ds_sf  <- sum(ps_ov$frame_category == "diff_start_same_frame")
  n_ds_df  <- sum(ps_ov$frame_category == "diff_start_diff_frame")
  n_ds_dfs <- sum(ps_ov$frame_category == "diff_start_diff_frame_with_frameshift")
  n_nc     <- sum(ps_ov$frame_category == "non_coding")
  n_nsc    <- sum(ps_ov$frame_category == "no_shared_cds")

  list(
    all = n_all, coding = n_both_coding, non_coding = n_non_coding,
    same_start = n_ss_if + n_ss_fs, diff_start = n_ds_sf + n_ds_df + n_ds_dfs,
    no_shared_cds = n_nsc,
    ss_if = n_ss_if, ss_fs = n_ss_fs,
    ds_sf = n_ds_sf, ds_df = n_ds_df, ds_dfs = n_ds_dfs,
    fs_ptc = count_ptc("same_start_frameshift"),
    fs_no_ptc = n_ss_fs - count_ptc("same_start_frameshift"),
    if_ptc = count_ptc("same_start_in_frame"),
    if_no_ptc = n_ss_if - count_ptc("same_start_in_frame"),
    sf_ptc = count_ptc("diff_start_same_frame"),
    sf_no_ptc = n_ds_sf - count_ptc("diff_start_same_frame"),
    df_ptc = count_ptc("diff_start_diff_frame"),
    df_no_ptc = n_ds_df - count_ptc("diff_start_diff_frame"),
    dfs_ptc = count_ptc("diff_start_diff_frame_with_frameshift"),
    dfs_no_ptc = n_ds_dfs - count_ptc("diff_start_diff_frame_with_frameshift")
  )
}

# Helper: format a dual-count label line
dual <- function(n_nmd, n_ctrl) {
  sprintf('<font color="#d95f0e">NMD: %s</font>  |  <font color="#3182bd">Ctrl: %s</font>',
          fmt(n_nmd), fmt(n_ctrl))
}

as_c2_key <- "c2_allsamples"
as_c4_key <- "c4_allsamples"
dot_combined <- NULL

if (requireNamespace("DiagrammeR", quietly = TRUE) &&
    as_c2_key %in% names(profiles_list) &&
    as_c4_key %in% names(profiles_list)) {

  nmd  <- compute_subset_counts(profiles_list[[as_c2_key]])
  ctrl <- compute_subset_counts(profiles_list[[as_c4_key]])

  # After gene-matching, NMD and Control have the same number of pairs
  n_matched <- nmd$all
  stopifnot(nmd$all == ctrl$all)

  if (!is.null(nmd) && !is.null(ctrl)) {

    # Build PTC subdivision nodes and edges for terminal categories
    ptc_nodes <- ""
    ptc_edges <- ""
    ptc_rank_items <- c("FPTC", "FNPTC")

    # Same-start in-frame → PTC subdivision (if either has PTC > 0)
    if (nmd$if_ptc > 0 || ctrl$if_ptc > 0) {
      ptc_nodes <- paste0(ptc_nodes, sprintf('
  IFPTC  [label=<<b>Same Frame + PTC</b><br/>%s>, fillcolor="#fcbba1", color="#cb181d"]
  IFNPTC [label=<<b>Same Frame, No PTC</b><br/>%s>, fillcolor="#c6dbef", color="#3182bd"]
', dual(nmd$if_ptc, ctrl$if_ptc), dual(nmd$if_no_ptc, ctrl$if_no_ptc)))
      ptc_edges <- paste0(ptc_edges, "  IF -> IFPTC\n  IF -> IFNPTC\n")
      ptc_rank_items <- c(ptc_rank_items, "IFPTC", "IFNPTC")
    }
    # Diff-start same-frame → PTC subdivision
    if (nmd$sf_ptc > 0 || ctrl$sf_ptc > 0) {
      ptc_nodes <- paste0(ptc_nodes, sprintf('
  SFPTC  [label=<<b>Same Frame + PTC</b><br/>%s>, fillcolor="#fcbba1", color="#cb181d"]
  SFNPTC [label=<<b>Same Frame, No PTC</b><br/>%s>, fillcolor="#c6dbef", color="#3182bd"]
', dual(nmd$sf_ptc, ctrl$sf_ptc), dual(nmd$sf_no_ptc, ctrl$sf_no_ptc)))
      ptc_edges <- paste0(ptc_edges, "  FP -> SFPTC\n  FP -> SFNPTC\n")
      ptc_rank_items <- c(ptc_rank_items, "SFPTC", "SFNPTC")
    }
    # Diff-start diff-frame → PTC subdivision
    if (nmd$df_ptc > 0 || ctrl$df_ptc > 0) {
      ptc_nodes <- paste0(ptc_nodes, sprintf('
  DFPTC  [label=<<b>Diff Frame + PTC</b><br/>%s>, fillcolor="#fcbba1", color="#cb181d"]
  DFNPTC [label=<<b>Diff Frame, No PTC</b><br/>%s>, fillcolor="#c6dbef", color="#3182bd"]
', dual(nmd$df_ptc, ctrl$df_ptc), dual(nmd$df_no_ptc, ctrl$df_no_ptc)))
      ptc_edges <- paste0(ptc_edges, "  FD -> DFPTC\n  FD -> DFNPTC\n")
      ptc_rank_items <- c(ptc_rank_items, "DFPTC", "DFNPTC")
    }
    # Diff-start diff-frame + FS → PTC subdivision
    if (nmd$dfs_ptc > 0 || ctrl$dfs_ptc > 0) {
      ptc_nodes <- paste0(ptc_nodes, sprintf('
  DFSPTC  [label=<<b>Diff Frame + FS + PTC</b><br/>%s>, fillcolor="#fcbba1", color="#cb181d"]
  DFSNPTC [label=<<b>Diff Frame + FS, No PTC</b><br/>%s>, fillcolor="#c6dbef", color="#3182bd"]
', dual(nmd$dfs_ptc, ctrl$dfs_ptc), dual(nmd$dfs_no_ptc, ctrl$dfs_no_ptc)))
      ptc_edges <- paste0(ptc_edges, "  FDFS -> DFSPTC\n  FDFS -> DFSNPTC\n")
      ptc_rank_items <- c(ptc_rank_items, "DFSPTC", "DFSNPTC")
    }

    ptc_rank <- sprintf("  {rank=same; %s}", paste(ptc_rank_items, collapse = "; "))

    dot_combined <- sprintf('
digraph pair_subsets {
  graph [rankdir=TB, fontname="Helvetica", nodesep=0.6, ranksep=0.9,
         size="16,20", label=<<b>Pair Subset Overview (all_samples)</b><br/>
         <font color="#d95f0e">NMD (C2)</font> | <font color="#3182bd">Control (C4)</font>>,
         labelloc=t, labeljust=c, fontsize=24]
  node  [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=20,
         margin="0.35,0.25", width=3.5]
  edge  [color="#666666", penwidth=2.0, fontname="Helvetica", fontsize=14]

  ALL [label=<<b>Gene-Matched Pairs</b><br/>%s pairs<br/><font point-size="16">(same genes, same references)</font>>,
       fillcolor="#e8f4f8", color="#2c7fb8"]

  COD [label=<<b>Both-Coding Pairs</b><br/>%s>,
       fillcolor="#c7e9c0", color="#31a354"]
  NC  [label=<<b>Non-Coding</b><br/>%s<br/><font point-size="15">(one/both lack CDS)</font>>,
       fillcolor="#f0f0f0", color="#999999"]

  SS  [label=<<b>Same Start Site</b><br/>%s>,
       fillcolor="#fff7bc", color="#d95f0e"]
  DS  [label=<<b>Different Start Site</b><br/>%s>,
       fillcolor="#fff7bc", color="#d95f0e"]
  NSC [label=<<b>No Shared CDS</b><br/>%s>,
       fillcolor="#f0f0f0", color="#999999"]

  IF  [label=<<b>Same Frame</b><br/>%s>,
       fillcolor="#c6dbef", color="#3182bd"]
  FS  [label=<<b>Frameshift</b><br/>%s>,
       fillcolor="#fcbba1", color="#cb181d"]
  FP  [label=<<b>Same Frame</b><br/>%s>,
       fillcolor="#c6dbef", color="#3182bd"]
  FD  [label=<<b>Diff Frame</b><br/>%s>,
       fillcolor="#fff7bc", color="#d95f0e"]
  FDFS [label=<<b>Diff Frame + FS</b><br/>%s>,
       fillcolor="#fcbba1", color="#cb181d"]

  FPTC  [label=<<b>Frameshift + PTC</b><br/>%s>,
         fillcolor="#fcbba1", color="#cb181d"]
  FNPTC [label=<<b>Frameshift, No PTC</b><br/>%s<br/><font point-size="15">(novel proteins)</font>>,
         fillcolor="#dadaeb", color="#6a51a3"]

  %s

  node [shape=plaintext, style="", fontsize=20, fontcolor="#555555"]

  A_ALL [label=<<b>Used by:</b> Event Frequency, Gain/Loss,<br/>IR, Divergence, Co-occurrence,<br/>Heterogeneity, Meta-analysis<br/><font point-size="17">(Sections 4-6)</font>>]
  A_COD [label=<<b>Used by:</b> PTC Rates,<br/>Frame Comparison<br/><font point-size="17">(Section 3)</font>>]
  A_SS  [label=<<b>Used by:</b> Splicing-Induced<br/>Frameshift Enrichment test<br/><font point-size="17">(Section 3)</font>>]
  A_FS  [label=<<b>Used by:</b> Frameshift Outcome,<br/>Frame-PTC Cross-Reference<br/><font point-size="17">(Section 3)</font>>]
  A_PTC [label=<<b>Used by:</b> PTC Distance vs logFC,<br/>GAM models<br/><font point-size="17">(Section 3)</font>>]
  A_NP  [label=<<b>Used by:</b> Novel Protein<br/>Analysis<br/><font point-size="17">(Section 3)</font>>]

  ALL -> COD
  ALL -> NC

  COD -> SS
  COD -> DS
  COD -> NSC

  SS -> IF
  SS -> FS

  DS -> FP
  DS -> FD
  DS -> FDFS

  FS -> FPTC
  FS -> FNPTC

  %s

  edge [style=dashed, color="#aaaaaa", penwidth=1.0, arrowhead=none]
  ALL -> A_ALL
  COD -> A_COD
  SS  -> A_SS
  FS  -> A_FS
  FPTC -> A_PTC
  FNPTC -> A_NP

  {rank=same; ALL; A_ALL}
  {rank=same; COD; NC; A_COD}
  {rank=same; SS; DS; NSC; A_SS}
  {rank=same; IF; FS; FP; FD; FDFS; A_FS}
  %s
}',
    fmt(n_matched),
    dual(nmd$coding, ctrl$coding),
    dual(nmd$non_coding, ctrl$non_coding),
    dual(nmd$same_start, ctrl$same_start),
    dual(nmd$diff_start, ctrl$diff_start),
    dual(nmd$no_shared_cds, ctrl$no_shared_cds),
    dual(nmd$ss_if, ctrl$ss_if),
    dual(nmd$ss_fs, ctrl$ss_fs),
    dual(nmd$ds_sf, ctrl$ds_sf),
    dual(nmd$ds_df, ctrl$ds_df),
    dual(nmd$ds_dfs, ctrl$ds_dfs),
    dual(nmd$fs_ptc, ctrl$fs_ptc),
    dual(nmd$fs_no_ptc, ctrl$fs_no_ptc),
    ptc_nodes,
    ptc_edges,
    ptc_rank
    )
  }
}
```

```{r pair-subset-combined, fig.width=16, fig.height=16}
if (!is.null(dot_combined)) DiagrammeR::grViz(dot_combined, width = 1400, height = 1200)
```

```{r compute-all, include=FALSE}
# =============================================================================
# Silent compute chunk: all cached_compute() calls + comparePairSets
# No output — results stored in set_results / cross_results for display sections
# =============================================================================
set_results <- list()
cross_results <- list()

for (key in names(profiles_list)) {
  profiles <- profiles_list[[key]]
  message(sprintf("Computing analyses for %s...", key))

  prof_summary <- summarizeProfiles(profiles)
  divergence <- cached_compute(paste0("div_", key),
    tryCatch(quantifyPairDivergence(profiles, structures, cds), error = function(e) NULL))
  regional_impact <- cached_compute(paste0("ri_", key),
    tryCatch(quantifyRegionalImpact(profiles, annotated_ue, cds), error = function(e) NULL))
  cooc <- cached_compute(paste0("cooc_", key),
    tryCatch(testCooccurrence(profiles), error = function(e) NULL))
  event_regions <- cached_compute(paste0("er_", key),
    tryCatch(mapEventsToRegions(profiles, annotated_ue, cds), error = function(e) NULL))
  ptc_assoc <- cached_compute(paste0("ptc_", key),
    tryCatch(testPtcAssociation(ptc, profiles), error = function(e) NULL))
  frame_walk <- cached_compute(paste0("fw_", key),
    tryCatch(analyzeFrameWalk(profiles, cds, ptc), error = function(e) NULL))
  frame_comp <- cached_compute(paste0("fc_", key),
    tryCatch(compareIsoformFrames(profiles, cds_exons, cds), error = function(e) NULL))

  # Dependent computations (not cached separately — fast from cached event_regions)
  orf_bound <- NULL; orf_impact <- NULL
  if (!is.null(event_regions) && nrow(event_regions) > 0) {
    orf_bound <- tryCatch(testOrfBoundarySusceptibility(event_regions, annotated_ue),
                          error = function(e) NULL)
    orf_impact <- tryCatch(summarizeOrfImpact(event_regions, cds),
                           error = function(e) NULL)
  }

  set_results[[key]] <- list(
    profiles = profiles, prof_summary = prof_summary, divergence = divergence,
    regional_impact = regional_impact,
    cooc = cooc, event_regions = event_regions, orf_bound = orf_bound,
    orf_impact = orf_impact, ptc_assoc = ptc_assoc, frame_walk = frame_walk,
    frame_comp = frame_comp, evt_freq = compute_event_freq(profiles)
  )
}

# comparePairSets per cell type (cached, reused by display sections + meta-analysis)
# Include DD_ALI for the dedicated section (profiles loaded from limma data)
all_cts_for_cps <- unique(c(cell_types, ali_ct))
for (ct in all_cts_for_cps) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(profiles_list) || !c4_key %in% names(profiles_list)) next

  message(sprintf("Computing NMD vs Control comparison for %s...", ct))
  ptc_c2 <- ptc[ptc$isoform_id %in% profiles_list[[c2_key]]$comparator_isoform_id, ]
  ptc_c4 <- ptc[ptc$isoform_id %in% profiles_list[[c4_key]]$comparator_isoform_id, ]

  cross_results[[ct]] <- cached_compute(
    paste0("cps_", ct_label),
    tryCatch(
      comparePairSets(profiles_list[[c2_key]], profiles_list[[c4_key]],
                       ptc_a = ptc_c2, ptc_b = ptc_c4,
                       paired = TRUE, verbose = FALSE),
      error = function(e) {
        message(sprintf("comparePairSets failed for %s: %s", ct, e$message))
        NULL
      }
    )
  )
}

# Cell types available for display (DD_ALI excluded — dedicated section below)
all_cts_with_profiles <- unique(c(cell_types, ali_ct))
display_cts <- setdiff(
  all_cts_with_profiles[sapply(all_cts_with_profiles, function(ct) {
    ct_label <- ct_to_label(ct)
    any(paste0(c("c2_", "c4_"), ct_label) %in% names(set_results))
  })],
  ali_ct
)
per_ct <- setdiff(display_cts, "all_samples")
c2_per_ct <- per_ct[sapply(per_ct, function(ct)
  paste0("c2_", ct_to_label(ct)) %in% names(set_results))]

# Main cell types for cross-CT analyses (excluding DD_ALI outlier)
main_per_ct <- intersect(c2_per_ct, main_cts)
ali_in_c2 <- ali_ct %in% c2_per_ct

message("All computations complete.")
```


# 3. PTC & Reading Frame

```{r sec3-ptc-compute}
# Compute PTC rates table (used by inline references and display chunk below)
ptc_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)

  # Restrict to coding pairs (both ref and comp have coding_status == "coding")
  coding_ids <- cds$isoform_id[cds$coding_status == "coding"]

  p_c2 <- set_results[[c2_key]]$profiles
  cod_c2 <- p_c2[p_c2$reference_isoform_id %in% coding_ids &
                   p_c2$comparator_isoform_id %in% coding_ids, ]

  p_c4 <- set_results[[c4_key]]$profiles
  cod_c4 <- p_c4[p_c4$reference_isoform_id %in% coding_ids &
                   p_c4$comparator_isoform_id %in% coding_ids, ]

  if (nrow(cod_c2) == 0 || nrow(cod_c4) == 0) return(NULL)

  # PTC rates among coding comparators
  ptc_c2 <- sum(cod_c2$comparator_isoform_id %in% ptc$isoform_id[ptc$has_ptc == TRUE])
  ptc_c4 <- sum(cod_c4$comparator_isoform_id %in% ptc$isoform_id[ptc$has_ptc == TRUE])
  n_c2 <- nrow(cod_c2)
  n_c4 <- nrow(cod_c4)

  # Fisher's exact test: PTC vs no-PTC × NMD vs Control
  ft <- tryCatch({
    mat <- matrix(c(ptc_c2, n_c2 - ptc_c2, ptc_c4, n_c4 - ptc_c4), nrow = 2)
    fisher.test(mat)
  }, error = function(e) NULL)

  data.frame(
    cell_type = ct,
    n_coding_pairs = n_c2,
    NMD_comp_ptc_n = ptc_c2,
    NMD_comp_ptc_pct = round(100 * ptc_c2 / n_c2, 1),
    Ctrl_comp_ptc_n = ptc_c4,
    Ctrl_comp_ptc_pct = round(100 * ptc_c4 / n_c4, 1),
    fisher_OR = if (!is.null(ft)) round(ft$estimate, 2) else NA_real_,
    p_value = if (!is.null(ft)) ft$p.value else NA_real_
  )
}))
if (!is.null(ptc_rows)) rownames(ptc_rows) <- NULL
ptc_or_range <- if (!is.null(ptc_rows)) {
  or_vals <- ptc_rows$fisher_OR[!is.na(ptc_rows$fisher_OR)]
  sprintf("OR %s--%s", round(min(or_vals)), round(max(or_vals)))
} else {
  "OR N/A"
}
ctrl_ptc_pcts <- if (!is.null(ptc_rows)) {
  round(median(ptc_rows$Ctrl_comp_ptc_pct, na.rm = TRUE))
} else {
  5
}
```

PTC enrichment (`r ptc_or_range`) is the single strongest signal distinguishing NMD-sensitive isoform
pairs from baseline structural variation. At baseline, ~`r ctrl_ptc_pcts`% of non-NMD comparator isoforms
contain a PTC. Reading frame is compared independently for each isoform at shared CDS regions,
distinguishing splicing-induced frameshifts (same start codon) from alternative initiation
(different start codons). NMD departs dramatically from Control in both PTC prevalence and
the consequences of frame disruption.

## PTC Rates

Premature termination codons (PTCs) are stop codons located >50 nucleotides upstream of the
last exon-exon junction, a position that triggers nonsense-mediated decay. This table compares
PTC prevalence in NMD comparator isoforms vs Control comparator isoforms, restricted to
**coding pairs** (both reference and comparator have a SQANTI CDS annotation). Non-coding
isoforms cannot harbour a PTC and are excluded from both the numerator and denominator.

PTC status is determined per-isoform using each isoform's own CDS (`computePtcStatus()` in
`Isopair/R/ptc.R`; see METHODS.md §Reading Frame and PTC Analysis). The 50-nt rule is
applied to each isoform's stop codon relative to its own last exon-exon junction.

**Columns:** **n_coding_pairs** = pairs where both reference and comparator are coding;
**NMD_comp_ptc_n / NMD_comp_ptc_pct** = number and % of NMD comparators with a PTC
(among coding pairs); **Ctrl_comp_ptc_n / Ctrl_comp_ptc_pct** = same for Control;
**fisher_OR** = odds ratio from Fisher's exact test (>1 = enriched in NMD);
**p_value** = Fisher's exact test p-value.

```{r sec3-ptc-rates, results='asis'}
if (!is.null(ptc_rows)) {
  knitr::kable(ptc_rows, row.names = FALSE,
               caption = "PTC Rates: NMD vs Control Comparators (coding pairs only)") |> print()
  cat("\n\n")
}
```

## Reading Frame Comparison

For each isoform pair, we compare the reading frame at shared CDS regions using each
isoform's own CDS annotation independently (`compareIsoformFrames()` in
`Isopair/R/ptc.R`; see METHODS.md §Reading Frame and PTC Analysis). CDS exon intervals
are extracted per isoform from the SQANTI GFF3 (`extractCdsExons()` in
`Isopair/R/annotate-regions.R`). For each pair, the function identifies genomic intervals
where both isoforms have CDS coverage, then computes the CDS phase (cumulative coding
bp mod 3) independently for each isoform at the 5' end of each shared interval.

The analysis first determines whether the two isoforms share the same start codon
(translation initiation site, with tolerance = 0), then evaluates whether shared CDS
regions are translated in the same reading frame. Every pair is assigned to exactly one
of six categories:

### Frame Category Counts

Each pair is classified into one of the following categories. The first four apply to
coding pairs with overlapping CDS; the last two are excluded from frame-specific analyses.

- **same_start_in_frame**: Both isoforms begin translation at the same position. All
  shared CDS regions are read in the same phase. Splicing differences between these
  isoforms affect sequence content but preserve the reading frame.
- **same_start_frameshift**: Both isoforms begin translation at the same position, but
  a splicing difference (e.g., a non-divisible-by-3 exon skip or boundary shift) causes
  downstream shared CDS to be read in a different phase. This is a splicing-induced
  frameshift.
- **diff_start_same_frame**: The two isoforms use different translation initiation
  sites, but shared CDS regions are read in the same phase (the start codon offset is a
  multiple of 3, or compensating splicing restores phase). No frameshift is implied.
- **diff_start_diff_frame**: The two isoforms use different translation initiation
  sites, and shared CDS regions are read in different phases with a constant phase offset.
  The frame difference is entirely attributable to the initiation site difference — all
  internal splicing differences are multiples of 3 (no frameshift).
- **diff_start_diff_frame_with_frameshift**: The two isoforms use different translation
  initiation sites, and the phase offset between them changes across shared CDS regions.
  This indicates a splicing-induced frameshift on top of the start codon offset.
- **non_coding**: One or both isoforms lack a CDS annotation (SQANTI coding_status ≠
  "coding"). These pairs cannot be evaluated for reading frame.
- **no_shared_cds**: Both isoforms are coding, but their CDS regions do not overlap
  genomically (e.g., entirely different exons are coding). No shared interval exists for
  phase comparison.

**Columns:** **cell_type** = cell type; **frame_category** = classification (see above);
**NMD_n / NMD_pct** = count and percentage of NMD (C2) pairs in this category;
**Ctrl_n / Ctrl_pct** = count and percentage of Control (C4) pairs. Percentages are
computed over all pairs in the respective set (including non_coding and no_shared_cds).

```{r sec3-frame-comp-detail, results='asis'}
# Detailed breakdown of frame categories
fc_detail <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  fc_c2 <- set_results[[c2_key]]$frame_comp
  fc_c4 <- set_results[[c4_key]]$frame_comp
  if (is.null(fc_c2) || is.null(fc_c4)) return(NULL)
  ps_c2 <- fc_c2$pair_summary
  ps_c4 <- fc_c4$pair_summary

  cats <- c("same_start_in_frame", "same_start_frameshift",
            "diff_start_same_frame", "diff_start_diff_frame",
            "diff_start_diff_frame_with_frameshift",
            "non_coding", "no_shared_cds")
  do.call(rbind, lapply(cats, function(cat) {
    data.frame(
      cell_type = ct,
      frame_category = cat,
      NMD_n = sum(ps_c2$frame_category == cat),
      NMD_pct = round(100 * sum(ps_c2$frame_category == cat) / nrow(ps_c2), 1),
      Ctrl_n = sum(ps_c4$frame_category == cat),
      Ctrl_pct = round(100 * sum(ps_c4$frame_category == cat) / nrow(ps_c4), 1)
    )
  }))
}))
if (!is.null(fc_detail)) {
  # Show all rows (36 = 6 categories × 6 cell types)
  widget <- datatable(fc_detail, caption = "Frame Category Counts: NMD vs Control",
                      rownames = FALSE, class = "compact stripe",
                      options = list(pageLength = nrow(fc_detail), dom = 'ft'))
  rendered <- htmltools::renderTags(widget)
  knitr::knit_meta_add(rendered$dependencies)
  cat(rendered$html)
  cat("\n")
}
```

```{r sec3-frame-comp-barchart, results='asis', fig.width=14, fig.height=10}
if (!is.null(fc_detail) && has_ggplot2) {
  coding_cats <- c("same_start_in_frame", "same_start_frameshift",
                   "diff_start_same_frame", "diff_start_diff_frame",
                   "diff_start_diff_frame_with_frameshift")
  cat_labels <- c(
    same_start_in_frame = "Same Start\nSame Frame",
    same_start_frameshift = "Same Start\nFrameshift",
    diff_start_same_frame = "Diff Start\nSame Frame",
    diff_start_diff_frame = "Diff Start\nDiff Frame",
    diff_start_diff_frame_with_frameshift = "Diff Start\nDiff Frame + FS"
  )

  ptc_pos_ids <- ptc$isoform_id[ptc$has_ptc == TRUE]

  # Build stacked data: PTC+ and No PTC per cell_type × category × comparison
  stacked_df <- do.call(rbind, lapply(display_cts, function(ct) {
    ct_label <- ct_to_label(ct)
    c2_key <- paste0("c2_", ct_label)
    c4_key <- paste0("c4_", ct_label)
    if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)

    do.call(rbind, lapply(c("c2", "c4"), function(comp) {
      key <- paste0(comp, "_", ct_label)
      fc <- set_results[[key]]$frame_comp
      if (is.null(fc)) return(NULL)
      ps <- fc$pair_summary
      total_pairs <- nrow(set_results[[key]]$profiles)
      comp_lbl <- if (comp == "c2") "NMD" else "Control"

      do.call(rbind, lapply(coding_cats, function(cat) {
        cat_pairs <- ps[ps$frame_category == cat, ]
        n_cat <- nrow(cat_pairs)
        n_ptc <- sum(cat_pairs$comparator_isoform_id %in% ptc_pos_ids)
        n_no_ptc <- n_cat - n_ptc
        pct_ptc <- 100 * n_ptc / max(total_pairs, 1)
        pct_no_ptc <- 100 * n_no_ptc / max(total_pairs, 1)

        rbind(
          data.frame(cell_type = ct, frame_category = cat,
                     category_label = cat_labels[cat],
                     comparison = comp_lbl, ptc_status = "PTC+",
                     n = n_ptc, pct = pct_ptc),
          data.frame(cell_type = ct, frame_category = cat,
                     category_label = cat_labels[cat],
                     comparison = comp_lbl, ptc_status = "No PTC",
                     n = n_no_ptc, pct = pct_no_ptc)
        )
      }))
    }))
  }))

  stacked_df$category_label <- factor(stacked_df$category_label,
    levels = cat_labels[coding_cats])
  # Stack order: No PTC on bottom, PTC+ on top
  stacked_df$ptc_status <- factor(stacked_df$ptc_status, levels = c("PTC+", "No PTC"))

  # Create fill variable combining comparison and PTC status
  stacked_df$fill_group <- paste(stacked_df$comparison, stacked_df$ptc_status, sep = " - ")
  stacked_df$fill_group <- factor(stacked_df$fill_group,
    levels = c("NMD - No PTC", "NMD - PTC+", "Control - No PTC", "Control - PTC+"))

  # Total count labels (above each bar)
  bar_totals <- aggregate(cbind(n, pct) ~ cell_type + frame_category + category_label + comparison,
                          data = stacked_df, FUN = sum)

  # Fisher's test per cell_type × category (same as before)
  p_labels <- do.call(rbind, lapply(unique(stacked_df$cell_type), function(ct) {
    do.call(rbind, lapply(coding_cats, function(cat) {
      sub <- fc_detail[fc_detail$cell_type == ct, ]
      n_c2 <- sub$NMD_n[sub$frame_category == cat]
      n_c4 <- sub$Ctrl_n[sub$frame_category == cat]
      total_c2 <- sum(sub$NMD_n)
      total_c4 <- sum(sub$Ctrl_n)
      ft <- tryCatch({
        mat <- matrix(c(n_c2, total_c2 - n_c2, n_c4, total_c4 - n_c4), nrow = 2)
        fisher.test(mat)
      }, error = function(e) list(p.value = NA))
      p <- ft$p.value
      stars <- if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else "ns"
      max_pct <- max(bar_totals$pct[bar_totals$cell_type == ct &
                                     bar_totals$frame_category == cat], na.rm = TRUE)
      data.frame(cell_type = ct, category_label = cat_labels[cat],
                 p_star = stars, max_pct = max_pct)
    }))
  }))
  p_labels$category_label <- factor(p_labels$category_label,
    levels = cat_labels[coding_cats])

  p_bar <- ggplot(stacked_df, aes(x = comparison, y = pct, fill = fill_group)) +
    geom_col(width = 0.7) +
    geom_text(data = bar_totals, aes(x = comparison, y = pct, label = n, fill = NULL),
              vjust = -0.3, size = 3) +
    geom_text(data = p_labels, aes(x = 1.5, y = max_pct + 4, label = p_star),
              inherit.aes = FALSE, size = 4, fontface = "bold") +
    facet_grid(cell_type ~ category_label, scales = "free_y") +
    scale_fill_manual(
      values = c("NMD - No PTC" = "#ef8a62", "NMD - PTC+" = "#8c2d04",
                 "Control - No PTC" = "#67a9cf", "Control - PTC+" = "#2166ac"),
      name = NULL,
      labels = c("NMD - No PTC" = "NMD (No PTC)", "NMD - PTC+" = "NMD (PTC+)",
                 "Control - No PTC" = "Control (No PTC)", "Control - PTC+" = "Control (PTC+)")
    ) +
    theme_minimal(base_size = 12) +
    theme(strip.text = element_text(size = 10, face = "bold"),
          axis.text.x = element_text(angle = 0, hjust = 0.5),
          legend.position = "bottom",
          panel.grid.major.x = element_blank()) +
    labs(x = NULL, y = "% of all pairs",
         title = "Frame Category Distribution: NMD vs Control",
         subtitle = "Stacked by PTC status. Numbers above bars = total pair counts. *** p<0.001, ** p<0.01, * p<0.05, ns (Fisher's exact test)")
  print(p_bar)
  cat("\n\n")
}
```

### PTC Mechanism Breakdown

For PTC-containing comparator isoforms, what is the mechanism by which the PTC arises?
Each PTC+ comparator is classified by its frame category, revealing the route to premature
termination:

- **Same start, frameshift**: Splicing disrupts the reading frame, creating a premature stop
  downstream of the frameshift boundary.
- **Same start, in-frame**: PTC exists within the original reading frame (e.g., a retained
  intron containing a stop codon, or an exon skip that joins in-frame segments with a stop
  between them).
- **Diff start, same frame**: Alternative translation initiation, but shared CDS regions
  are read in the same phase — PTC arises from the shorter or different ORF.
- **Diff start, diff frame**: Alternative initiation reads shared CDS in a different phase.
- **Diff start, diff frame + FS**: Alternative initiation plus an additional splicing-induced
  frameshift.

The table and figure show these mechanism frequencies for both NMD and Control PTC+
comparators. A Fisher's exact test for splicing-induced frameshift enrichment among
same-start pairs is also reported, since this is the primary NMD-generating mechanism.

```{r sec3-ptc-mechanism, results='asis', fig.width=14, fig.height=10}
coding_cats <- c("same_start_in_frame", "same_start_frameshift",
                 "diff_start_same_frame", "diff_start_diff_frame",
                 "diff_start_diff_frame_with_frameshift")
mech_labels <- c(
  same_start_in_frame = "Same Start\nIn Frame",
  same_start_frameshift = "Same Start\nFrameshift",
  diff_start_same_frame = "Diff Start\nSame Frame",
  diff_start_diff_frame = "Diff Start\nDiff Frame",
  diff_start_diff_frame_with_frameshift = "Diff Start\nDiff Frame + FS"
)
ptc_pos_ids <- ptc$isoform_id[ptc$has_ptc == TRUE]

mech_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  do.call(rbind, lapply(c("c2", "c4"), function(comp) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) return(NULL)
    fc <- set_results[[key]]$frame_comp
    if (is.null(fc) || is.null(fc$pair_summary)) return(NULL)
    ps <- fc$pair_summary

    # Restrict to PTC+ comparators in coding categories
    ps_coding <- ps[ps$frame_category %in% coding_cats, ]
    ps_ptc <- ps_coding[ps_coding$comparator_isoform_id %in% ptc_pos_ids, ]
    n_ptc_total <- nrow(ps_ptc)

    do.call(rbind, lapply(coding_cats, function(cat) {
      n_cat <- sum(ps_ptc$frame_category == cat)
      data.frame(
        cell_type = ct, comparison = comp_label(comp),
        frame_category = cat,
        n_ptc = n_cat,
        ptc_total = n_ptc_total,
        pct_of_ptc = round(100 * n_cat / max(n_ptc_total, 1), 1)
      )
    }))
  }))
}))

# Splicing-induced frameshift enrichment test (same-start pairs)
fs_test_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  fc_c2 <- set_results[[c2_key]]$frame_comp
  fc_c4 <- set_results[[c4_key]]$frame_comp
  if (is.null(fc_c2) || is.null(fc_c4)) return(NULL)
  ps_c2 <- fc_c2$pair_summary
  ps_c4 <- fc_c4$pair_summary

  n_ss_fs_c2 <- sum(ps_c2$frame_category == "same_start_frameshift")
  n_ss_if_c2 <- sum(ps_c2$frame_category == "same_start_in_frame")
  n_ss_fs_c4 <- sum(ps_c4$frame_category == "same_start_frameshift")
  n_ss_if_c4 <- sum(ps_c4$frame_category == "same_start_in_frame")
  fi <- tryCatch(
    fisher.test(matrix(c(n_ss_fs_c2, n_ss_if_c2, n_ss_fs_c4, n_ss_if_c4), nrow = 2)),
    error = function(e) NULL)

  data.frame(
    cell_type = ct,
    NMD_ss_total = n_ss_fs_c2 + n_ss_if_c2,
    NMD_fs_pct = round(100 * n_ss_fs_c2 / max(n_ss_fs_c2 + n_ss_if_c2, 1), 1),
    Ctrl_ss_total = n_ss_fs_c4 + n_ss_if_c4,
    Ctrl_fs_pct = round(100 * n_ss_fs_c4 / max(n_ss_fs_c4 + n_ss_if_c4, 1), 1),
    OR = if (!is.null(fi)) round(fi$estimate, 2) else NA_real_,
    p_value = if (!is.null(fi)) fi$p.value else NA_real_
  )
}))

if (!is.null(mech_rows)) {
  # Summary table: wide format showing NMD vs Control side by side
  mech_wide <- do.call(rbind, lapply(display_cts, function(ct) {
    nmd  <- mech_rows[mech_rows$cell_type == ct & mech_rows$comparison == "NMD", ]
    ctrl <- mech_rows[mech_rows$cell_type == ct & mech_rows$comparison == "Control", ]
    if (nrow(nmd) == 0 || nrow(ctrl) == 0) return(NULL)
    data.frame(
      cell_type = ct,
      frame_category = nmd$frame_category,
      NMD_n = nmd$n_ptc, NMD_pct = nmd$pct_of_ptc,
      Ctrl_n = ctrl$n_ptc, Ctrl_pct = ctrl$pct_of_ptc
    )
  }))
  rownames(mech_wide) <- NULL
  knitr::kable(mech_wide, row.names = FALSE,
    caption = "PTC Mechanism: Frame Category of PTC+ Comparators (NMD vs Control)") |> print()
  cat("\n\n")
}

if (!is.null(fs_test_rows)) {
  rownames(fs_test_rows) <- NULL
  knitr::kable(fs_test_rows, row.names = FALSE,
    caption = "Splicing-Induced Frameshift Enrichment (Same-Start Pairs): NMD vs Control") |> print()
  cat("\n\n")
}

# --- Bar chart: PTC mechanism distribution ---
if (!is.null(mech_rows) && has_ggplot2) {
  mech_rows$category_label <- factor(mech_labels[mech_rows$frame_category],
                                     levels = mech_labels[coding_cats])

  # Fisher's test per cell_type × category
  p_labels <- do.call(rbind, lapply(unique(mech_rows$cell_type), function(ct) {
    do.call(rbind, lapply(coding_cats, function(cat) {
      nmd  <- mech_rows[mech_rows$cell_type == ct & mech_rows$comparison == "NMD", ]
      ctrl <- mech_rows[mech_rows$cell_type == ct & mech_rows$comparison == "Control", ]
      n_nmd <- nmd$n_ptc[nmd$frame_category == cat]
      t_nmd <- nmd$ptc_total[nmd$frame_category == cat]
      n_ctrl <- ctrl$n_ptc[ctrl$frame_category == cat]
      t_ctrl <- ctrl$ptc_total[ctrl$frame_category == cat]
      if (length(n_nmd) == 0 || length(n_ctrl) == 0) return(NULL)

      ft <- tryCatch({
        mat <- matrix(c(n_nmd, t_nmd - n_nmd, n_ctrl, t_ctrl - n_ctrl), nrow = 2)
        fisher.test(mat)
      }, error = function(e) list(p.value = NA))
      p <- ft$p.value
      stars <- if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else "ns"
      max_pct <- max(mech_rows$pct_of_ptc[mech_rows$cell_type == ct &
                                           mech_rows$frame_category == cat], na.rm = TRUE)
      data.frame(cell_type = ct, category_label = mech_labels[cat],
                 p_star = stars, max_pct = max_pct)
    }))
  }))
  p_labels$category_label <- factor(p_labels$category_label, levels = mech_labels[coding_cats])

  p_mech <- ggplot(mech_rows, aes(x = comparison, y = pct_of_ptc, fill = comparison)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = n_ptc), vjust = -0.3, size = 3) +
    geom_text(data = p_labels, aes(x = 1.5, y = max_pct + 5, label = p_star),
              inherit.aes = FALSE, size = 4, fontface = "bold") +
    facet_grid(cell_type ~ category_label, scales = "free_y") +
    scale_fill_manual(values = c("NMD" = "#ef8a62", "Control" = "#67a9cf"), name = NULL) +
    theme_minimal(base_size = 12) +
    theme(strip.text = element_text(size = 10, face = "bold"),
          axis.text.x = element_text(angle = 0, hjust = 0.5),
          legend.position = "bottom",
          panel.grid.major.x = element_blank()) +
    labs(x = NULL, y = "% of PTC+ comparators",
         title = "PTC Mechanism: How Do PTCs Arise?",
         subtitle = "Distribution of PTC+ comparators across frame categories. Numbers = pair counts. *** p<0.001, ** p<0.01, * p<0.05 (Fisher's exact test)")
  print(p_mech)
  cat("\n\n")
}
```

## Frameshift Consequences

Among pairs that share the same start codon, splicing-induced frameshifts are common in both
NMD and Control comparisons, but their downstream consequences differ sharply. In NMD pairs,
frameshifts overwhelmingly produce PTCs that trigger decay. In Control pairs, most frameshifts
place the new stop codon in the last exon (0 downstream EJCs), producing stable proteins with
rewritten C-terminal sequences — productive alternative splicing generating protein diversity.

### Frameshift Outcome Distribution

Same-start coding pairs are partitioned into four mutually exclusive outcome categories, each
expressed as a percentage of all same-start coding pairs (so bars across categories sum to
100% within each cell type). This uses the cross-isoform reading frame comparison
(`compareIsoformFrames()` in `Isopair/R/ptc.R`; see METHODS.md §Reading Frame and PTC
Analysis). Only pairs classified as `same_start_in_frame` or `same_start_frameshift` are
included, since these share translation initiation and any frame difference is attributable
to splicing.

The four categories are:

- **In Frame**: Same start codon, no frameshift — splicing differences preserve reading frame.
- **Frameshift → PTC**: Frameshift produces a premature stop codon (>50 nt upstream of last
  EJC), triggering NMD.
- **Frameshift → Stop in Last Exon**: Frameshift without PTC; stop codon falls in the last
  exon (0 downstream EJCs), producing a stable protein with novel C-terminal sequence.
- **Frameshift → Other**: Frameshift without PTC and stop not in last exon.

```{r sec3-frameshift-outcome, results='asis', fig.width=14, fig.height=10}
outcome_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  rows <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) next
    fc <- set_results[[key]]$frame_comp
    if (is.null(fc) || is.null(fc$pair_summary)) next
    ps <- fc$pair_summary

    # Same-start coding pairs
    ss <- ps[ps$frame_category %in% c("same_start_in_frame", "same_start_frameshift"), ]
    n_same_start <- nrow(ss)
    n_in_frame <- sum(ss$frame_category == "same_start_in_frame")

    # Merge frameshifted pairs with PTC data
    fs_pairs <- ss[ss$frame_category == "same_start_frameshift", ]
    n_fs_ptc <- 0L; n_fs_stop_last <- 0L; n_fs_other <- 0L
    if (nrow(fs_pairs) > 0) {
      fs_ptc <- merge(fs_pairs, ptc[, c("isoform_id", "has_ptc", "stop_in_last_exon")],
                      by.x = "comparator_isoform_id", by.y = "isoform_id", all.x = TRUE)
      n_fs_ptc       <- sum(fs_ptc$has_ptc == TRUE, na.rm = TRUE)
      no_ptc_mask    <- !is.na(fs_ptc$has_ptc) & fs_ptc$has_ptc == FALSE
      n_fs_stop_last <- sum(no_ptc_mask & fs_ptc$stop_in_last_exon == TRUE, na.rm = TRUE)
      n_fs_other     <- sum(no_ptc_mask & (is.na(fs_ptc$stop_in_last_exon) |
                                            fs_ptc$stop_in_last_exon == FALSE))
    }

    rows[[comp]] <- data.frame(
      cell_type = ct, comparison = comp_label(comp),
      n_same_start = n_same_start,
      n_in_frame = n_in_frame,
      n_fs_ptc = n_fs_ptc,
      n_fs_stop_last = n_fs_stop_last,
      n_fs_other = n_fs_other
    )
  }
  do.call(rbind, rows)
}))

if (!is.null(outcome_rows)) {
  # Summary table with percentages
  outcome_tbl <- outcome_rows
  outcome_tbl$in_frame_pct <- round(100 * outcome_tbl$n_in_frame / outcome_tbl$n_same_start, 1)
  outcome_tbl$fs_ptc_pct <- round(100 * outcome_tbl$n_fs_ptc / outcome_tbl$n_same_start, 1)
  outcome_tbl$fs_stop_last_pct <- round(100 * outcome_tbl$n_fs_stop_last / outcome_tbl$n_same_start, 1)
  outcome_tbl$fs_other_pct <- round(100 * outcome_tbl$n_fs_other / outcome_tbl$n_same_start, 1)
  rownames(outcome_tbl) <- NULL
  knitr::kable(outcome_tbl[, c("cell_type", "comparison", "n_same_start",
                                "n_in_frame", "in_frame_pct",
                                "n_fs_ptc", "fs_ptc_pct",
                                "n_fs_stop_last", "fs_stop_last_pct",
                                "n_fs_other", "fs_other_pct")],
    row.names = FALSE,
    caption = "Frameshift Outcome Distribution: NMD vs Control (Same-Start Pairs)") |> print()
  cat("\n\n")
}

# --- Stacked bar chart: outcomes as % of same-start pairs ---
if (!is.null(outcome_rows) && has_ggplot2) {
  outcome_cats <- c("In Frame", "Frameshift\n→ PTC",
                    "Frameshift\n→ Stop in Last Exon", "Frameshift\n→ Other")

  outcome_plot_df <- do.call(rbind, lapply(seq_len(nrow(outcome_rows)), function(i) {
    r <- outcome_rows[i, ]
    n <- r$n_same_start
    data.frame(
      cell_type = r$cell_type,
      comparison = r$comparison,
      outcome = outcome_cats,
      n = c(r$n_in_frame, r$n_fs_ptc, r$n_fs_stop_last, r$n_fs_other),
      pct = 100 * c(r$n_in_frame, r$n_fs_ptc, r$n_fs_stop_last, r$n_fs_other) / max(n, 1)
    )
  }))
  outcome_plot_df$outcome <- factor(outcome_plot_df$outcome, levels = outcome_cats)

  # Fisher's test per cell_type × outcome category
  p_labels <- do.call(rbind, lapply(unique(outcome_rows$cell_type), function(ct) {
    nmd <- outcome_rows[outcome_rows$cell_type == ct & outcome_rows$comparison == "NMD", ]
    ctrl <- outcome_rows[outcome_rows$cell_type == ct & outcome_rows$comparison == "Control", ]
    if (nrow(nmd) == 0 || nrow(ctrl) == 0) return(NULL)

    count_cols <- c("n_in_frame", "n_fs_ptc", "n_fs_stop_last", "n_fs_other")
    do.call(rbind, lapply(seq_along(outcome_cats), function(j) {
      n_yes_nmd  <- nmd[[count_cols[j]]]
      n_yes_ctrl <- ctrl[[count_cols[j]]]
      n_tot_nmd  <- nmd$n_same_start
      n_tot_ctrl <- ctrl$n_same_start

      ft <- tryCatch({
        mat <- matrix(c(n_yes_nmd, n_tot_nmd - n_yes_nmd,
                        n_yes_ctrl, n_tot_ctrl - n_yes_ctrl), nrow = 2)
        fisher.test(mat)
      }, error = function(e) list(p.value = NA))
      p <- ft$p.value
      stars <- if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else "ns"

      max_pct <- max(outcome_plot_df$pct[outcome_plot_df$cell_type == ct &
                                          outcome_plot_df$outcome == outcome_cats[j]], na.rm = TRUE)
      data.frame(cell_type = ct, outcome = outcome_cats[j],
                 p_star = stars, max_pct = max_pct)
    }))
  }))
  p_labels$outcome <- factor(p_labels$outcome, levels = outcome_cats)

  # Bar totals for count labels
  bar_totals <- outcome_plot_df

  p_out <- ggplot(outcome_plot_df, aes(x = comparison, y = pct, fill = comparison)) +
    geom_col(width = 0.7) +
    geom_text(data = bar_totals, aes(x = comparison, y = pct, label = n, fill = NULL),
              vjust = -0.3, size = 3) +
    geom_text(data = p_labels, aes(x = 1.5, y = max_pct + 4, label = p_star),
              inherit.aes = FALSE, size = 4, fontface = "bold") +
    facet_grid(cell_type ~ outcome, scales = "free_y") +
    scale_fill_manual(values = c("NMD" = "#ef8a62", "Control" = "#67a9cf"),
                      name = NULL) +
    theme_minimal(base_size = 12) +
    theme(strip.text = element_text(size = 10, face = "bold"),
          axis.text.x = element_text(angle = 0, hjust = 0.5),
          legend.position = "bottom",
          panel.grid.major.x = element_blank()) +
    labs(x = NULL, y = "% of same-start pairs",
         title = "Frameshift Outcome Distribution: NMD vs Control",
         subtitle = "All categories sum to 100% within each cell type. Numbers above bars = pair counts. *** p<0.001, ** p<0.01, * p<0.05 (Fisher's exact test)")
  print(p_out)
  cat("\n\n")
}
```

In NMD comparisons, splicing-induced frameshifts overwhelmingly produce PTCs that trigger decay.
In Control comparisons, most frameshifts place the new stop codon in the last exon
(0 downstream EJCs), producing a stable mRNA with a novel C-terminal protein sequence.

### Stable Novel Proteins (Control)

Control (C4) pairs with same-start splicing-induced frameshifts and no PTC represent productive
alternative splicing: the frameshift rewrites the C-terminal protein sequence without triggering
NMD. The analysis below quantifies how much of each protein is novel — i.e., translated in a
non-original reading frame downstream of the first divergent CDS region.

```{r sec3-novel-protein}
# Compute novel protein metrics for C4 same_start_frameshift pairs without PTC
novel_protein_results <- list()

for (ct in display_cts) {
  if (ct == "all_samples") next
  ct_label <- ct_to_label(ct)
  c4_key <- paste0("c4_", ct_label)
  if (!c4_key %in% names(set_results)) next
  fc <- set_results[[c4_key]]$frame_comp
  if (is.null(fc)) next

  novel_protein_results[[ct]] <- cached_compute(paste0("novel_protein_c4_v2_", ct_label), {
    ps <- fc$pair_summary
    rd <- fc$region_detail

    # Same-start frameshift pairs
    fs_pairs <- ps[ps$frame_category == "same_start_frameshift", ]
    if (nrow(fs_pairs) == 0) return(NULL)

    # Merge with ptc: keep only no-PTC comparators
    fs_merged <- merge(fs_pairs,
                       ptc[, c("isoform_id", "has_ptc", "orf_length")],
                       by.x = "comparator_isoform_id", by.y = "isoform_id", all.x = TRUE)
    fs_merged <- fs_merged[!is.na(fs_merged$has_ptc) & fs_merged$has_ptc == FALSE, ]
    if (nrow(fs_merged) == 0) return(NULL)

    # Get strand from cds
    strand_lookup <- setNames(cds$strand, cds$isoform_id)

    # For each pair, find the 5'-most divergent CDS region as the frameshift boundary
    results <- lapply(seq_len(nrow(fs_merged)), function(i) {
      row <- fs_merged[i, ]
      comp_id <- row$comparator_isoform_id
      ref_id  <- row$reference_isoform_id
      s <- strand_lookup[comp_id]
      if (is.na(s)) return(NULL)

      # Get region detail for this pair: find first region where phases differ
      pair_rd <- rd[rd$reference_isoform_id == ref_id &
                    rd$comparator_isoform_id == comp_id, ]
      if (nrow(pair_rd) == 0) return(NULL)
      div_regions <- pair_rd[pair_rd$ref_phase_at_start != pair_rd$comp_phase_at_start, ]
      if (nrow(div_regions) == 0) return(NULL)

      # 5'-most divergent region boundary
      if (s == "+") {
        fs_boundary <- min(div_regions$region_start)
      } else {
        fs_boundary <- max(div_regions$region_end)
      }

      conserved_bp <- count_conserved_cds_bp(comp_id, fs_boundary, s, cds, structures)
      if (is.na(conserved_bp)) return(NULL)
      comp_orf <- row$orf_length
      novel_bp <- max(0L, comp_orf - conserved_bp)
      data.frame(
        gene_id = row$gene_id,
        comparator_isoform_id = comp_id,
        orf_length = comp_orf,
        conserved_bp = conserved_bp,
        novel_bp = novel_bp,
        novel_aa = novel_bp / 3,
        total_aa = comp_orf / 3,
        pct_novel = 100 * (novel_bp / 3) / (comp_orf / 3)
      )
    })
    do.call(rbind, results)
  })
}
```

**Columns:** **n_stable_novel** = C4 same-start frameshifts without PTC; **pct_of_same_start** = as % of all C4 same-start coding pairs; **median_novel_aa / iqr_novel_aa** = novel amino acids (translated in shifted frame); **median_total_aa** = total protein length; **median_pct_novel / iqr_pct_novel** = % of protein that is novel sequence.

```{r sec3-novel-protein-table, results='asis'}
novel_summary <- do.call(rbind, lapply(names(novel_protein_results), function(ct) {
  np <- novel_protein_results[[ct]]
  if (is.null(np) || nrow(np) == 0) return(NULL)
  ct_label <- ct_to_label(ct)
  c4_key <- paste0("c4_", ct_label)
  fc <- set_results[[c4_key]]$frame_comp
  n_same_start <- sum(fc$pair_summary$frame_category %in%
    c("same_start_in_frame", "same_start_frameshift"))
  data.frame(
    cell_type = ct,
    n_stable_novel = nrow(np),
    pct_of_same_start = round(100 * nrow(np) / n_same_start, 1),
    median_novel_aa = round(median(np$novel_aa), 0),
    iqr_novel_aa = paste0(round(quantile(np$novel_aa, 0.25), 0), "-",
                          round(quantile(np$novel_aa, 0.75), 0)),
    median_total_aa = round(median(np$total_aa), 0),
    median_pct_novel = round(median(np$pct_novel), 1),
    iqr_pct_novel = paste0(round(quantile(np$pct_novel, 0.25), 1), "-",
                           round(quantile(np$pct_novel, 0.75), 1))
  )
}))

if (!is.null(novel_summary)) {
  knitr::kable(novel_summary, caption = "Stable Novel Proteins: Control (C4) Same-Start Frameshifts Without PTC") |> print()
  cat("\n\n")
}
```

```{r sec3-novel-protein-plot, fig.height=5}
if (has_ggplot2) {
  novel_plot_data <- do.call(rbind, lapply(names(novel_protein_results), function(ct) {
    np <- novel_protein_results[[ct]]
    if (is.null(np) || nrow(np) == 0) return(NULL)
    data.frame(cell_type = ct, pct_novel = np$pct_novel)
  }))

  if (!is.null(novel_plot_data) && nrow(novel_plot_data) > 0) {
    p_novel <- ggplot(novel_plot_data, aes(x = pct_novel)) +
      geom_density(fill = "steelblue", alpha = 0.5) +
      facet_wrap(~cell_type, scales = "free_y") +
      theme_minimal() +
      labs(title = "Novel Protein Sequence in Stable Control Frameshifts",
           subtitle = "C4 unresolved frameshifts without PTC — % of protein in shifted reading frame",
           x = "% of protein that is novel sequence", y = "Density")
    print(p_novel)
  }
}
```

### Frameshift Examples: Real Gene Cases

Three real gene examples illustrating the major categories of frameshift outcomes among
same-start pairs. Panel A: a splicing-induced frameshift that produces a PTC, triggering NMD.
Panel B: a control pair classified as in-frame (same start codon, same reading frame).
Panel C: a same-start frameshift without PTC, producing a stable protein with novel
C-terminal sequence.

Each example is shown first as a full-gene overview, then zoomed to the region containing
the relevant splicing event(s).

```{r sec3-frameshift-examples, fig.width=12, fig.height=18}
if (has_ggplot2) {
  # Helper: resolve gene symbol from gene_id
  get_gene_sym <- function(gid) {
    sym <- gene_symbol_lookup[gid]
    if (is.na(sym) || sym == "") gid else sym
  }

  # Helper: compute zoom xlim around events, with padding
  compute_event_zoom <- function(events, padding_frac = 0.3) {
    if (is.null(events) || nrow(events) == 0) return(NULL)
    ev_min <- min(events$five_prime, events$three_prime, na.rm = TRUE)
    ev_max <- max(events$five_prime, events$three_prime, na.rm = TRUE)
    span <- ev_max - ev_min
    pad <- max(span * padding_frac, 500)
    c(ev_min - pad, ev_max + pad)
  }

  # Use DD cell type as representative (largest dataset)
  ex_ct <- "dd"
  c2_key <- paste0("c2_", ex_ct)
  c4_key <- paste0("c4_", ex_ct)

  fc_c2 <- set_results[[c2_key]]$frame_comp
  fc_c4 <- set_results[[c4_key]]$frame_comp

  example_plots <- list()

  # --- Panel A: Same-start frameshift → PTC (NMD) ---
  if (!is.null(fc_c2)) {
    ps_c2 <- fc_c2$pair_summary
    fs_c2 <- ps_c2[ps_c2$frame_category == "same_start_frameshift", ]
    fs_c2_ptc <- merge(fs_c2, ptc[ptc$has_ptc == TRUE, c("isoform_id", "has_ptc",
                       "ptc_distance", "n_downstream_ejcs")],
                       by.x = "comparator_isoform_id", by.y = "isoform_id")
    fs_c2_ptc <- merge(fs_c2_ptc,
                       structures[, c("isoform_id", "n_exons")],
                       by.x = "comparator_isoform_id", by.y = "isoform_id")
    names(fs_c2_ptc)[names(fs_c2_ptc) == "n_exons"] <- "comp_n_exons"
    fs_c2_ptc <- merge(fs_c2_ptc,
                       structures[, c("isoform_id", "n_exons")],
                       by.x = "reference_isoform_id", by.y = "isoform_id")
    names(fs_c2_ptc)[names(fs_c2_ptc) == "n_exons"] <- "ref_n_exons"
    fs_c2_ptc$exon_diff <- abs(fs_c2_ptc$ref_n_exons - fs_c2_ptc$comp_n_exons)
    candidates <- fs_c2_ptc[fs_c2_ptc$comp_n_exons >= 3 & fs_c2_ptc$comp_n_exons <= 10 &
                             fs_c2_ptc$exon_diff <= 1, ]
    if (nrow(candidates) > 0) {
      pick_a <- candidates[which.max(candidates$ptc_distance), ]
      ref_ex <- get_exon_coords(pick_a$reference_isoform_id, structures)
      comp_ex <- get_exon_coords(pick_a$comparator_isoform_id, structures)
      ref_cds_a <- get_cds_range(pick_a$reference_isoform_id, cds)
      comp_cds_a <- get_cds_range(pick_a$comparator_isoform_id, cds)
      str_row <- structures[structures$isoform_id == pick_a$reference_isoform_id, ]
      gene_strand <- str_row$strand[1]

      profs <- set_results[[c2_key]]$profiles
      pair_prof <- profs[profs$reference_isoform_id == pick_a$reference_isoform_id &
                         profs$comparator_isoform_id == pick_a$comparator_isoform_id, ]
      pair_events <- if (nrow(pair_prof) > 0) pair_prof$detailed_events[[1]] else NULL

      ptc_ann <- list()
      if (!is.null(comp_cds_a)) {
        ptc_pos <- if (gene_strand == "+") comp_cds_a$cds_stop else comp_cds_a$cds_start
        ptc_ann <- list(
          list(type = "ptc_marker", x = ptc_pos, track = "comp",
               label = "PTC", color = "#E74C3C")
        )
      }

      gene_sym_a <- get_gene_sym(pick_a$gene_id)
      title_a <- sprintf("A) Frameshift \u2192 PTC: %s (%s)", gene_sym_a, pick_a$gene_id)
      sub_a <- sprintf("Ref: %s | Comp: %s\nPTC distance = %d nt, %d downstream EJC(s)",
                       pick_a$reference_isoform_id, pick_a$comparator_isoform_id,
                       round(pick_a$ptc_distance), pick_a$n_downstream_ejcs)

      # Full view
      example_plots[["A_full"]] <- plot_ucsc_pair(
        ref_ex$starts, ref_ex$ends,
        comp_ex$starts, comp_ex$ends,
        gene_strand,
        ref_cds = ref_cds_a, comp_cds = comp_cds_a,
        ref_label = "Dominant\n(non-NMD)",
        comp_label = "NMD-sensitive",
        events = pair_events,
        annotations = ptc_ann,
        title = title_a, subtitle = sub_a
      )

      # Zoomed view
      zoom_a <- compute_event_zoom(pair_events)
      if (!is.null(zoom_a)) {
        example_plots[["A_zoom"]] <- plot_ucsc_pair(
          ref_ex$starts, ref_ex$ends,
          comp_ex$starts, comp_ex$ends,
          gene_strand,
          ref_cds = ref_cds_a, comp_cds = comp_cds_a,
          ref_label = "Dominant\n(non-NMD)",
          comp_label = "NMD-sensitive",
          events = pair_events,
          annotations = ptc_ann,
          title = paste0(title_a, " [ZOOM]"),
          subtitle = sub_a,
          xlim = zoom_a
        )
      }
    }
  }

  # --- Panel B: Same-start In-Frame (Control C4) — from Control (C4) ---
  pick_b <- NULL
  pick_b_key <- NULL
  for (search_ct in cell_types) {
    ct_lab <- tolower(gsub("_", "", search_ct))
    search_key <- paste0("c4_", ct_lab)
    if (!search_key %in% names(set_results)) next
    fc_search <- set_results[[search_key]]$frame_comp
    if (is.null(fc_search) || is.null(fc_search$pair_summary)) next
    ps_if <- fc_search$pair_summary[fc_search$pair_summary$frame_category == "same_start_in_frame", ]
    if (nrow(ps_if) == 0) next
    # Prefer pairs with a few splicing events for visual interest
    profs_search <- set_results[[search_key]]$profiles
    ps_if <- merge(ps_if, profs_search[, c("reference_isoform_id", "comparator_isoform_id", "n_events")],
                   by = c("reference_isoform_id", "comparator_isoform_id"))
    ps_if <- merge(ps_if,
                   structures[, c("isoform_id", "n_exons")],
                   by.x = "comparator_isoform_id", by.y = "isoform_id")
    candidates_b <- ps_if[ps_if$n_exons >= 3 & ps_if$n_exons <= 10 &
                           ps_if$n_events >= 1 & ps_if$n_events <= 3, ]
    if (nrow(candidates_b) > 0) {
      pick_b <- candidates_b[1, ]
      pick_b_key <- search_key
      break
    }
  }
  if (!is.null(pick_b)) {
    ref_ex_b <- get_exon_coords(pick_b$reference_isoform_id, structures)
    comp_ex_b <- get_exon_coords(pick_b$comparator_isoform_id, structures)
    ref_cds_b <- get_cds_range(pick_b$reference_isoform_id, cds)
    comp_cds_b <- get_cds_range(pick_b$comparator_isoform_id, cds)
    str_row_b <- structures[structures$isoform_id == pick_b$reference_isoform_id, ]
    gene_strand_b <- str_row_b$strand[1]

    profs_b <- set_results[[pick_b_key]]$profiles
    pair_prof_b <- profs_b[profs_b$reference_isoform_id == pick_b$reference_isoform_id &
                           profs_b$comparator_isoform_id == pick_b$comparator_isoform_id, ]
    pair_events_b <- if (nrow(pair_prof_b) > 0) pair_prof_b$detailed_events[[1]] else NULL

    gene_sym_b <- get_gene_sym(pick_b$gene_id)
    title_b <- sprintf("B) In-Frame Splicing (Control): %s (%s)",
                       gene_sym_b, pick_b$gene_id)
    sub_b <- sprintf("Ref: %s | Comp: %s\nSame start codon, splicing preserves reading frame",
                     pick_b$reference_isoform_id, pick_b$comparator_isoform_id)

    example_plots[["B_full"]] <- plot_ucsc_pair(
      ref_ex_b$starts, ref_ex_b$ends,
      comp_ex_b$starts, comp_ex_b$ends,
      gene_strand_b,
      ref_cds = ref_cds_b, comp_cds = comp_cds_b,
      ref_label = "Dominant\n(non-NMD)",
      comp_label = "Next-best\n(non-NMD)",
      events = pair_events_b,
      title = title_b, subtitle = sub_b
    )

    zoom_b <- compute_event_zoom(pair_events_b)
    if (!is.null(zoom_b)) {
      example_plots[["B_zoom"]] <- plot_ucsc_pair(
        ref_ex_b$starts, ref_ex_b$ends,
        comp_ex_b$starts, comp_ex_b$ends,
        gene_strand_b,
        ref_cds = ref_cds_b, comp_cds = comp_cds_b,
        ref_label = "Dominant\n(non-NMD)",
        comp_label = "Next-best\n(non-NMD)",
        events = pair_events_b,
        title = paste0(title_b, " [ZOOM]"),
        subtitle = sub_b,
        xlim = zoom_b
      )
    }
  }

  # --- Panel C: Same-start Frameshift → Novel Protein (Control, no PTC) ---
  np_dd <- novel_protein_results[["DD"]]
  if (!is.null(np_dd) && nrow(np_dd) > 0 && !is.null(fc_c4)) {
    np_cands <- np_dd[np_dd$pct_novel >= 40 & np_dd$pct_novel <= 80, ]
    if (nrow(np_cands) > 0) {
      np_cands <- merge(np_cands,
                        structures[, c("isoform_id", "n_exons")],
                        by.x = "comparator_isoform_id", by.y = "isoform_id")
      np_cands <- np_cands[np_cands$n_exons >= 3 & np_cands$n_exons <= 8, ]
    }
    if (nrow(np_cands) > 0) {
      np_cands <- np_cands[order(abs(np_cands$pct_novel - median(np_cands$pct_novel))), ]
      pick_c <- np_cands[1, ]

      ps_fc4 <- fc_c4$pair_summary
      pair_info <- ps_fc4[ps_fc4$comparator_isoform_id == pick_c$comparator_isoform_id &
                          ps_fc4$gene_id == pick_c$gene_id, ]
      if (nrow(pair_info) > 0) {
        ref_id_c <- pair_info$reference_isoform_id[1]
        ref_ex_c <- get_exon_coords(ref_id_c, structures)
        comp_ex_c <- get_exon_coords(pick_c$comparator_isoform_id, structures)
        ref_cds_c <- get_cds_range(ref_id_c, cds)
        comp_cds_c <- get_cds_range(pick_c$comparator_isoform_id, cds)
        str_row_c <- structures[structures$isoform_id == ref_id_c, ]
        gene_strand_c <- str_row_c$strand[1]

        profs_c <- set_results[[c4_key]]$profiles
        pair_prof_c <- profs_c[profs_c$reference_isoform_id == ref_id_c &
                               profs_c$comparator_isoform_id == pick_c$comparator_isoform_id, ]
        pair_events_c <- if (nrow(pair_prof_c) > 0) pair_prof_c$detailed_events[[1]] else NULL

        # Find first divergent CDS region from frame comparison
        rd_c <- fc_c4$region_detail
        pair_rd_c <- rd_c[rd_c$reference_isoform_id == ref_id_c &
                          rd_c$comparator_isoform_id == pick_c$comparator_isoform_id, ]
        div_rd_c <- pair_rd_c[pair_rd_c$ref_phase_at_start != pair_rd_c$comp_phase_at_start, ]
        fs_ann <- list()
        if (nrow(div_rd_c) > 0) {
          if (gene_strand_c == "+") {
            fs_pos <- min(div_rd_c$region_start)
          } else {
            fs_pos <- max(div_rd_c$region_end)
          }
          if (!is.null(comp_cds_c)) {
            novel_end <- if (gene_strand_c == "+") comp_cds_c$cds_stop else comp_cds_c$cds_start
            novel_start <- if (gene_strand_c == "+") fs_pos else novel_end
            novel_stop <- if (gene_strand_c == "+") novel_end else fs_pos
            fs_ann <- list(
              list(type = "frameshift_boundary", x = fs_pos, track = "comp",
                   label = "Frameshift", color = "#8E44AD"),
              list(type = "region_shade",
                   xmin = min(novel_start, novel_stop),
                   xmax = max(novel_start, novel_stop),
                   track = "comp", color = "#E67E22", alpha = 0.3,
                   label = sprintf("Novel protein (%d aa, %.0f%%)",
                                   round(pick_c$novel_aa), pick_c$pct_novel))
            )
          }
        }

        gene_sym_c <- get_gene_sym(pick_c$gene_id)
        title_c <- sprintf("C) Novel Protein \u2014 Stable Frameshift: %s (%s)",
                           gene_sym_c, pick_c$gene_id)
        sub_c <- sprintf("Ref: %s | Comp: %s\nSame-start frameshift, no PTC, stop in last exon \u2192 %d novel aa (%.0f%%)",
                         ref_id_c, pick_c$comparator_isoform_id,
                         round(pick_c$novel_aa), pick_c$pct_novel)

        example_plots[["C_full"]] <- plot_ucsc_pair(
          ref_ex_c$starts, ref_ex_c$ends,
          comp_ex_c$starts, comp_ex_c$ends,
          gene_strand_c,
          ref_cds = ref_cds_c, comp_cds = comp_cds_c,
          ref_label = "Dominant\n(non-NMD)",
          comp_label = "Next-best\n(non-NMD)",
          events = pair_events_c,
          annotations = fs_ann,
          title = title_c, subtitle = sub_c
        )

        zoom_c <- compute_event_zoom(pair_events_c)
        if (!is.null(zoom_c)) {
          example_plots[["C_zoom"]] <- plot_ucsc_pair(
            ref_ex_c$starts, ref_ex_c$ends,
            comp_ex_c$starts, comp_ex_c$ends,
            gene_strand_c,
            ref_cds = ref_cds_c, comp_cds = comp_cds_c,
            ref_label = "Dominant\n(non-NMD)",
            comp_label = "Next-best\n(non-NMD)",
            events = pair_events_c,
            annotations = fs_ann,
            title = paste0(title_c, " [ZOOM]"),
            subtitle = sub_c,
            xlim = zoom_c
          )
        }
      }
    }
  }

  # Combine panels: full + zoom pairs
  if (length(example_plots) > 0) {
    if (requireNamespace("patchwork", quietly = TRUE)) {
      combined <- patchwork::wrap_plots(example_plots, ncol = 1) +
        patchwork::plot_annotation(
          title = "Frameshift Consequences: Representative Gene Examples",
          theme = theme(plot.title = element_text(size = 14, face = "bold"))
        )
      print(combined)
    } else {
      for (pp in example_plots) print(pp)
    }
  }
}
```

## Stop Codon Distance to Last EJC

The distance from a transcript's stop codon to the last exon-exon junction (EJC) is a
key determinant of NMD susceptibility. When a stop codon sits >50 nt upstream of the last
EJC, the downstream junction(s) remain associated with the exon junction complex after
splicing, triggering NMD surveillance. Greater distance (more downstream EJCs) is generally
associated with more efficient NMD.

The distance is computed in mRNA coordinates as `last_EJC_position - stop_codon_position`.
Positive values indicate the stop codon is upstream of the last EJC (potential NMD trigger);
negative values indicate a stop codon downstream of the last EJC (normal termination).
Single-exon isoforms have no EJC and are excluded.

### Distribution of Stop-to-EJC Distance: All Comparators

Distribution of stop codon-to-last-EJC distances for **all** comparator isoforms (not just
PTC+), stratified by NMD and Control. This shows the full landscape of stop codon positioning
rather than conditioning on the PTC+ tail.

```{r sec3-stop-dist, results='asis', fig.height=6}
# Build stop-to-EJC distance data for all comparator isoforms
stop_dist_all <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  rows <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) next
    comp_ids <- set_results[[key]]$profiles$comparator_isoform_id
    ptc_sub <- ptc[ptc$isoform_id %in% comp_ids & !is.na(ptc$ptc_distance), ]
    if (nrow(ptc_sub) == 0) next
    rows[[comp]] <- data.frame(
      cell_type = ct,
      comparison = comp_label(comp),
      ptc_distance = ptc_sub$ptc_distance,
      has_ptc = ptc_sub$has_ptc
    )
  }
  do.call(rbind, rows)
}))

if (!is.null(stop_dist_all) && nrow(stop_dist_all) > 0 && has_ggplot2) {
  # Summary table
  dist_summary <- stop_dist_all |>
    group_by(cell_type, comparison) |>
    summarize(
      n_isoforms = n(),
      median_dist = round(median(ptc_distance), 0),
      pct_ptc_pos = round(100 * mean(has_ptc), 1),
      pct_negative = round(100 * mean(ptc_distance < 0), 1),
      .groups = "drop"
    )
  cat("**Columns:** **n_isoforms** = comparator isoforms with stop-to-EJC distance computed; ")
  cat("**median_dist** = median distance (nt); **pct_ptc_pos** = % with PTC (distance > 50 nt); ")
  cat("**pct_negative** = % with stop codon downstream of last EJC (normal termination).\n\n")
  knitr::kable(dist_summary, caption = "Stop Codon-to-Last EJC Distance: All Comparators") |> print()
  cat("\n\n")

  # Density plot
  p_dist <- ggplot(stop_dist_all, aes(x = ptc_distance, fill = comparison)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
    annotate("text", x = 50, y = Inf, label = "PTC threshold (50 nt)",
             hjust = -0.05, vjust = 1.5, size = 3, color = "red") +
    facet_wrap(~cell_type, scales = "free_y") +
    scale_x_continuous(limits = c(
      quantile(stop_dist_all$ptc_distance, 0.01, na.rm = TRUE),
      quantile(stop_dist_all$ptc_distance, 0.99, na.rm = TRUE))) +
    theme_minimal() +
    labs(title = "Stop Codon Distance to Last EJC: NMD vs Control Comparators",
         subtitle = "All comparator isoforms (coding, multi-exon). Dashed red = 50 nt PTC threshold.",
         x = "Distance from stop codon to last EJC (nt)", y = "Density", fill = "")
  print(p_dist)
  cat("\n\n")
}
```

### NMD Response vs Stop Codon Distance

Among NMD-sensitive comparator isoforms, is there a relationship between the distance from
the stop codon to the last EJC and the magnitude of upregulation upon NMD inhibition (logFC)?
Isoforms with stop codons further from the last EJC may be subject to more efficient NMD
surveillance and thus show stronger rescue (higher logFC) when NMD is inhibited.

```{r sec3-nmd-logfc-dist, results='asis', fig.height=6}
# Load mashr DE data to get logFC for NMD comparator isoforms
# Note: mashr logFC values are shrunken (empirical Bayes multivariate shrinkage),
# so magnitudes will be systematically smaller than limma logFC values.
de_dir <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/mashr"
de_date <- "2026.3.10"
ct_to_de_suffix <- c(AT = "at", DD = "dd", DD_ALI = "ddali",
                      DO = "doali", FB = "fb", MV = "mv")

# Collect NMD comparator isoforms with logFC and ptc_distance
nmd_logfc_dist <- do.call(rbind, lapply(setdiff(display_cts, "all_samples"), function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  if (!c2_key %in% names(set_results)) return(NULL)

  comp_ids <- set_results[[c2_key]]$profiles$comparator_isoform_id

  # Load DE results for this cell type
  de_suffix <- ct_to_de_suffix[ct]
  if (is.na(de_suffix)) return(NULL)
  de_file <- file.path(de_dir, sprintf("nmd_mashr_die_%s_%s.csv", de_suffix, de_date))
  if (!file.exists(de_file)) return(NULL)
  de <- read.csv(de_file)

  # Match comparator isoforms to DE logFC and PTC distance (mashr uses 'txid')
  de_sub <- de[de$txid %in% comp_ids, c("txid", "logFC")]
  ptc_sub <- ptc[ptc$isoform_id %in% comp_ids & !is.na(ptc$ptc_distance),
                  c("isoform_id", "ptc_distance", "has_ptc")]
  merged <- merge(de_sub, ptc_sub,
                   by.x = "txid", by.y = "isoform_id")
  if (nrow(merged) == 0) return(NULL)
  merged$cell_type <- ct
  merged
}))

if (!is.null(nmd_logfc_dist) && nrow(nmd_logfc_dist) > 0 && has_ggplot2) {
  # Scatter/hex plot: ptc_distance vs logFC
  p_logfc <- ggplot(nmd_logfc_dist, aes(x = ptc_distance, y = logFC)) +
    geom_point(alpha = 0.15, size = 0.8) +
    geom_smooth(method = "loess", se = TRUE, color = "steelblue") +
    geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
    facet_wrap(~cell_type, scales = "free") +
    scale_x_continuous(limits = c(
      quantile(nmd_logfc_dist$ptc_distance, 0.01, na.rm = TRUE),
      quantile(nmd_logfc_dist$ptc_distance, 0.99, na.rm = TRUE))) +
    theme_minimal() +
    labs(title = "NMD Response vs Stop Codon Distance to Last EJC",
         subtitle = "NMD comparator isoforms. Blue = loess smooth. Red dashed = 50 nt PTC threshold.",
         x = "Distance from stop codon to last EJC (nt)",
         y = "logFC (Smg1i vs DMSO)", fill = "")
  print(p_logfc)
  cat("\n\n")

  # Correlation summary
  cor_summary <- nmd_logfc_dist |>
    group_by(cell_type) |>
    summarize(
      n = n(),
      spearman_rho = round(cor(ptc_distance, logFC, method = "spearman",
                                use = "complete.obs"), 3),
      cor_p = cor.test(ptc_distance, logFC, method = "spearman")$p.value,
      .groups = "drop"
    )
  cat("**Columns:** **n** = NMD comparator isoforms with both logFC and stop-to-EJC distance; ")
  cat("**spearman_rho** = Spearman correlation between distance and logFC; ")
  cat("**cor_p** = p-value for the correlation.\n\n")
  knitr::kable(cor_summary,
               caption = "Correlation: Stop-to-EJC Distance vs logFC (NMD comparators)") |> print()
  cat("\n\n")
}
```

### PTC Position Metrics: Which Best Predicts NMD Response?

The loess curves above suggest a non-linear relationship between stop codon distance and
logFC. To better understand what drives NMD efficiency, we compare three ways of measuring
PTC position, using DD as a focal cell type:

- **PTC distance (nt)**: Raw nucleotide distance from stop codon to last EJC.
- **% CDS distance**: PTC distance normalized by ORF length (`ptc_distance / orf_length * 100`).
- **N downstream EJCs**: Number of exon-exon junctions downstream of the stop codon.

For each metric we fit a generalized additive model (GAM) to capture non-linear effects,
plus linear and quadratic models for comparison. All models are restricted to PTC+ isoforms
(distance > 50 nt) in the DD NMD comparison.

```{r sec3-ptc-position-metrics, fig.height=5, fig.width=10}
# Merge PTC features with logFC for DD NMD comparators
dd_c2_key <- "c2_dd"
dd_comp_ids <- set_results[[dd_c2_key]]$profiles$comparator_isoform_id

de_dd <- read.csv(file.path(de_dir, sprintf("nmd_mashr_die_dd_%s.csv", de_date)))
de_dd_sub <- de_dd[de_dd$txid %in% dd_comp_ids, c("txid", "logFC")]

ptc_dd <- ptc[ptc$isoform_id %in% dd_comp_ids, ]
dd_merged <- merge(de_dd_sub, ptc_dd, by.x = "txid", by.y = "isoform_id")
dd_merged <- dd_merged[!is.na(dd_merged$ptc_distance) & !is.na(dd_merged$orf_length) &
                         dd_merged$orf_length > 0, ]
dd_merged$pct_cds_distance <- 100 * dd_merged$ptc_distance / dd_merged$orf_length
dd_ptc_pos <- dd_merged[dd_merged$has_ptc == TRUE, ]
```

```{r sec3-ptc-panel-plot, fig.height=5, fig.width=12}
if (has_ggplot2 && nrow(dd_ptc_pos) > 0) {
  ptc_long <- rbind(
    data.frame(metric = "PTC distance (nt)", value = dd_ptc_pos$ptc_distance,
               logFC = dd_ptc_pos$logFC),
    data.frame(metric = "% CDS distance", value = dd_ptc_pos$pct_cds_distance,
               logFC = dd_ptc_pos$logFC),
    data.frame(metric = "N downstream EJCs", value = dd_ptc_pos$n_downstream_ejcs,
               logFC = dd_ptc_pos$logFC)
  )
  ptc_long$metric <- factor(ptc_long$metric,
    levels = c("PTC distance (nt)", "% CDS distance", "N downstream EJCs"))

  p_panel <- ggplot(ptc_long, aes(x = value, y = logFC)) +
    geom_point(alpha = 0.2, size = 0.8) +
    geom_smooth(method = "loess", span = 0.5, color = "steelblue", linewidth = 1.2) +
    facet_wrap(~metric, scales = "free_x") +
    theme_minimal() +
    labs(title = sprintf("DD NMD Comparators (PTC+, n = %d): logFC vs PTC Position Metrics",
                         nrow(dd_ptc_pos)),
         x = "Metric value", y = "logFC (Smg1i vs DMSO)")
  print(p_panel)
}
```

```{r sec3-ejc-boxplot, fig.height=5, fig.width=9}
if (has_ggplot2 && nrow(dd_ptc_pos) > 0) {
  ejc_counts <- dd_ptc_pos |>
    count(n_downstream_ejcs) |>
    mutate(label = paste0("n=", n))

  p_box <- ggplot(dd_ptc_pos, aes(x = factor(n_downstream_ejcs), y = logFC)) +
    geom_violin(fill = "lightblue", alpha = 0.5) +
    geom_boxplot(width = 0.15, outlier.alpha = 0.2) +
    geom_text(data = ejc_counts,
              aes(x = factor(n_downstream_ejcs),
                  y = max(dd_ptc_pos$logFC) + 0.3, label = label),
              size = 2.8) +
    theme_minimal() +
    labs(title = "DD NMD Comparators: logFC by Number of Downstream EJCs",
         subtitle = "logFC rises sharply from 1-4 EJCs, then plateaus",
         x = "Number of downstream EJCs", y = "logFC (Smg1i vs DMSO)")
  print(p_box)
}
```

```{r sec3-gam-comparison, results='asis'}
if (requireNamespace("mgcv", quietly = TRUE) && nrow(dd_ptc_pos) > 0) {
  n_ejc_unique <- length(unique(dd_ptc_pos$n_downstream_ejcs))

  # GAM models
  gam_dist <- mgcv::gam(logFC ~ s(ptc_distance), data = dd_ptc_pos)
  gam_pct  <- mgcv::gam(logFC ~ s(pct_cds_distance), data = dd_ptc_pos)
  gam_ejc  <- mgcv::gam(logFC ~ s(n_downstream_ejcs, k = min(n_ejc_unique, 10)),
                          data = dd_ptc_pos)
  gam_comb <- mgcv::gam(logFC ~ s(ptc_distance) +
                           s(n_downstream_ejcs, k = min(n_ejc_unique, 10)),
                          data = dd_ptc_pos)

  # Linear and quadratic models
  lm_dist_lin <- lm(logFC ~ ptc_distance, data = dd_ptc_pos)
  lm_pct_lin  <- lm(logFC ~ pct_cds_distance, data = dd_ptc_pos)
  lm_ejc_lin  <- lm(logFC ~ n_downstream_ejcs, data = dd_ptc_pos)
  lm_dist_q   <- lm(logFC ~ poly(ptc_distance, 2), data = dd_ptc_pos)
  lm_pct_q    <- lm(logFC ~ poly(pct_cds_distance, 2), data = dd_ptc_pos)
  lm_ejc_q    <- lm(logFC ~ poly(n_downstream_ejcs, 2), data = dd_ptc_pos)

  model_comparison <- data.frame(
    metric = c(rep("ptc_distance (nt)", 3), rep("pct_cds_distance (%)", 3),
               rep("n_downstream_ejcs", 3), "combined (dist + ejcs)"),
    model = c(rep(c("Linear", "Quadratic", "GAM"), 3), "GAM"),
    R_squared = round(c(
      summary(lm_dist_lin)$r.squared, summary(lm_dist_q)$r.squared,
        summary(gam_dist)$r.sq,
      summary(lm_pct_lin)$r.squared, summary(lm_pct_q)$r.squared,
        summary(gam_pct)$r.sq,
      summary(lm_ejc_lin)$r.squared, summary(lm_ejc_q)$r.squared,
        summary(gam_ejc)$r.sq,
      summary(gam_comb)$r.sq
    ), 3),
    AIC = round(c(
      AIC(lm_dist_lin), AIC(lm_dist_q), AIC(gam_dist),
      AIC(lm_pct_lin), AIC(lm_pct_q), AIC(gam_pct),
      AIC(lm_ejc_lin), AIC(lm_ejc_q), AIC(gam_ejc),
      AIC(gam_comb)
    ), 0)
  )
  knitr::kable(model_comparison,
               caption = sprintf("Model Comparison: PTC Position Metrics vs logFC (DD, n = %d PTC+ isoforms)",
                                  nrow(dd_ptc_pos))) |> print()
}
```

The number of downstream exon-exon junctions (`n_downstream_ejcs`) is the strongest single
predictor of NMD response magnitude, explaining ~15% of logFC variance in GAM models compared
to ~13% for raw nucleotide distance and ~8% for % CDS distance. Normalizing by CDS length
actually weakens the signal --- the absolute number of downstream junctions matters more than
the relative position within the transcript. The relationship is non-linear: logFC rises
sharply from 1 to ~4 downstream EJCs and then plateaus, consistent with EJC-based NMD
surveillance having diminishing returns beyond a threshold number of junctions. Quadratic
terms are highly significant for all three metrics, confirming the non-linear pattern.

## Frame-PTC Cross-Reference

PTC rates by frame category for both NMD and Control comparisons. This is the tabular form
of the PTC stacking in the Frame Category Distribution figure, showing how PTC prevalence
varies across frame categories and between comparison groups.

```{r sec3-frame-ptc-xref, results='asis'}
coding_cats <- c("same_start_in_frame", "same_start_frameshift",
                 "diff_start_same_frame", "diff_start_diff_frame",
                 "diff_start_diff_frame_with_frameshift")
ptc_pos_ids <- ptc$isoform_id[ptc$has_ptc == TRUE]

xref_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  do.call(rbind, lapply(c("c2", "c4"), function(comp) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) return(NULL)
    fc <- set_results[[key]]$frame_comp
    if (is.null(fc) || is.null(fc$pair_summary)) return(NULL)
    ps <- fc$pair_summary

    do.call(rbind, lapply(coding_cats, function(cat) {
      cat_pairs <- ps[ps$frame_category == cat, ]
      n_cat <- nrow(cat_pairs)
      n_ptc <- sum(cat_pairs$comparator_isoform_id %in% ptc_pos_ids)
      data.frame(cell_type = ct, comparison = comp_label(comp),
                 frame_category = cat, n_pairs = n_cat, n_ptc = n_ptc,
                 ptc_pct = round(100 * n_ptc / max(n_cat, 1), 1))
    }))
  }))
}))

if (!is.null(xref_rows)) {
  rownames(xref_rows) <- NULL
  knitr::kable(xref_rows, row.names = FALSE,
    caption = "PTC Rate by Frame Category: NMD vs Control") |> print()
  cat("\n\n")
}
```

## Frame Walk by Event Type {.tabset}

Breakdown of frameshift and compensatory rates by event type, comparing NMD and Control pairs.
Only events with CDS overlap are counted.

**Columns:** **event_type** = splicing event type; **n_cds_events** = number of events overlapping the CDS; **pct_frameshift** = % of those events that shift the reading frame (CDS overlap not divisible by 3); **pct_compensatory** = % that restore the frame after a prior shift. Suffixes **_NMD/_Ctrl** indicate the comparison group.

```{r sec3-frame-walk-by-event, results='asis'}
for (ct in display_cts) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results)) next
  cat(sprintf("### %s\n\n", ct))

  fw_tables <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) next
    fw <- set_results[[key]]$frame_walk
    if (is.null(fw) || nrow(fw$events) == 0L) next
    fw_evts <- fw$events
    fw_tables[[comp]] <- do.call(rbind, lapply(unique(fw_evts$event_type), function(et) {
      e <- fw_evts[fw_evts$event_type == et, ]
      data.frame(event_type = et,
                 n_cds_events = nrow(e),
                 pct_frameshift = round(100 * sum(e$is_frameshift) / nrow(e), 1),
                 pct_compensatory = round(100 * sum(e$is_compensatory) / nrow(e), 1))
    }))
  }

  if (length(fw_tables) == 2) {
    fw_merged <- merge(fw_tables[["c2"]], fw_tables[["c4"]],
                       by = "event_type", suffixes = c("_NMD", "_Ctrl"), all = TRUE)
    render_dt(fw_merged, caption = sprintf("Frame Walk by Event Type: NMD vs Control — %s", ct))
  } else if ("c2" %in% names(fw_tables)) {
    render_dt(fw_tables[["c2"]], caption = sprintf("Frame Walk by Event Type: NMD — %s", ct))
  }
  cat("\n\n")
}
```

# 4. Structural Differences

At baseline, non-NMD isoform pairs share ~50--58% of their exonic sequence and differ by a
median of ~700--800bp. IR and terminal events are the most prevalent event types, and the
next-best (comparator) isoform is typically shorter than the dominant (positive net_bp). This
represents the natural structural diversity between co-expressed isoforms from the same gene.

## Event Frequency {.tabset}

Prevalence of each splicing event type, measured as the percentage of isoform pairs containing
at least one instance of that event. NMD and Control are compared using Fisher's exact test per event type.

**Columns:** **event_type** = splicing event; **NMD_n/Ctrl_n** = number of pairs with the event; **NMD_pct/Ctrl_pct** = % of pairs; **OR** = odds ratio (>1 = more prevalent in NMD); **adj_p** = BH-adjusted p-value; **direction** = which group is enriched.

```{r sec4-event-freq, results='asis', fig.height=8}
for (ct in display_cts) {
  cps <- cross_results[[ct]]
  if (is.null(cps)) next
  ef <- cps$event_frequency
  if (is.null(ef) || isTRUE(ef$skipped)) next

  cat(sprintf("### %s\n\n", ct))

  ef_df <- do.call(rbind, lapply(ef$per_event, function(e) {
    dir_label <- gsub("enriched in A", "enriched in NMD",
                 gsub("enriched in B", "enriched in Control", e$direction))
    data.frame(
      event_type = e$event_type,
      NMD_n = e$count_a, NMD_pct = round(100 * e$count_a / e$n_a, 1),
      Ctrl_n = e$count_b, Ctrl_pct = round(100 * e$count_b / e$n_b, 1),
      OR = round(e$odds_ratio, 2),
      adj_p = e$adj_p_value,
      direction = dir_label
    )
  }))
  render_dt(ef_df, caption = sprintf("Event Type Prevalence: NMD vs Control — %s", ct))

  # Condensed one-liners for event count, profile categories, co-occurrence
  ec <- cps$event_count
  if (!is.null(ec) && !isTRUE(ec$skipped))
    cat(sprintf("Event count: Wilcoxon p = %s, d = %.3f (NMD median = %.0f, Control = %.0f)\n\n",
                format(ec$p_value, digits = 3), ec$cliffs_d, ec$median_a, ec$median_b))

  pc <- cps$profile_categories
  if (!is.null(pc) && !isTRUE(pc$skipped))
    cat(sprintf("Profile categories: Chi-square p = %s, V = %.3f\n\n",
                format(pc$p_value, digits = 3), pc$cramers_v))

  cooc_cps <- cps$cooccurrence
  if (!is.null(cooc_cps) && !isTRUE(cooc_cps$skipped))
    cat(sprintf("Co-occurrence structure is shared between NMD and Control (r = %.3f).\n\n",
                cooc_cps$correlation))

  # Paired tests (strongest within-gene evidence)
  pt <- cps$paired_tests
  if (!is.null(pt) && !isTRUE(pt$skipped)) {
    cat(sprintf("Paired analysis (%d gene-matched pairs): Wilcoxon p = %s; sign test: %.0f%% of genes have more events in NMD (p = %s)\n\n",
                pt$n_genes_overlap,
                format(pt$paired_wilcox$p_value, digits = 3),
                pt$sign_test$pct_a_more,
                format(pt$sign_test$p_value, digits = 3)))
  }
  cat("\n\n")
}
```

## Structural Divergence

Sequence similarity and exonic length differences between reference and comparator isoforms.
Higher pct_shared indicates more similar isoforms; larger exonic_diff indicates greater length disparity.

**Columns:** **NMD_med_pct_shared/Ctrl_med_pct_shared** = median % of exonic sequence shared between reference and comparator; **NMD_med_exonic_diff/Ctrl_med_exonic_diff** = median absolute exonic bp difference; **p_value** = Wilcoxon rank-sum test on pct_shared.

```{r sec4-divergence, results='asis'}
div_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  div_c2 <- set_results[[c2_key]]$divergence
  div_c4 <- set_results[[c4_key]]$divergence
  if (is.null(div_c2) || nrow(div_c2) == 0 || is.null(div_c4) || nrow(div_c4) == 0) return(NULL)
  wt <- tryCatch(wilcox.test(div_c2$pct_shared, div_c4$pct_shared), error = function(e) NULL)
  data.frame(
    cell_type = ct,
    NMD_med_pct_shared = round(median(div_c2$pct_shared, na.rm = TRUE), 1),
    Ctrl_med_pct_shared = round(median(div_c4$pct_shared, na.rm = TRUE), 1),
    NMD_med_exonic_diff = round(median(abs(div_c2$exonic_bp_diff)), 0),
    Ctrl_med_exonic_diff = round(median(abs(div_c4$exonic_bp_diff)), 0),
    p_value = if (!is.null(wt)) wt$p.value else NA_real_
  )
}))
if (!is.null(div_rows)) {
  knitr::kable(div_rows, caption = "Structural Divergence: NMD vs Control") |> print()
  cat("\n\n")
}
```

## Gain/Loss Balance

Net base pair change between reference and comparator isoforms. Computed as
`net_bp = bp_loss - bp_gain`, where LOSS and GAIN are from the comparator's perspective.
**Positive net_bp** means the comparator has lost more sequence than it gained (comparator is
shorter than the reference). **Negative net_bp** means the comparator has gained more sequence
(comparator is longer than the reference).

**Columns:** **NMD_med_net_bp/Ctrl_med_net_bp** = median net bp change; **wilcox_p** = Wilcoxon rank-sum test; **cliffs_d** = Cliff's delta effect size.

```{r sec4-gain-loss, results='asis'}
gl_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  cps <- cross_results[[ct]]
  if (is.null(cps)) return(NULL)
  gl <- cps$gain_loss
  if (is.null(gl) || isTRUE(gl$skipped)) return(NULL)
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  data.frame(
    cell_type = ct,
    NMD_med_net_bp = round(median(set_results[[c2_key]]$profiles$net_bp), 0),
    Ctrl_med_net_bp = round(median(set_results[[c4_key]]$profiles$net_bp), 0),
    wilcox_p = gl$p_value,
    cliffs_d = round(gl$cliffs_d, 3)
  )
}))
if (!is.null(gl_rows)) {
  knitr::kable(gl_rows, caption = "Net BP Change: NMD vs Control") |> print()
  cat("\n\n")
}
```

## Gain/Loss Direction by Event Type {.tabset}

For each event type, the direction indicates whether the comparator has GAINED or LOST that
feature relative to the reference. GAIN means the comparator has additional sequence (e.g., an
extra exon); LOSS means the comparator is missing sequence present in the reference.

**Columns:** **NMD_gain/NMD_loss** = event counts by direction in NMD pairs; **NMD_pct_gain** = % of events that are gains; **Ctrl_gain/Ctrl_loss/Ctrl_pct_gain** = same for Control; **fisher_p** = Fisher's exact test; **adj_p** = BH-adjusted p-value.

```{r sec4-gain-loss-direction, results='asis'}
for (ct in display_cts) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) next
  cat(sprintf("### %s\n\n", ct))

  events_c2 <- dplyr::bind_rows(set_results[[c2_key]]$profiles$detailed_events)
  events_c4 <- dplyr::bind_rows(set_results[[c4_key]]$profiles$detailed_events)

  if (nrow(events_c2) > 0 && nrow(events_c4) > 0) {
    dir_comp <- do.call(rbind, lapply(
      unique(c(events_c2$event_type, events_c4$event_type)), function(et) {
        e2 <- events_c2[events_c2$event_type == et, ]
        e4 <- events_c4[events_c4$event_type == et, ]
        g2 <- sum(e2$direction == "GAIN"); l2 <- sum(e2$direction == "LOSS")
        g4 <- sum(e4$direction == "GAIN"); l4 <- sum(e4$direction == "LOSS")
        ft <- if ((g2 + l2) > 0 && (g4 + l4) > 0)
          tryCatch(fisher.test(matrix(c(g2, l2, g4, l4), nrow = 2)), error = function(e) NULL) else NULL
        data.frame(
          event_type = et,
          NMD_gain = g2, NMD_loss = l2,
          NMD_pct_gain = round(100 * g2 / max(g2+l2, 1), 1),
          Ctrl_gain = g4, Ctrl_loss = l4,
          Ctrl_pct_gain = round(100 * g4 / max(g4+l4, 1), 1),
          fisher_p = if (!is.null(ft)) ft$p.value else NA_real_
        )
    }))
    dir_comp$adj_p <- p.adjust(dir_comp$fisher_p, "BH")
    render_dt(dir_comp, caption = sprintf("Gain/Loss Direction: NMD vs Control — %s", ct))
  }
  cat("\n\n")
}
```

## Intron Retention Analysis

Intron retention (IR) is the most prevalent event type at baseline, making it the dominant
structural difference between co-expressed isoforms from the same gene. IR events are detected
when one isoform has a single exon that spans the region covered by two or more exons (plus
the intervening intron) in the other isoform. Long-read sequencing may capture more IR than
short-read studies, since retained introns extend transcript length and are difficult to
assemble from short reads.

This section examines IR events (including IR, IR_diff, Partial_IR_5, Partial_IR_3) in
detail. **Direction reminder:** A GAIN means the comparator isoform has *retained* an intron
that is spliced out in the reference; a LOSS means the comparator has *lost* a retained intron
present in the reference (i.e., the comparator splices it out).

### IR Event Direction: Individual Types {.tabset}

GAIN/LOSS breakdown for each IR-family event type, stratified by NMD and Control.

```{r sec4-ir-individual, results='asis'}
# Define IR event types
ir_types <- c("IR", "IR_diff", "Partial_IR_5", "Partial_IR_3")

for (ct in display_cts) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) next
  cat(sprintf("#### %s\n\n", ct))

  events_c2 <- dplyr::bind_rows(set_results[[c2_key]]$profiles$detailed_events)
  events_c4 <- dplyr::bind_rows(set_results[[c4_key]]$profiles$detailed_events)

  ir_detail <- do.call(rbind, lapply(ir_types, function(et) {
    e2 <- events_c2[events_c2$event_type == et, ]
    e4 <- events_c4[events_c4$event_type == et, ]
    g2 <- sum(e2$direction == "GAIN"); l2 <- sum(e2$direction == "LOSS")
    g4 <- sum(e4$direction == "GAIN"); l4 <- sum(e4$direction == "LOSS")
    ft <- if ((g2 + l2) > 0 && (g4 + l4) > 0)
      tryCatch(fisher.test(matrix(c(g2, l2, g4, l4), nrow = 2)), error = function(e) NULL) else NULL
    data.frame(
      event_type = et,
      NMD_gain = g2, NMD_loss = l2, NMD_total = g2 + l2,
      NMD_pct_gain = round(100 * g2 / max(g2 + l2, 1), 1),
      Ctrl_gain = g4, Ctrl_loss = l4, Ctrl_total = g4 + l4,
      Ctrl_pct_gain = round(100 * g4 / max(g4 + l4, 1), 1),
      fisher_p = if (!is.null(ft)) ft$p.value else NA_real_
    )
  }))
  ir_detail$adj_p <- p.adjust(ir_detail$fisher_p, "BH")

  knitr::kable(ir_detail,
               caption = sprintf("IR Event Direction: NMD vs Control — %s", ct)) |> print()
  cat("\n\n")
}
```

### Cumulative IR Burden {.tabset}

Total IR-family events (IR + IR_diff + Partial_IR_5 + Partial_IR_3) aggregated across all
subtypes, comparing NMD and Control. This measures the overall intron retention burden
per pair and whether the direction (GAIN vs LOSS) differs between comparison groups.

```{r sec4-ir-cumulative, results='asis', fig.height=8}
ir_cumul_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)

  events_c2 <- dplyr::bind_rows(set_results[[c2_key]]$profiles$detailed_events)
  events_c4 <- dplyr::bind_rows(set_results[[c4_key]]$profiles$detailed_events)
  ir_c2 <- events_c2[events_c2$event_type %in% ir_types, ]
  ir_c4 <- events_c4[events_c4$event_type %in% ir_types, ]

  n_c2 <- nrow(set_results[[c2_key]]$profiles)
  n_c4 <- nrow(set_results[[c4_key]]$profiles)

  # Per-pair IR event counts
  ir_per_pair_c2 <- sapply(set_results[[c2_key]]$profiles$detailed_events, function(de)
    sum(de$event_type %in% ir_types))
  ir_per_pair_c4 <- sapply(set_results[[c4_key]]$profiles$detailed_events, function(de)
    sum(de$event_type %in% ir_types))

  wt <- tryCatch(wilcox.test(ir_per_pair_c2, ir_per_pair_c4), error = function(e) NULL)

  g2 <- sum(ir_c2$direction == "GAIN"); l2 <- sum(ir_c2$direction == "LOSS")
  g4 <- sum(ir_c4$direction == "GAIN"); l4 <- sum(ir_c4$direction == "LOSS")
  ft <- if ((g2 + l2) > 0 && (g4 + l4) > 0)
    tryCatch(fisher.test(matrix(c(g2, l2, g4, l4), nrow = 2)), error = function(e) NULL) else NULL

  data.frame(
    cell_type = ct,
    NMD_total_ir = nrow(ir_c2),
    NMD_per_pair = round(nrow(ir_c2) / n_c2, 2),
    NMD_pct_gain = round(100 * g2 / max(g2 + l2, 1), 1),
    Ctrl_total_ir = nrow(ir_c4),
    Ctrl_per_pair = round(nrow(ir_c4) / n_c4, 2),
    Ctrl_pct_gain = round(100 * g4 / max(g4 + l4, 1), 1),
    count_wilcox_p = if (!is.null(wt)) wt$p.value else NA_real_,
    dir_fisher_p = if (!is.null(ft)) ft$p.value else NA_real_
  )
}))

if (!is.null(ir_cumul_rows)) {
  cat("**Columns:** **NMD_total_ir/Ctrl_total_ir** = total IR-family events; ")
  cat("**NMD_per_pair/Ctrl_per_pair** = mean IR events per isoform pair; ")
  cat("**NMD_pct_gain/Ctrl_pct_gain** = % of IR events that are GAINs (comparator retains intron); ")
  cat("**count_wilcox_p** = Wilcoxon test on per-pair IR event counts (NMD vs Control); ")
  cat("**dir_fisher_p** = Fisher's exact test on GAIN vs LOSS proportions.\n\n")
  knitr::kable(ir_cumul_rows, caption = "Cumulative IR Burden: NMD vs Control") |> print()
  cat("\n\n")
}

```

## Distribution Plots

```{r sec4-dist-plots, results='asis', fig.height=8}
if (has_ggplot2) {
  # pct_shared from divergence
  pct_shared_all <- do.call(rbind, lapply(display_cts, function(ct) {
    ct_label <- ct_to_label(ct)
    rows <- list()
    for (comp in c("c2", "c4")) {
      key <- paste0(comp, "_", ct_label)
      if (!key %in% names(set_results)) next
      div <- set_results[[key]]$divergence
      if (is.null(div) || nrow(div) == 0) next
      rows[[comp]] <- data.frame(cell_type = ct, comparison = comp_label(comp),
                                 pct_shared = div$pct_shared)
    }
    do.call(rbind, rows)
  }))

  if (!is.null(pct_shared_all) && nrow(pct_shared_all) > 0) {
    p1 <- ggplot(pct_shared_all[!is.na(pct_shared_all$pct_shared), ],
                 aes(x = pct_shared, fill = comparison)) +
      geom_density(alpha = 0.5) +
      facet_wrap(~cell_type) +
      theme_minimal() +
      labs(title = "Sequence Shared (%): NMD vs Control",
           subtitle = "NMD shifted right = more similar (targeted changes)",
           x = "pct_shared", y = "Density", fill = "")
    print(p1)
    cat("\n\n")
  }

}
```

# 5. Cross-CT Consistency

## Event Frequency Heatmap

```{r sec6-heatmap, results='asis', fig.height=8}
if (length(main_per_ct) >= 2) {

  compute_freq_vec <- function(ct) {
    p <- profiles_list[[paste0("c2_", ct_to_label(ct))]]
    n <- nrow(p)
    n_pir5 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_5")))
    n_pir3 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_3")))
    data.frame(cell_type = ct, event_type = evt_types,
      pct = round(c(100*sum(p$tss_changed)/n, 100*sum(p$tes_changed)/n,
              100*sum(p$n_a5ss>0)/n, 100*sum(p$n_a3ss>0)/n,
              100*sum(p$n_se>0)/n, 100*sum(p$n_missing_internal>0)/n,
              100*sum(p$n_ir>0)/n, 100*sum(p$n_ir_diff>0)/n,
              100*n_pir5/n, 100*n_pir3/n), 1))
  }

  freq_matrix <- do.call(rbind, lapply(main_per_ct, compute_freq_vec))

  if (has_ggplot2) {
    p_heat <- ggplot(freq_matrix, aes(x = event_type, y = cell_type, fill = pct)) +
      geom_tile(color = "white") +
      geom_text(aes(label = sprintf("%.0f", pct)), size = 3) +
      scale_fill_gradient(low = "white", high = "steelblue", name = "% of pairs") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      labs(title = "NMD Event Frequency by Cell Type (pair-based %)",
           x = "Event Type", y = "Cell Type")
    print(p_heat)
    cat("\n\n")
  }

  freq_wide <- tidyr::pivot_wider(freq_matrix, names_from = event_type, values_from = pct)
  knitr::kable(freq_wide, caption = "NMD Event Frequency (% of pairs) by Cell Type") |> print()
  cat("\n\n")

} else {
  cat("Fewer than 2 main cell types with NMD profiles; skipping cross-cell-type analysis.\n\n")
}
```

## Heterogeneity Test

Tests whether the distribution of event types differs significantly across cell types.
A Pearson's chi-square test on the cell-type x event-type contingency table (raw pair
counts, not percentages). A significant result indicates that event type prevalence varies
across cell types beyond what would be expected by chance.

```{r sec6-heterogeneity, results='asis'}
if (length(main_per_ct) >= 2) {

  freq_count_list <- lapply(main_per_ct, function(ct) {
    p <- profiles_list[[paste0("c2_", ct_to_label(ct))]]
    n_pir5 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_5")))
    n_pir3 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_3")))
    c(sum(p$tss_changed), sum(p$tes_changed),
      sum(p$n_a5ss>0), sum(p$n_a3ss>0),
      sum(p$n_se>0), sum(p$n_missing_internal>0),
      sum(p$n_ir>0), sum(p$n_ir_diff>0),
      n_pir5, n_pir3)
  })
  freq_count_matrix <- do.call(rbind, freq_count_list)
  rownames(freq_count_matrix) <- main_per_ct
  colnames(freq_count_matrix) <- evt_types

  chi_main <- tryCatch(chisq.test(freq_count_matrix), error = function(e) NULL)
  if (!is.null(chi_main))
    cat(sprintf("**Main cell types** (AT, DD, DO, FB, MV): Chi-square = %.1f, df = %d, p = %s\n\n",
                chi_main$statistic, chi_main$parameter, format(chi_main$p.value, digits = 3)))

} else {
  cat("Fewer than 2 main cell types with NMD profiles; skipping heterogeneity test.\n\n")
}
```

# 6. Meta-Analysis

Random-effects meta-analysis pools NMD vs Control effect sizes across main cell types
(AT, DD, DO, FB, MV). DD_ALI is excluded due to reversed biological patterns that inflate
between-cell-type heterogeneity. all_samples is excluded as a superset.

## Pooled Effect Sizes Across Cell Types

Each metric from the NMD vs Control comparison is pooled across the five main cell types
using a random-effects model (metafor package, REML estimation). The table reports the
metrics discussed in earlier sections of this report, restricted to those with directional
interpretability.

**Effect size types and direction:**

- **Cliff's delta** (range -1 to +1): Nonparametric effect size for continuous comparisons.
  Positive values indicate NMD > Control; negative values indicate Control > NMD.
  - **event_count**: Positive = NMD pairs have more structural events per pair.
  - **gain_loss**: Positive = NMD pairs have higher net_bp (comparator shorter); negative = NMD comparator is longer.
- **Log odds ratio (log-OR)**: For binary outcomes. Positive = enriched in NMD; negative = enriched in Control.
  Exponentiate to get the odds ratio (e.g., log-OR of 2.3 = OR of ~10).
  - **ptc_rate**: Positive = PTC more frequent in NMD comparators.
  - **event_freq: {type}**: Positive = event type more prevalent in NMD pairs.

**Columns:** **metric** = analysis metric; **effect_size_type** = scale of the estimate;
**pooled_estimate** = meta-analytic effect size on the native scale (log-OR or Cliff's delta);
**OR** = exponentiated odds ratio for log-OR metrics (>1 = enriched in NMD; <1 = enriched in Control),
shown as "---" for Cliff's delta metrics;
**ci_lower/ci_upper** = 95% confidence interval on the native scale;
**p_value** = test of pooled effect = 0;
**I2** = heterogeneity (% of variance due to between-CT differences; >75% = high).

```{r sec6-meta, results='asis'}
if (requireNamespace("metafor", quietly = TRUE)) {
  # Use cached comparePairSets results — pool main CTs only (exclude DD_ALI + all_samples)
  cps_list <- cross_results[intersect(names(cross_results), main_cts)]
  cps_list <- cps_list[!sapply(cps_list, is.null)]

  if (length(cps_list) >= 2) {
    cat(sprintf("**Cell types pooled:** %s\n\n", paste(names(cps_list), collapse = ", ")))

    meta <- tryCatch(
      metaAnalyzePairSets(cps_list, verbose = FALSE),
      error = function(e) {
        cat(sprintf("Meta-analysis failed: %s\n\n", e$message))
        NULL
      }
    )

    if (!is.null(meta)) {
      # Only include metrics that are discussed in the report
      keep_metrics <- c("event_count", "gain_loss", "ptc_rate")
      scalar_sections <- intersect(
        names(meta)[vapply(meta, function(m) {
          !is.null(m$estimate) && !isTRUE(m$skipped)
        }, logical(1))],
        keep_metrics
      )

      meta_summary <- do.call(rbind, lapply(scalar_sections, function(mn) {
        m <- meta[[mn]]
        es_type <- switch(mn,
          event_count = "Cliff's delta",
          gain_loss = "Cliff's delta",
          ptc_rate = "log-OR",
          "---"
        )
        # For log-OR metrics, also show the exponentiated OR
        or_col <- if (es_type == "log-OR" && !is.null(m$pooled_or)) {
          sprintf("%.1f", m$pooled_or)
        } else { "---" }
        data.frame(
          metric = mn,
          effect_size_type = es_type,
          pooled_estimate = format(m$estimate, digits = 3),
          OR = or_col,
          ci_lower = format(m$ci_lower, digits = 3),
          ci_upper = format(m$ci_upper, digits = 3),
          p_value = format(m$p_value, digits = 3),
          I2 = sprintf("%.1f%%", m$I2)
        )
      }))

      # Add event_frequency per-event-type results (all log-OR)
      if (!is.null(meta$event_frequency) && is.list(meta$event_frequency)) {
        ef_rows <- do.call(rbind, lapply(names(meta$event_frequency), function(et) {
          m <- meta$event_frequency[[et]]
          if (isTRUE(m$skipped) || is.null(m$estimate)) return(NULL)
          or_col <- if (!is.null(m$pooled_or)) sprintf("%.2f", m$pooled_or) else
                      sprintf("%.2f", exp(m$estimate))
          data.frame(
            metric = paste0("event_freq: ", et),
            effect_size_type = "log-OR",
            pooled_estimate = format(m$estimate, digits = 3),
            OR = or_col,
            ci_lower = format(m$ci_lower, digits = 3),
            ci_upper = format(m$ci_upper, digits = 3),
            p_value = format(m$p_value, digits = 3),
            I2 = sprintf("%.1f%%", m$I2)
          )
        }))
        if (!is.null(ef_rows)) meta_summary <- rbind(meta_summary, ef_rows)
      }

      render_dt(meta_summary, caption = "Random-Effects Meta-Analysis (AT, DD, DO, FB, MV)")
    }
  } else {
    cat("Insufficient cell types for meta-analysis (need >= 2).\n\n")
  }
} else {
  cat("metafor package not available; skipping meta-analysis.\n\n")
}
```

# 7. DD_ALI

In AT, DD, DO, FB, and MV, NMD deviates from baseline in a consistent direction: more frameshifts,
more PTCs, more exon skipping, comparators longer, changes more targeted. DD_ALI reverses nearly
every one of these deviations --- its NMD transitions look more like an *amplification* of
baseline patterns than a departure from them.

Note: DD_ALI profiles shown here are from the **limma** classification (the original analysis),
since DD_ALI shows no meaningful PTC enrichment under any DE method and is excluded from all
main analyses. The comparison against main cell types uses mashr-classified results for the
main CTs.

```{r sec7-ddali-setup, include=FALSE}
ali_label <- ct_to_label(ali_ct)
c2_ali <- paste0("c2_", ali_label)
c4_ali <- paste0("c4_", ali_label)
has_ali <- c2_ali %in% names(set_results) && c4_ali %in% names(set_results)
```

## Key Metrics: DD_ALI vs Main Cell Types

```{r sec7-key-metrics, results='asis'}
if (has_ali) {

  range_str <- function(vals) {
    vals <- vals[!is.na(vals)]
    if (length(vals) == 0) return("\u2014")
    sprintf("%.0f\u2013%.0f", min(vals), max(vals))
  }

  ct_metrics <- lapply(main_per_ct, function(ct) {
    ct_label <- ct_to_label(ct)
    c2k <- paste0("c2_", ct_label); c4k <- paste0("c4_", ct_label)
    nmd_ptc <- set_results[[c2k]]$ptc_assoc
    ctrl_ptc <- set_results[[c4k]]$ptc_assoc
    nmd_div <- set_results[[c2k]]$divergence
    ctrl_div <- set_results[[c4k]]$divergence
    nmd_fw <- build_fw_row(c2k); ctrl_fw <- build_fw_row(c4k)
    list(
      nmd_ptc = if (!is.null(nmd_ptc)) 100 * nmd_ptc$comp_ptc_rate else NA,
      ctrl_ptc = if (!is.null(ctrl_ptc)) 100 * ctrl_ptc$comp_ptc_rate else NA,
      nmd_fw = if (!is.null(nmd_fw)) nmd_fw$pct_with_frameshift else NA,
      ctrl_fw = if (!is.null(ctrl_fw)) ctrl_fw$pct_with_frameshift else NA,
      nmd_net_bp = median(set_results[[c2k]]$profiles$net_bp),
      ctrl_net_bp = median(set_results[[c4k]]$profiles$net_bp),
      nmd_pct_shared = if (!is.null(nmd_div) && nrow(nmd_div) > 0)
        median(nmd_div$pct_shared, na.rm = TRUE) else NA,
      ctrl_pct_shared = if (!is.null(ctrl_div) && nrow(ctrl_div) > 0)
        median(ctrl_div$pct_shared, na.rm = TRUE) else NA
    )
  })

  ali_ptc <- set_results[[c2_ali]]$ptc_assoc
  ali_div <- set_results[[c2_ali]]$divergence
  ali_fw <- build_fw_row(c2_ali)
  ali_ptc_val <- if (!is.null(ali_ptc)) round(100 * ali_ptc$comp_ptc_rate, 1) else NA
  ali_fw_val <- if (!is.null(ali_fw)) ali_fw$pct_with_frameshift else NA
  ali_net_bp_val <- round(median(set_results[[c2_ali]]$profiles$net_bp), 0)
  ali_pct_shared_val <- if (!is.null(ali_div) && nrow(ali_div) > 0)
    round(median(ali_div$pct_shared, na.rm = TRUE), 1) else NA

  ddali_summary <- data.frame(
    metric = c("comp_ptc_%", "frameshift_%", "med_net_bp", "med_pct_shared"),
    Baseline_Ctrl = c(
      range_str(sapply(ct_metrics, `[[`, "ctrl_ptc")),
      range_str(sapply(ct_metrics, `[[`, "ctrl_fw")),
      range_str(sapply(ct_metrics, `[[`, "ctrl_net_bp")),
      range_str(sapply(ct_metrics, `[[`, "ctrl_pct_shared"))
    ),
    Main_CTs_NMD = c(
      range_str(sapply(ct_metrics, `[[`, "nmd_ptc")),
      range_str(sapply(ct_metrics, `[[`, "nmd_fw")),
      range_str(sapply(ct_metrics, `[[`, "nmd_net_bp")),
      range_str(sapply(ct_metrics, `[[`, "nmd_pct_shared"))
    ),
    DD_ALI_NMD = c(
      if (!is.na(ali_ptc_val)) sprintf("%.1f", ali_ptc_val) else "\u2014",
      if (!is.na(ali_fw_val)) sprintf("%.1f", ali_fw_val) else "\u2014",
      sprintf("%.0f", ali_net_bp_val),
      if (!is.na(ali_pct_shared_val)) sprintf("%.1f", ali_pct_shared_val) else "\u2014"
    ),
    DD_ALI_direction = c(
      "\u2248 baseline",
      "\u2248 baseline",
      "Amplified baseline",
      "Reversed"
    )
  )
  render_dt(ddali_summary, caption = "DD_ALI Reversal: Key Metrics")
  cat("\n\n")

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

## PTC & Frameshift

```{r sec7-ptc-frameshift, results='asis'}
if (has_ali) {

  ali_pa_c2 <- set_results[[c2_ali]]$ptc_assoc
  ali_pa_c4 <- set_results[[c4_ali]]$ptc_assoc
  ali_cps <- cross_results[[ali_ct]]
  ali_ptc_cps <- if (!is.null(ali_cps)) ali_cps$ptc_rate else NULL

  # Need ali_fw for this chunk too
  ali_fw <- build_fw_row(c2_ali)

  ptc_row <- data.frame(
    metric = c("Comparator PTC %", "Frameshift %", "Unresolved %"),
    NMD = c(
      round(100 * ali_pa_c2$comp_ptc_rate, 1),
      if (!is.null(ali_fw)) ali_fw$pct_with_frameshift else NA,
      if (!is.null(ali_fw)) ali_fw$pct_unresolved else NA
    ),
    Control = c(
      round(100 * ali_pa_c4$comp_ptc_rate, 1),
      {fw4 <- build_fw_row(c4_ali); if (!is.null(fw4)) fw4$pct_with_frameshift else NA},
      {if (!is.null(fw4)) fw4$pct_unresolved else NA}
    )
  )
  knitr::kable(ptc_row, caption = "DD_ALI: PTC & Frameshift Rates") |> print()
  cat("\n")

  if (!is.null(ali_ptc_cps) && !isTRUE(ali_ptc_cps$skipped))
    cat(sprintf("PTC Fisher's exact: OR = %.2f, p = %s\n\n",
                ali_ptc_cps$odds_ratio, format(ali_ptc_cps$p_value, digits = 3)))

  # Stop-to-EJC distance for DD_ALI comparators
  ali_comp_c2 <- set_results[[c2_ali]]$profiles$comparator_isoform_id
  ali_comp_c4 <- set_results[[c4_ali]]$profiles$comparator_isoform_id
  ali_ptc_c2 <- ptc[ptc$isoform_id %in% ali_comp_c2 & !is.na(ptc$ptc_distance), ]
  ali_ptc_c4 <- ptc[ptc$isoform_id %in% ali_comp_c4 & !is.na(ptc$ptc_distance), ]
  if (nrow(ali_ptc_c2) > 0 && nrow(ali_ptc_c4) > 0) {
    wt <- wilcox.test(ali_ptc_c2$ptc_distance, ali_ptc_c4$ptc_distance)
    cat(sprintf("Stop-to-EJC distance (all comparators): NMD median = %.0f nt, Control = %.0f nt, Wilcoxon p = %s\n\n",
                median(ali_ptc_c2$ptc_distance), median(ali_ptc_c4$ptc_distance),
                format(wt$p.value, digits = 3)))
  }

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

## Event Frequency

```{r sec7-event-freq, results='asis'}
if (has_ali) {

  ali_cps <- cross_results[[ali_ct]]
  ali_ef <- if (!is.null(ali_cps)) ali_cps$event_frequency else NULL
  if (!is.null(ali_ef) && !isTRUE(ali_ef$skipped)) {
    ef_df <- do.call(rbind, lapply(ali_ef$per_event, function(e) {
      dir_label <- gsub("enriched in A", "enriched in NMD",
                   gsub("enriched in B", "enriched in Control", e$direction))
      data.frame(
        event_type = e$event_type,
        NMD_n = e$count_a, NMD_pct = round(100 * e$count_a / e$n_a, 1),
        Ctrl_n = e$count_b, Ctrl_pct = round(100 * e$count_b / e$n_b, 1),
        OR = round(e$odds_ratio, 2),
        adj_p = e$adj_p_value,
        direction = dir_label
      )
    }))
    render_dt(ef_df, caption = "DD_ALI: Event Type Prevalence NMD vs Control")
    cat("\n")
  }

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

In main cell types, SE is enriched in NMD. In DD_ALI, SE is enriched in *Control*,
consistent with a fundamentally different NMD mechanism.

## Structural Divergence & Gain/Loss

```{r sec7-divergence, results='asis'}
if (has_ali) {

  ali_div_c2 <- set_results[[c2_ali]]$divergence
  ali_div_c4 <- set_results[[c4_ali]]$divergence
  if (!is.null(ali_div_c2) && nrow(ali_div_c2) > 0 &&
      !is.null(ali_div_c4) && nrow(ali_div_c4) > 0) {
    div_row <- data.frame(
      metric = c("med_pct_shared", "med_|exonic_diff|", "med_net_bp"),
      NMD = c(
        round(median(ali_div_c2$pct_shared, na.rm = TRUE), 1),
        round(median(abs(ali_div_c2$exonic_bp_diff)), 0),
        round(median(set_results[[c2_ali]]$profiles$net_bp), 0)
      ),
      Control = c(
        round(median(ali_div_c4$pct_shared, na.rm = TRUE), 1),
        round(median(abs(ali_div_c4$exonic_bp_diff)), 0),
        round(median(set_results[[c4_ali]]$profiles$net_bp), 0)
      )
    )
    knitr::kable(div_row, caption = "DD_ALI: Structural Divergence") |> print()
    cat("\n")
  }

  ali_cps <- cross_results[[ali_ct]]
  ali_gl <- if (!is.null(ali_cps)) ali_cps$gain_loss else NULL
  if (!is.null(ali_gl) && !isTRUE(ali_gl$skipped))
    cat(sprintf("Gain/loss: Wilcoxon p = %s, d = %.3f\n\n",
                format(ali_gl$p_value, digits = 3), ali_gl$cliffs_d))

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

In main cell types, NMD pairs share MORE sequence than baseline (~70% vs ~55%). DD_ALI
reverses this: NMD pairs share LESS sequence (~28%). DD_ALI NMD comparators are also
much longer than their reference (large positive net_bp), amplifying the baseline pattern
rather than reversing it.

## Gene Overlap with Main CTs

```{r sec7-gene-overlap, results='asis'}
if (has_ali) {

  ali_genes <- unique(profiles_list[[c2_ali]]$gene_id)
  main_gene_lists <- lapply(main_per_ct, function(ct)
    unique(profiles_list[[paste0("c2_", ct_to_label(ct))]]$gene_id))
  names(main_gene_lists) <- main_per_ct

  ali_overlap <- data.frame(
    cell_type = main_per_ct,
    ddali_genes = length(ali_genes),
    shared = sapply(main_per_ct, function(ct)
      length(intersect(ali_genes, main_gene_lists[[ct]]))),
    pct_of_ddali = sapply(main_per_ct, function(ct)
      round(100 * length(intersect(ali_genes, main_gene_lists[[ct]])) / length(ali_genes), 1))
  )
  render_dt(ali_overlap, caption = "DD_ALI Gene Overlap with Main CTs")
  cat("\n\n")

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

DD_ALI NMD-sensitive isoforms resemble highly divergent baseline isoforms (low sequence
sharing, high retention) rather than the targeted exon-skipping pattern seen in other cell
types. This suggests DD_ALI NMD targets arise through terminal/retention mechanisms rather
than frameshift-driven PTC introduction, possibly reflecting ALI culture conditions.

# 8. Key Figures


```{r sec9-figures, eval=has_ggplot2, fig.height=8}
if (has_ggplot2 && length(profiles_list) > 0) {

  # --- Figure 1: Event frequency by comparison type ---
  event_freq <- do.call(rbind, lapply(names(profiles_list), function(key) {
    p <- profiles_list[[key]]
    parts <- strsplit(key, "_")[[1]]
    s <- summarizeProfiles(p)
    data.frame(
      comparison = comp_label(parts[1]),
      cell_type = paste(parts[-1], collapse = "_"),
      event_type = names(s$event_counts),
      count = as.integer(s$event_counts),
      total = s$total_events
    )
  }))
  event_freq$pct <- 100 * event_freq$count / event_freq$total

  p1 <- ggplot(event_freq, aes(x = event_type, y = pct, fill = comparison)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~cell_type, scales = "free_y") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "Event Type Distribution by Comparison",
         subtitle = "Raw event counts (not pair-based percentages — see Section 4 tables)",
         x = "Event Type", y = "Percentage of Total Events",
         fill = "Comparison")
  print(p1)

}
```

# 9. Session Info

```{r session-info}
sessionInfo()
```
