---
title: "NMD Isoform Transitions Analysis"
subtitle: "Isopair Wrapper — Isocall Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(Isopair)
library(dplyr)
library(tibble)
library(tidyr)
library(DT)
library(htmltools)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
if (has_ggplot2) library(ggplot2)
source("visualization_functions.R")

# DT rendering helper for results='asis' chunks
# Uses renderTags() + knit_meta_add() to register JS/CSS dependencies with knitr,
# so widgets render even when they are not the last expression in a chunk.
render_dt <- function(df, caption = NULL, ...) {
  widget <- datatable(df, caption = caption, rownames = FALSE,
                      class = "compact stripe",
                      options = list(pageLength = 25, dom = 'ft'), ...)
  rendered <- htmltools::renderTags(widget)
  knitr::knit_meta_add(rendered$dependencies)
  cat(rendered$html)
  cat("\n")
}

# Caching infrastructure: wrap expensive computations with cached_compute()
cached_compute <- function(key, expr, cache_dir = file.path(data_dir, "analysis_cache"),
                           force_recompute = FALSE) {
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
  cache_file <- file.path(cache_dir, paste0(key, ".rds"))
  if (!force_recompute && file.exists(cache_file)) {
    return(readRDS(cache_file))
  }
  result <- force(expr)
  saveRDS(result, cache_file)
  result
}

# Utility functions
ct_to_label <- function(ct) tolower(gsub("_", "", ct))

evt_types <- c("Alt_TSS", "Alt_TES", "A5SS", "A3SS", "SE",
               "Missing_Internal", "IR", "IR_diff",
               "Partial_IR_5", "Partial_IR_3")

compute_event_freq <- function(profiles) {
  n_pairs <- nrow(profiles)
  n_pir5 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_5")))
  n_pir3 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_3")))
  counts <- c(sum(profiles$tss_changed), sum(profiles$tes_changed),
              sum(profiles$n_a5ss > 0), sum(profiles$n_a3ss > 0),
              sum(profiles$n_se > 0), sum(profiles$n_missing_internal > 0),
              sum(profiles$n_ir > 0), sum(profiles$n_ir_diff > 0),
              n_pir5, n_pir3)
  data.frame(event_type = evt_types,
             n_pairs_with_event = counts,
             pct_of_pairs = round(100 * counts / n_pairs, 1))
}

# Display labels: internal c2/c4 → user-facing NMD/Control
comp_label <- function(comp) {
  switch(toupper(comp), C2 = "NMD", C4 = "Control", toupper(comp))
}

parse_key <- function(key) {
  parts <- strsplit(key, "_")[[1]]
  list(comp = comp_label(parts[1]),
       ct = paste(parts[-1], collapse = "_"))
}

# Count CDS bp in the comparator isoform upstream (5') of the first frameshift boundary.
# Used to determine how much of the protein is conserved vs novel after a frameshift.
count_conserved_cds_bp <- function(comp_id, fs_boundary, strand, cds_df, str_df) {
  cds_row <- cds_df[cds_df$isoform_id == comp_id, ]
  str_row <- str_df[str_df$isoform_id == comp_id, ]
  if (nrow(cds_row) == 0 || nrow(str_row) == 0) return(NA_integer_)
  cds_start <- cds_row$cds_start[1]
  cds_stop  <- cds_row$cds_stop[1]
  orf_len   <- cds_row$orf_length[1]
  e_starts  <- str_row$exon_starts[[1]]
  e_ends    <- str_row$exon_ends[[1]]

  # Order exons 5'→3'
  if (strand == "-") {
    ord <- order(e_starts, decreasing = TRUE)
  } else {
    ord <- order(e_starts)
  }
  e_starts <- e_starts[ord]
  e_ends   <- e_ends[ord]

  conserved <- 0L
  for (i in seq_along(e_starts)) {
    es <- e_starts[i]
    ee <- e_ends[i]
    # Intersect exon with CDS range
    cs <- max(es, cds_start)
    ce <- min(ee, cds_stop)
    if (cs > ce) next  # No CDS overlap

    if (strand == "+") {
      if (fs_boundary <= cs) break
      if (fs_boundary >= ce) {
        conserved <- conserved + (ce - cs + 1L)
      } else {
        conserved <- conserved + (fs_boundary - cs)
        break
      }
    } else {
      if (fs_boundary >= ce) break
      if (fs_boundary <= cs) {
        conserved <- conserved + (ce - cs + 1L)
      } else {
        conserved <- conserved + (ce - fs_boundary)
        break
      }
    }
  }
  conserved <- min(conserved, orf_len)
  as.integer(conserved)
}

# Build frame walk summary row for any profile key
# Restrict frame walk summary to pairs with coding references (frame_resolved not NA)
build_fw_row <- function(key) {
  fw <- set_results[[key]]$frame_walk
  if (is.null(fw) || is.null(fw$profile_summary) || nrow(fw$profile_summary) == 0) return(NULL)
  ps <- fw$profile_summary
  # Restrict to coding pairs: frame_resolved is NA when reference is non-coding
  ps <- ps[!is.na(ps$frame_resolved), ]
  if (nrow(ps) == 0) return(NULL)
  n <- nrow(ps)
  has_shift <- ps$n_frameshifts > 0
  n_shift <- sum(has_shift, na.rm = TRUE)
  n_unres <- sum(has_shift & ps$frame_resolved == FALSE, na.rm = TRUE)
  n_comp <- sum(ps$n_compensatory > 0, na.rm = TRUE)
  tibble::tibble(
    n_pairs = n,
    n_with_frameshift = n_shift,
    pct_with_frameshift = round(100 * n_shift / n, 1),
    n_unresolved = n_unres,
    pct_unresolved = round(100 * n_unres / n, 1),
    n_compensated = n_comp,
    pct_compensated = if (n_shift > 0) round(100 * n_comp / n_shift, 1) else 0,
    median_pct_orf_frameshifted = round(
      median(ps$pct_orf_frameshifted[ps$n_frameshifts > 0], na.rm = TRUE), 1)
  )
}
```

# 1. Setup

This report analyzes structural differences between isoform pairs using the Isopair package.
Main analyses cover all_samples (union NMD / intersection non-NMD) plus AT, DD, FB, and MV.
DD_ALI is analyzed separately (Section 6) due to reversed biological patterns.

```{r load-data}
data_dir <- "data"

# Infrastructure
structures     <- readRDS(file.path(data_dir, "structures.rds"))
union_exons    <- readRDS(file.path(data_dir, "union_exons.rds"))
ue_mapping     <- readRDS(file.path(data_dir, "isoform_union_mapping.rds"))
cds            <- readRDS(file.path(data_dir, "cds.rds"))
annotated_ue   <- readRDS(file.path(data_dir, "annotated_ue.rds"))
ptc            <- readRDS(file.path(data_dir, "ptc.rds"))

# Study data
expr_mat       <- readRDS(file.path(data_dir, "expression_data.rds"))
sample_meta    <- readRDS(file.path(data_dir, "sample_metadata.rds"))
gene_map       <- readRDS(file.path(data_dir, "gene_map.rds"))
nmd_class      <- readRDS(file.path(data_dir, "nmd_classification.rds"))

# Cell types with profiles
cell_types <- c("all_samples", "AT", "DD", "DD_ALI", "FB", "MV")
comparisons <- c("c2", "c4")

# DD_ALI separation: main cell types for cross-CT analyses and meta-analysis
main_cts <- c("AT", "DD", "FB", "MV")
ali_ct <- "DD_ALI"

# Load all available profiles
profiles_list <- list()
pairs_list    <- list()
for (ct in cell_types) {
  ct_label <- tolower(gsub("_", "", ct))
  for (comp in comparisons) {
    key <- paste(comp, ct_label, sep = "_")
    prof_file <- file.path(data_dir, sprintf("profiles_%s.rds", key))
    pair_file <- file.path(data_dir, sprintf("pairs_%s.rds", key))
    if (file.exists(prof_file)) {
      profiles_list[[key]] <- readRDS(prof_file)
      if (file.exists(pair_file)) pairs_list[[key]] <- readRDS(pair_file)
    }
  }
}

cat(sprintf("Loaded %d profile sets across %d cell types\n",
            length(profiles_list), length(cell_types)))
```

```{r gene-match}
# Bidirectional gene-matching: restrict NMD and Control to the intersection of
# (gene_id, reference_isoform_id) pairs present in BOTH sets. After this step,
# every gene in NMD has a corresponding Control pair and vice versa.
gene_match_summary <- list()
for (ct in cell_types) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(profiles_list) || !c4_key %in% names(profiles_list)) next

  nmd_before <- nrow(profiles_list[[c2_key]])
  control_before <- nrow(profiles_list[[c4_key]])

  # Intersection of (gene_id, reference_isoform_id)
  c2_gene_refs <- unique(profiles_list[[c2_key]][, c("gene_id", "reference_isoform_id")])
  c4_gene_refs <- unique(profiles_list[[c4_key]][, c("gene_id", "reference_isoform_id")])
  shared_refs <- inner_join(c2_gene_refs, c4_gene_refs, by = c("gene_id", "reference_isoform_id"))

  profiles_list[[c2_key]] <- semi_join(profiles_list[[c2_key]], shared_refs,
                                        by = c("gene_id", "reference_isoform_id"))
  profiles_list[[c4_key]] <- semi_join(profiles_list[[c4_key]], shared_refs,
                                        by = c("gene_id", "reference_isoform_id"))

  gene_match_summary[[ct]] <- data.frame(
    cell_type = ct,
    nmd_before = nmd_before,
    control_before = control_before,
    matched_genes = nrow(shared_refs),
    nmd_after = nrow(profiles_list[[c2_key]]),
    control_after = nrow(profiles_list[[c4_key]])
  )
}
gene_match_df <- do.call(rbind, gene_match_summary)
```

# 2. Data Flow

```{r data-flow-numbers}
# --- Gather numbers ---
n_isoforms <- nrow(expr_mat)
n_samples <- ncol(expr_mat)
n_dmso <- sum(sample_meta$treatment == "DMSO")
n_smg1i <- sum(sample_meta$treatment == "Smg1i")
n_genes_struct <- n_distinct(structures$gene_id)
coding_n <- sum(cds$coding_status == "coding")
unknown_n <- sum(cds$coding_status == "unknown")

nmd_summary <- do.call(rbind, lapply(names(nmd_class), function(ct) {
  data.frame(cell_type = ct,
             n_nmd = length(nmd_class[[ct]]$nmd),
             n_non_nmd = length(nmd_class[[ct]]$non_nmd))
}))
total_nmd <- length(unique(unlist(lapply(nmd_class, `[[`, "nmd"))))
total_non_nmd <- length(unique(unlist(lapply(nmd_class, `[[`, "non_nmd"))))

total_nmd_before <- sum(gene_match_df$nmd_before)
total_ctrl_before <- sum(gene_match_df$control_before)
total_nmd_after <- sum(gene_match_df$nmd_after)
total_ctrl_after <- sum(gene_match_df$control_after)
total_matched_genes <- sum(gene_match_df$matched_genes)
```

```{r data-flow-chart}
# --- Flowchart (Graphviz via DiagrammeR) ---
fmt <- function(x) format(x, big.mark = ",")

dot_spec <- sprintf('
digraph flowchart {
  graph [rankdir=TB, fontname="Helvetica", nodesep=1.6, ranksep=1.4, size="13,18"]
  node  [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=23, margin="0.65,0.5", width=6.5]
  edge  [color="#666666", penwidth=2.5]

  A [label=<<b>Isoform Expression</b><br/>%s isoforms × %d samples<br/>%d DMSO + %d Smg1i | 6 cell types>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  B [label=<<b>NMD Classification</b><br/>NMD-sensitive: %s isoforms<br/>Non-NMD: %s isoforms>,
     fillcolor="#fff7bc", color="#d95f0e"]

  C [label=<<b>Pair Construction</b><br/>%s NMD pairs | %s Control pairs>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  D [label=<<b>Gene-Matching</b><br/>%s shared genes<br/>%s NMD | %s Control pairs>,
     fillcolor="#c7e9c0", color="#31a354"]

  E [label=<<b>NMD</b><br/>%s pairs>,
     fillcolor="#fee0d2", color="#de2d26", width=4.5]

  F [label=<<b>Control</b><br/>%s pairs>,
     fillcolor="#deebf7", color="#3182bd", width=4.5]

  G [label=<<b>Gene-Matched Comparison</b><br/>%s genes | %d cell types>,
     fillcolor="#c7e9c0", color="#31a354"]

  A -> B -> C -> D
  D -> E
  D -> F
  {rank=same; E; F}
  E -> G
  F -> G
}',
  fmt(n_isoforms), n_samples, n_dmso, n_smg1i,
  fmt(total_nmd), fmt(total_non_nmd),
  fmt(total_nmd_before), fmt(total_ctrl_before),
  fmt(total_matched_genes), fmt(total_nmd_after), fmt(total_ctrl_after),
  fmt(total_nmd_after),
  fmt(total_ctrl_after),
  fmt(total_matched_genes), nrow(gene_match_df)
)

if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(dot_spec, width = 1000, height = 900)
} else {
  message("DiagrammeR not installed — flowchart not rendered")
}
```

## NMD Classification

Isoforms are classified per cell type using limma-voom differential expression results.
NMD-sensitive: adj.P.Val < 0.05 & logFC > 0 (upregulated by SMG1 inhibition, i.e., rescued from NMD decay). Non-NMD: adj.P.Val > 0.50 (no evidence of NMD regulation).

**Columns:** **cell_type** = cell type used for DE testing; **n_nmd** = number of isoforms classified as NMD-sensitive in that cell type; **n_non_nmd** = number of isoforms classified as non-NMD.

```{r nmd-classification-table, results='asis'}
render_dt(nmd_summary, caption = "NMD Classification Summary")
```

## Gene-Matching Detail

To enable paired NMD vs Control comparisons, we restrict both sets to genes where the same
reference isoform appears in both NMD and Control pairs. This ensures that any differences
between NMD and Control comparators are not confounded by different reference isoforms or gene sets.

**Columns:** **nmd_before/control_before** = pair counts before matching; **matched_genes** = genes with shared (gene_id, reference_isoform_id) in both sets; **nmd_after/control_after** = pair counts after restricting to matched genes.

```{r gene-matching-table, results='asis'}
render_dt(gene_match_df, caption = "Gene-Matching Summary (bidirectional, per cell type)")
```

## Profile Sets (after gene-matching)

```{r profile-sets-compute}
prof_overview <- do.call(rbind, lapply(names(profiles_list), function(key) {
  p <- profiles_list[[key]]
  parts <- strsplit(key, "_")[[1]]
  data.frame(comparison = comp_label(parts[1]),
             cell_type = paste(parts[-1], collapse = "_"),
             n_profiles = nrow(p),
             n_genes = n_distinct(p$gene_id),
             total_events = sum(p$n_events),
             mean_events = round(mean(p$n_events), 2),
             pct_recon_pass = round(100 * mean(p$reconstruction_status == "PASS",
                                                na.rm = TRUE), 1))
}))
```

Each profile set represents isoform pairs analyzed for structural differences using Isopair.
NMD pairs compare an NMD-sensitive comparator to a non-NMD reference; Control pairs compare
the next-best non-NMD isoform to the same reference.

**Columns:** **comparison** = NMD or Control; **cell_type** = cell type (or all_samples for union/intersection); **n_profiles** = number of isoform pairs; **n_genes** = unique genes represented; **total_events** = sum of structural differences across all pairs; **mean_events** = average structural differences per pair; **pct_recon_pass** = % of pairs where the comparator isoform can be perfectly reconstructed from the reference plus detected events (quality control metric; 100% expected).

```{r profile-sets-table, results='asis'}
render_dt(prof_overview, caption = "Profile Set Summary (gene-matched)")
```

## Coding Status

NMD is a translation-dependent pathway: only protein-coding transcripts with an open reading
frame can harbour a premature termination codon. We therefore expect NMD-sensitive isoform pairs
to be overwhelmingly coding. This table characterizes the coding status of reference and
comparator isoforms in each profile set, based on SQANTI CDS annotations. A small fraction of
isoforms have unknown coding status (no CDS annotation in SQANTI); these are excluded from
downstream analyses that require reading-frame information (e.g., frameshift analysis).

```{r coding-status}
coding_tbl <- do.call(rbind, lapply(names(profiles_list), function(key) {
  p <- profiles_list[[key]]
  parts <- strsplit(key, "_")[[1]]
  comp <- comp_label(parts[1])
  ct <- paste(parts[-1], collapse = "_")

  ref_ids <- unique(p$reference_isoform_id)
  cmp_ids <- unique(p$comparator_isoform_id)

  ref_coding <- sum(ref_ids %in% cds$isoform_id[cds$coding_status == "coding"])
  cmp_coding <- sum(cmp_ids %in% cds$isoform_id[cds$coding_status == "coding"])

  # Pairs where both are coding
  ref_status <- cds$coding_status[match(p$reference_isoform_id, cds$isoform_id)]
  cmp_status <- cds$coding_status[match(p$comparator_isoform_id, cds$isoform_id)]
  both_coding <- sum(ref_status == "coding" & cmp_status == "coding", na.rm = TRUE)

  data.frame(
    Comparison = comp,
    Cell_Type = ct,
    N_Pairs = nrow(p),
    Ref_Coding = ref_coding,
    Ref_Pct = round(100 * ref_coding / length(ref_ids), 1),
    Comp_Coding = cmp_coding,
    Comp_Pct = round(100 * cmp_coding / length(cmp_ids), 1),
    Both_Coding = both_coding,
    Both_Pct = round(100 * both_coding / nrow(p), 1)
  )
}))
```

**Columns:** **Comparison** = NMD or Control; **Cell_Type** = cell type; **N_Pairs** = total
isoform pairs; **Ref_Coding / Ref_Pct** = number and percentage of unique reference isoforms
with a CDS annotation; **Comp_Coding / Comp_Pct** = same for comparator isoforms;
**Both_Coding / Both_Pct** = pairs where both reference and comparator are coding.

```{r coding-status-table, results='asis'}
render_dt(coding_tbl, caption = "Coding Status of Reference and Comparator Isoforms")
```

```{r compute-all, include=FALSE}
# =============================================================================
# Silent compute chunk: all cached_compute() calls + comparePairSets
# No output — results stored in set_results / cross_results for display sections
# =============================================================================
set_results <- list()
cross_results <- list()

for (key in names(profiles_list)) {
  profiles <- profiles_list[[key]]
  message(sprintf("Computing analyses for %s...", key))

  prof_summary <- summarizeProfiles(profiles)
  divergence <- cached_compute(paste0("div_", key),
    tryCatch(quantifyPairDivergence(profiles, structures, cds), error = function(e) NULL))
  regional_impact <- cached_compute(paste0("ri_", key),
    tryCatch(quantifyRegionalImpact(profiles, annotated_ue, cds), error = function(e) NULL))
  cooc <- cached_compute(paste0("cooc_", key),
    tryCatch(testCooccurrence(profiles), error = function(e) NULL))
  event_regions <- cached_compute(paste0("er_", key),
    tryCatch(mapEventsToRegions(profiles, annotated_ue, cds), error = function(e) NULL))
  ptc_assoc <- cached_compute(paste0("ptc_", key),
    tryCatch(testPtcAssociation(ptc, profiles), error = function(e) NULL))
  frame_walk <- cached_compute(paste0("fw_", key),
    tryCatch(analyzeFrameWalk(profiles, cds, ptc), error = function(e) NULL))

  # Dependent computations (not cached separately — fast from cached event_regions)
  orf_bound <- NULL; orf_impact <- NULL
  if (!is.null(event_regions) && nrow(event_regions) > 0) {
    orf_bound <- tryCatch(testOrfBoundarySusceptibility(event_regions, annotated_ue),
                          error = function(e) NULL)
    orf_impact <- tryCatch(summarizeOrfImpact(event_regions, cds),
                           error = function(e) NULL)
  }

  set_results[[key]] <- list(
    profiles = profiles, prof_summary = prof_summary, divergence = divergence,
    regional_impact = regional_impact,
    cooc = cooc, event_regions = event_regions, orf_bound = orf_bound,
    orf_impact = orf_impact, ptc_assoc = ptc_assoc, frame_walk = frame_walk,
    evt_freq = compute_event_freq(profiles)
  )
}

# comparePairSets per cell type (cached, reused by display sections + meta-analysis)
for (ct in cell_types) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(profiles_list) || !c4_key %in% names(profiles_list)) next

  message(sprintf("Computing NMD vs Control comparison for %s...", ct))
  ptc_c2 <- ptc[ptc$isoform_id %in% profiles_list[[c2_key]]$comparator_isoform_id, ]
  ptc_c4 <- ptc[ptc$isoform_id %in% profiles_list[[c4_key]]$comparator_isoform_id, ]

  cross_results[[ct]] <- cached_compute(
    paste0("cps_", ct_label),
    tryCatch(
      comparePairSets(profiles_list[[c2_key]], profiles_list[[c4_key]],
                       ptc_a = ptc_c2, ptc_b = ptc_c4,
                       paired = TRUE, verbose = FALSE),
      error = function(e) {
        message(sprintf("comparePairSets failed for %s: %s", ct, e$message))
        NULL
      }
    )
  )
}

# Cell types available for display (DD_ALI excluded — dedicated section below)
display_cts <- setdiff(
  cell_types[sapply(cell_types, function(ct) {
    ct_label <- ct_to_label(ct)
    any(paste0(c("c2_", "c4_"), ct_label) %in% names(set_results))
  })],
  ali_ct
)
per_ct <- setdiff(display_cts, "all_samples")
c2_per_ct <- per_ct[sapply(per_ct, function(ct)
  paste0("c2_", ct_to_label(ct)) %in% names(set_results))]

# Main cell types for cross-CT analyses (excluding DD_ALI outlier)
main_per_ct <- intersect(c2_per_ct, main_cts)
ali_in_c2 <- ali_ct %in% c2_per_ct

message("All computations complete.")
```


# 3. PTC & Reading Frame

PTC enrichment (OR 12--19) is the single strongest signal distinguishing NMD-sensitive isoform
pairs from baseline structural variation. At baseline, ~5% of non-NMD comparator isoforms
contain a PTC and ~30% of pairs involve a frameshift --- the background rate of frame disruption
in normal isoform variation. NMD departs dramatically.

## PTC Rates

Premature termination codons (PTCs) are stop codons located >50 nucleotides upstream of the
last exon-exon junction, a position that triggers nonsense-mediated decay. This table compares
PTC prevalence in NMD comparator isoforms vs Control comparator isoforms.

**Columns:** **n_pairs** = gene-matched isoform pairs; **NMD_comp_ptc_pct** = % of NMD comparator isoforms containing a PTC; **Ctrl_comp_ptc_pct** = % of Control comparators with a PTC; **fisher_OR** = odds ratio from Fisher's exact test (>1 = enriched in NMD); **p_value** = Fisher's exact test p-value.

```{r sec3-ptc-rates, results='asis'}
ptc_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  pa_c2 <- set_results[[c2_key]]$ptc_assoc
  pa_c4 <- set_results[[c4_key]]$ptc_assoc
  cps <- cross_results[[ct]]
  ptc_cps <- if (!is.null(cps)) cps$ptc_rate else NULL
  if (is.null(pa_c2) || is.null(pa_c4)) return(NULL)
  data.frame(
    cell_type = ct,
    n_pairs = nrow(set_results[[c2_key]]$profiles),
    NMD_comp_ptc_pct = round(100 * pa_c2$comp_ptc_rate, 1),
    Ctrl_comp_ptc_pct = round(100 * pa_c4$comp_ptc_rate, 1),
    fisher_OR = if (!is.null(ptc_cps) && !isTRUE(ptc_cps$skipped)) round(ptc_cps$odds_ratio, 2) else NA_real_,
    p_value = if (!is.null(ptc_cps) && !isTRUE(ptc_cps$skipped)) ptc_cps$p_value else NA_real_
  )
}))
if (!is.null(ptc_rows)) {
  knitr::kable(ptc_rows, caption = "PTC Rates: NMD vs Control Comparators") |> print()
  cat("\n\n")
}
```

## Frameshift Summary

A **frameshift** is detected when a splicing event alters the CDS by a number of base pairs
not divisible by 3, shifting the downstream reading frame. Only events that can affect the CDS
are considered (SE, Missing_Internal, A5SS, A3SS, Partial_IR_5, Partial_IR_3); IR and terminal
events are excluded. Isopair walks through events in 5'-to-3' order, tracking the cumulative
reading frame offset.

A frameshift is **resolved** if subsequent events restore the reading frame to its original
phase (cumulative offset returns to 0 by the end of the transcript). An **unresolved** frameshift
means the reading frame remains shifted through the remainder of the CDS, typically leading to
a premature termination codon. A **compensatory** event is one that restores the frame after a
prior shift (e.g., a second non-divisible-by-3 event that cancels the first).

**Columns:** **NMD_shift_pct/Ctrl_shift_pct** = % of coding pairs with at least one frameshift; **NMD_unres_pct/Ctrl_unres_pct** = % of coding pairs with unresolved frameshifts (frame remains shifted); **fisher_OR** = odds ratio for frameshift presence (NMD vs Control, >1 = enriched in NMD); **p_value** = Fisher's exact test.

```{r sec3-frameshift, results='asis'}
fw_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  fw_c2 <- build_fw_row(c2_key)
  fw_c4 <- build_fw_row(c4_key)
  if (is.null(fw_c2) || is.null(fw_c4)) return(NULL)

  # Fisher's exact on frameshift presence: shift vs no-shift, NMD vs Control
  ps_c2 <- set_results[[c2_key]]$frame_walk$profile_summary
  ps_c4 <- set_results[[c4_key]]$frame_walk$profile_summary
  ps_c2 <- ps_c2[!is.na(ps_c2$frame_resolved), ]
  ps_c4 <- ps_c4[!is.na(ps_c4$frame_resolved), ]
  n_shift_c2 <- sum(ps_c2$n_frameshifts > 0, na.rm = TRUE)
  n_noshift_c2 <- sum(ps_c2$n_frameshifts == 0, na.rm = TRUE)
  n_shift_c4 <- sum(ps_c4$n_frameshifts > 0, na.rm = TRUE)
  n_noshift_c4 <- sum(ps_c4$n_frameshifts == 0, na.rm = TRUE)
  fi <- tryCatch(fisher.test(matrix(c(n_shift_c2, n_noshift_c2, n_shift_c4, n_noshift_c4), nrow = 2)),
                 error = function(e) NULL)
  data.frame(
    cell_type = ct,
    NMD_shift_pct = fw_c2$pct_with_frameshift,
    NMD_unres_pct = fw_c2$pct_unresolved,
    Ctrl_shift_pct = fw_c4$pct_with_frameshift,
    Ctrl_unres_pct = fw_c4$pct_unresolved,
    fisher_OR = if (!is.null(fi)) round(fi$estimate, 2) else NA_real_,
    p_value = if (!is.null(fi)) fi$p.value else NA_real_
  )
}))
if (!is.null(fw_rows)) {
  knitr::kable(fw_rows, caption = "Frameshift Rates: NMD vs Control") |> print()
  cat("\n\n")
}
```

## Frameshift Consequences

Frameshifts are common in both NMD and Control comparisons (~50-60% of coding pairs), but
PTC enrichment in NMD is massive (OR 15-26). The gap is explained by what happens *after*
a frameshift: most Control frameshifts produce a new stop codon in the last exon
(0 downstream EJCs), so the transcript escapes NMD surveillance. These represent stable
proteins with rewritten C-terminal sequences — a form of productive alternative splicing
generating protein diversity. The funnel below traces pairs from coding status through
frameshift and PTC outcomes.

### Frameshift → PTC Funnel

**Columns:** **n_coding** = pairs where reference is coding; **n_frameshift** = pairs with ≥1 frameshift event; **n_unresolved** = frameshifts not compensated by a downstream event; **n_ptc** = unresolved with PTC (stop >50 nt upstream of last EJC); **n_no_ptc** = unresolved without PTC; **n_ptc_unknown** = unresolved with unknown CDS (no PTC call); **n_stop_last_exon** = among no-PTC unresolved, stop codon in last exon; **n_res_ptc** = resolved frameshifts with PTC; **n_nofs_ptc** = no-frameshift pairs with PTC.

```{r sec3-frameshift-funnel, results='asis'}
funnel_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  rows <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) next
    fw <- set_results[[key]]$frame_walk
    if (is.null(fw) || is.null(fw$profile_summary)) next
    ps <- fw$profile_summary
    ps <- ps[!is.na(ps$frame_resolved), ]
    if (nrow(ps) == 0) next

    # Merge with ptc to get PTC status for comparators
    ps_ptc <- merge(ps, ptc[, c("isoform_id", "has_ptc", "n_downstream_ejcs", "stop_in_last_exon")],
                    by.x = "comparator_isoform_id", by.y = "isoform_id", all.x = TRUE)

    n_coding     <- nrow(ps_ptc)
    has_fs       <- ps_ptc$n_frameshifts > 0
    n_frameshift <- sum(has_fs)
    n_no_fs      <- n_coding - n_frameshift

    # Unresolved = frameshift + frame NOT restored
    unresolved   <- has_fs & ps_ptc$frame_resolved == FALSE
    n_unresolved <- sum(unresolved)

    # Among unresolved: PTC breakdown (exclude unknown CDS)
    unres_ptc_known   <- unresolved & !is.na(ps_ptc$has_ptc)
    n_ptc_unknown     <- sum(unresolved & is.na(ps_ptc$has_ptc))
    n_unres_ptc       <- sum(unres_ptc_known & ps_ptc$has_ptc == TRUE)
    n_unres_no_ptc    <- sum(unres_ptc_known & ps_ptc$has_ptc == FALSE)

    # Among no-PTC unresolved: stop in last exon
    no_ptc_mask       <- unres_ptc_known & ps_ptc$has_ptc == FALSE
    n_stop_last_exon  <- sum(no_ptc_mask & ps_ptc$stop_in_last_exon == TRUE, na.rm = TRUE)

    # Resolved frameshifts with PTC
    resolved <- has_fs & ps_ptc$frame_resolved == TRUE
    n_res_ptc <- sum(resolved & ps_ptc$has_ptc == TRUE, na.rm = TRUE)

    # No-frameshift pairs with PTC
    n_nofs_ptc <- sum(!has_fs & ps_ptc$has_ptc == TRUE, na.rm = TRUE)

    rows[[comp]] <- data.frame(
      cell_type = ct, comparison = comp_label(comp),
      n_coding = n_coding, n_frameshift = n_frameshift,
      n_unresolved = n_unresolved,
      n_ptc = n_unres_ptc, n_no_ptc = n_unres_no_ptc,
      n_ptc_unknown = n_ptc_unknown,
      n_stop_last_exon = n_stop_last_exon,
      n_res_ptc = n_res_ptc, n_nofs_ptc = n_nofs_ptc
    )
  }
  do.call(rbind, rows)
}))

if (!is.null(funnel_rows)) {
  # Verification: funnel arithmetic
  check <- with(funnel_rows, all(n_ptc + n_no_ptc + n_ptc_unknown == n_unresolved))
  if (!check) cat("**WARNING:** Funnel arithmetic check failed (n_ptc + n_no_ptc + n_ptc_unknown != n_unresolved)\n\n")

  knitr::kable(funnel_rows, caption = "Frameshift → PTC Funnel: NMD vs Control") |> print()
  cat("\n\n")
}
```

In NMD comparisons, unresolved frameshifts overwhelmingly produce PTCs that trigger decay.
In Control comparisons, most unresolved frameshifts place the new stop codon in the last exon
(0 downstream EJCs), producing a stable mRNA with a novel C-terminal protein sequence.

### Stable Novel Proteins (Control)

Control (C4) pairs with unresolved frameshifts and no PTC represent productive alternative
splicing: the frameshift rewrites the C-terminal protein sequence without triggering NMD.
The analysis below quantifies how much of each protein is novel — i.e., translated in a
non-original reading frame downstream of the first frameshift boundary.

```{r sec3-novel-protein}
# Compute novel protein metrics for C4 unresolved frameshifts without PTC
novel_protein_results <- list()

for (ct in display_cts) {
  if (ct == "all_samples") next
  ct_label <- ct_to_label(ct)
  c4_key <- paste0("c4_", ct_label)
  if (!c4_key %in% names(set_results)) next
  fw <- set_results[[c4_key]]$frame_walk
  if (is.null(fw)) next

  novel_protein_results[[ct]] <- cached_compute(paste0("novel_protein_c4_", ct_label), {
    ps <- fw$profile_summary
    ps <- ps[!is.na(ps$frame_resolved), ]

    # Unresolved frameshifts
    ps_unres <- ps[ps$n_frameshifts > 0 & ps$frame_resolved == FALSE, ]
    if (nrow(ps_unres) == 0) return(NULL)

    # Merge with ptc: keep only no-PTC comparators (coding, no PTC)
    ps_merged <- merge(ps_unres,
                       ptc[, c("isoform_id", "has_ptc", "orf_length")],
                       by.x = "comparator_isoform_id", by.y = "isoform_id", all.x = TRUE)
    ps_merged <- ps_merged[!is.na(ps_merged$has_ptc) & ps_merged$has_ptc == FALSE, ]
    if (nrow(ps_merged) == 0) return(NULL)

    # Get first frameshift event per pair from events table
    evts <- fw$events
    evts_fs <- evts[evts$is_frameshift == TRUE, ]
    if (nrow(evts_fs) == 0) return(NULL)

    # First frameshift per pair (events are in 5'→3' order)
    first_fs <- evts_fs |>
      group_by(gene_id, reference_isoform_id, comparator_isoform_id) |>
      slice(1) |>
      ungroup()

    # Merge with our target pairs
    ps_merged <- merge(ps_merged, first_fs[, c("gene_id", "reference_isoform_id",
                       "comparator_isoform_id", "genomic_start", "genomic_end")],
                       by = c("gene_id", "reference_isoform_id", "comparator_isoform_id"))
    if (nrow(ps_merged) == 0) return(NULL)

    # Get strand from structures (or cds)
    strand_lookup <- setNames(cds$strand, cds$isoform_id)

    # Compute conserved/novel bp per pair
    results <- lapply(seq_len(nrow(ps_merged)), function(i) {
      row <- ps_merged[i, ]
      comp_id <- row$comparator_isoform_id
      s <- strand_lookup[comp_id]
      if (is.na(s)) return(NULL)

      # Frameshift boundary: plus → genomic_end, minus → genomic_start
      fs_boundary <- if (s == "+") row$genomic_end else row$genomic_start

      conserved_bp <- count_conserved_cds_bp(comp_id, fs_boundary, s, cds, structures)
      if (is.na(conserved_bp)) return(NULL)
      comp_orf <- row$orf_length
      novel_bp <- max(0L, comp_orf - conserved_bp)
      data.frame(
        gene_id = row$gene_id,
        comparator_isoform_id = comp_id,
        orf_length = comp_orf,
        conserved_bp = conserved_bp,
        novel_bp = novel_bp,
        novel_aa = novel_bp / 3,
        total_aa = comp_orf / 3,
        pct_novel = 100 * (novel_bp / 3) / (comp_orf / 3)
      )
    })
    do.call(rbind, results)
  })
}
```

**Columns:** **n_stable_novel** = C4 unresolved frameshifts without PTC (coding comparators only); **pct_of_coding** = as % of all C4 coding pairs; **median_novel_aa / iqr_novel_aa** = novel amino acids (translated in shifted frame); **median_total_aa** = total protein length; **median_pct_novel / iqr_pct_novel** = % of protein that is novel sequence.

```{r sec3-novel-protein-table, results='asis'}
novel_summary <- do.call(rbind, lapply(names(novel_protein_results), function(ct) {
  np <- novel_protein_results[[ct]]
  if (is.null(np) || nrow(np) == 0) return(NULL)
  ct_label <- ct_to_label(ct)
  c4_key <- paste0("c4_", ct_label)
  fw <- set_results[[c4_key]]$frame_walk
  n_coding <- sum(!is.na(fw$profile_summary$frame_resolved))
  data.frame(
    cell_type = ct,
    n_stable_novel = nrow(np),
    pct_of_coding = round(100 * nrow(np) / n_coding, 1),
    median_novel_aa = round(median(np$novel_aa), 0),
    iqr_novel_aa = paste0(round(quantile(np$novel_aa, 0.25), 0), "-",
                          round(quantile(np$novel_aa, 0.75), 0)),
    median_total_aa = round(median(np$total_aa), 0),
    median_pct_novel = round(median(np$pct_novel), 1),
    iqr_pct_novel = paste0(round(quantile(np$pct_novel, 0.25), 1), "-",
                           round(quantile(np$pct_novel, 0.75), 1))
  )
}))

if (!is.null(novel_summary)) {
  knitr::kable(novel_summary, caption = "Stable Novel Proteins: Control (C4) Unresolved Frameshifts Without PTC") |> print()
  cat("\n\n")
}
```

```{r sec3-novel-protein-plot, fig.height=5}
if (has_ggplot2) {
  novel_plot_data <- do.call(rbind, lapply(names(novel_protein_results), function(ct) {
    np <- novel_protein_results[[ct]]
    if (is.null(np) || nrow(np) == 0) return(NULL)
    data.frame(cell_type = ct, pct_novel = np$pct_novel)
  }))

  if (!is.null(novel_plot_data) && nrow(novel_plot_data) > 0) {
    p_novel <- ggplot(novel_plot_data, aes(x = pct_novel)) +
      geom_density(fill = "steelblue", alpha = 0.5) +
      facet_wrap(~cell_type, scales = "free_y") +
      theme_minimal() +
      labs(title = "Novel Protein Sequence in Stable Control Frameshifts",
           subtitle = "C4 unresolved frameshifts without PTC — % of protein in shifted reading frame",
           x = "% of protein that is novel sequence", y = "Density")
    print(p_novel)
  }
}
```

### Frameshift Fate Flowchart

The flowcharts below trace the fate of all coding pairs through the frameshift → PTC
funnel, using the all_samples (union NMD / intersection non-NMD) data. Each box shows
the count at that stage.

```{r sec3-frameshift-flowchart-nmd, fig.width=12, fig.height=10}
if (has_ggplot2 && !is.null(funnel_rows)) {
  nmd_row  <- funnel_rows[funnel_rows$cell_type == "all_samples" &
                           funnel_rows$comparison == "NMD", ]
  ctrl_row <- funnel_rows[funnel_rows$cell_type == "all_samples" &
                           funnel_rows$comparison == "Control", ]

  build_funnel_stages <- function(r, side_label) {
    n_no_fs    <- r$n_coding - r$n_frameshift
    n_resolved <- r$n_frameshift - r$n_unresolved
    bw <- 3.0  # box width
    bh <- 1.0  # box height
    data.frame(
      label = c(
        paste0(side_label, "\nCoding Pairs"),
        "No Frameshift",
        "Has Frameshift",
        "Resolved\n(Compensated)",
        "Unresolved",
        "PTC",
        "No PTC",
        "Unknown CDS",
        "Stop in\nLast Exon"
      ),
      n = c(r$n_coding, n_no_fs, r$n_frameshift, n_resolved,
            r$n_unresolved, r$n_ptc, r$n_no_ptc, r$n_ptc_unknown,
            r$n_stop_last_exon),
      x = c(0, -3, 3, -1, 4, 1, 4, 7, 4),
      y = c(10, 7.5, 7.5, 5, 5, 2.5, 2.5, 2.5, 0),
      width  = bw,
      height = bh,
      fill = c("#D5E8D4", "#DAE8FC", "#DAE8FC", "#FFF2CC",
               "#FFF2CC", "#F8CECC", "#D5E8D4", "#E6E6E6", "#D5E8D4"),
      stringsAsFactors = FALSE
    )
  }

  build_funnel_arrows <- function(side_label) {
    top <- paste0(side_label, "\nCoding Pairs")
    data.frame(
      from_label = c(top, top,
                     "Has Frameshift", "Has Frameshift",
                     "Unresolved", "Unresolved", "Unresolved",
                     "No PTC"),
      to_label = c("No Frameshift", "Has Frameshift",
                   "Resolved\n(Compensated)", "Unresolved",
                   "PTC", "No PTC", "Unknown CDS",
                   "Stop in\nLast Exon"),
      stringsAsFactors = FALSE
    )
  }

  nmd_stages <- build_funnel_stages(nmd_row, "NMD")
  nmd_arrows <- build_funnel_arrows("NMD")
  print(plot_funnel(nmd_stages, nmd_arrows, title = "NMD (C2): Frameshift Fate — all_samples"))
}
```

```{r sec3-frameshift-flowchart-ctrl, fig.width=12, fig.height=10}
if (has_ggplot2 && !is.null(funnel_rows) && nrow(ctrl_row) > 0) {
  ctrl_stages <- build_funnel_stages(ctrl_row, "Control")
  ctrl_arrows <- build_funnel_arrows("Control")
  print(plot_funnel(ctrl_stages, ctrl_arrows, title = "Control (C4): Frameshift Fate — all_samples"))
}
```

### Frameshift Examples: Real Gene Cases

Three real gene examples illustrating the major categories of frameshift outcomes.
Panel A: a frameshift that produces a PTC, triggering NMD. Panel B: a pair with
compensatory frameshift events that restore the reading frame. Panel C: a frameshift
without PTC, producing a stable protein with novel C-terminal sequence.

```{r sec3-frameshift-examples, fig.width=12, fig.height=10}
if (has_ggplot2) {
  # Use DD cell type as representative (largest dataset)
  ex_ct <- "dd"
  c2_key <- paste0("c2_", ex_ct)
  c4_key <- paste0("c4_", ex_ct)

  fw_c2 <- set_results[[c2_key]]$frame_walk
  fw_c4 <- set_results[[c4_key]]$frame_walk

  example_plots <- list()

  # --- Panel A: Frameshift → PTC (NMD) ---
  # Find a pair with 1 frameshift event, similar exon count between ref/comp
  if (!is.null(fw_c2)) {
    ps_c2 <- fw_c2$profile_summary
    ps_c2 <- ps_c2[!is.na(ps_c2$frame_resolved) & ps_c2$n_frameshifts == 1 &
                    ps_c2$frame_resolved == FALSE & ps_c2$n_cds_events == 1, ]
    ps_c2_ptc <- merge(ps_c2, ptc[ptc$has_ptc == TRUE, c("isoform_id", "has_ptc",
                       "ptc_distance", "n_downstream_ejcs")],
                       by.x = "comparator_isoform_id", by.y = "isoform_id")
    # Get exon counts for both ref and comp to find structurally similar pairs
    ps_c2_ptc <- merge(ps_c2_ptc,
                       structures[, c("isoform_id", "n_exons")],
                       by.x = "comparator_isoform_id", by.y = "isoform_id")
    names(ps_c2_ptc)[names(ps_c2_ptc) == "n_exons"] <- "comp_n_exons"
    ps_c2_ptc <- merge(ps_c2_ptc,
                       structures[, c("isoform_id", "n_exons")],
                       by.x = "reference_isoform_id", by.y = "isoform_id")
    names(ps_c2_ptc)[names(ps_c2_ptc) == "n_exons"] <- "ref_n_exons"
    ps_c2_ptc$exon_diff <- abs(ps_c2_ptc$ref_n_exons - ps_c2_ptc$comp_n_exons)
    # Prefer: same exon count (diff <= 1), moderate exon number, clear PTC signal
    candidates <- ps_c2_ptc[ps_c2_ptc$comp_n_exons >= 3 & ps_c2_ptc$comp_n_exons <= 10 &
                             ps_c2_ptc$exon_diff <= 1, ]
    if (nrow(candidates) > 0) {
      # Among similar pairs, pick one with clear PTC distance
      pick_a <- candidates[which.max(candidates$ptc_distance), ]
      ref_ex <- get_exon_coords(pick_a$reference_isoform_id, structures)
      comp_ex <- get_exon_coords(pick_a$comparator_isoform_id, structures)
      ref_cds_a <- get_cds_range(pick_a$reference_isoform_id, cds)
      comp_cds_a <- get_cds_range(pick_a$comparator_isoform_id, cds)
      str_row <- structures[structures$isoform_id == pick_a$reference_isoform_id, ]
      gene_strand <- str_row$strand[1]

      # Get events for this pair
      profs <- set_results[[c2_key]]$profiles
      pair_prof <- profs[profs$reference_isoform_id == pick_a$reference_isoform_id &
                         profs$comparator_isoform_id == pick_a$comparator_isoform_id, ]
      pair_events <- if (nrow(pair_prof) > 0) pair_prof$detailed_events[[1]] else NULL

      # Build PTC annotation — approximate PTC position from comparator CDS stop
      ptc_ann <- list()
      if (!is.null(comp_cds_a)) {
        ptc_pos <- if (gene_strand == "+") comp_cds_a$cds_stop else comp_cds_a$cds_start
        ptc_ann <- list(
          list(type = "ptc_marker", x = ptc_pos, track = "comp",
               label = "PTC", color = "#E74C3C")
        )
      }

      example_plots[["A"]] <- plot_ucsc_pair(
        ref_ex$starts, ref_ex$ends,
        comp_ex$starts, comp_ex$ends,
        gene_strand,
        ref_cds = ref_cds_a, comp_cds = comp_cds_a,
        ref_label = "Dominant\n(non-NMD)",
        comp_label = "NMD-sensitive",
        events = pair_events,
        annotations = ptc_ann,
        title = paste0("A) Frameshift \u2192 PTC (", pick_a$gene_id, ")"),
        subtitle = sprintf("Unresolved frameshift, PTC distance = %d nt, %d downstream EJC(s)",
                           round(pick_a$ptc_distance), pick_a$n_downstream_ejcs)
      )
    }
  }

  # --- Panel B: Compensatory Frameshift (resolved) ---
  # Search across all cell types and comparisons to find a good example
  pick_b <- NULL
  pick_b_key <- NULL
  for (search_key in names(set_results)) {
    fw_search <- set_results[[search_key]]$frame_walk
    if (is.null(fw_search) || is.null(fw_search$profile_summary)) next
    ps_comp <- fw_search$profile_summary
    ps_comp <- ps_comp[!is.na(ps_comp$frame_resolved) & ps_comp$n_frameshifts >= 2 &
                       ps_comp$frame_resolved == TRUE, ]
    if (nrow(ps_comp) == 0) next
    ps_comp <- merge(ps_comp,
                     structures[, c("isoform_id", "n_exons")],
                     by.x = "comparator_isoform_id", by.y = "isoform_id")
    candidates_b <- ps_comp[ps_comp$n_exons >= 3 & ps_comp$n_exons <= 10, ]
    if (nrow(candidates_b) > 0) {
      # Prefer n_frameshifts == 2 for clarity
      prefer <- candidates_b[candidates_b$n_frameshifts == 2, ]
      if (nrow(prefer) > 0) candidates_b <- prefer
      pick_b <- candidates_b[1, ]
      pick_b_key <- search_key
      break
    }
  }
  if (!is.null(pick_b)) {
    ref_ex_b <- get_exon_coords(pick_b$reference_isoform_id, structures)
    comp_ex_b <- get_exon_coords(pick_b$comparator_isoform_id, structures)
    ref_cds_b <- get_cds_range(pick_b$reference_isoform_id, cds)
    comp_cds_b <- get_cds_range(pick_b$comparator_isoform_id, cds)
    str_row_b <- structures[structures$isoform_id == pick_b$reference_isoform_id, ]
    gene_strand_b <- str_row_b$strand[1]

    profs_b <- set_results[[pick_b_key]]$profiles
    pair_prof_b <- profs_b[profs_b$reference_isoform_id == pick_b$reference_isoform_id &
                           profs_b$comparator_isoform_id == pick_b$comparator_isoform_id, ]
    pair_events_b <- if (nrow(pair_prof_b) > 0) pair_prof_b$detailed_events[[1]] else NULL

    parsed_b <- parse_key(pick_b_key)
    comp_label_b <- if (parsed_b$comp == "NMD") "NMD-sensitive" else "Next-best\n(non-NMD)"

    example_plots[["B"]] <- plot_ucsc_pair(
      ref_ex_b$starts, ref_ex_b$ends,
      comp_ex_b$starts, comp_ex_b$ends,
      gene_strand_b,
      ref_cds = ref_cds_b, comp_cds = comp_cds_b,
      ref_label = "Dominant\n(non-NMD)",
      comp_label = comp_label_b,
      events = pair_events_b,
      title = paste0("B) Compensatory Frameshifts (", pick_b$gene_id, ")"),
      subtitle = sprintf("%d frameshifts, frame restored \u2192 reading frame preserved",
                         pick_b$n_frameshifts)
    )
  }

  # --- Panel C: Frameshift → Novel Protein (Control, no PTC) ---
  # Use novel_protein_results from the earlier computation
  np_dd <- novel_protein_results[["DD"]]
  if (!is.null(np_dd) && nrow(np_dd) > 0 && !is.null(fw_c4)) {
    # Pick a pair with moderate pct_novel (40-80%) and reasonable exon count
    np_cands <- np_dd[np_dd$pct_novel >= 40 & np_dd$pct_novel <= 80, ]
    if (nrow(np_cands) > 0) {
      np_cands <- merge(np_cands,
                        structures[, c("isoform_id", "n_exons")],
                        by.x = "comparator_isoform_id", by.y = "isoform_id")
      np_cands <- np_cands[np_cands$n_exons >= 3 & np_cands$n_exons <= 8, ]
    }
    if (nrow(np_cands) > 0) {
      # Pick one near the median pct_novel
      np_cands <- np_cands[order(abs(np_cands$pct_novel - median(np_cands$pct_novel))), ]
      pick_c <- np_cands[1, ]

      # Get the pair's reference isoform from frame_walk
      ps_c4 <- fw_c4$profile_summary
      pair_info <- ps_c4[ps_c4$comparator_isoform_id == pick_c$comparator_isoform_id &
                         ps_c4$gene_id == pick_c$gene_id, ]
      if (nrow(pair_info) > 0) {
        ref_id_c <- pair_info$reference_isoform_id[1]
        ref_ex_c <- get_exon_coords(ref_id_c, structures)
        comp_ex_c <- get_exon_coords(pick_c$comparator_isoform_id, structures)
        ref_cds_c <- get_cds_range(ref_id_c, cds)
        comp_cds_c <- get_cds_range(pick_c$comparator_isoform_id, cds)
        str_row_c <- structures[structures$isoform_id == ref_id_c, ]
        gene_strand_c <- str_row_c$strand[1]

        profs_c <- set_results[[c4_key]]$profiles
        pair_prof_c <- profs_c[profs_c$reference_isoform_id == ref_id_c &
                               profs_c$comparator_isoform_id == pick_c$comparator_isoform_id, ]
        pair_events_c <- if (nrow(pair_prof_c) > 0) pair_prof_c$detailed_events[[1]] else NULL

        # Get frameshift boundary for novel protein shading
        evts_c <- fw_c4$events
        evts_fs_c <- evts_c[evts_c$is_frameshift == TRUE &
                            evts_c$comparator_isoform_id == pick_c$comparator_isoform_id &
                            evts_c$gene_id == pick_c$gene_id, ]
        fs_ann <- list()
        if (nrow(evts_fs_c) > 0) {
          first_fs <- evts_fs_c[1, ]
          fs_pos <- if (gene_strand_c == "+") first_fs$genomic_end else first_fs$genomic_start
          # Novel region: from frameshift boundary to CDS end
          if (!is.null(comp_cds_c)) {
            novel_end <- if (gene_strand_c == "+") comp_cds_c$cds_stop else comp_cds_c$cds_start
            novel_start <- if (gene_strand_c == "+") fs_pos else novel_end
            novel_stop <- if (gene_strand_c == "+") novel_end else fs_pos
            fs_ann <- list(
              list(type = "frameshift_boundary", x = fs_pos, track = "comp",
                   label = "Frameshift", color = "#8E44AD"),
              list(type = "region_shade",
                   xmin = min(novel_start, novel_stop),
                   xmax = max(novel_start, novel_stop),
                   track = "comp", color = "#E67E22", alpha = 0.3,
                   label = sprintf("Novel protein (%d aa, %.0f%%)",
                                   round(pick_c$novel_aa), pick_c$pct_novel))
            )
          }
        }

        example_plots[["C"]] <- plot_ucsc_pair(
          ref_ex_c$starts, ref_ex_c$ends,
          comp_ex_c$starts, comp_ex_c$ends,
          gene_strand_c,
          ref_cds = ref_cds_c, comp_cds = comp_cds_c,
          ref_label = "Dominant\n(non-NMD)",
          comp_label = "Next-best\n(non-NMD)",
          events = pair_events_c,
          annotations = fs_ann,
          title = paste0("C) Novel Protein \u2014 Stable Frameshift (",
                         pick_c$gene_id, ")"),
          subtitle = sprintf("Unresolved frameshift, no PTC, stop in last exon \u2192 %d novel aa (%.0f%% of protein)",
                             round(pick_c$novel_aa), pick_c$pct_novel)
        )
      }
    }
  }

  # Combine panels
  if (length(example_plots) > 0) {
    if (requireNamespace("patchwork", quietly = TRUE)) {
      combined <- patchwork::wrap_plots(example_plots, ncol = 1) +
        patchwork::plot_annotation(
          title = "Frameshift Consequences: Representative Gene Examples",
          theme = theme(plot.title = element_text(size = 14, face = "bold"))
        )
      print(combined)
    } else {
      for (pp in example_plots) print(pp)
    }
  }
}
```

## Stop Codon Distance to Last EJC

The distance from a transcript's stop codon to the last exon-exon junction (EJC) is a
key determinant of NMD susceptibility. When a stop codon sits >50 nt upstream of the last
EJC, the downstream junction(s) remain associated with the exon junction complex after
splicing, triggering NMD surveillance. Greater distance (more downstream EJCs) is generally
associated with more efficient NMD.

The distance is computed in mRNA coordinates as `last_EJC_position - stop_codon_position`.
Positive values indicate the stop codon is upstream of the last EJC (potential NMD trigger);
negative values indicate a stop codon downstream of the last EJC (normal termination).
Single-exon isoforms have no EJC and are excluded.

### Distribution of Stop-to-EJC Distance: All Comparators

Distribution of stop codon-to-last-EJC distances for **all** comparator isoforms (not just
PTC+), stratified by NMD and Control. This shows the full landscape of stop codon positioning
rather than conditioning on the PTC+ tail.

```{r sec3-stop-dist, results='asis', fig.height=6}
# Build stop-to-EJC distance data for all comparator isoforms
stop_dist_all <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  rows <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) next
    comp_ids <- set_results[[key]]$profiles$comparator_isoform_id
    ptc_sub <- ptc[ptc$isoform_id %in% comp_ids & !is.na(ptc$ptc_distance), ]
    if (nrow(ptc_sub) == 0) next
    rows[[comp]] <- data.frame(
      cell_type = ct,
      comparison = comp_label(comp),
      ptc_distance = ptc_sub$ptc_distance,
      has_ptc = ptc_sub$has_ptc
    )
  }
  do.call(rbind, rows)
}))

if (!is.null(stop_dist_all) && nrow(stop_dist_all) > 0 && has_ggplot2) {
  # Summary table
  dist_summary <- stop_dist_all |>
    group_by(cell_type, comparison) |>
    summarize(
      n_isoforms = n(),
      median_dist = round(median(ptc_distance), 0),
      pct_ptc_pos = round(100 * mean(has_ptc), 1),
      pct_negative = round(100 * mean(ptc_distance < 0), 1),
      .groups = "drop"
    )
  cat("**Columns:** **n_isoforms** = comparator isoforms with stop-to-EJC distance computed; ")
  cat("**median_dist** = median distance (nt); **pct_ptc_pos** = % with PTC (distance > 50 nt); ")
  cat("**pct_negative** = % with stop codon downstream of last EJC (normal termination).\n\n")
  knitr::kable(dist_summary, caption = "Stop Codon-to-Last EJC Distance: All Comparators") |> print()
  cat("\n\n")

  # Density plot
  p_dist <- ggplot(stop_dist_all, aes(x = ptc_distance, fill = comparison)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
    annotate("text", x = 50, y = Inf, label = "PTC threshold (50 nt)",
             hjust = -0.05, vjust = 1.5, size = 3, color = "red") +
    facet_wrap(~cell_type, scales = "free_y") +
    scale_x_continuous(limits = c(
      quantile(stop_dist_all$ptc_distance, 0.01, na.rm = TRUE),
      quantile(stop_dist_all$ptc_distance, 0.99, na.rm = TRUE))) +
    theme_minimal() +
    labs(title = "Stop Codon Distance to Last EJC: NMD vs Control Comparators",
         subtitle = "All comparator isoforms (coding, multi-exon). Dashed red = 50 nt PTC threshold.",
         x = "Distance from stop codon to last EJC (nt)", y = "Density", fill = "")
  print(p_dist)
  cat("\n\n")
}
```

### NMD Response vs Stop Codon Distance

Among NMD-sensitive comparator isoforms, is there a relationship between the distance from
the stop codon to the last EJC and the magnitude of upregulation upon NMD inhibition (logFC)?
Isoforms with stop codons further from the last EJC may be subject to more efficient NMD
surveillance and thus show stronger rescue (higher logFC) when NMD is inhibited.

```{r sec3-nmd-logfc-dist, results='asis', fig.height=6}
# Load DE data to get logFC for NMD comparator isoforms
de_dir <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge"
de_date <- "2026.3.1"
ct_to_de_suffix <- c(AT = "at", DD = "dd", DD_ALI = "ddali",
                      DO = "do", FB = "fb", MV = "mv")

# Collect NMD comparator isoforms with logFC and ptc_distance
nmd_logfc_dist <- do.call(rbind, lapply(setdiff(display_cts, "all_samples"), function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  if (!c2_key %in% names(set_results)) return(NULL)

  comp_ids <- set_results[[c2_key]]$profiles$comparator_isoform_id

  # Load DE results for this cell type
  de_suffix <- ct_to_de_suffix[ct]
  if (is.na(de_suffix)) return(NULL)
  de_file <- file.path(de_dir, sprintf("nmd_isocall_dge_%s_%s.csv", de_suffix, de_date))
  if (!file.exists(de_file)) return(NULL)
  de <- read.csv(de_file)

  # Match comparator isoforms to DE logFC and PTC distance
  de_sub <- de[de$transcript_id %in% comp_ids, c("transcript_id", "logFC")]
  ptc_sub <- ptc[ptc$isoform_id %in% comp_ids & !is.na(ptc$ptc_distance),
                  c("isoform_id", "ptc_distance", "has_ptc")]
  merged <- merge(de_sub, ptc_sub,
                   by.x = "transcript_id", by.y = "isoform_id")
  if (nrow(merged) == 0) return(NULL)
  merged$cell_type <- ct
  merged
}))

if (!is.null(nmd_logfc_dist) && nrow(nmd_logfc_dist) > 0 && has_ggplot2) {
  # Scatter/hex plot: ptc_distance vs logFC
  p_logfc <- ggplot(nmd_logfc_dist, aes(x = ptc_distance, y = logFC)) +
    geom_point(alpha = 0.15, size = 0.8) +
    geom_smooth(method = "loess", se = TRUE, color = "steelblue") +
    geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
    facet_wrap(~cell_type, scales = "free") +
    scale_x_continuous(limits = c(
      quantile(nmd_logfc_dist$ptc_distance, 0.01, na.rm = TRUE),
      quantile(nmd_logfc_dist$ptc_distance, 0.99, na.rm = TRUE))) +
    theme_minimal() +
    labs(title = "NMD Response vs Stop Codon Distance to Last EJC",
         subtitle = "NMD comparator isoforms. Blue = loess smooth. Red dashed = 50 nt PTC threshold.",
         x = "Distance from stop codon to last EJC (nt)",
         y = "logFC (Smg1i vs DMSO)", fill = "")
  print(p_logfc)
  cat("\n\n")

  # Correlation summary
  cor_summary <- nmd_logfc_dist |>
    group_by(cell_type) |>
    summarize(
      n = n(),
      spearman_rho = round(cor(ptc_distance, logFC, method = "spearman",
                                use = "complete.obs"), 3),
      cor_p = cor.test(ptc_distance, logFC, method = "spearman")$p.value,
      .groups = "drop"
    )
  cat("**Columns:** **n** = NMD comparator isoforms with both logFC and stop-to-EJC distance; ")
  cat("**spearman_rho** = Spearman correlation between distance and logFC; ")
  cat("**cor_p** = p-value for the correlation.\n\n")
  knitr::kable(cor_summary,
               caption = "Correlation: Stop-to-EJC Distance vs logFC (NMD comparators)") |> print()
  cat("\n\n")
}
```

### PTC Position Metrics: Which Best Predicts NMD Response?

The loess curves above suggest a non-linear relationship between stop codon distance and
logFC. To better understand what drives NMD efficiency, we compare three ways of measuring
PTC position, using DD as a focal cell type:

- **PTC distance (nt)**: Raw nucleotide distance from stop codon to last EJC.
- **% CDS distance**: PTC distance normalized by ORF length (`ptc_distance / orf_length * 100`).
- **N downstream EJCs**: Number of exon-exon junctions downstream of the stop codon.

For each metric we fit a generalized additive model (GAM) to capture non-linear effects,
plus linear and quadratic models for comparison. All models are restricted to PTC+ isoforms
(distance > 50 nt) in the DD NMD comparison.

```{r sec3-ptc-position-metrics, fig.height=5, fig.width=10}
# Merge PTC features with logFC for DD NMD comparators
dd_c2_key <- "c2_dd"
dd_comp_ids <- set_results[[dd_c2_key]]$profiles$comparator_isoform_id

de_dd <- read.csv(file.path(de_dir, sprintf("nmd_isocall_dge_dd_%s.csv", de_date)))
de_dd_sub <- de_dd[de_dd$transcript_id %in% dd_comp_ids, c("transcript_id", "logFC")]

ptc_dd <- ptc[ptc$isoform_id %in% dd_comp_ids, ]
dd_merged <- merge(de_dd_sub, ptc_dd, by.x = "transcript_id", by.y = "isoform_id")
dd_merged <- dd_merged[!is.na(dd_merged$ptc_distance) & !is.na(dd_merged$orf_length) &
                         dd_merged$orf_length > 0, ]
dd_merged$pct_cds_distance <- 100 * dd_merged$ptc_distance / dd_merged$orf_length
dd_ptc_pos <- dd_merged[dd_merged$has_ptc == TRUE, ]
```

```{r sec3-ptc-panel-plot, fig.height=5, fig.width=12}
if (has_ggplot2 && nrow(dd_ptc_pos) > 0) {
  ptc_long <- rbind(
    data.frame(metric = "PTC distance (nt)", value = dd_ptc_pos$ptc_distance,
               logFC = dd_ptc_pos$logFC),
    data.frame(metric = "% CDS distance", value = dd_ptc_pos$pct_cds_distance,
               logFC = dd_ptc_pos$logFC),
    data.frame(metric = "N downstream EJCs", value = dd_ptc_pos$n_downstream_ejcs,
               logFC = dd_ptc_pos$logFC)
  )
  ptc_long$metric <- factor(ptc_long$metric,
    levels = c("PTC distance (nt)", "% CDS distance", "N downstream EJCs"))

  p_panel <- ggplot(ptc_long, aes(x = value, y = logFC)) +
    geom_point(alpha = 0.2, size = 0.8) +
    geom_smooth(method = "loess", span = 0.5, color = "steelblue", linewidth = 1.2) +
    facet_wrap(~metric, scales = "free_x") +
    theme_minimal() +
    labs(title = sprintf("DD NMD Comparators (PTC+, n = %d): logFC vs PTC Position Metrics",
                         nrow(dd_ptc_pos)),
         x = "Metric value", y = "logFC (Smg1i vs DMSO)")
  print(p_panel)
}
```

```{r sec3-ejc-boxplot, fig.height=5, fig.width=9}
if (has_ggplot2 && nrow(dd_ptc_pos) > 0) {
  ejc_counts <- dd_ptc_pos |>
    count(n_downstream_ejcs) |>
    mutate(label = paste0("n=", n))

  p_box <- ggplot(dd_ptc_pos, aes(x = factor(n_downstream_ejcs), y = logFC)) +
    geom_violin(fill = "lightblue", alpha = 0.5) +
    geom_boxplot(width = 0.15, outlier.alpha = 0.2) +
    geom_text(data = ejc_counts,
              aes(x = factor(n_downstream_ejcs),
                  y = max(dd_ptc_pos$logFC) + 0.3, label = label),
              size = 2.8) +
    theme_minimal() +
    labs(title = "DD NMD Comparators: logFC by Number of Downstream EJCs",
         subtitle = "logFC rises sharply from 1-4 EJCs, then plateaus",
         x = "Number of downstream EJCs", y = "logFC (Smg1i vs DMSO)")
  print(p_box)
}
```

```{r sec3-gam-comparison, results='asis'}
if (requireNamespace("mgcv", quietly = TRUE) && nrow(dd_ptc_pos) > 0) {
  n_ejc_unique <- length(unique(dd_ptc_pos$n_downstream_ejcs))

  # GAM models
  gam_dist <- mgcv::gam(logFC ~ s(ptc_distance), data = dd_ptc_pos)
  gam_pct  <- mgcv::gam(logFC ~ s(pct_cds_distance), data = dd_ptc_pos)
  gam_ejc  <- mgcv::gam(logFC ~ s(n_downstream_ejcs, k = min(n_ejc_unique, 10)),
                          data = dd_ptc_pos)
  gam_comb <- mgcv::gam(logFC ~ s(ptc_distance) +
                           s(n_downstream_ejcs, k = min(n_ejc_unique, 10)),
                          data = dd_ptc_pos)

  # Linear and quadratic models
  lm_dist_lin <- lm(logFC ~ ptc_distance, data = dd_ptc_pos)
  lm_pct_lin  <- lm(logFC ~ pct_cds_distance, data = dd_ptc_pos)
  lm_ejc_lin  <- lm(logFC ~ n_downstream_ejcs, data = dd_ptc_pos)
  lm_dist_q   <- lm(logFC ~ poly(ptc_distance, 2), data = dd_ptc_pos)
  lm_pct_q    <- lm(logFC ~ poly(pct_cds_distance, 2), data = dd_ptc_pos)
  lm_ejc_q    <- lm(logFC ~ poly(n_downstream_ejcs, 2), data = dd_ptc_pos)

  model_comparison <- data.frame(
    metric = c(rep("ptc_distance (nt)", 3), rep("pct_cds_distance (%)", 3),
               rep("n_downstream_ejcs", 3), "combined (dist + ejcs)"),
    model = c(rep(c("Linear", "Quadratic", "GAM"), 3), "GAM"),
    R_squared = round(c(
      summary(lm_dist_lin)$r.squared, summary(lm_dist_q)$r.squared,
        summary(gam_dist)$r.sq,
      summary(lm_pct_lin)$r.squared, summary(lm_pct_q)$r.squared,
        summary(gam_pct)$r.sq,
      summary(lm_ejc_lin)$r.squared, summary(lm_ejc_q)$r.squared,
        summary(gam_ejc)$r.sq,
      summary(gam_comb)$r.sq
    ), 3),
    AIC = round(c(
      AIC(lm_dist_lin), AIC(lm_dist_q), AIC(gam_dist),
      AIC(lm_pct_lin), AIC(lm_pct_q), AIC(gam_pct),
      AIC(lm_ejc_lin), AIC(lm_ejc_q), AIC(gam_ejc),
      AIC(gam_comb)
    ), 0)
  )
  knitr::kable(model_comparison,
               caption = sprintf("Model Comparison: PTC Position Metrics vs logFC (DD, n = %d PTC+ isoforms)",
                                  nrow(dd_ptc_pos))) |> print()
}
```

The number of downstream exon-exon junctions (`n_downstream_ejcs`) is the strongest single
predictor of NMD response magnitude, explaining ~15% of logFC variance in GAM models compared
to ~13% for raw nucleotide distance and ~8% for % CDS distance. Normalizing by CDS length
actually weakens the signal --- the absolute number of downstream junctions matters more than
the relative position within the transcript. The relationship is non-linear: logFC rises
sharply from 1 to ~4 downstream EJCs and then plateaus, consistent with EJC-based NMD
surveillance having diminishing returns beyond a threshold number of junctions. Quadratic
terms are highly significant for all three metrics, confirming the non-linear pattern.

## Frame-PTC Cross-Reference

Among NMD pairs with unresolved frameshifts, what fraction also have a PTC annotated in
the comparator isoform? High concordance confirms that frameshifts and PTCs are mechanistically
linked (frameshifts introduce PTCs downstream).

```{r sec3-frame-ptc-xref, results='asis'}
for (ct in display_cts) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  if (!c2_key %in% names(set_results)) next
  fw <- set_results[[c2_key]]$frame_walk
  if (is.null(fw) || is.null(fw$profile_summary)) next
  ps <- fw$profile_summary
  n_unresolved <- sum(ps$frame_resolved == FALSE & !is.na(ps$frame_resolved))
  if (n_unresolved > 0 && any(!is.na(ps$comparator_has_ptc))) {
    n_unres_ptc <- sum(ps$frame_resolved == FALSE & ps$comparator_has_ptc == TRUE, na.rm = TRUE)
    cat(sprintf("**%s**: Of NMD pairs with unresolved frameshifts, %d/%d (%.1f%%) have a PTC in the comparator.\n",
                ct, n_unres_ptc, n_unresolved, 100 * n_unres_ptc / n_unresolved))
  }
}
cat("\n\n")
```

## Frame Walk by Event Type {.tabset}

Breakdown of frameshift and compensatory rates by event type, comparing NMD and Control pairs.
Only events with CDS overlap are counted.

**Columns:** **event_type** = splicing event type; **n_cds_events** = number of events overlapping the CDS; **pct_frameshift** = % of those events that shift the reading frame (CDS overlap not divisible by 3); **pct_compensatory** = % that restore the frame after a prior shift. Suffixes **_NMD/_Ctrl** indicate the comparison group.

```{r sec3-frame-walk-by-event, results='asis'}
for (ct in display_cts) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results)) next
  cat(sprintf("### %s\n\n", ct))

  fw_tables <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_", ct_label)
    if (!key %in% names(set_results)) next
    fw <- set_results[[key]]$frame_walk
    if (is.null(fw) || nrow(fw$events) == 0L) next
    fw_evts <- fw$events
    fw_tables[[comp]] <- do.call(rbind, lapply(unique(fw_evts$event_type), function(et) {
      e <- fw_evts[fw_evts$event_type == et, ]
      data.frame(event_type = et,
                 n_cds_events = nrow(e),
                 pct_frameshift = round(100 * sum(e$is_frameshift) / nrow(e), 1),
                 pct_compensatory = round(100 * sum(e$is_compensatory) / nrow(e), 1))
    }))
  }

  if (length(fw_tables) == 2) {
    fw_merged <- merge(fw_tables[["c2"]], fw_tables[["c4"]],
                       by = "event_type", suffixes = c("_NMD", "_Ctrl"), all = TRUE)
    render_dt(fw_merged, caption = sprintf("Frame Walk by Event Type: NMD vs Control — %s", ct))
  } else if ("c2" %in% names(fw_tables)) {
    render_dt(fw_tables[["c2"]], caption = sprintf("Frame Walk by Event Type: NMD — %s", ct))
  }
  cat("\n\n")
}
```

# 4. Structural Differences

At baseline, non-NMD isoform pairs share ~50--58% of their exonic sequence and differ by a
median of ~700--800bp. IR and terminal events are the most prevalent event types, and the
next-best (comparator) isoform is typically shorter than the dominant (positive net_bp). This
represents the natural structural diversity between co-expressed isoforms from the same gene.

## Event Frequency {.tabset}

Prevalence of each splicing event type, measured as the percentage of isoform pairs containing
at least one instance of that event. NMD and Control are compared using Fisher's exact test per event type.

**Columns:** **event_type** = splicing event; **NMD_n/Ctrl_n** = number of pairs with the event; **NMD_pct/Ctrl_pct** = % of pairs; **OR** = odds ratio (>1 = more prevalent in NMD); **adj_p** = BH-adjusted p-value; **direction** = which group is enriched.

```{r sec4-event-freq, results='asis', fig.height=8}
for (ct in display_cts) {
  cps <- cross_results[[ct]]
  if (is.null(cps)) next
  ef <- cps$event_frequency
  if (is.null(ef) || isTRUE(ef$skipped)) next

  cat(sprintf("### %s\n\n", ct))

  ef_df <- do.call(rbind, lapply(ef$per_event, function(e) {
    dir_label <- gsub("enriched in A", "enriched in NMD",
                 gsub("enriched in B", "enriched in Control", e$direction))
    data.frame(
      event_type = e$event_type,
      NMD_n = e$count_a, NMD_pct = round(100 * e$count_a / e$n_a, 1),
      Ctrl_n = e$count_b, Ctrl_pct = round(100 * e$count_b / e$n_b, 1),
      OR = round(e$odds_ratio, 2),
      adj_p = e$adj_p_value,
      direction = dir_label
    )
  }))
  render_dt(ef_df, caption = sprintf("Event Type Prevalence: NMD vs Control — %s", ct))

  # Condensed one-liners for event count, profile categories, co-occurrence
  ec <- cps$event_count
  if (!is.null(ec) && !isTRUE(ec$skipped))
    cat(sprintf("Event count: Wilcoxon p = %s, d = %.3f (NMD median = %.0f, Control = %.0f)\n\n",
                format(ec$p_value, digits = 3), ec$cliffs_d, ec$median_a, ec$median_b))

  pc <- cps$profile_categories
  if (!is.null(pc) && !isTRUE(pc$skipped))
    cat(sprintf("Profile categories: Chi-square p = %s, V = %.3f\n\n",
                format(pc$p_value, digits = 3), pc$cramers_v))

  cooc_cps <- cps$cooccurrence
  if (!is.null(cooc_cps) && !isTRUE(cooc_cps$skipped))
    cat(sprintf("Co-occurrence structure is shared between NMD and Control (r = %.3f).\n\n",
                cooc_cps$correlation))

  # Paired tests (strongest within-gene evidence)
  pt <- cps$paired_tests
  if (!is.null(pt) && !isTRUE(pt$skipped)) {
    cat(sprintf("Paired analysis (%d gene-matched pairs): Wilcoxon p = %s; sign test: %.0f%% of genes have more events in NMD (p = %s)\n\n",
                pt$n_genes_overlap,
                format(pt$paired_wilcox$p_value, digits = 3),
                pt$sign_test$pct_a_more,
                format(pt$sign_test$p_value, digits = 3)))
  }
  cat("\n\n")
}
```

## Structural Divergence

Sequence similarity and exonic length differences between reference and comparator isoforms.
Higher pct_shared indicates more similar isoforms; larger exonic_diff indicates greater length disparity.

**Columns:** **NMD_med_pct_shared/Ctrl_med_pct_shared** = median % of exonic sequence shared between reference and comparator; **NMD_med_exonic_diff/Ctrl_med_exonic_diff** = median absolute exonic bp difference; **p_value** = Wilcoxon rank-sum test on pct_shared.

```{r sec4-divergence, results='asis'}
div_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  div_c2 <- set_results[[c2_key]]$divergence
  div_c4 <- set_results[[c4_key]]$divergence
  if (is.null(div_c2) || nrow(div_c2) == 0 || is.null(div_c4) || nrow(div_c4) == 0) return(NULL)
  wt <- tryCatch(wilcox.test(div_c2$pct_shared, div_c4$pct_shared), error = function(e) NULL)
  data.frame(
    cell_type = ct,
    NMD_med_pct_shared = round(median(div_c2$pct_shared, na.rm = TRUE), 1),
    Ctrl_med_pct_shared = round(median(div_c4$pct_shared, na.rm = TRUE), 1),
    NMD_med_exonic_diff = round(median(abs(div_c2$exonic_bp_diff)), 0),
    Ctrl_med_exonic_diff = round(median(abs(div_c4$exonic_bp_diff)), 0),
    p_value = if (!is.null(wt)) wt$p.value else NA_real_
  )
}))
if (!is.null(div_rows)) {
  knitr::kable(div_rows, caption = "Structural Divergence: NMD vs Control") |> print()
  cat("\n\n")
}
```

## Gain/Loss Balance

Net base pair change between reference and comparator isoforms. Computed as
`net_bp = bp_loss - bp_gain`, where LOSS and GAIN are from the comparator's perspective.
**Positive net_bp** means the comparator has lost more sequence than it gained (comparator is
shorter than the reference). **Negative net_bp** means the comparator has gained more sequence
(comparator is longer than the reference).

**Columns:** **NMD_med_net_bp/Ctrl_med_net_bp** = median net bp change; **wilcox_p** = Wilcoxon rank-sum test; **cliffs_d** = Cliff's delta effect size.

```{r sec4-gain-loss, results='asis'}
gl_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  cps <- cross_results[[ct]]
  if (is.null(cps)) return(NULL)
  gl <- cps$gain_loss
  if (is.null(gl) || isTRUE(gl$skipped)) return(NULL)
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  data.frame(
    cell_type = ct,
    NMD_med_net_bp = round(median(set_results[[c2_key]]$profiles$net_bp), 0),
    Ctrl_med_net_bp = round(median(set_results[[c4_key]]$profiles$net_bp), 0),
    wilcox_p = gl$p_value,
    cliffs_d = round(gl$cliffs_d, 3)
  )
}))
if (!is.null(gl_rows)) {
  knitr::kable(gl_rows, caption = "Net BP Change: NMD vs Control") |> print()
  cat("\n\n")
}
```

## Gain/Loss Direction by Event Type {.tabset}

For each event type, the direction indicates whether the comparator has GAINED or LOST that
feature relative to the reference. GAIN means the comparator has additional sequence (e.g., an
extra exon); LOSS means the comparator is missing sequence present in the reference.

**Columns:** **NMD_gain/NMD_loss** = event counts by direction in NMD pairs; **NMD_pct_gain** = % of events that are gains; **Ctrl_gain/Ctrl_loss/Ctrl_pct_gain** = same for Control; **fisher_p** = Fisher's exact test; **adj_p** = BH-adjusted p-value.

```{r sec4-gain-loss-direction, results='asis'}
for (ct in display_cts) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) next
  cat(sprintf("### %s\n\n", ct))

  events_c2 <- dplyr::bind_rows(set_results[[c2_key]]$profiles$detailed_events)
  events_c4 <- dplyr::bind_rows(set_results[[c4_key]]$profiles$detailed_events)

  if (nrow(events_c2) > 0 && nrow(events_c4) > 0) {
    dir_comp <- do.call(rbind, lapply(
      unique(c(events_c2$event_type, events_c4$event_type)), function(et) {
        e2 <- events_c2[events_c2$event_type == et, ]
        e4 <- events_c4[events_c4$event_type == et, ]
        g2 <- sum(e2$direction == "GAIN"); l2 <- sum(e2$direction == "LOSS")
        g4 <- sum(e4$direction == "GAIN"); l4 <- sum(e4$direction == "LOSS")
        ft <- if ((g2 + l2) > 0 && (g4 + l4) > 0)
          tryCatch(fisher.test(matrix(c(g2, l2, g4, l4), nrow = 2)), error = function(e) NULL) else NULL
        data.frame(
          event_type = et,
          NMD_gain = g2, NMD_loss = l2,
          NMD_pct_gain = round(100 * g2 / max(g2+l2, 1), 1),
          Ctrl_gain = g4, Ctrl_loss = l4,
          Ctrl_pct_gain = round(100 * g4 / max(g4+l4, 1), 1),
          fisher_p = if (!is.null(ft)) ft$p.value else NA_real_
        )
    }))
    dir_comp$adj_p <- p.adjust(dir_comp$fisher_p, "BH")
    render_dt(dir_comp, caption = sprintf("Gain/Loss Direction: NMD vs Control — %s", ct))
  }
  cat("\n\n")
}
```

## Intron Retention Analysis

Intron retention (IR) is the most prevalent event type at baseline, making it the dominant
structural difference between co-expressed isoforms from the same gene. IR events are detected
when one isoform has a single exon that spans the region covered by two or more exons (plus
the intervening intron) in the other isoform. Long-read sequencing may capture more IR than
short-read studies, since retained introns extend transcript length and are difficult to
assemble from short reads.

This section examines IR events (including IR, IR_diff, Partial_IR_5, Partial_IR_3) in
detail. **Direction reminder:** A GAIN means the comparator isoform has *retained* an intron
that is spliced out in the reference; a LOSS means the comparator has *lost* a retained intron
present in the reference (i.e., the comparator splices it out).

### IR Event Direction: Individual Types {.tabset}

GAIN/LOSS breakdown for each IR-family event type, stratified by NMD and Control.

```{r sec4-ir-individual, results='asis'}
# Define IR event types
ir_types <- c("IR", "IR_diff", "Partial_IR_5", "Partial_IR_3")

for (ct in display_cts) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) next
  cat(sprintf("#### %s\n\n", ct))

  events_c2 <- dplyr::bind_rows(set_results[[c2_key]]$profiles$detailed_events)
  events_c4 <- dplyr::bind_rows(set_results[[c4_key]]$profiles$detailed_events)

  ir_detail <- do.call(rbind, lapply(ir_types, function(et) {
    e2 <- events_c2[events_c2$event_type == et, ]
    e4 <- events_c4[events_c4$event_type == et, ]
    g2 <- sum(e2$direction == "GAIN"); l2 <- sum(e2$direction == "LOSS")
    g4 <- sum(e4$direction == "GAIN"); l4 <- sum(e4$direction == "LOSS")
    ft <- if ((g2 + l2) > 0 && (g4 + l4) > 0)
      tryCatch(fisher.test(matrix(c(g2, l2, g4, l4), nrow = 2)), error = function(e) NULL) else NULL
    data.frame(
      event_type = et,
      NMD_gain = g2, NMD_loss = l2, NMD_total = g2 + l2,
      NMD_pct_gain = round(100 * g2 / max(g2 + l2, 1), 1),
      Ctrl_gain = g4, Ctrl_loss = l4, Ctrl_total = g4 + l4,
      Ctrl_pct_gain = round(100 * g4 / max(g4 + l4, 1), 1),
      fisher_p = if (!is.null(ft)) ft$p.value else NA_real_
    )
  }))
  ir_detail$adj_p <- p.adjust(ir_detail$fisher_p, "BH")

  knitr::kable(ir_detail,
               caption = sprintf("IR Event Direction: NMD vs Control — %s", ct)) |> print()
  cat("\n\n")
}
```

### Cumulative IR Burden {.tabset}

Total IR-family events (IR + IR_diff + Partial_IR_5 + Partial_IR_3) aggregated across all
subtypes, comparing NMD and Control. This measures the overall intron retention burden
per pair and whether the direction (GAIN vs LOSS) differs between comparison groups.

```{r sec4-ir-cumulative, results='asis', fig.height=8}
ir_cumul_rows <- do.call(rbind, lapply(display_cts, function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)

  events_c2 <- dplyr::bind_rows(set_results[[c2_key]]$profiles$detailed_events)
  events_c4 <- dplyr::bind_rows(set_results[[c4_key]]$profiles$detailed_events)
  ir_c2 <- events_c2[events_c2$event_type %in% ir_types, ]
  ir_c4 <- events_c4[events_c4$event_type %in% ir_types, ]

  n_c2 <- nrow(set_results[[c2_key]]$profiles)
  n_c4 <- nrow(set_results[[c4_key]]$profiles)

  # Per-pair IR event counts
  ir_per_pair_c2 <- sapply(set_results[[c2_key]]$profiles$detailed_events, function(de)
    sum(de$event_type %in% ir_types))
  ir_per_pair_c4 <- sapply(set_results[[c4_key]]$profiles$detailed_events, function(de)
    sum(de$event_type %in% ir_types))

  wt <- tryCatch(wilcox.test(ir_per_pair_c2, ir_per_pair_c4), error = function(e) NULL)

  g2 <- sum(ir_c2$direction == "GAIN"); l2 <- sum(ir_c2$direction == "LOSS")
  g4 <- sum(ir_c4$direction == "GAIN"); l4 <- sum(ir_c4$direction == "LOSS")
  ft <- if ((g2 + l2) > 0 && (g4 + l4) > 0)
    tryCatch(fisher.test(matrix(c(g2, l2, g4, l4), nrow = 2)), error = function(e) NULL) else NULL

  data.frame(
    cell_type = ct,
    NMD_total_ir = nrow(ir_c2),
    NMD_per_pair = round(nrow(ir_c2) / n_c2, 2),
    NMD_pct_gain = round(100 * g2 / max(g2 + l2, 1), 1),
    Ctrl_total_ir = nrow(ir_c4),
    Ctrl_per_pair = round(nrow(ir_c4) / n_c4, 2),
    Ctrl_pct_gain = round(100 * g4 / max(g4 + l4, 1), 1),
    count_wilcox_p = if (!is.null(wt)) wt$p.value else NA_real_,
    dir_fisher_p = if (!is.null(ft)) ft$p.value else NA_real_
  )
}))

if (!is.null(ir_cumul_rows)) {
  cat("**Columns:** **NMD_total_ir/Ctrl_total_ir** = total IR-family events; ")
  cat("**NMD_per_pair/Ctrl_per_pair** = mean IR events per isoform pair; ")
  cat("**NMD_pct_gain/Ctrl_pct_gain** = % of IR events that are GAINs (comparator retains intron); ")
  cat("**count_wilcox_p** = Wilcoxon test on per-pair IR event counts (NMD vs Control); ")
  cat("**dir_fisher_p** = Fisher's exact test on GAIN vs LOSS proportions.\n\n")
  knitr::kable(ir_cumul_rows, caption = "Cumulative IR Burden: NMD vs Control") |> print()
  cat("\n\n")
}

```

## Distribution Plots

```{r sec4-dist-plots, results='asis', fig.height=8}
if (has_ggplot2) {
  # pct_shared from divergence
  pct_shared_all <- do.call(rbind, lapply(display_cts, function(ct) {
    ct_label <- ct_to_label(ct)
    rows <- list()
    for (comp in c("c2", "c4")) {
      key <- paste0(comp, "_", ct_label)
      if (!key %in% names(set_results)) next
      div <- set_results[[key]]$divergence
      if (is.null(div) || nrow(div) == 0) next
      rows[[comp]] <- data.frame(cell_type = ct, comparison = comp_label(comp),
                                 pct_shared = div$pct_shared)
    }
    do.call(rbind, rows)
  }))

  if (!is.null(pct_shared_all) && nrow(pct_shared_all) > 0) {
    p1 <- ggplot(pct_shared_all[!is.na(pct_shared_all$pct_shared), ],
                 aes(x = pct_shared, fill = comparison)) +
      geom_density(alpha = 0.5) +
      facet_wrap(~cell_type) +
      theme_minimal() +
      labs(title = "Sequence Shared (%): NMD vs Control",
           subtitle = "NMD shifted right = more similar (targeted changes)",
           x = "pct_shared", y = "Density", fill = "")
    print(p1)
    cat("\n\n")
  }

}
```

# 5. Cross-CT Consistency

## Event Frequency Heatmap

```{r sec6-heatmap, results='asis', fig.height=8}
if (length(main_per_ct) >= 2) {

  compute_freq_vec <- function(ct) {
    p <- profiles_list[[paste0("c2_", ct_to_label(ct))]]
    n <- nrow(p)
    n_pir5 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_5")))
    n_pir3 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_3")))
    data.frame(cell_type = ct, event_type = evt_types,
      pct = round(c(100*sum(p$tss_changed)/n, 100*sum(p$tes_changed)/n,
              100*sum(p$n_a5ss>0)/n, 100*sum(p$n_a3ss>0)/n,
              100*sum(p$n_se>0)/n, 100*sum(p$n_missing_internal>0)/n,
              100*sum(p$n_ir>0)/n, 100*sum(p$n_ir_diff>0)/n,
              100*n_pir5/n, 100*n_pir3/n), 1))
  }

  freq_matrix <- do.call(rbind, lapply(main_per_ct, compute_freq_vec))

  if (has_ggplot2) {
    p_heat <- ggplot(freq_matrix, aes(x = event_type, y = cell_type, fill = pct)) +
      geom_tile(color = "white") +
      geom_text(aes(label = sprintf("%.0f", pct)), size = 3) +
      scale_fill_gradient(low = "white", high = "steelblue", name = "% of pairs") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      labs(title = "NMD Event Frequency by Cell Type (pair-based %)",
           x = "Event Type", y = "Cell Type")
    print(p_heat)
    cat("\n\n")
  }

  freq_wide <- tidyr::pivot_wider(freq_matrix, names_from = event_type, values_from = pct)
  knitr::kable(freq_wide, caption = "NMD Event Frequency (% of pairs) by Cell Type") |> print()
  cat("\n\n")

} else {
  cat("Fewer than 2 main cell types with NMD profiles; skipping cross-cell-type analysis.\n\n")
}
```

## Heterogeneity Test

Tests whether the distribution of event types differs significantly across cell types.
A Pearson's chi-square test on the cell-type x event-type contingency table (raw pair
counts, not percentages). A significant result indicates that event type prevalence varies
across cell types beyond what would be expected by chance.

```{r sec6-heterogeneity, results='asis'}
if (length(main_per_ct) >= 2) {

  freq_count_list <- lapply(main_per_ct, function(ct) {
    p <- profiles_list[[paste0("c2_", ct_to_label(ct))]]
    n_pir5 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_5")))
    n_pir3 <- sum(sapply(p$detailed_events, function(de) any(de$event_type == "Partial_IR_3")))
    c(sum(p$tss_changed), sum(p$tes_changed),
      sum(p$n_a5ss>0), sum(p$n_a3ss>0),
      sum(p$n_se>0), sum(p$n_missing_internal>0),
      sum(p$n_ir>0), sum(p$n_ir_diff>0),
      n_pir5, n_pir3)
  })
  freq_count_matrix <- do.call(rbind, freq_count_list)
  rownames(freq_count_matrix) <- main_per_ct
  colnames(freq_count_matrix) <- evt_types

  chi_main <- tryCatch(chisq.test(freq_count_matrix), error = function(e) NULL)
  if (!is.null(chi_main))
    cat(sprintf("**Main cell types** (AT, DD, FB, MV): Chi-square = %.1f, df = %d, p = %s\n\n",
                chi_main$statistic, chi_main$parameter, format(chi_main$p.value, digits = 3)))

} else {
  cat("Fewer than 2 main cell types with NMD profiles; skipping heterogeneity test.\n\n")
}
```

# 6. DD_ALI

In AT, DD, FB, and MV, NMD deviates from baseline in a consistent direction: more frameshifts,
more PTCs, more exon skipping, comparators longer, changes more targeted. DD_ALI reverses nearly
every one of these deviations --- its NMD transitions look more like an *amplification* of
baseline patterns than a departure from them.

```{r sec7-ddali-setup, include=FALSE}
ali_label <- ct_to_label(ali_ct)
c2_ali <- paste0("c2_", ali_label)
c4_ali <- paste0("c4_", ali_label)
has_ali <- c2_ali %in% names(set_results) && c4_ali %in% names(set_results)
```

## Key Metrics: DD_ALI vs Main Cell Types

```{r sec7-key-metrics, results='asis'}
if (has_ali) {

  range_str <- function(vals) {
    vals <- vals[!is.na(vals)]
    if (length(vals) == 0) return("\u2014")
    sprintf("%.0f\u2013%.0f", min(vals), max(vals))
  }

  ct_metrics <- lapply(main_per_ct, function(ct) {
    ct_label <- ct_to_label(ct)
    c2k <- paste0("c2_", ct_label); c4k <- paste0("c4_", ct_label)
    nmd_ptc <- set_results[[c2k]]$ptc_assoc
    ctrl_ptc <- set_results[[c4k]]$ptc_assoc
    nmd_div <- set_results[[c2k]]$divergence
    ctrl_div <- set_results[[c4k]]$divergence
    nmd_fw <- build_fw_row(c2k); ctrl_fw <- build_fw_row(c4k)
    list(
      nmd_ptc = if (!is.null(nmd_ptc)) 100 * nmd_ptc$comp_ptc_rate else NA,
      ctrl_ptc = if (!is.null(ctrl_ptc)) 100 * ctrl_ptc$comp_ptc_rate else NA,
      nmd_fw = if (!is.null(nmd_fw)) nmd_fw$pct_with_frameshift else NA,
      ctrl_fw = if (!is.null(ctrl_fw)) ctrl_fw$pct_with_frameshift else NA,
      nmd_net_bp = median(set_results[[c2k]]$profiles$net_bp),
      ctrl_net_bp = median(set_results[[c4k]]$profiles$net_bp),
      nmd_pct_shared = if (!is.null(nmd_div) && nrow(nmd_div) > 0)
        median(nmd_div$pct_shared, na.rm = TRUE) else NA,
      ctrl_pct_shared = if (!is.null(ctrl_div) && nrow(ctrl_div) > 0)
        median(ctrl_div$pct_shared, na.rm = TRUE) else NA
    )
  })

  ali_ptc <- set_results[[c2_ali]]$ptc_assoc
  ali_div <- set_results[[c2_ali]]$divergence
  ali_fw <- build_fw_row(c2_ali)
  ali_ptc_val <- if (!is.null(ali_ptc)) round(100 * ali_ptc$comp_ptc_rate, 1) else NA
  ali_fw_val <- if (!is.null(ali_fw)) ali_fw$pct_with_frameshift else NA
  ali_net_bp_val <- round(median(set_results[[c2_ali]]$profiles$net_bp), 0)
  ali_pct_shared_val <- if (!is.null(ali_div) && nrow(ali_div) > 0)
    round(median(ali_div$pct_shared, na.rm = TRUE), 1) else NA

  ddali_summary <- data.frame(
    metric = c("comp_ptc_%", "frameshift_%", "med_net_bp", "med_pct_shared"),
    Baseline_Ctrl = c(
      range_str(sapply(ct_metrics, `[[`, "ctrl_ptc")),
      range_str(sapply(ct_metrics, `[[`, "ctrl_fw")),
      range_str(sapply(ct_metrics, `[[`, "ctrl_net_bp")),
      range_str(sapply(ct_metrics, `[[`, "ctrl_pct_shared"))
    ),
    Main_CTs_NMD = c(
      range_str(sapply(ct_metrics, `[[`, "nmd_ptc")),
      range_str(sapply(ct_metrics, `[[`, "nmd_fw")),
      range_str(sapply(ct_metrics, `[[`, "nmd_net_bp")),
      range_str(sapply(ct_metrics, `[[`, "nmd_pct_shared"))
    ),
    DD_ALI_NMD = c(
      if (!is.na(ali_ptc_val)) sprintf("%.1f", ali_ptc_val) else "\u2014",
      if (!is.na(ali_fw_val)) sprintf("%.1f", ali_fw_val) else "\u2014",
      sprintf("%.0f", ali_net_bp_val),
      if (!is.na(ali_pct_shared_val)) sprintf("%.1f", ali_pct_shared_val) else "\u2014"
    ),
    DD_ALI_direction = c(
      "\u2248 baseline",
      "\u2248 baseline",
      "Amplified baseline",
      "Reversed"
    )
  )
  render_dt(ddali_summary, caption = "DD_ALI Reversal: Key Metrics")
  cat("\n\n")

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

## PTC & Frameshift

```{r sec7-ptc-frameshift, results='asis'}
if (has_ali) {

  ali_pa_c2 <- set_results[[c2_ali]]$ptc_assoc
  ali_pa_c4 <- set_results[[c4_ali]]$ptc_assoc
  ali_cps <- cross_results[[ali_ct]]
  ali_ptc_cps <- if (!is.null(ali_cps)) ali_cps$ptc_rate else NULL

  # Need ali_fw for this chunk too
  ali_fw <- build_fw_row(c2_ali)

  ptc_row <- data.frame(
    metric = c("Comparator PTC %", "Frameshift %", "Unresolved %"),
    NMD = c(
      round(100 * ali_pa_c2$comp_ptc_rate, 1),
      if (!is.null(ali_fw)) ali_fw$pct_with_frameshift else NA,
      if (!is.null(ali_fw)) ali_fw$pct_unresolved else NA
    ),
    Control = c(
      round(100 * ali_pa_c4$comp_ptc_rate, 1),
      {fw4 <- build_fw_row(c4_ali); if (!is.null(fw4)) fw4$pct_with_frameshift else NA},
      {if (!is.null(fw4)) fw4$pct_unresolved else NA}
    )
  )
  knitr::kable(ptc_row, caption = "DD_ALI: PTC & Frameshift Rates") |> print()
  cat("\n")

  if (!is.null(ali_ptc_cps) && !isTRUE(ali_ptc_cps$skipped))
    cat(sprintf("PTC Fisher's exact: OR = %.2f, p = %s\n\n",
                ali_ptc_cps$odds_ratio, format(ali_ptc_cps$p_value, digits = 3)))

  # Stop-to-EJC distance for DD_ALI comparators
  ali_comp_c2 <- set_results[[c2_ali]]$profiles$comparator_isoform_id
  ali_comp_c4 <- set_results[[c4_ali]]$profiles$comparator_isoform_id
  ali_ptc_c2 <- ptc[ptc$isoform_id %in% ali_comp_c2 & !is.na(ptc$ptc_distance), ]
  ali_ptc_c4 <- ptc[ptc$isoform_id %in% ali_comp_c4 & !is.na(ptc$ptc_distance), ]
  if (nrow(ali_ptc_c2) > 0 && nrow(ali_ptc_c4) > 0) {
    wt <- wilcox.test(ali_ptc_c2$ptc_distance, ali_ptc_c4$ptc_distance)
    cat(sprintf("Stop-to-EJC distance (all comparators): NMD median = %.0f nt, Control = %.0f nt, Wilcoxon p = %s\n\n",
                median(ali_ptc_c2$ptc_distance), median(ali_ptc_c4$ptc_distance),
                format(wt$p.value, digits = 3)))
  }

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

## Event Frequency

```{r sec7-event-freq, results='asis'}
if (has_ali) {

  ali_cps <- cross_results[[ali_ct]]
  ali_ef <- if (!is.null(ali_cps)) ali_cps$event_frequency else NULL
  if (!is.null(ali_ef) && !isTRUE(ali_ef$skipped)) {
    ef_df <- do.call(rbind, lapply(ali_ef$per_event, function(e) {
      dir_label <- gsub("enriched in A", "enriched in NMD",
                   gsub("enriched in B", "enriched in Control", e$direction))
      data.frame(
        event_type = e$event_type,
        NMD_n = e$count_a, NMD_pct = round(100 * e$count_a / e$n_a, 1),
        Ctrl_n = e$count_b, Ctrl_pct = round(100 * e$count_b / e$n_b, 1),
        OR = round(e$odds_ratio, 2),
        adj_p = e$adj_p_value,
        direction = dir_label
      )
    }))
    render_dt(ef_df, caption = "DD_ALI: Event Type Prevalence NMD vs Control")
    cat("\n")
  }

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

In main cell types, SE is enriched in NMD. In DD_ALI, SE is enriched in *Control*,
consistent with a fundamentally different NMD mechanism.

## Structural Divergence & Gain/Loss

```{r sec7-divergence, results='asis'}
if (has_ali) {

  ali_div_c2 <- set_results[[c2_ali]]$divergence
  ali_div_c4 <- set_results[[c4_ali]]$divergence
  if (!is.null(ali_div_c2) && nrow(ali_div_c2) > 0 &&
      !is.null(ali_div_c4) && nrow(ali_div_c4) > 0) {
    div_row <- data.frame(
      metric = c("med_pct_shared", "med_|exonic_diff|", "med_net_bp"),
      NMD = c(
        round(median(ali_div_c2$pct_shared, na.rm = TRUE), 1),
        round(median(abs(ali_div_c2$exonic_bp_diff)), 0),
        round(median(set_results[[c2_ali]]$profiles$net_bp), 0)
      ),
      Control = c(
        round(median(ali_div_c4$pct_shared, na.rm = TRUE), 1),
        round(median(abs(ali_div_c4$exonic_bp_diff)), 0),
        round(median(set_results[[c4_ali]]$profiles$net_bp), 0)
      )
    )
    knitr::kable(div_row, caption = "DD_ALI: Structural Divergence") |> print()
    cat("\n")
  }

  ali_cps <- cross_results[[ali_ct]]
  ali_gl <- if (!is.null(ali_cps)) ali_cps$gain_loss else NULL
  if (!is.null(ali_gl) && !isTRUE(ali_gl$skipped))
    cat(sprintf("Gain/loss: Wilcoxon p = %s, d = %.3f\n\n",
                format(ali_gl$p_value, digits = 3), ali_gl$cliffs_d))

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

In main cell types, NMD pairs share MORE sequence than baseline (~70% vs ~55%). DD_ALI
reverses this: NMD pairs share LESS sequence (~28%). DD_ALI NMD comparators are also
much longer than their reference (large positive net_bp), amplifying the baseline pattern
rather than reversing it.

## Gene Overlap with Main CTs

```{r sec7-gene-overlap, results='asis'}
if (has_ali) {

  ali_genes <- unique(profiles_list[[c2_ali]]$gene_id)
  main_gene_lists <- lapply(main_per_ct, function(ct)
    unique(profiles_list[[paste0("c2_", ct_to_label(ct))]]$gene_id))
  names(main_gene_lists) <- main_per_ct

  ali_overlap <- data.frame(
    cell_type = main_per_ct,
    ddali_genes = length(ali_genes),
    shared = sapply(main_per_ct, function(ct)
      length(intersect(ali_genes, main_gene_lists[[ct]]))),
    pct_of_ddali = sapply(main_per_ct, function(ct)
      round(100 * length(intersect(ali_genes, main_gene_lists[[ct]])) / length(ali_genes), 1))
  )
  render_dt(ali_overlap, caption = "DD_ALI Gene Overlap with Main CTs")
  cat("\n\n")

} else {
  cat("DD_ALI profiles not available.\n\n")
}
```

DD_ALI NMD-sensitive isoforms resemble highly divergent baseline isoforms (low sequence
sharing, high retention) rather than the targeted exon-skipping pattern seen in other cell
types. This suggests DD_ALI NMD targets arise through terminal/retention mechanisms rather
than frameshift-driven PTC introduction, possibly reflecting ALI culture conditions.

# 7. Meta-Analysis

Random-effects meta-analysis pools NMD vs Control effect sizes across main cell types
(AT, DD, FB, MV). DD_ALI is excluded due to reversed biological patterns that inflate
between-cell-type heterogeneity. all_samples is excluded as a superset.

## Pooled Effect Sizes Across Cell Types

Each metric from the NMD vs Control comparison is pooled across the four main cell types
using a random-effects model (metafor package, REML estimation). The table reports the
metrics discussed in earlier sections of this report, restricted to those with directional
interpretability.

**Effect size types and direction:**

- **Cliff's delta** (range -1 to +1): Nonparametric effect size for continuous comparisons.
  Positive values indicate NMD > Control; negative values indicate Control > NMD.
  - **event_count**: Positive = NMD pairs have more structural events per pair.
  - **gain_loss**: Positive = NMD pairs have higher net_bp (comparator shorter); negative = NMD comparator is longer.
- **Log odds ratio (log-OR)**: For binary outcomes. Positive = enriched in NMD; negative = enriched in Control.
  Exponentiate to get the odds ratio (e.g., log-OR of 2.3 = OR of ~10).
  - **ptc_rate**: Positive = PTC more frequent in NMD comparators.
  - **event_freq: {type}**: Positive = event type more prevalent in NMD pairs.

**Columns:** **metric** = analysis metric; **effect_size_type** = scale of the estimate;
**pooled_estimate** = meta-analytic effect size on the native scale (log-OR or Cliff's delta);
**OR** = exponentiated odds ratio for log-OR metrics (>1 = enriched in NMD; <1 = enriched in Control),
shown as "—" for Cliff's delta metrics;
**ci_lower/ci_upper** = 95% confidence interval on the native scale;
**p_value** = test of pooled effect = 0;
**I2** = heterogeneity (% of variance due to between-CT differences; >75% = high).

```{r sec8-meta, results='asis'}
if (requireNamespace("metafor", quietly = TRUE)) {
  # Use cached comparePairSets results — pool main CTs only (exclude DD_ALI + all_samples)
  cps_list <- cross_results[intersect(names(cross_results), main_cts)]
  cps_list <- cps_list[!sapply(cps_list, is.null)]

  if (length(cps_list) >= 2) {
    cat(sprintf("**Cell types pooled:** %s\n\n", paste(names(cps_list), collapse = ", ")))

    meta <- tryCatch(
      metaAnalyzePairSets(cps_list, verbose = FALSE),
      error = function(e) {
        cat(sprintf("Meta-analysis failed: %s\n\n", e$message))
        NULL
      }
    )

    if (!is.null(meta)) {
      # Only include metrics that are discussed in the report
      keep_metrics <- c("event_count", "gain_loss", "ptc_rate")
      scalar_sections <- intersect(
        names(meta)[vapply(meta, function(m) {
          !is.null(m$estimate) && !isTRUE(m$skipped)
        }, logical(1))],
        keep_metrics
      )

      meta_summary <- do.call(rbind, lapply(scalar_sections, function(mn) {
        m <- meta[[mn]]
        es_type <- switch(mn,
          event_count = "Cliff's delta",
          gain_loss = "Cliff's delta",
          ptc_rate = "log-OR",
          "—"
        )
        # For log-OR metrics, also show the exponentiated OR
        or_col <- if (es_type == "log-OR" && !is.null(m$pooled_or)) {
          sprintf("%.1f", m$pooled_or)
        } else { "—" }
        data.frame(
          metric = mn,
          effect_size_type = es_type,
          pooled_estimate = format(m$estimate, digits = 3),
          OR = or_col,
          ci_lower = format(m$ci_lower, digits = 3),
          ci_upper = format(m$ci_upper, digits = 3),
          p_value = format(m$p_value, digits = 3),
          I2 = sprintf("%.1f%%", m$I2)
        )
      }))

      # Add event_frequency per-event-type results (all log-OR)
      if (!is.null(meta$event_frequency) && is.list(meta$event_frequency)) {
        ef_rows <- do.call(rbind, lapply(names(meta$event_frequency), function(et) {
          m <- meta$event_frequency[[et]]
          if (isTRUE(m$skipped) || is.null(m$estimate)) return(NULL)
          or_col <- if (!is.null(m$pooled_or)) sprintf("%.2f", m$pooled_or) else
                      sprintf("%.2f", exp(m$estimate))
          data.frame(
            metric = paste0("event_freq: ", et),
            effect_size_type = "log-OR",
            pooled_estimate = format(m$estimate, digits = 3),
            OR = or_col,
            ci_lower = format(m$ci_lower, digits = 3),
            ci_upper = format(m$ci_upper, digits = 3),
            p_value = format(m$p_value, digits = 3),
            I2 = sprintf("%.1f%%", m$I2)
          )
        }))
        if (!is.null(ef_rows)) meta_summary <- rbind(meta_summary, ef_rows)
      }

      render_dt(meta_summary, caption = "Random-Effects Meta-Analysis (AT, DD, FB, MV)")
    }
  } else {
    cat("Insufficient cell types for meta-analysis (need >= 2).\n\n")
  }
} else {
  cat("metafor package not available; skipping meta-analysis.\n\n")
}
```

# 8. Key Figures

```{r sec9-figures, eval=has_ggplot2, fig.height=8}
if (has_ggplot2 && length(profiles_list) > 0) {

  # --- Figure 1: Event frequency by comparison type ---
  event_freq <- do.call(rbind, lapply(names(profiles_list), function(key) {
    p <- profiles_list[[key]]
    parts <- strsplit(key, "_")[[1]]
    s <- summarizeProfiles(p)
    data.frame(
      comparison = comp_label(parts[1]),
      cell_type = paste(parts[-1], collapse = "_"),
      event_type = names(s$event_counts),
      count = as.integer(s$event_counts),
      total = s$total_events
    )
  }))
  event_freq$pct <- 100 * event_freq$count / event_freq$total

  p1 <- ggplot(event_freq, aes(x = event_type, y = pct, fill = comparison)) +
    geom_bar(stat = "identity", position = "dodge") +
    facet_wrap(~cell_type, scales = "free_y") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = "Event Type Distribution by Comparison",
         subtitle = "Raw event counts (not pair-based percentages — see Section 4 tables)",
         x = "Event Type", y = "Percentage of Total Events",
         fill = "Comparison")
  print(p1)

}
```

# 9. Session Info

```{r session-info}
sessionInfo()
```
