---
title: "NMD Isoform Transitions: Publication Report (mashr Classification)"
subtitle: "Isopair Wrapper — Isocall Pipeline, mashr DIE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
library(Isopair)
library(dplyr)
library(ggplot2)
library(knitr)
library(DT)
source("visualization_functions.R")

# Helper: render interactive DT table (sortable, searchable)
render_dt <- function(df, caption = NULL, pageLength = 15, ...) {
  datatable(df, caption = caption, rownames = FALSE,
            options = list(pageLength = pageLength, scrollX = TRUE),
            ...)
}

# ---------------------------------------------------------------------------
# Publication theme and color palette
# ---------------------------------------------------------------------------
theme_pub <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
theme_set(theme_pub)

pal <- c(NMD = "#ef8a62", Control = "#67a9cf")


# Figure output
fig_dir <- "figures"
dir.create(fig_dir, showWarnings = FALSE)
table_dir <- "tables"
dir.create(table_dir, showWarnings = FALSE)

save_fig <- function(p, name, width = 8, height = 5) {
  ggsave(file.path(fig_dir, paste0(name, ".pdf")), p, width = width, height = height)
  ggsave(file.path(fig_dir, paste0(name, ".png")), p, width = width, height = height, dpi = 300)
  print(p)
}

# ---------------------------------------------------------------------------
# Helper functions (adapted from 03_nmd_analysis_mashr.Rmd)
# ---------------------------------------------------------------------------
ct_to_label <- function(ct) tolower(gsub("_", "", ct))

comp_label <- function(comp) {
  switch(toupper(comp), C2 = "NMD", C4 = "Control", toupper(comp))
}

evt_types <- c("Alt_TSS", "Alt_TES", "A5SS", "A3SS", "SE",
               "Missing_Internal", "IR", "IR_diff",
               "Partial_IR_5", "Partial_IR_3")

compute_event_freq <- function(profiles) {
  n_pairs <- nrow(profiles)
  n_pir5 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_5")))
  n_pir3 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_3")))
  counts <- c(sum(profiles$tss_changed), sum(profiles$tes_changed),
              sum(profiles$n_a5ss > 0), sum(profiles$n_a3ss > 0),
              sum(profiles$n_se > 0), sum(profiles$n_missing_internal > 0),
              sum(profiles$n_ir > 0), sum(profiles$n_ir_diff > 0),
              n_pir5, n_pir3)
  data.frame(event_type = evt_types,
             n_pairs_with_event = counts,
             pct_of_pairs = round(100 * counts / n_pairs, 1))
}

compute_ptc_rates_row <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  coding_ids <- cds$isoform_id[cds$coding_status == "coding"]
  p_c2 <- set_results[[c2_key]]$profiles
  cod_c2 <- p_c2[p_c2$reference_isoform_id %in% coding_ids &
                   p_c2$comparator_isoform_id %in% coding_ids, ]
  p_c4 <- set_results[[c4_key]]$profiles
  cod_c4 <- p_c4[p_c4$reference_isoform_id %in% coding_ids &
                   p_c4$comparator_isoform_id %in% coding_ids, ]
  if (nrow(cod_c2) == 0 || nrow(cod_c4) == 0) return(NULL)
  ptc_c2 <- sum(cod_c2$comparator_isoform_id %in% ptc$isoform_id[ptc$has_ptc == TRUE])
  ptc_c4 <- sum(cod_c4$comparator_isoform_id %in% ptc$isoform_id[ptc$has_ptc == TRUE])
  n_c2 <- nrow(cod_c2); n_c4 <- nrow(cod_c4)
  ft <- tryCatch({
    mat <- matrix(c(ptc_c2, n_c2 - ptc_c2, ptc_c4, n_c4 - ptc_c4), nrow = 2)
    fisher.test(mat)
  }, error = function(e) NULL)
  data.frame(
    cell_type = ct,
    n_coding_pairs_NMD = n_c2,
    n_coding_pairs_Ctrl = n_c4,
    NMD_comp_ptc_n = ptc_c2,
    NMD_comp_ptc_pct = round(100 * ptc_c2 / n_c2, 1),
    Ctrl_comp_ptc_n = ptc_c4,
    Ctrl_comp_ptc_pct = round(100 * ptc_c4 / n_c4, 1),
    fisher_OR = if (!is.null(ft)) round(ft$estimate, 2) else NA_real_,
    p_value = if (!is.null(ft)) ft$p.value else NA_real_
  )
}

compute_frame_category_rows <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  fc_c2 <- set_results[[c2_key]]$frame_comp
  fc_c4 <- set_results[[c4_key]]$frame_comp
  if (is.null(fc_c2) || is.null(fc_c4)) return(NULL)
  ps_c2 <- fc_c2$pair_summary
  ps_c4 <- fc_c4$pair_summary
  cats <- c("same_start_in_frame", "same_start_frameshift",
            "diff_start_same_frame", "diff_start_diff_frame",
            "diff_start_diff_frame_with_frameshift",
            "non_coding", "no_shared_cds")
  # Denominator: coding pairs only (exclude non_coding and no_shared_cds)
  n_coding_c2 <- sum(!ps_c2$frame_category %in% c("non_coding", "no_shared_cds"))
  n_coding_c4 <- sum(!ps_c4$frame_category %in% c("non_coding", "no_shared_cds"))
  do.call(rbind, lapply(cats, function(cat) {
    is_coding_cat <- !cat %in% c("non_coding", "no_shared_cds")
    data.frame(
      cell_type = ct,
      frame_category = cat,
      NMD_n = sum(ps_c2$frame_category == cat),
      NMD_pct = if (is_coding_cat) round(100 * sum(ps_c2$frame_category == cat) / n_coding_c2, 1) else NA_real_,
      Ctrl_n = sum(ps_c4$frame_category == cat),
      Ctrl_pct = if (is_coding_cat) round(100 * sum(ps_c4$frame_category == cat) / n_coding_c4, 1) else NA_real_
    )
  }))
}

compute_divergence_row <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  div_c2 <- set_results[[c2_key]]$divergence
  div_c4 <- set_results[[c4_key]]$divergence
  if (is.null(div_c2) || nrow(div_c2) == 0 || is.null(div_c4) || nrow(div_c4) == 0) return(NULL)
  data.frame(
    cell_type = ct,
    NMD_n = nrow(div_c2),
    NMD_median_pct_shared = round(median(div_c2$pct_shared, na.rm = TRUE), 1),
    Ctrl_n = nrow(div_c4),
    Ctrl_median_pct_shared = round(median(div_c4$pct_shared, na.rm = TRUE), 1)
  )
}
```

# 0. Data Loading and Gene-Matching

## Data Provenance Overview

All analyses in this report begin from **gene-matched isoform pair sets**. The
data pipeline is:

1. **01_prepare_data_mashr.R**: Loads the isocall count matrix
   (`nmd_isocall.count_matrix.txt`, 645,273 isoforms x 38 samples), excludes
   2 outlier samples (DO donor 029T: `Sample25_DO_029T_DMSO` and
   `Sample26_DO_029T_Smg1i`, identified as PCA outlier), applies TMM
   normalization and condition-stratified 5% isoform filtering, then classifies
   isoforms as NMD-sensitive (`nmd_responsive == TRUE` in mashr DIE results) or
   non-NMD (`adj.P.Val > 0.50`). Final dataset: 36 samples across 6 cell types.
   Outputs: `expression_data.rds`, `nmd_classification.rds`,
   `sample_metadata.rds`, `gene_map.rds`.

2. **02_build_profiles_mashr.R**: Generates isoform pairs —
   **C4 (Control)**: top-two non-NMD isoforms by DMSO CPM per gene
   (`generatePairsExpression(..., method = "top_two")`);
   **C2 (NMD)**: same C4 reference paired with top NMD isoform by Smg1i CPM
   (via `identifyDominantIsoforms()` + `inner_join` on `gene_id`).
   Builds splicing choice profiles with `Isopair::buildProfiles()`.
   Outputs: `profiles_c{2,4}_*.rds`, `pairs_c{2,4}_*.rds`.

3. **03_nmd_analysis_mashr.Rmd**: Applies **gene-matching** (restricts C2 and C4
   to genes with shared `(gene_id, reference_isoform_id)` in both sets), then
   runs Isopair analysis functions on the matched profiles. Caches results to
   `analysis_cache/`.

4. **This report (05)**: Loads profiles from step 2, re-applies gene-matching
   (step 3), loads cached analysis results, and verifies consistency. All
   downstream analyses use gene-matched profiles exclusively.

```{r flowchart-data, include=FALSE}
# ── Gather numbers for the flowchart ──
# These are computed fresh from the loaded data objects to ensure accuracy.
# Raw count matrix dimensions are hardcoded (not loaded here to save memory).
n_raw_isoforms <- 645273L
n_raw_samples  <- 38L
n_excluded     <- 2L
```

```{r flowchart, fig.width=10, fig.height=10}
# ── Data provenance ──
# All numbers derived from objects loaded in subsequent chunks.
# Raw count matrix dimensions from isocall pipeline output.
# Filtered/classified counts from 01_prepare_data_mashr.R outputs.
# Pair counts from 02_build_profiles_mashr.R outputs.
# Gene-matched counts computed in load-profiles chunk below.
#
# We load the key objects here just for the flowchart numbers, then
# the full load happens in the subsequent chunks.

fmt <- function(x) format(x, big.mark = ",")

# Load just what we need for counts
.fc_expr   <- readRDS(file.path("data_mashr", "expression_data.rds"))
.fc_sm     <- readRDS(file.path("data_mashr", "sample_metadata.rds"))
.fc_nmd    <- readRDS(file.path("data_mashr", "nmd_classification.rds"))
.fc_gm     <- readRDS(file.path("data_mashr", "gene_map.rds"))

n_filtered   <- nrow(.fc_expr)
n_samples    <- ncol(.fc_expr)
n_dmso       <- sum(.fc_sm$treatment == "DMSO")
n_smg1i      <- sum(.fc_sm$treatment == "Smg1i")
n_nmd_as     <- length(.fc_nmd[["all_samples"]]$nmd)
n_nonnmd_as  <- length(.fc_nmd[["all_samples"]]$non_nmd)

# How many genes have both NMD and non-NMD isoforms?
nmd_genes <- unique(.fc_gm$gene_id[.fc_gm$isoform_id %in% .fc_nmd[["all_samples"]]$nmd])
nonnmd_genes <- unique(.fc_gm$gene_id[.fc_gm$isoform_id %in% .fc_nmd[["all_samples"]]$non_nmd])
# Genes with ≥2 non-NMD (needed for C4)
nonnmd_per_gene <- table(.fc_gm$gene_id[.fc_gm$isoform_id %in% .fc_nmd[["all_samples"]]$non_nmd])
genes_2plus_nonnmd <- names(nonnmd_per_gene[nonnmd_per_gene >= 2])
# Genes eligible for both C2 and C4
genes_eligible <- intersect(nmd_genes, genes_2plus_nonnmd)
n_genes_eligible <- length(genes_eligible)

# Pair counts (all_samples, before gene-matching)
.fc_c2 <- readRDS(file.path("data_mashr", "profiles_c2_allsamples.rds"))
.fc_c4 <- readRDS(file.path("data_mashr", "profiles_c4_allsamples.rds"))
n_c2_before <- nrow(.fc_c2)
n_c4_before <- nrow(.fc_c4)

# Gene-matched counts
.shared <- dplyr::inner_join(
  unique(.fc_c2[, c("gene_id", "reference_isoform_id")]),
  unique(.fc_c4[, c("gene_id", "reference_isoform_id")]),
  by = c("gene_id", "reference_isoform_id"))
n_matched_genes <- nrow(.shared)
n_c2_matched <- nrow(dplyr::semi_join(.fc_c2, .shared, by = c("gene_id", "reference_isoform_id")))
n_c4_matched <- nrow(dplyr::semi_join(.fc_c4, .shared, by = c("gene_id", "reference_isoform_id")))

rm(.fc_expr, .fc_sm, .fc_nmd, .fc_gm, .fc_c2, .fc_c4, .shared)

dot_spec <- sprintf('
digraph flowchart {
  graph [rankdir=TB, fontname="Helvetica", nodesep=1.6, ranksep=1.4, size="13,16"]
  node  [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=23, margin="0.65,0.5", width=6.5]
  edge  [color="#666666", penwidth=2.5]

  A [label=<<b>Isocall Count Matrix</b><br/>%s isoforms × %d samples>,
     fillcolor="#f0f0f0", color="#636363"]

  B [label=<<b>Sample Exclusion</b><br/>Remove DO donor 029T (PCA outlier)<br/>%s isoforms × %d samples>,
     fillcolor="#fee0d2", color="#de2d26"]

  C [label=<<b>5%% Filter + filterByExpr</b><br/>%s isoforms pass<br/>%d DMSO + %d Smg1i | 6 cell types>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  D [label=<<b>NMD Classification (mashr)</b><br/>NMD-sensitive: %s isoforms (union across CTs)<br/>Non-NMD: %s isoforms (intersection across CTs)>,
     fillcolor="#fff7bc", color="#d95f0e"]

  D2 [label=<<b>Eligible Genes</b><br/>%s genes have ≥1 NMD isoform AND ≥2 non-NMD isoforms<br/>(required for both C2 and C4 pair construction)>,
      fillcolor="#fff7bc", color="#d95f0e"]

  E [label=<<b>Pair Construction (all_samples)</b><br/>C2 (NMD): %s pairs (1 per gene)<br/>C4 (Control): %s pairs (1 per gene)>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  F [label=<<b>Gene-Matching</b><br/>%s genes with shared (gene, reference) in both C2 and C4<br/>C2: %s pairs | C4: %s pairs>,
     fillcolor="#c7e9c0", color="#31a354"]

  A -> B -> C -> D -> D2 -> E -> F
}',
  fmt(n_raw_isoforms), n_raw_samples,
  fmt(n_raw_isoforms), n_samples,
  fmt(n_filtered), n_dmso, n_smg1i,
  fmt(n_nmd_as), fmt(n_nonnmd_as),
  fmt(n_genes_eligible),
  fmt(n_c2_before), fmt(n_c4_before),
  fmt(n_matched_genes), fmt(n_c2_matched), fmt(n_c4_matched)
)

if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(dot_spec, width = 1000, height = 700)
} else {
  message("DiagrammeR not installed -- flowchart not rendered")
}
```

```{r load-data}
# ==============================================================================
# STAGE 1: Load raw data objects
# ==============================================================================

# ── 01_prepare_data_mashr.R outputs ──
# Source: isocall count matrix → TMM normalization → 5% filter → mashr NMD classification
data_dir  <- "data_mashr"
cache_dir <- file.path(data_dir, "analysis_cache")

expr_mat    <- readRDS(file.path(data_dir, "expression_data.rds"))
sample_meta <- readRDS(file.path(data_dir, "sample_metadata.rds"))
gene_map    <- readRDS(file.path(data_dir, "gene_map.rds"))
nmd_class   <- readRDS(file.path(data_dir, "nmd_classification.rds"))

# ── Infrastructure objects (isoform-intrinsic, DE-independent) ──
# Originally from Isopair pipeline scripts 02-05, copied to data_mashr/ by
# 02_build_profiles_mashr.R for self-containment.
# structures.rds: extractIsoformStructures() — exon coordinates per isoform
# cds.rds: extractCds() — CDS boundaries per isoform
# ptc.rds: testPtcAssociation() infrastructure — PTC status per isoform
# union_exons.rds: buildUnionExons() — merged exon models per gene
# isoform_union_mapping.rds: mapIsoformsToUnion() — isoform-to-union-exon mapping
# annotated_ue.rds: annotateUnionExons() — union exons with CDS/UTR labels
# cds_exons.rds: CDS mapped to individual exons
structures   <- readRDS(file.path(data_dir, "structures.rds"))
cds          <- readRDS(file.path(data_dir, "cds.rds"))
ptc          <- readRDS(file.path(data_dir, "ptc.rds"))
union_exons  <- readRDS(file.path(data_dir, "union_exons.rds"))
ue_mapping   <- readRDS(file.path(data_dir, "isoform_union_mapping.rds"))
annotated_ue <- readRDS(file.path(data_dir, "annotated_ue.rds"))
cds_exons    <- readRDS(file.path(data_dir, "cds_exons.rds"))

# ── Gene symbol lookup from mashr DE files ──
de_dir_mashr <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/mashr"
gene_symbol_lookup <- character(0)
for (de_f in list.files(de_dir_mashr, pattern = "^nmd_mashr_die_.*\\.csv$", full.names = TRUE)) {
  de_tmp <- read.csv(de_f, stringsAsFactors = FALSE)[, c("gene_id", "hgnc_symbol")]
  de_tmp <- de_tmp[!duplicated(de_tmp$gene_id) & !is.na(de_tmp$hgnc_symbol) &
                    de_tmp$hgnc_symbol != "", ]
  new_entries <- setdiff(de_tmp$gene_id, names(gene_symbol_lookup))
  gene_symbol_lookup[de_tmp$gene_id[de_tmp$gene_id %in% new_entries]] <-
    de_tmp$hgnc_symbol[de_tmp$gene_id %in% new_entries]
}

get_gene_sym <- function(gid) {
  sym <- gene_symbol_lookup[gid]
  if (is.na(sym) || sym == "") gid else sym
}

# Cell types (mashr: DO included, DD_ALI excluded from main)
cell_types  <- c("all_samples", "AT", "DD", "DO", "FB", "MV")
comparisons <- c("c2", "c4")
main_cts    <- c("AT", "DD", "DO", "FB", "MV")
```

```{r load-profiles}
# ==============================================================================
# STAGE 2: Load profiles and apply gene-matching
# ==============================================================================
#
# profiles_c{2,4}_*.rds are the output of 02_build_profiles_mashr.R. C2 pairs
# share their reference isoform with C4 (by construction in script 02), but C4
# may contain genes that have no NMD isoform (i.e., genes with >=2 non-NMD
# isoforms but no mashr-classified NMD isoform). Gene-matching restricts both
# sets to the intersection of (gene_id, reference_isoform_id), ensuring fair
# NMD vs Control comparison on identical gene sets.

profiles_list <- list()
for (ct in cell_types) {
  ct_label <- ct_to_label(ct)
  for (comp in comparisons) {
    key <- paste(comp, ct_label, sep = "_")
    prof_file <- file.path(data_dir, sprintf("profiles_%s.rds", key))
    if (file.exists(prof_file)) profiles_list[[key]] <- readRDS(prof_file)
  }
}

# ── Gene-matching ──
# Replicates the logic from 03_nmd_analysis_mashr.Rmd lines 743-775.
# For each cell type, restrict C2 and C4 profiles to genes where both have
# a pair with the same (gene_id, reference_isoform_id).
gene_match_summary <- list()
for (ct in cell_types) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(profiles_list) || !c4_key %in% names(profiles_list)) next

  n_c2_before <- nrow(profiles_list[[c2_key]])
  n_c4_before <- nrow(profiles_list[[c4_key]])

  c2_gene_refs <- unique(profiles_list[[c2_key]][, c("gene_id", "reference_isoform_id")])
  c4_gene_refs <- unique(profiles_list[[c4_key]][, c("gene_id", "reference_isoform_id")])
  shared_refs  <- dplyr::inner_join(c2_gene_refs, c4_gene_refs,
                                     by = c("gene_id", "reference_isoform_id"))

  profiles_list[[c2_key]] <- dplyr::semi_join(profiles_list[[c2_key]], shared_refs,
                                               by = c("gene_id", "reference_isoform_id"))
  profiles_list[[c4_key]] <- dplyr::semi_join(profiles_list[[c4_key]], shared_refs,
                                               by = c("gene_id", "reference_isoform_id"))

  gene_match_summary[[ct]] <- data.frame(
    cell_type      = ct,
    c2_before      = n_c2_before,
    c4_before      = n_c4_before,
    matched_genes  = nrow(shared_refs),
    c2_after       = nrow(profiles_list[[c2_key]]),
    c4_after       = nrow(profiles_list[[c4_key]])
  )
}
gene_match_df <- do.call(rbind, gene_match_summary)
```

```{r build-set-results}
# ==============================================================================
# STAGE 3: Build set_results from gene-matched profiles + cached analyses
# ==============================================================================
#
# set_results[[key]] is the primary data structure for all downstream analyses.
# $profiles: gene-matched splicing choice profiles (from STAGE 2 above)
# $divergence: quantifyPairDivergence() output (cached by script 03)
# $frame_comp: compareIsoformFrames() output (cached by script 03)
# $frame_walk: analyzeFrameWalk() output (cached by script 03)
# $ptc_assoc: testPtcAssociation() output (cached by script 03)
# $evt_freq: computed here from gene-matched profiles

set_results <- list()
for (key in names(profiles_list)) {
  profiles <- profiles_list[[key]]
  set_results[[key]] <- list(
    profiles   = profiles,
    divergence = tryCatch(readRDS(file.path(cache_dir, paste0("div_", key, ".rds"))), error = function(e) NULL),
    frame_comp = tryCatch(readRDS(file.path(cache_dir, paste0("fc_",  key, ".rds"))), error = function(e) NULL),
    frame_walk = tryCatch(readRDS(file.path(cache_dir, paste0("fw_",  key, ".rds"))), error = function(e) NULL),
    ptc_assoc  = tryCatch(readRDS(file.path(cache_dir, paste0("ptc_", key, ".rds"))), error = function(e) NULL),
    evt_freq   = compute_event_freq(profiles)
  )
}
```

```{r verify-cache-consistency}
# ==============================================================================
# STAGE 4: Verify cached analysis objects match gene-matched profiles
# ==============================================================================
#
# The cached div/fc/fw/ptc objects were computed by script 03 from gene-matched
# profiles. We verify that the row counts in the cached objects match our
# gene-matched profiles_list. A mismatch would indicate the cache is stale or
# was computed from a different pair set.

verification_rows <- list()
for (key in names(set_results)) {
  sr <- set_results[[key]]
  n_profiles <- nrow(sr$profiles)

  # divergence: 1 row per pair (should exactly match profiles)
  n_div <- if (!is.null(sr$divergence)) nrow(sr$divergence) else NA
  # frame_comp$pair_summary: 1 row per pair (should exactly match)
  n_fc <- if (!is.null(sr$frame_comp)) nrow(sr$frame_comp$pair_summary) else NA
  # frame_walk$profile_summary: 1 row per pair (should exactly match)
  n_fw <- if (!is.null(sr$frame_walk)) nrow(sr$frame_walk$profile_summary) else NA
  # ptc_assoc$pair_summary: 1 row per pair where PTC analysis was possible
  #   (both isoforms coding with EJC data). Expected to be <= n_profiles.
  n_ptc <- if (!is.null(sr$ptc_assoc)) nrow(sr$ptc_assoc$pair_summary) else NA

  # Exact match expected for div, fc, fw
  div_ok <- is.na(n_div) || n_div == n_profiles
  fc_ok  <- is.na(n_fc)  || n_fc  == n_profiles
  fw_ok  <- is.na(n_fw)  || n_fw  == n_profiles
  # ptc: subset is OK (non-coding pairs excluded), but should not exceed profiles
  ptc_ok <- is.na(n_ptc) || n_ptc <= n_profiles

  verification_rows[[key]] <- data.frame(
    key = key, n_profiles = n_profiles,
    n_div = n_div, n_fc = n_fc, n_fw = n_fw, n_ptc = n_ptc,
    exact_match = div_ok & fc_ok & fw_ok,
    ptc_subset_ok = ptc_ok
  )
}
verification_df <- do.call(rbind, verification_rows)

if (!all(verification_df$exact_match)) {
  warning("CACHE MISMATCH DETECTED — div/fc/fw row counts do not match ",
          "gene-matched profiles. Affected keys: ",
          paste(verification_df$key[!verification_df$exact_match], collapse = ", "))
}
if (!all(verification_df$ptc_subset_ok)) {
  warning("PTC CACHE ANOMALY — ptc_assoc has MORE rows than profiles for: ",
          paste(verification_df$key[!verification_df$ptc_subset_ok], collapse = ", "))
}
```

```{r load-meta-and-precomputes}
# ==============================================================================
# STAGE 5: Load mashr meta-analysis logFC and precomputed objects
# ==============================================================================

# ── mashr meta-analysis logFC ──
# Source: mashr model fit on isocall DIE (6 conditions).
# PosteriorMean averaged across 5 main conditions (excluding DD_ALI).
# Used in Goal 1 to characterize NMD response magnitude.
mashr_model_path <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/mashr/mashr_isoform_model_2026.3.10.rds"
mashr_model <- readRDS(mashr_model_path)
post_mean <- ashr::get_pm(mashr_model)
meta_cols <- c("Smg1i_in_DD", "Smg1i_in_AT", "Smg1i_in_DO_ALI",
               "Smg1i_in_FB", "Smg1i_in_MV")
meta_logfc <- data.frame(
  txid = rownames(post_mean),
  mean_logFC = rowMeans(post_mean[, meta_cols, drop = FALSE])
)

# ── C4 protein comparison (gene-matched, used in productive frameshifts) ──
# Source: 04b_all_c4_protein_comparison.R
# Note: Goal 5 now uses unmatched data from 05d_protein_domain_analysis.R
c4_protein <- readRDS(file.path(cache_dir, "all_c4_protein_comparison_allsamples.rds"))

# ── Productive frameshift precompute ──
# Source: 04_productive_frameshift_precompute.R
# Input: fc_c4_allsamples.rds + fw_c4_allsamples.rds (gene-matched, from script 03)
#        + SQANTI protein FASTA + EBI/biomaRt API queries
# Output: characterization of same-start frameshift C4 pairs without PTC
pf_data <- readRDS(file.path(cache_dir, "productive_frameshift_allsamples.rds"))
```

```{r data-dimensions}
# ==============================================================================
# Summary: Gene-matching and data dimensions
# ==============================================================================
cat("=== Gene-Matching Summary ===\n")
render_dt(gene_match_df, caption = "Gene-Matching: Before and After")

cat(sprintf("\n=== Data Dimensions ===\n"))
cat(sprintf("Gene-matched profile keys: %d\n", length(profiles_list)))
cat(sprintf("C2 all_samples pairs (gene-matched): %d\n", nrow(profiles_list[["c2_allsamples"]])))
cat(sprintf("C4 all_samples pairs (gene-matched): %d\n", nrow(profiles_list[["c4_allsamples"]])))
cat(sprintf("PTC records: %d isoforms\n", nrow(ptc)))
cat(sprintf("CDS records: %d isoforms\n", nrow(cds)))
cat(sprintf("Meta-analysis logFC: %d isoforms\n", nrow(meta_logfc)))
cat(sprintf("C4 protein comparison (gene-matched): %d pairs\n", nrow(c4_protein$comparison)))
cat(sprintf("Productive frameshifts: %d pairs\n", nrow(pf_data$characterization)))

cat("\n=== Cache Verification ===\n")
render_dt(verification_df, caption = "Cache Consistency Check")
```


# Event Type Glossary {.tabset}

Isopair detects 12 structural event types organized into 4 categories. Each tab below
shows a schematic diagram of the event alongside its definition. In all diagrams, the
**Reference** (top) is the dominant non-NMD isoform and the **Comparator** (bottom) is
the isoform being compared. Events are always described from the comparator's perspective:
a **GAIN** means the comparator has additional sequence relative to the reference, and a
**LOSS** means the comparator is missing sequence present in the reference.

**Corrections from the exploratory report (03):** A5SS and A3SS schematics were swapped;
IR_diff_5 and IR_diff_3 schematics showed the retaining exon extending beyond the reference
when the code classifies "diff" when the retaining exon falls short.

```{r glossary, results='asis', fig.width=8, fig.height=3}
glossary_plot <- function(ref_s, ref_e, comp_s, comp_e, title, subtitle,
                          ref_cds = NULL, comp_cds = NULL, events = NULL) {
  plot_ucsc_pair(ref_s, ref_e, comp_s, comp_e, strand = "+",
                 ref_cds = ref_cds, comp_cds = comp_cds,
                 ref_label = "Reference", comp_label = "Comparator",
                 events = events, title = title, subtitle = subtitle)
}

# ---- Terminal Events ----

cat("## Alt_TSS\n\n")
cat("**Alternative Transcription Start Site.** The two isoforms begin at different\n")
cat("genomic positions. Detected when the 5'-most exon of one isoform does not overlap\n")
cat("the 5'-most exon of the other. Direction: GAIN if the comparator's TSS is further\n")
cat("upstream (extending the transcript); LOSS if it is further downstream (truncating it).\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(250, 400, 700), comp_e = c(350, 550, 850),
  title = "Alt_TSS: Alternative Transcription Start Site",
  subtitle = "Comparator starts at a different position (here, downstream = LOSS)")
print(p); cat("\n\n")

cat("## Alt_TES\n\n")
cat("**Alternative Transcription End Site.** The two isoforms end at different\n")
cat("genomic positions. Detected when the 3'-most exon of one isoform does not overlap\n")
cat("the 3'-most exon of the other. Direction: GAIN if the comparator extends further;\n")
cat("LOSS if it ends earlier.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 900),
  comp_s = c(100, 400, 700), comp_e = c(200, 550, 800),
  title = "Alt_TES: Alternative Transcription End Site",
  subtitle = "Comparator ends earlier (LOSS)")
print(p); cat("\n\n")

# ---- Boundary Events ----

cat("## A5SS\n\n")
cat("**Alternative 5' Splice Site.** An internal exon has a different **donor** (exon_end\n")
cat("on + strand) boundary while sharing the same **acceptor** (exon_start) boundary.\n")
cat("The exon is extended or truncated at its downstream/3' end.\n")
cat("Direction: GAIN if the comparator's exon is longer; LOSS if shorter.\n\n")
# A5SS: same acceptor (exon_start=400), different donor (exon_end: 550 vs 600)
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400, 700), comp_e = c(200, 600, 850),
  title = "A5SS: Alternative 5' Splice Site",
  subtitle = "Same acceptor (start), different donor (end). Comparator exon extends further (GAIN)")
print(p); cat("\n\n")

cat("## A3SS\n\n")
cat("**Alternative 3' Splice Site.** An internal exon has a different **acceptor** (exon_start\n")
cat("on + strand) boundary while sharing the same **donor** (exon_end) boundary.\n")
cat("The exon is extended or truncated at its upstream/5' end.\n")
cat("Direction: GAIN if the comparator's exon is longer; LOSS if shorter.\n\n")
# A3SS: different acceptor (exon_start: 400 vs 350), same donor (exon_end=550)
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 350, 700), comp_e = c(200, 550, 850),
  title = "A3SS: Alternative 3' Splice Site",
  subtitle = "Same donor (end), different acceptor (start). Comparator exon extends further (GAIN)")
print(p); cat("\n\n")

cat("## Partial_IR_5\n\n")
cat("**Partial Intron Retention (5' end).** An intron is partially retained: one isoform's\n")
cat("exon extends into the intron from its 5' side, covering what is intronic sequence in\n")
cat("the other isoform. The 3' boundary of the retaining exon falls within the intron.\n")
cat("Direction: GAIN if the comparator retains; LOSS if the reference retains.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400, 700), comp_e = c(200, 620, 850),
  title = "Partial_IR_5: Partial Intron Retention (5' side)",
  subtitle = "Comparator exon extends into the intron from the 5' end (GAIN)")
print(p); cat("\n\n")

cat("## Partial_IR_3\n\n")
cat("**Partial Intron Retention (3' end).** An intron is partially retained: one isoform's\n")
cat("exon extends into the intron from its 3' side, covering what is intronic sequence in\n")
cat("the other isoform. The 5' boundary of the retaining exon falls within the intron.\n")
cat("Direction: GAIN if the comparator retains; LOSS if the reference retains.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400, 640), comp_e = c(200, 550, 850),
  title = "Partial_IR_3: Partial Intron Retention (3' side)",
  subtitle = "Comparator exon extends into the intron from the 3' end (GAIN)")
print(p); cat("\n\n")

# ---- Exon-Level Events ----

cat("## SE (Skipped Exon)\n\n")
cat("**Skipped Exon.** An exon present in one isoform is completely absent from the\n")
cat("other. The flanking exons splice directly to each other, skipping over the exon.\n")
cat("Direction: GAIN if the comparator includes the exon (reference skips it);\n")
cat("LOSS if the comparator skips it (reference includes it).\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 600, 800), ref_e  = c(200, 500, 700, 900),
  comp_s = c(100, 400, 800),       comp_e = c(200, 500, 900),
  title = "SE: Skipped Exon",
  subtitle = "Comparator skips an internal exon present in the reference (LOSS)")
print(p); cat("\n\n")

cat("## Missing_Internal\n\n")
cat("**Missing Internal Exon.** Similar to SE but arising from a multi-exon skip:\n")
cat("one isoform is missing one or more internal exons that are present in the other.\n")
cat("Often co-occurs with other Missing_Internal events from the same splice decision.\n")
cat("Direction: GAIN if the comparator has the exon; LOSS if the comparator lacks it.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 350, 550, 750, 950), ref_e  = c(200, 450, 650, 850, 1050),
  comp_s = c(100, 950),                 comp_e = c(200, 1050),
  title = "Missing_Internal: Missing Internal Exon(s)",
  subtitle = "Comparator skips multiple internal exons (LOSS x 3)")
print(p); cat("\n\n")

# ---- Intron Retention Events ----

cat("## IR\n\n")
cat("**Full Intron Retention.** One isoform has a single exon that spans the region\n")
cat("covered by two or more exons (plus the intervening intron) in the other isoform.\n")
cat("The retaining exon boundaries match exactly. This is a full retention where the\n")
cat("entire intron is included in the retaining isoform's exon.\n")
cat("Direction: GAIN if the comparator retains the intron; LOSS if the reference retains it.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400),       comp_e = c(200, 850),
  title = "IR: Full Intron Retention",
  subtitle = "Comparator retains intron -- single exon spans two exons + intron in ref (GAIN)")
print(p); cat("\n\n")

cat("## IR_diff_5\n\n")
cat("**Intron Retention with 5' Boundary Difference.** Like IR, but the retaining\n")
cat("exon's 5' boundary differs from the first overlapping exon in the other isoform.\n")
cat("The intron is retained, but the retaining exon starts after the first reference\n")
cat("overlap exon (falls short at the 5' end).\n\n")
# IR_diff_5: comp retaining exon starts AFTER first ref overlap exon start
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 430),       comp_e = c(200, 850),
  title = "IR_diff_5: Intron Retention with 5' Boundary Difference",
  subtitle = "Retaining exon starts AFTER ref overlap exon (falls short at 5' end)")
print(p); cat("\n\n")

cat("## IR_diff_3\n\n")
cat("**Intron Retention with 3' Boundary Difference.** Like IR, but the retaining\n")
cat("exon's 3' boundary differs from the last overlapping exon in the other isoform.\n")
cat("The intron is retained, but the retaining exon ends before the last reference\n")
cat("overlap exon (falls short at the 3' end).\n\n")
# IR_diff_3: comp retaining exon ends BEFORE last ref overlap exon end
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400),       comp_e = c(200, 820),
  title = "IR_diff_3: Intron Retention with 3' Boundary Difference",
  subtitle = "Retaining exon ends BEFORE ref overlap exon (falls short at 3' end)")
print(p); cat("\n\n")

cat("## IR_diff_5_3\n\n")
cat("**Intron Retention with Both Boundary Differences.** Like IR, but both the 5'\n")
cat("and 3' boundaries of the retaining exon differ from the flanking exons in the\n")
cat("other isoform. The intron is retained but the retaining exon falls short at\n")
cat("both ends.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 430),       comp_e = c(200, 820),
  title = "IR_diff_5_3: Intron Retention with Both Boundary Differences",
  subtitle = "Retaining exon falls short at both 5' and 3' ends")
print(p); cat("\n\n")
```

In the IR section of this report, **IR_diff** refers to the aggregate of IR_diff_5,
IR_diff_3, and IR_diff_5_3 events.


# 1. PTC Enrichment Validates NMD Classification {#goal1}

**Goal 1:** Demonstrate that NMD-responsive comparator isoforms are enriched for premature
termination codons (PTCs), validating the mashr-based NMD classification.

**Methods:** A PTC is defined as a stop codon >50 nt upstream of the last exon-junction
complex (EJC). Fisher's exact test compares PTC rates between NMD (C2) and Control (C4)
comparator isoforms, restricted to pairs where both reference and comparator are coding.
PTC position metrics (distance to EJC, % CDS, number of downstream EJCs) characterize NMD
response strength using mashr meta-analysis posterior means averaged across 5 conditions
(DD, AT, DO_ALI, FB, MV).

## PTC Rates

```{r goal1-ptc-rates-subsetting}
# ── Data provenance ──
# Starting data: gene-matched profiles (profiles_list → set_results$profiles)
#   Pipeline: 01 (mashr NMD classification) → 02 (pair generation + buildProfiles)
#   → gene-matching (this report, STAGE 2) → set_results (STAGE 3)
#
# Subsetting for PTC rate calculation:
#   1. Start with gene-matched C2 (NMD) and C4 (Control) profiles
#   2. Restrict to "coding pairs": both reference AND comparator isoform must have
#      coding_status == "coding" in cds.rds (from Isopair extractCds() on SQANTI CDS)
#   3. Numerator: comparator isoforms with has_ptc == TRUE in ptc.rds
#      PTC definition: stop codon >50 nt upstream of the last exon-junction complex (EJC)
#   4. Denominator: total coding pairs (from step 2)
#   5. Fisher's exact test: 2×2 table of (PTC+/PTC−) × (NMD/Control)

# Show the subsetting explicitly
coding_ids <- cds$isoform_id[cds$coding_status == "coding"]
n_gm_c2 <- nrow(set_results[["c2_allsamples"]]$profiles)
n_gm_c4 <- nrow(set_results[["c4_allsamples"]]$profiles)

p_c2 <- set_results[["c2_allsamples"]]$profiles
cod_c2 <- p_c2[p_c2$reference_isoform_id %in% coding_ids &
                p_c2$comparator_isoform_id %in% coding_ids, ]
p_c4 <- set_results[["c4_allsamples"]]$profiles
cod_c4 <- p_c4[p_c4$reference_isoform_id %in% coding_ids &
                p_c4$comparator_isoform_id %in% coding_ids, ]
```

Starting from `r fmt(n_gm_c2)` gene-matched NMD (C2) pairs and
`r fmt(n_gm_c4)` gene-matched Control (C4) pairs, we restrict to **coding
pairs** where both the reference and comparator isoform are classified as
protein-coding by SQANTI (`coding_status == "coding"` in `cds.rds`). This
yields `r fmt(nrow(cod_c2))` coding NMD pairs and `r fmt(nrow(cod_c4))` coding
Control pairs. The PTC rate is the fraction of these coding comparator isoforms
whose stop codon falls >50 nt upstream of the last exon-junction complex (EJC),
as computed by `Isopair::testPtcAssociation()`.

```{r goal1-ptc-rates}
ptc_rates <- compute_ptc_rates_row("all_samples")
kable(ptc_rates, caption = "Table 1: PTC Rates — all_samples",
      digits = c(0, 0, 0, 0, 1, 0, 1, 2, 2))
write.csv(ptc_rates, file.path(table_dir, "table1_ptc_rates_allsamples.csv"), row.names = FALSE)
```

Columns: **cell_type** = sample set; **n_coding_pairs_NMD/Ctrl** = number of
coding pairs (both ref and comp coding) in NMD (C2) / Control (C4);
**NMD_comp_ptc_n/pct** = count and percentage of NMD comparators with PTC;
**Ctrl_comp_ptc_n/pct** = same for Control; **fisher_OR** = odds ratio from
Fisher's exact test; **p_value** = Fisher p-value.

## Figure 1: Stop Codon Distance to Last EJC

```{r goal1-fig1-stop-dist-pop}
# ── Population derivation ──
# Start from gene-matched comparators, filter to those with non-NA ptc_distance.
# This is broader than Table 1 (not restricted to coding pairs).
fig1_pop <- list()
for (comp in c("c2", "c4")) {
  key <- paste0(comp, "_allsamples")
  comp_ids <- set_results[[key]]$profiles$comparator_isoform_id
  n_total <- length(comp_ids)
  ptc_sub <- ptc[ptc$isoform_id %in% comp_ids & !is.na(ptc$ptc_distance), ]
  fig1_pop[[comp]] <- list(n_total = n_total, n_with_dist = nrow(ptc_sub))
}
```

Starting from `r fmt(fig1_pop$c2$n_total)` gene-matched NMD comparators,
`r fmt(fig1_pop$c2$n_with_dist)` have a non-NA `ptc_distance` in `ptc.rds`
(i.e., have an annotated CDS with a stop codon and ≥1 exon-junction complex).
For Control: `r fmt(fig1_pop$c4$n_with_dist)` of `r fmt(fig1_pop$c4$n_total)`.
This is a broader set than Table 1, which additionally requires **both**
reference and comparator to be coding (pair-level filter). Here only the
comparator needs `ptc_distance` (isoform-level filter).

```{r goal1-fig1-stop-dist, fig.width=8, fig.height=5}
stop_dist_all <- {
  rows <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_allsamples")
    comp_ids <- set_results[[key]]$profiles$comparator_isoform_id
    ptc_sub <- ptc[ptc$isoform_id %in% comp_ids & !is.na(ptc$ptc_distance), ]
    if (nrow(ptc_sub) == 0) next
    rows[[comp]] <- data.frame(
      comparison = comp_label(comp),
      ptc_distance = ptc_sub$ptc_distance,
      has_ptc = ptc_sub$has_ptc)
  }
  do.call(rbind, rows)
}

fig1 <- ggplot(stop_dist_all, aes(x = ptc_distance, fill = comparison)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
  annotate("text", x = 50, y = Inf, label = "PTC threshold (50 nt)",
           hjust = -0.05, vjust = 1.5, size = 3, color = "red") +
  scale_x_continuous(limits = c(
    quantile(stop_dist_all$ptc_distance, 0.01, na.rm = TRUE),
    quantile(stop_dist_all$ptc_distance, 0.99, na.rm = TRUE))) +
  scale_fill_manual(values = pal) +
  labs(title = "Stop Codon Distance to Last EJC: NMD vs Control Comparators",
       subtitle = "All comparator isoforms with annotated stop codon + EJC. Dashed red = 50 nt PTC threshold.",
       x = "Distance from stop codon to last EJC (nt)", y = "Density", fill = "")
save_fig(fig1, "fig1_stop_codon_distance")
```

## Figure 2: NMD Response vs Stop Codon Distance

```{r goal1-fig2-logfc-dist, fig.width=8, fig.height=5}
# ── Data provenance ──
# Starting data: gene-matched C2 all_samples profiles
# Step 1: Extract comparator_isoform_id from gene-matched C2 profiles
# Step 2: Look up ptc features (ptc_distance, has_ptc, orf_length, n_downstream_ejcs)
#   from ptc.rds for those comparators
# Step 3: Merge with meta_logfc — mashr PosteriorMean averaged across 5 conditions
#   (DD, AT, DO_ALI, FB, MV). Source: mashr_isoform_model_2026.3.10.rds
# Result: scatter of NMD response magnitude vs stop codon position

comp_ids_as <- set_results[["c2_allsamples"]]$profiles$comparator_isoform_id
ptc_sub <- ptc[ptc$isoform_id %in% comp_ids_as & !is.na(ptc$ptc_distance),
                c("isoform_id", "ptc_distance", "has_ptc", "orf_length", "n_downstream_ejcs")]
n_with_logfc <- sum(comp_ids_as %in% meta_logfc$txid)
nmd_logfc_dist <- merge(meta_logfc[meta_logfc$txid %in% comp_ids_as, ],
                         ptc_sub, by.x = "txid", by.y = "isoform_id")
```

Of `r fmt(length(comp_ids_as))` gene-matched NMD comparators,
`r fmt(nrow(ptc_sub))` have non-NA `ptc_distance` and
`r fmt(n_with_logfc)` have a mashr posterior mean logFC. After merging
(requiring both), `r fmt(nrow(nmd_logfc_dist))` comparators remain.
`r if (length(comp_ids_as) - n_with_logfc > 0) paste0("The ", fmt(length(comp_ids_as) - n_with_logfc), " comparators absent from the mashr model are novel isoforms not tested for differential expression.") else "All comparators have mashr logFC values."`

```{r goal1-fig2-logfc-dist-plot, fig.width=8, fig.height=5}
fig2 <- ggplot(nmd_logfc_dist, aes(x = ptc_distance, y = mean_logFC)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "loess", se = TRUE, color = "steelblue") +
  geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(limits = c(
    quantile(nmd_logfc_dist$ptc_distance, 0.01, na.rm = TRUE),
    quantile(nmd_logfc_dist$ptc_distance, 0.99, na.rm = TRUE))) +
  labs(title = sprintf("NMD Response vs Stop Codon Distance (n = %d)", nrow(nmd_logfc_dist)),
       subtitle = "mashr meta-analysis mean logFC (5 conditions). Blue = loess.",
       x = "Distance from stop codon to last EJC (nt)",
       y = "Mean logFC (mashr)")
save_fig(fig2, "fig2_nmd_response_vs_distance")

# Correlation summary
cor_test <- cor.test(nmd_logfc_dist$ptc_distance, nmd_logfc_dist$mean_logFC,
                      method = "spearman")
kable(data.frame(n = nrow(nmd_logfc_dist),
                 spearman_rho = round(cor_test$estimate, 3),
                 p_value = cor_test$p.value),
      caption = "Spearman correlation: Stop-to-EJC distance vs mean logFC")
```

## Figure 3: NMD Strength by Downstream EJC Count

```{r goal1-fig3-pop}
as_ptc_pos <- nmd_logfc_dist[nmd_logfc_dist$has_ptc == TRUE, ]
as_ptc_pos$logFC <- as_ptc_pos$mean_logFC
```

Restricted to the `r fmt(nrow(as_ptc_pos))` **PTC-positive** comparators from
Figure 2's population of `r fmt(nrow(nmd_logfc_dist))` (`has_ptc == TRUE`),
binned by number of downstream EJCs.

```{r goal1-fig3-ejc-boxplot, fig.width=9, fig.height=5}
# Bin EJC counts >= 7 into "7+"
as_ptc_pos$ejc_bin <- ifelse(as_ptc_pos$n_downstream_ejcs >= 7,
                              "7+", as.character(as_ptc_pos$n_downstream_ejcs))
as_ptc_pos$ejc_bin <- factor(as_ptc_pos$ejc_bin,
                              levels = c(as.character(1:6), "7+"))

ejc_counts <- as_ptc_pos |>
  count(ejc_bin) |>
  mutate(label = paste0("n=", n))

fig3 <- ggplot(as_ptc_pos, aes(x = ejc_bin, y = logFC)) +
  geom_violin(fill = "lightblue", alpha = 0.5) +
  geom_boxplot(width = 0.15, outlier.alpha = 0.2) +
  geom_text(data = ejc_counts,
            aes(x = ejc_bin,
                y = max(as_ptc_pos$logFC, na.rm = TRUE) + 0.3, label = label),
            size = 2.8) +
  labs(title = "NMD Comparators: Mean logFC by Number of Downstream EJCs",
       subtitle = "mashr meta-analysis. logFC rises sharply from 1-4 EJCs, then plateaus.",
       x = "Number of Downstream EJCs", y = "Mean logFC (mashr)")
save_fig(fig3, "fig3_ejc_vs_nmd_strength", width = 9)
```

## Table 2: PTC Rates Across Cell Types

```{r goal1-table2-ct-ptc}
# ── Data provenance ──
# Same pipeline as Table 1 (compute_ptc_rates_row), iterated across main_cts.
# Each cell type uses its own gene-matched profiles from set_results.
ptc_ct <- do.call(rbind, lapply(c("all_samples", main_cts), compute_ptc_rates_row))
render_dt(ptc_ct, caption = "Table 2: PTC Rates Across Cell Types")
write.csv(ptc_ct, file.path(table_dir, "table2_ptc_rates_by_ct.csv"), row.names = FALSE)
```

Same columns as Table 1, one row per cell type. all_samples uses the union
(NMD) and intersection (non-NMD) across AT, DD, DO, FB, MV.


# 2. Splicing Mechanisms Producing PTCs {#goal2}

**Goal 2:** Identify the splicing mechanisms that produce PTCs in NMD-responsive isoforms.

**Methods:** Frame comparison (`compareIsoformFrames()`) classifies each pair by start codon
and reading frame relationship. Frame walk (`analyzeFrameWalk()`) identifies which splicing
events cause frameshifts. The Sankey diagram traces flows from splice event type through frame
category to PTC outcome for NMD comparators.

## Figure 4: Frame Category Distribution

```{r goal2-fig4-pop}
# Count excluded categories for prose
fc_c2_ps <- set_results[["c2_allsamples"]]$frame_comp$pair_summary
fc_c4_ps <- set_results[["c4_allsamples"]]$frame_comp$pair_summary
n_noncoding_c2 <- sum(fc_c2_ps$frame_category %in% c("non_coding", "no_shared_cds"))
n_noncoding_c4 <- sum(fc_c4_ps$frame_category %in% c("non_coding", "no_shared_cds"))
```

`compareIsoformFrames()` assigns each pair a frame category based on start codon
and reading frame relationship. Of `r fmt(nrow(fc_c2_ps))` NMD pairs,
`r fmt(n_noncoding_c2)` are classified as `non_coding` or `no_shared_cds` and
excluded from the figure (Control: `r fmt(n_noncoding_c4)` of
`r fmt(nrow(fc_c4_ps))` excluded). The denominator (y-axis) is the number of
**coding pairs** (`r fmt(nrow(fc_c2_ps) - n_noncoding_c2)` NMD,
`r fmt(nrow(fc_c4_ps) - n_noncoding_c4)` Control), so the 5 categories sum
to 100%.

```{r goal2-fig4-frame-category, fig.width=10, fig.height=6}
# ── Data provenance ──
# Starting data: gene-matched profiles → cached frame comparison
# set_results[[key]]$frame_comp: compareIsoformFrames() output, cached by script 03
#   Input to compareIsoformFrames(): gene-matched profiles + cds_exons.rds + cds.rds
#   $pair_summary contains frame_category per pair (same_start_in_frame, same_start_frameshift, etc.)
# ptc.rds$has_ptc: PTC status per isoform (isoform-intrinsic, independent of pairing)
# Denominator (total_pairs): nrow(gene-matched profiles) — ensures consistent
#   percentages between profiles and frame_comp (verified in STAGE 4)

coding_cats <- c("same_start_in_frame", "same_start_frameshift",
                 "diff_start_same_frame", "diff_start_diff_frame",
                 "diff_start_diff_frame_with_frameshift")
cat_labels <- c(
  same_start_in_frame = "Same Start\nSame Frame",
  same_start_frameshift = "Same Start\nFrameshift",
  diff_start_same_frame = "Diff Start\nSame Frame",
  diff_start_diff_frame = "Diff Start\nDiff Frame",
  diff_start_diff_frame_with_frameshift = "Diff Start\nDiff Frame + FS")

ptc_pos_ids <- ptc$isoform_id[ptc$has_ptc == TRUE]

# 4-color palette: NMD/Control × PTC+/No PTC
pal4 <- c("NMD PTC+"    = "#d6604d",   # dark orange-red
          "NMD No PTC"  = "#fddbc7",   # light orange
          "Ctrl PTC+"   = "#4393c3",   # dark blue
          "Ctrl No PTC" = "#d1e5f0")   # light blue

stacked_df <- do.call(rbind, lapply(c("c2", "c4"), function(comp) {
  key <- paste0(comp, "_allsamples")
  fc <- set_results[[key]]$frame_comp
  if (is.null(fc)) return(NULL)
  ps <- fc$pair_summary
  # Denominator: coding pairs only (exclude non_coding and no_shared_cds)
  coding_pairs_total <- sum(!ps$frame_category %in% c("non_coding", "no_shared_cds"))
  comp_lbl <- comp_label(comp)
  comp_short <- ifelse(comp_lbl == "NMD", "NMD", "Ctrl")

  do.call(rbind, lapply(coding_cats, function(cat) {
    cat_pairs <- ps[ps$frame_category == cat, ]
    n_cat <- nrow(cat_pairs)
    n_ptc <- sum(cat_pairs$comparator_isoform_id %in% ptc_pos_ids)
    n_no_ptc <- n_cat - n_ptc
    pct_ptc <- 100 * n_ptc / max(coding_pairs_total, 1)
    pct_no_ptc <- 100 * n_no_ptc / max(coding_pairs_total, 1)
    rbind(
      data.frame(frame_category = cat, category_label = cat_labels[cat],
                 comparison = comp_lbl, ptc_status = "PTC+",
                 fill_group = paste(comp_short, "PTC+"),
                 n = n_ptc, pct = pct_ptc),
      data.frame(frame_category = cat, category_label = cat_labels[cat],
                 comparison = comp_lbl, ptc_status = "No PTC",
                 fill_group = paste(comp_short, "No PTC"),
                 n = n_no_ptc, pct = pct_no_ptc))
  }))
}))

stacked_df$category_label <- factor(stacked_df$category_label,
  levels = cat_labels[coding_cats])
stacked_df$fill_group <- factor(stacked_df$fill_group,
  levels = c("NMD PTC+", "NMD No PTC", "Ctrl PTC+", "Ctrl No PTC"))

fig4 <- ggplot(stacked_df, aes(x = category_label, y = pct, fill = fill_group)) +
  geom_col() +
  facet_wrap(~comparison) +
  scale_fill_manual(values = pal4) +
  labs(title = "Frame Category Distribution: NMD vs Control (all_samples)",
       subtitle = "% of total pairs. PTC+ = comparator has premature termination codon.",
       x = "Frame Category", y = "% of Total Pairs", fill = "") +
  theme(axis.text.x = element_text(size = 8))
save_fig(fig4, "fig4_frame_category_distribution", width = 10, height = 6)
```

## Table 3: Frame Category Counts

```{r goal2-table3-frame-cats}
fc_rows <- compute_frame_category_rows("all_samples")
render_dt(fc_rows, caption = "Table 3: Frame Category Distribution — all_samples")
write.csv(fc_rows, file.path(table_dir, "table3_frame_categories.csv"), row.names = FALSE)
```

Columns: **cell_type**; **frame_category** = reading frame relationship between
reference and comparator (from `Isopair::compareIsoformFrames()`);
**NMD_n/NMD_pct** = count and percentage in NMD (C2) pairs;
**Ctrl_n/Ctrl_pct** = same for Control (C4). The denominator for percentages is
**coding pairs** (excluding `non_coding` and `no_shared_cds`), consistent with
Figure 4 — the 5 coding categories sum to 100%. The `non_coding` and
`no_shared_cds` rows show counts only (no percentage).

Note: the coding pair count here differs from Table 1 because `compareIsoformFrames()`
classifies some individually-coding pairs as `no_shared_cds` (both isoforms are
coding but their CDS regions don't overlap). Table 1's "coding pairs" filter
requires only that both isoforms have `coding_status == "coding"` in `cds.rds`.

## How PTCs Arise: Frameshift vs Non-Frameshift Mechanisms

```{r goal2-ptc-mechanisms}
# ── Data provenance ──
# Source: computed inline in this report (not a separate script)
# Methods: METHODS.md § "PTC-Causing Event Attribution"
#
# For each PTC+ NMD comparator, identify the PTC-CAUSING splice event:
#   - Frameshift PTCs: from analyzeFrameWalk() (Isopair::analyzeFrameWalk,
#     cached as fw_c2_allsamples.rds by 03_nmd_analysis_mashr.Rmd)
#   - Non-frameshift PTCs: the event whose genomic coordinates (from
#     buildProfiles() detailed_events) contain the comparator's cds_stop
#     (= the PTC genomic position, from Isopair::extractCds → cds.rds)
#
# cds_stop in cds.rds is the genomic position of the stop codon for each isoform.
# For PTC isoforms, SQANTI annotates the actual (premature) stop, so cds_stop IS
# the PTC position.

fc_c2 <- set_results[["c2_allsamples"]]$frame_comp
fw_c2 <- set_results[["c2_allsamples"]]$frame_walk
prof_c2 <- set_results[["c2_allsamples"]]$profiles

fc_c2_ps <- fc_c2$pair_summary
fc_c2_ps$has_ptc <- fc_c2_ps$comparator_isoform_id %in% ptc_pos_ids

frameshift_cats <- c("same_start_frameshift",
                     "diff_start_diff_frame_with_frameshift")
fc_c2_ps$is_frameshift_cat <- fc_c2_ps$frame_category %in% frameshift_cats

# PTC+ pairs by mechanism
ptc_plus <- fc_c2_ps[fc_c2_ps$has_ptc == TRUE, ]
n_ptc_via_fs    <- sum(ptc_plus$is_frameshift_cat)
n_ptc_via_nonfs <- sum(!ptc_plus$is_frameshift_cat)
n_ptc_total     <- nrow(ptc_plus)

# Also count all coding NMD pairs for context
n_coding_nmd <- sum(!fc_c2_ps$frame_category %in% c("non_coding", "no_shared_cds"))

# ── Attribute PTC-causing event for each PTC+ pair ──
cds_stop_lookup <- setNames(cds$cds_stop, cds$isoform_id)

# Build lookup: comparator_id → row index in prof_c2
comp_to_idx <- setNames(seq_len(nrow(prof_c2)), prof_c2$comparator_isoform_id)

# Frameshift PTC-causing events (from frame_walk)
fw_fs <- fw_c2$events[fw_c2$events$is_frameshift == TRUE, ]

ptc_event_rows <- list()
for (i in seq_len(nrow(ptc_plus))) {
  comp_id <- ptc_plus$comparator_isoform_id[i]
  is_fs <- ptc_plus$is_frameshift_cat[i]

  if (is_fs) {
    # Frameshift PTC: use the frameshift-causing event(s) from frame_walk
    fs_match <- fw_fs[fw_fs$comparator_isoform_id == comp_id, ]
    if (nrow(fs_match) > 0) {
      # Use the first frameshift event (primary cause)
      ptc_event_rows[[length(ptc_event_rows) + 1]] <- data.frame(
        comparator_id = comp_id,
        mechanism = "Frameshift",
        ptc_causing_event = fs_match$event_type[1],
        stringsAsFactors = FALSE)
    }
  } else {
    # Non-frameshift PTC: find the event whose coordinates contain cds_stop
    prof_idx <- comp_to_idx[comp_id]
    if (is.na(prof_idx)) next
    de <- prof_c2$detailed_events[[prof_idx]]
    if (is.null(de) || nrow(de) == 0) next
    ptc_pos_genomic <- cds_stop_lookup[comp_id]
    if (is.na(ptc_pos_genomic)) next

    # Check which event's coordinates span the PTC position
    de$ev_min <- pmin(de$five_prime, de$three_prime)
    de$ev_max <- pmax(de$five_prime, de$three_prime)
    containing <- de[ptc_pos_genomic >= de$ev_min & ptc_pos_genomic <= de$ev_max, ]

    if (nrow(containing) > 0) {
      ptc_event_rows[[length(ptc_event_rows) + 1]] <- data.frame(
        comparator_id = comp_id,
        mechanism = "In-frame stop",
        ptc_causing_event = containing$event_type[1],
        stringsAsFactors = FALSE)
    } else {
      ptc_event_rows[[length(ptc_event_rows) + 1]] <- data.frame(
        comparator_id = comp_id,
        mechanism = "In-frame stop",
        ptc_causing_event = "Unattributed",
        stringsAsFactors = FALSE)
    }
  }
}

ptc_events_df <- do.call(rbind, ptc_event_rows)
n_attributed <- sum(ptc_events_df$ptc_causing_event != "Unattributed")
n_unattributed <- sum(ptc_events_df$ptc_causing_event == "Unattributed")
```

Of `r fmt(n_coding_nmd)` coding NMD pairs, `r fmt(n_ptc_total)` have a PTC in
the comparator. These arise through two mechanisms:

- **Frameshift → PTC**: `r fmt(n_ptc_via_fs)` pairs
  (`r round(100 * n_ptc_via_fs / n_ptc_total, 1)`%) — a splicing event shifts
  the reading frame, and the new frame encounters a premature stop codon. The
  PTC-causing event is identified by `analyzeFrameWalk()`.
- **In-frame stop → PTC**: `r fmt(n_ptc_via_nonfs)` pairs
  (`r round(100 * n_ptc_via_nonfs / n_ptc_total, 1)`%) — the comparator is
  in the same reading frame but a structural change (e.g., a gained exon or
  retained intron) directly introduces an in-frame stop codon. The PTC-causing
  event is identified by finding the event whose genomic coordinates contain the
  comparator's stop codon position (`cds_stop` in `cds.rds`).

Of `r fmt(nrow(ptc_events_df))` PTC+ pairs with attribution attempted,
`r fmt(n_attributed)` were successfully attributed to a specific splice event
and `r fmt(n_unattributed)` could not be attributed (PTC position did not fall
within any event's coordinates).

## Figure 5: PTC-Causing Splice Events (Sankey)

This figure shows the complete mapping from PTC-causing splice event → mechanism
(Frameshift or In-frame stop) for all attributed PTC+ NMD comparators. For
frameshift PTCs, the event is the one identified by `analyzeFrameWalk()`. For
in-frame stop PTCs, it is the event whose coordinates contain the stop codon.

```{r goal2-fig5-sankey}
# Shorten event type labels
ptc_events_df$event_short <- ptc_events_df$ptc_causing_event
ptc_events_df$event_short[ptc_events_df$event_short == "Missing_Internal"] <- "Missing Int."
ptc_events_df$event_short[ptc_events_df$event_short == "Partial_IR_5"] <- "Partial IR 5'"
ptc_events_df$event_short[ptc_events_df$event_short == "Partial_IR_3"] <- "Partial IR 3'"
ptc_events_df$event_short[ptc_events_df$event_short == "Alt_TSS"] <- "Alt TSS"
ptc_events_df$event_short[ptc_events_df$event_short == "Alt_TES"] <- "Alt TES"
ptc_events_df$event_short[ptc_events_df$event_short == "IR_diff"] <- "IR diff"

# Exclude unattributed for Sankey
ptc_attr <- ptc_events_df[ptc_events_df$ptc_causing_event != "Unattributed", ]

# Aggregate: event_type → mechanism
sankey_agg <- ptc_attr |>
  group_by(event_short, mechanism) |>
  summarise(freq = n(), .groups = "drop") |>
  filter(freq >= 3)

# Build networkD3 data structures
nodes_vec <- unique(c(sankey_agg$event_short, sankey_agg$mechanism))
nodes <- data.frame(name = nodes_vec, stringsAsFactors = FALSE)

links <- data.frame(
  source = match(sankey_agg$event_short, nodes_vec) - 1L,
  target = match(sankey_agg$mechanism, nodes_vec) - 1L,
  value  = sankey_agg$freq
)

# Node colors: Frameshift red, In-frame stop blue, event types grey
node_colors <- ifelse(nodes$name == "Frameshift", "#d6604d",
               ifelse(nodes$name == "In-frame stop", "#4393c3", "#bdbdbd"))

color_scale <- paste0(
  'd3.scaleOrdinal().domain([',
  paste0('"', nodes$name, '"', collapse = ','), ']).range([',
  paste0('"', node_colors, '"', collapse = ','), '])')

sankey_widget <- networkD3::sankeyNetwork(
  Links = links, Nodes = nodes,
  Source = "source", Target = "target", Value = "value",
  NodeID = "name", fontSize = 14, nodeWidth = 30, nodePadding = 15,
  colourScale = color_scale, sinksRight = TRUE,
  margin = list(left = 120, right = 20, top = 10, bottom = 10))

# Move source node labels to the left of the rectangles
htmlwidgets::onRender(sankey_widget, '
  function(el, x) {
    d3.select(el).selectAll(".node text")
      .filter(function(d) { return d.sourceLinks.length > 0 && d.targetLinks.length === 0; })
      .attr("x", -6)
      .attr("text-anchor", "end");
  }
')
```

## Table 4a: PTC-Causing Events by Mechanism

```{r goal2-fig5-summary}
# Summary table of PTC-causing events by mechanism
ptc_event_summary <- ptc_attr |>
  group_by(mechanism, ptc_causing_event) |>
  summarise(n = n(), .groups = "drop") |>
  arrange(mechanism, desc(n))
ptc_event_summary$pct <- round(100 * ptc_event_summary$n / nrow(ptc_attr), 1)

render_dt(ptc_event_summary, caption = "Table 4a: PTC-Causing Events by Mechanism (NMD)")
write.csv(ptc_event_summary, file.path(table_dir, "table4a_ptc_causing_events.csv"),
          row.names = FALSE)
```

Columns: **mechanism** = Frameshift or In-frame stop; **ptc_causing_event** =
splice event type that caused the PTC; **n** = number of pairs; **pct** =
percentage of all attributed PTCs.

## Table 4b: PTC-Causing Event Proportions vs Control Baseline

```{r goal2-table4b-baseline}
# ── Data provenance ──
# NMD: PTC-causing events attributed above (both frameshift and in-frame stop)
# Control baseline: all splicing events across all C4 pairs
# Comparison: proportion of each event type among PTC-causing NMD events vs
#   all Control events (the baseline frequency in normal alternative splicing)

# NMD PTC-causing event frequencies (attributed only)
ptc_evt_freq <- sort(table(ptc_attr$ptc_causing_event), decreasing = TRUE)
ptc_evt_df <- data.frame(
  event_type = names(ptc_evt_freq),
  n_ptc_events = as.integer(ptc_evt_freq),
  pct_of_ptc = round(100 * as.integer(ptc_evt_freq) / nrow(ptc_attr), 1),
  stringsAsFactors = FALSE)

# Control baseline: all events across all C4 pairs
ctrl_all_events <- do.call(rbind, set_results[["c4_allsamples"]]$profiles$detailed_events)
ctrl_event_freq <- table(ctrl_all_events$event_type)
ctrl_total <- sum(ctrl_event_freq)

# Merge
ptc_evt_df$n_ctrl_events <- as.integer(ctrl_event_freq[ptc_evt_df$event_type])
ptc_evt_df$n_ctrl_events[is.na(ptc_evt_df$n_ctrl_events)] <- 0L
ptc_evt_df$pct_ctrl <- round(100 * ptc_evt_df$n_ctrl_events / ctrl_total, 1)
ptc_evt_df$enrichment <- round(ptc_evt_df$pct_of_ptc / pmax(ptc_evt_df$pct_ctrl, 0.1), 1)

# Fisher's exact test per event type: is this event over-represented among
# PTC-causing events vs all Control events?
n_ptc_total <- nrow(ptc_attr)
ptc_evt_df$fisher_p <- NA_real_
ptc_evt_df$direction <- NA_character_
for (i in seq_len(nrow(ptc_evt_df))) {
  a <- ptc_evt_df$n_ptc_events[i]      # this event in PTC-causing
  b <- n_ptc_total - a                   # other events in PTC-causing
  c_val <- ptc_evt_df$n_ctrl_events[i]  # this event in Control
  d_val <- ctrl_total - c_val            # other events in Control
  ft <- tryCatch(fisher.test(matrix(c(a, b, c_val, d_val), nrow = 2)),
                 error = function(e) NULL)
  if (!is.null(ft)) {
    ptc_evt_df$fisher_p[i] <- signif(ft$p.value, 3)
    ptc_evt_df$direction[i] <- if (ft$p.value > 0.05) "Non-significant"
      else if (ptc_evt_df$enrichment[i] > 1) "Enriched in PTC-causing"
      else "Depleted in PTC-causing"
  }
}

render_dt(ptc_evt_df, caption = "Table 4b: PTC-Causing Events (NMD) vs All Events (Control)")
write.csv(ptc_evt_df, file.path(table_dir, "table4b_ptc_event_enrichment.csv"),
          row.names = FALSE)
```

Columns: **event_type** = splice event type; **n_ptc_events** = count of
PTC-causing events of this type; **pct_of_ptc** = % of all
`r fmt(n_ptc_total)` attributed PTC-causing events; **n_ctrl_events** = count in
gene-matched Control (C4) pairs; **pct_ctrl** = % of `r fmt(ctrl_total)` Control
events; **enrichment** = fold enrichment (% PTC / % Ctrl); **fisher_p** =
Fisher's exact test p-value (2×2: this event vs others × PTC-causing vs
Control); **direction** = Enriched/Depleted in PTC-causing, or Non-significant
(p > 0.05).


# 3. Non-PTC NMD Triggers {#goal3}

**Goal 3:** Characterize NMD-responsive comparator isoforms that lack a PTC.

**Methods:** Among coding NMD (C2) comparators, a substantial fraction lack a PTC (stop
codon >50 nt upstream of last EJC). We categorize these PTC-negative NMD isoforms by:
(a) borderline PTC position (stop codon with 1+ downstream EJCs but within 50 nt),
(b) 3'UTR exonic length, (c) frame category, and (d) NMD response magnitude.

```{r goal3-identify-ptc-neg}
# ── Data provenance ──
# Starting data: gene-matched C2 all_samples profiles → cached PTC association
# ptc_c2_as = set_results[["c2_allsamples"]]$ptc_assoc
#   Source: testPtcAssociation(ptc, gene-matched profiles), cached by script 03
#   $pair_summary: one row per gene-matched pair, with comp_has_ptc flag
# Step 1: Filter to coding pairs (both ref and comp in cds$coding_status == "coding")
# Step 2: Split into PTC+ and PTC- NMD comparators

ptc_c2_as <- set_results[["c2_allsamples"]]$ptc_assoc
pair_summary <- ptc_c2_as$pair_summary

# Restrict to coding pairs (same filter as Table 1)
coding_ids_g3 <- cds$isoform_id[cds$coding_status == "coding"]
coding_pairs_c2 <- pair_summary[pair_summary$comparator_isoform_id %in% coding_ids_g3 &
                                  pair_summary$reference_isoform_id %in% coding_ids_g3, ]

n_ptc_pos <- sum(coding_pairs_c2$comp_has_ptc == TRUE)
n_ptc_neg <- sum(coding_pairs_c2$comp_has_ptc == FALSE)
```

Of the `r fmt(nrow(set_results[["c2_allsamples"]]$profiles))` gene-matched NMD
pairs, `r fmt(nrow(coding_pairs_c2))` are coding pairs (both reference and
comparator classified as protein-coding). Among these,
`r fmt(n_ptc_pos)` (`r round(100 * n_ptc_pos / nrow(coding_pairs_c2), 1)`%)
have a PTC (comparator stop codon >50 nt upstream of last EJC) and
`r fmt(n_ptc_neg)` (`r round(100 * n_ptc_neg / nrow(coding_pairs_c2), 1)`%)
do not. The remainder of this section characterizes these
`r fmt(n_ptc_neg)` **PTC-negative NMD comparators**.

## Categorization by Stop Codon Position

```{r goal3-stop-position}
# ── Data provenance ──
# Starting data: ptc_neg_ids (PTC-negative NMD comparators identified above)
# ptc.rds: per-isoform annotations (n_downstream_ejcs, ptc_distance, etc.)
#   Source: Isopair testPtcAssociation() infrastructure — isoform-intrinsic
# Step 1: Look up ptc features for PTC-neg comparators
# Step 2: Categorize by n_downstream_ejcs (borderline vs stop-in-last-exon)

ptc_neg_ids <- coding_pairs_c2$comparator_isoform_id[coding_pairs_c2$comp_has_ptc == FALSE]
ptc_neg_info <- ptc[ptc$isoform_id %in% ptc_neg_ids, ]

# Categorize
n_borderline <- sum(ptc_neg_info$n_downstream_ejcs > 0, na.rm = TRUE)
n_stop_last_exon <- sum(ptc_neg_info$n_downstream_ejcs == 0, na.rm = TRUE)
n_na <- sum(is.na(ptc_neg_info$n_downstream_ejcs))

stop_cat <- data.frame(
  category = c("Borderline (1+ downstream EJCs, distance <= 50 nt)",
               "Stop in last exon (0 downstream EJCs)",
               "Missing annotation"),
  n = c(n_borderline, n_stop_last_exon, n_na),
  pct = round(100 * c(n_borderline, n_stop_last_exon, n_na) / length(ptc_neg_ids), 1))

kable(stop_cat, caption = "Table 5: Categorization of PTC-Negative NMD Comparators")
```

## Figure 6: 3'UTR Length by PTC Mechanism

```{r goal3-fig6-pop}
# ── Data provenance ──
# Source: computed inline in this report using cached divergence + cds.rds
# Methods: METHODS.md § "3'UTR Length Analysis with PTC Correction"
# Key functions:
#   Isopair::quantifyPairDivergence() → div_c2/div_c4 (utr3_bp_comp column)
#   Isopair::.computeRegionalBp() → computes 3'UTR as exonic bp after cds_stop
#   Reference-based 3'UTR: computed inline using ref cds_stop on comp exon structure
# Starting data: gene-matched C2 and C4 profiles → cached divergence
# PTC mechanism classification: "same stop" vs "diff stop"
#   same_stop: comparator and reference share the same cds_stop position.
#     The PTC arises because a 3'UTR splicing event repositioned the last EJC
#     downstream, making the existing stop codon >50 nt upstream of the new EJC.
#     The 3'UTR length measurement (from cds_stop to tx end) is correct and
#     unaffected by the PTC classification.
#   diff_stop: comparator has a different cds_stop (= premature stop from
#     frameshift or in-frame stop in the CDS). The measured 3'UTR is inflated
#     because it includes the former CDS region between PTC and normal stop.
#     For these, we also compute a "reference-based 3'UTR": exonic bp downstream
#     of the REFERENCE's cds_stop in the COMPARATOR's exon structure. This is
#     only valid when the reference stop position falls within the comparator's
#     exons (86% of cases).

div_c2 <- set_results[["c2_allsamples"]]$divergence
div_c4 <- set_results[["c4_allsamples"]]$divergence
ptc_pos_comp_ids <- coding_pairs_c2$comparator_isoform_id[coding_pairs_c2$comp_has_ptc == TRUE]

# Classify PTC+ by same vs different stop codon position
cds_stop_lookup_g3 <- setNames(cds$cds_stop, cds$isoform_id)
strand_lookup_g3 <- setNames(cds$strand, cds$isoform_id)

ptc_pos_ps <- coding_pairs_c2[coding_pairs_c2$comp_has_ptc == TRUE, ]
ptc_pos_ps$comp_stop <- cds_stop_lookup_g3[ptc_pos_ps$comparator_isoform_id]
ptc_pos_ps$ref_stop <- cds_stop_lookup_g3[ptc_pos_ps$reference_isoform_id]
ptc_pos_ps$same_stop <- ptc_pos_ps$comp_stop == ptc_pos_ps$ref_stop

same_stop_ids <- ptc_pos_ps$comparator_isoform_id[ptc_pos_ps$same_stop == TRUE]
diff_stop_ids <- ptc_pos_ps$comparator_isoform_id[ptc_pos_ps$same_stop == FALSE]

# For diff-stop pairs, compute reference-based 3'UTR
diff_stop_ps <- ptc_pos_ps[ptc_pos_ps$same_stop == FALSE, ]
ref_based_utr3 <- setNames(rep(NA_integer_, nrow(diff_stop_ps)), diff_stop_ps$comparator_isoform_id)

for (i in seq_len(nrow(diff_stop_ps))) {
  comp_id <- diff_stop_ps$comparator_isoform_id[i]
  ref_stop_pos <- diff_stop_ps$ref_stop[i]
  comp_strand <- strand_lookup_g3[comp_id]
  if (is.na(comp_strand) || is.na(ref_stop_pos)) next

  str_row <- structures[structures$isoform_id == comp_id, ]
  if (nrow(str_row) == 0) next
  exon_s <- str_row$exon_starts[[1]]
  exon_e <- str_row$exon_ends[[1]]

  # Check ref stop is in comparator exons
  in_exon <- any(ref_stop_pos >= exon_s & ref_stop_pos <= exon_e)
  if (!in_exon) next

  utr3 <- 0L
  for (k in seq_along(exon_s)) {
    if (comp_strand == "+") {
      if (exon_e[k] > ref_stop_pos) {
        post_start <- max(exon_s[k], ref_stop_pos + 1L)
        utr3 <- utr3 + as.integer(exon_e[k] - post_start + 1L)
      }
    } else {
      if (exon_s[k] < ref_stop_pos) {
        pre_end <- min(exon_e[k], ref_stop_pos - 1L)
        utr3 <- utr3 + as.integer(pre_end - exon_s[k] + 1L)
      }
    }
  }
  ref_based_utr3[comp_id] <- utr3
}

n_eligible <- sum(!is.na(ref_based_utr3))
n_ineligible <- sum(is.na(ref_based_utr3))

# Lookup for comparator 3'UTR from divergence
utr3_comp_lk <- setNames(div_c2$utr3_bp_comp, div_c2$comparator_isoform_id)

# Build 5-group data
utr3_data <- rbind(
  data.frame(group = "Same-stop PTC+\n(3'UTR splicing)",
    utr3_bp = utr3_comp_lk[same_stop_ids]),
  data.frame(group = "Diff-stop PTC+\n(own stop, inflated)",
    utr3_bp = utr3_comp_lk[diff_stop_ids]),
  data.frame(group = "Diff-stop PTC+\n(ref stop, corrected)",
    utr3_bp = ref_based_utr3[!is.na(ref_based_utr3)]),
  data.frame(group = "NMD PTC-",
    utr3_bp = utr3_comp_lk[ptc_neg_ids]),
  data.frame(group = "Control",
    utr3_bp = div_c4$utr3_bp_comp)
)
utr3_data <- utr3_data[!is.na(utr3_data$utr3_bp) & utr3_data$utr3_bp > 0, ]
utr3_data$group <- factor(utr3_data$group,
  levels = c("Same-stop PTC+\n(3'UTR splicing)",
             "Diff-stop PTC+\n(own stop, inflated)",
             "Diff-stop PTC+\n(ref stop, corrected)",
             "NMD PTC-", "Control"))
utr3_n <- table(utr3_data$group)
```

PTC+ comparators are split by whether the stop codon position is the **same** as
the reference (`cds_stop` identical) or **different** (premature stop from
frameshift or in-frame stop in CDS):

- **Same-stop PTC+** (`r fmt(utr3_n[[1]])`): The stop codon is the same as the
  reference. A 3'UTR splicing event repositioned the last EJC, making the
  existing stop >50 nt upstream of the new EJC. The 3'UTR length is correctly
  measured. *Example: TBCCD1 — comp 3'UTR=1,698 bp, ref 3'UTR=910 bp, same stop codon,
  3 downstream EJCs from 3'UTR splicing.*
- **Diff-stop PTC+ (own stop)** (`r fmt(utr3_n[[2]])`): The stop codon moved
  upstream (PTC). The measured 3'UTR is inflated — it includes former CDS.
  *Example: TRIM41 — comp ORF=1,146 aa vs ref ORF=1,893 aa; comp 3'UTR=1,758 bp
  but much of this is former CDS.*
- **Diff-stop PTC+ (ref stop)** (`r fmt(utr3_n[[3]])`): Same pairs as above, but
  3'UTR is measured from the **reference's** stop codon position projected onto
  the comparator's exon structure. `r fmt(n_eligible)` of `r fmt(n_eligible + n_ineligible)`
  pairs were eligible (reference stop falls within comparator exons).
- **NMD PTC−** (`r fmt(utr3_n[[4]])`): No PTC. 3'UTR is correctly measured.
  *Example: GPAT3 — comp 3'UTR=1,108 bp, ref 3'UTR=1,108 bp.*
- **Control** (`r fmt(utr3_n[[5]])`): C4 comparators.
  *Example: SART1 — comp 3'UTR=1,107 bp, ref 3'UTR=1,107 bp.*

```{r goal3-fig6-utr3-density, fig.width=10, fig.height=6}
pal5_utr3 <- c("Same-stop PTC+\n(3'UTR splicing)" = "#d6604d",
               "Diff-stop PTC+\n(own stop, inflated)" = "#fc8d59",
               "Diff-stop PTC+\n(ref stop, corrected)" = "#fee08b",
               "NMD PTC-" = "#d9ef8b",
               "Control" = "#67a9cf")

fig6 <- ggplot(utr3_data, aes(x = utr3_bp, fill = group)) +
  geom_density(alpha = 0.5) +
  scale_x_log10(labels = scales::comma) +
  scale_fill_manual(values = pal5_utr3) +
  labs(title = "3'UTR Length by PTC Mechanism",
       subtitle = "Correcting for CDS truncation reveals PTC-neg NMD isoforms have normal 3'UTR lengths",
       x = "3'UTR length (bp, log scale)", y = "Density", fill = "")
save_fig(fig6, "fig6_utr3_length_density", width = 10, height = 6)
```

## Table 6: 3'UTR Length Comparisons (Wilcoxon)

```{r goal3-table6-utr3-tests}
grps <- split(utr3_data$utr3_bp, utr3_data$group)

comparisons_list <- list(
  c("Same-stop PTC+\n(3'UTR splicing)", "Control"),
  c("Diff-stop PTC+\n(own stop, inflated)", "Control"),
  c("Diff-stop PTC+\n(ref stop, corrected)", "Control"),
  c("NMD PTC-", "Control"),
  c("Same-stop PTC+\n(3'UTR splicing)", "Diff-stop PTC+\n(ref stop, corrected)")
)

tests <- do.call(rbind, lapply(comparisons_list, function(pair) {
  g1 <- grps[[pair[1]]]
  g2 <- grps[[pair[2]]]
  wt <- wilcox.test(g1, g2)
  data.frame(
    group1 = gsub("\n", " ", pair[1]),
    n_group1 = length(g1), median_group1 = round(median(g1)),
    group2 = gsub("\n", " ", pair[2]),
    n_group2 = length(g2), median_group2 = round(median(g2)),
    p_value = signif(wt$p.value, 3),
    stringsAsFactors = FALSE)
}))

render_dt(tests, caption = "Table 6: 3'UTR Length Comparisons (Wilcoxon rank-sum test)")
write.csv(tests, file.path(table_dir, "table6_utr3_wilcoxon.csv"), row.names = FALSE)
```

Columns: **group1/group2** = the two groups being compared (see Figure 6 for
group definitions); **n_group1/n_group2** = number of isoforms in each group;
**median_group1/median_group2** = median 3'UTR length (bp);
**p_value** = Wilcoxon rank-sum test p-value.
```

## Illustrative Example: 3'UTR Measurement Correction (CMAS)

The figure below shows the CMAS gene pair (reference ENST00000229329.7 vs NMD
comparator ENST00000534981.5) to illustrate the difference between the two
3'UTR measurements. The comparator has a skipped exon that causes a frameshift,
creating a PTC at genomic position 22,062,284. The reference stop codon is at
22,065,311.

```{r goal3-utr3-example, fig.width=10, fig.height=10}
# CMAS: + strand, ref has 8 exons, comp has 7 (skipped exon 22061281-22061452)
# Same CDS start (22046304), different stop (ref: 22065311, comp: 22062284)

ref_s <- c(22046218, 22055149, 22055455, 22058567, 22060832, 22061281, 22062281, 22065121)
ref_e <- c(22046563, 22055291, 22055610, 22058700, 22060926, 22061452, 22062434, 22065668)
comp_s <- c(22046225, 22055149, 22055455, 22058567, 22060832, 22062281, 22065121)
comp_e <- c(22046563, 22055291, 22055610, 22058700, 22060926, 22062434, 22065674)

# Skipped exon event (comparator lost exon 22061281-22061452)
se_event <- data.frame(
  five_prime = 22061281, three_prime = 22061452,
  event_type = "SE", direction = "LOSS")

# Annotations: PTC marker on comparator, reference stop marker on both
ptc_marker <- list(type = "ptc_marker", track = "comp",
  x = 22062284, label = "PTC", color = "#E74C3C")
ref_stop_marker <- list(type = "ptc_marker", track = "ref",
  x = 22065311, label = "Stop", color = "#27AE60")

# Panel A: "Own stop" — comparator CDS ends at PTC (22062284)
p1 <- plot_ucsc_pair(
  ref_s, ref_e, comp_s, comp_e, strand = "+",
  ref_cds = list(cds_start = 22046304, cds_stop = 22065311),
  comp_cds = list(cds_start = 22046304, cds_stop = 22062284),
  ref_label = "Reference (ENST00000229329.7)",
  comp_label = "NMD Comparator (ENST00000534981.5)",
  events = se_event,
  annotations = list(ptc_marker, ref_stop_marker),
  title = "A) 3'UTR at own stop (inflated)",
  subtitle = "Comparator CDS ends at PTC. Measured 3'UTR (thin) includes former CDS downstream of PTC.")

# Panel B: "Ref stop" — comparator CDS extends to reference stop (22065311)
ref_stop_comp <- list(type = "ptc_marker", track = "comp",
  x = 22065311, label = "Ref Stop", color = "#27AE60")

p2 <- plot_ucsc_pair(
  ref_s, ref_e, comp_s, comp_e, strand = "+",
  ref_cds = list(cds_start = 22046304, cds_stop = 22065311),
  comp_cds = list(cds_start = 22046304, cds_stop = 22065311),
  ref_label = "Reference (ENST00000229329.7)",
  comp_label = "NMD Comparator (ENST00000534981.5)",
  events = se_event,
  annotations = list(ptc_marker, ref_stop_marker, ref_stop_comp),
  title = "B) 3'UTR at reference stop (corrected)",
  subtitle = "Comparator CDS extended to reference stop. 3'UTR = only true downstream sequence.")

gridExtra::grid.arrange(p1, p2, ncol = 1)
```

In panel A, the skipped exon (SE LOSS bracket) causes a frameshift that creates
a PTC (red triangle on comparator). The comparator's CDS (thick) ends at the
PTC, and everything downstream is measured as 3'UTR (thin) — inflating the
length. The reference stop codon (green triangle) is further downstream. In
panel B, the comparator's CDS boundary is set to the reference's stop position
(green triangle on both tracks), giving the "natural" 3'UTR length.

**Key finding:** PTC-negative NMD comparators have 3'UTR lengths indistinguishable
from Control (p = `r signif(tests$p_value[tests$comparison == "NMD PTC- vs Control"], 2)`).
Long 3'UTR length does not appear to be an independent NMD trigger in this
dataset. The apparently long 3'UTRs in PTC+ isoforms are explained by two
mechanisms: (1) for same-stop PTC+ isoforms, the 3'UTR is genuinely extended by
the same splicing event that repositioned the EJC; (2) for diff-stop PTC+
isoforms, the 3'UTR measurement is inflated by inclusion of former CDS sequence
downstream of the premature stop.

## 3'UTR Splicing and NMD

The question is whether 3'UTR splicing contributes to NMD in our data,
and if so, through what mechanism — PTC creation (EJC repositioning) or a
PTC-independent pathway (e.g., long 3'UTR-mediated UPF1 recruitment).

```{r goal3-utr3-splicing}
# ── Data provenance ──
# Source: computed inline using count_utr3_splicing() defined in this chunk
# Methods: METHODS.md § "3'UTR Splicing Enrichment"
# Starting data: gene-matched C2 and C4 profiles (detailed_events from
#   Isopair::buildProfiles() → compareSplicingChoices) + cds.rds (Isopair::extractCds)

cds_lookup <- setNames(cds$cds_stop, cds$isoform_id)
strand_lookup <- setNames(cds$strand, cds$isoform_id)

count_utr3_splicing <- function(profiles, comp_id_subset = NULL) {
  if (!is.null(comp_id_subset)) {
    idx <- which(profiles$comparator_isoform_id %in% comp_id_subset)
  } else {
    idx <- seq_len(nrow(profiles))
  }
  n_with <- 0L
  n_total <- 0L
  for (i in idx) {
    comp_id <- profiles$comparator_isoform_id[i]
    de <- profiles$detailed_events[[i]]
    comp_stop <- cds_lookup[comp_id]
    comp_strand <- strand_lookup[comp_id]
    if (is.na(comp_stop) || is.na(comp_strand)) next
    n_total <- n_total + 1L
    if (is.null(de) || nrow(de) == 0) next
    if (comp_strand == "+") {
      utr3_ev <- de[de$five_prime >= comp_stop | de$three_prime >= comp_stop, ]
    } else {
      utr3_ev <- de[de$five_prime <= comp_stop | de$three_prime <= comp_stop, ]
    }
    if (nrow(utr3_ev) > 0) n_with <- n_with + 1L
  }
  list(n_with = n_with, n_total = n_total,
       pct = round(100 * n_with / max(n_total, 1), 1))
}

prof_c4_as <- set_results[["c4_allsamples"]]$profiles

utr3_ptcneg <- count_utr3_splicing(prof_c2, ptc_neg_ids)
utr3_ptcpos <- count_utr3_splicing(prof_c2, ptc_pos_comp_ids)
utr3_ctrl   <- count_utr3_splicing(prof_c4_as)

ft_utr3_neg <- fisher.test(matrix(c(
  utr3_ptcneg$n_with, utr3_ptcneg$n_total - utr3_ptcneg$n_with,
  utr3_ctrl$n_with, utr3_ctrl$n_total - utr3_ctrl$n_with), nrow = 2))
ft_utr3_pos <- fisher.test(matrix(c(
  utr3_ptcpos$n_with, utr3_ptcpos$n_total - utr3_ptcpos$n_with,
  utr3_ctrl$n_with, utr3_ctrl$n_total - utr3_ctrl$n_with), nrow = 2))

utr3_summary <- data.frame(
  group = c("NMD PTC-", "NMD PTC+", "Control"),
  n_with_utr3_event = c(utr3_ptcneg$n_with, utr3_ptcpos$n_with, utr3_ctrl$n_with),
  n_total = c(utr3_ptcneg$n_total, utr3_ptcpos$n_total, utr3_ctrl$n_total),
  pct = c(utr3_ptcneg$pct, utr3_ptcpos$pct, utr3_ctrl$pct),
  vs_control_OR = c(round(as.numeric(ft_utr3_neg$estimate), 2),
                     round(as.numeric(ft_utr3_pos$estimate), 2), NA),
  vs_control_p = c(signif(ft_utr3_neg$p.value, 3),
                    signif(ft_utr3_pos$p.value, 3), NA))

kable(utr3_summary,
      caption = "3'UTR Splicing Prevalence by Group",
      col.names = c("Group", "With 3'UTR Event", "Total", "%",
                     "OR vs Control", "p vs Control"))
```

Columns: **Group**; **With 3'UTR Event** = pairs where ≥1 splice event has
coordinates in the 3'UTR (downstream of `cds_stop` on + strand);
**OR/p vs Control** = Fisher's exact test comparing to Control (C4).

```{r goal3-utr3-same-stop-verification}
# ── Direct verification: 3'UTR splicing creates PTCs via EJC repositioning ──
# For same-stop PTC+ pairs (645), the stop codon is identical to the reference.
# The PTC status arises because the comparator has more exon-exon junctions
# downstream of the stop codon (from 3'UTR splicing), repositioning the last EJC.
# We can verify by comparing n_downstream_ejcs between ref and comp.

same_stop_ps <- ptc_pos_ps[ptc_pos_ps$same_stop == TRUE, ]
ptc_ejc_lookup <- setNames(ptc$n_downstream_ejcs, ptc$isoform_id)

same_stop_ps$comp_ejcs <- ptc_ejc_lookup[same_stop_ps$comparator_isoform_id]
same_stop_ps$ref_ejcs <- ptc_ejc_lookup[same_stop_ps$reference_isoform_id]
same_stop_ps$ejc_diff <- same_stop_ps$comp_ejcs - same_stop_ps$ref_ejcs

n_comp_more_ejcs <- sum(same_stop_ps$ejc_diff > 0, na.rm = TRUE)
n_same_ejcs <- sum(same_stop_ps$ejc_diff == 0, na.rm = TRUE)
n_comp_fewer_ejcs <- sum(same_stop_ps$ejc_diff < 0, na.rm = TRUE)
n_valid <- sum(!is.na(same_stop_ps$ejc_diff))
```

### 3'UTR Splicing Creates PTCs via EJC Repositioning

Among the `r fmt(nrow(same_stop_ps))` same-stop PTC+ pairs (where the
comparator has the same stop codon as the reference but is classified as PTC+),
we directly verified the mechanism by comparing the number of exon-exon
junctions downstream of the stop codon:

- **`r n_comp_more_ejcs`** pairs (`r round(100 * n_comp_more_ejcs / n_valid, 1)`%):
  comparator has MORE downstream EJCs than reference — confirming that 3'UTR
  splicing added new junctions, repositioning the last EJC and creating the PTC
- **`r n_same_ejcs`** pairs: same number of downstream EJCs (PTC may arise from
  junction repositioning rather than addition)
- **`r n_comp_fewer_ejcs`** pairs: comparator has fewer EJCs (unexpected —
  may reflect annotation differences)

### Conclusions: 3'UTR Mechanisms in NMD

Three lines of evidence address whether 3'UTR length or splicing triggers
PTC-independent NMD:

1. **3'UTR length is NOT enriched in PTC-negative NMD** (Figure 6: PTC-neg
   median = Control median, p = `r signif(tests$p_value[tests$group1 == "NMD PTC-"], 2)`).
   Long 3'UTR does not appear to be an independent NMD trigger in this dataset.

2. **3'UTR splicing that triggers NMD does so by creating PTCs** (EJC
   repositioning), not through a PTC-independent pathway. The
   `r fmt(nrow(same_stop_ps))` same-stop PTC+ pairs are direct evidence:
   `r round(100 * n_comp_more_ejcs / n_valid, 1)`% have additional downstream
   EJCs in the comparator from 3'UTR splicing events.

3. **3'UTR splicing prevalence** is
   `r if (ft_utr3_neg$p.value < 0.05) "modestly enriched" else "not enriched"`
   in PTC-negative NMD vs Control (OR =
   `r round(as.numeric(ft_utr3_neg$estimate), 2)`, p =
   `r signif(ft_utr3_neg$p.value, 3)`), but the effect is far smaller than the
   uORF enrichment shown below.

## 5'UTR uORF Analysis

Upstream open reading frames (uORFs) in the 5'UTR are a known mechanism that
can trigger NMD independently of PTCs in the main CDS. We scanned the 5'UTR
sequence of each comparator isoform for uORFs (AUG followed by an in-frame stop
codon, minimum 3 codons / 9 nt). uORFs whose stop codon falls at or beyond the
main CDS start are classified as **overlapping** — a particularly potent NMD
trigger.

```{r goal3-uorf}
# ── Data provenance ──
# Source script: 05b_uorf_analysis.R → analysis_cache/uorf_analysis.rds
# Methods: METHODS.md § "5'UTR uORF Analysis"
#
# Pipeline:
#   1. 5'UTR boundaries computed by compute_utr5_length() in 05b_uorf_analysis.R
#      Input: genomic CDS coordinates (cds.rds) + exon structure (structures.rds)
#      + strand: CDS start = cds_start (plus) or cds_stop (minus)
#      Validation: 98.4% of isoforms have ATG at computed CDS start position
#   2. 5'UTR sequences extracted from SQANTI corrected transcript FASTA
#      (nmd_lungcells_corrected.fasta) via awk-based subset extraction
#   3. uORF detection by find_uorfs() in 05b_uorf_analysis.R:
#      gregexpr("ATG") finds all AUG positions in 5'UTR; for each, walks in-frame
#      through full transcript to first stop codon (TAA/TAG/TGA). Min 3 codons.
#      Overlapping uORFs: stop codon at or past main CDS start.
#   4. Groups: PTC_neg, PTC_pos, Control (same coding pair definitions as Goal 3)

uorf_data <- readRDS(file.path(cache_dir, "uorf_analysis.rds"))
uorf <- uorf_data$per_isoform
uorf$group <- factor(uorf$group, levels = c("PTC_neg", "PTC_pos", "Control"))
```

```{r goal3-uorf-summary}
uorf_summary <- uorf |>
  group_by(group) |>
  summarise(
    n = n(),
    median_utr5_bp = median(utr5_length),
    pct_with_uorf = round(100 * mean(n_uorfs > 0), 1),
    median_n_uorfs = median(n_uorfs),
    mean_n_uorfs = round(mean(n_uorfs), 1),
    pct_overlapping = round(100 * mean(n_overlapping_uorfs > 0), 1),
    .groups = "drop")

render_dt(uorf_summary, caption = "Table 7a: uORF Prevalence by Group")
```

Columns: **group** = PTC_neg (NMD without PTC), PTC_pos (NMD with PTC), or
Control (C4); **n** = number of isoforms; **median_utr5_bp** = median 5'UTR
length; **pct_with_uorf** = % with ≥1 uORF (≥3 codons); **median/mean_n_uorfs** =
uORFs per isoform; **pct_overlapping** = % with ≥1 overlapping uORF (stop
codon at or past CDS start).

```{r goal3-uorf-tests}
# ── Data provenance ──
# Source: uorf_data$per_isoform (from 05b_uorf_analysis.R)
# Methods: METHODS.md § "5'UTR uORF Analysis"
# Test: Fisher's exact test on 2x2 table (group × has_uorf) for each pair of groups
# "any uORF": isoform has ≥1 uORF (≥3 codons) in the 5'UTR
# "overlapping uORF": isoform has ≥1 uORF whose stop codon falls at or past CDS start
# n per group: PTC_neg=1,860, PTC_pos=1,458, Control=3,238

uorf_tests <- list()
test_pairs <- list(
  c("PTC_neg", "Control", "any uORF"),
  c("PTC_pos", "Control", "any uORF"),
  c("PTC_neg", "PTC_pos", "any uORF"),
  c("PTC_neg", "Control", "overlapping uORF")
)

for (tp in test_pairs) {
  sub <- uorf[uorf$group %in% tp[1:2], ]
  sub$group <- droplevels(factor(sub$group))
  if (tp[3] == "any uORF") {
    tbl <- table(sub$group, sub$n_uorfs > 0)
  } else {
    tbl <- table(sub$group, sub$n_overlapping_uorfs > 0)
  }
  ft <- tryCatch(fisher.test(tbl), error = function(e) NULL)
  if (!is.null(ft)) {
    n1 <- sum(sub$group == tp[1])
    n2 <- sum(sub$group == tp[2])
    uorf_tests[[length(uorf_tests) + 1]] <- data.frame(
      comparison = paste(tp[1], "vs", tp[2]),
      test = tp[3],
      n_group1 = n1, n_group2 = n2,
      OR = round(as.numeric(ft$estimate), 2),
      p_value = signif(as.numeric(ft$p.value), 3),
      stringsAsFactors = FALSE)
  }
}
uorf_test_df <- do.call(rbind, uorf_tests)
render_dt(uorf_test_df, caption = "Table 7b: uORF Enrichment Tests (Fisher's exact)")
```

Columns: **comparison** = group pair tested; **test** = "any uORF" (≥1 uORF) or
"overlapping uORF" (≥1 uORF with stop at/past CDS start); **n_group1/n_group2** =
group sizes; **OR** = Fisher's exact test odds ratio; **p_value** = Fisher p-value.

```{r goal3-uorf-density, fig.width=10, fig.height=6}
# Density plot (better than dodged histogram for different group sizes)
uorf_plot <- uorf[uorf$utr5_length > 0, ]
uorf_plot$group_label <- factor(uorf_plot$group,
  levels = c("PTC_neg", "PTC_pos", "Control"),
  labels = c("NMD PTC-", "NMD PTC+", "Control"))

fig_uorf <- ggplot(uorf_plot, aes(x = n_uorfs + 0.5, fill = group_label)) +
  geom_density(alpha = 0.5) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50)) +
  scale_fill_manual(values = c("NMD PTC-" = "#fddbc7", "NMD PTC+" = "#ef8a62",
                                "Control" = "#67a9cf")) +
  labs(title = "uORF Count Distribution by Group",
       subtitle = "PTC-negative NMD isoforms have dramatically more uORFs (median 8 vs 1 in Control)",
       x = "Number of uORFs (+0.5, log scale)", y = "Density", fill = "")
save_fig(fig_uorf, "fig_uorf_count_distribution", width = 10, height = 6)
```

**Key finding:** PTC-negative NMD comparators have dramatically more uORFs than
Control (96.2% vs 63.3% with any uORF, OR =
`r uorf_test_df$OR[uorf_test_df$comparison == "PTC_neg vs Control" & uorf_test_df$test == "any uORF"]`,
p < 10^-180^). They also have longer 5'UTRs (median
`r uorf_summary$median_utr5_bp[uorf_summary$group == "PTC_neg"]` bp vs
`r uorf_summary$median_utr5_bp[uorf_summary$group == "Control"]` bp in Control),
which provides more sequence for uORFs. Overlapping uORFs (stop codon at or past
main CDS start) are also enriched (61.5% vs 27.9%, OR =
`r uorf_test_df$OR[uorf_test_df$comparison == "PTC_neg vs Control" & uorf_test_df$test == "overlapping uORF"]`).

Combined with the 3'UTR length finding (PTC-neg NMD isoforms have normal 3'UTR
lengths), this points to **5'UTR uORFs as the dominant non-PTC NMD mechanism**
in this dataset, rather than long 3'UTR-mediated NMD.

**Caveat:** The high uORF count is partly driven by longer 5'UTRs (more sequence =
more chance of ATG + in-frame stop). The enrichment should be interpreted in
conjunction with the 5'UTR length difference.

**Limitations:** uORF detection is sequence-based and does not confirm
translation. Not all sequence-predicted uORFs are functional NMD triggers.
Other non-PTC NMD mechanisms (3'UTR secondary structure, EJC-independent UPF1
recruitment) cannot be ruled out from structural data alone.


# 4. Baseline Splicing and NMD-Enriched Events {#goal4}

**Goal 4:** Characterize baseline alternative splicing (Control C4 pairs) and identify which
splicing event types are enriched in NMD vs Control.

**Methods:** Structural profiles quantify splicing event types for each pair. Sequence
divergence (`quantifyPairDivergence()`) measures exonic sequence shared between isoforms.
NMD vs Control enrichment is tested by Fisher's exact test per event type.

## Figure 7: Sequence Shared — NMD vs Control

```{r goal4-fig7-pct-shared, fig.width=8, fig.height=5}
# ── Data provenance ──
# Starting data: gene-matched profiles → cached divergence
# div_c2 = set_results[["c2_allsamples"]]$divergence
#   Source: quantifyPairDivergence(gene-matched profiles, structures, cds),
#   cached by script 03
#   $pct_shared: percentage of exonic sequence shared between ref and comp
# div_c4: same for Control
# Result: density comparison of sequence similarity between NMD and Control pairs

pct_shared_data <- rbind(
  data.frame(comparison = "NMD",     pct_shared = div_c2$pct_shared),
  data.frame(comparison = "Control", pct_shared = div_c4$pct_shared))
pct_shared_data <- pct_shared_data[!is.na(pct_shared_data$pct_shared), ]

fig7 <- ggplot(pct_shared_data, aes(x = pct_shared, fill = comparison)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = pal) +
  labs(title = "Exonic Sequence Shared (%): NMD vs Control",
       subtitle = "NMD pairs shifted right = more similar (targeted splicing changes)",
       x = "Sequence Shared (%)", y = "Density", fill = "")
save_fig(fig7, "fig7_pct_shared_density")

# Summary stats
shared_summary <- pct_shared_data |>
  group_by(comparison) |>
  summarise(n = n(), median = round(median(pct_shared), 1),
            mean = round(mean(pct_shared), 1), .groups = "drop")
render_dt(shared_summary, caption = "Table 7c: Sequence Shared Summary")
```

Columns: **comparison** = NMD or Control; **n** = number of pairs; **median/mean** =
median and mean percentage of exonic sequence shared between reference and
comparator (from `quantifyPairDivergence()`).

## Table 8a: Splicing Event Enrichment — NMD vs Control

For each event type, the denominator is all gene-matched pairs (not restricted
to coding) — a pair is counted if it has ≥1 event of that type. Fisher's exact
test compares the proportion of pairs with each event type between NMD (C2) and
Control (C4), both on the same gene-matched set.

```{r goal4-table8a-event-enrichment}
# ── Data provenance ──
# Starting data: gene-matched profiles (set_results[[key]]$profiles)
# compute_event_freq(profiles) counts pairs with each event type using:
#   - Boolean columns (tss_changed, tes_changed, n_a5ss > 0, etc.) from profiles
#   - Partial_IR_5/3 from detailed_events list-column
#   These are computed by buildProfiles → compareSplicingChoices
# Fisher's exact test: event prevalence in NMD vs Control on gene-matched sets

ef_c2 <- compute_event_freq(set_results[["c2_allsamples"]]$profiles)
ef_c4 <- compute_event_freq(set_results[["c4_allsamples"]]$profiles)
n_c2 <- nrow(set_results[["c2_allsamples"]]$profiles)
n_c4 <- nrow(set_results[["c4_allsamples"]]$profiles)

enrichment <- merge(ef_c2, ef_c4, by = "event_type", suffixes = c("_NMD", "_Ctrl"))

# Fisher's exact test per event type
enrichment$fisher_OR <- NA_real_
enrichment$fisher_p <- NA_real_
for (i in seq_len(nrow(enrichment))) {
  a <- enrichment$n_pairs_with_event_NMD[i]
  b <- n_c2 - a
  c <- enrichment$n_pairs_with_event_Ctrl[i]
  d <- n_c4 - c
  ft <- tryCatch(fisher.test(matrix(c(a, b, c, d), nrow = 2)), error = function(e) NULL)
  if (!is.null(ft)) {
    enrichment$fisher_OR[i] <- round(ft$estimate, 2)
    enrichment$fisher_p[i] <- ft$p.value
  }
}

enrichment$direction <- ifelse(enrichment$fisher_p > 0.05, "Non-significant",
  ifelse(enrichment$fisher_OR > 1, "Enriched in NMD", "Enriched in Control"))
enrichment <- enrichment[order(enrichment$fisher_p), ]

enrich_display <- enrichment[, c("event_type", "n_pairs_with_event_NMD", "pct_of_pairs_NMD",
                                 "n_pairs_with_event_Ctrl", "pct_of_pairs_Ctrl",
                                 "fisher_OR", "fisher_p", "direction")]
enrich_display$fisher_p <- signif(enrich_display$fisher_p, 3)
render_dt(enrich_display,
          caption = "Table 8a: Splicing Event Type Enrichment — NMD vs Control")
write.csv(enrichment, file.path(table_dir, "table8a_event_enrichment.csv"), row.names = FALSE)
```

Columns: **event_type** = splicing event type (see Event Type Glossary);
**n_pairs_with_event_NMD** = number of gene-matched NMD (C2) pairs with ≥1
event of this type; **pct_of_pairs_NMD** = percentage of all
`r fmt(n_c2)` NMD pairs; **n_pairs_with_event_Ctrl/pct_of_pairs_Ctrl** =
same for Control (C4, `r fmt(n_c4)` pairs); **fisher_OR** = Fisher's exact
test odds ratio (>1 = NMD-enriched); **fisher_p** = p-value; **direction** =
NMD-enriched or Control-enriched.

## Table 8b: Gain/Loss Direction by Event Type

For each event type, the comparator can GAIN sequence (comparator has additional
sequence relative to reference) or LOSE sequence (comparator is missing sequence
present in reference). The table below shows event counts and the percentage of
events in each direction, separately for NMD and Control pairs.

```{r goal4-table8b-gain-loss}
# ── Data provenance ──
# Starting data: gene-matched profiles (set_results[[key]]$profiles)
# $detailed_events: list-column, one data.frame per pair from buildProfiles
#   Each event row has event_type and direction (GAIN or LOSS)
#   Direction is from comparator's perspective (see Event Type Glossary)
# For each comparison (NMD, Control): count events by type × direction,
#   compute % GAIN and % LOSS within each event type

get_gain_loss_pct <- function(profiles, label) {
  all_events <- do.call(rbind, profiles$detailed_events)
  if (is.null(all_events) || nrow(all_events) == 0) return(NULL)
  gl <- all_events |>
    group_by(event_type, direction) |>
    summarise(n = n(), .groups = "drop")
  # Compute total per event type for percentages
  totals <- gl |> group_by(event_type) |> summarise(total = sum(n), .groups = "drop")
  gl <- merge(gl, totals, by = "event_type")
  gl$pct <- round(100 * gl$n / gl$total, 1)
  gl$comparison <- label
  gl
}

gl_c2 <- get_gain_loss_pct(set_results[["c2_allsamples"]]$profiles, "NMD")
gl_c4 <- get_gain_loss_pct(set_results[["c4_allsamples"]]$profiles, "Control")
gl_all <- rbind(gl_c2, gl_c4)

# Create a readable wide format: event_type | NMD_GAIN | NMD_GAIN_% | NMD_LOSS | NMD_LOSS_% | ...
gl_wide <- gl_all |>
  tidyr::pivot_wider(
    id_cols = event_type,
    names_from = c(comparison, direction),
    values_from = c(n, pct),
    values_fill = 0) |>
  arrange(event_type)

# Select columns and add Fisher test: NMD vs Control gain/loss proportions
col_order <- c("event_type",
  "n_NMD_GAIN", "n_NMD_LOSS", "pct_NMD_GAIN",
  "n_Control_GAIN", "n_Control_LOSS", "pct_Control_GAIN")
existing_cols <- intersect(col_order, names(gl_wide))
gl_display <- gl_wide[, existing_cols, drop = FALSE]

# Fisher's exact test per event type: is the GAIN/LOSS ratio different in NMD vs Control?
gl_display$fisher_p <- NA_real_
gl_display$direction_diff <- NA_character_
for (i in seq_len(nrow(gl_display))) {
  ng <- gl_display$n_NMD_GAIN[i]
  nl <- gl_display$n_NMD_LOSS[i]
  cg <- gl_display$n_Control_GAIN[i]
  cl <- gl_display$n_Control_LOSS[i]
  if (ng + nl > 0 && cg + cl > 0) {
    ft <- tryCatch(fisher.test(matrix(c(ng, nl, cg, cl), nrow = 2)), error = function(e) NULL)
    if (!is.null(ft)) {
      gl_display$fisher_p[i] <- signif(ft$p.value, 3)
      pct_nmd <- 100 * ng / (ng + nl)
      pct_ctrl <- 100 * cg / (cg + cl)
      gl_display$direction_diff[i] <- if (ft$p.value > 0.05) "Non-significant"
        else if (pct_nmd > pct_ctrl) "More GAIN in NMD"
        else "More LOSS in NMD"
    }
  }
}

render_dt(gl_display, caption = "Table 8b: Gain/Loss Events by Type — NMD vs Control")
write.csv(gl_display, file.path(table_dir, "table8b_gain_loss_events.csv"), row.names = FALSE)
```

Columns: **event_type** = splicing event type; **n_NMD_GAIN/n_NMD_LOSS** =
count of NMD events with GAIN/LOSS direction (from comparator's perspective);
**pct_NMD_GAIN** = percentage of NMD events of that type with GAIN direction;
**n_Control_GAIN/n_Control_LOSS/pct_Control_GAIN** = same for Control;
**fisher_p** = Fisher's exact test p-value comparing GAIN/LOSS ratio between
NMD and Control; **direction_diff** = "More GAIN in NMD", "More LOSS in NMD",
or "Non-significant" (p > 0.05).


# 5. Protein-Level Consequences of Normal Alternative Splicing {#goal5}

**Goal 5:** Characterize how normal (non-NMD) alternative splicing affects
protein sequence content and function. The C4 Control pairs — dominant vs
next-best non-NMD isoforms per gene — represent the most common alternative
splicing in these cell types. When a cell uses the alternative isoform, how
different is the protein? Are functional domains affected? Is there mass-spec
evidence that the predicted novel peptides exist?

**Scope:** Unlike Goals 1-4 which use gene-matched NMD vs Control pairs, this
section analyzes **all non-NMD C4 pairs** (not gene-matched) — every gene with
≥2 non-NMD isoforms passing the 5%/filterByExpr filters. This comprises
23,132 unique coding pairs (30,902 unique isoforms) across cell types:
all_samples (8,835), AT (8,600), DD (8,239), DO (9,173), FB (8,068), MV (7,946).
This provides the broadest view of normal isoform-level protein diversity.

**Methods:** For all C4 coding pairs (both reference and comparator
protein-coding), we compare protein sequences from the SQANTI corrected protein
FASTA. The current analysis uses character-by-character prefix comparison to
find the first amino acid position where reference and comparator diverge
(`actual_shared_aa`). Domain analysis maps Pfam/InterPro domain positions (from
biomaRt, gene-level) to this protein divergence point to classify domains as
conserved, disrupted/lost, or gained. Mass-spec validation queries novel-region
tryptic peptides against PeptideAtlas (2.53M human peptides, local lookup).

*Note: The prefix comparison correctly identifies divergence for frameshifts
(everything downstream is genuinely novel) but may overcount novel protein for
in-frame events where proteins re-converge after the event.
`Isopair::mapSpliceToProtein()` provides precise per-event protein coordinate
mapping and is used in the v2 analysis pipeline. Isoform-specific domain
detection via `Isopair::runHmmscan()` with Pfam-A is available for future
refinement of domain positions.*

## Figure 8: Novel Protein Content — All C4 Comparators

```{r goal5-fig8-pop}
# ── Data provenance ──
# Source: 05d_protein_domain_analysis.R → protein_domain_full_analysis.rds
# Uses UNMATCHED C4 pairs (all genes with ≥2 non-NMD isoforms), deduplicated
# across cell types. all_samples subset shown here.
# Protein comparison: character-by-character prefix comparison of SQANTI protein
#   sequences. actual_shared_aa = position where ref and comp first diverge.
#   pct_novel_protein = % of comparator protein after divergence point.

prot_full <- readRDS(file.path(cache_dir, "protein_domain_full_analysis.rds"))
prot_as <- prot_full$protein_comparisons[prot_full$protein_comparisons$cell_type == "allsamples", ]
prot_as <- prot_as[!is.na(prot_as$actual_shared_aa), ]
n_c4_total_unmatched <- nrow(prot_as)
```

Of `r fmt(n_c4_total_unmatched)` coding C4 pairs (all_samples, unmatched — all
genes with ≥2 non-NMD isoforms), protein sequences were compared by prefix
divergence. Frame category comes from `Isopair::compareIsoformFrames()`. Pairs
classified as `non_coding` or `no_shared_cds` are excluded from the frame
category facets below.

```{r goal5-fig8-novel-protein, fig.width=10, fig.height=6}
comp_data <- prot_as

# Simplify frame category labels for plot
comp_data$frame_label <- factor(
  comp_data$frame_category,
  levels = c("same_start_in_frame", "same_start_frameshift",
             "diff_start_same_frame", "diff_start_diff_frame",
             "diff_start_diff_frame_with_frameshift"),
  labels = c("Same Start\nSame Frame", "Same Start\nFrameshift",
             "Diff Start\nSame Frame", "Diff Start\nDiff Frame",
             "Diff Start\nDiff Frame + FS"))
comp_data <- comp_data[!is.na(comp_data$frame_label), ]

fig8 <- ggplot(comp_data, aes(x = pct_novel_protein, fill = frame_label)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~frame_label, scales = "free_y", ncol = 3) +
  labs(title = "Novel Protein Content: All C4 (Control) Coding Comparators",
       subtitle = sprintf("n = %d pairs. Protein comparison via prefix divergence.", nrow(comp_data)),
       x = "Novel Protein (%)", y = "Density") +
  theme(legend.position = "none")
save_fig(fig8, "fig8_novel_protein_all_c4", width = 10, height = 6)

# Summary table
novel_summary <- comp_data |>
  group_by(frame_category) |>
  summarise(
    n = n(),
    median_novel_pct = round(median(pct_novel_protein, na.rm = TRUE), 1),
    mean_novel_pct = round(mean(pct_novel_protein, na.rm = TRUE), 1),
    pct_with_any_novel = round(100 * mean(pct_novel_protein > 0, na.rm = TRUE), 1),
    .groups = "drop")
```

## Table 9: Novel Protein Summary by Frame Category

```{r goal5-table9-novel-summary}
render_dt(novel_summary, caption = "Table 9: Novel Protein Content by Frame Category (All C4 Pairs)")
write.csv(novel_summary, file.path(table_dir, "table9_novel_protein_summary.csv"), row.names = FALSE)
```

Columns: **frame_category** = reading frame relationship (from
`compareIsoformFrames()`); **n** = number of coding C4 pairs in this category;
**median/mean_novel_pct** = median and mean percentage of novel protein
(from character-by-character prefix comparison of SQANTI protein sequences);
**pct_with_any_novel** = percentage of pairs with any novel protein content.

## Table 10: Domain Disruption by Normal Alternative Splicing

```{r goal5-domain-load}
# ── Data provenance ──
# Source script: 05f_protein_analysis_v2.R → protein_analysis_v2.rds
# Methods: METHODS.md § domain analysis (to be updated)
#
# Pipeline:
#   1. All coding C4 pairs (unmatched), deduplicated across cell types
#   2. Isopair::mapSpliceToProtein() → protein-level event regions per splice event
#   3. Pfam domains detected by hmmscan on actual isoform protein sequences
#      (isoform-specific positions, not gene-level)
#   4. Domain-event intersection: domains overlapping splice event protein regions
#      classified as disrupted_lost (LOSS events) or gained_intact (GAIN events)
#   5. Over-representation: Fisher's exact test per domain vs reference prevalence

prot_v2 <- readRDS(file.path(cache_dir, "protein_analysis_v2.rds"))
dom_ann_v2 <- prot_v2$domain_effects
protein_events <- prot_v2$protein_events
v2_meta <- prot_v2$metadata
```

```{r goal5-domain-summary}
# Compute domain summary from v2 domain effects (isoform-specific hmmscan)
n_pairs_analyzed <- n_c4_total_unmatched
n_cds_events <- v2_meta$n_cds_events
n_pairs_with_cds_events <- length(unique(paste(
  protein_events$reference_isoform_id, protein_events$comparator_isoform_id)))
n_domain_overlaps <- nrow(dom_ann_v2)
n_pairs_disrupted <- length(unique(paste(
  dom_ann_v2$comparator_isoform_id[dom_ann_v2$effect == "disrupted_lost"],
  dom_ann_v2$reference_isoform_id[dom_ann_v2$effect == "disrupted_lost"])))
n_pairs_gained <- length(unique(paste(
  dom_ann_v2$comparator_isoform_id[dom_ann_v2$effect == "gained_intact"],
  dom_ann_v2$reference_isoform_id[dom_ann_v2$effect == "gained_intact"])))

dom_disposition <- data.frame(
  category = c("At least one Pfam domain disrupted/lost",
               "At least one Pfam domain gained intact",
               "Pairs with CDS events but no domain overlap",
               "Pairs without CDS-overlapping events"),
  n_pairs = c(n_pairs_disrupted,
              n_pairs_gained,
              n_pairs_with_cds_events - n_pairs_disrupted - n_pairs_gained +
                length(intersect(
                  unique(paste(dom_ann_v2$comparator_isoform_id[dom_ann_v2$effect == "disrupted_lost"],
                               dom_ann_v2$reference_isoform_id[dom_ann_v2$effect == "disrupted_lost"])),
                  unique(paste(dom_ann_v2$comparator_isoform_id[dom_ann_v2$effect == "gained_intact"],
                               dom_ann_v2$reference_isoform_id[dom_ann_v2$effect == "gained_intact"])))),
              n_pairs_analyzed - n_pairs_with_cds_events),
  pct = NA_real_)
dom_disposition$pct <- round(100 * dom_disposition$n_pairs / n_pairs_analyzed, 1)

kable(dom_disposition, caption = "Table 10: Domain Effects — All Coding C4 Pairs (hmmscan/Pfam)")
```

Of `r fmt(n_pairs_analyzed)` coding C4 pairs (all_samples, unmatched),
`r fmt(n_pairs_with_cds_events)` have CDS-overlapping splice events (mapped
by `Isopair::mapSpliceToProtein()`). Domain positions were detected on each
isoform's actual protein sequence using hmmscan against Pfam-A (28,307 isoforms
with ≥1 domain). **`r fmt(n_pairs_disrupted)` pairs have at least one Pfam
domain disrupted or lost** by a splice event, and `r fmt(n_pairs_gained)`
have at least one domain gained intact.

Columns: **category** = domain effect status; **n_pairs** = count;
**pct** = percentage of all `r fmt(n_pairs_analyzed)` coding C4 pairs.

## Table 10b: Domain Disruption by Cell Type

```{r goal5-domain-ct}
# Per-CT: map v2 domain effects back to cell types using ct_membership
ct_mem <- prot_v2$ct_membership
ct_dom_rows <- list()
for (ct in c("allsamples", "at", "dd", "do", "fb", "mv")) {
  ct_keys <- names(ct_mem)[sapply(ct_mem, function(x) ct %in% x)]
  if (length(ct_keys) == 0) next
  ct_eff <- dom_ann_v2[paste(dom_ann_v2$reference_isoform_id,
    dom_ann_v2$comparator_isoform_id, sep = "|") %in% ct_keys, ]
  n_coding <- sum(ct_keys %in% paste(prot_v2$protein_events$reference_isoform_id,
    prot_v2$protein_events$comparator_isoform_id, sep = "|"))
  # Use total coding pairs for this CT from profiles
  n_coding_total <- length(ct_keys)
  n_disrupted <- length(unique(paste(
    ct_eff$comparator_isoform_id[ct_eff$effect == "disrupted_lost"],
    ct_eff$reference_isoform_id[ct_eff$effect == "disrupted_lost"])))
  ct_dom_rows[[ct]] <- data.frame(
    cell_type = toupper(ct), n_coding_pairs = n_coding_total,
    n_with_domain_events = nrow(ct_eff),
    n_disrupted = n_disrupted,
    pct_disrupted_of_total = round(100 * n_disrupted / max(n_coding_total, 1), 1))
}
ct_dom_summary <- do.call(rbind, ct_dom_rows)

render_dt(ct_dom_summary, caption = "Table 10b: Domain Disruption Rates by Cell Type")
```

Columns: **cell_type**; **n_coding_pairs** = total coding C4 pairs (unmatched);
**n_with_domain_events** = domain-event overlaps in this CT;
**n_disrupted** = pairs with ≥1 disrupted/lost domain; **pct_disrupted_of_total** =
% of all coding pairs with disruption.

## Table 10c: Most Frequently Disrupted Domains

```{r goal5-domain-top}
# Top disrupted Pfam domains from v2 (hmmscan, isoform-specific)
disrupted_v2 <- dom_ann_v2[dom_ann_v2$effect == "disrupted_lost", ]
top_disrupted <- disrupted_v2 |>
  dplyr::group_by(pfam_acc, domain_name) |>
  dplyr::summarise(n_pairs_disrupted = dplyr::n_distinct(
    paste(comparator_isoform_id, reference_isoform_id)), .groups = "drop") |>
  dplyr::arrange(desc(n_pairs_disrupted)) |>
  head(20)
top_disrupted$pct_of_disrupted <- round(100 * top_disrupted$n_pairs_disrupted /
  n_pairs_disrupted, 1)

render_dt(top_disrupted, caption = "Table 10c: Most Frequently Disrupted Pfam Domains (Top 20)")
```

Columns: **pfam_acc** = Pfam accession (from hmmscan on isoform protein
sequences); **domain_name** = Pfam domain name; **n_pairs_disrupted** = number
of C4 pairs where this domain overlaps a splice event's protein region;
**pct_of_disrupted** = percentage of all `r fmt(n_pairs_disrupted)` pairs with
any domain disruption.

## Figure 9: Productive Frameshifts

```{r goal5-fig9-pop}
pf_char <- pf_data$characterization
n_ss_fs <- sum(novel_summary$frame_category == "same_start_frameshift")
n_ss_fs_count <- if (n_ss_fs > 0) novel_summary$n[novel_summary$frame_category == "same_start_frameshift"] else NA
```

A special case of protein-level impact: `r fmt(n_ss_fs_count)` coding C4 pairs
(all_samples, unmatched) are `same_start_frameshift` (same start codon,
frameshift in CDS). Among the gene-matched subset,
`r fmt(nrow(pf_char))` lack a PTC in the comparator — these are **productive
frameshifts** that produce a shared N-terminal domain followed by a completely
novel C-terminal protein sequence. The detailed characterization below
(domain annotations, mass-spec) is from the gene-matched productive frameshifts
precompute (`04_productive_frameshift_precompute.R`).

```{r goal5-fig9-pf-mechanisms, fig.width=8, fig.height=5}
# ── Data provenance ──
# Source: 04_productive_frameshift_precompute.R → productive_frameshift_allsamples.rds
# Methods: METHODS.md § productive frameshifts section (to be added)

event_types_raw <- pf_char$frameshift_event_types[!is.na(pf_char$frameshift_event_types)]
all_event_types <- unlist(strsplit(event_types_raw, ","))
event_freq <- sort(table(all_event_types), decreasing = TRUE)

event_df <- data.frame(
  event_type = names(event_freq),
  count = as.integer(event_freq),
  stringsAsFactors = FALSE)
event_df$pct <- round(100 * event_df$count / length(event_types_raw), 1)

fig9 <- ggplot(event_df, aes(x = reorder(event_type, -count), y = count)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = count), vjust = -0.3, size = 3.5) +
  labs(title = sprintf("Frameshift-Causing Event Types in Productive Frameshifts (n = %d)",
                       nrow(pf_char)),
       x = "Splicing Event Type", y = "Number of Pairs")
save_fig(fig9, "fig9_productive_frameshift_events")
```

Columns: **event_type** = splice event type that caused the frameshift;
**count** = number of productive frameshift pairs with this event; **pct** =
percentage of productive frameshifts (one pair may contribute multiple events).

## Table 11: Mass Spectrometry Validation (PeptideAtlas)

For each C4 coding pair where the comparator has novel protein sequence (after
the divergence point `actual_shared_aa`), we generated tryptic peptides from
the novel region by in silico digestion (cut after K/R, not before P; peptides
7-30 AA). Peptides also present in the reference protein were excluded — only
**splice-specific** peptides unique to the comparator were queried. These were
matched against the Human PeptideAtlas (2.53 million observed peptide sequences)
by exact string match.

```{r goal5-table11-massspec}
# ── Data provenance ──
# Source: inline computation using protein comparisons from 05d +
#   PeptideAtlas (resources/peptideatlas/human_peptides_sorted.txt)
# Methods: METHODS.md to be updated
# Pipeline:
#   1. For each unique C4 pair, extract comparator protein after actual_shared_aa
#   2. In silico tryptic digest → peptides 7-30 AA
#   3. Remove peptides also found in reference protein → splice-specific peptides
#   4. Exact match against PeptideAtlas human peptide database (local lookup)

# Use v2 mass-spec results (event-based regions, PeptideAtlas local lookup)
ms_pa <- prot_v2$massspec

n_with_novel <- sum(ms_pa$n_novel_peptides > 0 | ms_pa$n_pa_hits > 0)
n_with_hits <- sum(ms_pa$n_pa_hits > 0)
total_peptides <- sum(ms_pa$n_novel_peptides)
total_hits <- sum(ms_pa$n_pa_hits)
```

```{r goal5-table11-summary}
ms_summary_tbl <- data.frame(
  metric = c("Pairs with novel protein region",
             "Pairs with unique tryptic peptides",
             "Pairs with PeptideAtlas hits",
             "Total splice-specific peptides queried",
             "Total PeptideAtlas hits"),
  n = c(nrow(ms_pa[ms_pa$n_novel_peptides > 0, ]),
        nrow(ms_pa),
        n_with_hits,
        total_peptides,
        total_hits),
  pct = c(NA, NA,
          round(100 * n_with_hits / nrow(ms_pa), 1),
          NA,
          round(100 * total_hits / total_peptides, 1)))

kable(ms_summary_tbl, caption = "Table 11a: Mass Spectrometry Validation Summary",
      col.names = c("Metric", "N", "%"))
```

Columns: **Metric** = description; **N** = count; **%** = percentage (of pairs
with unique peptides for hit rate, of total peptides for peptide hit rate).

```{r goal5-table11-top-hits}
# Top pairs by number of PeptideAtlas hits
# Add gene_id from protein_events (from v2)
gene_lk_ms <- setNames(protein_events$gene_id, protein_events$comparator_isoform_id)
gene_lk_ms <- gene_lk_ms[!duplicated(names(gene_lk_ms))]
ms_pa$gene_id <- gene_lk_ms[ms_pa$comparator_isoform_id]
ms_pa$gene_symbol <- gene_symbol_lookup[ms_pa$gene_id]
ms_hits <- ms_pa[ms_pa$n_pa_hits > 0, ]
ms_hits <- ms_hits[order(-ms_hits$n_pa_hits), ]

render_dt(head(ms_hits[, c("gene_symbol", "comparator_isoform_id",
                            "n_novel_peptides", "n_pa_hits")], 50),
          caption = "Table 11b: Top 50 Pairs by PeptideAtlas Hits")
```

Columns: **gene_symbol**; **comparator_isoform_id**; **n_novel_peptides** =
splice-specific tryptic peptides from event-affected regions (not in reference);
**n_pa_hits** = number of those peptides found in PeptideAtlas.

**Key finding:** `r fmt(n_with_hits)` of `r fmt(nrow(ms_pa))` C4 pairs with
splice-specific peptides (`r round(100 * n_with_hits / nrow(ms_pa), 1)`%) have
at least one peptide previously detected by mass spectrometry. This is direct
evidence that normal alternative splicing produces translated proteins with
novel peptide sequences at scale.


# 6. Cross-Cell-Type Consistency Summary {#summary}

```{r sec6-summary-table}
# ── Data provenance ──
# Starting data: gene-matched profiles and cached analyses across all cell types
# Each row computed via helper functions that access set_results[[key]]:
#   - Pair counts: nrow(gene-matched profiles)
#   - PTC rates: compute_ptc_rates_row() — coding pairs from gene-matched profiles + ptc.rds
#   - Divergence: compute_divergence_row() — from cached quantifyPairDivergence()
#   - Productive frameshifts: from pf_data precompute (all_samples only)
# All helpers operate on gene-matched data (profiles_list was matched in STAGE 2)

summary_rows <- lapply(c("all_samples", main_cts), function(ct) {
  ct_label <- ct_to_label(ct)

  # NMD and Control pair counts
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  n_nmd  <- if (c2_key %in% names(set_results)) nrow(set_results[[c2_key]]$profiles) else NA
  n_ctrl <- if (c4_key %in% names(set_results)) nrow(set_results[[c4_key]]$profiles) else NA

  # PTC rates
  ptc_row <- compute_ptc_rates_row(ct)
  ptc_rate_nmd  <- if (!is.null(ptc_row)) ptc_row$NMD_comp_ptc_pct else NA
  ptc_rate_ctrl <- if (!is.null(ptc_row)) ptc_row$Ctrl_comp_ptc_pct else NA
  ptc_or        <- if (!is.null(ptc_row)) ptc_row$fisher_OR else NA

  # PTC-neg %
  ptc_neg_pct <- if (!is.null(ptc_row)) round(100 - ptc_row$NMD_comp_ptc_pct, 1) else NA

  # Median pct_shared for C4
  div_row <- compute_divergence_row(ct)
  med_shared <- if (!is.null(div_row)) div_row$Ctrl_median_pct_shared else NA

  # Productive frameshifts (only available for all_samples)
  n_pf <- if (ct == "all_samples") nrow(pf_data$characterization) else NA

  data.frame(
    cell_type = ct,
    N_NMD_pairs = n_nmd,
    N_Ctrl_pairs = n_ctrl,
    PTC_rate_NMD_pct = ptc_rate_nmd,
    PTC_rate_Ctrl_pct = ptc_rate_ctrl,
    PTC_OR = ptc_or,
    PTC_neg_pct = ptc_neg_pct,
    Median_pct_shared_C4 = med_shared,
    N_productive_FS = n_pf)
})

summary_table <- do.call(rbind, summary_rows)
render_dt(summary_table, caption = "Table 12: Cross-Cell-Type Summary")
write.csv(summary_table, file.path(table_dir, "table12_cross_ct_summary.csv"), row.names = FALSE)
```

Columns: **cell_type** = sample set; **N_NMD/N_Ctrl_pairs** = gene-matched pair
counts; **PTC_rate_NMD/Ctrl_pct** = PTC rate among coding comparators;
**PTC_OR** = Fisher OR for PTC enrichment; **PTC_neg_pct** = percentage of
coding NMD comparators without PTC; **Median_pct_shared_C4** = median exonic
sequence shared in Control pairs (from `quantifyPairDivergence()`);
**N_productive_FS** = productive frameshifts (all_samples only).

# Session Info

```{r session-info}
sessionInfo()
```
