---
title: "NMD analysis: current GENCODE-only scopes (2026-06-15)"
subtitle: "Canonical scopes for Figure 3 (D/E/F) + Figure 4 (A/B at all-3-ENST + coding-CDS n=190; C/D at ref-AUG-traceable re-intersected n=1,166/1,166). Branched from 05_final_report_mashr.Rmd; that report is now legacy."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)

NMD_ROOT        <- Sys.getenv("NMD_ROOT",
                              "/Users/petecastaldi/claude_projects/nmd")
ISOPAIR_WRAPPER <- file.path(NMD_ROOT, "results/isoform_transitions/Version_6.0/isopair_wrapper")
DM              <- file.path(ISOPAIR_WRAPPER, "data_mashr")
FASTA_PATH      <- file.path(NMD_ROOT, "sqanti/nmd_lungcells/results/nmd_lungcells_corrected.fasta")

stopifnot(
  "NMD_ROOT does not resolve to an existing directory" = dir.exists(NMD_ROOT),
  "ISOPAIR_WRAPPER does not resolve"                   = dir.exists(ISOPAIR_WRAPPER),
  "DM does not resolve"                                = dir.exists(DM),
  "FASTA_PATH does not exist (needed for §4b Kozak PWM)" = file.exists(FASTA_PATH)
)

library(Isopair)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

stopifnot(packageVersion("Isopair") >= "0.99.4")

source(file.path(ISOPAIR_WRAPPER, "analysis_functions.R"))

profiles_c2 <- readRDS(file.path(DM, "profiles_c2_allsamples.rds"))
profiles_c4 <- readRDS(file.path(DM, "profiles_c4_allsamples.rds"))
cds         <- as.data.table(readRDS(file.path(DM, "cds.rds")))
structures  <- as.data.table(readRDS(file.path(DM, "structures.rds")))
ref_atg     <- readRDS(file.path(DM, "analysis_cache/ref_atg_analysis.rds"))
utr5_refaug <- readRDS(file.path(DM, "analysis_cache/utr5_features_refaug.rds"))
utr5_all    <- readRDS(file.path(DM, "analysis_cache/utr5_features_all.rds"))
expr_mat    <- readRDS(file.path(DM, "expression_data.rds"))
nmd_class   <- readRDS(file.path(DM, "nmd_classification.rds"))
dmso_samples <- readRDS(file.path(DM, "dmso_samples.rds"))

need_in <- function(obj, cols, label) {
  missing <- setdiff(cols, colnames(obj))
  if (length(missing) > 0)
    stop(sprintf("%s missing columns: %s", label,
                  paste(missing, collapse = ", ")))
  invisible(TRUE)
}

# profiles_c{2,4} schema
need_in(profiles_c2,
        c("gene_id", "reference_isoform_id", "comparator_isoform_id"),
        "profiles_c2")
need_in(profiles_c4,
        c("gene_id", "reference_isoform_id", "comparator_isoform_id"),
        "profiles_c4")

# cds schema (used in §2a own-stop conversion, §2d CDS length, §4 TD2 ORF)
need_in(cds,
        c("isoform_id", "coding_status", "cds_start", "cds_stop",
          "strand", "orf_length"),
        "cds")

# structures schema (exon coords for genomicToTranscript + tx-length)
need_in(structures,
        c("isoform_id", "exon_starts", "exon_ends", "strand"),
        "structures")

# utr5 caches
need_in(utr5_refaug$isoform_features,
        c("isoform_id", "utr5_length", "longest_orf_nt"),
        "utr5_refaug$isoform_features")
need_in(utr5_all$isoform_features,
        c("isoform_id", "utr5_length", "longest_orf_nt"),
        "utr5_all$isoform_features")

# ref_atg slots + columns
stopifnot("ref_atg missing c2 slot" = "c2" %in% names(ref_atg))
stopifnot("ref_atg missing c4 slot" = "c4" %in% names(ref_atg))
need_in(as.data.frame(ref_atg$c2),
        c("comparator_isoform_id", "reference_isoform_id", "gene_id",
          "category", "comp_orf_length", "ref_orf_length",
          "comp_orf_length_from_ref_atg", "ref_atg_genomic",
          "ref_atg_exonic_in_comp", "original_ptc",
          "utr3_via_non_ptc_stop_nt"),
        "ref_atg$c2")
need_in(as.data.frame(ref_atg$c4),
        c("comparator_isoform_id", "reference_isoform_id", "gene_id",
          "category", "utr3_via_non_ptc_stop_nt"),
        "ref_atg$c4")

format_p <- function(p, digits = 2) {
  if (is.na(p)) return("NA")
  if (p < 1e-3) formatC(p, format = "e", digits = digits)
  else formatC(signif(p, digits + 1), format = "g")
}
```

```{r mechanism-helpers, include=FALSE}
# Inline-replicated from figures/lib/mechanism_class.R (per v4 plan C2: kept
# inline so the Rmd is self-contained; the two copies are simple enough to
# compare in one diff if they ever drift).

MECHANISM_CLASS_LEVELS <- c(
  "NMD+/PTC+",
  "NMD+/PTC- ORF same",
  "NMD+/PTC- ORF extended",
  "NMD+/PTC- ORF truncated",
  "Ref AUG absent"
)

MECHANISM_CLASS_4_LEVELS <- c(
  "NMD+/PTC+",
  "NMD+/PTC- ORF match",
  "NMD+/PTC- ORF diff",
  "Ref AUG absent"
)

REF_AUG_PRESENT_CATEGORIES <- c(
  "effectively_ptc", "no_downstream_ejc", "truncated_no_ejc"
)

mechanism_class <- function(category, comp_orf_length, ref_orf_length) {
  stopifnot(length(category) == length(comp_orf_length),
            length(category) == length(ref_orf_length))
  orf_diff <- comp_orf_length - ref_orf_length
  cls <- rep(NA_character_, length(category))
  cls[is.na(category) | !category %in% REF_AUG_PRESENT_CATEGORIES] <-
    "Ref AUG absent"
  cls[!is.na(category) & category == "effectively_ptc"] <- "NMD+/PTC+"
  ndj_mask <- !is.na(category) & category == "no_downstream_ejc"
  trc_mask <- !is.na(category) & category == "truncated_no_ejc"
  cls[ndj_mask & !is.na(orf_diff) & orf_diff == 0] <- "NMD+/PTC- ORF same"
  cls[ndj_mask & !is.na(orf_diff) & orf_diff > 0]  <- "NMD+/PTC- ORF extended"
  cls[trc_mask & !is.na(orf_diff) & orf_diff < 0]  <- "NMD+/PTC- ORF truncated"
  # ndj/trc with NA orf_diff: route to PTC- (mech4 merges to NMD+/PTC-).
  # Synced with figures/lib/mechanism_class.R.
  cls[ndj_mask & is.na(orf_diff)] <- "NMD+/PTC- ORF same"
  cls[trc_mask & is.na(orf_diff)] <- "NMD+/PTC- ORF truncated"
  cls[is.na(cls)] <- "Ref AUG absent"
  factor(cls, levels = MECHANISM_CLASS_LEVELS)
}

mechanism_class_4 <- function(cls) {
  cls_char <- as.character(cls)
  out <- ifelse(cls_char %in% c("NMD+/PTC- ORF extended",
                                "NMD+/PTC- ORF truncated"),
                "NMD+/PTC- ORF diff",
                ifelse(cls_char == "NMD+/PTC- ORF same",
                       "NMD+/PTC- ORF match",
                       cls_char))
  factor(out, levels = MECHANISM_CLASS_4_LEVELS)
}

ref_source <- function(reference_isoform_id) {
  out <- ifelse(grepl("^ENST", reference_isoform_id), "GENCODE", "TD2")
  factor(out, levels = c("GENCODE", "TD2"))
}

TES_CONCORDANT_THRESHOLD_NT <- 50L
tes_concordant <- function(comp_tx_len, ref_tx_len,
                           threshold = TES_CONCORDANT_THRESHOLD_NT) {
  abs(comp_tx_len - ref_tx_len) <= threshold
}
```

# Overview {.unnumbered}

This report supersedes `05_final_report_mashr.Rmd` (now legacy-bannered) for the
two canonical scopes used in the manuscript's Figure 3 and Figure 4:

1. **all-3-ENST + coding-CDS (n=190)** — both NMD+ and Control comparators
   have ENST reference isoforms with a complete coding CDS. Drives Figure 3
   D/E/F and Figure 4 Panels A/B + the `CDSand3UTR_GENCODEonly` supplement.
2. **ref-AUG-traceable, re-intersected on `(gene_id, reference_isoform_id)`
   after category filter (n=1,166/1,166)** — drives Figure 4 Panels C/D and
   the `TD2BiasEvidence` supplement (`n=1,166` broad, `n=492` occult-PTC).

Analyses kept in the legacy Rmd (per v4 plan): per-cell-type reclassification,
two-pass TD2→ref-AUG handoff, the "85% combined" framework, the n=1,904/22,335
GENCODE-only 3'UTR outside-set, and all ORF / model analyses
(`06_orf_analysis_mashr.Rmd`).

# §1 — Isopair construction and pair sets

The Stage-2 gene-matched pair cohort (`pop_BC`) is the foundation of every
analysis in this report. Construction proceeds in three stages, replicated
inline so the cohort can be regenerated without re-running upstream scripts:

1. **Upstream isoform calling and filtering** (`01_prepare_data_mashr.R`) —
   the long-read isocall count matrix (`nmd_isocall.count_matrix.txt`,
   `r format(645273L, big.mark=",")` isoforms × 38 samples; constant defined
   at the head of `01_prepare_data_mashr.R`) is reduced by
   removing the 2 DO outlier samples, applying TMM normalization, condition-
   stratified 5% isoform filter, and `edgeR::filterByExpr`. Output:
   `expression_data.rds` with `r format(nrow(expr_mat), big.mark=",")`
   filtered isoforms × `r ncol(expr_mat)` retained samples (6 cell types
   in the matrix; the 4-CT manuscript scope — AT, DD, FB, MV — covers
   **26 of those 36 samples**: AT 6 (3 DMSO + 3 SMG1i), DD 8 (4 + 4),
   FB 6 (3 + 3), MV 6 (3 + 3). The other 10 samples — 6 DD_ALI + 4 DO —
   are in the matrix but excluded from downstream pair construction and
   mashr classification per the manuscript scope).

2. **Pair construction** (`02_build_profiles_mashr.R`) — `profiles_c2`
   pairs the top NMD-sensitive isoform per gene against its dominant non-NMD
   reference; `profiles_c4` pairs the top-two non-NMD isoforms per gene.
   Cache: `profiles_c{2,4}_allsamples.rds` (c2 = `r format(nrow(profiles_c2), big.mark=",")`
   pre-match rows; c4 = `r format(nrow(profiles_c4), big.mark=",")`).

3. **Gene-matching** (inline below; replicates `03_nmd_analysis_mashr.Rmd`
   Stage-2 filter) — restrict both arms to the shared `(gene_id,
   reference_isoform_id)` set so every gene contributes both a NMD pair and
   a Control pair under the same reference.

```{r sec1-pop-bc, include=FALSE}
pop_bc_shared <- dplyr::inner_join(
  unique(profiles_c2[, c("gene_id", "reference_isoform_id")]),
  unique(profiles_c4[, c("gene_id", "reference_isoform_id")]),
  by = c("gene_id", "reference_isoform_id")
)
pop_bc_c2 <- dplyr::semi_join(profiles_c2, pop_bc_shared,
                              by = c("gene_id", "reference_isoform_id"))
pop_bc_c4 <- dplyr::semi_join(profiles_c4, pop_bc_shared,
                              by = c("gene_id", "reference_isoform_id"))

ref_ids_pop  <- unique(pop_bc_c2$reference_isoform_id)
nmd_ids_pop  <- unique(pop_bc_c2$comparator_isoform_id)
ctrl_ids_pop <- unique(pop_bc_c4$comparator_isoform_id)
```

`pop_BC` = `r format(nrow(pop_bc_shared), big.mark=",")` shared `(gene,
reference)` pairs; `r format(nrow(pop_bc_c2), big.mark=",")` NMD-comp pairs
and `r format(nrow(pop_bc_c4), big.mark=",")` Control-comp pairs — i.e.,
every gene contributes exactly one of each.

### Cohort cascade — pop_BC through Section A / Section C / TD2 branches

```{r sec1-flowchart-data, include=FALSE}
# Pull canonical counts for the flowchart (NMD classification + pair
# construction stages) from the loaded caches. The §2 / §3 scope counts and
# the TD2 broad / occult-PTC scope counts are taken from the canonical Pass-7
# verifier manifest at reproducibility/verify_pass7_new_rmd.R — these are
# guarded against drift by both verifiers.
n_filtered    <- nrow(expr_mat)                       # 123,790 isoforms
n_samples_4ct <- 26L                                  # 4-CT manuscript scope
n_dmso_4ct    <- 13L                                  # AT 3 + DD 4 + FB 3 + MV 3
n_smg1i_4ct   <- 13L
n_nmd_as      <- length(nmd_class[["all_samples"]]$nmd)
n_nonnmd_as   <- length(nmd_class[["all_samples"]]$non_nmd)
n_c2_before   <- nrow(profiles_c2)
n_c4_before   <- nrow(profiles_c4)
fmt <- function(x) format(x, big.mark = ",")
```

```{r sec1-flowchart, fig.width=11, fig.height=14, echo=FALSE}
dot_spec <- sprintf('
digraph cohort_cascade {
  graph [rankdir=TB, fontname="Helvetica", nodesep=0.5, ranksep=0.6, splines=ortho]
  node  [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=12, margin="0.18,0.10"]
  edge  [color="#666666", penwidth=1.6, fontname="Helvetica", fontsize=10, fontcolor="#333"]

  // ─── Upstream chain ────────────────────────────────────────────────
  A [label=<<b>4-CT manuscript-scope cohort</b><br/>
            %d samples (AT 6  ·  DD 8  ·  FB 6  ·  MV 6)<br/>
            %s expressed isoforms (after TMM + 5%% filter)>,
     fillcolor="#e8eef4", color="#2f4858"]

  B [label=<<b>mashr NMD classification</b><br/>
            NMD-sensitive: %s isoforms (union across CTs)<br/>
            Non-NMD: %s isoforms (intersection across CTs)>,
     fillcolor="#fff7bc", color="#d95f0e"]

  C [label=<<b>Pair construction</b> (Isopair)<br/>
            c4: top-two non-NMD isoforms per gene<br/>
            c2: top NMD isoform vs the same reference<br/>
            <font point-size="11">c2 pre-match: %s pairs  ·  c4 pre-match: %s pairs</font>>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  HUB [label=<<b><font point-size="15">pop_BC</font></b><br/>
              Stage-2 gene-matched on (gene, reference)<br/>
              <font point-size="14"><b>3,009 NMD pairs    /    3,009 Control pairs</b></font>>,
       fillcolor="#1f3a5f", fontcolor="white", color="#0a1a30", penwidth=2.0]

  A -> B [label=" per-CT mashr + 4-CT union "]
  B -> C [label=" generatePairsExpression "]
  C -> HUB [label=" inner_join on (gene, reference) "]

  // ─── Section A branch (left) ───────────────────────────────────────
  SA_H [label=<<b>SECTION A scope</b><br/>
                all-3-ENST + coding-CDS<br/>
                <font point-size="11">re-intersect on (gene, reference)</font>>,
        fillcolor="#ef8a62", color="#7d4030", fontcolor="#1a0a05"]
  SA1 [label=<+ all-3-ENST (ref AND comp)<br/>
              re-intersect → <b>301 / 301</b>>,
       fillcolor="#f5b294", color="#a2604a"]
  SA2 [label=<+ coding-CDS everywhere<br/>
              re-intersect → <b>190 / 190</b>>,
       fillcolor="#f5b294", color="#a2604a"]
  SA_CL [label=<<b>own-GENCODE-stop classifier</b><br/>
                50-nt last-EJC rule>,
         fillcolor="#fff7bc", color="#d95f0e"]
  SA_PTC  [label=<<b>NMD+/PTC+</b><br/>n = 72  (37.9%%)>,
           fillcolor="#ef8a62", color="#7d4030"]
  SA_NPTC [label=<<b>NMD+/PTC−</b><br/>n = 118>,
           fillcolor="#fee08b", color="#a07a30"]
  SA_CTL  [label=<<b>Control</b><br/>n = 190>,
           fillcolor="#92c5de", color="#406080"]
  SA_STAT [label=<Fisher OR = 28.2  ·  p = 1.88e−20  ·  18× enrichment>,
           shape=plaintext, fillcolor="white"]
  SA_F3 [label=<<b>Figure 3</b><br/>Panels D / E / F<br/>mechanism 38 / 23 / 8>,
         fillcolor="#dddddd", color="#666"]
  SA_F4 [label=<<b>Figure 4</b><br/>Panels A / B<br/>5\\\'UTR length + ORF>,
         fillcolor="#dddddd", color="#666"]
  SA_SUP [label=<<b>CDS / 3\\\'UTR supplement</b><br/>Row 1 (A / B / C)<br/>n = 190>,
          fillcolor="#dddddd", color="#666"]

  HUB -> SA_H [label="  ENST + coding-CDS  ", color="#7d4030"]
  SA_H -> SA1 -> SA2 -> SA_CL -> {SA_PTC, SA_NPTC, SA_CTL} [color="#7d4030"]
  SA_CL -> SA_STAT [style=dotted, arrowhead=none, color="#999"]
  {SA_PTC, SA_NPTC, SA_CTL} -> SA_F3 [style=dashed, color="#888"]
  {SA_PTC, SA_NPTC, SA_CTL} -> SA_F4 [style=dashed, color="#888"]
  {SA_PTC, SA_NPTC, SA_CTL} -> SA_SUP [style=dashed, color="#888"]

  // ─── Section C branch (right) ──────────────────────────────────────
  SC_H [label=<<b>SECTION C scope</b><br/>
                ENST-reference + ref-AUG-traceable<br/>
                <font point-size="11">re-intersect AFTER category filter</font>>,
        fillcolor="#3690c0", color="#1f5570", fontcolor="white"]
  SC1 [label=<+ ENST-reference only<br/>
              pop_BC ENST-ref subset → <b>2,098 / 2,098</b>>,
       fillcolor="#a6cee3", color="#406080"]
  SC2 [label=<+ ref-AUG-traceable category per side<br/>
              (effectively_ptc / no_downstream_ejc / truncated_no_ejc)<br/>
              pre re-intersect: <b>1,659 NMD  /  1,286 Control</b>>,
       fillcolor="#a6cee3", color="#406080"]
  SC3 [label=<re-intersect on (gene, ref) → <b>1,166 / 1,166</b>>,
       fillcolor="#a6cee3", color="#406080"]
  SC_CL [label=<<b>mechanism_class_4 classifier</b><br/>
                (figures/lib/mechanism_class.R)>,
         fillcolor="#fff7bc", color="#d95f0e"]
  SC_PTC  [label=<<b>NMD+/PTC+</b><br/>n = 1,050  (90.1%%)>,
           fillcolor="#ef8a62", color="#7d4030"]
  SC_NPTC [label=<<b>NMD+/PTC−</b><br/>n = 116>,
           fillcolor="#fee08b", color="#a07a30"]
  SC_CTL  [label=<<b>Control</b><br/>n = 1,166>,
           fillcolor="#92c5de", color="#406080"]
  SC_F4 [label=<<b>Figure 4</b><br/>Panels C / D<br/>ref-AUG 5\\\'UTR + ORF>,
         fillcolor="#dddddd", color="#666"]
  SC_SUP [label=<<b>CDS / 3\\\'UTR supplement</b><br/>Row 2 (D / E / F)<br/>n = 1,166>,
          fillcolor="#dddddd", color="#666"]

  HUB -> SC_H [label="  ENST-reference  ", color="#1f5570"]
  SC_H -> SC1 -> SC2 -> SC3 -> SC_CL -> {SC_PTC, SC_NPTC, SC_CTL} [color="#1f5570"]
  {SC_PTC, SC_NPTC, SC_CTL} -> SC_F4 [style=dashed, color="#888"]
  {SC_PTC, SC_NPTC, SC_CTL} -> SC_SUP [style=dashed, color="#888"]

  // ─── TD2 cascade (branches off Section C scope) ────────────────────
  TD2_BR [label=<<b>TD2BiasEvidence — broad</b><br/>
                  Section C ∩ TD2 CDS ∩ ref-AUG ORF ∩ ref AUG exonic<br/>
                  <b>n = 1,166</b>>,
          fillcolor="#a48bbf", color="#4a3556"]
  TD2_OC [label=<<b>TD2BiasEvidence — occult-PTC</b><br/>
                  broad ∩ effectively_ptc ∩ original_ptc = FALSE<br/>
                  <b>n = 492</b><br/>
                  <font point-size="11">TD2/ref ORF 7.73×  ·  ref &gt; TD2 Kozak 384 / 492 (78%%)<br/>
                  paired p = 2.6e−38  ·  TD2 downstream of ref AUG 99.0%%</font>>,
          fillcolor="#7a3f8f", color="#3a1e44", fontcolor="white"]

  SC3 -> TD2_BR [label="  + TD2 CDS available  ", color="#4a3556"]
  TD2_BR -> TD2_OC [label="  + occult-PTC filter  ", color="#4a3556"]

  // ─── Rank constraints — encourage Section A and Section C side-by-side
  { rank=same; SA_H; SC_H }
  { rank=same; SA1; SC1 }
  { rank=same; SA2; SC2 }
  { rank=same; SA_CL; SC_CL }
  { rank=same; SA_PTC, SA_NPTC, SA_CTL; SC_PTC, SC_NPTC, SC_CTL }
}',
  n_samples_4ct, fmt(n_filtered),
  fmt(n_nmd_as), fmt(n_nonnmd_as),
  fmt(n_c2_before), fmt(n_c4_before))

if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(dot_spec, width = 1100, height = 1400)
} else {
  message("DiagrammeR not installed — flowchart not rendered")
}
```

The figure-consumer endpoints (gray boxes) tie each cohort to the figures and
supplements that consume it: Figure 3 D/E/F and Figure 4 A/B + the
CDS/3'UTR supplement Row 1 from Section A; Figure 4 C/D + the
CDS/3'UTR supplement Row 2 from Section C; and the TD2BiasEvidence
supplement from the n=1,166 → n=492 occult-PTC chain.

## §1a — Transcript-length comparison

Lengths are recomputed inline from `structures.rds`. The legacy Rmd reported
slightly different values because it used the full pre-match c4 set; the
pop_BC-restricted Control median is the correct comparator.

```{r sec1-tx-length, include=FALSE}
tx_lengths_all <- setNames(
  vapply(seq_len(nrow(structures)), function(i)
    sum(structures$exon_ends[[i]] - structures$exon_starts[[i]] + 1L),
    integer(1)),
  structures$isoform_id
)

# pop_BC is gene-matched: each gene contributes exactly one (ref, NMD comp,
# Control comp) triplet. Build the triplet table so Wilcoxon tests are
# PROPERLY PAIRED on the gene-matched design (Pass 2 / M1 fix).
pop_bc_triplets <- merge(
  pop_bc_c2[, c("gene_id", "reference_isoform_id", "comparator_isoform_id")],
  pop_bc_c4[, c("gene_id", "reference_isoform_id", "comparator_isoform_id")],
  by = c("gene_id", "reference_isoform_id"),
  suffixes = c("_nmd", "_ctrl"))
pop_bc_triplets$ref_len  <- tx_lengths_all[pop_bc_triplets$reference_isoform_id]
pop_bc_triplets$nmd_len  <- tx_lengths_all[pop_bc_triplets$comparator_isoform_id_nmd]
pop_bc_triplets$ctrl_len <- tx_lengths_all[pop_bc_triplets$comparator_isoform_id_ctrl]
pop_bc_triplets <- pop_bc_triplets[
  complete.cases(pop_bc_triplets[, c("ref_len", "nmd_len", "ctrl_len")]), ]

ref_lens  <- pop_bc_triplets$ref_len
nmd_lens  <- pop_bc_triplets$nmd_len
ctrl_lens <- pop_bc_triplets$ctrl_len

len_summary <- data.frame(
  Group   = c("Reference (non-NMD)", "NMD comparator", "Control comparator"),
  n       = c(length(ref_lens), length(nmd_lens), length(ctrl_lens)),
  Median  = c(median(ref_lens), median(nmd_lens), median(ctrl_lens)),
  Q25     = c(unname(quantile(ref_lens, 0.25)),
              unname(quantile(nmd_lens, 0.25)),
              unname(quantile(ctrl_lens, 0.25))),
  Q75     = c(unname(quantile(ref_lens, 0.75)),
              unname(quantile(nmd_lens, 0.75)),
              unname(quantile(ctrl_lens, 0.75))),
  row.names = NULL
)
len_pw <- data.frame(
  Comparison = c("NMD vs Reference", "NMD vs Control", "Control vs Reference"),
  p_value    = c(
    wilcox.test(nmd_lens,  ref_lens,  paired = TRUE, exact = FALSE)$p.value,
    wilcox.test(nmd_lens,  ctrl_lens, paired = TRUE, exact = FALSE)$p.value,
    wilcox.test(ctrl_lens, ref_lens,  paired = TRUE, exact = FALSE)$p.value
  )
)
```

Median transcript length (nt):
**Reference** = `r format(median(ref_lens), big.mark=",")`,
**NMD comparator** = `r format(median(nmd_lens), big.mark=",")`,
**Control comparator** = `r format(median(ctrl_lens), big.mark=",")`.

Paired Wilcoxon (each gene contributes one (ref, NMD comp, Control comp)
triplet, so paired tests are the correct comparator on this gene-matched
design): NMD vs Reference `r format_p(len_pw$p_value[1])`;
NMD vs Control `r format_p(len_pw$p_value[2])`;
Control vs Reference `r format_p(len_pw$p_value[3])`.

```{r sec1-tx-length-table, echo=FALSE}
knitr::kable(len_summary, caption = "Transcript length by isoform role (pop_BC scope, nt)",
             digits = 0, format.args = list(big.mark = ","))
knitr::kable(len_pw, caption = "Paired Wilcoxon: transcript length (gene-matched pairs)")
```

## §1b — Annotation status by role

```{r sec1-annotation, include=FALSE}
is_novel_iso <- function(ids) grepl("[.]novel[0-9]+$", ids)

ann_table <- data.frame(
  Role = c("Reference (non-NMD)", "NMD comparator", "Control comparator"),
  n    = c(length(ref_ids_pop), length(nmd_ids_pop), length(ctrl_ids_pop)),
  ENST = c(sum(!is_novel_iso(ref_ids_pop)),
           sum(!is_novel_iso(nmd_ids_pop)),
           sum(!is_novel_iso(ctrl_ids_pop))),
  Novel = c(sum(is_novel_iso(ref_ids_pop)),
            sum(is_novel_iso(nmd_ids_pop)),
            sum(is_novel_iso(ctrl_ids_pop)))
)
ann_table$pct_ENST  <- round(100 * ann_table$ENST  / ann_table$n, 1)
ann_table$pct_Novel <- round(100 * ann_table$Novel / ann_table$n, 1)
```

References are predominantly GENCODE-annotated
(`r ann_table$pct_ENST[ann_table$Role == "Reference (non-NMD)"]`% ENST,
`r format(ann_table$ENST[ann_table$Role == "Reference (non-NMD)"], big.mark=",")` /
`r format(ann_table$n[ann_table$Role == "Reference (non-NMD)"], big.mark=",")`),
while NMD comparators are predominantly novel
(`r ann_table$pct_Novel[ann_table$Role == "NMD comparator"]`%) — consistent with
NMD-sensitive transcripts being underrepresented in reference annotations.

```{r sec1-annotation-table, echo=FALSE}
knitr::kable(ann_table, caption = "Annotation status by role (pop_BC scope)")
```

The Reference %ENST and the NMD-comparator %Novel define the scope reduction
that motivates §2 (the GENCODE-only n=190 scope: where ALL three isoforms are
ENST-annotated, which constrains the cohort sharply).

## §1c — Per-gene isoform diversity and reference-isoform dominance

The §4¶1 descriptive numbers (median expressed-isoforms per gene; reference-
isoform share of parent-gene expression) are computed here from the same
pop_BC universe used for transcript-length in §1a, so the SF
*PairSetDescriptives* draws from these canonical chunks rather than
recomputing.

```{r sec1-iso-per-gene, include=FALSE}
# Expressed non-NMD isoforms per gene across the pop_BC universe.
# Filter: gene in pop_BC; isoform has a row in the TMM-normalized count matrix
# (expression_data.rds); NOT classified as NMD-sensitive by mashr 4-CT union.
nmd_iso_set <- nmd_class[["all_samples"]]$nmd
expr_iso    <- rownames(expr_mat)

sec1_iso <- data.table::as.data.table(structures)[
  gene_id %in% pop_bc_shared$gene_id &
    isoform_id %in% expr_iso &
    !isoform_id %in% nmd_iso_set,
  .(isoform_id, gene_id)]

iso_per_gene <- sec1_iso[, .(n_isoforms = .N), by = gene_id]
# Genes with zero matches keep a 0 count (rare; preserves denominator).
all_pop_genes <- data.table::data.table(gene_id = unique(pop_bc_shared$gene_id))
iso_per_gene <- merge(all_pop_genes, iso_per_gene, by = "gene_id", all.x = TRUE)
iso_per_gene[is.na(n_isoforms), n_isoforms := 0L]

iso_med <- as.integer(median(iso_per_gene$n_isoforms))
iso_iqr <- as.integer(IQR(iso_per_gene$n_isoforms))
```

Across the `r format(nrow(iso_per_gene), big.mark=",")` pop_BC genes, the
median number of expressed non-NMD isoforms per gene is **`r iso_med`**
(IQR = `r iso_iqr`). The pop_BC ≥3-isoform inclusion criterion sets the floor.

```{r sec1-ref-share, include=FALSE}
# Reference-isoform DMSO 4-CT mean expression as a fraction of the gene's
# total non-NMD DMSO 4-CT mean expression. The same denominator (non-NMD
# expressed isoforms of the gene) used in sec1-iso-per-gene.
dmso_4ct_samples <- unique(c(dmso_samples$AT, dmso_samples$DD,
                              dmso_samples$FB, dmso_samples$MV))
stopifnot(all(dmso_4ct_samples %in% colnames(expr_mat)))

dmso_means <- rowMeans(expr_mat[, dmso_4ct_samples, drop = FALSE])
sec1_iso[, mean_dmso := dmso_means[isoform_id]]
sec1_iso[is.na(mean_dmso), mean_dmso := 0]
gene_total_dmso <- sec1_iso[, .(gene_total = sum(mean_dmso)), by = gene_id]

ref_share <- data.table::data.table(
  gene_id              = pop_bc_c2$gene_id,
  reference_isoform_id = pop_bc_c2$reference_isoform_id
)
ref_share <- merge(ref_share, gene_total_dmso, by = "gene_id", all.x = TRUE)
ref_share[, ref_dmso := dmso_means[reference_isoform_id]]
ref_share[is.na(ref_dmso), ref_dmso := 0]
ref_share[, share := ifelse(gene_total > 0, ref_dmso / gene_total, NA_real_)]

ref_share_med_pct  <- round(100 * median(ref_share$share, na.rm = TRUE))
ref_share_ge50_pct <- round(100 * mean(ref_share$share >= 0.50, na.rm = TRUE))
```

The median reference-isoform DMSO 4-CT mean expression captures
**`r ref_share_med_pct`%** of the parent gene's total non-NMD expression;
**`r ref_share_ge50_pct`%** of references are above the 50% threshold.

## §1d — Figure 3 Panels A/B/C

Panels A (pair-set concept), B (sequence similarity), and C (event prevalence)
characterize the pop_BC pair structure. Panel-level methodology is in
`figures/multipanel/figure3_isopair_and_ptc/`.

```{r sec1-fig3abc, echo=FALSE, fig.cap=c("Figure 3 Panel A — Isoform pair construction (schematic)", "Figure 3 Panel B — Sequence similarity to reference", "Figure 3 Panel C — Splice event prevalence"), out.width="100%"}
knitr::include_graphics(c(
  file.path(NMD_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelA_pair_concept.png"),
  file.path(NMD_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelB_sequence_similarity.png"),
  file.path(NMD_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelC_event_prevalence.png")
))
```

# §2 — PTC analysis at the GENCODE-only scope (n=190)

The Section A scope is the most stringent "complete GENCODE-only" subset:
every isoform in the pair (Reference, NMD comparator, Control comparator) is
GENCODE-annotated (ENST prefix) AND has a curated coding CDS. This eliminates
TD2-derived CDS calls from the comparison and gives the cleanest possible
PTC-vs-non-PTC contrast.

**Scope construction** (replicated from
[`figures/multipanel/figure4_ptcneg_and_model/data_export.R`](../../../../figures/multipanel/figure4_ptcneg_and_model/data_export.R) §"Section A scope"):

```{r sec2-scope, include=FALSE}
profiles_c2_dt <- as.data.table(profiles_c2)
profiles_c4_dt <- as.data.table(profiles_c4)
cds_dt         <- cds

is_enst  <- function(x) grepl("^ENST", x)
make_key <- function(d) paste(d$gene_id, d$reference_isoform_id, sep = "::")

# Step 1 — pop_BC (gene-matched on shared (gene, ref))
pop_bc_keys <- intersect(unique(make_key(profiles_c2_dt)),
                          unique(make_key(profiles_c4_dt)))
pop_bc_c2   <- profiles_c2_dt[make_key(profiles_c2_dt) %in% pop_bc_keys]
pop_bc_c4   <- profiles_c4_dt[make_key(profiles_c4_dt) %in% pop_bc_keys]

# Step 2 — all-ENST on both reference AND comparator
pop_bc_c2_E <- pop_bc_c2[is_enst(reference_isoform_id) &
                           is_enst(comparator_isoform_id)]
pop_bc_c4_E <- pop_bc_c4[is_enst(reference_isoform_id) &
                           is_enst(comparator_isoform_id)]
joint1      <- intersect(make_key(pop_bc_c2_E), make_key(pop_bc_c4_E))
pop_bc_c2_E <- pop_bc_c2_E[make_key(pop_bc_c2_E) %in% joint1]
pop_bc_c4_E <- pop_bc_c4_E[make_key(pop_bc_c4_E) %in% joint1]

# Step 3 — coding-everywhere (curated CDS on both sides)
coding_ids <- cds_dt[coding_status == "coding", isoform_id]
have_cds   <- function(d) d$reference_isoform_id %in% coding_ids &
                          d$comparator_isoform_id %in% coding_ids
pop_bc_c2_E2 <- pop_bc_c2_E[have_cds(pop_bc_c2_E)]
pop_bc_c4_E2 <- pop_bc_c4_E[have_cds(pop_bc_c4_E)]

# Step 4 — re-intersect on (gene, ref) after the coding filter, so both arms
#          share the same reference set
joint2 <- intersect(make_key(pop_bc_c2_E2), make_key(pop_bc_c4_E2))
gencode_all3_c2 <- pop_bc_c2_E2[make_key(pop_bc_c2_E2) %in% joint2]
gencode_all3_c4 <- pop_bc_c4_E2[make_key(pop_bc_c4_E2) %in% joint2]

sec2_cascade <- data.frame(
  Step = c("pop_BC (gene-matched)",
           "+ all-ENST (ref & comp)",
           "  re-intersect on (gene, ref)",
           "+ coding-CDS everywhere",
           "  re-intersect on (gene, ref)"),
  c2_n = c(nrow(pop_bc_c2),
           length(unique(c(make_key(pop_bc_c2_E), make_key(pop_bc_c4_E)))),
           nrow(pop_bc_c2_E),
           nrow(pop_bc_c2_E2),
           nrow(gencode_all3_c2)),
  c4_n = c(nrow(pop_bc_c4),
           NA,
           nrow(pop_bc_c4_E),
           nrow(pop_bc_c4_E2),
           nrow(gencode_all3_c4))
)
```

Filter cascade from pop_BC down to the n=190 Section A scope:

```{r sec2-scope-table, echo=FALSE}
knitr::kable(sec2_cascade,
             caption = sprintf("Section A scope cascade — pop_BC → all-ENST → coding-everywhere → gencode_all3 (n=%s / %s)",
                               format(nrow(gencode_all3_c2), big.mark=","),
                               format(nrow(gencode_all3_c4), big.mark=",")),
             format.args = list(big.mark = ","),
             col.names = c("Filter step", "c2 rows", "c4 rows"))
```

`gencode_all3_c2` = `r nrow(gencode_all3_c2)` NMD pairs;
`gencode_all3_c4` = `r nrow(gencode_all3_c4)` Control pairs. Both arms share
the same `r length(joint2)` `(gene, reference)` pairs.

**Why the cohort drops from 2,098/2,098 ENST-reference pairs to 190 coding-
everywhere pairs:** the ENST-on-comparator filter is the tightest constraint
— most NMD comparators in `pop_BC` are novel isoforms
(`r ann_table$pct_Novel[ann_table$Role == "NMD comparator"]`% per §1b), so
requiring ENST on the NMD-comparator side eliminates the bulk of the cohort.
Requiring a curated CDS on top reduces further to 190. This stringent scope
defines the Section A panels in Figure 3 (D/E/F) and Figure 4 (A/B), plus
the `CDSand3UTR_GENCODEonly` supplement.

## §2a — PTC determination

At this scope every isoform is GENCODE-annotated with a curated coding CDS, so
PTC status is determined per comparator from its own annotated stop's distance
to the last exon-exon junction (50-nt rule). No ref-AUG projection is needed.
This is the methodology used by Figure 3 Panels D/E/F and is replicated inline
from
[`figures/multipanel/figure3_isopair_and_ptc/data_export.R`](../../../../figures/multipanel/figure3_isopair_and_ptc/data_export.R).

```{r sec2a-ptc, include=FALSE}
coding_cds <- cds[coding_status == "coding"]
struct_idx <- match(coding_cds$isoform_id, structures$isoform_id)

own_stop_tx_lookup <- (function() {
  stop_tx <- rep(NA_integer_, nrow(coding_cds))
  for (i in seq_len(nrow(coding_cds))) {
    si <- struct_idx[i]
    if (is.na(si)) next
    strand <- coding_cds$strand[i]
    if (is.na(strand) || strand == "") next
    stop_g <- if (strand == "+") coding_cds$cds_stop[i] else coding_cds$cds_start[i]
    stop_tx[i] <- Isopair::genomicToTranscript(
      stop_g, structures$exon_starts[[si]], structures$exon_ends[[si]], strand)
  }
  data.table(isoform_id = coding_cds$isoform_id,
             own_stop_tx_pos = stop_tx,
             strand = coding_cds$strand)
})()

junctions_for <- function(iso_id) {
  si <- match(iso_id, structures$isoform_id)
  if (is.na(si)) return(integer(0))
  exon_starts <- structures$exon_starts[[si]]
  exon_ends   <- structures$exon_ends[[si]]
  strand      <- structures$strand[si]
  if (is.na(strand) || strand == "") return(integer(0))
  exon_lens <- exon_ends - exon_starts + 1L
  if (strand == "-") exon_lens <- rev(exon_lens)
  if (length(exon_lens) <= 1L) return(integer(0))
  cumsum(exon_lens[-length(exon_lens)])
}

augment_own_stop <- function(pop) {
  merge(pop[, .(gene_id, reference_isoform_id, comparator_isoform_id)],
        own_stop_tx_lookup,
        by.x = "comparator_isoform_id", by.y = "isoform_id",
        all.x = TRUE)
}
compute_ptc_info <- function(d) {
  d[, last_ejc_tx_pos := vapply(comparator_isoform_id,
                                 function(iso) {
                                   junc <- junctions_for(iso)
                                   if (length(junc) == 0L) NA_integer_ else max(junc)
                                 }, integer(1))]
  d[, distance := last_ejc_tx_pos - own_stop_tx_pos]
  d[, has_ptc := !is.na(distance) & distance > 50L]
  d
}

trace_c2 <- compute_ptc_info(augment_own_stop(gencode_all3_c2))
trace_c4 <- compute_ptc_info(augment_own_stop(gencode_all3_c4))

n_ptc_nmd     <- sum(trace_c2$has_ptc, na.rm = TRUE)
n_ptc_ctrl    <- sum(trace_c4$has_ptc, na.rm = TRUE)
n_avail_nmd   <- sum(!is.na(trace_c2$distance))
n_avail_ctrl  <- sum(!is.na(trace_c4$distance))
ptc_rate_nmd  <- 100 * n_ptc_nmd  / n_avail_nmd
ptc_rate_ctrl <- 100 * n_ptc_ctrl / n_avail_ctrl
fold_enrich   <- ptc_rate_nmd / ptc_rate_ctrl

ptc_fisher <- fisher.test(matrix(c(
  n_ptc_nmd,            n_avail_nmd  - n_ptc_nmd,
  n_ptc_ctrl,           n_avail_ctrl - n_ptc_ctrl
), nrow = 2, byrow = TRUE))

pop_ptc_c2 <- merge(
  gencode_all3_c2,
  trace_c2[has_ptc == TRUE,
           .(comparator_isoform_id, own_stop_tx_pos, last_ejc_tx_pos, distance)],
  by = "comparator_isoform_id"
)
```

**PTC+ rates** (own GENCODE stop + 50-nt rule, n=`r nrow(gencode_all3_c2)`
NMD pairs / `r nrow(gencode_all3_c4)` Control pairs):

- NMD: `r n_ptc_nmd` / `r n_avail_nmd` = **`r sprintf("%.1f", ptc_rate_nmd)`%**
- Control: `r n_ptc_ctrl` / `r n_avail_ctrl` = **`r sprintf("%.1f", ptc_rate_ctrl)`%**
- Fold enrichment: **`r sprintf("%.1f×", fold_enrich)`**
- Fisher: OR = `r sprintf("%.1f", ptc_fisher$estimate)`, p = `r format_p(ptc_fisher$p.value)`

```{r sec2a-fig3-panelD, echo=FALSE, fig.cap="Figure 3 Panel D — Stop-to-last-EJC distance at the n=190 scope", out.width="80%"}
knitr::include_graphics(
  file.path(NMD_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelD_stop_codon_distance.png")
)
```

## §2b — PTC-causing event attribution

For each of the `r n_ptc_nmd` PTC+ NMD pairs, we attribute the splice event
that introduced the PTC. Two attribution functions are applied:

- `attribute_ptc_events()` — matches a `detailed_events` row to the stop
  position via `fw_events` (forward-walked junctions); handles the diff-stop
  case where the comparator's PTC is at a different genomic position from
  the reference's stop.
- `attribute_3utr_splice()` — for same-stop pairs (comparator's own GENCODE
  stop equals the reference's stop), looks for splice events DOWNSTREAM of
  the stop in the 3'UTR.

For the small set of pairs where both functions return a `"direct"` row, the
same-stop call takes precedence (a 3'UTR splice is a more specific
attribution than a diff-stop event when the stops coincide).

```{r sec2b-attribution, include=FALSE}
fc_c2 <- readRDS(file.path(DM, "analysis_cache/fc_c2_allsamples.rds"))
fw_c2 <- readRDS(file.path(DM, "analysis_cache/fw_c2_allsamples.rds"))

own_stop_genomic_lookup <- setNames(
  ifelse(coding_cds$strand == "+", coding_cds$cds_stop, coding_cds$cds_start),
  coding_cds$isoform_id)
own_atg_genomic_lookup <- setNames(
  ifelse(coding_cds$strand == "+", coding_cds$cds_start, coding_cds$cds_stop),
  coding_cds$isoform_id)
strand_lookup <- setNames(coding_cds$strand, coding_cds$isoform_id)

ep_ids <- pop_ptc_c2$comparator_isoform_id
fc_ps <- as.data.table(fc_c2$pair_summary)
fc_ep <- fc_ps[comparator_isoform_id %in% ep_ids]
frameshift_cats <- c("same_start_frameshift",
                     "diff_start_diff_frame_with_frameshift")
fc_ep[, is_frameshift_cat := frame_category %in% frameshift_cats]

ptc_genomic_pos <- own_stop_genomic_lookup[fc_ep$comparator_isoform_id]
atg_genomic_pos <- own_atg_genomic_lookup[fc_ep$comparator_isoform_id]
strand_vec      <- strand_lookup[fc_ep$comparator_isoform_id]
names(ptc_genomic_pos) <- fc_ep$comparator_isoform_id
names(atg_genomic_pos) <- fc_ep$comparator_isoform_id
names(strand_vec)      <- fc_ep$comparator_isoform_id
is_fs_lookup <- setNames(fc_ep$is_frameshift_cat, fc_ep$comparator_isoform_id)

attr_b <- as.data.table(attribute_ptc_events(
  pairs           = fc_ep[, .(comparator_isoform_id, reference_isoform_id)],
  fw_events       = fw_c2$events,
  profiles        = as.data.frame(pop_ptc_c2),
  ptc_genomic_pos = ptc_genomic_pos,
  atg_genomic_pos = atg_genomic_pos,
  strand_vec      = strand_vec,
  is_frameshift_vec = is_fs_lookup
))
n_direct_b     <- sum(attr_b$attribution == "direct")
n_unresolved_b <- sum(attr_b$attribution == "unresolved")

# Same-stop 3'UTR splice handling
fc_ep[, strand_s     := strand_lookup[comparator_isoform_id]]
fc_ep[, comp_stop_s  := own_stop_genomic_lookup[comparator_isoform_id]]
fc_ep[, ref_stop_s   := own_stop_genomic_lookup[reference_isoform_id]]
fc_ep[, same_stop_s  := comp_stop_s == ref_stop_s]
same_stop_b <- fc_ep[same_stop_s == TRUE]

ss_events_b <- data.table()
if (nrow(same_stop_b) > 0) {
  ss_stop_vec   <- setNames(same_stop_b$comp_stop_s, same_stop_b$comparator_isoform_id)
  ss_strand_vec <- setNames(same_stop_b$strand_s,    same_stop_b$comparator_isoform_id)
  ss_events_b <- as.data.table(attribute_3utr_splice(
    pairs            = same_stop_b[, .(comparator_isoform_id, reference_isoform_id)],
    profiles         = as.data.frame(pop_ptc_c2),
    stop_genomic_pos = ss_stop_vec,
    strand_vec       = ss_strand_vec
  ))[attribution == "direct"]
}
n_same_stop_direct <- nrow(ss_events_b)

# Combine: same-stop precedence over diff-stop
same_stop_ids <- if (nrow(ss_events_b) > 0) ss_events_b$comparator_isoform_id else character(0)
diffstop_b    <- attr_b[attribution == "direct" &
                        !comparator_isoform_id %in% same_stop_ids]
all_attr_new  <- rbind(diffstop_b, ss_events_b, fill = TRUE)
all_attr_new  <- all_attr_new[attribution == "direct"]
all_attr_new$event_short <- shorten_event_labels(all_attr_new$ptc_causing_event,
                                                  collapse_rare = 3)
n_unique_direct <- nrow(all_attr_new)

# Mechanism breakdown (38 / 23 / 8 target)
all_attr_new[, mechanism := factor(mechanism,
                                   levels = c("Frameshift", "In-frame stop", "3'UTR splice"))]
mech_tab <- as.data.frame(table(all_attr_new$mechanism, useNA = "ifany"))
names(mech_tab) <- c("Mechanism", "n")
```

**Attribution accounting:**
`r n_direct_b` direct (from `attribute_ptc_events`) +
`r n_same_stop_direct` same-stop 3'UTR splice =
`r n_direct_b + n_same_stop_direct` total attribution rows →
`r n_unique_direct` unique direct (same-stop takes precedence over diff-stop;
`r n_direct_b + n_same_stop_direct - n_unique_direct` rows were attributed by
both functions, with the same-stop call kept) →
`r n_ptc_nmd - n_unique_direct` unresolved out of `r n_ptc_nmd` PTC+ pairs
(`r sprintf("%.1f%%", 100 * n_unique_direct / n_ptc_nmd)` attribution coverage).

The pre-combine `attribute_ptc_events` pass left `r n_unresolved_b`
unresolved; the same-stop pass rescued
`r n_unresolved_b - (n_ptc_nmd - n_unique_direct)`
of those, leaving `r n_ptc_nmd - n_unique_direct` finally unresolved.

**Mechanism breakdown** of the `r n_unique_direct` attributed PTC events:

```{r sec2b-mech-table, echo=FALSE}
knitr::kable(mech_tab,
             caption = sprintf("Mechanism breakdown for the %d attributed PTC events (3 of 72 PTC+ unresolved; Section A n=190 scope)", n_unique_direct),
             col.names = c("Mechanism", "n"))
```

```{r sec2b-fig3-panelE-F, echo=FALSE, fig.cap=c("Figure 3 Panel E — PTC-causing event attribution (per-event Fisher table)", "Figure 3 Panel F — Mechanism breakdown × event type"), out.width="80%"}
knitr::include_graphics(c(
  file.path(NMD_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelE_ptc_event_attribution.png"),
  file.path(NMD_ROOT, "figures/multipanel/figure3_isopair_and_ptc/figure3_panelF_mechanism_breakdown.png")
))
```

## §2c — 5'UTR length + longest 5'UTR ORF at the n=190 scope

At the n=190 all-3-ENST scope, the 5'UTR is the comparator's annotated 5'UTR
(every isoform is GENCODE-annotated so the TD2-anchored `utr5_features_all`
cache value equals the GENCODE 5'UTR length at this scope; the Rmd reads it
from cache rather than recomputing). Three groups (using §2a `has_ptc`):

- **NMD+/PTC+** (`r n_ptc_nmd` pairs) — `has_ptc == TRUE`
- **NMD+/PTC−** (`r n_avail_nmd - n_ptc_nmd` pairs) — `has_ptc == FALSE`
- **Control** (`r n_avail_ctrl` pairs) — `gencode_all3_c4` comparators

The Control arm is a single group (not PTC-stratified) because PTC status is
a property of the NMD-comparator. This matches the figure-side panel
construction.

Interpretive caution: at the n=190 scope, NMD+/PTC+ comparators have markedly
short 5'UTRs (median 84.5 nt vs Control 138 nt) — this is most likely an
alternative-TSS effect (PTC+ ENSTs at this scope tend to be 5'-truncated
alternative transcripts) rather than 5'UTR cis-element divergence per se.
The NMD+/PTC− 5'UTR enrichment is therefore the more biologically meaningful
contrast for uORF-mediated NMD.

```{r sec2c-helpers, include=FALSE}
hodges_lehmann <- function(x, y) {
  x <- x[!is.na(x)]; y <- y[!is.na(y)]
  if (length(x) < 2L || length(y) < 2L) return(c(NA_real_, NA_real_, NA_real_, NA_real_))
  wt <- tryCatch(wilcox.test(x, y, conf.int = TRUE, exact = FALSE),
                  error = function(e) NULL)
  if (is.null(wt)) return(c(NA_real_, NA_real_, NA_real_, NA_real_))
  c(unname(wt$estimate), unname(wt$conf.int[1]), unname(wt$conf.int[2]),
    unname(wt$p.value))
}
cliffs_delta <- function(x, y) {
  x <- x[!is.na(x)]; y <- y[!is.na(y)]
  if (length(x) == 0L || length(y) == 0L) return(NA_real_)
  (sum(outer(x, y, ">")) - sum(outer(x, y, "<"))) / (length(x) * length(y))
}
pairwise_stats <- function(groups_list, lookup) {
  combos <- combn(names(groups_list), 2, simplify = FALSE)
  do.call(rbind, lapply(combos, function(p) {
    x <- lookup[groups_list[[p[1]]]]; y <- lookup[groups_list[[p[2]]]]
    x <- x[!is.na(x)]; y <- y[!is.na(y)]
    hl <- hodges_lehmann(x, y)
    d  <- cliffs_delta(x, y)
    data.frame(group_x = p[1], group_y = p[2],
               n_x = length(x), n_y = length(y),
               median_x = round(median(x), 1), median_y = round(median(y), 1),
               hl_shift = round(hl[1], 1),
               hl_ci = sprintf("[%.1f, %.1f]", hl[2], hl[3]),
               cliffs_delta = round(d, 3),
               wilcox_p = hl[4],
               stringsAsFactors = FALSE)
  }))
}
group_descriptives <- function(groups_list, lookup, order_names) {
  out <- do.call(rbind, lapply(order_names, function(g) {
    v <- lookup[groups_list[[g]]]
    v <- v[!is.na(v)]
    data.frame(group = g, n = length(v),
               median = round(median(v), 1),
               q25 = round(unname(quantile(v, 0.25)), 1),
               q75 = round(unname(quantile(v, 0.75)), 1),
               stringsAsFactors = FALSE)
  }))
  rownames(out) <- NULL
  out
}
```

```{r sec2c-secA-groups, include=FALSE}
ptc_pos_ids <- trace_c2[has_ptc == TRUE,  comparator_isoform_id]
ptc_neg_ids <- trace_c2[has_ptc == FALSE, comparator_isoform_id]
sec_A_groups <- list(
  "NMD+/PTC+" = ptc_pos_ids,
  "NMD+/PTC-" = ptc_neg_ids,
  "Control"   = gencode_all3_c4$comparator_isoform_id
)

u5_all_dt <- as.data.table(utr5_all$isoform_features)
u5_len_lookup <- setNames(as.numeric(u5_all_dt$utr5_length),    u5_all_dt$isoform_id)
u5_orf_lookup <- setNames(as.numeric(u5_all_dt$longest_orf_nt), u5_all_dt$isoform_id)

secA_utr5_desc <- group_descriptives(sec_A_groups, u5_len_lookup,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secA_utr5_pw   <- pairwise_stats(sec_A_groups, u5_len_lookup)
secA_orf_desc  <- group_descriptives(sec_A_groups, u5_orf_lookup,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secA_orf_pw    <- pairwise_stats(sec_A_groups, u5_orf_lookup)

secA_utr5_pw$wilcox_p_fmt <- vapply(secA_utr5_pw$wilcox_p, format_p, character(1))
secA_orf_pw$wilcox_p_fmt  <- vapply(secA_orf_pw$wilcox_p,  format_p, character(1))
```

### Panel A — 5'UTR length

```{r sec2c-panelA-tables, echo=FALSE}
knitr::kable(secA_utr5_desc,
             caption = "5'UTR length (nt) — Section A descriptives",
             col.names = c("Group", "n", "Median", "Q25", "Q75"))
knitr::kable(secA_utr5_pw[, c("group_x","group_y","n_x","n_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "5'UTR length — pairwise stats (Hodges–Lehmann shift, Cliff's δ, Wilcoxon p)",
             col.names = c("X","Y","n_X","n_Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

### Panel B — Longest 5'UTR ORF

```{r sec2c-panelB-tables, echo=FALSE}
knitr::kable(secA_orf_desc,
             caption = "Longest 5'UTR ORF (nt) — Section A descriptives",
             col.names = c("Group", "n", "Median", "Q25", "Q75"))
knitr::kable(secA_orf_pw[, c("group_x","group_y","n_x","n_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "Longest 5'UTR ORF — pairwise stats",
             col.names = c("X","Y","n_X","n_Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

```{r sec2c-fig4AB, echo=FALSE, fig.cap=c("Figure 4 Panel A — 5'UTR length at n=190 scope", "Figure 4 Panel B — Longest 5'UTR ORF at n=190 scope"), out.width="90%"}
knitr::include_graphics(c(
  file.path(NMD_ROOT, "figures/multipanel/figure4_ptcneg_and_model/figure4_panelA_5utr_length_all3enst.png"),
  file.path(NMD_ROOT, "figures/multipanel/figure4_ptcneg_and_model/figure4_panelB_longest_5utr_orf_all3enst.png")
))
```

## §2d — 3'UTR / CDS sanity at n=190

Three additional structural features for the same `sec_A_groups`: CDS length,
translation-based 3'UTR (`tx_len − own_stop_tx_pos`), and bias-corrected
3'UTR via the `utr3_via_non_ptc_stop_nt` field of `ref_atg` (walks past PTCs
to find the next non-PTC stop). The non-PTC measure is the more meaningful
3'UTR-length comparator because for PTC+ pairs the translation-based 3'UTR
inflates with PTC distance and is no longer comparable to the Control 3'UTR.

```{r sec2d-features, include=FALSE}
own_atg_tx_lookup <- (function() {
  atg_tx <- rep(NA_integer_, nrow(coding_cds))
  for (i in seq_len(nrow(coding_cds))) {
    si <- struct_idx[i]; if (is.na(si)) next
    strand <- coding_cds$strand[i]
    if (is.na(strand) || strand == "") next
    atg_g <- if (strand == "+") coding_cds$cds_start[i] else coding_cds$cds_stop[i]
    atg_tx[i] <- Isopair::genomicToTranscript(
      atg_g, structures$exon_starts[[si]], structures$exon_ends[[si]], strand)
  }
  setNames(atg_tx, coding_cds$isoform_id)
})()

tx_len_lookup <- setNames(
  vapply(seq_len(nrow(structures)), function(i)
    sum(structures$exon_ends[[i]] - structures$exon_starts[[i]] + 1L),
    integer(1)),
  structures$isoform_id
)

# Build per-comparator features for sec_A_groups
ra_c2_lookup <- setNames(
  as.data.table(ref_atg$c2)$utr3_via_non_ptc_stop_nt,
  as.data.table(ref_atg$c2)$comparator_isoform_id
)
ra_c4_lookup <- setNames(
  as.data.table(ref_atg$c4)$utr3_via_non_ptc_stop_nt,
  as.data.table(ref_atg$c4)$comparator_isoform_id
)

compute_secA_feat <- function(comp_ids, ra_lookup) {
  own_stop <- own_stop_tx_lookup$own_stop_tx_pos[match(comp_ids, own_stop_tx_lookup$isoform_id)]
  own_atg  <- own_atg_tx_lookup[comp_ids]
  tx_len   <- tx_len_lookup[comp_ids]
  data.table(
    comparator_isoform_id = comp_ids,
    cds_nt   = pmax(0L, own_stop - own_atg + 1L),
    utr3_nt  = pmax(0L, tx_len - own_stop),
    utr3_via_non_ptc_stop_nt = ra_lookup[comp_ids]
  )
}

feat_ptc_pos <- compute_secA_feat(ptc_pos_ids, ra_c2_lookup)
feat_ptc_neg <- compute_secA_feat(ptc_neg_ids, ra_c2_lookup)
feat_ctrl    <- compute_secA_feat(gencode_all3_c4$comparator_isoform_id, ra_c4_lookup)
# At the n=190 all-3-ENST + coding-CDS scope every comparator's own GENCODE
# stop is non-PTC by classification (PTC+ pairs were segregated via §2a's
# has_ptc flag), so for PTC- and Control the bias-corrected 3'UTR length IS
# the translation-based measure. Overwrite unconditionally (matches the
# figure-side CDSand3UTR_GENCODEonly/data_export.R convention).
feat_ptc_neg[, utr3_via_non_ptc_stop_nt := utr3_nt]
feat_ctrl[,    utr3_via_non_ptc_stop_nt := utr3_nt]

cds_lookup <- setNames(
  c(feat_ptc_pos$cds_nt, feat_ptc_neg$cds_nt, feat_ctrl$cds_nt),
  c(feat_ptc_pos$comparator_isoform_id,
    feat_ptc_neg$comparator_isoform_id,
    feat_ctrl$comparator_isoform_id)
)
utr3_lookup <- setNames(
  c(feat_ptc_pos$utr3_nt, feat_ptc_neg$utr3_nt, feat_ctrl$utr3_nt),
  c(feat_ptc_pos$comparator_isoform_id,
    feat_ptc_neg$comparator_isoform_id,
    feat_ctrl$comparator_isoform_id)
)
utr3_npstop_lookup <- setNames(
  c(feat_ptc_pos$utr3_via_non_ptc_stop_nt,
    feat_ptc_neg$utr3_via_non_ptc_stop_nt,
    feat_ctrl$utr3_via_non_ptc_stop_nt),
  c(feat_ptc_pos$comparator_isoform_id,
    feat_ptc_neg$comparator_isoform_id,
    feat_ctrl$comparator_isoform_id)
)

secA_cds_desc  <- group_descriptives(sec_A_groups, cds_lookup,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secA_cds_pw    <- pairwise_stats(sec_A_groups, cds_lookup)
secA_utr3_desc <- group_descriptives(sec_A_groups, utr3_lookup,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secA_utr3_pw   <- pairwise_stats(sec_A_groups, utr3_lookup)
secA_utr3np_desc <- group_descriptives(sec_A_groups, utr3_npstop_lookup,
                                        c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secA_utr3np_pw   <- pairwise_stats(sec_A_groups, utr3_npstop_lookup)

secA_cds_pw$wilcox_p_fmt    <- vapply(secA_cds_pw$wilcox_p,    format_p, character(1))
secA_utr3_pw$wilcox_p_fmt   <- vapply(secA_utr3_pw$wilcox_p,   format_p, character(1))
secA_utr3np_pw$wilcox_p_fmt <- vapply(secA_utr3np_pw$wilcox_p, format_p, character(1))
```

### CDS length

```{r sec2d-cds, echo=FALSE}
knitr::kable(secA_cds_desc, caption = "CDS length (nt)",
             col.names = c("Group","n","Median","Q25","Q75"))
knitr::kable(secA_cds_pw[, c("group_x","group_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "CDS length — pairwise stats",
             col.names = c("X","Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

### 3'UTR length (translation-based)

```{r sec2d-utr3, echo=FALSE}
knitr::kable(secA_utr3_desc, caption = "3'UTR length, translation-based (own stop → tx end)",
             col.names = c("Group","n","Median","Q25","Q75"))
knitr::kable(secA_utr3_pw[, c("group_x","group_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "3'UTR translation-based — pairwise stats",
             col.names = c("X","Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

### 3'UTR length (bias-corrected, walks past PTCs)

```{r sec2d-utr3np, echo=FALSE}
knitr::kable(secA_utr3np_desc,
             caption = "3'UTR length, bias-corrected (non-PTC-stop based)",
             col.names = c("Group","n","Median","Q25","Q75"))
knitr::kable(secA_utr3np_pw[, c("group_x","group_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "3'UTR bias-corrected — pairwise stats",
             col.names = c("X","Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

```{r sec2d-supp, echo=FALSE, fig.cap="CDSand3UTR_GENCODEonly supplement — 2×3 layout. Row 1 (Section A, n=190): A CDS length, B 3'UTR translation-based, C 3'UTR non-PTC-stop (bias-corrected). Row 2 (Section C, n=1,166, ref-AUG-projected): D CDS length, E 3'UTR translation-based, F 3'UTR non-PTC-stop. The bias-correction pattern (PTC+ inflation under translation-based collapses under non-PTC-stop measure) replicates across both scopes.", out.width="100%"}
knitr::include_graphics(
  file.path(NMD_ROOT, "figures/SupplementalFigures/CDSand3UTR_GENCODEonly/figure_s_cds_and_3utr.png")
)
```

# §3 — Larger-scope replication at ref-AUG-traceable (n=1,166)

Section C extends Section A by relaxing the all-3-ENST + coding-CDS
constraints and instead routing through the ref-AUG-projected ORF
classification. This roughly 6× larger cohort lets us check that Section A's
PTC findings replicate at the larger scope without relying on TD2's CDS calls
on novel comparators.

**Scope construction** (replicated from
[`figures/multipanel/figure4_ptcneg_and_model/data_export.R`](../../../../figures/multipanel/figure4_ptcneg_and_model/data_export.R) §"Section C scope"):

```{r sec3-scope, include=FALSE}
ref_atg_c2 <- as.data.table(ref_atg$c2)
ref_atg_c4 <- as.data.table(ref_atg$c4)

# Step 1 — ENST-reference subset of pop_BC (no constraint on comparator)
sec_C_c2_base <- pop_bc_c2[is_enst(reference_isoform_id)]
sec_C_c4_base <- pop_bc_c4[is_enst(reference_isoform_id)]

# Step 2 — restrict ref_atg tables to the ENST-reference pop_BC comparators
c2_full <- ref_atg_c2[comparator_isoform_id %in%
                        sec_C_c2_base$comparator_isoform_id]
c4_full <- ref_atg_c4[comparator_isoform_id %in%
                        sec_C_c4_base$comparator_isoform_id]

# Step 3 — ref-AUG-traceable category filter (per side)
TR    <- REF_AUG_PRESENT_CATEGORIES   # from mechanism-helpers chunk
c2_tr <- c2_full[category %in% TR]
c4_tr <- c4_full[category %in% TR]

# Step 4 — re-intersect on (gene, reference) AFTER the category filter so
#          NMD and Control panels share the same 1:1 gene-matched universe
shared_keys <- merge(
  unique(c2_tr[, .(gene_id, reference_isoform_id)]),
  unique(c4_tr[, .(gene_id, reference_isoform_id)]),
  by = c("gene_id", "reference_isoform_id")
)
refaug_matched_c2 <- merge(c2_tr, shared_keys,
                            by = c("gene_id", "reference_isoform_id"))
refaug_matched_c4 <- merge(c4_tr, shared_keys,
                            by = c("gene_id", "reference_isoform_id"))

# Classification: 5-group → 4-group → 3-group merged
refaug_matched_c2[, mech5 := mechanism_class(category, comp_orf_length, ref_orf_length)]
refaug_matched_c2[, mech4 := mechanism_class_4(mech5)]
refaug_matched_c2[, group3 := data.table::fcase(
  mech4 == "NMD+/PTC+",                                       "NMD+/PTC+",
  mech4 %in% c("NMD+/PTC- ORF match", "NMD+/PTC- ORF diff"),  "NMD+/PTC-",
  default = "Ref AUG absent"
)]

sec3_cascade <- data.frame(
  Step = c("pop_BC (gene-matched)",
           "+ ENST-reference (no comp constraint)",
           "+ ref-AUG-traceable category (per side)",
           "  re-intersect on (gene, ref)"),
  c2_n = c(nrow(pop_bc_c2),
           nrow(sec_C_c2_base),
           nrow(c2_tr),
           nrow(refaug_matched_c2)),
  c4_n = c(nrow(pop_bc_c4),
           nrow(sec_C_c4_base),
           nrow(c4_tr),
           nrow(refaug_matched_c4))
)

sec3_class <- as.data.frame(table(refaug_matched_c2$group3,
                                   useNA = "ifany"))
names(sec3_class) <- c("group", "n")
```

Filter cascade from pop_BC down to the n=1,166 Section C scope:

```{r sec3-scope-table, echo=FALSE}
knitr::kable(sec3_cascade,
             caption = sprintf("Section C scope cascade — pop_BC → ENST-ref → ref-AUG-traceable → re-intersect (n=%s / %s)",
                               format(nrow(refaug_matched_c2), big.mark=","),
                               format(nrow(refaug_matched_c4), big.mark=",")),
             format.args = list(big.mark = ","),
             col.names = c("Filter step", "c2 rows", "c4 rows"))
```

`refaug_matched_c2` = `r format(nrow(refaug_matched_c2), big.mark=",")` NMD
pairs; `refaug_matched_c4` = `r format(nrow(refaug_matched_c4), big.mark=",")`
Control pairs. Both arms share the same
`r format(nrow(shared_keys), big.mark=",")` `(gene, reference)` pairs.

**Mechanism classification** of the NMD arm (`mechanism_class_4()` on the
ref-AUG-projected ORF category + ORF-length comparison):

```{r sec3-class-table, echo=FALSE}
knitr::kable(sec3_class,
             caption = "Section C NMD-arm 3-group classification",
             col.names = c("Group", "n"),
             format.args = list(big.mark = ","))
```

PTC+ rate at this scope: `r format(sec3_class$n[sec3_class$group == "NMD+/PTC+"], big.mark=",")` /
`r format(nrow(refaug_matched_c2), big.mark=",")` =
`r round(100 * sec3_class$n[sec3_class$group == "NMD+/PTC+"] / nrow(refaug_matched_c2), 1)`%.
This is higher than Section A's PTC+ rate because the classifier definitions
differ — see §5 for reconciliation.

## §3a — 5'UTR length + longest 5'UTR ORF at Section C

At the n=1,166 scope, most NMD comparators are novel (TD2-anchored CDS not
reliable for 5'UTR measurement) — so we use `utr5_features_refaug`, which is
the 5'UTR delimited by the ref-AUG-projected start. Three groups (from §3
classification): NMD+/PTC+, NMD+/PTC− (ORF match + ORF diff merged), Control.

```{r sec3a-secC-groups, include=FALSE}
sec_C_groups <- list(
  "NMD+/PTC+" = refaug_matched_c2[group3 == "NMD+/PTC+", comparator_isoform_id],
  "NMD+/PTC-" = refaug_matched_c2[group3 == "NMD+/PTC-", comparator_isoform_id],
  "Control"   = refaug_matched_c4$comparator_isoform_id
)

u5_refaug_dt <- as.data.table(utr5_refaug$isoform_features)
u5_len_ra <- setNames(as.numeric(u5_refaug_dt$utr5_length),    u5_refaug_dt$isoform_id)
u5_orf_ra <- setNames(as.numeric(u5_refaug_dt$longest_orf_nt), u5_refaug_dt$isoform_id)

secC_utr5_desc <- group_descriptives(sec_C_groups, u5_len_ra,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secC_utr5_pw   <- pairwise_stats(sec_C_groups, u5_len_ra)
secC_orf_desc  <- group_descriptives(sec_C_groups, u5_orf_ra,
                                      c("NMD+/PTC+", "NMD+/PTC-", "Control"))
secC_orf_pw    <- pairwise_stats(sec_C_groups, u5_orf_ra)
secC_utr5_pw$wilcox_p_fmt <- vapply(secC_utr5_pw$wilcox_p, format_p, character(1))
secC_orf_pw$wilcox_p_fmt  <- vapply(secC_orf_pw$wilcox_p,  format_p, character(1))
```

### Panel C — 5'UTR length (ref-AUG-anchored)

```{r sec3a-panelC-tables, echo=FALSE}
knitr::kable(secC_utr5_desc, caption = "5'UTR length (nt) — Section C descriptives",
             col.names = c("Group","n","Median","Q25","Q75"))
knitr::kable(secC_utr5_pw[, c("group_x","group_y","n_x","n_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "5'UTR length — pairwise stats",
             col.names = c("X","Y","n_X","n_Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

### Panel D — Longest 5'UTR ORF (ref-AUG-anchored)

```{r sec3a-panelD-tables, echo=FALSE}
knitr::kable(secC_orf_desc, caption = "Longest 5'UTR ORF (nt) — Section C descriptives",
             col.names = c("Group","n","Median","Q25","Q75"))
knitr::kable(secC_orf_pw[, c("group_x","group_y","n_x","n_y","median_x","median_y","hl_shift","hl_ci","cliffs_delta","wilcox_p_fmt")],
             caption = "Longest 5'UTR ORF — pairwise stats",
             col.names = c("X","Y","n_X","n_Y","Med X","Med Y","HL shift","HL 95% CI","Cliff's δ","Wilcoxon p"))
```

```{r sec3a-fig4CD, echo=FALSE, fig.cap=c("Figure 4 Panel C — 5'UTR length at n=1,166 scope", "Figure 4 Panel D — Longest 5'UTR ORF at n=1,166 scope"), out.width="90%"}
knitr::include_graphics(c(
  file.path(NMD_ROOT, "figures/multipanel/figure4_ptcneg_and_model/figure4_panelC_5utr_length_refaug.png"),
  file.path(NMD_ROOT, "figures/multipanel/figure4_ptcneg_and_model/figure4_panelD_longest_5utr_orf_refaug.png")
))
```

## §3b — PTC rate at Section C

Already computed inline in the §3 scope construction:
`r format(sec3_class$n[sec3_class$group == "NMD+/PTC+"], big.mark=",")` PTC+ /
`r format(nrow(refaug_matched_c2), big.mark=",")` =
**`r round(100 * sec3_class$n[sec3_class$group == "NMD+/PTC+"] / nrow(refaug_matched_c2), 1)`%** at this scope (vs 37.9% at the n=190
all-3-ENST scope). The classifier definitions differ —
see §5 for reconciliation.

# §4 — TD2 bias evidence (TD2BiasEvidence supplement)

Scope (broad): `refaug_matched_c2 ∩ TD2 CDS available ∩ ref-AUG ORF computable
∩ ref AUG exonic` → n=1,166 broad.
Scope (occult-PTC): broad `∩ effectively_ptc ∩ original_ptc == FALSE` → n=492.

**Scope construction.** Broad (n=1,166) = `refaug_matched_c2` ∩ TD2 CDS
available ∩ ref-AUG ORF computable ∩ ref AUG exonic. Occult-PTC (n=492) =
broad ∩ `category == "effectively_ptc"` ∩ `original_ptc == FALSE` — pairs
where ref-AUG tracing revealed a PTC that TD2 missed. This is the population
in which TD2's PTC-avoidance bias is most pronounced.

```{r sec4-scope, include=FALSE}
c2_full_dt <- as.data.table(refaug_matched_c2)
comp_cds   <- cds[match(c2_full_dt$comparator_isoform_id, isoform_id)]
c2_full_dt[, strand_use      := comp_cds$strand]
c2_full_dt[, td2_atg_genomic := ifelse(comp_cds$strand == "+",
                                       comp_cds$cds_start, comp_cds$cds_stop)]
c2_full_dt[, td2_orf_nt      := comp_cds$orf_length * 3L]
c2_full_dt[, ref_atg_orf_nt  := comp_orf_length_from_ref_atg]

broad <- c2_full_dt[!is.na(td2_orf_nt) & !is.na(ref_atg_orf_nt) &
                     ref_atg_exonic_in_comp == TRUE]
```

`broad` = `r format(nrow(broad), big.mark=",")` pairs (matches the
n=1,166 TD2BiasEvidence supplement scope).

```{r sec4-compute, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(FASTA_PATH)}
genomic_to_tx <- function(gp, es, ee, s) {
  if (s == "-") { es <- rev(es); ee <- rev(ee) }
  cum <- 0L
  for (j in seq_along(es)) {
    if (s == "+" && gp >= es[j] && gp <= ee[j]) return(cum + (gp - es[j]) + 1L)
    if (s == "-" && gp >= es[j] && gp <= ee[j]) return(cum + (ee[j] - gp) + 1L)
    cum <- cum + (ee[j] - es[j] + 1L)
  }
  NA_integer_
}

bs <- structures[match(broad$comparator_isoform_id, isoform_id)]
ref_tx <- integer(nrow(broad)); td2_tx <- integer(nrow(broad))
for (i in seq_len(nrow(broad))) {
  es <- bs$exon_starts[[i]]; ee <- bs$exon_ends[[i]]
  s  <- broad$strand_use[i]
  if (length(es) == 0L || is.na(broad$ref_atg_genomic[i]) ||
      is.na(broad$td2_atg_genomic[i])) {
    ref_tx[i] <- NA_integer_; td2_tx[i] <- NA_integer_; next
  }
  ref_tx[i] <- genomic_to_tx(broad$ref_atg_genomic[i], es, ee, s)
  td2_tx[i] <- genomic_to_tx(broad$td2_atg_genomic[i], es, ee, s)
}
broad[, ref_atg_tx_pos := ref_tx]
broad[, td2_atg_tx_pos := td2_tx]
broad[, distance_nt   := td2_atg_tx_pos - ref_atg_tx_pos]
broad_clean <- broad[!is.na(ref_atg_tx_pos) & !is.na(td2_atg_tx_pos)]

# Kozak PWM scoring — load sequences for broad_clean comparators from the
# nmd_lungcells corrected FASTA via awk (avoid full Biostrings dependency).
load_sequences <- function(isoform_ids, fasta_path) {
  ids <- unique(isoform_ids)
  id_file <- tempfile(fileext = ".txt")
  writeLines(ids, id_file); on.exit(unlink(id_file), add = TRUE)
  awk_cmd <- sprintf(
    "awk 'BEGIN{while((getline id < \"%s\") > 0) want[id]=1} /^>/{if(seq && found) print hdr \"\\t\" seq; hdr=substr($1,2); found=(hdr in want); seq=\"\"; next} found{seq=seq$0} END{if(seq && found) print hdr \"\\t\" seq}' '%s'",
    id_file, fasta_path)
  raw <- system(awk_cmd, intern = TRUE)
  split_raw <- strsplit(raw, "\t", fixed = TRUE)
  setNames(toupper(vapply(split_raw, `[`, character(1), 2)),
                   vapply(split_raw, `[`, character(1), 1))
}
seqs <- load_sequences(broad_clean$comparator_isoform_id, FASTA_PATH)
pwm_seqs <- character(nrow(broad_clean))
for (k in seq_len(nrow(broad_clean))) {
  cid <- broad_clean$comparator_isoform_id[k]
  if (cid %in% names(seqs)) pwm_seqs[k] <- seqs[cid]
}
v_ref <- !is.na(broad_clean$ref_atg_tx_pos) & pwm_seqs != ""
v_td2 <- !is.na(broad_clean$td2_atg_tx_pos) & pwm_seqs != ""
ref_pwm <- rep(NA_real_, nrow(broad_clean)); td2_pwm <- rep(NA_real_, nrow(broad_clean))
if (sum(v_ref) > 0) {
  r <- Isopair::scoreKozakPWM(pwm_seqs[v_ref], broad_clean$ref_atg_tx_pos[v_ref])
  ref_pwm[v_ref] <- r$score
}
if (sum(v_td2) > 0) {
  r <- Isopair::scoreKozakPWM(pwm_seqs[v_td2], broad_clean$td2_atg_tx_pos[v_td2])
  td2_pwm[v_td2] <- r$score
}
broad_clean[, ref_kozak := ref_pwm]
broad_clean[, td2_kozak := td2_pwm]

# Broad-scope summary stats — ratio convention TD2/ref-AUG (matches the
# TD2BiasEvidence supplement). For PTC+ pairs TD2 picks a downstream ATG
# that extends past the PTC, so TD2 ORF > ref-AUG ORF on average.
broad_orf_ratio <- median(broad_clean$td2_orf_nt) / median(broad_clean$ref_atg_orf_nt)
broad_koz_valid  <- !is.na(broad_clean$ref_kozak) & !is.na(broad_clean$td2_kozak)
broad_n_kozak    <- sum(broad_koz_valid)
broad_n_ref_gt   <- sum(broad_clean$ref_kozak[broad_koz_valid] >
                         broad_clean$td2_kozak[broad_koz_valid])
broad_n_ref_eq   <- sum(broad_clean$ref_kozak[broad_koz_valid] ==
                         broad_clean$td2_kozak[broad_koz_valid])
broad_n_nontied  <- broad_n_kozak - broad_n_ref_eq
broad_koz_w      <- wilcox.test(broad_clean$ref_kozak[broad_koz_valid],
                                 broad_clean$td2_kozak[broad_koz_valid],
                                 paired = TRUE)
broad_pos_eq    <- sum(broad_clean$distance_nt == 0L)
broad_pos_down  <- sum(broad_clean$distance_nt > 0L)
broad_pos_up    <- sum(broad_clean$distance_nt < 0L)
broad_pos_pct_down <- 100 * broad_pos_down / nrow(broad_clean)

# Occult-PTC subset
occ <- broad_clean[category == "effectively_ptc" & original_ptc == FALSE]
occ_orf_ratio <- median(occ$td2_orf_nt) / median(occ$ref_atg_orf_nt)
occ_koz_valid <- !is.na(occ$ref_kozak) & !is.na(occ$td2_kozak)
occ_n_kozak   <- sum(occ_koz_valid)
occ_n_ref_gt  <- sum(occ$ref_kozak[occ_koz_valid] > occ$td2_kozak[occ_koz_valid])
occ_n_ref_eq  <- sum(occ$ref_kozak[occ_koz_valid] == occ$td2_kozak[occ_koz_valid])
occ_n_nontied <- occ_n_kozak - occ_n_ref_eq
occ_koz_w     <- wilcox.test(occ$ref_kozak[occ_koz_valid],
                              occ$td2_kozak[occ_koz_valid], paired = TRUE)
occ_pos_down  <- sum(occ$distance_nt > 0L)
occ_pos_pct_down <- 100 * occ_pos_down / nrow(occ)
```

## §4a — TD2 vs ref-AUG ORF length

| Scope | n | TD2 median ORF (nt) | ref-AUG median ORF (nt) | TD2/ref ratio |
|---|---|---|---|---|
| Broad | `r format(nrow(broad_clean), big.mark=",")` | `r format(median(broad_clean$td2_orf_nt), big.mark=",")` | `r format(median(broad_clean$ref_atg_orf_nt), big.mark=",")` | **`r sprintf("%.2f×", broad_orf_ratio)`** |
| Occult-PTC | `r format(nrow(occ), big.mark=",")` | `r format(median(occ$td2_orf_nt), big.mark=",")` | `r format(median(occ$ref_atg_orf_nt), big.mark=",")` | **`r sprintf("%.2f×", occ_orf_ratio)`** |

## §4b — Paired Kozak PWM at ref AUG vs TD2 ATG

The "ref > TD2 (% of all)" column counts pairs where the ref-AUG Kozak score
exceeds TD2's — pooled over all pairs including ties (broad scope has many
pairs where TD2 == ref AUG, distance 0). The "% of non-tied" column drops
ties so the directional preference is unobscured. The paired-Wilcoxon p-value
is computed on the raw score vector, ignoring ties as `wilcox.test` does.

| Scope | n | ref > TD2 | ties | ref > TD2 (% of all) | ref > TD2 (% of non-tied) | Paired-Wilcoxon p |
|---|---|---|---|---|---|---|
| Broad | `r format(broad_n_kozak, big.mark=",")` | `r format(broad_n_ref_gt, big.mark=",")` | `r format(broad_n_ref_eq, big.mark=",")` | `r sprintf("%.1f%%", 100 * broad_n_ref_gt / broad_n_kozak)` | `r sprintf("%.1f%%", 100 * broad_n_ref_gt / broad_n_nontied)` | `r format_p(broad_koz_w$p.value)` |
| Occult-PTC | `r format(occ_n_kozak, big.mark=",")` | `r format(occ_n_ref_gt, big.mark=",")` | `r format(occ_n_ref_eq, big.mark=",")` | `r sprintf("%.1f%%", 100 * occ_n_ref_gt / occ_n_kozak)` | `r sprintf("%.1f%%", 100 * occ_n_ref_gt / occ_n_nontied)` | `r format_p(occ_koz_w$p.value)` |

## §4c — TD2 ATG position relative to ref AUG

| Scope | n | TD2 == ref | TD2 downstream | TD2 upstream | % downstream |
|---|---|---|---|---|---|
| Broad | `r format(nrow(broad_clean), big.mark=",")` | `r format(broad_pos_eq, big.mark=",")` | `r format(broad_pos_down, big.mark=",")` | `r format(broad_pos_up, big.mark=",")` | **`r sprintf("%.1f%%", broad_pos_pct_down)`** |
| Occult-PTC | `r format(nrow(occ), big.mark=",")` | `r format(sum(occ$distance_nt == 0L), big.mark=",")` | `r format(occ_pos_down, big.mark=",")` | `r format(sum(occ$distance_nt < 0L), big.mark=",")` | **`r sprintf("%.1f%%", occ_pos_pct_down)`** |

In the occult-PTC scope, **TD2 selects an ATG downstream of the ref-AUG in
`r sprintf("%.1f%%", occ_pos_pct_down)` of pairs**, with a Kozak score lower
than the ref-AUG in `r sprintf("%.1f%%", 100 * occ_n_ref_gt / occ_n_kozak)`
of pairs (paired-Wilcoxon `r format_p(occ_koz_w$p.value)`). The TD2 ORF is
`r sprintf("%.2f×", occ_orf_ratio)` LONGER than the ref-AUG-projected ORF
(ratio of medians) — TD2 picks a downstream ATG that extends past the PTC into an
in-frame coding region, producing a longer ORF that lacks the EJC-downstream
stop that defines a PTC. Together these three signals describe TD2's
PTC-avoidance bias: TD2 systematically routes around occult PTCs by picking a
downstream ATG, even at the cost of a lower Kozak score.

```{r sec4-supp, echo=FALSE, fig.cap="TD2BiasEvidence supplement — 2×3 layout: broad scope (row 1, n=1,166) + occult-PTC scope (row 2, n=492)", out.width="100%"}
knitr::include_graphics(
  file.path(NMD_ROOT, "figures/SupplementalFigures/TD2BiasEvidence/figure_s_td2_bias.png")
)
```

# §5 — Cumulative accounting [verification chokepoint]

Recompute from §2 / §3 result objects so the totals in this report can be
audited end-to-end without trusting upstream caches.

```{r sec5-accounting, include=FALSE}
secA_ptc_pos  <- length(ptc_pos_ids)
secA_ptc_neg  <- length(ptc_neg_ids)
secA_total    <- nrow(gencode_all3_c2)
stopifnot(
  "§2a NMD denominator drift"   = n_avail_nmd  == secA_total,
  "§2a Control denominator drift" = n_avail_ctrl == nrow(gencode_all3_c4),
  "§2a PTC+ + PTC- != Section A total" = secA_ptc_pos + secA_ptc_neg == secA_total,
  "§2b mechanism table sum != unique direct attributions" =
    sum(mech_tab$n) == n_unique_direct
)

get_n <- function(df, g) {
  v <- df$n[df$group == g]
  if (length(v) == 0L) 0L else v
}
secC_ptc_pos  <- get_n(sec3_class, "NMD+/PTC+")
secC_ptc_neg  <- get_n(sec3_class, "NMD+/PTC-")
secC_ref_abs  <- get_n(sec3_class, "Ref AUG absent")
secC_total    <- nrow(refaug_matched_c2)
stopifnot(
  "§3 class sum != Section C total" =
    secC_ptc_pos + secC_ptc_neg + secC_ref_abs == secC_total,
  "§3a Control denominator != Section C total" =
    length(sec_C_groups$Control) == secC_total
)

acct <- data.frame(
  Scope = c("Section A (n=190)", "Section C (n=1,166)"),
  PTC_pos = c(secA_ptc_pos, secC_ptc_pos),
  PTC_neg = c(secA_ptc_neg, secC_ptc_neg),
  Ref_AUG_absent = c(NA_integer_, secC_ref_abs),
  Total = c(secA_total, secC_total),
  PTC_pos_rate_pct = c(round(100 * secA_ptc_pos / secA_total, 1),
                       round(100 * secC_ptc_pos / secC_total, 1))
)
```

```{r sec5-acct-table, echo=FALSE}
knitr::kable(acct,
             caption = "Cumulative accounting at both canonical scopes",
             col.names = c("Scope", "PTC+", "PTC−", "Ref AUG absent",
                           "Total", "PTC+ rate (%)"))
```

- **Section A (n=`r secA_total`):** `r secA_ptc_pos` PTC+ and `r secA_ptc_neg` PTC− sum to `r secA_total`. PTC+ rate **`r round(100 * secA_ptc_pos / secA_total, 1)`%**.
- **Section C (n=`r format(secC_total, big.mark=",")`):** `r format(secC_ptc_pos, big.mark=",")` PTC+, `r secC_ptc_neg` PTC−, and `r secC_ref_abs` Ref-AUG-absent sum to `r format(secC_total, big.mark=",")`. PTC+ rate **`r round(100 * secC_ptc_pos / secC_total, 1)`%**.

**Why the two scopes give different PTC+ rates** is a methodological — not
biological — divergence:

| Difference | Section A | Section C |
|---|---|---|
| Cohort filter | all-3-ENST + coding-CDS (most stringent) | ENST-reference only (broader) |
| PTC classifier | own-GENCODE-stop + 50-nt rule | ref-AUG-projected `category == "effectively_ptc"` |
| Comparator CDS source | curated GENCODE | TD2 (heavily debiased via ref-AUG projection) |
| Selection-on-PTC | none — every coding-CDS comparator measured | TR-category filter requires ref-AUG-traceable ORF — pairs where ref-AUG points at "no_downstream_ejc" or "truncated_no_ejc" classify as PTC− regardless of comparator's own stop |

The Section C 90% PTC+ rate is enriched by the TR-category filter, which
removes pairs where ref-AUG tracing finds no PTC and no stop near the EJC.
The Section A 38% PTC+ rate has no such pre-filter: it's the raw PTC rate
across all GENCODE-only NMD pairs. The two rates are not directly
comparable. The point of measuring both is to show that the §2-§4 5'UTR /
ORF / Kozak signals replicate at both scopes.

# §6 — What is intentionally NOT in this Rmd

These analyses remain in `05_final_report_mashr.Rmd` (legacy-bannered) or
`06_orf_analysis_mashr.Rmd` (legacy-bannered) — they are retained for
reproducibility of the prior submission and for sensitivity analyses:

- Reclassification two-pass (TD2 → ref-AUG handoff at the old scope) — legacy
  `05_final_report_mashr.Rmd`.
- Per-cell-type PTC stratification at the old scope — legacy
  `05_final_report_mashr.Rmd`.
- "85% combined" framework — legacy `05_final_report_mashr.Rmd`.
- n=1,904 / n=22,335 GENCODE-only 3'UTR outside-set — legacy
  `05_final_report_mashr.Rmd`.
- ORF / deep-learning model analyses — `06_orf_analysis_mashr.Rmd` (in this
  repo, legacy). The canonical home of the manuscript deep-learning model is
  the **linked repo `peter4244/NMD_orf_model_v5_4ct`**: `03_train.py`,
  `config.yaml`, `model.py`, `METHODS.md`, plus DeepSHAP / KernelSHAP scripts.
  The "ORF identification is the bottleneck" headline result lives there.

Cross-cell-type consistency checks are out of scope for this Rmd; per-CT
mashr CSVs in `shortread_dge/mashr/` and `isocall_dge/mashr/` are the source
of truth for CT-stratified numbers.
