---
title: "NMD Isoform Transitions: Publication Report (mashr Classification)"
subtitle: "Isopair Wrapper — Isocall Pipeline, mashr DIE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
library(Isopair)
library(dplyr)
library(ggplot2)
library(knitr)
library(DT)
source("visualization_functions.R")

# Helper: render interactive DT table (sortable, searchable)
render_dt <- function(df, caption = NULL, pageLength = 15, ...) {
  datatable(df, caption = caption, rownames = FALSE,
            options = list(pageLength = pageLength, scrollX = TRUE),
            ...)
}

# ---------------------------------------------------------------------------
# Publication theme and color palette
# ---------------------------------------------------------------------------
theme_pub <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
theme_set(theme_pub)

pal <- c(NMD = "#ef8a62", Control = "#67a9cf")


# Figure output
fig_dir <- "figures"
dir.create(fig_dir, showWarnings = FALSE)
table_dir <- "tables"
dir.create(table_dir, showWarnings = FALSE)

save_fig <- function(p, name, width = 8, height = 5) {
  ggsave(file.path(fig_dir, paste0(name, ".pdf")), p, width = width, height = height)
  ggsave(file.path(fig_dir, paste0(name, ".png")), p, width = width, height = height, dpi = 300)
  print(p)
}

# ---------------------------------------------------------------------------
# Helper functions (adapted from 03_nmd_analysis_mashr.Rmd)
# ---------------------------------------------------------------------------
ct_to_label <- function(ct) tolower(gsub("_", "", ct))

comp_label <- function(comp) {
  switch(toupper(comp), C2 = "NMD", C4 = "Control", toupper(comp))
}

evt_types <- c("Alt_TSS", "Alt_TES", "A5SS", "A3SS", "SE",
               "Missing_Internal", "IR", "IR_diff",
               "Partial_IR_5", "Partial_IR_3")

compute_event_freq <- function(profiles) {
  n_pairs <- nrow(profiles)
  n_pir5 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_5")))
  n_pir3 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_3")))
  counts <- c(sum(profiles$tss_changed), sum(profiles$tes_changed),
              sum(profiles$n_a5ss > 0), sum(profiles$n_a3ss > 0),
              sum(profiles$n_se > 0), sum(profiles$n_missing_internal > 0),
              sum(profiles$n_ir > 0), sum(profiles$n_ir_diff > 0),
              n_pir5, n_pir3)
  data.frame(event_type = evt_types,
             n_pairs_with_event = counts,
             pct_of_pairs = round(100 * counts / n_pairs, 1))
}

compute_ptc_rates_row <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  coding_ids <- cds$isoform_id[cds$coding_status == "coding"]
  p_c2 <- set_results[[c2_key]]$profiles
  cod_c2 <- p_c2[p_c2$reference_isoform_id %in% coding_ids &
                   p_c2$comparator_isoform_id %in% coding_ids, ]
  p_c4 <- set_results[[c4_key]]$profiles
  cod_c4 <- p_c4[p_c4$reference_isoform_id %in% coding_ids &
                   p_c4$comparator_isoform_id %in% coding_ids, ]
  if (nrow(cod_c2) == 0 || nrow(cod_c4) == 0) return(NULL)
  ptc_c2 <- sum(cod_c2$comparator_isoform_id %in% ptc$isoform_id[ptc$has_ptc == TRUE])
  ptc_c4 <- sum(cod_c4$comparator_isoform_id %in% ptc$isoform_id[ptc$has_ptc == TRUE])
  n_c2 <- nrow(cod_c2); n_c4 <- nrow(cod_c4)
  ft <- tryCatch({
    mat <- matrix(c(ptc_c2, n_c2 - ptc_c2, ptc_c4, n_c4 - ptc_c4), nrow = 2)
    fisher.test(mat)
  }, error = function(e) NULL)
  data.frame(
    cell_type = ct,
    n_coding_pairs_NMD = n_c2,
    n_coding_pairs_Ctrl = n_c4,
    NMD_comp_ptc_n = ptc_c2,
    NMD_comp_ptc_pct = round(100 * ptc_c2 / n_c2, 1),
    Ctrl_comp_ptc_n = ptc_c4,
    Ctrl_comp_ptc_pct = round(100 * ptc_c4 / n_c4, 1),
    fisher_OR = if (!is.null(ft)) round(ft$estimate, 2) else NA_real_,
    p_value = if (!is.null(ft)) ft$p.value else NA_real_
  )
}

compute_frame_category_rows <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  fc_c2 <- set_results[[c2_key]]$frame_comp
  fc_c4 <- set_results[[c4_key]]$frame_comp
  if (is.null(fc_c2) || is.null(fc_c4)) return(NULL)
  ps_c2 <- fc_c2$pair_summary
  ps_c4 <- fc_c4$pair_summary
  cats <- c("same_start_in_frame", "same_start_frameshift",
            "diff_start_same_frame", "diff_start_diff_frame",
            "diff_start_diff_frame_with_frameshift",
            "non_coding", "no_shared_cds")
  # Denominator: coding pairs only (exclude non_coding and no_shared_cds)
  n_coding_c2 <- sum(!ps_c2$frame_category %in% c("non_coding", "no_shared_cds"))
  n_coding_c4 <- sum(!ps_c4$frame_category %in% c("non_coding", "no_shared_cds"))
  do.call(rbind, lapply(cats, function(cat) {
    is_coding_cat <- !cat %in% c("non_coding", "no_shared_cds")
    data.frame(
      cell_type = ct,
      frame_category = cat,
      NMD_n = sum(ps_c2$frame_category == cat),
      NMD_pct = if (is_coding_cat) round(100 * sum(ps_c2$frame_category == cat) / n_coding_c2, 1) else NA_real_,
      Ctrl_n = sum(ps_c4$frame_category == cat),
      Ctrl_pct = if (is_coding_cat) round(100 * sum(ps_c4$frame_category == cat) / n_coding_c4, 1) else NA_real_
    )
  }))
}

compute_divergence_row <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  div_c2 <- set_results[[c2_key]]$divergence
  div_c4 <- set_results[[c4_key]]$divergence
  if (is.null(div_c2) || nrow(div_c2) == 0 || is.null(div_c4) || nrow(div_c4) == 0) return(NULL)
  data.frame(
    cell_type = ct,
    NMD_n = nrow(div_c2),
    NMD_median_pct_shared = round(median(div_c2$pct_shared, na.rm = TRUE), 1),
    Ctrl_n = nrow(div_c4),
    Ctrl_median_pct_shared = round(median(div_c4$pct_shared, na.rm = TRUE), 1)
  )
}
```

# 0. Data Loading and Gene-Matching

## Data Provenance Overview

All analyses in this report begin from **gene-matched isoform pair sets**. The
data pipeline is:

1. **01_prepare_data_mashr.R**: Loads the isocall count matrix
   (`nmd_isocall.count_matrix.txt`, 645,273 isoforms x 38 samples), excludes
   2 outlier samples (DO donor 029T: `Sample25_DO_029T_DMSO` and
   `Sample26_DO_029T_Smg1i`, identified as PCA outlier), applies TMM
   normalization and condition-stratified 5% isoform filtering, then classifies
   isoforms as NMD-sensitive (`nmd_responsive == TRUE` in mashr DIE results) or
   non-NMD (`adj.P.Val > 0.50`). Final dataset: 36 samples across 6 cell types.
   Outputs: `expression_data.rds`, `nmd_classification.rds`,
   `sample_metadata.rds`, `gene_map.rds`.

2. **02_build_profiles_mashr.R**: Generates isoform pairs —
   **C4 (Control)**: top-two non-NMD isoforms by DMSO CPM per gene
   (`generatePairsExpression(..., method = "top_two")`);
   **C2 (NMD)**: same C4 reference paired with top NMD isoform by Smg1i CPM
   (via `identifyDominantIsoforms()` + `inner_join` on `gene_id`).
   Builds splicing choice profiles with `Isopair::buildProfiles()`.
   Outputs: `profiles_c{2,4}_*.rds`, `pairs_c{2,4}_*.rds`.

3. **03_nmd_analysis_mashr.Rmd**: Applies **gene-matching** (restricts C2 and C4
   to genes with shared `(gene_id, reference_isoform_id)` in both sets), then
   runs Isopair analysis functions on the matched profiles. Caches results to
   `analysis_cache/`.

4. **This report (05)**: Loads profiles from step 2, re-applies gene-matching
   (step 3), loads cached analysis results, and verifies consistency. All
   downstream analyses use gene-matched profiles exclusively.

```{r flowchart-data, include=FALSE}
# ── Gather numbers for the flowchart ──
# These are computed fresh from the loaded data objects to ensure accuracy.
# Raw count matrix dimensions are hardcoded (not loaded here to save memory).
n_raw_isoforms <- 645273L
n_raw_samples  <- 38L
n_excluded     <- 2L
```

```{r flowchart, fig.width=10, fig.height=10}
# ── Data provenance ──
# All numbers derived from objects loaded in subsequent chunks.
# Raw count matrix dimensions from isocall pipeline output.
# Filtered/classified counts from 01_prepare_data_mashr.R outputs.
# Pair counts from 02_build_profiles_mashr.R outputs.
# Gene-matched counts computed in load-profiles chunk below.
#
# We load the key objects here just for the flowchart numbers, then
# the full load happens in the subsequent chunks.

fmt <- function(x) format(x, big.mark = ",")

# Load just what we need for counts
.fc_expr   <- readRDS(file.path("data_mashr", "expression_data.rds"))
.fc_sm     <- readRDS(file.path("data_mashr", "sample_metadata.rds"))
.fc_nmd    <- readRDS(file.path("data_mashr", "nmd_classification.rds"))
.fc_gm     <- readRDS(file.path("data_mashr", "gene_map.rds"))

n_filtered   <- nrow(.fc_expr)
n_samples    <- ncol(.fc_expr)
n_dmso       <- sum(.fc_sm$treatment == "DMSO")
n_smg1i      <- sum(.fc_sm$treatment == "Smg1i")
n_nmd_as     <- length(.fc_nmd[["all_samples"]]$nmd)
n_nonnmd_as  <- length(.fc_nmd[["all_samples"]]$non_nmd)

# How many genes have both NMD and non-NMD isoforms?
nmd_genes <- unique(.fc_gm$gene_id[.fc_gm$isoform_id %in% .fc_nmd[["all_samples"]]$nmd])
nonnmd_genes <- unique(.fc_gm$gene_id[.fc_gm$isoform_id %in% .fc_nmd[["all_samples"]]$non_nmd])
# Genes with ≥2 non-NMD (needed for C4)
nonnmd_per_gene <- table(.fc_gm$gene_id[.fc_gm$isoform_id %in% .fc_nmd[["all_samples"]]$non_nmd])
genes_2plus_nonnmd <- names(nonnmd_per_gene[nonnmd_per_gene >= 2])
# Genes eligible for both C2 and C4
genes_eligible <- intersect(nmd_genes, genes_2plus_nonnmd)
n_genes_eligible <- length(genes_eligible)

# Pair counts (all_samples, before gene-matching)
.fc_c2 <- readRDS(file.path("data_mashr", "profiles_c2_allsamples.rds"))
.fc_c4 <- readRDS(file.path("data_mashr", "profiles_c4_allsamples.rds"))
n_c2_before <- nrow(.fc_c2)
n_c4_before <- nrow(.fc_c4)

# Gene-matched counts
.shared <- dplyr::inner_join(
  unique(.fc_c2[, c("gene_id", "reference_isoform_id")]),
  unique(.fc_c4[, c("gene_id", "reference_isoform_id")]),
  by = c("gene_id", "reference_isoform_id"))
n_matched_genes <- nrow(.shared)
n_c2_matched <- nrow(dplyr::semi_join(.fc_c2, .shared, by = c("gene_id", "reference_isoform_id")))
n_c4_matched <- nrow(dplyr::semi_join(.fc_c4, .shared, by = c("gene_id", "reference_isoform_id")))

rm(.fc_expr, .fc_sm, .fc_nmd, .fc_gm, .fc_c2, .fc_c4, .shared)

dot_spec <- sprintf('
digraph flowchart {
  graph [rankdir=TB, fontname="Helvetica", nodesep=2.0, ranksep=1.8, size="16,20"]
  node  [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=30, margin="0.8,0.6", width=8]
  edge  [color="#666666", penwidth=3.0]

  A [label=<<b>Isocall Count Matrix</b><br/>%s isoforms × %d samples>,
     fillcolor="#f0f0f0", color="#636363"]

  B [label=<<b>Sample Exclusion</b><br/>Remove DO donor 029T (PCA outlier)<br/>%s isoforms × %d samples>,
     fillcolor="#fee0d2", color="#de2d26"]

  C [label=<<b>5%% Filter + filterByExpr</b><br/>%s isoforms pass<br/>%d DMSO + %d Smg1i | 6 cell types>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  D [label=<<b>NMD Classification (mashr)</b><br/>NMD-sensitive: %s isoforms (union across CTs)<br/>Non-NMD: %s isoforms (intersection across CTs)>,
     fillcolor="#fff7bc", color="#d95f0e"]

  D2 [label=<<b>Eligible Genes</b><br/>%s genes have ≥1 NMD isoform AND ≥2 non-NMD isoforms<br/>(required for both C2 and C4 pair construction)>,
      fillcolor="#fff7bc", color="#d95f0e"]

  E [label=<<b>Pair Construction (all_samples)</b><br/>C2 (NMD): %s pairs (1 per gene)<br/>C4 (Control): %s pairs (1 per gene)>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  F [label=<<b>Gene-Matching</b><br/>%s genes with shared (gene, reference) in both C2 and C4<br/>C2: %s pairs | C4: %s pairs>,
     fillcolor="#c7e9c0", color="#31a354"]

  A -> B -> C -> D -> D2 -> E -> F
}',
  fmt(n_raw_isoforms), n_raw_samples,
  fmt(n_raw_isoforms), n_samples,
  fmt(n_filtered), n_dmso, n_smg1i,
  fmt(n_nmd_as), fmt(n_nonnmd_as),
  fmt(n_genes_eligible),
  fmt(n_c2_before), fmt(n_c4_before),
  fmt(n_matched_genes), fmt(n_c2_matched), fmt(n_c4_matched)
)

if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(dot_spec, width = 1200, height = 900)
} else {
  message("DiagrammeR not installed -- flowchart not rendered")
}
```

```{r load-data}
# ==============================================================================
# STAGE 1: Load raw data objects
# ==============================================================================

# ── 01_prepare_data_mashr.R outputs ──
# Source: isocall count matrix → TMM normalization → 5% filter → mashr NMD classification
data_dir  <- "data_mashr"
cache_dir <- file.path(data_dir, "analysis_cache")

expr_mat    <- readRDS(file.path(data_dir, "expression_data.rds"))
sample_meta <- readRDS(file.path(data_dir, "sample_metadata.rds"))
gene_map    <- readRDS(file.path(data_dir, "gene_map.rds"))
nmd_class   <- readRDS(file.path(data_dir, "nmd_classification.rds"))

# ── Infrastructure objects (isoform-intrinsic, DE-independent) ──
# Originally from Isopair pipeline scripts 02-05, copied to data_mashr/ by
# 02_build_profiles_mashr.R for self-containment.
# structures.rds: extractIsoformStructures() — exon coordinates per isoform
# cds.rds: extractCds() — CDS boundaries per isoform
# ptc.rds: testPtcAssociation() infrastructure — PTC status per isoform
# union_exons.rds: buildUnionExons() — merged exon models per gene
# isoform_union_mapping.rds: mapIsoformsToUnion() — isoform-to-union-exon mapping
# annotated_ue.rds: annotateUnionExons() — union exons with CDS/UTR labels
# cds_exons.rds: CDS mapped to individual exons
structures   <- readRDS(file.path(data_dir, "structures.rds"))
cds          <- readRDS(file.path(data_dir, "cds.rds"))
ptc          <- readRDS(file.path(data_dir, "ptc.rds"))
union_exons  <- readRDS(file.path(data_dir, "union_exons.rds"))
ue_mapping   <- readRDS(file.path(data_dir, "isoform_union_mapping.rds"))
annotated_ue <- readRDS(file.path(data_dir, "annotated_ue.rds"))
cds_exons    <- readRDS(file.path(data_dir, "cds_exons.rds"))

# ── Gene symbol lookup from mashr DE files ──
de_dir_mashr <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/mashr"
gene_symbol_lookup <- character(0)
for (de_f in list.files(de_dir_mashr, pattern = "^nmd_mashr_die_.*\\.csv$", full.names = TRUE)) {
  de_tmp <- read.csv(de_f, stringsAsFactors = FALSE)[, c("gene_id", "hgnc_symbol")]
  de_tmp <- de_tmp[!duplicated(de_tmp$gene_id) & !is.na(de_tmp$hgnc_symbol) &
                    de_tmp$hgnc_symbol != "", ]
  new_entries <- setdiff(de_tmp$gene_id, names(gene_symbol_lookup))
  gene_symbol_lookup[de_tmp$gene_id[de_tmp$gene_id %in% new_entries]] <-
    de_tmp$hgnc_symbol[de_tmp$gene_id %in% new_entries]
}

get_gene_sym <- function(gid) {
  sym <- gene_symbol_lookup[gid]
  if (is.na(sym) || sym == "") gid else sym
}

# Cell types (mashr: DO included, DD_ALI excluded from main)
cell_types  <- c("all_samples", "AT", "DD", "DO", "FB", "MV")
comparisons <- c("c2", "c4")
main_cts    <- c("AT", "DD", "DO", "FB", "MV")
```

```{r load-profiles}
# ==============================================================================
# STAGE 2: Load profiles and apply gene-matching
# ==============================================================================
#
# profiles_c{2,4}_*.rds are the output of 02_build_profiles_mashr.R. C2 pairs
# share their reference isoform with C4 (by construction in script 02), but C4
# may contain genes that have no NMD isoform (i.e., genes with >=2 non-NMD
# isoforms but no mashr-classified NMD isoform). Gene-matching restricts both
# sets to the intersection of (gene_id, reference_isoform_id), ensuring fair
# NMD vs Control comparison on identical gene sets.

profiles_list <- list()
for (ct in cell_types) {
  ct_label <- ct_to_label(ct)
  for (comp in comparisons) {
    key <- paste(comp, ct_label, sep = "_")
    prof_file <- file.path(data_dir, sprintf("profiles_%s.rds", key))
    if (file.exists(prof_file)) profiles_list[[key]] <- readRDS(prof_file)
  }
}

# ── Gene-matching ──
# Replicates the logic from 03_nmd_analysis_mashr.Rmd lines 743-775.
# For each cell type, restrict C2 and C4 profiles to genes where both have
# a pair with the same (gene_id, reference_isoform_id).
gene_match_summary <- list()
for (ct in cell_types) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(profiles_list) || !c4_key %in% names(profiles_list)) next

  n_c2_before <- nrow(profiles_list[[c2_key]])
  n_c4_before <- nrow(profiles_list[[c4_key]])

  c2_gene_refs <- unique(profiles_list[[c2_key]][, c("gene_id", "reference_isoform_id")])
  c4_gene_refs <- unique(profiles_list[[c4_key]][, c("gene_id", "reference_isoform_id")])
  shared_refs  <- dplyr::inner_join(c2_gene_refs, c4_gene_refs,
                                     by = c("gene_id", "reference_isoform_id"))

  profiles_list[[c2_key]] <- dplyr::semi_join(profiles_list[[c2_key]], shared_refs,
                                               by = c("gene_id", "reference_isoform_id"))
  profiles_list[[c4_key]] <- dplyr::semi_join(profiles_list[[c4_key]], shared_refs,
                                               by = c("gene_id", "reference_isoform_id"))

  gene_match_summary[[ct]] <- data.frame(
    cell_type      = ct,
    c2_before      = n_c2_before,
    c4_before      = n_c4_before,
    matched_genes  = nrow(shared_refs),
    c2_after       = nrow(profiles_list[[c2_key]]),
    c4_after       = nrow(profiles_list[[c4_key]])
  )
}
gene_match_df <- do.call(rbind, gene_match_summary)
```

```{r build-set-results}
# ==============================================================================
# STAGE 3: Build set_results from gene-matched profiles + cached analyses
# ==============================================================================
#
# set_results[[key]] is the primary data structure for all downstream analyses.
# $profiles: gene-matched splicing choice profiles (from STAGE 2 above)
# $divergence: quantifyPairDivergence() output (cached by script 03)
# $frame_comp: compareIsoformFrames() output (cached by script 03)
# $frame_walk: analyzeFrameWalk() output (cached by script 03)
# $ptc_assoc: testPtcAssociation() output (cached by script 03)
# $evt_freq: computed here from gene-matched profiles

set_results <- list()
for (key in names(profiles_list)) {
  profiles <- profiles_list[[key]]
  set_results[[key]] <- list(
    profiles   = profiles,
    divergence = tryCatch(readRDS(file.path(cache_dir, paste0("div_", key, ".rds"))), error = function(e) NULL),
    frame_comp = tryCatch(readRDS(file.path(cache_dir, paste0("fc_",  key, ".rds"))), error = function(e) NULL),
    frame_walk = tryCatch(readRDS(file.path(cache_dir, paste0("fw_",  key, ".rds"))), error = function(e) NULL),
    ptc_assoc  = tryCatch(readRDS(file.path(cache_dir, paste0("ptc_", key, ".rds"))), error = function(e) NULL),
    evt_freq   = compute_event_freq(profiles)
  )
}

# Load divergence for all_samples (used by multiple sections)
div_c2 <- set_results[["c2_allsamples"]]$divergence
div_c4 <- set_results[["c4_allsamples"]]$divergence
```

```{r verify-cache-consistency}
# ==============================================================================
# STAGE 4: Verify cached analysis objects match gene-matched profiles
# ==============================================================================
#
# The cached div/fc/fw/ptc objects were computed by script 03 from gene-matched
# profiles. We verify that the row counts in the cached objects match our
# gene-matched profiles_list. A mismatch would indicate the cache is stale or
# was computed from a different pair set.

verification_rows <- list()
for (key in names(set_results)) {
  sr <- set_results[[key]]
  n_profiles <- nrow(sr$profiles)

  # divergence: 1 row per pair (should exactly match profiles)
  n_div <- if (!is.null(sr$divergence)) nrow(sr$divergence) else NA
  # frame_comp$pair_summary: 1 row per pair (should exactly match)
  n_fc <- if (!is.null(sr$frame_comp)) nrow(sr$frame_comp$pair_summary) else NA
  # frame_walk$profile_summary: 1 row per pair (should exactly match)
  n_fw <- if (!is.null(sr$frame_walk)) nrow(sr$frame_walk$profile_summary) else NA
  # ptc_assoc$pair_summary: 1 row per pair where PTC analysis was possible
  #   (both isoforms coding with EJC data). Expected to be <= n_profiles.
  n_ptc <- if (!is.null(sr$ptc_assoc)) nrow(sr$ptc_assoc$pair_summary) else NA

  # Exact match expected for div, fc, fw
  div_ok <- is.na(n_div) || n_div == n_profiles
  fc_ok  <- is.na(n_fc)  || n_fc  == n_profiles
  fw_ok  <- is.na(n_fw)  || n_fw  == n_profiles
  # ptc: subset is OK (non-coding pairs excluded), but should not exceed profiles
  ptc_ok <- is.na(n_ptc) || n_ptc <= n_profiles

  verification_rows[[key]] <- data.frame(
    key = key, n_profiles = n_profiles,
    n_div = n_div, n_fc = n_fc, n_fw = n_fw, n_ptc = n_ptc,
    exact_match = div_ok & fc_ok & fw_ok,
    ptc_subset_ok = ptc_ok
  )
}
verification_df <- do.call(rbind, verification_rows)

if (!all(verification_df$exact_match)) {
  warning("CACHE MISMATCH DETECTED — div/fc/fw row counts do not match ",
          "gene-matched profiles. Affected keys: ",
          paste(verification_df$key[!verification_df$exact_match], collapse = ", "))
}
if (!all(verification_df$ptc_subset_ok)) {
  warning("PTC CACHE ANOMALY — ptc_assoc has MORE rows than profiles for: ",
          paste(verification_df$key[!verification_df$ptc_subset_ok], collapse = ", "))
}
```

```{r load-meta-and-precomputes}
# ==============================================================================
# STAGE 5: Load mashr meta-analysis logFC and precomputed objects
# ==============================================================================

# ── mashr meta-analysis logFC ──
# Source: mashr model fit on isocall DIE (6 conditions).
# PosteriorMean averaged across 5 main conditions (excluding DD_ALI).
# Used in Goal 1 to characterize NMD response magnitude.
mashr_model_path <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/mashr/mashr_isoform_model_2026.3.10.rds"
mashr_model <- readRDS(mashr_model_path)
post_mean <- ashr::get_pm(mashr_model)
meta_cols <- c("Smg1i_in_DD", "Smg1i_in_AT", "Smg1i_in_DO_ALI",
               "Smg1i_in_FB", "Smg1i_in_MV")
meta_logfc <- data.frame(
  txid = rownames(post_mean),
  mean_logFC = rowMeans(post_mean[, meta_cols, drop = FALSE])
)

# ── C4 protein comparison (gene-matched, used in productive frameshifts) ──
# Source: 04b_all_c4_protein_comparison.R
# Note: Goal 5 now uses unmatched data from 05d_protein_domain_analysis.R
c4_protein <- readRDS(file.path(cache_dir, "all_c4_protein_comparison_allsamples.rds"))

# ── Productive frameshift precompute ──
# Source: 04_productive_frameshift_precompute.R
# Input: fc_c4_allsamples.rds + fw_c4_allsamples.rds (gene-matched, from script 03)
#        + SQANTI protein FASTA + EBI/biomaRt API queries
# Output: characterization of same-start frameshift C4 pairs without PTC
pf_data <- readRDS(file.path(cache_dir, "productive_frameshift_allsamples.rds"))
```

```{r data-dimensions}
# ==============================================================================
# Summary: Gene-matching and data dimensions
# ==============================================================================
cat("=== Gene-Matching Summary ===\n")
render_dt(gene_match_df, caption = "Gene-Matching: Before and After")

cat(sprintf("\n=== Data Dimensions ===\n"))
cat(sprintf("Gene-matched profile keys: %d\n", length(profiles_list)))
cat(sprintf("C2 all_samples pairs (gene-matched): %d\n", nrow(profiles_list[["c2_allsamples"]])))
cat(sprintf("C4 all_samples pairs (gene-matched): %d\n", nrow(profiles_list[["c4_allsamples"]])))
cat(sprintf("PTC records: %d isoforms\n", nrow(ptc)))
cat(sprintf("CDS records: %d isoforms\n", nrow(cds)))
cat(sprintf("Meta-analysis logFC: %d isoforms\n", nrow(meta_logfc)))
cat(sprintf("C4 protein comparison (gene-matched): %d pairs\n", nrow(c4_protein$comparison)))
cat(sprintf("Productive frameshifts: %d pairs\n", nrow(pf_data$characterization)))

cat("\n=== Cache Verification ===\n")
render_dt(verification_df, caption = "Cache Consistency Check")
```


# Event Type Glossary {.tabset}

Isopair detects 12 structural event types organized into 4 categories. Each tab below
shows a schematic diagram of the event alongside its definition. In all diagrams, the
**Reference** (top) is the dominant non-NMD isoform and the **Comparator** (bottom) is
the isoform being compared. Events are always described from the comparator's perspective:
a **GAIN** means the comparator has additional sequence relative to the reference, and a
**LOSS** means the comparator is missing sequence present in the reference.

**Corrections from the exploratory report (03):** A5SS and A3SS schematics were swapped;
IR_diff_5 and IR_diff_3 schematics showed the retaining exon extending beyond the reference
when the code classifies "diff" when the retaining exon falls short.

```{r glossary, results='asis', fig.width=8, fig.height=3}
glossary_plot <- function(ref_s, ref_e, comp_s, comp_e, title, subtitle,
                          ref_cds = NULL, comp_cds = NULL, events = NULL) {
  plot_ucsc_pair(ref_s, ref_e, comp_s, comp_e, strand = "+",
                 ref_cds = ref_cds, comp_cds = comp_cds,
                 ref_label = "Reference", comp_label = "Comparator",
                 events = events, title = title, subtitle = subtitle)
}

# ---- Terminal Events ----

cat("## Alt_TSS\n\n")
cat("**Alternative Transcription Start Site.** The two isoforms begin at different\n")
cat("genomic positions. Detected when the 5'-most exon of one isoform does not overlap\n")
cat("the 5'-most exon of the other. Direction: GAIN if the comparator's TSS is further\n")
cat("upstream (extending the transcript); LOSS if it is further downstream (truncating it).\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(250, 400, 700), comp_e = c(350, 550, 850),
  title = "Alt_TSS: Alternative Transcription Start Site",
  subtitle = "Comparator starts at a different position (here, downstream = LOSS)")
print(p); cat("\n\n")

cat("## Alt_TES\n\n")
cat("**Alternative Transcription End Site.** The two isoforms end at different\n")
cat("genomic positions. Detected when the 3'-most exon of one isoform does not overlap\n")
cat("the 3'-most exon of the other. Direction: GAIN if the comparator extends further;\n")
cat("LOSS if it ends earlier.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 900),
  comp_s = c(100, 400, 700), comp_e = c(200, 550, 800),
  title = "Alt_TES: Alternative Transcription End Site",
  subtitle = "Comparator ends earlier (LOSS)")
print(p); cat("\n\n")

# ---- Boundary Events ----

cat("## A5SS\n\n")
cat("**Alternative 5' Splice Site.** An internal exon has a different **donor** (exon_end\n")
cat("on + strand) boundary while sharing the same **acceptor** (exon_start) boundary.\n")
cat("The exon is extended or truncated at its downstream/3' end.\n")
cat("Direction: GAIN if the comparator's exon is longer; LOSS if shorter.\n\n")
# A5SS: same acceptor (exon_start=400), different donor (exon_end: 550 vs 600)
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400, 700), comp_e = c(200, 600, 850),
  title = "A5SS: Alternative 5' Splice Site",
  subtitle = "Same acceptor (start), different donor (end). Comparator exon extends further (GAIN)")
print(p); cat("\n\n")

cat("## A3SS\n\n")
cat("**Alternative 3' Splice Site.** An internal exon has a different **acceptor** (exon_start\n")
cat("on + strand) boundary while sharing the same **donor** (exon_end) boundary.\n")
cat("The exon is extended or truncated at its upstream/5' end.\n")
cat("Direction: GAIN if the comparator's exon is longer; LOSS if shorter.\n\n")
# A3SS: different acceptor (exon_start: 400 vs 350), same donor (exon_end=550)
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 350, 700), comp_e = c(200, 550, 850),
  title = "A3SS: Alternative 3' Splice Site",
  subtitle = "Same donor (end), different acceptor (start). Comparator exon extends further (GAIN)")
print(p); cat("\n\n")

cat("## Partial_IR_5\n\n")
cat("**Partial Intron Retention (5' end).** An intron is partially retained: one isoform's\n")
cat("exon extends into the intron from its 5' side, covering what is intronic sequence in\n")
cat("the other isoform. The 3' boundary of the retaining exon falls within the intron.\n")
cat("Direction: GAIN if the comparator retains; LOSS if the reference retains.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400, 700), comp_e = c(200, 620, 850),
  title = "Partial_IR_5: Partial Intron Retention (5' side)",
  subtitle = "Comparator exon extends into the intron from the 5' end (GAIN)")
print(p); cat("\n\n")

cat("## Partial_IR_3\n\n")
cat("**Partial Intron Retention (3' end).** An intron is partially retained: one isoform's\n")
cat("exon extends into the intron from its 3' side, covering what is intronic sequence in\n")
cat("the other isoform. The 5' boundary of the retaining exon falls within the intron.\n")
cat("Direction: GAIN if the comparator retains; LOSS if the reference retains.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400, 640), comp_e = c(200, 550, 850),
  title = "Partial_IR_3: Partial Intron Retention (3' side)",
  subtitle = "Comparator exon extends into the intron from the 3' end (GAIN)")
print(p); cat("\n\n")

# ---- Exon-Level Events ----

cat("## SE (Skipped Exon)\n\n")
cat("**Skipped Exon.** An exon present in one isoform is completely absent from the\n")
cat("other. The flanking exons splice directly to each other, skipping over the exon.\n")
cat("Direction: GAIN if the comparator includes the exon (reference skips it);\n")
cat("LOSS if the comparator skips it (reference includes it).\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 600, 800), ref_e  = c(200, 500, 700, 900),
  comp_s = c(100, 400, 800),       comp_e = c(200, 500, 900),
  title = "SE: Skipped Exon",
  subtitle = "Comparator skips an internal exon present in the reference (LOSS)")
print(p); cat("\n\n")

cat("## Missing_Internal\n\n")
cat("**Missing Internal Exon.** Similar to SE but arising from a multi-exon skip:\n")
cat("one isoform is missing one or more internal exons that are present in the other.\n")
cat("Often co-occurs with other Missing_Internal events from the same splice decision.\n")
cat("Direction: GAIN if the comparator has the exon; LOSS if the comparator lacks it.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 350, 550, 750, 950), ref_e  = c(200, 450, 650, 850, 1050),
  comp_s = c(100, 950),                 comp_e = c(200, 1050),
  title = "Missing_Internal: Missing Internal Exon(s)",
  subtitle = "Comparator skips multiple internal exons (LOSS x 3)")
print(p); cat("\n\n")

# ---- Intron Retention Events ----

cat("## IR\n\n")
cat("**Full Intron Retention.** One isoform has a single exon that spans the region\n")
cat("covered by two or more exons (plus the intervening intron) in the other isoform.\n")
cat("The retaining exon boundaries match exactly. This is a full retention where the\n")
cat("entire intron is included in the retaining isoform's exon.\n")
cat("Direction: GAIN if the comparator retains the intron; LOSS if the reference retains it.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400),       comp_e = c(200, 850),
  title = "IR: Full Intron Retention",
  subtitle = "Comparator retains intron -- single exon spans two exons + intron in ref (GAIN)")
print(p); cat("\n\n")

cat("## IR_diff_5\n\n")
cat("**Intron Retention with 5' Boundary Difference.** Like IR, but the retaining\n")
cat("exon's 5' boundary differs from the first overlapping exon in the other isoform.\n")
cat("The intron is retained, but the retaining exon starts after the first reference\n")
cat("overlap exon (falls short at the 5' end).\n\n")
# IR_diff_5: comp retaining exon starts AFTER first ref overlap exon start
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 430),       comp_e = c(200, 850),
  title = "IR_diff_5: Intron Retention with 5' Boundary Difference",
  subtitle = "Retaining exon starts AFTER ref overlap exon (falls short at 5' end)")
print(p); cat("\n\n")

cat("## IR_diff_3\n\n")
cat("**Intron Retention with 3' Boundary Difference.** Like IR, but the retaining\n")
cat("exon's 3' boundary differs from the last overlapping exon in the other isoform.\n")
cat("The intron is retained, but the retaining exon ends before the last reference\n")
cat("overlap exon (falls short at the 3' end).\n\n")
# IR_diff_3: comp retaining exon ends BEFORE last ref overlap exon end
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400),       comp_e = c(200, 820),
  title = "IR_diff_3: Intron Retention with 3' Boundary Difference",
  subtitle = "Retaining exon ends BEFORE ref overlap exon (falls short at 3' end)")
print(p); cat("\n\n")

cat("## IR_diff_5_3\n\n")
cat("**Intron Retention with Both Boundary Differences.** Like IR, but both the 5'\n")
cat("and 3' boundaries of the retaining exon differ from the flanking exons in the\n")
cat("other isoform. The intron is retained but the retaining exon falls short at\n")
cat("both ends.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 430),       comp_e = c(200, 820),
  title = "IR_diff_5_3: Intron Retention with Both Boundary Differences",
  subtitle = "Retaining exon falls short at both 5' and 3' ends")
print(p); cat("\n\n")
```

In the IR section of this report, **IR_diff** refers to the aggregate of IR_diff_5,
IR_diff_3, and IR_diff_5_3 events.


# 1. Baseline Splicing and NMD-Enriched Events {#goal1-baseline}

**Goal 1:** Characterize baseline alternative splicing (Control C4 pairs) and identify which
splicing event types are enriched in NMD vs Control.

**Methods:** Structural profiles quantify splicing event types for each pair. Sequence
divergence (`quantifyPairDivergence()`) measures exonic sequence shared between isoforms.
NMD vs Control enrichment is tested by Fisher's exact test per event type.

## Figure 1: Sequence Shared — NMD vs Control

```{r goal1-fig1-pct-shared, fig.width=8, fig.height=5}
# ── Data provenance ──
# Starting data: gene-matched profiles → cached divergence
# div_c2 = set_results[["c2_allsamples"]]$divergence
#   Source: quantifyPairDivergence(gene-matched profiles, structures, cds),
#   cached by script 03
#   $pct_shared: percentage of exonic sequence shared between ref and comp
# div_c4: same for Control
# Result: density comparison of sequence similarity between NMD and Control pairs

pct_shared_data <- rbind(
  data.frame(comparison = "NMD",     pct_shared = div_c2$pct_shared),
  data.frame(comparison = "Control", pct_shared = div_c4$pct_shared))
pct_shared_data <- pct_shared_data[!is.na(pct_shared_data$pct_shared), ]

fig1_shared <- ggplot(pct_shared_data, aes(x = pct_shared, fill = comparison)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = pal) +
  labs(title = "Exonic Sequence Shared (%): NMD vs Control",
       subtitle = "NMD pairs shifted right = more similar (targeted splicing changes)",
       x = "Sequence Shared (%)", y = "Density", fill = "")
save_fig(fig1_shared, "fig1_pct_shared_density")

# Summary stats
shared_summary <- pct_shared_data |>
  group_by(comparison) |>
  summarise(n = n(), median = round(median(pct_shared), 1),
            mean = round(mean(pct_shared), 1), .groups = "drop")
render_dt(shared_summary, caption = "Table 1a: Sequence Shared Summary")
```

Columns: **comparison** = NMD or Control; **n** = number of pairs; **median/mean** =
median and mean percentage of exonic sequence shared between reference and
comparator (from `Isopair::quantifyPairDivergence()`).

## Table 1b: Splicing Event Enrichment — NMD vs Control

For each event type, the denominator is all gene-matched pairs (not restricted
to coding) — a pair is counted if it has ≥1 event of that type. Fisher's exact
test compares the proportion of pairs with each event type between NMD (C2) and
Control (C4), both on the same gene-matched set.

```{r goal1-table1b-event-enrichment}
# ── Data provenance ──
# Starting data: gene-matched profiles (set_results[[key]]$profiles)
# compute_event_freq(profiles) counts pairs with each event type using:
#   - Boolean columns (tss_changed, tes_changed, n_a5ss > 0, etc.) from profiles
#   - Partial_IR_5/3 from detailed_events list-column
#   These are computed by buildProfiles → compareSplicingChoices
# Fisher's exact test: event prevalence in NMD vs Control on gene-matched sets

ef_c2 <- compute_event_freq(set_results[["c2_allsamples"]]$profiles)
ef_c4 <- compute_event_freq(set_results[["c4_allsamples"]]$profiles)
n_c2 <- nrow(set_results[["c2_allsamples"]]$profiles)
n_c4 <- nrow(set_results[["c4_allsamples"]]$profiles)

enrichment <- merge(ef_c2, ef_c4, by = "event_type", suffixes = c("_NMD", "_Ctrl"))

# Fisher's exact test per event type
enrichment$fisher_OR <- NA_real_
enrichment$fisher_p <- NA_real_
for (i in seq_len(nrow(enrichment))) {
  a <- enrichment$n_pairs_with_event_NMD[i]
  b <- n_c2 - a
  c <- enrichment$n_pairs_with_event_Ctrl[i]
  d <- n_c4 - c
  ft <- tryCatch(fisher.test(matrix(c(a, b, c, d), nrow = 2)), error = function(e) NULL)
  if (!is.null(ft)) {
    enrichment$fisher_OR[i] <- round(ft$estimate, 2)
    enrichment$fisher_p[i] <- ft$p.value
  }
}

enrichment$direction <- ifelse(enrichment$fisher_p > 0.05, "Non-significant",
  ifelse(enrichment$fisher_OR > 1, "Enriched in NMD", "Enriched in Control"))
enrichment <- enrichment[order(enrichment$fisher_p), ]

enrich_display <- enrichment[, c("event_type", "n_pairs_with_event_NMD", "pct_of_pairs_NMD",
                                 "n_pairs_with_event_Ctrl", "pct_of_pairs_Ctrl",
                                 "fisher_OR", "fisher_p", "direction")]
enrich_display$fisher_p <- signif(enrich_display$fisher_p, 3)
render_dt(enrich_display,
          caption = "Table 1b: Splicing Event Type Enrichment — NMD vs Control")
write.csv(enrichment, file.path(table_dir, "table1b_event_enrichment.csv"), row.names = FALSE)
```

Columns: **event_type** = splicing event type (see Event Type Glossary);
**n_pairs_with_event_NMD** = number of gene-matched NMD (C2) pairs with ≥1
event of this type; **pct_of_pairs_NMD** = percentage of all
`r fmt(n_c2)` NMD pairs; **n_pairs_with_event_Ctrl/pct_of_pairs_Ctrl** =
same for Control (C4, `r fmt(n_c4)` pairs); **fisher_OR** = Fisher's exact
test odds ratio (>1 = NMD-enriched); **fisher_p** = p-value; **direction** =
Enriched in NMD, Enriched in Control, or Non-significant (p > 0.05).

### Figure 1b: Splicing Event Prevalence — NMD vs Control

```{r goal1-fig1b-event-barplot, fig.width=10, fig.height=6}
# ── Data provenance ──
# enrichment: from goal1-table1b-event-enrichment chunk
# Paired barplot of event prevalence (% of pairs) by comparison type

bar_df <- rbind(
  data.frame(event_type = enrichment$event_type,
             comparison = "NMD",
             pct = enrichment$pct_of_pairs_NMD),
  data.frame(event_type = enrichment$event_type,
             comparison = "Control",
             pct = enrichment$pct_of_pairs_Ctrl))

# Order events by NMD prevalence (descending)
evt_order <- enrichment$event_type[order(-enrichment$pct_of_pairs_NMD)]
bar_df$event_type <- factor(bar_df$event_type, levels = evt_order)

# Significance stars
sig_df <- data.frame(
  event_type = factor(enrichment$event_type, levels = evt_order),
  y = pmax(enrichment$pct_of_pairs_NMD, enrichment$pct_of_pairs_Ctrl) + 2,
  label = ifelse(enrichment$fisher_p < 0.001, "***",
           ifelse(enrichment$fisher_p < 0.01, "**",
             ifelse(enrichment$fisher_p < 0.05, "*", "ns"))))

fig1b <- ggplot(bar_df, aes(x = event_type, y = pct, fill = comparison)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(data = sig_df, aes(x = event_type, y = y, label = label),
            inherit.aes = FALSE, size = 4, fontface = "bold") +
  scale_fill_manual(values = pal) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Splicing Event Prevalence: NMD vs Control",
       subtitle = sprintf("NMD: %s pairs, Control: %s pairs. Significance: Fisher's exact test (* p<0.05, ** p<0.01, *** p<0.001).",
                           fmt(n_c2), fmt(n_c4)),
       x = "", y = "% of Pairs with Event", fill = "Comparison") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

save_fig(fig1b, "fig1b_event_prevalence_barplot", width = 10, height = 6)
```

## Table 1c: Gain/Loss Direction by Event Type

For each event type, the comparator can GAIN sequence (comparator has additional
sequence relative to reference) or LOSE sequence (comparator is missing sequence
present in reference). The table below shows event counts and the percentage of
events in each direction, separately for NMD and Control pairs.

```{r goal1-table1c-gain-loss}
# ── Data provenance ──
# Starting data: gene-matched profiles (set_results[[key]]$profiles)
# $detailed_events: list-column, one data.frame per pair from buildProfiles
#   Each event row has event_type and direction (GAIN or LOSS)
#   Direction is from comparator's perspective (see Event Type Glossary)
# For each comparison (NMD, Control): count events by type × direction,
#   compute % GAIN and % LOSS within each event type

get_gain_loss_pct <- function(profiles, label) {
  all_events <- do.call(rbind, profiles$detailed_events)
  if (is.null(all_events) || nrow(all_events) == 0) return(NULL)
  gl <- all_events |>
    group_by(event_type, direction) |>
    summarise(n = n(), .groups = "drop")
  totals <- gl |> group_by(event_type) |> summarise(total = sum(n), .groups = "drop")
  gl <- merge(gl, totals, by = "event_type")
  gl$pct <- round(100 * gl$n / gl$total, 1)
  gl$comparison <- label
  gl
}

gl_c2 <- get_gain_loss_pct(set_results[["c2_allsamples"]]$profiles, "NMD")
gl_c4 <- get_gain_loss_pct(set_results[["c4_allsamples"]]$profiles, "Control")
gl_all <- rbind(gl_c2, gl_c4)

gl_wide <- gl_all |>
  tidyr::pivot_wider(
    id_cols = event_type,
    names_from = c(comparison, direction),
    values_from = c(n, pct),
    values_fill = 0) |>
  arrange(event_type)

col_order <- c("event_type",
  "n_NMD_GAIN", "n_NMD_LOSS", "pct_NMD_GAIN",
  "n_Control_GAIN", "n_Control_LOSS", "pct_Control_GAIN")
existing_cols <- intersect(col_order, names(gl_wide))
gl_display <- gl_wide[, existing_cols, drop = FALSE]

gl_display$fisher_p <- NA_real_
gl_display$direction_diff <- NA_character_
for (i in seq_len(nrow(gl_display))) {
  ng <- gl_display$n_NMD_GAIN[i]
  nl <- gl_display$n_NMD_LOSS[i]
  cg <- gl_display$n_Control_GAIN[i]
  cl <- gl_display$n_Control_LOSS[i]
  if (ng + nl > 0 && cg + cl > 0) {
    ft <- tryCatch(fisher.test(matrix(c(ng, nl, cg, cl), nrow = 2)), error = function(e) NULL)
    if (!is.null(ft)) {
      gl_display$fisher_p[i] <- signif(ft$p.value, 3)
      pct_nmd <- 100 * ng / (ng + nl)
      pct_ctrl <- 100 * cg / (cg + cl)
      gl_display$direction_diff[i] <- if (ft$p.value > 0.05) "Non-significant"
        else if (pct_nmd > pct_ctrl) "More GAIN in NMD"
        else "More LOSS in NMD"
    }
  }
}

render_dt(gl_display, caption = "Table 1c: Gain/Loss Events by Type — NMD vs Control")
write.csv(gl_display, file.path(table_dir, "table1c_gain_loss_events.csv"), row.names = FALSE)
```

Columns: **event_type** = splicing event type; **n_NMD_GAIN/n_NMD_LOSS** =
count of NMD events with GAIN/LOSS direction (from comparator's perspective);
**pct_NMD_GAIN** = percentage of NMD events of that type with GAIN direction;
**n_Control_GAIN/n_Control_LOSS/pct_Control_GAIN** = same for Control;
**fisher_p** = Fisher's exact test p-value comparing GAIN/LOSS ratio between
NMD and Control; **direction_diff** = "More GAIN in NMD", "More LOSS in NMD",
or "Non-significant" (p > 0.05).

### Figure 1c: GAIN Direction by Event Type — NMD vs Control

```{r goal1-fig1c-gain-loss-barplot, fig.width=10, fig.height=6}
# ── Data provenance ──
# gl_display: from goal1-table1c-gain-loss chunk
# Paired barplot of % GAIN direction by event type, NMD vs Control

bar1c <- rbind(
  data.frame(event_type = gl_display$event_type,
             comparison = "NMD",
             pct_gain = gl_display$pct_NMD_GAIN),
  data.frame(event_type = gl_display$event_type,
             comparison = "Control",
             pct_gain = gl_display$pct_Control_GAIN))

# Order events by NMD % GAIN (descending)
evt_order_1c <- gl_display$event_type[order(-gl_display$pct_NMD_GAIN)]
bar1c$event_type <- factor(bar1c$event_type, levels = evt_order_1c)

# Significance stars
sig1c <- data.frame(
  event_type = factor(gl_display$event_type, levels = evt_order_1c),
  y = pmax(gl_display$pct_NMD_GAIN, gl_display$pct_Control_GAIN) + 3,
  label = ifelse(is.na(gl_display$fisher_p), "",
           ifelse(gl_display$fisher_p < 0.001, "***",
             ifelse(gl_display$fisher_p < 0.01, "**",
               ifelse(gl_display$fisher_p < 0.05, "*", "ns")))))

fig1c <- ggplot(bar1c, aes(x = event_type, y = pct_gain, fill = comparison)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(data = sig1c, aes(x = event_type, y = y, label = label),
            inherit.aes = FALSE, size = 4, fontface = "bold") +
  scale_fill_manual(values = pal) +
  scale_y_continuous(limits = c(0, 105), expand = expansion(mult = c(0, 0))) +
  labs(title = "GAIN Direction by Event Type: NMD vs Control",
       subtitle = "% of events where comparator GAINS sequence relative to reference. Significance: Fisher's exact test.",
       x = "", y = "% Events with GAIN Direction", fill = "Comparison") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

save_fig(fig1c, "fig1c_gain_direction_barplot", width = 10, height = 6)
```

# 2. PTC Enrichment Validates NMD Classification {#goal2}

**Goal 2:** Demonstrate that NMD-responsive comparator isoforms are enriched for premature
termination codons (PTCs), validating the mashr-based NMD classification.

**Methods:** A PTC is defined as a stop codon >50 nt upstream of the last exon-junction
complex (EJC). Fisher's exact test compares PTC rates between NMD (C2) and Control (C4)
comparator isoforms, restricted to pairs where both reference and comparator are coding.
PTC position metrics (distance to EJC, % CDS, number of downstream EJCs) characterize NMD
response strength using mashr meta-analysis posterior means averaged across 5 conditions
(DD, AT, DO_ALI, FB, MV).

## PTC Rates

```{r goal1-ptc-rates-subsetting}
# ── Data provenance ──
# Starting data: gene-matched profiles (profiles_list → set_results$profiles)
#   Pipeline: 01 (mashr NMD classification) → 02 (pair generation + buildProfiles)
#   → gene-matching (this report, STAGE 2) → set_results (STAGE 3)
#
# Subsetting for PTC rate calculation:
#   1. Start with gene-matched C2 (NMD) and C4 (Control) profiles
#   2. Restrict to "coding pairs": both reference AND comparator isoform must have
#      coding_status == "coding" in cds.rds (from Isopair extractCds() on SQANTI CDS)
#   3. Numerator: comparator isoforms with has_ptc == TRUE in ptc.rds
#      PTC definition: stop codon >50 nt upstream of the last exon-junction complex (EJC)
#   4. Denominator: total coding pairs (from step 2)
#   5. Fisher's exact test: 2×2 table of (PTC+/PTC−) × (NMD/Control)

# Show the subsetting explicitly
coding_ids <- cds$isoform_id[cds$coding_status == "coding"]
n_gm_c2 <- nrow(set_results[["c2_allsamples"]]$profiles)
n_gm_c4 <- nrow(set_results[["c4_allsamples"]]$profiles)

p_c2 <- set_results[["c2_allsamples"]]$profiles
cod_c2 <- p_c2[p_c2$reference_isoform_id %in% coding_ids &
                p_c2$comparator_isoform_id %in% coding_ids, ]
p_c4 <- set_results[["c4_allsamples"]]$profiles
cod_c4 <- p_c4[p_c4$reference_isoform_id %in% coding_ids &
                p_c4$comparator_isoform_id %in% coding_ids, ]
```

Starting from `r fmt(n_gm_c2)` gene-matched NMD (C2) pairs and
`r fmt(n_gm_c4)` gene-matched Control (C4) pairs, we restrict to **coding
pairs** where both the reference and comparator isoform are classified as
protein-coding by SQANTI (`coding_status == "coding"` in `cds.rds`). This
yields `r fmt(nrow(cod_c2))` coding NMD pairs and `r fmt(nrow(cod_c4))` coding
Control pairs. The PTC rate is the fraction of these coding comparator isoforms
whose stop codon falls >50 nt upstream of the last exon-junction complex (EJC),
as computed by `Isopair::testPtcAssociation()`.

```{r goal1-ptc-rates}
ptc_rates <- compute_ptc_rates_row("all_samples")
kable(ptc_rates, caption = "Table 2: PTC Rates — all_samples",
      digits = c(0, 0, 0, 0, 1, 0, 1, 2, 2))
write.csv(ptc_rates, file.path(table_dir, "table2_ptc_rates_allsamples.csv"), row.names = FALSE)
```

Columns: **cell_type** = sample set; **n_coding_pairs_NMD/Ctrl** = number of
coding pairs (both ref and comp coding) in NMD (C2) / Control (C4);
**NMD_comp_ptc_n/pct** = count and percentage of NMD comparators with PTC;
**Ctrl_comp_ptc_n/pct** = same for Control; **fisher_OR** = odds ratio from
Fisher's exact test; **p_value** = Fisher p-value.

## Figure 2: Stop Codon Distance to Last EJC

```{r goal1-fig1-stop-dist-pop}
# ── Population derivation ──
# Start from gene-matched comparators, filter to those with non-NA ptc_distance.
# This is broader than Table 2 (not restricted to coding pairs).
fig2_pop <- list()
for (comp in c("c2", "c4")) {
  key <- paste0(comp, "_allsamples")
  comp_ids <- set_results[[key]]$profiles$comparator_isoform_id
  n_total <- length(comp_ids)
  ptc_sub <- ptc[ptc$isoform_id %in% comp_ids & !is.na(ptc$ptc_distance), ]
  fig2_pop[[comp]] <- list(n_total = n_total, n_with_dist = nrow(ptc_sub))
}
```

Starting from `r fmt(fig2_pop$c2$n_total)` gene-matched NMD comparators,
`r fmt(fig2_pop$c2$n_with_dist)` have a non-NA `ptc_distance` in `ptc.rds`
(i.e., have an annotated CDS with a stop codon and ≥1 exon-junction complex).
For Control: `r fmt(fig2_pop$c4$n_with_dist)` of `r fmt(fig2_pop$c4$n_total)`.
This is a broader set than Table 2, which additionally requires **both**
reference and comparator to be coding (pair-level filter). Here only the
comparator needs `ptc_distance` (isoform-level filter).

```{r goal1-fig1-stop-dist, fig.width=8, fig.height=5}
stop_dist_all <- {
  rows <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_allsamples")
    comp_ids <- set_results[[key]]$profiles$comparator_isoform_id
    ptc_sub <- ptc[ptc$isoform_id %in% comp_ids & !is.na(ptc$ptc_distance), ]
    if (nrow(ptc_sub) == 0) next
    rows[[comp]] <- data.frame(
      comparison = comp_label(comp),
      ptc_distance = ptc_sub$ptc_distance,
      has_ptc = ptc_sub$has_ptc)
  }
  do.call(rbind, rows)
}

fig1 <- ggplot(stop_dist_all, aes(x = ptc_distance, fill = comparison)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
  annotate("text", x = 50, y = Inf, label = "PTC threshold (50 nt)",
           hjust = -0.05, vjust = 1.5, size = 3, color = "red") +
  scale_x_continuous(limits = c(
    quantile(stop_dist_all$ptc_distance, 0.01, na.rm = TRUE),
    quantile(stop_dist_all$ptc_distance, 0.99, na.rm = TRUE))) +
  scale_fill_manual(values = pal) +
  labs(title = "Stop Codon Distance to Last EJC: NMD vs Control Comparators",
       subtitle = "All comparator isoforms with annotated stop codon + EJC. Dashed red = 50 nt PTC threshold.",
       x = "Distance from stop codon to last EJC (nt)", y = "Density", fill = "")
save_fig(fig1, "fig2_stop_codon_distance")
```

## Figure 3: NMD Response vs Stop Codon Distance

```{r goal1-fig2-logfc-dist, fig.width=8, fig.height=5}
# ── Data provenance ──
# Starting data: gene-matched C2 all_samples profiles
# Step 1: Extract comparator_isoform_id from gene-matched C2 profiles
# Step 2: Look up ptc features (ptc_distance, has_ptc, orf_length, n_downstream_ejcs)
#   from ptc.rds for those comparators
# Step 3: Merge with meta_logfc — mashr PosteriorMean averaged across 5 conditions
#   (DD, AT, DO_ALI, FB, MV). Source: mashr_isoform_model_2026.3.10.rds
# Result: scatter of NMD response magnitude vs stop codon position

comp_ids_as <- set_results[["c2_allsamples"]]$profiles$comparator_isoform_id
ptc_sub <- ptc[ptc$isoform_id %in% comp_ids_as & !is.na(ptc$ptc_distance),
                c("isoform_id", "ptc_distance", "has_ptc", "orf_length", "n_downstream_ejcs")]
n_with_logfc <- sum(comp_ids_as %in% meta_logfc$txid)
nmd_logfc_dist <- merge(meta_logfc[meta_logfc$txid %in% comp_ids_as, ],
                         ptc_sub, by.x = "txid", by.y = "isoform_id")
```

Of `r fmt(length(comp_ids_as))` gene-matched NMD comparators,
`r fmt(nrow(ptc_sub))` have non-NA `ptc_distance` and
`r fmt(n_with_logfc)` have a mashr posterior mean logFC. After merging
(requiring both), `r fmt(nrow(nmd_logfc_dist))` comparators remain.
`r if (length(comp_ids_as) - n_with_logfc > 0) paste0("The ", fmt(length(comp_ids_as) - n_with_logfc), " comparators absent from the mashr model are novel isoforms not tested for differential expression.") else "All comparators have mashr logFC values."`

```{r goal1-fig2-logfc-dist-plot, fig.width=8, fig.height=5}
fig2 <- ggplot(nmd_logfc_dist, aes(x = ptc_distance, y = mean_logFC)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "loess", se = TRUE, color = "steelblue") +
  geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(limits = c(
    quantile(nmd_logfc_dist$ptc_distance, 0.01, na.rm = TRUE),
    quantile(nmd_logfc_dist$ptc_distance, 0.99, na.rm = TRUE))) +
  labs(title = sprintf("NMD Response vs Stop Codon Distance (n = %d)", nrow(nmd_logfc_dist)),
       subtitle = "mashr meta-analysis mean logFC (5 conditions). Blue = loess.",
       x = "Distance from stop codon to last EJC (nt)",
       y = "Mean logFC (mashr)")
save_fig(fig2, "fig3_nmd_response_vs_distance")

# Correlation summary
cor_test <- cor.test(nmd_logfc_dist$ptc_distance, nmd_logfc_dist$mean_logFC,
                      method = "spearman")
kable(data.frame(n = nrow(nmd_logfc_dist),
                 spearman_rho = round(cor_test$estimate, 3),
                 p_value = cor_test$p.value),
      caption = "Spearman correlation: Stop-to-EJC distance vs mean logFC")
```

## Figure 4: NMD Strength by Downstream EJC Count

```{r goal1-fig3-pop}
as_ptc_pos <- nmd_logfc_dist[nmd_logfc_dist$has_ptc == TRUE, ]
as_ptc_pos$logFC <- as_ptc_pos$mean_logFC
```

Restricted to the `r fmt(nrow(as_ptc_pos))` **PTC-positive** comparators from
Figure 3's population of `r fmt(nrow(nmd_logfc_dist))` (`has_ptc == TRUE`),
binned by number of downstream EJCs.

```{r goal1-fig3-ejc-boxplot, fig.width=9, fig.height=5}
# Bin EJC counts >= 7 into "7+"
as_ptc_pos$ejc_bin <- ifelse(as_ptc_pos$n_downstream_ejcs >= 7,
                              "7+", as.character(as_ptc_pos$n_downstream_ejcs))
as_ptc_pos$ejc_bin <- factor(as_ptc_pos$ejc_bin,
                              levels = c(as.character(1:6), "7+"))

ejc_counts <- as_ptc_pos |>
  count(ejc_bin) |>
  mutate(label = paste0("n=", n))

fig3 <- ggplot(as_ptc_pos, aes(x = ejc_bin, y = logFC)) +
  geom_violin(fill = "lightblue", alpha = 0.5) +
  geom_boxplot(width = 0.15, outlier.alpha = 0.2) +
  geom_text(data = ejc_counts,
            aes(x = ejc_bin,
                y = max(as_ptc_pos$logFC, na.rm = TRUE) + 0.3, label = label),
            size = 2.8) +
  labs(title = "NMD Comparators: Mean logFC by Number of Downstream EJCs",
       subtitle = "mashr meta-analysis. logFC rises sharply from 1-4 EJCs, then plateaus.",
       x = "Number of Downstream EJCs", y = "Mean logFC (mashr)")
save_fig(fig3, "fig4_ejc_vs_nmd_strength", width = 9)
```

## Table 3: PTC Rates Across Cell Types

```{r goal1-table2-ct-ptc}
# ── Data provenance ──
# Same pipeline as Table 2 (compute_ptc_rates_row), iterated across main_cts.
# Each cell type uses its own gene-matched profiles from set_results.
ptc_ct <- do.call(rbind, lapply(c("all_samples", main_cts), compute_ptc_rates_row))
render_dt(ptc_ct, caption = "Table 3: PTC Rates Across Cell Types")
write.csv(ptc_ct, file.path(table_dir, "table3_ptc_rates_by_ct.csv"), row.names = FALSE)
```

Same columns as Table 2, one row per cell type. all_samples uses the union
(NMD) and intersection (non-NMD) across AT, DD, DO, FB, MV.


# 3. Splicing Mechanisms Producing PTCs {#goal3}

**Goal 3:** Identify the splicing mechanisms that produce PTCs in NMD-responsive isoforms.

**Methods:** Frame comparison (`compareIsoformFrames()`) classifies each pair by start codon
and reading frame relationship. Frame walk (`analyzeFrameWalk()`) identifies which splicing
events cause frameshifts. The Sankey diagram traces flows from splice event type through frame
category to PTC outcome for NMD comparators.

## Figure 5: Frame Category Distribution

```{r goal2-fig4-pop}
# Count excluded categories for prose
fc_c2_ps <- set_results[["c2_allsamples"]]$frame_comp$pair_summary
fc_c4_ps <- set_results[["c4_allsamples"]]$frame_comp$pair_summary
n_noncoding_c2 <- sum(fc_c2_ps$frame_category %in% c("non_coding", "no_shared_cds"))
n_noncoding_c4 <- sum(fc_c4_ps$frame_category %in% c("non_coding", "no_shared_cds"))
```

`compareIsoformFrames()` assigns each pair a frame category based on start codon
and reading frame relationship. Of `r fmt(nrow(fc_c2_ps))` NMD pairs,
`r fmt(n_noncoding_c2)` are classified as `non_coding` or `no_shared_cds` and
excluded from the figure (Control: `r fmt(n_noncoding_c4)` of
`r fmt(nrow(fc_c4_ps))` excluded). The denominator (y-axis) is the number of
**coding pairs** (`r fmt(nrow(fc_c2_ps) - n_noncoding_c2)` NMD,
`r fmt(nrow(fc_c4_ps) - n_noncoding_c4)` Control), so the 5 categories sum
to 100%.

```{r goal2-fig4-frame-category, fig.width=10, fig.height=6}
# ── Data provenance ──
# Starting data: gene-matched profiles → cached frame comparison
# set_results[[key]]$frame_comp: compareIsoformFrames() output, cached by script 03
#   Input to compareIsoformFrames(): gene-matched profiles + cds_exons.rds + cds.rds
#   $pair_summary contains frame_category per pair (same_start_in_frame, same_start_frameshift, etc.)
# ptc.rds$has_ptc: PTC status per isoform (isoform-intrinsic, independent of pairing)
# Denominator (total_pairs): nrow(gene-matched profiles) — ensures consistent
#   percentages between profiles and frame_comp (verified in STAGE 4)

coding_cats <- c("same_start_in_frame", "same_start_frameshift",
                 "diff_start_same_frame", "diff_start_diff_frame",
                 "diff_start_diff_frame_with_frameshift")
cat_labels <- c(
  same_start_in_frame = "Same Start\nSame Frame",
  same_start_frameshift = "Same Start\nFrameshift",
  diff_start_same_frame = "Diff Start\nSame Frame",
  diff_start_diff_frame = "Diff Start\nDiff Frame",
  diff_start_diff_frame_with_frameshift = "Diff Start\nDiff Frame + FS")

ptc_pos_ids <- ptc$isoform_id[ptc$has_ptc == TRUE]

# 4-color palette: NMD/Control × PTC+/No PTC
pal4 <- c("NMD PTC+"    = "#d6604d",   # dark orange-red
          "NMD No PTC"  = "#fddbc7",   # light orange
          "Ctrl PTC+"   = "#4393c3",   # dark blue
          "Ctrl No PTC" = "#d1e5f0")   # light blue

stacked_df <- do.call(rbind, lapply(c("c2", "c4"), function(comp) {
  key <- paste0(comp, "_allsamples")
  fc <- set_results[[key]]$frame_comp
  if (is.null(fc)) return(NULL)
  ps <- fc$pair_summary
  # Denominator: coding pairs only (exclude non_coding and no_shared_cds)
  coding_pairs_total <- sum(!ps$frame_category %in% c("non_coding", "no_shared_cds"))
  comp_lbl <- comp_label(comp)
  comp_short <- ifelse(comp_lbl == "NMD", "NMD", "Ctrl")

  do.call(rbind, lapply(coding_cats, function(cat) {
    cat_pairs <- ps[ps$frame_category == cat, ]
    n_cat <- nrow(cat_pairs)
    n_ptc <- sum(cat_pairs$comparator_isoform_id %in% ptc_pos_ids)
    n_no_ptc <- n_cat - n_ptc
    pct_ptc <- 100 * n_ptc / max(coding_pairs_total, 1)
    pct_no_ptc <- 100 * n_no_ptc / max(coding_pairs_total, 1)
    rbind(
      data.frame(frame_category = cat, category_label = cat_labels[cat],
                 comparison = comp_lbl, ptc_status = "PTC+",
                 fill_group = paste(comp_short, "PTC+"),
                 n = n_ptc, pct = pct_ptc),
      data.frame(frame_category = cat, category_label = cat_labels[cat],
                 comparison = comp_lbl, ptc_status = "No PTC",
                 fill_group = paste(comp_short, "No PTC"),
                 n = n_no_ptc, pct = pct_no_ptc))
  }))
}))

stacked_df$category_label <- factor(stacked_df$category_label,
  levels = cat_labels[coding_cats])
stacked_df$fill_group <- factor(stacked_df$fill_group,
  levels = c("NMD PTC+", "NMD No PTC", "Ctrl PTC+", "Ctrl No PTC"))

fig4 <- ggplot(stacked_df, aes(x = category_label, y = pct, fill = fill_group)) +
  geom_col() +
  facet_wrap(~comparison) +
  scale_fill_manual(values = pal4) +
  labs(title = "Frame Category Distribution: NMD vs Control (all_samples)",
       subtitle = "% of total pairs. PTC+ = comparator has premature termination codon.",
       x = "Frame Category", y = "% of Total Pairs", fill = "") +
  theme(axis.text.x = element_text(size = 8))
save_fig(fig4, "fig5_frame_category_distribution", width = 10, height = 6)
```

## Table 4: Frame Category Counts

```{r goal2-table3-frame-cats}
fc_rows <- compute_frame_category_rows("all_samples")
render_dt(fc_rows, caption = "Table 4: Frame Category Distribution — all_samples")
write.csv(fc_rows, file.path(table_dir, "table4_frame_categories.csv"), row.names = FALSE)
```

Columns: **cell_type**; **frame_category** = reading frame relationship between
reference and comparator (from `Isopair::compareIsoformFrames()`);
**NMD_n/NMD_pct** = count and percentage in NMD (C2) pairs;
**Ctrl_n/Ctrl_pct** = same for Control (C4). The denominator for percentages is
**coding pairs** (excluding `non_coding` and `no_shared_cds`), consistent with
Figure 5 — the 5 coding categories sum to 100%. The `non_coding` and
`no_shared_cds` rows show counts only (no percentage).

Note: the coding pair count here differs from Table 2 because `compareIsoformFrames()`
classifies some individually-coding pairs as `no_shared_cds` (both isoforms are
coding but their CDS regions don't overlap). Table 2's "coding pairs" filter
requires only that both isoforms have `coding_status == "coding"` in `cds.rds`.

## How PTCs Arise: Frameshift vs Non-Frameshift Mechanisms

```{r goal2-ptc-mechanisms}
# ── Data provenance ──
# Source: computed inline in this report (not a separate script)
# Methods: METHODS.md § "PTC-Causing Event Attribution"
#
# For each PTC+ NMD comparator, identify the PTC-CAUSING splice event:
#   - Frameshift PTCs: from analyzeFrameWalk() (Isopair::analyzeFrameWalk,
#     cached as fw_c2_allsamples.rds by 03_nmd_analysis_mashr.Rmd)
#   - Non-frameshift PTCs: the event whose genomic coordinates (from
#     buildProfiles() detailed_events) contain the strand-aware PTC position
#     (from Isopair::extractCds → cds.rds)
#
# Strand-aware PTC position:
#   + strand: PTC = cds_stop (larger genomic coordinate = 3' end of CDS)
#   - strand: PTC = cds_start (smaller genomic coordinate = 3' end on minus strand)
# For PTC isoforms, SQANTI annotates the actual (premature) stop, so cds_start/
# cds_stop IS the PTC position (depending on strand).

fc_c2 <- set_results[["c2_allsamples"]]$frame_comp
fw_c2 <- set_results[["c2_allsamples"]]$frame_walk
prof_c2 <- set_results[["c2_allsamples"]]$profiles

fc_c2_ps <- fc_c2$pair_summary
fc_c2_ps$has_ptc <- fc_c2_ps$comparator_isoform_id %in% ptc_pos_ids

frameshift_cats <- c("same_start_frameshift",
                     "diff_start_diff_frame_with_frameshift")
fc_c2_ps$is_frameshift_cat <- fc_c2_ps$frame_category %in% frameshift_cats

# PTC+ pairs by mechanism — restrict to coding pairs (both ref and comp coding)
# to match the coding_pairs_c2 base used in Goals 3-4 and the cumulative flowchart
coding_ids_g2 <- cds$isoform_id[cds$coding_status == "coding"]
ptc_plus <- fc_c2_ps[fc_c2_ps$has_ptc == TRUE &
  fc_c2_ps$comparator_isoform_id %in% coding_ids_g2 &
  fc_c2_ps$reference_isoform_id %in% coding_ids_g2, ]
n_ptc_via_fs    <- sum(ptc_plus$is_frameshift_cat)
n_ptc_via_nonfs <- sum(!ptc_plus$is_frameshift_cat)
n_ptc_total     <- nrow(ptc_plus)

# Also count all coding NMD pairs for context (same coding filter as ptc_plus)
fc_c2_coding <- fc_c2_ps[fc_c2_ps$comparator_isoform_id %in% coding_ids_g2 &
  fc_c2_ps$reference_isoform_id %in% coding_ids_g2, ]
n_coding_nmd <- nrow(fc_c2_coding)

# ── Strand-aware PTC position lookups ──
# On + strand the stop codon is at cds_stop (larger coordinate).
# On - strand the stop codon is at cds_start (smaller coordinate).
cds_start_lookup <- setNames(cds$cds_start, cds$isoform_id)
cds_stop_lookup  <- setNames(cds$cds_stop, cds$isoform_id)
strand_lookup    <- setNames(cds$strand, cds$isoform_id)

# Build lookup: comparator_id → row index in prof_c2
comp_to_idx <- setNames(seq_len(nrow(prof_c2)), prof_c2$comparator_isoform_id)

# Frameshift PTC-causing events (from frame_walk)
fw_fs <- fw_c2$events[fw_c2$events$is_frameshift == TRUE, ]

ptc_event_rows <- list()
for (i in seq_len(nrow(ptc_plus))) {
  comp_id <- ptc_plus$comparator_isoform_id[i]
  is_fs <- ptc_plus$is_frameshift_cat[i]

  if (is_fs) {
    # Frameshift PTC: use the frameshift-causing event from frame_walk
    fs_match <- fw_fs[fw_fs$comparator_isoform_id == comp_id, ]
    if (nrow(fs_match) > 0) {
      ptc_event_rows[[length(ptc_event_rows) + 1]] <- data.frame(
        comparator_id = comp_id,
        mechanism = "Frameshift",
        ptc_causing_event = fs_match$event_type[1],
        attribution = "direct",
        stringsAsFactors = FALSE)
    } else {
      # Frameshift category but no is_frameshift event in frame_walk.
      # This indicates a disagreement between compareIsoformFrames() and
      # analyzeFrameWalk() — do not guess; label as unresolved.
      ptc_event_rows[[length(ptc_event_rows) + 1]] <- data.frame(
        comparator_id = comp_id,
        mechanism = "Frameshift",
        ptc_causing_event = "Unresolved",
        attribution = "unresolved",
        stringsAsFactors = FALSE)
    }
  } else {
    # Non-frameshift PTC: find the event containing the strand-aware PTC position
    prof_idx <- comp_to_idx[comp_id]
    if (is.na(prof_idx)) next
    de <- prof_c2$detailed_events[[prof_idx]]
    if (is.null(de) || nrow(de) == 0) next

    # Strand-aware PTC genomic position
    comp_strand <- strand_lookup[comp_id]
    ptc_pos_genomic <- if (!is.na(comp_strand) && comp_strand == "-") {
      cds_start_lookup[comp_id]
    } else {
      cds_stop_lookup[comp_id]
    }
    if (is.na(ptc_pos_genomic)) next

    # Check which event's coordinates span the PTC position
    de$ev_min <- pmin(de$five_prime, de$three_prime)
    de$ev_max <- pmax(de$five_prime, de$three_prime)
    containing <- de[ptc_pos_genomic >= de$ev_min & ptc_pos_genomic <= de$ev_max, ]

    if (nrow(containing) > 0) {
      ptc_event_rows[[length(ptc_event_rows) + 1]] <- data.frame(
        comparator_id = comp_id,
        mechanism = "In-frame stop",
        ptc_causing_event = containing$event_type[1],
        attribution = "direct",
        stringsAsFactors = FALSE)
    } else {
      ptc_event_rows[[length(ptc_event_rows) + 1]] <- data.frame(
        comparator_id = comp_id,
        mechanism = "In-frame stop",
        ptc_causing_event = "Unresolved",
        attribution = "unresolved",
        stringsAsFactors = FALSE)
    }
  }
}

ptc_events_df <- do.call(rbind, ptc_event_rows)
n_attributed <- sum(ptc_events_df$attribution == "direct")
n_unresolved <- sum(ptc_events_df$attribution == "unresolved")
```

Of `r fmt(n_coding_nmd)` coding NMD pairs, `r fmt(n_ptc_total)` have a PTC in
the comparator. These arise through two mechanisms:

- **Frameshift → PTC**: `r fmt(n_ptc_via_fs)` pairs
  (`r round(100 * n_ptc_via_fs / n_ptc_total, 1)`%) — a splicing event shifts
  the reading frame, and the new frame encounters a premature stop codon. The
  PTC-causing event is identified by `analyzeFrameWalk()`.
- **In-frame stop → PTC**: `r fmt(n_ptc_via_nonfs)` pairs
  (`r round(100 * n_ptc_via_nonfs / n_ptc_total, 1)`%) — the comparator is
  in the same reading frame but a structural change (e.g., a gained exon or
  retained intron) directly introduces an in-frame stop codon. The PTC-causing
  event is identified by finding the event whose genomic coordinates contain the
  comparator's stop codon position.

### PTC position and strand awareness

The PTC genomic position depends on strand: for + strand genes, the stop codon
is at `cds_stop` (the larger genomic coordinate); for - strand genes, it is at
`cds_start` (the smaller genomic coordinate, which is the 3' end in
transcription direction). This mirrors the logic in `Isopair::computePtcStatus`.

### Attribution results

Of `r fmt(nrow(ptc_events_df))` PTC+ pairs, `r fmt(n_attributed)` were
directly attributed (PTC position inside event coordinates or frameshift event
from `analyzeFrameWalk`)`r if (n_unresolved > 0) paste0(", and ", fmt(n_unresolved), " remain unresolved")``r if (n_unresolved == 0) ""`.
`r if (n_unresolved > 0) "Unresolved cases include both in-frame stop pairs where the PTC falls outside all event coordinates and frameshift-category pairs where analyzeFrameWalk() found no frameshift event (indicating a classification disagreement between compareIsoformFrames() and analyzeFrameWalk())."`

```{r goal2-ptc-diagnostic, eval = n_unresolved > 0}
# ── Diagnostic: characterize all unresolved PTC attributions ──
# Two types of unresolved:
#   1. In-frame stop: PTC position outside all event coordinates
#   2. Frameshift: frame_category says frameshift but analyzeFrameWalk() found
#      no is_frameshift event (classification disagreement)
# For each, report mechanism, strand, frame_category, and distance to nearest event.

unresolved <- ptc_events_df[ptc_events_df$attribution == "unresolved", ]

diag_rows <- list()
for (i in seq_len(nrow(unresolved))) {
  uid <- unresolved$comparator_id[i]
  mech <- unresolved$mechanism[i]

  prof_idx <- comp_to_idx[uid]
  fc_row <- fc_c2_ps[fc_c2_ps$comparator_isoform_id == uid, ]
  gene_id <- if (nrow(fc_row) > 0) fc_row$gene_id[1] else NA_character_
  frame_cat <- if (nrow(fc_row) > 0) fc_row$frame_category[1] else NA_character_
  comp_strand <- strand_lookup[uid]

  # PTC position and distance to nearest event (if events exist)
  ptc_pos <- if (!is.na(comp_strand) && comp_strand == "-") {
    cds_start_lookup[uid]
  } else {
    cds_stop_lookup[uid]
  }

  nearest_evt <- NA_character_; min_dist <- NA_real_; n_ev <- 0L
  if (!is.na(prof_idx)) {
    de <- prof_c2$detailed_events[[prof_idx]]
    if (!is.null(de) && nrow(de) > 0) {
      n_ev <- nrow(de)
      if (!is.na(ptc_pos)) {
        de$ev_min <- pmin(de$five_prime, de$three_prime)
        de$ev_max <- pmax(de$five_prime, de$three_prime)
        nearest_idx <- which.min(pmin(abs(ptc_pos - de$ev_min),
                                      abs(ptc_pos - de$ev_max)))
        min_dist <- min(abs(ptc_pos - c(de$ev_min, de$ev_max)))
        nearest_evt <- de$event_type[nearest_idx]
      }
    }
  }

  diag_rows[[length(diag_rows) + 1]] <- data.frame(
    comparator_id = uid, gene_id = gene_id, mechanism = mech,
    strand = comp_strand, frame_category = frame_cat,
    ptc_pos = if (is.na(ptc_pos)) NA_real_ else ptc_pos,
    nearest_event_type = nearest_evt,
    dist_to_nearest_boundary = min_dist,
    n_events = n_ev,
    stringsAsFactors = FALSE)
}
diag_df <- do.call(rbind, diag_rows)
diag_df <- diag_df[order(diag_df$mechanism, diag_df$dist_to_nearest_boundary), ]

render_dt(diag_df,
  caption = paste0("Diagnostic: ", nrow(diag_df),
    " Unresolved PTC Attributions"))
```

`r if (n_unresolved > 0) paste0("The diagnostic table shows ", fmt(n_unresolved), " unresolved cases. **Mechanism** distinguishes the two types: *In-frame stop* pairs have a PTC outside all event coordinates (dist_to_nearest_boundary shows how far); *Frameshift* pairs have a frameshift frame_category but analyzeFrameWalk() found no is_frameshift event, indicating a classification disagreement between the two Isopair functions.")`

```{r goal2-ptc-fs-test}
# ── Data provenance ──
# n_ptc_via_fs, n_ptc_total: from goal2-ptc-mechanisms chunk
# Baseline: frameshift rate among ALL coding NMD pairs (not just PTC+)
# fc_c2_ps$is_frameshift_cat: TRUE for frameshift frame categories
# Binomial test: is the frameshift proportion among PTC+ pairs different
#   from the frameshift proportion among all coding NMD pairs?

n_fs_all_nmd <- sum(fc_c2_coding$is_frameshift_cat)
baseline_fs_rate <- n_fs_all_nmd / n_coding_nmd

bt_fs <- binom.test(n_ptc_via_fs, n_ptc_total, p = baseline_fs_rate)
```

Frameshifts account for **`r fmt(n_ptc_via_fs)`** of **`r fmt(n_ptc_total)`**
PTC events in NMD pairs (**`r round(100 * n_ptc_via_fs / n_ptc_total, 1)`%**),
with in-frame stops accounting for the remaining
**`r round(100 * n_ptc_via_nonfs / n_ptc_total, 1)`%** — the two mechanisms
contribute roughly equally. However, frameshifts are significantly enriched
among PTC+ pairs relative to the overall frameshift rate among all coding NMD
pairs (`r round(100 * baseline_fs_rate, 1)`%; binomial test p =
`r signif(bt_fs$p.value, 3)`).

## Figure 6: PTC-Causing Splice Events (Sankey)

```{r goal2-fig5-sankey}
# ── Data provenance ──
# Combines three sources of PTC attribution:
#   1. Diff-stop frameshift: ptc_events_df (from analyzeFrameWalk)
#   2. Diff-stop in-frame: ptc_events_df (from coordinate containment)
#   3. Same-stop (3'UTR splice): attributed here using nearest-to-stop 3'UTR event
# Same-stop attribution: for each same-stop pair, find splice events downstream
# of the stop codon and attribute to the nearest one (the junction most likely
# to place the critical EJC within the >50 nt window).

# ── Identify same-stop PTC+ pairs (strand-aware) ──
# This duplicates the strand-aware stop comparison from goal3-fig6-pop,
# which has not yet run. Uses the same logic: + strand → cds_stop, - strand → cds_start.
ptc_plus_sankey <- ptc_plus  # from goal2-ptc-mechanisms
ptc_plus_sankey$strand_s <- strand_lookup[ptc_plus_sankey$comparator_isoform_id]
ptc_plus_sankey$comp_stop_s <- ifelse(ptc_plus_sankey$strand_s == "-",
  cds_start_lookup[ptc_plus_sankey$comparator_isoform_id],
  cds_stop_lookup[ptc_plus_sankey$comparator_isoform_id])
ptc_plus_sankey$ref_stop_s <- ifelse(ptc_plus_sankey$strand_s == "-",
  cds_start_lookup[ptc_plus_sankey$reference_isoform_id],
  cds_stop_lookup[ptc_plus_sankey$reference_isoform_id])
ptc_plus_sankey$same_stop_s <- ptc_plus_sankey$comp_stop_s == ptc_plus_sankey$ref_stop_s

same_stop_sankey <- ptc_plus_sankey[which(ptc_plus_sankey$same_stop_s == TRUE), ]

# ── Attribute same-stop pairs to nearest 3'UTR event ──
same_stop_attr <- list()
for (i in seq_len(nrow(same_stop_sankey))) {
  comp_id <- same_stop_sankey$comparator_isoform_id[i]
  stop_pos <- same_stop_sankey$comp_stop_s[i]
  comp_strand <- same_stop_sankey$strand_s[i]

  prof_idx <- comp_to_idx[comp_id]
  if (is.na(prof_idx)) next
  de <- prof_c2$detailed_events[[prof_idx]]
  if (is.null(de) || nrow(de) == 0) next

  # Find events downstream of stop in transcription direction
  if (comp_strand == "+") {
    utr3_ev <- de[de$five_prime >= stop_pos | de$three_prime >= stop_pos, ]
    if (nrow(utr3_ev) == 0) next
    ev_starts <- pmin(utr3_ev$five_prime, utr3_ev$three_prime)
    nearest_idx <- which.min(ev_starts - stop_pos)
  } else {
    utr3_ev <- de[de$five_prime <= stop_pos | de$three_prime <= stop_pos, ]
    if (nrow(utr3_ev) == 0) next
    ev_ends <- pmax(utr3_ev$five_prime, utr3_ev$three_prime)
    nearest_idx <- which.min(stop_pos - ev_ends)
  }

  same_stop_attr[[length(same_stop_attr) + 1]] <- data.frame(
    comparator_id = comp_id,
    mechanism = "3'UTR splice",
    ptc_causing_event = utr3_ev$event_type[nearest_idx],
    attribution = "direct",
    stringsAsFactors = FALSE)
}
same_stop_events <- do.call(rbind, same_stop_attr)
n_same_stop_attributed <- nrow(same_stop_events)

# ── Combine all attributed pairs ──
# Diff-stop attributed (from goal2-ptc-mechanisms), excluding same-stop pairs.
# Same-stop pairs may have been "directly" attributed in goal2 if the stop
# position fell inside an event by coincidence — these belong in 3'UTR splice.
# Pairs with unknown ref stop (coding_status="unknown") are kept in diff-stop.
same_stop_comp_ids <- same_stop_sankey$comparator_isoform_id
diffstop_attr <- ptc_events_df[ptc_events_df$attribution == "direct" &
  !ptc_events_df$comparator_id %in% same_stop_comp_ids, ]
# Same-stop attributed (from above)
all_attr <- rbind(diffstop_attr, same_stop_events)

# Shorten event type labels
all_attr$event_short <- all_attr$ptc_causing_event
all_attr$event_short[all_attr$event_short == "Missing_Internal"] <- "Missing Int."
all_attr$event_short[all_attr$event_short == "Partial_IR_5"] <- "Partial IR 5'"
all_attr$event_short[all_attr$event_short == "Partial_IR_3"] <- "Partial IR 3'"
all_attr$event_short[all_attr$event_short == "Alt_TSS"] <- "Alt TSS"
all_attr$event_short[all_attr$event_short == "Alt_TES"] <- "Alt TES"
all_attr$event_short[all_attr$event_short == "IR_diff"] <- "IR diff"
all_attr$event_short[all_attr$event_short == "IR_diff_3"] <- "IR diff 3'"
all_attr$event_short[all_attr$event_short == "IR_diff_5"] <- "IR diff 5'"

# Collapse rare event types (<3 occurrences) into "Other"
evt_counts <- table(all_attr$event_short)
rare_events <- names(evt_counts[evt_counts < 3])
all_attr$event_short[all_attr$event_short %in% rare_events] <- "Other"

# Also make ptc_attr available for Table 5a/5b (diff-stop attributed only)
ptc_attr <- all_attr[all_attr$mechanism != "3'UTR splice", ]

# Aggregate: event_type → mechanism
sankey_agg <- all_attr |>
  group_by(event_short, mechanism) |>
  summarise(freq = n(), .groups = "drop")

# Build networkD3 data structures
nodes_vec <- unique(c(sankey_agg$event_short, sankey_agg$mechanism))
nodes <- data.frame(name = nodes_vec, stringsAsFactors = FALSE)

links <- data.frame(
  source = match(sankey_agg$event_short, nodes_vec) - 1L,
  target = match(sankey_agg$mechanism, nodes_vec) - 1L,
  value  = sankey_agg$freq
)

# Node colors: mechanism nodes colored, event types grey
mech_colors <- c("Frameshift" = "#d6604d",
                 "In-frame stop" = "#4393c3",
                 "3'UTR splice" = "#66c2a5")
node_colors <- ifelse(nodes$name %in% names(mech_colors),
                      mech_colors[nodes$name], "#bdbdbd")

color_scale <- paste0(
  'd3.scaleOrdinal().domain([',
  paste0('"', nodes$name, '"', collapse = ','), ']).range([',
  paste0('"', node_colors, '"', collapse = ','), '])')

networkD3::sankeyNetwork(
  Links = links, Nodes = nodes,
  Source = "source", Target = "target", Value = "value",
  NodeID = "name", fontSize = 14, nodeWidth = 30, nodePadding = 15,
  colourScale = color_scale, sinksRight = TRUE)
```

This figure maps each PTC-causing splice event to one of three mechanisms:
**Frameshift** (red; `r fmt(sum(all_attr$mechanism == "Frameshift"))` pairs) —
a splice event shifts the reading frame, introducing a new stop codon;
**In-frame stop** (blue; `r fmt(sum(all_attr$mechanism == "In-frame stop"))`) —
a splice event introduces an in-frame stop codon directly;
**3'UTR splice** (green; `r fmt(n_same_stop_attributed)`) —
the comparator shares the reference's stop codon, but a downstream splice event
repositions an EJC, making the normal stop "premature." For 3'UTR splice pairs,
the attributed event is the nearest downstream event to the stop codon.
`r { n_same_stop_unresolved_in_goal2 <- sum(ptc_events_df$attribution == "unresolved" & ptc_events_df$comparator_id %in% same_stop_comp_ids); n_diffstop_unresolved <- n_unresolved - n_same_stop_unresolved_in_goal2; n_same_no_event <- nrow(same_stop_sankey) - n_same_stop_attributed; paste0(fmt(n_diffstop_unresolved), " unresolved diff-stop pairs and ", fmt(n_same_no_event), " same-stop pairs without identifiable 3'UTR events are excluded (", fmt(nrow(all_attr)), " of ", fmt(nrow(ptc_plus)), " PTC+ pairs shown).") }`

## Table 5a: PTC-Causing Events by Mechanism

```{r goal2-fig5-summary}
# Summary table of PTC-causing events by mechanism
# Denominator: all directly attributed pairs (excludes unresolved)
ptc_event_summary <- ptc_attr |>
  group_by(mechanism, ptc_causing_event) |>
  summarise(n = n(), .groups = "drop") |>
  arrange(mechanism, desc(n))
ptc_event_summary$pct <- round(100 * ptc_event_summary$n / nrow(ptc_attr), 1)

render_dt(ptc_event_summary, caption = "Table 5a: PTC-Causing Events by Mechanism (NMD)")
write.csv(ptc_event_summary, file.path(table_dir, "table5a_ptc_causing_events.csv"),
          row.names = FALSE)
```

Columns: **mechanism** = Frameshift or In-frame stop; **ptc_causing_event** =
splice event type that caused the PTC; **n** = number of pairs; **pct** =
percentage of all `r fmt(nrow(ptc_attr))` attributed PTCs.

## Table 5b: PTC-Causing Event Proportions vs Control Baseline

```{r goal2-table4b-baseline}
# ── Data provenance ──
# NMD: PTC-causing events attributed above (both frameshift and in-frame stop)
# Control baseline: all splicing events across all C4 pairs
# Comparison: proportion of each event type among PTC-causing NMD events vs
#   all Control events (the baseline frequency in normal alternative splicing)

# NMD PTC-causing event frequencies (attributed pairs: direct + largest_event)
n_ptc_attr <- nrow(ptc_attr)
ptc_evt_freq <- sort(table(ptc_attr$ptc_causing_event), decreasing = TRUE)
ptc_evt_df <- data.frame(
  event_type = names(ptc_evt_freq),
  n_ptc_events = as.integer(ptc_evt_freq),
  pct_of_ptc = round(100 * as.integer(ptc_evt_freq) / n_ptc_attr, 1),
  stringsAsFactors = FALSE)

# Control baseline: all events across all C4 pairs
ctrl_all_events <- do.call(rbind, set_results[["c4_allsamples"]]$profiles$detailed_events)
ctrl_event_freq <- table(ctrl_all_events$event_type)
ctrl_total <- sum(ctrl_event_freq)

# Merge
ptc_evt_df$n_ctrl_events <- as.integer(ctrl_event_freq[ptc_evt_df$event_type])
ptc_evt_df$n_ctrl_events[is.na(ptc_evt_df$n_ctrl_events)] <- 0L
ptc_evt_df$pct_ctrl <- round(100 * ptc_evt_df$n_ctrl_events / ctrl_total, 1)
ptc_evt_df$enrichment <- round(ptc_evt_df$pct_of_ptc / pmax(ptc_evt_df$pct_ctrl, 0.1), 1)

# Fisher's exact test per event type: is this event over-represented among
# PTC-causing events vs all Control events?
ptc_evt_df$fisher_p <- NA_real_
ptc_evt_df$direction <- NA_character_
for (i in seq_len(nrow(ptc_evt_df))) {
  a <- ptc_evt_df$n_ptc_events[i]      # this event in PTC-causing
  b <- n_ptc_attr - a                    # other events in PTC-causing
  c_val <- ptc_evt_df$n_ctrl_events[i]  # this event in Control
  d_val <- ctrl_total - c_val            # other events in Control
  ft <- tryCatch(fisher.test(matrix(c(a, b, c_val, d_val), nrow = 2)),
                 error = function(e) NULL)
  if (!is.null(ft)) {
    ptc_evt_df$fisher_p[i] <- signif(ft$p.value, 3)
    ptc_evt_df$direction[i] <- if (ft$p.value > 0.05) "Non-significant"
      else if (ptc_evt_df$enrichment[i] > 1) "Enriched in PTC-causing"
      else "Depleted in PTC-causing"
  }
}

render_dt(ptc_evt_df, caption = "Table 5b: PTC-Causing Events (NMD) vs All Events (Control)")
write.csv(ptc_evt_df, file.path(table_dir, "table5b_ptc_event_enrichment.csv"),
          row.names = FALSE)
```

Columns: **event_type** = splice event type; **n_ptc_events** = count of
PTC-causing events of this type; **pct_of_ptc** = % of all
`r fmt(n_ptc_attr)` attributed PTC-causing events; **n_ctrl_events** = count in
gene-matched Control (C4) pairs; **pct_ctrl** = % of `r fmt(ctrl_total)` Control
events; **enrichment** = fold enrichment (% PTC / % Ctrl); **fisher_p** =
Fisher's exact test p-value (2×2: this event vs others × PTC-causing vs
Control); **direction** = Enriched/Depleted in PTC-causing, or Non-significant
(p > 0.05).


# 4. Non-PTC NMD Triggers {#goal4}

**Goal 4:** Characterize NMD-responsive comparator isoforms that lack a PTC.

**Methods:** Among coding NMD (C2) comparators, a substantial fraction lack a PTC (stop
codon >50 nt upstream of last EJC). We categorize these PTC-negative NMD isoforms by:
(a) borderline PTC position (stop codon with 1+ downstream EJCs but within 50 nt),
(b) 3'UTR exonic length, (c) frame category, and (d) NMD response magnitude.

```{r goal3-identify-ptc-neg}
# ── Data provenance ──
# Starting data: gene-matched C2 all_samples profiles → cached PTC association
# ptc_c2_as = set_results[["c2_allsamples"]]$ptc_assoc
#   Source: testPtcAssociation(ptc, gene-matched profiles), cached by script 03
#   $pair_summary: one row per gene-matched pair, with comp_has_ptc flag
# Step 1: Filter to coding pairs (both ref and comp in cds$coding_status == "coding")
# Step 2: Split into PTC+ and PTC- NMD comparators

ptc_c2_as <- set_results[["c2_allsamples"]]$ptc_assoc
pair_summary <- ptc_c2_as$pair_summary

# Restrict to coding pairs (same filter as Table 2)
coding_ids_g3 <- cds$isoform_id[cds$coding_status == "coding"]
coding_pairs_c2 <- pair_summary[pair_summary$comparator_isoform_id %in% coding_ids_g3 &
                                  pair_summary$reference_isoform_id %in% coding_ids_g3, ]

n_ptc_pos <- sum(coding_pairs_c2$comp_has_ptc == TRUE)
n_ptc_neg <- sum(coding_pairs_c2$comp_has_ptc == FALSE)
```

Of the `r fmt(nrow(set_results[["c2_allsamples"]]$profiles))` gene-matched NMD
pairs, `r fmt(nrow(coding_pairs_c2))` are coding pairs (both reference and
comparator classified as protein-coding). Among these,
`r fmt(n_ptc_pos)` (`r round(100 * n_ptc_pos / nrow(coding_pairs_c2), 1)`%)
have a PTC (comparator stop codon >50 nt upstream of last EJC) and
`r fmt(n_ptc_neg)` (`r round(100 * n_ptc_neg / nrow(coding_pairs_c2), 1)`%)
do not. The remainder of this section characterizes these
`r fmt(n_ptc_neg)` **PTC-negative NMD comparators**.

## Categorization by Stop Codon Position

```{r goal3-stop-position}
# ── Data provenance ──
# Starting data: ptc_neg_ids (PTC-negative NMD comparators identified above)
# ptc.rds: per-isoform annotations (n_downstream_ejcs, ptc_distance, etc.)
#   Source: Isopair testPtcAssociation() infrastructure — isoform-intrinsic
# Step 1: Look up ptc features for PTC-neg comparators
# Step 2: Categorize by n_downstream_ejcs (borderline vs stop-in-last-exon)

ptc_neg_ids <- coding_pairs_c2$comparator_isoform_id[coding_pairs_c2$comp_has_ptc == FALSE]
ptc_neg_info <- ptc[ptc$isoform_id %in% ptc_neg_ids, ]

# Categorize
n_borderline <- sum(ptc_neg_info$n_downstream_ejcs > 0, na.rm = TRUE)
n_stop_last_exon <- sum(ptc_neg_info$n_downstream_ejcs == 0, na.rm = TRUE)
n_na <- sum(is.na(ptc_neg_info$n_downstream_ejcs))

stop_cat <- data.frame(
  category = c("Borderline (1+ downstream EJCs, distance <= 50 nt)",
               "Stop in last exon (0 downstream EJCs)",
               "Missing annotation"),
  n = c(n_borderline, n_stop_last_exon, n_na),
  pct = round(100 * c(n_borderline, n_stop_last_exon, n_na) / length(ptc_neg_ids), 1))

kable(stop_cat, caption = "Table 6: Categorization of PTC-Negative NMD Comparators")
```

## Figure 7: 3'UTR Length by PTC Mechanism

```{r goal3-fig6-pop}
# ── Data provenance ──
# Source: computed inline in this report using cached divergence + cds.rds
# Methods: METHODS.md § "3'UTR Length Analysis with PTC Correction"
# Key functions:
#   Isopair::quantifyPairDivergence() → div_c2/div_c4 (utr3_bp_comp column)
#   Isopair::.computeRegionalBp() → computes 3'UTR as exonic bp after cds_stop
#   Reference-based 3'UTR: computed inline using ref cds_stop on comp exon structure
# Starting data: gene-matched C2 and C4 profiles → cached divergence
# PTC mechanism classification: "same stop" vs "diff stop"
#   same_stop: comparator and reference share the same cds_stop position.
#     The PTC arises because a 3'UTR splicing event repositioned the last EJC
#     downstream, making the existing stop codon >50 nt upstream of the new EJC.
#     The 3'UTR length measurement (from cds_stop to tx end) is correct and
#     unaffected by the PTC classification.
#   diff_stop: comparator has a different cds_stop (= premature stop from
#     frameshift or in-frame stop in the CDS). The measured 3'UTR is inflated
#     because it includes the former CDS region between PTC and normal stop.
#     For these, we also compute a "reference-based 3'UTR": exonic bp downstream
#     of the REFERENCE's cds_stop in the COMPARATOR's exon structure. This is
#     only valid when the reference stop position falls within the comparator's
#     exons (86% of cases).

div_c2 <- set_results[["c2_allsamples"]]$divergence
div_c4 <- set_results[["c4_allsamples"]]$divergence
ptc_pos_comp_ids <- coding_pairs_c2$comparator_isoform_id[coding_pairs_c2$comp_has_ptc == TRUE]

# Classify PTC+ by same vs different stop codon position
# Strand-aware: + strand stop = cds_stop; - strand stop = cds_start
cds_start_lookup_g3 <- setNames(cds$cds_start, cds$isoform_id)
cds_stop_lookup_g3  <- setNames(cds$cds_stop, cds$isoform_id)
strand_lookup_g3    <- setNames(cds$strand, cds$isoform_id)

ptc_pos_ps <- coding_pairs_c2[coding_pairs_c2$comp_has_ptc == TRUE, ]
ptc_pos_ps$strand <- strand_lookup_g3[ptc_pos_ps$comparator_isoform_id]
ptc_pos_ps$comp_stop <- ifelse(ptc_pos_ps$strand == "-",
  cds_start_lookup_g3[ptc_pos_ps$comparator_isoform_id],
  cds_stop_lookup_g3[ptc_pos_ps$comparator_isoform_id])
ptc_pos_ps$ref_stop <- ifelse(ptc_pos_ps$strand == "-",
  cds_start_lookup_g3[ptc_pos_ps$reference_isoform_id],
  cds_stop_lookup_g3[ptc_pos_ps$reference_isoform_id])
ptc_pos_ps$same_stop <- ptc_pos_ps$comp_stop == ptc_pos_ps$ref_stop

same_stop_ids <- ptc_pos_ps$comparator_isoform_id[ptc_pos_ps$same_stop == TRUE]
diff_stop_ids <- ptc_pos_ps$comparator_isoform_id[ptc_pos_ps$same_stop == FALSE]

# For diff-stop pairs, compute reference-based 3'UTR
diff_stop_ps <- ptc_pos_ps[ptc_pos_ps$same_stop == FALSE, ]
ref_based_utr3 <- setNames(rep(NA_integer_, nrow(diff_stop_ps)), diff_stop_ps$comparator_isoform_id)

for (i in seq_len(nrow(diff_stop_ps))) {
  comp_id <- diff_stop_ps$comparator_isoform_id[i]
  ref_stop_pos <- diff_stop_ps$ref_stop[i]
  comp_strand <- strand_lookup_g3[comp_id]
  if (is.na(comp_strand) || is.na(ref_stop_pos)) next

  str_row <- structures[structures$isoform_id == comp_id, ]
  if (nrow(str_row) == 0) next
  exon_s <- str_row$exon_starts[[1]]
  exon_e <- str_row$exon_ends[[1]]

  # Check ref stop is in comparator exons
  in_exon <- any(ref_stop_pos >= exon_s & ref_stop_pos <= exon_e)
  if (!in_exon) next

  utr3 <- 0L
  for (k in seq_along(exon_s)) {
    if (comp_strand == "+") {
      if (exon_e[k] > ref_stop_pos) {
        post_start <- max(exon_s[k], ref_stop_pos + 1L)
        utr3 <- utr3 + as.integer(exon_e[k] - post_start + 1L)
      }
    } else {
      if (exon_s[k] < ref_stop_pos) {
        pre_end <- min(exon_e[k], ref_stop_pos - 1L)
        utr3 <- utr3 + as.integer(pre_end - exon_s[k] + 1L)
      }
    }
  }
  ref_based_utr3[comp_id] <- utr3
}

n_eligible <- sum(!is.na(ref_based_utr3))
n_ineligible <- sum(is.na(ref_based_utr3))

# Lookup for comparator 3'UTR from divergence
utr3_comp_lk <- setNames(div_c2$utr3_bp_comp, div_c2$comparator_isoform_id)

# Build 5-group data
utr3_data <- rbind(
  data.frame(group = "Same-stop PTC+\n(3'UTR splicing)",
    utr3_bp = utr3_comp_lk[same_stop_ids]),
  data.frame(group = "Diff-stop PTC+\n(own stop, inflated)",
    utr3_bp = utr3_comp_lk[diff_stop_ids]),
  data.frame(group = "Diff-stop PTC+\n(ref stop, corrected)",
    utr3_bp = ref_based_utr3[!is.na(ref_based_utr3)]),
  data.frame(group = "NMD PTC-",
    utr3_bp = utr3_comp_lk[ptc_neg_ids]),
  data.frame(group = "Control",
    utr3_bp = div_c4$utr3_bp_comp)
)
utr3_data <- utr3_data[!is.na(utr3_data$utr3_bp) & utr3_data$utr3_bp > 0, ]
utr3_data$group <- factor(utr3_data$group,
  levels = c("Same-stop PTC+\n(3'UTR splicing)",
             "Diff-stop PTC+\n(own stop, inflated)",
             "Diff-stop PTC+\n(ref stop, corrected)",
             "NMD PTC-", "Control"))
utr3_n <- table(utr3_data$group)
```

PTC+ comparators are split by whether the strand-aware stop codon position is
the **same** as the reference or **different** (premature stop from frameshift
or in-frame stop in CDS). Stop codon position: `cds_stop` on + strand,
`cds_start` on - strand.

- **Same-stop PTC+** (`r fmt(utr3_n[[1]])`): The stop codon is the same as the
  reference. A 3'UTR splicing event repositioned the last EJC, making the
  existing stop >50 nt upstream of the new EJC. The 3'UTR length is correctly
  measured.
- **Diff-stop PTC+ (own stop)** (`r fmt(utr3_n[[2]])`): The stop codon moved
  upstream (PTC). The measured 3'UTR is inflated — it includes former CDS.
- **Diff-stop PTC+ (ref stop)** (`r fmt(utr3_n[[3]])`): Same pairs as above, but
  3'UTR is measured from the **reference's** stop codon position projected onto
  the comparator's exon structure. `r fmt(n_eligible)` of `r fmt(n_eligible + n_ineligible)`
  pairs were eligible (reference stop falls within comparator exons).
- **NMD PTC−** (`r fmt(utr3_n[[4]])`): No PTC. 3'UTR is correctly measured.
  *Example: GPAT3 — comp 3'UTR=1,108 bp, ref 3'UTR=1,108 bp.*
- **Control** (`r fmt(utr3_n[[5]])`): C4 comparators.
  *Example: SART1 — comp 3'UTR=1,107 bp, ref 3'UTR=1,107 bp.*

```{r goal3-fig6-utr3-density, fig.width=10, fig.height=6}
pal5_utr3 <- c("Same-stop PTC+\n(3'UTR splicing)" = "#d6604d",
               "Diff-stop PTC+\n(own stop, inflated)" = "#fc8d59",
               "Diff-stop PTC+\n(ref stop, corrected)" = "#fee08b",
               "NMD PTC-" = "#d9ef8b",
               "Control" = "#67a9cf")

fig6 <- ggplot(utr3_data, aes(x = utr3_bp, fill = group)) +
  geom_density(alpha = 0.5) +
  scale_x_log10(labels = scales::comma) +
  scale_fill_manual(values = pal5_utr3) +
  labs(title = "3'UTR Length by PTC Mechanism",
       subtitle = "Correcting for CDS truncation reveals PTC-neg NMD isoforms have normal 3'UTR lengths",
       x = "3'UTR length (bp, log scale)", y = "Density", fill = "")
save_fig(fig6, "fig7_utr3_length_density", width = 10, height = 6)
```

## Table 7: 3'UTR Length Comparisons (Wilcoxon)

```{r goal3-table6-utr3-tests}
grps <- split(utr3_data$utr3_bp, utr3_data$group)

comparisons_list <- list(
  c("Same-stop PTC+\n(3'UTR splicing)", "Control"),
  c("Diff-stop PTC+\n(own stop, inflated)", "Control"),
  c("Diff-stop PTC+\n(ref stop, corrected)", "Control"),
  c("NMD PTC-", "Control"),
  c("Same-stop PTC+\n(3'UTR splicing)", "Diff-stop PTC+\n(ref stop, corrected)")
)

tests <- do.call(rbind, lapply(comparisons_list, function(pair) {
  g1 <- grps[[pair[1]]]
  g2 <- grps[[pair[2]]]
  wt <- wilcox.test(g1, g2)
  U <- as.numeric(wt$statistic)
  rbc <- 1 - (2 * U) / (length(g1) * length(g2))
  data.frame(
    group1 = gsub("\n", " ", pair[1]),
    n_group1 = length(g1), median_group1 = round(median(g1)),
    group2 = gsub("\n", " ", pair[2]),
    n_group2 = length(g2), median_group2 = round(median(g2)),
    median_diff_bp = round(median(g1) - median(g2)),
    rank_biserial_r = round(rbc, 3),
    p_value = signif(wt$p.value, 3),
    stringsAsFactors = FALSE)
}))

render_dt(tests, caption = "Table 7: 3'UTR Length Comparisons (Wilcoxon rank-sum test)")
write.csv(tests, file.path(table_dir, "table7_utr3_wilcoxon.csv"), row.names = FALSE)
```

Columns: **group1/group2** = the two groups being compared (see Figure 7 for
group definitions); **n_group1/n_group2** = number of isoforms in each group;
**median_group1/median_group2** = median 3'UTR length (bp);
**median_diff_bp** = difference in medians (group1 − group2);
**rank_biserial_r** = rank-biserial correlation (effect size; positive =
group1 longer); **p_value** = Wilcoxon rank-sum test p-value.

## Illustrative Example: 3'UTR Measurement Correction (CMAS)

The figure below shows the CMAS gene pair (reference ENST00000229329.7 vs NMD
comparator ENST00000534981.5) to illustrate the difference between the two
3'UTR measurements. The comparator has a skipped exon that causes a frameshift,
creating a PTC at genomic position 22,062,284. The reference stop codon is at
22,065,311.

```{r goal3-utr3-example, fig.width=10, fig.height=10}
# CMAS: + strand, ref has 8 exons, comp has 7 (skipped exon 22061281-22061452)
# Same CDS start (22046304), different stop (ref: 22065311, comp: 22062284)

ref_s <- c(22046218, 22055149, 22055455, 22058567, 22060832, 22061281, 22062281, 22065121)
ref_e <- c(22046563, 22055291, 22055610, 22058700, 22060926, 22061452, 22062434, 22065668)
comp_s <- c(22046225, 22055149, 22055455, 22058567, 22060832, 22062281, 22065121)
comp_e <- c(22046563, 22055291, 22055610, 22058700, 22060926, 22062434, 22065674)

# Skipped exon event (comparator lost exon 22061281-22061452)
se_event <- data.frame(
  five_prime = 22061281, three_prime = 22061452,
  event_type = "SE", direction = "LOSS")

# Annotations: PTC marker on comparator, reference stop marker on both
ptc_marker <- list(type = "ptc_marker", track = "comp",
  x = 22062284, label = "PTC", color = "#E74C3C")
ref_stop_marker <- list(type = "ptc_marker", track = "ref",
  x = 22065311, label = "Stop", color = "#27AE60")

# Panel A: "Own stop" — comparator CDS ends at PTC (22062284)
p1 <- plot_ucsc_pair(
  ref_s, ref_e, comp_s, comp_e, strand = "+",
  ref_cds = list(cds_start = 22046304, cds_stop = 22065311),
  comp_cds = list(cds_start = 22046304, cds_stop = 22062284),
  ref_label = "Reference (ENST00000229329.7)",
  comp_label = "NMD Comparator (ENST00000534981.5)",
  events = se_event,
  annotations = list(ptc_marker, ref_stop_marker),
  title = "A) 3'UTR at own stop (inflated)",
  subtitle = "Comparator CDS ends at PTC. Measured 3'UTR (thin) includes former CDS downstream of PTC.")

# Panel B: "Ref stop" — comparator CDS extends to reference stop (22065311)
ref_stop_comp <- list(type = "ptc_marker", track = "comp",
  x = 22065311, label = "Ref Stop", color = "#27AE60")

p2 <- plot_ucsc_pair(
  ref_s, ref_e, comp_s, comp_e, strand = "+",
  ref_cds = list(cds_start = 22046304, cds_stop = 22065311),
  comp_cds = list(cds_start = 22046304, cds_stop = 22065311),
  ref_label = "Reference (ENST00000229329.7)",
  comp_label = "NMD Comparator (ENST00000534981.5)",
  events = se_event,
  annotations = list(ptc_marker, ref_stop_marker, ref_stop_comp),
  title = "B) 3'UTR at reference stop (corrected)",
  subtitle = "Comparator CDS extended to reference stop. 3'UTR = only true downstream sequence.")

gridExtra::grid.arrange(p1, p2, ncol = 1)
```

In panel A, the skipped exon (SE LOSS bracket) causes a frameshift that creates
a PTC (red triangle on comparator). The comparator's CDS (thick) ends at the
PTC, and everything downstream is measured as 3'UTR (thin) — inflating the
length. The reference stop codon (green triangle) is further downstream. In
panel B, the comparator's CDS boundary is set to the reference's stop position
(green triangle on both tracks), giving the "natural" 3'UTR length.

**Key finding (Figure 7, Table 7):** PTC-negative NMD comparators have 3'UTR lengths
indistinguishable from Control (p = `r signif(tests$p_value[tests$group1 == "NMD PTC-"], 2)`).
Long 3'UTR length does not appear to be an independent NMD trigger in this
dataset. The apparently long 3'UTRs in PTC+ isoforms are explained by two
mechanisms: (1) for same-stop PTC+ isoforms, the 3'UTR is genuinely extended by
the same splicing event that repositioned the EJC; (2) for diff-stop PTC+
isoforms, the 3'UTR measurement is inflated by inclusion of former CDS sequence
downstream of the premature stop.

## 3'UTR Splicing and NMD

The question is whether 3'UTR splicing contributes to NMD in our data,
and if so, through what mechanism — PTC creation (EJC repositioning) or a
PTC-independent pathway (e.g., long 3'UTR-mediated UPF1 recruitment).

```{r goal3-utr3-splicing}
# ── Data provenance ──
# Source: computed inline using count_utr3_splicing() defined in this chunk
# Methods: METHODS.md § "3'UTR Splicing Enrichment"
# Starting data: gene-matched C2 and C4 profiles (detailed_events from
#   Isopair::buildProfiles() → compareSplicingChoices) + cds.rds (Isopair::extractCds)

# Strand-aware stop codon position for 3'UTR event detection
# + strand: stop = cds_stop; - strand: stop = cds_start
cds_start_lookup_utr3 <- setNames(cds$cds_start, cds$isoform_id)
cds_stop_lookup_utr3  <- setNames(cds$cds_stop, cds$isoform_id)
strand_lookup <- setNames(cds$strand, cds$isoform_id)

count_utr3_splicing <- function(profiles, comp_id_subset = NULL) {
  if (!is.null(comp_id_subset)) {
    idx <- which(profiles$comparator_isoform_id %in% comp_id_subset)
  } else {
    idx <- seq_len(nrow(profiles))
  }
  n_with <- 0L
  n_total <- 0L
  for (i in idx) {
    comp_id <- profiles$comparator_isoform_id[i]
    de <- profiles$detailed_events[[i]]
    comp_strand <- strand_lookup[comp_id]
    comp_stop <- if (!is.na(comp_strand) && comp_strand == "-") {
      cds_start_lookup_utr3[comp_id]
    } else {
      cds_stop_lookup_utr3[comp_id]
    }
    if (is.na(comp_stop) || is.na(comp_strand)) next
    n_total <- n_total + 1L
    if (is.null(de) || nrow(de) == 0) next
    if (comp_strand == "+") {
      utr3_ev <- de[de$five_prime >= comp_stop | de$three_prime >= comp_stop, ]
    } else {
      utr3_ev <- de[de$five_prime <= comp_stop | de$three_prime <= comp_stop, ]
    }
    if (nrow(utr3_ev) > 0) n_with <- n_with + 1L
  }
  list(n_with = n_with, n_total = n_total,
       pct = round(100 * n_with / max(n_total, 1), 1))
}

prof_c4_as <- set_results[["c4_allsamples"]]$profiles

utr3_ptcneg <- count_utr3_splicing(prof_c2, ptc_neg_ids)
utr3_ptcpos <- count_utr3_splicing(prof_c2, ptc_pos_comp_ids)
utr3_ctrl   <- count_utr3_splicing(prof_c4_as)

ft_utr3_neg <- fisher.test(matrix(c(
  utr3_ptcneg$n_with, utr3_ptcneg$n_total - utr3_ptcneg$n_with,
  utr3_ctrl$n_with, utr3_ctrl$n_total - utr3_ctrl$n_with), nrow = 2))
ft_utr3_pos <- fisher.test(matrix(c(
  utr3_ptcpos$n_with, utr3_ptcpos$n_total - utr3_ptcpos$n_with,
  utr3_ctrl$n_with, utr3_ctrl$n_total - utr3_ctrl$n_with), nrow = 2))

utr3_summary <- data.frame(
  group = c("NMD PTC-", "NMD PTC+", "Control"),
  n_with_utr3_event = c(utr3_ptcneg$n_with, utr3_ptcpos$n_with, utr3_ctrl$n_with),
  n_total = c(utr3_ptcneg$n_total, utr3_ptcpos$n_total, utr3_ctrl$n_total),
  pct = c(utr3_ptcneg$pct, utr3_ptcpos$pct, utr3_ctrl$pct),
  vs_control_OR = c(round(as.numeric(ft_utr3_neg$estimate), 2),
                     round(as.numeric(ft_utr3_pos$estimate), 2), NA),
  vs_control_p = c(signif(ft_utr3_neg$p.value, 3),
                    signif(ft_utr3_pos$p.value, 3), NA))

kable(utr3_summary,
      caption = "3'UTR Splicing Prevalence by Group",
      col.names = c("Group", "With 3'UTR Event", "Total", "%",
                     "OR vs Control", "p vs Control"))
```

Columns: **Group**; **With 3'UTR Event** = pairs where ≥1 splice event has
coordinates in the 3'UTR (downstream of the strand-aware stop codon position);
**OR/p vs Control** = Fisher's exact test comparing to Control (C4).

```{r goal3-utr3-same-stop-verification}
# ── Direct verification: 3'UTR splicing creates PTCs via EJC repositioning ──
# For same-stop PTC+ pairs, the stop codon is identical to the reference.
# The PTC status arises because the comparator has more exon-exon junctions
# downstream of the stop codon (from 3'UTR splicing), repositioning the last EJC.
# We can verify by comparing n_downstream_ejcs between ref and comp.

same_stop_ps <- ptc_pos_ps[ptc_pos_ps$same_stop == TRUE, ]
ptc_ejc_lookup <- setNames(ptc$n_downstream_ejcs, ptc$isoform_id)

same_stop_ps$comp_ejcs <- ptc_ejc_lookup[same_stop_ps$comparator_isoform_id]
same_stop_ps$ref_ejcs <- ptc_ejc_lookup[same_stop_ps$reference_isoform_id]
same_stop_ps$ejc_diff <- same_stop_ps$comp_ejcs - same_stop_ps$ref_ejcs

n_comp_more_ejcs <- sum(same_stop_ps$ejc_diff > 0, na.rm = TRUE)
n_same_ejcs <- sum(same_stop_ps$ejc_diff == 0, na.rm = TRUE)
n_comp_fewer_ejcs <- sum(same_stop_ps$ejc_diff < 0, na.rm = TRUE)
n_valid <- sum(!is.na(same_stop_ps$ejc_diff))
```

### 3'UTR Splicing Creates PTCs via EJC Repositioning

Among the `r fmt(nrow(same_stop_ps))` same-stop PTC+ pairs (where the
comparator has the same stop codon as the reference but is classified as PTC+),
we directly verified the mechanism by comparing the number of exon-exon
junctions downstream of the stop codon:

- **`r n_comp_more_ejcs`** pairs (`r round(100 * n_comp_more_ejcs / n_valid, 1)`%):
  comparator has MORE downstream EJCs than reference — confirming that 3'UTR
  splicing added new junctions, repositioning the last EJC and creating the PTC
- **`r n_same_ejcs`** pairs: same number of downstream EJCs (PTC may arise from
  junction repositioning rather than addition)
- **`r n_comp_fewer_ejcs`** pairs: comparator has fewer EJCs (unexpected —
  may reflect annotation differences)

### Conclusions: 3'UTR Mechanisms in NMD

Three lines of evidence address whether 3'UTR length or splicing triggers
PTC-independent NMD:

1. **No NMD subgroup has substantially longer 3'UTRs than Control when measured
   from the appropriate stop codon** (Table 7). PTC-negative NMD comparators
   have the same median 3'UTR as Control (p =
   `r signif(tests$p_value[tests$group1 == "NMD PTC-"], 2)`). Same-stop PTC+
   comparators are also indistinguishable from Control (p =
   `r signif(tests$p_value[tests$group1 == "Same-stop PTC+ (3'UTR splicing)" & tests$group2 == "Control"], 2)`).
   The diff-stop PTC+ group measured at own stop has longer 3'UTRs (p < 10^-54^),
   but this is an artifact of measuring from the PTC rather than the true stop.
   When corrected using the reference stop position, the effect is small (median
   difference `r tests$median_diff_bp[tests$group1 == "Diff-stop PTC+ (ref stop, corrected)" & tests$group2 == "Control"]`
   bp, r = `r tests$rank_biserial_r[tests$group1 == "Diff-stop PTC+ (ref stop, corrected)" & tests$group2 == "Control"]`,
   p = `r signif(tests$p_value[tests$group1 == "Diff-stop PTC+ (ref stop, corrected)" & tests$group2 == "Control"], 2)`).

2. **3'UTR splicing that triggers NMD does so by creating PTCs** (EJC
   repositioning), not through a PTC-independent pathway. The
   `r fmt(nrow(same_stop_ps))` same-stop PTC+ pairs are direct evidence:
   `r round(100 * n_comp_more_ejcs / n_valid, 1)`% have additional downstream
   EJCs in the comparator from 3'UTR splicing events.

3. **3'UTR splicing prevalence** is actually **depleted** in PTC-negative NMD
   vs Control (OR =
   `r round(as.numeric(ft_utr3_neg$estimate), 2)`, p =
   `r signif(ft_utr3_neg$p.value, 3)`). This is consistent with our finding
   that 3'UTR splicing creates PTCs — isoforms that trigger NMD without a PTC
   would be expected to have *less* 3'UTR splicing, not more.

## 5'UTR uORF Analysis

Upstream open reading frames (uORFs) in the 5'UTR are a known mechanism that
can trigger NMD independently of PTCs in the main CDS. We scanned the 5'UTR
sequence of each comparator isoform for uORFs (AUG followed by an in-frame stop
codon, minimum 3 codons / 9 nt). uORFs whose stop codon falls at or beyond the
main CDS start are classified as **overlapping** — a particularly potent NMD
trigger.

```{r goal3-uorf}
# ── Data provenance ──
# Source script: 05b_uorf_analysis.R → analysis_cache/uorf_analysis.rds
# Methods: METHODS.md § "5'UTR uORF Analysis"
#
# Pipeline:
#   1. 5'UTR boundaries computed by compute_utr5_length() in 05b_uorf_analysis.R
#      Input: genomic CDS coordinates (cds.rds) + exon structure (structures.rds)
#      + strand: CDS start = cds_start (plus) or cds_stop (minus)
#      Validation: 98.4% of isoforms have ATG at computed CDS start position
#   2. 5'UTR sequences extracted from SQANTI corrected transcript FASTA
#      (nmd_lungcells_corrected.fasta) via awk-based subset extraction
#   3. uORF detection by find_uorfs() in 05b_uorf_analysis.R:
#      gregexpr("ATG") finds all AUG positions in 5'UTR; for each, walks in-frame
#      through full transcript to first stop codon (TAA/TAG/TGA). Min 3 codons.
#      Overlapping uORFs: stop codon at or past main CDS start.
#   4. Groups: PTC_neg, PTC_pos, Control (same coding pair definitions as Goal 3)

uorf_data <- readRDS(file.path(cache_dir, "uorf_analysis.rds"))
uorf <- uorf_data$per_isoform
uorf$group <- factor(uorf$group, levels = c("PTC_neg", "PTC_pos", "Control"))
```

```{r goal3-uorf-summary}
uorf_summary <- uorf |>
  group_by(group) |>
  summarise(
    n = n(),
    median_utr5_bp = median(utr5_length),
    pct_with_uorf = round(100 * mean(n_uorfs > 0), 1),
    median_n_uorfs = median(n_uorfs),
    mean_n_uorfs = round(mean(n_uorfs), 1),
    pct_overlapping = round(100 * mean(n_overlapping_uorfs > 0), 1),
    .groups = "drop")

render_dt(uorf_summary, caption = "Table 8a: uORF Prevalence by Group")
```

Columns: **group** = PTC_neg (NMD without PTC), PTC_pos (NMD with PTC), or
Control (C4); **n** = number of isoforms; **median_utr5_bp** = median 5'UTR
length; **pct_with_uorf** = % with ≥1 uORF (≥3 codons); **median/mean_n_uorfs** =
uORFs per isoform; **pct_overlapping** = % with ≥1 overlapping uORF (stop
codon at or past CDS start).

```{r goal3-uorf-tests}
# ── Data provenance ──
# Source: uorf_data$per_isoform (from 05b_uorf_analysis.R)
# Methods: METHODS.md § "5'UTR uORF Analysis"
# Test: Fisher's exact test on 2x2 table (group × has_uorf) for each pair of groups
# "any uORF": isoform has ≥1 uORF (≥3 codons) in the 5'UTR
# "overlapping uORF": isoform has ≥1 uORF whose stop codon falls at or past CDS start
# n per group: PTC_neg=1,860, PTC_pos=1,458, Control=3,238

uorf_tests <- list()
test_pairs <- list(
  c("PTC_neg", "Control", "any uORF"),
  c("PTC_pos", "Control", "any uORF"),
  c("PTC_neg", "PTC_pos", "any uORF"),
  c("PTC_neg", "Control", "overlapping uORF")
)

for (tp in test_pairs) {
  sub <- uorf[uorf$group %in% tp[1:2], ]
  sub$group <- droplevels(factor(sub$group))
  if (tp[3] == "any uORF") {
    tbl <- table(sub$group, sub$n_uorfs > 0)
  } else {
    tbl <- table(sub$group, sub$n_overlapping_uorfs > 0)
  }
  ft <- tryCatch(fisher.test(tbl), error = function(e) NULL)
  if (!is.null(ft)) {
    n1 <- sum(sub$group == tp[1])
    n2 <- sum(sub$group == tp[2])
    uorf_tests[[length(uorf_tests) + 1]] <- data.frame(
      comparison = paste(tp[1], "vs", tp[2]),
      test = tp[3],
      n_group1 = n1, n_group2 = n2,
      OR = round(as.numeric(ft$estimate), 2),
      p_value = signif(as.numeric(ft$p.value), 3),
      stringsAsFactors = FALSE)
  }
}
uorf_test_df <- do.call(rbind, uorf_tests)
render_dt(uorf_test_df, caption = "Table 8b: uORF Enrichment Tests (Fisher's exact)")
```

Columns: **comparison** = group pair tested; **test** = "any uORF" (≥1 uORF) or
"overlapping uORF" (≥1 uORF with stop at/past CDS start); **n_group1/n_group2** =
group sizes; **OR** = Fisher's exact test odds ratio; **p_value** = Fisher p-value.

## Figure 8: uORF Count Distribution

```{r goal3-uorf-density, fig.width=10, fig.height=6}
# Density plot (better than dodged histogram for different group sizes)
uorf_plot <- uorf[uorf$utr5_length > 0, ]
uorf_plot$group_label <- factor(uorf_plot$group,
  levels = c("PTC_neg", "PTC_pos", "Control"),
  labels = c("NMD PTC-", "NMD PTC+", "Control"))

fig_uorf <- ggplot(uorf_plot, aes(x = n_uorfs + 0.5, fill = group_label)) +
  geom_density(alpha = 0.5) +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20, 50)) +
  scale_fill_manual(values = c("NMD PTC-" = "#fddbc7", "NMD PTC+" = "#ef8a62",
                                "Control" = "#67a9cf")) +
  labs(title = "uORF Count Distribution by Group",
       subtitle = "PTC-negative NMD isoforms have dramatically more uORFs (median 8 vs 1 in Control)",
       x = "Number of uORFs (+0.5, log scale)", y = "Density", fill = "")
save_fig(fig_uorf, "fig8_uorf_count_distribution", width = 10, height = 6)
```

**Key finding:** PTC-negative NMD comparators have dramatically more uORFs than
Control (96.2% vs 63.3% with any uORF, OR =
`r uorf_test_df$OR[uorf_test_df$comparison == "PTC_neg vs Control" & uorf_test_df$test == "any uORF"]`,
p < 10^-180^). They also have longer 5'UTRs (median
`r uorf_summary$median_utr5_bp[uorf_summary$group == "PTC_neg"]` bp vs
`r uorf_summary$median_utr5_bp[uorf_summary$group == "Control"]` bp in Control),
which provides more sequence for uORFs. Overlapping uORFs (stop codon at or past
main CDS start) are also enriched (61.5% vs 27.9%, OR =
`r uorf_test_df$OR[uorf_test_df$comparison == "PTC_neg vs Control" & uorf_test_df$test == "overlapping uORF"]`).

Combined with the 3'UTR length finding (PTC-neg NMD isoforms have normal 3'UTR
lengths), this points to **5'UTR uORFs as the dominant non-PTC NMD mechanism**
in this dataset, rather than long 3'UTR-mediated NMD.

**Caveat:** The high uORF count is partly driven by longer 5'UTRs (more sequence =
more chance of ATG + in-frame stop). The enrichment should be interpreted in
conjunction with the 5'UTR length difference.

## Figure 9: uORF Enrichment by Frame Category

```{r goal3-uorf-by-frame, fig.width=12, fig.height=6}
# ── Data provenance ──
# Source: uorf_data + fc_c2/fc_c4 pair_summary (frame categories)
# Merges uORF counts with frame category classification to show that
# diff_start_same_frame PTC-neg isoforms drive the uORF enrichment.

fc_c2_ps <- set_results[["c2_allsamples"]]$frame_comp$pair_summary
fc_c4_ps <- set_results[["c4_allsamples"]]$frame_comp$pair_summary

uorf_nmd_fc <- merge(uorf[uorf$group %in% c("PTC_neg", "PTC_pos"), ],
  fc_c2_ps[, c("comparator_isoform_id", "frame_category")],
  by.x = "isoform_id", by.y = "comparator_isoform_id", all.x = TRUE)
uorf_ctrl_fc <- merge(uorf[uorf$group == "Control", ],
  fc_c4_ps[, c("comparator_isoform_id", "frame_category")],
  by.x = "isoform_id", by.y = "comparator_isoform_id", all.x = TRUE)
uorf_by_fc <- rbind(uorf_nmd_fc, uorf_ctrl_fc)

# Filter to coding frame categories with reasonable sample sizes
coding_cats_plot <- c("same_start_in_frame", "same_start_frameshift",
  "diff_start_same_frame", "diff_start_diff_frame_with_frameshift")
uorf_by_fc <- uorf_by_fc[uorf_by_fc$frame_category %in% coding_cats_plot &
  !is.na(uorf_by_fc$frame_category), ]

uorf_by_fc$frame_label <- factor(uorf_by_fc$frame_category,
  levels = coding_cats_plot,
  labels = c("Same Start\nIn Frame", "Same Start\nFrameshift",
             "Diff Start\nSame Frame", "Diff Start\nDiff Frame + FS"))
uorf_by_fc$group_label <- factor(uorf_by_fc$group,
  levels = c("PTC_neg", "PTC_pos", "Control"),
  labels = c("NMD PTC-", "NMD PTC+", "Control"))

# Panel A: 5'UTR length by frame category and group
fig_utr5_fc <- ggplot(uorf_by_fc, aes(x = group_label, y = utr5_length, fill = group_label)) +
  geom_boxplot(outlier.alpha = 0.1) +
  facet_wrap(~frame_label, nrow = 1) +
  scale_fill_manual(values = c("NMD PTC-" = "#fddbc7", "NMD PTC+" = "#ef8a62",
                                "Control" = "#67a9cf")) +
  scale_y_log10() +
  labs(title = "5'UTR Length by Frame Category and NMD Status",
       subtitle = "Diff-start same-frame PTC-neg isoforms have dramatically longer 5'UTRs",
       x = "", y = "5'UTR length (bp, log scale)", fill = "") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
save_fig(fig_utr5_fc, "fig9_utr5_by_frame_category", width = 12, height = 6)
```

```{r goal3-uorf-by-frame-table}
# Summary table
uorf_fc_summary <- uorf_by_fc |>
  group_by(frame_label, group_label) |>
  summarise(
    n = n(),
    median_utr5_bp = round(median(utr5_length)),
    median_uorfs = round(median(n_uorfs)),
    pct_with_uorf = round(100 * mean(n_uorfs > 0), 1),
    .groups = "drop")

render_dt(uorf_fc_summary, caption = "Table 8c: uORF Prevalence by Frame Category and Group")
```

Columns: **frame_label** = reading frame category (from `compareIsoformFrames()`);
**group_label** = NMD PTC-, NMD PTC+, or Control; **n** = number of isoforms;
**median_utr5_bp** = median 5'UTR length; **median_uorfs** = median uORF count;
**pct_with_uorf** = % with ≥1 uORF.

```{r goal3-uorf-by-frame-tests}
# Wilcoxon test: 5'UTR length PTC_neg vs Control, per frame category
uorf_fc_tests <- do.call(rbind, lapply(levels(uorf_by_fc$frame_label), function(fc) {
  neg <- uorf_by_fc$utr5_length[uorf_by_fc$frame_label == fc & uorf_by_fc$group_label == "NMD PTC-"]
  ctrl <- uorf_by_fc$utr5_length[uorf_by_fc$frame_label == fc & uorf_by_fc$group_label == "Control"]
  if (length(neg) < 5 || length(ctrl) < 5) return(NULL)
  wt <- wilcox.test(neg, ctrl)
  data.frame(
    frame_category = fc,
    n_ptc_neg = length(neg), median_ptc_neg = round(median(neg)),
    n_control = length(ctrl), median_control = round(median(ctrl)),
    wilcox_p = signif(wt$p.value, 3),
    stringsAsFactors = FALSE)
}))

kable(uorf_fc_tests, caption = "Table 8d: 5'UTR Length — PTC-neg vs Control by Frame Category (Wilcoxon)")
```

Columns: **frame_category**; **n_ptc_neg/n_control** = group sizes;
**median_ptc_neg/median_control** = median 5'UTR length (bp);
**wilcox_p** = Wilcoxon rank-sum test p-value.

The **diff-start same-frame** PTC-negative NMD isoforms are the primary drivers
of the uORF enrichment: 99.1% have uORFs (median 9) with a median 5'UTR of
779 bp, compared to 81.9% (median 4, 5'UTR 446 bp) in Control. This is
consistent with the use of an alternative downstream start codon: the original
start codon becomes a uORF in the extended 5'UTR, and translation from the
downstream AUG produces the same reading frame but a shorter protein. The
resulting overlapping uORF (from the original AUG) is a potent NMD trigger.

**Limitations:** uORF detection is sequence-based and does not confirm
translation. Not all sequence-predicted uORFs are functional NMD triggers.
Other non-PTC NMD mechanisms (3'UTR secondary structure, EJC-independent UPF1
recruitment) cannot be ruled out from structural data alone.

## Cumulative NMD Mechanism Attribution

```{r goal4-cumulative-attribution, fig.width=10, fig.height=8}
# ── Data provenance ──
# Synthesizes across Goals 2-4 to provide a single accounting of NMD mechanisms.
# All counts derive from coding_pairs_c2 (gene-matched coding NMD pairs).
# PTC+/PTC-: from goal3-identify-ptc-neg (coding_pairs_c2$comp_has_ptc)
# Same-stop/diff-stop: from goal3-fig6-pop (strand-aware stop comparison)
# Frameshift/in-frame: from goal2-ptc-mechanisms (frame_category)

n_coding_total <- nrow(coding_pairs_c2)
n_same_stop <- length(same_stop_ids)
n_diff_stop <- length(diff_stop_ids)

n_diffstop_fs <- sum(ptc_plus$comparator_isoform_id %in% diff_stop_ids &
                       ptc_plus$is_frameshift_cat)
n_diffstop_inframe <- length(diff_stop_ids) - n_diffstop_fs

# ── Flowchart: NMD mechanism hierarchy ──
# Using DiagrammeR for a top-down flowchart
library(DiagrammeR)

grViz(sprintf('
digraph nmd_flowchart {
  graph [rankdir=TB, fontname="Helvetica", nodesep=0.6, ranksep=0.8]
  node [shape=box, style="filled,rounded", fontname="Helvetica", fontsize=12,
        width=2.5, height=0.6]

  all [label="All NMD+ coding pairs\\n%s", fillcolor="#f0f0f0"]

  ptc_pos [label="PTC+\\n%s (%s%%)", fillcolor="#fee0d2"]
  ptc_neg [label="PTC-\\n%s (%s%%)", fillcolor="#deebf7"]

  same_stop [label="Same stop codon\\n(3-prime UTR splice)\\n%s (%s%%)",
             fillcolor="#fc9272"]
  new_stop  [label="New stop codon\\n%s (%s%%)", fillcolor="#fc9272"]

  fs_new    [label="Frameshift-induced\\n%s (%s%%)", fillcolor="#fb6a4a"]
  if_new    [label="In-frame splice event\\n%s (%s%%)", fillcolor="#fb6a4a"]

  all -> ptc_pos
  all -> ptc_neg
  ptc_pos -> same_stop
  ptc_pos -> new_stop
  new_stop -> fs_new
  new_stop -> if_new
}',
  fmt(n_coding_total),
  fmt(n_ptc_pos), round(100 * n_ptc_pos / n_coding_total, 1),
  fmt(n_ptc_neg), round(100 * n_ptc_neg / n_coding_total, 1),
  fmt(n_same_stop), round(100 * n_same_stop / n_coding_total, 1),
  fmt(n_diff_stop), round(100 * n_diff_stop / n_coding_total, 1),
  fmt(n_diffstop_fs), round(100 * n_diffstop_fs / n_coding_total, 1),
  fmt(n_diffstop_inframe), round(100 * n_diffstop_inframe / n_coding_total, 1)
))
```

The flowchart shows the complete hierarchy for `r fmt(n_coding_total)` coding
gene-matched NMD pairs. The PTC+ branch (`r fmt(n_ptc_pos)` pairs,
`r round(100 * n_ptc_pos / n_coding_total, 1)`%) represents definitive
structural mechanisms: for each pair we identify the specific splice event
responsible (Figure 6). The PTC- branch (`r fmt(n_ptc_neg)` pairs,
`r round(100 * n_ptc_neg / n_coding_total, 1)`%) lacks individual-level
causal attribution, but shows group-level enrichment for 5'UTR uORFs
(Figure 8, Tables 8a-8d) and no evidence for 3'UTR length as an independent
trigger (Table 7).

**PTC+: New stop codon** (`r fmt(n_diff_stop)`,
`r round(100 * n_diff_stop / n_coding_total, 1)`%): A splice event introduces
a premature stop codon not present in the reference. Subdivides into
frameshift-induced (`r fmt(n_diffstop_fs)`) — a splice event shifts the reading
frame, and the new frame encounters a stop — and in-frame splice events
(`r fmt(n_diffstop_inframe)`) — the event directly introduces an in-frame stop
codon.

**PTC+: Same stop / 3'UTR splice** (`r fmt(n_same_stop)`,
`r round(100 * n_same_stop / n_coding_total, 1)`%): The comparator and
reference share the same stop codon. A 3'UTR splicing event repositions an EJC
>50 nt downstream, making the normal stop "premature" under the EJC model.

**PTC-** (`r fmt(n_ptc_neg)`,
`r round(100 * n_ptc_neg / n_coding_total, 1)`%): No individual causal
attribution. As a group, PTC-negative NMD comparators have longer 5'UTRs
(median `r uorf_summary$median_utr5_bp[uorf_summary$group == "PTC_neg"]` bp vs
`r uorf_summary$median_utr5_bp[uorf_summary$group == "Control"]` bp in Control)
and more uORFs (median `r uorf_summary$median_n_uorfs[uorf_summary$group == "PTC_neg"]` vs
`r uorf_summary$median_n_uorfs[uorf_summary$group == "Control"]`; OR =
`r round(1 / uorf_test_df$OR[uorf_test_df$comparison == "PTC_neg vs Control" & uorf_test_df$test == "any uORF"], 1)`,
p < 10^-180^), consistent with uORF-mediated NMD. Overlapping uORFs have been
shown to strongly repress main ORF expression through a combination of NMD and
translational inhibition (Luo & Bhalla 2018), and uORF-containing transcripts
are enriched among NMD targets (Nyikó et al. 2009; Ow et al. 2023). However,
the relative contribution of NMD vs translational repression has not been
cleanly separated, and individual attribution is not possible from sequence
data alone.


# 5. Protein-Level Consequences of Normal Alternative Splicing {#goal5}

**Goal:** When a cell uses its second-most-abundant isoform instead of the
dominant one, how different is the protein? This section examines the dominant vs
next-best non-NMD isoform for each gene. Unlike Goals 1-4 (gene-matched NMD vs
Control), this uses **all non-NMD isoform comparisons** — every gene with ≥2
non-NMD isoforms, deduplicated across cell types.

## Analysis Overview

```{r goal5-load}
# ── Data provenance ──
# Source: 05f_protein_analysis_v2.R → protein_analysis_v2.rds
# Pipeline:
#   1. All coding C4 isoform comparisons (unmatched), deduplicated
#   2. Isopair::mapSpliceToProtein() → protein-level regions per splice event
#   3. hmmscan (Pfam-A) → domain positions on each isoform's protein sequence
#   4. Domain-event intersection → disrupted/gained classification
#   5. Isopair::testDomainEnrichment() → Fisher's test vs reference prevalence
#   6. PeptideAtlas → mass-spec validation of splice-specific peptides

prot_v2 <- readRDS(file.path(cache_dir, "protein_analysis_v2.rds"))
pe <- prot_v2$protein_events
dom <- prot_v2$domain_effects
parsed_doms <- prot_v2$parsed_doms  # per-isoform hmmscan domain annotations

# Mass-spec: use the quality-enriched reanalysis (massspec_peptideatlas_v2.rds)
# as the single source of truth. prot_v2$massspec is from the initial run and
# has slightly different hit counts due to PeptideAtlas version differences.
ms <- readRDS(file.path(cache_dir, "massspec_peptideatlas_v2.rds"))

# Key counts (single source of truth for this section)
n_total       <- prot_v2$metadata$n_unique_pairs
n_with_events <- length(unique(paste(pe$reference_isoform_id, pe$comparator_isoform_id)))
n_events      <- nrow(pe)
n_dom_overlaps <- nrow(dom)
n_unique_doms  <- length(unique(dom$pfam_acc))
n_disrupted   <- length(unique(paste(dom$comparator_isoform_id[dom$effect == "disrupted_lost"],
                                      dom$reference_isoform_id[dom$effect == "disrupted_lost"])))
n_gained      <- length(unique(paste(dom$comparator_isoform_id[dom$effect == "gained_intact"],
                                      dom$reference_isoform_id[dom$effect == "gained_intact"])))
n_with_domains <- length(unique(paste(dom$comparator_isoform_id, dom$reference_isoform_id)))
n_overrep_tested <- nrow(prot_v2$overrep_tests)
n_overrep_sig    <- sum(prot_v2$overrep_tests$significant, na.rm = TRUE)
n_ms_queried  <- nrow(ms)
n_ms_hits     <- sum(ms$n_pa_hits > 0)
n_peptides    <- sum(ms$n_novel_peptides)
n_pep_hits    <- sum(ms$n_pa_hits)

# Pair keys for joining pe/dom/ms
pe$pair_key  <- paste(pe$reference_isoform_id, pe$comparator_isoform_id, sep = "|")
dom$pair_key <- paste(dom$reference_isoform_id, dom$comparator_isoform_id, sep = "|")
# ms already has pair_key from 05f_protein_analysis_v2.R

# MS hit lookup (pair keys with ≥1 PeptideAtlas hit)
ms_hit_keys <- ms$pair_key[ms$n_pa_hits > 0]

# Gene symbol annotation for ms
gene_lk <- setNames(pe$gene_id, pe$comparator_isoform_id)
gene_lk <- gene_lk[!duplicated(names(gene_lk))]
ms$gene_id <- gene_lk[ms$comparator_isoform_id]
ms$gene_symbol <- gene_symbol_lookup[ms$gene_id]
ms_hits <- ms[ms$n_pa_hits > 0, ]

# Cell-type membership (protein analysis uses lowercase labels)
ct_membership <- prot_v2$ct_membership
ct_map <- c(AT = "at", DD = "dd", DO = "do", FB = "fb", MV = "mv")
```

The unit tracked through this section is always **isoform comparisons** (one per
gene: dominant vs next-best non-NMD isoform).

```{r goal5-flowchart}
# Count domain-affected pairs with MS support (for flowchart node)
dom_pair_keys <- unique(dom$pair_key)
n_dom_with_ms <- sum(dom_pair_keys %in% ms_hit_keys)

# C4 pairs in all_samples (unique, no CT duplication)
n_c4_allsamples <- nrow(readRDS(file.path("data_mashr", "profiles_c4_allsamples.rds")))

dot_g5 <- sprintf('
digraph goal5 {
  graph [rankdir=TB, fontname="Helvetica", nodesep=2.0, ranksep=1.8, size="16,28"]
  node  [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=28, margin="0.8,0.6", width=11]
  edge  [color="#666666", penwidth=3.0]

  Z [label=<<b>%s Non-NMD Isoforms</b> (from Section 0)<br/>→ C4 pairs: top 2 non-NMD isoforms per gene<br/>%s unique pairs in all_samples>,
     fillcolor="#f0f0f0", color="#969696", style="rounded,filled,dashed"]
  A [label=<<b>%s Unique Coding Isoform Comparisons</b><br/>(C4 pairs pooled across 6 CT sets, both isoforms protein-coding,<br/>deduplicated — some genes gain pairs in per-CT analyses)>,
     fillcolor="#f0f0f0", color="#636363"]
  B [label=<<b>%s Have CDS-Overlapping Splice Events</b><br/>(%s without CDS events excluded)>,
     fillcolor="#e8f4f8", color="#2c7fb8"]
  C [label=<<b>%s Have ≥1 Pfam Domain Affected</b><br/>(%s disrupted/lost | %s gained)>,
     fillcolor="#fff7bc", color="#d95f0e"]
  D [label=<<b>%s / %s Domains Over-Represented</b><br/>(FDR &lt; 0.05 vs reference prevalence)>,
     fillcolor="#c7e9c0", color="#31a354"]
  E [label=<<b>%s Domain-Affected Pairs Have Mass-Spec Support</b><br/>(splice-specific peptide in PeptideAtlas)>,
     fillcolor="#dadaeb", color="#6a51a3"]
  F [label=<<b>Cell-Type-Specific Analysis</b><br/>(domain overlap by CT, direction × CT interaction)>,
     fillcolor="#fde0dd", color="#c51b8a"]

  Z -> A [label="  pool CTs, coding filter, dedup", fontsize=22, fontname="Helvetica"]
  A -> B -> C -> D
  C -> E
  B -> F
}',
  fmt(n_nonnmd_as),
  fmt(n_c4_allsamples),
  fmt(n_total),
  fmt(n_with_events), fmt(n_total - n_with_events),
  fmt(n_with_domains), fmt(n_disrupted), fmt(n_gained),
  fmt(n_overrep_sig), fmt(n_overrep_tested),
  fmt(n_dom_with_ms))

DiagrammeR::grViz(dot_g5, width = 1200, height = 1100)
```

## Figure 10: Domain Effects of Alternative Splicing

Of `r fmt(n_total)` coding isoform comparisons, `r fmt(n_with_events)`
(`r round(100 * n_with_events / n_total, 1)`%) have ≥1 CDS-overlapping splice
event. Among those, `r fmt(n_with_domains)` (`r round(100 * n_with_domains / n_total, 1)`%
of total) overlap ≥1 Pfam domain, with `r fmt(n_disrupted)` disrupted and
`r fmt(n_gained)` gained.

```{r goal5-domain-bar, fig.width=10, fig.height=5}
n_both <- length(intersect(
  unique(paste(dom$comparator_isoform_id[dom$effect == "disrupted_lost"],
               dom$reference_isoform_id[dom$effect == "disrupted_lost"])),
  unique(paste(dom$comparator_isoform_id[dom$effect == "gained_intact"],
               dom$reference_isoform_id[dom$effect == "gained_intact"]))))

bar_data <- data.frame(
  category = c("No CDS-overlapping events",
               "CDS event, no domain overlap",
               "Both disrupted & gained",
               "Gained only",
               "Disrupted/lost only"),
  n = c(n_total - n_with_events,
        n_with_events - n_with_domains,
        n_both,
        n_gained - n_both,
        n_disrupted - n_both))
bar_data$pct <- round(100 * bar_data$n / n_total, 1)
bar_data$category <- factor(bar_data$category, levels = bar_data$category)
bar_data$label <- sprintf("%s (%.1f%%)", format(bar_data$n, big.mark = ","), bar_data$pct)

ggplot(bar_data, aes(x = category, y = n, fill = category)) +
  geom_col() +
  geom_text(aes(label = label), hjust = -0.1, size = 4) +
  scale_fill_manual(values = c(
    "Disrupted/lost only" = "#d6604d",
    "Gained only" = "#4393c3",
    "Both disrupted & gained" = "#8856a7",
    "CDS event, no domain overlap" = "#bdbdbd",
    "No CDS-overlapping events" = "#f0f0f0")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3))) +
  labs(title = "Domain Effects of Alternative Splicing",
       subtitle = sprintf("Denominator: %s coding isoform comparisons. Categories are mutually exclusive.",
                           fmt(n_total)),
       x = "", y = "Number of Isoform Comparisons") +
  theme(legend.position = "none") +
  coord_flip()
```

## Table 8: Protein-Level Analysis Summary

```{r goal5-summary-table}
# ── Data provenance ──
# All counts from goal5-load (single source of truth for this section).
# n_total, n_with_events, n_with_domains, n_disrupted, n_gained, n_both:
#   computed from pe and dom in goal5-load and goal5-domain-bar chunks.
# n_ms_queried, n_ms_hits: from massspec_peptideatlas_v2.rds.
# n_events, n_dom_overlaps, n_unique_doms: event- and domain-level counts.

pooled_summary <- data.frame(
  metric = c("Total coding isoform comparisons",
             "  — with ≥1 CDS-overlapping splice event",
             "  — without CDS events",
             "Total CDS-overlapping events",
             "Pairs with ≥1 Pfam domain affected",
             "  — disrupted/lost only",
             "  — gained only",
             "  — both disrupted & gained",
             "Total domain-event overlaps",
             "Unique Pfam domains affected",
             "Domains over-represented (FDR < 0.05)",
             "Pairs with mass-spec support"),
  n = c(n_total,
        n_with_events,
        n_total - n_with_events,
        n_events,
        n_with_domains,
        n_disrupted - n_both,
        n_gained - n_both,
        n_both,
        n_dom_overlaps,
        n_unique_doms,
        n_overrep_sig,
        n_ms_hits),
  pct = c(NA,
          round(100 * n_with_events / n_total, 1),
          round(100 * (n_total - n_with_events) / n_total, 1),
          NA,
          round(100 * n_with_domains / n_total, 1),
          round(100 * (n_disrupted - n_both) / n_total, 1),
          round(100 * (n_gained - n_both) / n_total, 1),
          round(100 * n_both / n_total, 1),
          NA, NA, NA,
          round(100 * n_ms_hits / n_ms_queried, 1)))

render_dt(pooled_summary,
  caption = "Table 8: Protein-Level Analysis Summary (Pooled)")
```

Columns: **metric**; **n** = count; **pct** = percentage of the
`r fmt(n_total)` coding isoform comparisons for pair-level rows, or of the
`r fmt(n_ms_queried)` comparisons with splice-specific peptides for the
mass-spec row. Event- and domain-level counts have no percentage.

## Event-Level Gating: GAIN vs LOSS Domain Overlap

GAIN and LOSS events may interact differently with protein domains. A LOSS event
removes sequence from the comparator, potentially **disrupting** an existing
domain. A GAIN event adds sequence, potentially **introducing** an intact domain
that the reference lacks.

```{r goal5-direction-gating}
# ── Data provenance ──
# pe = prot_v2$protein_events (CDS-overlapping splice events, has direction column)
# dom = prot_v2$domain_effects (domain-event overlaps, has effect column)
# Join: pe events → dom overlaps via (reference_isoform_id, comparator_isoform_id,
#        event_type, event_protein_start/protein_start_aa, event_protein_end/protein_end_aa)

# Create event key for pe
pe$event_key <- paste(pe$reference_isoform_id, pe$comparator_isoform_id,
                      pe$event_type, pe$protein_start_aa, pe$protein_end_aa,
                      sep = "|")

# Create matching key for dom
dom$event_key <- paste(dom$reference_isoform_id, dom$comparator_isoform_id,
                       dom$event_type, dom$event_protein_start, dom$event_protein_end,
                       sep = "|")

# Flag each pe event as overlapping any domain or not
pe$has_domain_overlap <- pe$event_key %in% dom$event_key

# Separate by direction
gain_events <- pe[pe$direction == "GAIN", ]
loss_events <- pe[pe$direction == "LOSS", ]

n_gain <- nrow(gain_events)
n_loss <- nrow(loss_events)
n_gain_dom <- sum(gain_events$has_domain_overlap)
n_loss_dom <- sum(loss_events$has_domain_overlap)
pct_gain_dom <- round(100 * n_gain_dom / n_gain, 1)
pct_loss_dom <- round(100 * n_loss_dom / n_loss, 1)

# Fisher's exact test: domain overlap rate for GAIN vs LOSS
ft_direction <- fisher.test(matrix(c(n_gain_dom, n_gain - n_gain_dom,
                                     n_loss_dom, n_loss - n_loss_dom), nrow = 2))
```

GAIN events overlap at least one Pfam domain at a rate of **`r pct_gain_dom`%**
(`r fmt(n_gain_dom)` / `r fmt(n_gain)`), compared to **`r pct_loss_dom`%**
(`r fmt(n_loss_dom)` / `r fmt(n_loss)`) for LOSS events (Fisher's exact OR =
`r round(ft_direction$estimate, 2)`, p = `r signif(ft_direction$p.value, 3)`).
GAIN events preferentially target domain-containing protein regions.

### Figure 11: Domain Overlap Rate by Direction

```{r goal5-direction-bar, fig.width=8, fig.height=4}
dir_bar <- data.frame(
  direction = c("GAIN", "LOSS"),
  pct = c(pct_gain_dom, pct_loss_dom),
  n = c(n_gain, n_loss),
  n_dom = c(n_gain_dom, n_loss_dom))
dir_bar$label <- sprintf("%.1f%%\n(%s / %s)", dir_bar$pct,
                          fmt(dir_bar$n_dom), fmt(dir_bar$n))

fig11 <- ggplot(dir_bar, aes(x = direction, y = pct, fill = direction)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = label), vjust = -0.3, size = 4) +
  scale_fill_manual(values = c(GAIN = "#4393c3", LOSS = "#d6604d")) +
  scale_y_continuous(limits = c(0, max(dir_bar$pct) * 1.25),
                     expand = expansion(mult = c(0, 0))) +
  labs(title = "Domain Overlap Rate by Splice Event Direction",
       subtitle = sprintf("Fisher OR = %.2f, p = %s. Total events: %s GAIN, %s LOSS.",
                           ft_direction$estimate, signif(ft_direction$p.value, 3),
                           fmt(n_gain), fmt(n_loss)),
       x = "Event Direction", y = "% of Events Overlapping ≥1 Pfam Domain") +
  theme(legend.position = "none")

save_fig(fig11, "fig11_domain_overlap_by_direction")
```

## Domain Enrichment by Direction

The following tables test which specific Pfam domains are over-represented
among disrupted and gained effects, and compare the two.

### Table 9: Disrupted Domain Enrichment

```{r goal5-enrichment}
# ── Data provenance ──
# overrep = prot_v2$overrep_tests (Fisher enrichment of disrupted domains)
# MS annotation: count disrupted pairs per domain that also have PeptideAtlas support
overrep <- prot_v2$overrep_tests

# Add MS support columns
disrupted_dom <- dom[dom$effect == "disrupted_lost", ]
disrupted_ms <- disrupted_dom |>
  dplyr::group_by(pfam_acc) |>
  dplyr::summarise(
    n_ms_supported = dplyr::n_distinct(pair_key[pair_key %in% ms_hit_keys]),
    .groups = "drop")
overrep <- merge(overrep, disrupted_ms, by = "pfam_acc", all.x = TRUE)
overrep$n_ms_supported[is.na(overrep$n_ms_supported)] <- 0L
overrep$pct_ms <- round(100 * overrep$n_ms_supported / pmax(overrep$n_disrupted, 1), 1)

overrep_sig <- overrep[overrep$significant == TRUE & !is.na(overrep$significant), ]
overrep_sig <- overrep_sig[order(overrep_sig$fisher_p), ]
overrep_top <- head(overrep_sig, 30)
overrep_top$fisher_OR <- round(overrep_top$fisher_OR, 2)
overrep_top$padj <- signif(overrep_top$padj, 3)

render_dt(overrep_top[, c("pfam_acc", "domain_name", "n_ref", "n_disrupted",
  "n_ms_supported", "pct_ms", "fisher_OR", "padj")],
  caption = "Table 9: Top 30 Over-Represented Disrupted Domains (FDR < 0.05)")
```

Of `r fmt(n_overrep_tested)` Pfam domains tested, **`r fmt(n_overrep_sig)` are
significantly over-represented** among disrupted domains (FDR < 0.05). The test
compares each domain's disruption rate to its prevalence among reference
(dominant) isoforms.

Columns: **pfam_acc** = Pfam accession; **domain_name**; **n_ref** = reference
isoforms with this domain; **n_disrupted** = comparisons where this domain
overlaps a splice event; **n_ms_supported** = disrupted pairs with ≥1
PeptideAtlas hit for splice-specific peptides; **pct_ms** = % of disrupted pairs
with MS support; **fisher_OR** = odds ratio; **padj** = BH-adjusted p-value.

### Table 9b: Gained Domain Enrichment

```{r goal5-gained-enrichment}
# ── Data provenance ──
# dom$effect == "gained_intact": events where a GAIN introduces a domain
# Denominator: comparator-side domain prevalence from parsed_doms (hmmscan)
# n_total: total unique coding pairs (from prot_v2$metadata$n_unique_pairs)
# MS annotation: count gained pairs per domain that also have PeptideAtlas support

gained <- dom[dom$effect == "gained_intact", ]
gained_counts <- gained |>
  dplyr::group_by(pfam_acc) |>
  dplyr::summarise(
    n_gained = dplyr::n_distinct(paste(reference_isoform_id, comparator_isoform_id)),
    domain_name = domain_name[1],
    .groups = "drop")

n_any_gained <- length(unique(paste(
  gained$reference_isoform_id, gained$comparator_isoform_id)))

# Add MS support columns
gained_ms <- gained |>
  dplyr::group_by(pfam_acc) |>
  dplyr::summarise(
    n_ms_supported = dplyr::n_distinct(pair_key[pair_key %in% ms_hit_keys]),
    .groups = "drop")

# Comparator-side domain prevalence: how many comparator isoforms carry each domain?
comp_ids <- unique(pe$comparator_isoform_id)
comp_dom_counts <- parsed_doms[parsed_doms$isoform_id %in% comp_ids, ] |>
  dplyr::group_by(pfam_acc) |>
  dplyr::summarise(n_comp = dplyr::n_distinct(isoform_id), .groups = "drop")

gained_enrich <- merge(gained_counts, comp_dom_counts, by = "pfam_acc", all.x = TRUE)
gained_enrich$n_comp[is.na(gained_enrich$n_comp)] <- 0L
gained_enrich <- merge(gained_enrich, gained_ms, by = "pfam_acc", all.x = TRUE)
gained_enrich$n_ms_supported[is.na(gained_enrich$n_ms_supported)] <- 0L
gained_enrich$pct_ms <- round(100 * gained_enrich$n_ms_supported /
                                pmax(gained_enrich$n_gained, 1), 1)

# Fisher's exact test per domain
# 2×2: (gained / not gained) × (has domain / no domain)
# Denominator: comparator isoforms with this domain (n_comp)
gained_enrich$fisher_p <- NA_real_
gained_enrich$fisher_OR <- NA_real_
for (i in seq_len(nrow(gained_enrich))) {
  a <- gained_enrich$n_gained[i]
  b <- max(gained_enrich$n_comp[i] - a, 0L)
  c_val <- max(n_any_gained - a, 0L)
  d_val <- max(n_total - a - b - c_val, 0L)
  ft <- tryCatch(fisher.test(matrix(c(a, c_val, b, d_val), nrow = 2)),
                 error = function(e) NULL)
  if (!is.null(ft)) {
    gained_enrich$fisher_p[i] <- ft$p.value
    gained_enrich$fisher_OR[i] <- as.numeric(ft$estimate)
  }
}
gained_enrich$padj <- p.adjust(gained_enrich$fisher_p, method = "BH")
gained_enrich$significant <- gained_enrich$padj < 0.05
gained_enrich <- gained_enrich[order(gained_enrich$fisher_p), ]

n_gained_tested <- nrow(gained_enrich)
n_gained_sig <- sum(gained_enrich$significant, na.rm = TRUE)

gained_sig <- gained_enrich[gained_enrich$significant == TRUE &
                              !is.na(gained_enrich$significant), ]
gained_top <- head(gained_sig, 30)
gained_top$fisher_OR <- round(gained_top$fisher_OR, 2)
gained_top$padj <- signif(gained_top$padj, 3)

render_dt(gained_top[, c("pfam_acc", "domain_name", "n_comp", "n_gained",
  "n_ms_supported", "pct_ms", "fisher_OR", "padj")],
  caption = "Table 9b: Top 30 Over-Represented Gained Domains (FDR < 0.05)")
```

Of `r fmt(n_gained_tested)` Pfam domains tested, **`r fmt(n_gained_sig)` are
significantly over-represented** among gained-intact domains (FDR < 0.05). The
test compares each domain's rate of being gained intact to its prevalence among
comparator isoforms (the isoforms that carry the gained domain).

Columns: **pfam_acc** = Pfam accession; **domain_name**; **n_comp** = comparator
isoforms carrying this domain (baseline prevalence); **n_gained** = comparisons
where this domain is gained intact; **n_ms_supported** = gained pairs with ≥1
PeptideAtlas hit; **pct_ms** = % of gained pairs with MS support;
**fisher_OR** = odds ratio; **padj** = BH-adjusted p-value.

### Table 9c: Disrupted vs Gained Comparison

```{r goal5-disrupted-vs-gained}
# ── Data provenance ──
# For each domain with ≥5 affected pairs (disrupted or gained):
# Test whether the gained fraction exceeds the overall baseline.
# Baseline = overall gained fraction across all domain-affected pairs.

# Merge disrupted and gained counts
dg_compare <- merge(
  overrep[, c("pfam_acc", "domain_name", "n_disrupted")],
  gained_counts[, c("pfam_acc", "n_gained")],
  by = "pfam_acc", all = TRUE)
dg_compare$n_disrupted[is.na(dg_compare$n_disrupted)] <- 0L
dg_compare$n_gained[is.na(dg_compare$n_gained)] <- 0L

# Fill domain_name from gained_counts where missing
missing_name <- is.na(dg_compare$domain_name)
if (any(missing_name)) {
  name_lk <- setNames(gained_counts$domain_name, gained_counts$pfam_acc)
  dg_compare$domain_name[missing_name] <- name_lk[dg_compare$pfam_acc[missing_name]]
}

dg_compare$n_total_affected <- dg_compare$n_disrupted + dg_compare$n_gained
dg_compare$pct_gained <- round(100 * dg_compare$n_gained / dg_compare$n_total_affected, 1)

# Filter to domains with ≥5 affected pairs
dg_testable <- dg_compare[dg_compare$n_total_affected >= 5, ]

# Overall baseline: fraction of all domain effects that are gained
baseline_gained <- n_any_gained / (n_any_gained +
  length(unique(paste(
    dom$comparator_isoform_id[dom$effect == "disrupted_lost"],
    dom$reference_isoform_id[dom$effect == "disrupted_lost"]))))

# One-sided binomial test per domain: is gained fraction > baseline?
dg_testable$binom_p <- NA_real_
for (i in seq_len(nrow(dg_testable))) {
  bt <- tryCatch(
    binom.test(dg_testable$n_gained[i], dg_testable$n_total_affected[i],
               p = baseline_gained, alternative = "greater"),
    error = function(e) NULL)
  if (!is.null(bt)) dg_testable$binom_p[i] <- bt$p.value
}
dg_testable$padj <- p.adjust(dg_testable$binom_p, method = "BH")
dg_testable$significant <- dg_testable$padj < 0.05
dg_testable <- dg_testable[order(-dg_testable$pct_gained), ]

n_dg_tested <- nrow(dg_testable)
n_dg_sig <- sum(dg_testable$significant, na.rm = TRUE)

dg_display <- dg_testable[, c("pfam_acc", "domain_name", "n_disrupted",
  "n_gained", "n_total_affected", "pct_gained", "padj")]
dg_display$padj <- signif(dg_display$padj, 3)

render_dt(dg_display,
  caption = sprintf("Table 9c: Disrupted vs Gained by Domain (≥5 affected pairs, baseline gained = %.1f%%)",
                    100 * baseline_gained))
```

Of `r fmt(n_dg_tested)` domains with ≥5 affected pairs, **`r fmt(n_dg_sig)`**
have a gained fraction significantly above the overall baseline of
**`r round(100 * baseline_gained, 1)`%** (one-sided binomial, BH-adjusted
p < 0.05).

Columns: **pfam_acc** = Pfam accession; **domain_name**; **n_disrupted** =
pairs where the domain is disrupted; **n_gained** = pairs where the domain is
gained intact; **n_total_affected** = disrupted + gained; **pct_gained** =
percentage gained; **padj** = BH-adjusted binomial p-value.

## Mass Spectrometry Integration

Domain overlap results are cross-referenced with PeptideAtlas mass-spectrometry
data to assess whether domain-affecting splice events produce detectable protein
products.

### Table 10: Domain × Mass-Spec Cross-Tabulation

```{r goal5-domain-ms-crosstab}
# ── Data provenance ──
# Universe: pairs with CDS-overlapping events (n_with_events)
# dom$pair_key: pairs with domain overlap
# ms_hit_keys: pairs with ≥1 PeptideAtlas hit (from ms loaded in goal5-load)
# 2×2: (domain affected / not) × (MS hit / not)

cds_pair_keys <- unique(pe$pair_key)
has_dom <- cds_pair_keys %in% dom_pair_keys
has_ms  <- cds_pair_keys %in% ms_hit_keys

a <- sum(has_dom & has_ms)
b <- sum(has_dom & !has_ms)
c_val <- sum(!has_dom & has_ms)
d_val <- sum(!has_dom & !has_ms)

ft_dom_ms <- fisher.test(matrix(c(a, c_val, b, d_val), nrow = 2))

crosstab <- data.frame(
  category = c("Domain affected", "No domain affected", "Total"),
  ms_hit = c(a, c_val, a + c_val),
  no_ms_hit = c(b, d_val, b + d_val),
  total = c(a + b, c_val + d_val, a + b + c_val + d_val))

kable(crosstab,
  caption = "Table 10: Domain × Mass-Spec Cross-Tabulation",
  col.names = c("", "MS Hit", "No MS Hit", "Total"))
```

Among `r fmt(n_with_events)` pairs with CDS-overlapping events,
domain-affected pairs have a mass-spec hit rate of
**`r round(100 * a / (a + b), 1)`%** vs **`r round(100 * c_val / (c_val + d_val), 1)`%**
for pairs without domain overlap (Fisher OR =
`r round(ft_dom_ms$estimate, 2)`, p = `r signif(ft_dom_ms$p.value, 3)`).

Columns: **MS Hit** = pairs with ≥1 PeptideAtlas-confirmed splice-specific
peptide; **No MS Hit** = no PeptideAtlas match; rows stratified by whether the
pair has ≥1 Pfam domain affected by alternative splicing.

### Mass-Spec Validation Summary

```{r goal5-massspec-summary}
# ── Data provenance ──
# Source: massspec_peptideatlas_v2.rds loaded in goal5-load chunk (single source of truth)
# Pipeline: splice-specific peptides from event-affected protein regions
#   → exact match against Human PeptideAtlas 2024-01 (1.5M peptides with
#   n_observations, n_samples, best_probability per peptide)

ms_summary <- data.frame(
  metric = c("Comparisons with splice-specific peptides",
             "Comparisons with ≥1 PeptideAtlas hit",
             "  — with ≥2 confirmed peptides",
             "  — with ≥4 confirmed peptides",
             "  — high-confidence (best peptide in ≥10 samples)",
             "Total splice-specific peptides queried",
             "Total peptides found in PeptideAtlas"),
  n = c(n_ms_queried, n_ms_hits,
        sum(ms_hits$n_pa_hits >= 2),
        sum(ms_hits$n_pa_hits >= 4),
        sum(ms_hits$max_n_samples >= 10),
        n_peptides, n_pep_hits),
  pct_of_queried = c(NA, round(100 * n_ms_hits / n_ms_queried, 1),
        round(100 * sum(ms_hits$n_pa_hits >= 2) / n_ms_queried, 1),
        round(100 * sum(ms_hits$n_pa_hits >= 4) / n_ms_queried, 1),
        round(100 * sum(ms_hits$max_n_samples >= 10) / n_ms_queried, 1),
        NA, round(100 * n_pep_hits / n_peptides, 1)))

kable(ms_summary, caption = "Table 10a: Mass Spectrometry Validation Summary",
      col.names = c("Metric", "N", "% of queried"))
```

Columns: **Metric**; **N** = count; **% of queried** = percentage of the
`r fmt(n_ms_queried)` comparisons with splice-specific peptides (first block)
or of total peptides (last row).

## Cell-Type Domain Analysis

The analyses above use deduplicated pairs pooled across cell types. Here we
stratify by cell type to assess whether domain overlap patterns vary across
the five lung cell populations. Pairs appear in all cell types where they were
observed, so cell types are **not independent** — a pair shared across three
CTs contributes to each. This inflates per-CT sample sizes; interpret
cross-CT comparisons as descriptive rather than formally independent tests.

### Table 9d: Domain Overlap by Cell Type

```{r goal5-ct-domain}
# ── Data provenance ──
# ct_membership: named list from prot_v2, key = "ref_id|comp_id", value = CT labels
#   (lowercase: "at", "dd", "do", "fb", "mv")
# pe$pair_key, dom$pair_key: pair-level keys matching ct_membership names
# Subsetting pe and dom by ct_membership to compute per-CT metrics

ct_domain_rows <- lapply(main_cts, function(ct) {
  ct_lower <- ct_map[ct]
  ct_keys <- names(ct_membership)[sapply(ct_membership, function(x) ct_lower %in% x)]

  pe_ct <- pe[pe$pair_key %in% ct_keys, ]
  dom_ct <- dom[dom$pair_key %in% ct_keys, ]

  n_pairs_ct <- length(ct_keys)
  n_cds_ct <- length(unique(pe_ct$pair_key))
  n_dom_ct <- length(unique(dom_ct$pair_key))
  n_dis_ct <- length(unique(dom_ct$pair_key[dom_ct$effect == "disrupted_lost"]))
  n_gai_ct <- length(unique(dom_ct$pair_key[dom_ct$effect == "gained_intact"]))

  data.frame(
    cell_type = ct,
    n_pairs = n_pairs_ct,
    n_cds_events = n_cds_ct,
    n_domain_overlap = n_dom_ct,
    pct_domain_overlap = round(100 * n_dom_ct / max(n_cds_ct, 1), 1),
    n_disrupted = n_dis_ct,
    n_gained = n_gai_ct)
})
ct_domain_table <- do.call(rbind, ct_domain_rows)

render_dt(ct_domain_table,
  caption = "Table 9d: Domain Overlap Summary by Cell Type")
```

Columns: **cell_type**; **n_pairs** = coding isoform pairs in this CT;
**n_cds_events** = pairs with ≥1 CDS-overlapping splice event;
**n_domain_overlap** = pairs with ≥1 Pfam domain affected;
**pct_domain_overlap** = % of CDS-event pairs with domain overlap;
**n_disrupted** = pairs with ≥1 disrupted domain; **n_gained** = pairs with ≥1
gained domain.

### Figure 11b: Direction × Cell Type

```{r goal5-ct-direction, fig.width=10, fig.height=5}
# ── Data provenance ──
# Expand events to CT level via ct_membership, then compute domain overlap rate
# by CT × direction. CMH test for overall direction × domain overlap association
# stratified by cell type.

ct_dir_rows <- list()
for (ct in main_cts) {
  ct_lower <- ct_map[ct]
  ct_keys <- names(ct_membership)[sapply(ct_membership, function(x) ct_lower %in% x)]
  pe_ct <- pe[pe$pair_key %in% ct_keys, ]

  for (dir in c("GAIN", "LOSS")) {
    events_dir <- pe_ct[pe_ct$direction == dir, ]
    n_dir <- nrow(events_dir)
    n_dom_dir <- sum(events_dir$has_domain_overlap)
    ct_dir_rows[[length(ct_dir_rows) + 1]] <- data.frame(
      cell_type = ct, direction = dir,
      n_events = n_dir, n_dom = n_dom_dir,
      pct_dom = round(100 * n_dom_dir / max(n_dir, 1), 1))
  }
}
ct_dir_df <- do.call(rbind, ct_dir_rows)

fig11b <- ggplot(ct_dir_df, aes(x = cell_type, y = pct_dom, fill = direction)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = sprintf("%.0f%%", pct_dom)),
            position = position_dodge(width = 0.7), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c(GAIN = "#4393c3", LOSS = "#d6604d")) +
  scale_y_continuous(limits = c(0, max(ct_dir_df$pct_dom) * 1.2),
                     expand = expansion(mult = c(0, 0))) +
  labs(title = "Domain Overlap Rate by Direction and Cell Type",
       subtitle = "% of CDS-overlapping events that overlap ≥1 Pfam domain",
       x = "Cell Type", y = "% Events with Domain Overlap", fill = "Direction") +
  theme(legend.position = "bottom")

save_fig(fig11b, "fig11b_direction_by_celltype", width = 10, height = 5)

# CMH test: direction × domain overlap, stratified by CT
# Build 3D array: 2 (direction) × 2 (domain overlap) × 5 (CT)
cmh_array <- array(0L, dim = c(2, 2, length(main_cts)),
  dimnames = list(direction = c("GAIN", "LOSS"),
                  domain = c("overlap", "no_overlap"),
                  ct = main_cts))

for (i in seq_along(main_cts)) {
  ct <- main_cts[i]
  ct_lower <- ct_map[ct]
  ct_keys <- names(ct_membership)[sapply(ct_membership, function(x) ct_lower %in% x)]
  pe_ct <- pe[pe$pair_key %in% ct_keys, ]

  for (dir in c("GAIN", "LOSS")) {
    events_dir <- pe_ct[pe_ct$direction == dir, ]
    cmh_array[dir, "overlap", i] <- sum(events_dir$has_domain_overlap)
    cmh_array[dir, "no_overlap", i] <- sum(!events_dir$has_domain_overlap)
  }
}

cmh_test <- mantelhaen.test(cmh_array)
```

GAIN events consistently show higher domain overlap rates than LOSS events
across all cell types (Figure 11b). A Cochran-Mantel-Haenszel test confirms
this association is significant when stratified by cell type (common OR =
`r round(cmh_test$estimate, 2)`, p = `r signif(cmh_test$p.value, 3)`).

**Note:** Because pairs can appear in multiple cell types, CT strata are not
fully independent. The CMH test assumes independent strata; the p-value should
be interpreted conservatively.

### Table 9e: Cell-Type-Specific Domain Enrichment

```{r goal5-ct-domain-enrichment, fig.width=10, fig.height=8}
# ── Data provenance ──
# For each Pfam domain: build a 2×5 table (affected/not × 5 CTs).
# Denominator per CT: pairs with CDS events in that CT.
# Test: chi-squared with Monte Carlo simulation (handles sparse cells).
# H0: the rate at which domain X is affected is the same across all CTs.
# BH correction across tested domains.

# Pre-compute per-CT CDS-event pair sets and domain-affected pair sets
ct_cds_keys <- list()
ct_dom_affected <- list()  # pair_key → pfam_acc affected
for (ct in main_cts) {
  ct_lower <- ct_map[ct]
  ct_keys <- names(ct_membership)[sapply(ct_membership, function(x) ct_lower %in% x)]
  ct_cds_keys[[ct]] <- unique(pe$pair_key[pe$pair_key %in% ct_keys])
  dom_ct <- dom[dom$pair_key %in% ct_keys, ]
  ct_dom_affected[[ct]] <- dom_ct[, c("pair_key", "pfam_acc")]
}

# Get all domains with ≥10 total affected pairs across CTs AND ≥2 CTs with ≥1
all_doms <- unique(dom$pfam_acc)
dom_ct_counts <- list()
for (d in all_doms) {
  cts_with <- 0L
  total <- 0L
  per_ct <- integer(length(main_cts))
  names(per_ct) <- main_cts
  for (j in seq_along(main_cts)) {
    ct <- main_cts[j]
    n_aff <- length(unique(ct_dom_affected[[ct]]$pair_key[
      ct_dom_affected[[ct]]$pfam_acc == d]))
    per_ct[j] <- n_aff
    total <- total + n_aff
    if (n_aff > 0) cts_with <- cts_with + 1L
  }
  if (total >= 10 && cts_with >= 2) {
    dom_ct_counts[[d]] <- per_ct
  }
}

n_ct_enrich_tested <- length(dom_ct_counts)

# Run chi-squared tests
ct_enrich_results <- list()
set.seed(42)  # reproducibility for Monte Carlo
for (d in names(dom_ct_counts)) {
  affected <- dom_ct_counts[[d]]
  n_cds <- sapply(main_cts, function(ct) length(ct_cds_keys[[ct]]))
  not_affected <- n_cds - affected

  # 2×5 matrix: rows = affected/not, cols = CTs
  mat <- rbind(affected, not_affected)

  # Monte Carlo for p-value (handles sparse cells)
  ct_mc <- tryCatch(
    chisq.test(mat, simulate.p.value = TRUE, B = 10000),
    error = function(e) NULL)
  # Standard chi-squared for residuals (stdres not available with simulation)
  ct_std <- tryCatch(
    chisq.test(mat),
    error = function(e) NULL)

  if (!is.null(ct_mc) && !is.null(ct_std)) {
    std_res <- ct_std$stdres[1, ]
    ct_enrich_results[[d]] <- data.frame(
      pfam_acc = d,
      chisq = round(ct_mc$statistic, 2),
      p_value = ct_mc$p.value,
      max_residual_ct = names(which.max(std_res)),
      max_residual = round(max(std_res), 2),
      min_residual_ct = names(which.min(std_res)),
      min_residual = round(min(std_res), 2),
      stringsAsFactors = FALSE)
  }
}

ct_enrich_df <- do.call(rbind, ct_enrich_results)
ct_enrich_df$padj <- p.adjust(ct_enrich_df$p_value, method = "BH")
ct_enrich_df$significant <- ct_enrich_df$padj < 0.05
ct_enrich_df <- ct_enrich_df[order(ct_enrich_df$p_value), ]

# Add domain name lookup
dom_name_lk <- setNames(dom$domain_name, dom$pfam_acc)
dom_name_lk <- dom_name_lk[!duplicated(names(dom_name_lk))]
ct_enrich_df$domain_name <- dom_name_lk[ct_enrich_df$pfam_acc]

n_ct_enrich_sig <- sum(ct_enrich_df$significant, na.rm = TRUE)

# Add per-CT percentage columns to ALL tested domains (needed for display)
for (ct in main_cts) {
  ct_enrich_df[[paste0("pct_", ct)]] <- sapply(ct_enrich_df$pfam_acc, function(d) {
    n_aff <- dom_ct_counts[[d]][ct]
    n_cds <- length(ct_cds_keys[[ct]])
    round(100 * n_aff / max(n_cds, 1), 2)
  })
}

ct_enrich_sig <- ct_enrich_df[ct_enrich_df$significant == TRUE &
                                !is.na(ct_enrich_df$significant), ]

display_cols <- c("pfam_acc", "domain_name",
  paste0("pct_", main_cts), "chisq", "padj",
  "max_residual_ct", "max_residual", "min_residual_ct", "min_residual")

if (nrow(ct_enrich_sig) > 0) {
  ct_enrich_display <- head(ct_enrich_sig[, display_cols], 30)
  ct_enrich_display$padj <- signif(ct_enrich_display$padj, 3)
  render_dt(ct_enrich_display,
    caption = sprintf("Table 9e: Domains with Cell-Type-Specific Enrichment (FDR < 0.05, %s tested)",
                      fmt(n_ct_enrich_tested)))
} else {
  cat("No domains with significant cell-type heterogeneity at FDR < 0.05.\n")
}
```

Of `r fmt(n_ct_enrich_tested)` Pfam domains tested (≥10 affected pairs across
≥2 cell types), **`r fmt(n_ct_enrich_sig)`** show significant heterogeneity in
their rate of being affected across cell types (chi-squared with Monte Carlo
simulation, BH-adjusted p < 0.05).

Columns: **pfam_acc** = Pfam accession; **domain_name**; **pct_AT** through
**pct_MV** = % of CDS-event pairs in each CT where this domain is affected;
**chisq** = chi-squared statistic; **padj** = BH-adjusted p-value;
**max_residual_ct** = cell type with the largest positive standardized residual
(most enriched); **max_residual** = that residual value; **min_residual_ct** =
cell type with the largest negative residual (most depleted);
**min_residual** = that residual value.

### Figure 11c: Cell-Type-Specific Domain Heatmap

```{r goal5-ct-domain-heatmap, fig.width=10, fig.height=8}
# ── Data provenance ──
# Standardized residuals from the chi-squared tests above.
# Positive residual = domain affected more than expected in that CT.
# Negative residual = affected less than expected.

if (nrow(ct_enrich_sig) > 0) {
  # Build residual matrix for top significant domains (up to 30)
  top_doms <- head(ct_enrich_sig$pfam_acc, 30)

  resid_rows <- list()
  for (d in top_doms) {
    affected <- dom_ct_counts[[d]]
    n_cds <- sapply(main_cts, function(ct) length(ct_cds_keys[[ct]]))
    not_affected <- n_cds - affected
    mat <- rbind(affected, not_affected)
    ct_test <- chisq.test(mat)  # standard chi-sq for stdres
    resid_rows[[d]] <- ct_test$stdres[1, ]
  }

  resid_mat <- do.call(rbind, resid_rows)
  rownames(resid_mat) <- paste0(dom_name_lk[top_doms], " (", top_doms, ")")

  # Reshape for ggplot
  resid_df <- data.frame(
    domain = rep(rownames(resid_mat), ncol(resid_mat)),
    cell_type = rep(colnames(resid_mat), each = nrow(resid_mat)),
    residual = as.vector(resid_mat),
    stringsAsFactors = FALSE)
  resid_df$domain <- factor(resid_df$domain, levels = rev(rownames(resid_mat)))

  # Cap residuals at ±4 for color scale
  resid_df$residual_capped <- pmin(pmax(resid_df$residual, -4), 4)

  fig11c <- ggplot(resid_df, aes(x = cell_type, y = domain, fill = residual_capped)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = round(residual, 1)), size = 3) +
    scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                         midpoint = 0, limits = c(-4, 4),
                         name = "Std. Residual") +
    labs(title = "Cell-Type-Specific Domain Enrichment",
         subtitle = sprintf("Top %d domains with significant heterogeneity (FDR < 0.05). Positive = enriched, negative = depleted.",
                            length(top_doms)),
         x = "Cell Type", y = "") +
    theme(axis.text.y = element_text(size = 9),
          panel.grid = element_blank())

  save_fig(fig11c, "fig11c_ct_domain_heatmap", width = 10, height = max(4, length(top_doms) * 0.35))
} else {
  cat("No domains with significant cell-type heterogeneity at FDR < 0.05.\n")
}
```

The heatmap displays standardized residuals from the chi-squared tests. A
positive residual (red) indicates the domain is affected more frequently than
expected in that cell type; a negative residual (blue) indicates less frequently.
Residuals beyond ±2 are conventionally considered noteworthy.

**Non-independence caveat:** Pairs shared across cell types contribute to
multiple columns. This inflates chi-squared statistics and may produce
anti-conservative p-values. The results identify candidate CT-specific domain
patterns, but formal confirmation would require CT-exclusive pair analysis or a
mixed-effects framework.

# 6. Cross-Cell-Type Summary {#summary}

```{r sec6-summary-table}
# ── Data provenance ──
# Starting data: gene-matched profiles and cached analyses across all cell types
# Each row computed via helper functions that access set_results[[key]]:
#   - Pair counts: nrow(gene-matched profiles)
#   - PTC rates: compute_ptc_rates_row() — coding pairs from gene-matched profiles + ptc.rds
#   - Divergence: compute_divergence_row() — from cached quantifyPairDivergence()
#   - Productive frameshifts: from pf_data precompute (all_samples only)
# All helpers operate on gene-matched data (profiles_list was matched in STAGE 2)

summary_rows <- lapply(c("all_samples", main_cts), function(ct) {
  ct_label <- ct_to_label(ct)

  # NMD and Control pair counts
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  n_nmd  <- if (c2_key %in% names(set_results)) nrow(set_results[[c2_key]]$profiles) else NA
  n_ctrl <- if (c4_key %in% names(set_results)) nrow(set_results[[c4_key]]$profiles) else NA

  # PTC rates
  ptc_row <- compute_ptc_rates_row(ct)
  ptc_rate_nmd  <- if (!is.null(ptc_row)) ptc_row$NMD_comp_ptc_pct else NA
  ptc_rate_ctrl <- if (!is.null(ptc_row)) ptc_row$Ctrl_comp_ptc_pct else NA
  ptc_or        <- if (!is.null(ptc_row)) ptc_row$fisher_OR else NA

  # PTC-neg %
  ptc_neg_pct <- if (!is.null(ptc_row)) round(100 - ptc_row$NMD_comp_ptc_pct, 1) else NA

  # Median pct_shared for C4
  div_row <- compute_divergence_row(ct)
  med_shared <- if (!is.null(div_row)) div_row$Ctrl_median_pct_shared else NA

  # Productive frameshifts (only available for all_samples)
  n_pf <- if (ct == "all_samples") nrow(pf_data$characterization) else NA

  data.frame(
    cell_type = ct,
    N_NMD_pairs = n_nmd,
    N_Ctrl_pairs = n_ctrl,
    PTC_rate_NMD_pct = ptc_rate_nmd,
    PTC_rate_Ctrl_pct = ptc_rate_ctrl,
    PTC_OR = ptc_or,
    PTC_neg_pct = ptc_neg_pct,
    Median_pct_shared_C4 = med_shared,
    N_productive_FS = n_pf)
})

summary_table <- do.call(rbind, summary_rows)
render_dt(summary_table, caption = "Table 11: Cross-Cell-Type Summary")
write.csv(summary_table, file.path(table_dir, "table11_cross_ct_summary.csv"), row.names = FALSE)
```

Columns: **cell_type** = sample set; **N_NMD/N_Ctrl_pairs** = gene-matched pair
counts; **PTC_rate_NMD/Ctrl_pct** = PTC rate among coding comparators;
**PTC_OR** = Fisher OR for PTC enrichment; **PTC_neg_pct** = percentage of
coding NMD comparators without PTC; **Median_pct_shared_C4** = median exonic
sequence shared in Control pairs (from `quantifyPairDivergence()`);
**N_productive_FS** = productive frameshifts (all_samples only).

# Session Info

```{r session-info}
sessionInfo()
```
