---
title: "NMD as Post-Transcriptional Reading Frame Selection"
subtitle: "Isopair Wrapper — Isocall Pipeline, mashr Classification"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
library(Isopair)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(DT)
library(pROC)
library(glmnet)
library(patchwork)
source("visualization_functions.R")

# Helper: render interactive DT table (sortable, searchable)
render_dt <- function(df, caption = NULL, pageLength = 15, ...) {
  datatable(df, caption = caption, rownames = FALSE,
            options = list(pageLength = pageLength, scrollX = TRUE),
            ...)
}

# ---------------------------------------------------------------------------
# Publication theme and color palette
# ---------------------------------------------------------------------------
theme_pub <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
theme_set(theme_pub)

pal <- c(NMD = "#ef8a62", Control = "#67a9cf")


# Figure output
fig_dir <- "figures"
dir.create(fig_dir, showWarnings = FALSE)
table_dir <- "tables"
dir.create(table_dir, showWarnings = FALSE)

save_fig <- function(p, name, width = 8, height = 5) {
  ggsave(file.path(fig_dir, paste0(name, ".pdf")), p, width = width, height = height)
  ggsave(file.path(fig_dir, paste0(name, ".png")), p, width = width, height = height, dpi = 300)
  print(p)
}

# ---------------------------------------------------------------------------
# Helper functions (adapted from 03_nmd_analysis_mashr.Rmd)
# ---------------------------------------------------------------------------
ct_to_label <- function(ct) tolower(gsub("_", "", ct))

comp_label <- function(comp) {
  switch(toupper(comp), C2 = "NMD", C4 = "Control", toupper(comp))
}

evt_types <- c("Alt_TSS", "Alt_TES", "A5SS", "A3SS", "SE",
               "Missing_Internal", "IR", "IR_diff",
               "Partial_IR_5", "Partial_IR_3")

compute_event_freq <- function(profiles) {
  n_pairs <- nrow(profiles)
  n_pir5 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_5")))
  n_pir3 <- sum(sapply(profiles$detailed_events, function(de)
    any(de$event_type == "Partial_IR_3")))
  counts <- c(sum(profiles$tss_changed), sum(profiles$tes_changed),
              sum(profiles$n_a5ss > 0), sum(profiles$n_a3ss > 0),
              sum(profiles$n_se > 0), sum(profiles$n_missing_internal > 0),
              sum(profiles$n_ir > 0), sum(profiles$n_ir_diff > 0),
              n_pir5, n_pir3)
  data.frame(event_type = evt_types,
             n_pairs_with_event = counts,
             pct_of_pairs = round(100 * counts / n_pairs, 1))
}

compute_ptc_rates_row <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  coding_ids <- cds$isoform_id[cds$coding_status == "coding"]
  p_c2 <- set_results[[c2_key]]$profiles
  cod_c2 <- p_c2[p_c2$reference_isoform_id %in% coding_ids &
                   p_c2$comparator_isoform_id %in% coding_ids, ]
  p_c4 <- set_results[[c4_key]]$profiles
  cod_c4 <- p_c4[p_c4$reference_isoform_id %in% coding_ids &
                   p_c4$comparator_isoform_id %in% coding_ids, ]
  if (nrow(cod_c2) == 0 || nrow(cod_c4) == 0) return(NULL)
  ptc_c2 <- sum(cod_c2$comparator_isoform_id %in% ptc$isoform_id[ptc$has_ptc == TRUE])
  ptc_c4 <- sum(cod_c4$comparator_isoform_id %in% ptc$isoform_id[ptc$has_ptc == TRUE])
  n_c2 <- nrow(cod_c2); n_c4 <- nrow(cod_c4)
  ft <- tryCatch({
    mat <- matrix(c(ptc_c2, n_c2 - ptc_c2, ptc_c4, n_c4 - ptc_c4), nrow = 2)
    fisher.test(mat)
  }, error = function(e) NULL)
  data.frame(
    cell_type = ct,
    n_coding_pairs_NMD = n_c2,
    n_coding_pairs_Ctrl = n_c4,
    NMD_comp_ptc_n = ptc_c2,
    NMD_comp_ptc_pct = round(100 * ptc_c2 / n_c2, 1),
    Ctrl_comp_ptc_n = ptc_c4,
    Ctrl_comp_ptc_pct = round(100 * ptc_c4 / n_c4, 1),
    fisher_OR = if (!is.null(ft)) round(ft$estimate, 2) else NA_real_,
    p_value = if (!is.null(ft)) ft$p.value else NA_real_
  )
}

compute_frame_category_rows <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  fc_c2 <- set_results[[c2_key]]$frame_comp
  fc_c4 <- set_results[[c4_key]]$frame_comp
  if (is.null(fc_c2) || is.null(fc_c4)) return(NULL)
  ps_c2 <- fc_c2$pair_summary
  ps_c4 <- fc_c4$pair_summary
  cats <- c("same_start_in_frame", "same_start_frameshift",
            "diff_start_same_frame", "diff_start_diff_frame",
            "diff_start_diff_frame_with_frameshift",
            "non_coding", "no_shared_cds")
  # Denominator: coding pairs only (exclude non_coding and no_shared_cds)
  n_coding_c2 <- sum(!ps_c2$frame_category %in% c("non_coding", "no_shared_cds"))
  n_coding_c4 <- sum(!ps_c4$frame_category %in% c("non_coding", "no_shared_cds"))
  do.call(rbind, lapply(cats, function(cat) {
    is_coding_cat <- !cat %in% c("non_coding", "no_shared_cds")
    data.frame(
      cell_type = ct,
      frame_category = cat,
      NMD_n = sum(ps_c2$frame_category == cat),
      NMD_pct = if (is_coding_cat) round(100 * sum(ps_c2$frame_category == cat) / n_coding_c2, 1) else NA_real_,
      Ctrl_n = sum(ps_c4$frame_category == cat),
      Ctrl_pct = if (is_coding_cat) round(100 * sum(ps_c4$frame_category == cat) / n_coding_c4, 1) else NA_real_
    )
  }))
}

compute_divergence_row <- function(ct) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(set_results) || !c4_key %in% names(set_results)) return(NULL)
  div_c2 <- set_results[[c2_key]]$divergence
  div_c4 <- set_results[[c4_key]]$divergence
  if (is.null(div_c2) || nrow(div_c2) == 0 || is.null(div_c4) || nrow(div_c4) == 0) return(NULL)
  data.frame(
    cell_type = ct,
    NMD_n = nrow(div_c2),
    NMD_median_pct_shared = round(median(div_c2$pct_shared, na.rm = TRUE), 1),
    Ctrl_n = nrow(div_c4),
    Ctrl_median_pct_shared = round(median(div_c4$pct_shared, na.rm = TRUE), 1)
  )
}
```

# 0. Data Loading and Gene-Matching

## Data Provenance Overview

All analyses in this report begin from **gene-matched isoform pair sets**. The
data pipeline is:

1. **01_prepare_data_mashr.R**: Loads the isocall count matrix
   (`nmd_isocall.count_matrix.txt`, 645,273 isoforms x 38 samples), excludes
   2 outlier samples (DO donor 029T: `Sample25_DO_029T_DMSO` and
   `Sample26_DO_029T_Smg1i`, identified as PCA outlier), applies TMM
   normalization and condition-stratified 5% isoform filtering, then classifies
   isoforms as NMD-sensitive (`nmd_responsive == TRUE` in mashr DIE results) or
   non-NMD (`adj.P.Val > 0.50`). Final dataset: 36 samples across 6 cell types.
   Outputs: `expression_data.rds`, `nmd_classification.rds`,
   `sample_metadata.rds`, `gene_map.rds`.

2. **02_build_profiles_mashr.R**: Generates isoform pairs —
   **C4 (Control)**: top-two non-NMD isoforms by DMSO CPM per gene
   (`generatePairsExpression(..., method = "top_two")`);
   **C2 (NMD)**: same C4 reference paired with top NMD isoform by Smg1i CPM
   (via `identifyDominantIsoforms()` + `inner_join` on `gene_id`).
   Builds splicing choice profiles with `Isopair::buildProfiles()`.
   Outputs: `profiles_c{2,4}_*.rds`, `pairs_c{2,4}_*.rds`.

3. **03_nmd_analysis_mashr.Rmd**: Applies **gene-matching** (restricts C2 and C4
   to genes with shared `(gene_id, reference_isoform_id)` in both sets), then
   runs Isopair analysis functions on the matched profiles. Caches results to
   `analysis_cache/`.

4. **This report (05)**: Loads profiles from step 2, re-applies gene-matching
   (step 3), loads cached analysis results, and verifies consistency. All
   downstream analyses use gene-matched profiles exclusively.

```{r flowchart-data, include=FALSE}
# ── Gather numbers for the flowchart ──
# These are computed fresh from the loaded data objects to ensure accuracy.
# Raw count matrix dimensions are hardcoded (not loaded here to save memory).
n_raw_isoforms <- 645273L
n_raw_samples  <- 38L
n_excluded     <- 2L
```

```{r flowchart, fig.width=10, fig.height=10}
# ── Data provenance ──
# All numbers derived from objects loaded in subsequent chunks.
# Raw count matrix dimensions from isocall pipeline output.
# Filtered/classified counts from 01_prepare_data_mashr.R outputs.
# Pair counts from 02_build_profiles_mashr.R outputs.
# Gene-matched counts computed in load-profiles chunk below.
#
# We load the key objects here just for the flowchart numbers, then
# the full load happens in the subsequent chunks.

fmt <- function(x) format(x, big.mark = ",")

# Load just what we need for counts
.fc_expr   <- readRDS(file.path("data_mashr", "expression_data.rds"))
.fc_sm     <- readRDS(file.path("data_mashr", "sample_metadata.rds"))
.fc_nmd    <- readRDS(file.path("data_mashr", "nmd_classification.rds"))
.fc_gm     <- readRDS(file.path("data_mashr", "gene_map.rds"))

n_filtered   <- nrow(.fc_expr)
n_samples    <- ncol(.fc_expr)
n_dmso       <- sum(.fc_sm$treatment == "DMSO")
n_smg1i      <- sum(.fc_sm$treatment == "Smg1i")
n_nmd_as     <- length(.fc_nmd[["all_samples"]]$nmd)
n_nonnmd_as  <- length(.fc_nmd[["all_samples"]]$non_nmd)

# How many genes have both NMD and non-NMD isoforms?
nmd_genes <- unique(.fc_gm$gene_id[.fc_gm$isoform_id %in% .fc_nmd[["all_samples"]]$nmd])
nonnmd_genes <- unique(.fc_gm$gene_id[.fc_gm$isoform_id %in% .fc_nmd[["all_samples"]]$non_nmd])
# Genes with ≥2 non-NMD (needed for C4)
nonnmd_per_gene <- table(.fc_gm$gene_id[.fc_gm$isoform_id %in% .fc_nmd[["all_samples"]]$non_nmd])
genes_2plus_nonnmd <- names(nonnmd_per_gene[nonnmd_per_gene >= 2])
# Genes eligible for both C2 and C4
genes_eligible <- intersect(nmd_genes, genes_2plus_nonnmd)
n_genes_eligible <- length(genes_eligible)

# Pair counts (all_samples, before gene-matching)
.fc_c2 <- readRDS(file.path("data_mashr", "profiles_c2_allsamples.rds"))
.fc_c4 <- readRDS(file.path("data_mashr", "profiles_c4_allsamples.rds"))
n_c2_before <- nrow(.fc_c2)
n_c4_before <- nrow(.fc_c4)

# Gene-matched counts
.shared <- dplyr::inner_join(
  unique(.fc_c2[, c("gene_id", "reference_isoform_id")]),
  unique(.fc_c4[, c("gene_id", "reference_isoform_id")]),
  by = c("gene_id", "reference_isoform_id"))
n_matched_genes <- nrow(.shared)
n_c2_matched <- nrow(dplyr::semi_join(.fc_c2, .shared, by = c("gene_id", "reference_isoform_id")))
n_c4_matched <- nrow(dplyr::semi_join(.fc_c4, .shared, by = c("gene_id", "reference_isoform_id")))

rm(.fc_expr, .fc_sm, .fc_nmd, .fc_gm, .fc_c2, .fc_c4, .shared)

dot_spec <- sprintf('
digraph flowchart {
  graph [rankdir=TB, fontname="Helvetica", nodesep=2.0, ranksep=1.8, size="16,20"]
  node  [shape=box, style="rounded,filled", fontname="Helvetica", fontsize=30, margin="0.8,0.6", width=8]
  edge  [color="#666666", penwidth=3.0]

  A [label=<<b>Isocall Count Matrix</b><br/>%s isoforms × %d samples>,
     fillcolor="#f0f0f0", color="#636363"]

  B [label=<<b>Sample Exclusion</b><br/>Remove DO donor 029T (PCA outlier)<br/>%s isoforms × %d samples>,
     fillcolor="#fee0d2", color="#de2d26"]

  C [label=<<b>5%% Filter + filterByExpr</b><br/>%s isoforms pass<br/>%d DMSO + %d Smg1i | 6 cell types>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  D [label=<<b>NMD Classification (mashr)</b><br/>NMD-sensitive: %s isoforms (union across CTs)<br/>Non-NMD: %s isoforms (intersection across CTs)>,
     fillcolor="#fff7bc", color="#d95f0e"]

  D2 [label=<<b>Eligible Genes</b><br/>%s genes have ≥1 NMD isoform AND ≥2 non-NMD isoforms<br/>(required for both C2 and C4 pair construction)>,
      fillcolor="#fff7bc", color="#d95f0e"]

  E [label=<<b>Pair Construction (all_samples)</b><br/>C2 (NMD): %s pairs (1 per gene)<br/>C4 (Control): %s pairs (1 per gene)>,
     fillcolor="#e8f4f8", color="#2c7fb8"]

  F [label=<<b>Gene-Matching</b><br/>%s genes with shared (gene, reference) in both C2 and C4<br/>C2: %s pairs | C4: %s pairs>,
     fillcolor="#c7e9c0", color="#31a354"]

  A -> B -> C -> D -> D2 -> E -> F
}',
  fmt(n_raw_isoforms), n_raw_samples,
  fmt(n_raw_isoforms), n_samples,
  fmt(n_filtered), n_dmso, n_smg1i,
  fmt(n_nmd_as), fmt(n_nonnmd_as),
  fmt(n_genes_eligible),
  fmt(n_c2_before), fmt(n_c4_before),
  fmt(n_matched_genes), fmt(n_c2_matched), fmt(n_c4_matched)
)

if (requireNamespace("DiagrammeR", quietly = TRUE)) {
  DiagrammeR::grViz(dot_spec, width = 1200, height = 900)
} else {
  message("DiagrammeR not installed -- flowchart not rendered")
}
```

```{r load-data}
# ==============================================================================
# STAGE 1: Load raw data objects
# ==============================================================================

# ── 01_prepare_data_mashr.R outputs ──
# Source: isocall count matrix → TMM normalization → 5% filter → mashr NMD classification
data_dir  <- "data_mashr"
cache_dir <- file.path(data_dir, "analysis_cache")

expr_mat    <- readRDS(file.path(data_dir, "expression_data.rds"))
sample_meta <- readRDS(file.path(data_dir, "sample_metadata.rds"))
gene_map    <- readRDS(file.path(data_dir, "gene_map.rds"))
nmd_class   <- readRDS(file.path(data_dir, "nmd_classification.rds"))

# ── Infrastructure objects (isoform-intrinsic, DE-independent) ──
# Originally from Isopair pipeline scripts 02-05, copied to data_mashr/ by
# 02_build_profiles_mashr.R for self-containment.
# structures.rds: extractIsoformStructures() — exon coordinates per isoform
# cds.rds: extractCds() — CDS boundaries per isoform
# ptc.rds: testPtcAssociation() infrastructure — PTC status per isoform
# union_exons.rds: buildUnionExons() — merged exon models per gene
# isoform_union_mapping.rds: mapIsoformsToUnion() — isoform-to-union-exon mapping
# annotated_ue.rds: annotateUnionExons() — union exons with CDS/UTR labels
# cds_exons.rds: CDS mapped to individual exons
structures   <- readRDS(file.path(data_dir, "structures.rds"))
cds          <- readRDS(file.path(data_dir, "cds.rds"))
ptc          <- readRDS(file.path(data_dir, "ptc.rds"))
union_exons  <- readRDS(file.path(data_dir, "union_exons.rds"))
ue_mapping   <- readRDS(file.path(data_dir, "isoform_union_mapping.rds"))
annotated_ue <- readRDS(file.path(data_dir, "annotated_ue.rds"))
cds_exons    <- readRDS(file.path(data_dir, "cds_exons.rds"))

# ── Gene symbol lookup from mashr DE files ──
de_dir_mashr <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/mashr"
gene_symbol_lookup <- character(0)
for (de_f in list.files(de_dir_mashr, pattern = "^nmd_mashr_die_.*\\.csv$", full.names = TRUE)) {
  de_tmp <- read.csv(de_f, stringsAsFactors = FALSE)[, c("gene_id", "hgnc_symbol")]
  de_tmp <- de_tmp[!duplicated(de_tmp$gene_id) & !is.na(de_tmp$hgnc_symbol) &
                    de_tmp$hgnc_symbol != "", ]
  new_entries <- setdiff(de_tmp$gene_id, names(gene_symbol_lookup))
  gene_symbol_lookup[de_tmp$gene_id[de_tmp$gene_id %in% new_entries]] <-
    de_tmp$hgnc_symbol[de_tmp$gene_id %in% new_entries]
}

get_gene_sym <- function(gid) {
  sym <- gene_symbol_lookup[gid]
  if (is.na(sym) || sym == "") gid else sym
}

# Cell types (mashr: DO included, DD_ALI excluded from main)
cell_types  <- c("all_samples", "AT", "DD", "DO", "FB", "MV")
comparisons <- c("c2", "c4")
main_cts    <- c("AT", "DD", "DO", "FB", "MV")
```

```{r load-profiles}
# ==============================================================================
# STAGE 2: Load profiles and apply gene-matching
# ==============================================================================
#
# profiles_c{2,4}_*.rds are the output of 02_build_profiles_mashr.R. C2 pairs
# share their reference isoform with C4 (by construction in script 02), but C4
# may contain genes that have no NMD isoform (i.e., genes with >=2 non-NMD
# isoforms but no mashr-classified NMD isoform). Gene-matching restricts both
# sets to the intersection of (gene_id, reference_isoform_id), ensuring fair
# NMD vs Control comparison on identical gene sets.

profiles_list <- list()
for (ct in cell_types) {
  ct_label <- ct_to_label(ct)
  for (comp in comparisons) {
    key <- paste(comp, ct_label, sep = "_")
    prof_file <- file.path(data_dir, sprintf("profiles_%s.rds", key))
    if (file.exists(prof_file)) profiles_list[[key]] <- readRDS(prof_file)
  }
}

# ── Gene-matching ──
# Replicates the logic from 03_nmd_analysis_mashr.Rmd lines 743-775.
# For each cell type, restrict C2 and C4 profiles to genes where both have
# a pair with the same (gene_id, reference_isoform_id).
gene_match_summary <- list()
for (ct in cell_types) {
  ct_label <- ct_to_label(ct)
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  if (!c2_key %in% names(profiles_list) || !c4_key %in% names(profiles_list)) next

  n_c2_before <- nrow(profiles_list[[c2_key]])
  n_c4_before <- nrow(profiles_list[[c4_key]])

  c2_gene_refs <- unique(profiles_list[[c2_key]][, c("gene_id", "reference_isoform_id")])
  c4_gene_refs <- unique(profiles_list[[c4_key]][, c("gene_id", "reference_isoform_id")])
  shared_refs  <- dplyr::inner_join(c2_gene_refs, c4_gene_refs,
                                     by = c("gene_id", "reference_isoform_id"))

  profiles_list[[c2_key]] <- dplyr::semi_join(profiles_list[[c2_key]], shared_refs,
                                               by = c("gene_id", "reference_isoform_id"))
  profiles_list[[c4_key]] <- dplyr::semi_join(profiles_list[[c4_key]], shared_refs,
                                               by = c("gene_id", "reference_isoform_id"))

  gene_match_summary[[ct]] <- data.frame(
    cell_type      = ct,
    c2_before      = n_c2_before,
    c4_before      = n_c4_before,
    matched_genes  = nrow(shared_refs),
    c2_after       = nrow(profiles_list[[c2_key]]),
    c4_after       = nrow(profiles_list[[c4_key]])
  )
}
gene_match_df <- do.call(rbind, gene_match_summary)
```

```{r build-set-results}
# ==============================================================================
# STAGE 3: Build set_results from gene-matched profiles + cached analyses
# ==============================================================================
#
# set_results[[key]] is the primary data structure for all downstream analyses.
# $profiles: gene-matched splicing choice profiles (from STAGE 2 above)
# $divergence: quantifyPairDivergence() output (cached by script 03)
# $frame_comp: compareIsoformFrames() output (cached by script 03)
# $frame_walk: analyzeFrameWalk() output (cached by script 03)
# $ptc_assoc: testPtcAssociation() output (cached by script 03)
# $evt_freq: computed here from gene-matched profiles

set_results <- list()
for (key in names(profiles_list)) {
  profiles <- profiles_list[[key]]
  set_results[[key]] <- list(
    profiles   = profiles,
    divergence = tryCatch(readRDS(file.path(cache_dir, paste0("div_", key, ".rds"))), error = function(e) NULL),
    frame_comp = tryCatch(readRDS(file.path(cache_dir, paste0("fc_",  key, ".rds"))), error = function(e) NULL),
    frame_walk = tryCatch(readRDS(file.path(cache_dir, paste0("fw_",  key, ".rds"))), error = function(e) NULL),
    ptc_assoc  = tryCatch(readRDS(file.path(cache_dir, paste0("ptc_", key, ".rds"))), error = function(e) NULL),
    evt_freq   = compute_event_freq(profiles)
  )
}

# Load divergence for all_samples (used by multiple sections)
div_c2 <- set_results[["c2_allsamples"]]$divergence
div_c4 <- set_results[["c4_allsamples"]]$divergence
```

```{r verify-cache-consistency}
# ==============================================================================
# STAGE 4: Verify cached analysis objects match gene-matched profiles
# ==============================================================================
#
# The cached div/fc/fw/ptc objects were computed by script 03 from gene-matched
# profiles. We verify that the row counts in the cached objects match our
# gene-matched profiles_list. A mismatch would indicate the cache is stale or
# was computed from a different pair set.

verification_rows <- list()
for (key in names(set_results)) {
  sr <- set_results[[key]]
  n_profiles <- nrow(sr$profiles)

  # divergence: 1 row per pair (should exactly match profiles)
  n_div <- if (!is.null(sr$divergence)) nrow(sr$divergence) else NA
  # frame_comp$pair_summary: 1 row per pair (should exactly match)
  n_fc <- if (!is.null(sr$frame_comp)) nrow(sr$frame_comp$pair_summary) else NA
  # frame_walk$profile_summary: 1 row per pair (should exactly match)
  n_fw <- if (!is.null(sr$frame_walk)) nrow(sr$frame_walk$profile_summary) else NA
  # ptc_assoc$pair_summary: 1 row per pair where PTC analysis was possible
  #   (both isoforms coding with EJC data). Expected to be <= n_profiles.
  n_ptc <- if (!is.null(sr$ptc_assoc)) nrow(sr$ptc_assoc$pair_summary) else NA

  # Exact match expected for div, fc, fw
  div_ok <- is.na(n_div) || n_div == n_profiles
  fc_ok  <- is.na(n_fc)  || n_fc  == n_profiles
  fw_ok  <- is.na(n_fw)  || n_fw  == n_profiles
  # ptc: subset is OK (non-coding pairs excluded), but should not exceed profiles
  ptc_ok <- is.na(n_ptc) || n_ptc <= n_profiles

  verification_rows[[key]] <- data.frame(
    key = key, n_profiles = n_profiles,
    n_div = n_div, n_fc = n_fc, n_fw = n_fw, n_ptc = n_ptc,
    exact_match = div_ok & fc_ok & fw_ok,
    ptc_subset_ok = ptc_ok
  )
}
verification_df <- do.call(rbind, verification_rows)

if (!all(verification_df$exact_match)) {
  warning("CACHE MISMATCH DETECTED — div/fc/fw row counts do not match ",
          "gene-matched profiles. Affected keys: ",
          paste(verification_df$key[!verification_df$exact_match], collapse = ", "))
}
if (!all(verification_df$ptc_subset_ok)) {
  warning("PTC CACHE ANOMALY — ptc_assoc has MORE rows than profiles for: ",
          paste(verification_df$key[!verification_df$ptc_subset_ok], collapse = ", "))
}
```

```{r load-meta-and-precomputes}
# ==============================================================================
# STAGE 5: Load mashr meta-analysis logFC and precomputed objects
# ==============================================================================

# ── mashr meta-analysis logFC ──
# Source: mashr model fit on isocall DIE (6 conditions).
# PosteriorMean averaged across 5 main conditions (excluding DD_ALI).
# Used in Goal 1 to characterize NMD response magnitude.
mashr_model_path <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/mashr/mashr_isoform_model_2026.3.10.rds"
mashr_model <- readRDS(mashr_model_path)
post_mean <- ashr::get_pm(mashr_model)
meta_cols <- c("Smg1i_in_DD", "Smg1i_in_AT", "Smg1i_in_DO_ALI",
               "Smg1i_in_FB", "Smg1i_in_MV")
meta_logfc <- data.frame(
  txid = rownames(post_mean),
  mean_logFC = rowMeans(post_mean[, meta_cols, drop = FALSE])
)

# ── C4 protein comparison (gene-matched, used in productive frameshifts) ──
# Source: 04b_all_c4_protein_comparison.R
# Note: Goal 5 now uses unmatched data from 05d_protein_domain_analysis.R
c4_protein <- readRDS(file.path(cache_dir, "all_c4_protein_comparison_allsamples.rds"))

# ── Productive frameshift precompute ──
# Source: 04_productive_frameshift_precompute.R
# Input: fc_c4_allsamples.rds + fw_c4_allsamples.rds (gene-matched, from script 03)
#        + SQANTI protein FASTA + EBI/biomaRt API queries
# Output: characterization of same-start frameshift C4 pairs without PTC
pf_data <- readRDS(file.path(cache_dir, "productive_frameshift_allsamples.rds"))
```

```{r data-dimensions}
# ==============================================================================
# Summary: Gene-matching and data dimensions
# ==============================================================================
cat("=== Gene-Matching Summary ===\n")
render_dt(gene_match_df, caption = "Gene-Matching: Before and After")

cat(sprintf("\n=== Data Dimensions ===\n"))
cat(sprintf("Gene-matched profile keys: %d\n", length(profiles_list)))
cat(sprintf("C2 all_samples pairs (gene-matched): %d\n", nrow(profiles_list[["c2_allsamples"]])))
cat(sprintf("C4 all_samples pairs (gene-matched): %d\n", nrow(profiles_list[["c4_allsamples"]])))
cat(sprintf("PTC records: %d isoforms\n", nrow(ptc)))
cat(sprintf("CDS records: %d isoforms\n", nrow(cds)))
cat(sprintf("Meta-analysis logFC: %d isoforms\n", nrow(meta_logfc)))
cat(sprintf("C4 protein comparison (gene-matched): %d pairs\n", nrow(c4_protein$comparison)))
cat(sprintf("Productive frameshifts: %d pairs\n", nrow(pf_data$characterization)))

cat("\n=== Cache Verification ===\n")
render_dt(verification_df, caption = "Cache Consistency Check")
```


# Event Type Glossary {.tabset}

Isopair detects 12 structural event types organized into 4 categories. Each tab below
shows a schematic diagram of the event alongside its definition. In all diagrams, the
**Reference** (top) is the dominant non-NMD isoform and the **Comparator** (bottom) is
the isoform being compared. Events are always described from the comparator's perspective:
a **GAIN** means the comparator has additional sequence relative to the reference, and a
**LOSS** means the comparator is missing sequence present in the reference.

**Corrections from the exploratory report (03):** A5SS and A3SS schematics were swapped;
IR_diff_5 and IR_diff_3 schematics showed the retaining exon extending beyond the reference
when the code classifies "diff" when the retaining exon falls short.

```{r glossary, results='asis', fig.width=8, fig.height=3}
glossary_plot <- function(ref_s, ref_e, comp_s, comp_e, title, subtitle,
                          ref_cds = NULL, comp_cds = NULL, events = NULL) {
  plot_ucsc_pair(ref_s, ref_e, comp_s, comp_e, strand = "+",
                 ref_cds = ref_cds, comp_cds = comp_cds,
                 ref_label = "Reference", comp_label = "Comparator",
                 events = events, title = title, subtitle = subtitle)
}

# ---- Terminal Events ----

cat("## Alt_TSS\n\n")
cat("**Alternative Transcription Start Site.** The two isoforms begin at different\n")
cat("genomic positions. Detected when the 5'-most exon of one isoform does not overlap\n")
cat("the 5'-most exon of the other. Direction: GAIN if the comparator's TSS is further\n")
cat("upstream (extending the transcript); LOSS if it is further downstream (truncating it).\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(250, 400, 700), comp_e = c(350, 550, 850),
  title = "Alt_TSS: Alternative Transcription Start Site",
  subtitle = "Comparator starts at a different position (here, downstream = LOSS)")
print(p); cat("\n\n")

cat("## Alt_TES\n\n")
cat("**Alternative Transcription End Site.** The two isoforms end at different\n")
cat("genomic positions. Detected when the 3'-most exon of one isoform does not overlap\n")
cat("the 3'-most exon of the other. Direction: GAIN if the comparator extends further;\n")
cat("LOSS if it ends earlier.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 900),
  comp_s = c(100, 400, 700), comp_e = c(200, 550, 800),
  title = "Alt_TES: Alternative Transcription End Site",
  subtitle = "Comparator ends earlier (LOSS)")
print(p); cat("\n\n")

# ---- Boundary Events ----

cat("## A5SS\n\n")
cat("**Alternative 5' Splice Site.** An internal exon has a different **donor** (exon_end\n")
cat("on + strand) boundary while sharing the same **acceptor** (exon_start) boundary.\n")
cat("The exon is extended or truncated at its downstream/3' end.\n")
cat("Direction: GAIN if the comparator's exon is longer; LOSS if shorter.\n\n")
# A5SS: same acceptor (exon_start=400), different donor (exon_end: 550 vs 600)
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400, 700), comp_e = c(200, 600, 850),
  title = "A5SS: Alternative 5' Splice Site",
  subtitle = "Same acceptor (start), different donor (end). Comparator exon extends further (GAIN)")
print(p); cat("\n\n")

cat("## A3SS\n\n")
cat("**Alternative 3' Splice Site.** An internal exon has a different **acceptor** (exon_start\n")
cat("on + strand) boundary while sharing the same **donor** (exon_end) boundary.\n")
cat("The exon is extended or truncated at its upstream/5' end.\n")
cat("Direction: GAIN if the comparator's exon is longer; LOSS if shorter.\n\n")
# A3SS: different acceptor (exon_start: 400 vs 350), same donor (exon_end=550)
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 350, 700), comp_e = c(200, 550, 850),
  title = "A3SS: Alternative 3' Splice Site",
  subtitle = "Same donor (end), different acceptor (start). Comparator exon extends further (GAIN)")
print(p); cat("\n\n")

cat("## Partial_IR_5\n\n")
cat("**Partial Intron Retention (5' end).** An intron is partially retained: one isoform's\n")
cat("exon extends into the intron from its 5' side, covering what is intronic sequence in\n")
cat("the other isoform. The 3' boundary of the retaining exon falls within the intron.\n")
cat("Direction: GAIN if the comparator retains; LOSS if the reference retains.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400, 700), comp_e = c(200, 620, 850),
  title = "Partial_IR_5: Partial Intron Retention (5' side)",
  subtitle = "Comparator exon extends into the intron from the 5' end (GAIN)")
print(p); cat("\n\n")

cat("## Partial_IR_3\n\n")
cat("**Partial Intron Retention (3' end).** An intron is partially retained: one isoform's\n")
cat("exon extends into the intron from its 3' side, covering what is intronic sequence in\n")
cat("the other isoform. The 5' boundary of the retaining exon falls within the intron.\n")
cat("Direction: GAIN if the comparator retains; LOSS if the reference retains.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400, 640), comp_e = c(200, 550, 850),
  title = "Partial_IR_3: Partial Intron Retention (3' side)",
  subtitle = "Comparator exon extends into the intron from the 3' end (GAIN)")
print(p); cat("\n\n")

# ---- Exon-Level Events ----

cat("## SE (Skipped Exon)\n\n")
cat("**Skipped Exon.** An exon present in one isoform is completely absent from the\n")
cat("other. The flanking exons splice directly to each other, skipping over the exon.\n")
cat("Direction: GAIN if the comparator includes the exon (reference skips it);\n")
cat("LOSS if the comparator skips it (reference includes it).\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 600, 800), ref_e  = c(200, 500, 700, 900),
  comp_s = c(100, 400, 800),       comp_e = c(200, 500, 900),
  title = "SE: Skipped Exon",
  subtitle = "Comparator skips an internal exon present in the reference (LOSS)")
print(p); cat("\n\n")

cat("## Missing_Internal\n\n")
cat("**Missing Internal Exon.** Similar to SE but arising from a multi-exon skip:\n")
cat("one isoform is missing one or more internal exons that are present in the other.\n")
cat("Often co-occurs with other Missing_Internal events from the same splice decision.\n")
cat("Direction: GAIN if the comparator has the exon; LOSS if the comparator lacks it.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 350, 550, 750, 950), ref_e  = c(200, 450, 650, 850, 1050),
  comp_s = c(100, 950),                 comp_e = c(200, 1050),
  title = "Missing_Internal: Missing Internal Exon(s)",
  subtitle = "Comparator skips multiple internal exons (LOSS x 3)")
print(p); cat("\n\n")

# ---- Intron Retention Events ----

cat("## IR\n\n")
cat("**Full Intron Retention.** One isoform has a single exon that spans the region\n")
cat("covered by two or more exons (plus the intervening intron) in the other isoform.\n")
cat("The retaining exon boundaries match exactly. This is a full retention where the\n")
cat("entire intron is included in the retaining isoform's exon.\n")
cat("Direction: GAIN if the comparator retains the intron; LOSS if the reference retains it.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400),       comp_e = c(200, 850),
  title = "IR: Full Intron Retention",
  subtitle = "Comparator retains intron -- single exon spans two exons + intron in ref (GAIN)")
print(p); cat("\n\n")

cat("## IR_diff_5\n\n")
cat("**Intron Retention with 5' Boundary Difference.** Like IR, but the retaining\n")
cat("exon's 5' boundary differs from the first overlapping exon in the other isoform.\n")
cat("The intron is retained, but the retaining exon starts after the first reference\n")
cat("overlap exon (falls short at the 5' end).\n\n")
# IR_diff_5: comp retaining exon starts AFTER first ref overlap exon start
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 430),       comp_e = c(200, 850),
  title = "IR_diff_5: Intron Retention with 5' Boundary Difference",
  subtitle = "Retaining exon starts AFTER ref overlap exon (falls short at 5' end)")
print(p); cat("\n\n")

cat("## IR_diff_3\n\n")
cat("**Intron Retention with 3' Boundary Difference.** Like IR, but the retaining\n")
cat("exon's 3' boundary differs from the last overlapping exon in the other isoform.\n")
cat("The intron is retained, but the retaining exon ends before the last reference\n")
cat("overlap exon (falls short at the 3' end).\n\n")
# IR_diff_3: comp retaining exon ends BEFORE last ref overlap exon end
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 400),       comp_e = c(200, 820),
  title = "IR_diff_3: Intron Retention with 3' Boundary Difference",
  subtitle = "Retaining exon ends BEFORE ref overlap exon (falls short at 3' end)")
print(p); cat("\n\n")

cat("## IR_diff_5_3\n\n")
cat("**Intron Retention with Both Boundary Differences.** Like IR, but both the 5'\n")
cat("and 3' boundaries of the retaining exon differ from the flanking exons in the\n")
cat("other isoform. The intron is retained but the retaining exon falls short at\n")
cat("both ends.\n\n")
p <- glossary_plot(
  ref_s  = c(100, 400, 700), ref_e  = c(200, 550, 850),
  comp_s = c(100, 430),       comp_e = c(200, 820),
  title = "IR_diff_5_3: Intron Retention with Both Boundary Differences",
  subtitle = "Retaining exon falls short at both 5' and 3' ends")
print(p); cat("\n\n")
```

In the IR section of this report, **IR_diff** refers to the aggregate of IR_diff_5,
IR_diff_3, and IR_diff_5_3 events.


# 1. Transcriptional Diversity and Isoform Optionality {#sec1-diversity}

Before asking what triggers NMD, we establish the transcriptional landscape
that NMD operates on. Genes produce diverse isoforms through combinatorial
TSS, internal splicing, and TES choices. This section characterizes
isoform diversity within the gene-matched pair set — the same genes analyzed
throughout this report. Diversity and TSS/TES analyses use all expressed
isoforms (coding and non-coding); CDS-dependent analyses are restricted to
coding isoforms.

**Data source:** All expressed isoforms for genes in the gene-matched pair
set (structures.rds, expression_data.rds). CDS analyses additionally use
cds.rds (coding isoforms only).

```{r sec1-diversity-data}
# ── Data provenance ──
# Source: structures.rds (extractIsoformStructures), cds.rds (extractCds),
#   expression_data.rds (TMM-normalized count matrix from 01_prepare_data_mashr.R)
# Population: All expressed isoforms for genes in gene-matched pair set.
#   Gene set defined by profiles_list (loaded in STAGE 2, gene-matched).

# Genes in the gene-matched pair set
pair_genes <- unique(profiles_list[["c2_allsamples"]]$gene_id)

# All expressed isoforms for these genes (coding + non-coding)
sec1_structures <- structures %>%
  filter(gene_id %in% pair_genes,
         isoform_id %in% rownames(expr_mat))

# Coding subset (for CDS-dependent analyses later)
coding_structures <- sec1_structures %>%
  filter(isoform_id %in% cds$isoform_id[cds$coding_status == "coding"])

coding_cds <- cds %>%
  filter(isoform_id %in% coding_structures$isoform_id)

cat("Gene-matched pair genes:", length(pair_genes), "\n")
cat("All expressed isoforms (these genes):", nrow(sec1_structures), "\n")
cat("  of which coding:", nrow(coding_structures), "\n")

# ── Isoform count per gene (all isoforms) ──
iso_per_gene <- sec1_structures %>%
  group_by(gene_id) %>%
  summarise(n_isoforms = n(), .groups = "drop")

cat("\nIsoforms per gene (all expressed):\n")
print(summary(iso_per_gene$n_isoforms))
cat("Genes with ≥2 isoforms:", sum(iso_per_gene$n_isoforms >= 2), "\n")
cat("Genes with ≥5 isoforms:", sum(iso_per_gene$n_isoforms >= 5), "\n")
```

## Isoform Count Distribution

```{r sec1-isoform-count-dist, fig.width=8, fig.height=5}
fig_iso_count <- ggplot(iso_per_gene, aes(x = n_isoforms)) +
  geom_histogram(binwidth = 1, fill = "#4292c6", color = "white", alpha = 0.8) +
  scale_x_continuous(limits = c(0, 30), breaks = seq(0, 30, 5)) +
  labs(title = "Expressed Isoform Count per Gene",
       subtitle = sprintf("Gene-matched pair set | Median: %d isoforms | %s genes with ≥2",
                          median(iso_per_gene$n_isoforms),
                          format(sum(iso_per_gene$n_isoforms >= 2), big.mark = ",")),
       x = "Number of expressed isoforms", y = "Number of genes")
save_fig(fig_iso_count, "fig0a_isoform_count_distribution")
```

## TSS and TES Diversity

Isoform diversity arises from three combinatorial axes: which transcription
start site (TSS) is used, which internal exons are included, and which
transcription end site (TES) is used. We quantify diversity at each axis
and ask whether TSS and TES choices are coupled or independent.

```{r sec1-tss-tes-diversity, fig.width=10, fig.height=8}
# Multi-isoform genes only (≥2 coding isoforms)
multi_iso_genes <- iso_per_gene$gene_id[iso_per_gene$n_isoforms >= 2]

# ── Per-gene diversity counts (all isoforms, not just coding) ──
tss_tes <- sec1_structures %>%
  filter(gene_id %in% multi_iso_genes) %>%
  group_by(gene_id, strand) %>%
  summarise(
    n_isoforms = n(),
    n_unique_tss = n_distinct(ifelse(strand == "+", tx_start, tx_end)),
    n_unique_tes = n_distinct(ifelse(strand == "+", tx_end, tx_start)),
    .groups = "drop"
  )

cat("TSS diversity (multi-isoform coding genes, n =", nrow(tss_tes), "):\n")
cat("  Median distinct TSS per gene:", median(tss_tes$n_unique_tss), "\n")
cat("  Median distinct TES per gene:", median(tss_tes$n_unique_tes), "\n")

# ── Within-gene TSS-TES independence ──
# The key question: for a given gene, does knowing which TSS is used
# predict which TES is used? We compute per-gene Spearman rho between
# TSS and TES positions across isoforms.
within_gene_cor <- sec1_structures %>%
  filter(gene_id %in% multi_iso_genes) %>%
  mutate(
    tss = ifelse(strand == "+", tx_start, tx_end),
    tes = ifelse(strand == "+", tx_end, tx_start)
  ) %>%
  group_by(gene_id) %>%
  filter(n() >= 3, n_distinct(tss) >= 2, n_distinct(tes) >= 2) %>%
  summarise(
    n_iso = n(),
    rho_tss_tes = cor(tss, tes, method = "spearman"),
    .groups = "drop"
  )

within_gene_rho <- median(within_gene_cor$rho_tss_tes, na.rm = TRUE)
cat("\nWithin-gene TSS-TES coupling (n =", nrow(within_gene_cor), "genes with ≥3 iso):\n")
cat("  Median Spearman rho:", round(within_gene_rho, 3), "\n")
cat("  Mean Spearman rho:", round(mean(within_gene_cor$rho_tss_tes, na.rm = TRUE), 3), "\n")
cat("  IQR:", round(quantile(within_gene_cor$rho_tss_tes, 0.25, na.rm = TRUE), 3),
    "to", round(quantile(within_gene_cor$rho_tss_tes, 0.75, na.rm = TRUE), 3), "\n")

# ── Combination usage ──
# If TSS and TES were fully independent and all combinations viable,
# genes would use ~100% of TSS×TES pairs. Selective usage indicates
# the exon path constrains which TSS-TES pairings are realized.
tss_tes_combos <- sec1_structures %>%
  filter(gene_id %in% multi_iso_genes) %>%
  mutate(
    tss = ifelse(strand == "+", tx_start, tx_end),
    tes = ifelse(strand == "+", tx_end, tx_start)
  ) %>%
  group_by(gene_id) %>%
  summarise(
    n_tss = n_distinct(tss),
    n_tes = n_distinct(tes),
    n_combos = n_distinct(paste(tss, tes)),
    max_possible = n_distinct(tss) * n_distinct(tes),
    pct_used = round(100 * n_distinct(paste(tss, tes)) /
                       (n_distinct(tss) * n_distinct(tes)), 1),
    .groups = "drop"
  ) %>%
  filter(max_possible > 1)

cat("\nTSS×TES combination usage (n =", nrow(tss_tes_combos), "genes):\n")
cat("  Median % of possible combinations used:", round(median(tss_tes_combos$pct_used), 1), "%\n")
cat("  Mean:", round(mean(tss_tes_combos$pct_used), 1), "%\n")

# ── Figures ──

# Panel A: TSS and TES count distributions
tss_tes_long <- tss_tes %>%
  select(gene_id, TSS = n_unique_tss, TES = n_unique_tes) %>%
  pivot_longer(-gene_id, names_to = "feature", values_to = "count")

p_diversity <- ggplot(tss_tes_long, aes(x = count, fill = feature)) +
  geom_histogram(binwidth = 1, position = "dodge", alpha = 0.7) +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, 5)) +
  scale_fill_manual(values = c(TSS = "#ef8a62", TES = "#67a9cf")) +
  labs(title = "A. TSS and TES Diversity per Gene",
       subtitle = sprintf("Multi-isoform coding genes (n = %s)",
                          format(nrow(tss_tes), big.mark = ",")),
       x = "Number of distinct positions", y = "Number of genes", fill = "")

# Panel B: Within-gene TSS-TES correlation distribution
p_within <- ggplot(within_gene_cor, aes(x = rho_tss_tes)) +
  geom_histogram(binwidth = 0.05, fill = "#756bb1", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = within_gene_rho, color = "#e6550d", linewidth = 1) +
  annotate("text", x = within_gene_rho + 0.05, y = Inf, vjust = 2,
           label = sprintf("median = %.3f", within_gene_rho),
           color = "#e6550d", size = 3.5, hjust = 0) +
  labs(title = "B. Within-Gene TSS-TES Coupling",
       subtitle = sprintf("Per-gene Spearman rho (n = %s genes with ≥3 isoforms)",
                          format(nrow(within_gene_cor), big.mark = ",")),
       x = "Spearman rho (TSS position vs TES position)", y = "Number of genes")

# Panel C: Combination usage
p_combos <- ggplot(tss_tes_combos, aes(x = pct_used)) +
  geom_histogram(binwidth = 5, fill = "#31a354", color = "white", alpha = 0.8) +
  geom_vline(xintercept = median(tss_tes_combos$pct_used), color = "#e6550d", linewidth = 1) +
  annotate("text", x = median(tss_tes_combos$pct_used) + 2, y = Inf, vjust = 2,
           label = sprintf("median = %.0f%%", median(tss_tes_combos$pct_used)),
           color = "#e6550d", size = 3.5, hjust = 0) +
  labs(title = "C. TSS × TES Combination Usage",
       subtitle = "% of possible TSS-TES pairs actually observed",
       x = "Combinations used (%)", y = "Number of genes")

fig_tss_tes <- p_diversity / (p_within | p_combos)
save_fig(fig_tss_tes, "fig0b_tss_tes_diversity", width = 10, height = 8)
```

Within any given gene, the TSS and TES positions are largely independent:
the median within-gene Spearman rho is `r round(within_gene_rho, 3)`,
centered tightly around zero. Knowing which promoter a transcript initiates
from does not predict where it will terminate.

Yet genes use only `r round(median(tss_tes_combos$pct_used))`% of the
TSS×TES combinations that would be possible if every start could pair with
every end. This apparent contradiction resolves when we consider the exon
path: internal splicing decisions constrain which TSS-TES pairings produce
viable transcripts. TSS and TES are chosen independently of *each other*,
but the internal exon structure acts as a filter — not all combinations
produce stable, processable transcripts.

## TSS–CDS and TES–CDS Coupling

```{r sec1-tss-cds-coupling}
# TSS determines CDS start; TES determines CDS stop
coupling <- coding_structures %>%
  filter(gene_id %in% multi_iso_genes) %>%
  inner_join(coding_cds %>% select(isoform_id, cds_start, cds_stop),
             by = "isoform_id") %>%
  mutate(
    tss = ifelse(strand == "+", tx_start, tx_end),
    tes = ifelse(strand == "+", tx_end, tx_start),
    cds_5prime = ifelse(strand == "+", cds_start, cds_stop),
    cds_3prime = ifelse(strand == "+", cds_stop, cds_start)
  )

# Per-gene correlations (genes with ≥3 isoforms for meaningful correlation)
gene_cors <- coupling %>%
  group_by(gene_id) %>%
  filter(n() >= 3) %>%
  summarise(
    n_iso = n(),
    rho_tss_cds5 = cor(tss, cds_5prime, method = "spearman"),
    rho_tes_cds3 = cor(tes, cds_3prime, method = "spearman"),
    .groups = "drop"
  )

cat("TSS → CDS start coupling (per-gene Spearman rho):\n")
cat("  Median:", round(median(gene_cors$rho_tss_cds5, na.rm = TRUE), 3), "\n")
cat("  Mean:", round(mean(gene_cors$rho_tss_cds5, na.rm = TRUE), 3), "\n")

cat("\nTES → CDS stop coupling:\n")
cat("  Median:", round(median(gene_cors$rho_tes_cds3, na.rm = TRUE), 3), "\n")
cat("  Mean:", round(mean(gene_cors$rho_tes_cds3, na.rm = TRUE), 3), "\n")
```

The TSS strongly determines the CDS start position, and the TES strongly
determines the CDS stop position. The two ends of the protein are set by
the two ends of the transcript — through the exon path, not directly.

## CDS Start Diversity and Reading Frame Multiplicity

```{r sec1-cds-start-diversity}
# How many distinct CDS start positions per gene? (coding isoforms only)
cds_diversity <- coding_structures %>%
  filter(gene_id %in% multi_iso_genes) %>%
  inner_join(coding_cds %>% select(isoform_id, cds_start, cds_stop),
             by = "isoform_id") %>%
  mutate(cds_5prime = ifelse(strand == "+", cds_start, cds_stop)) %>%
  group_by(gene_id) %>%
  filter(n() >= 2) %>%
  summarise(
    n_isoforms = n(),
    n_cds_starts = n_distinct(cds_5prime),
    cds_starts = list(unique(cds_5prime)),
    .groups = "drop"
  )

cat("CDS start diversity (multi-coding-isoform genes, n =", nrow(cds_diversity), "):\n")
cat("  Genes with ≥2 distinct CDS starts:",
    sum(cds_diversity$n_cds_starts >= 2),
    sprintf("(%.1f%%)\n", 100 * mean(cds_diversity$n_cds_starts >= 2)))
cat("  Genes with ≥3 distinct CDS starts:",
    sum(cds_diversity$n_cds_starts >= 3),
    sprintf("(%.1f%%)\n", 100 * mean(cds_diversity$n_cds_starts >= 3)))
cat("  Median distinct CDS starts:", median(cds_diversity$n_cds_starts), "\n")

# Of genes with multiple CDS starts, how many use different reading frames?
# Two CDS starts are in different frames if their positions differ by a non-multiple of 3.
cds_diversity$has_diff_frame <- sapply(cds_diversity$cds_starts, function(s) {
  if (length(s) < 2) return(FALSE)
  diffs <- outer(s, s, "-")
  diffs <- diffs[upper.tri(diffs)]
  any(diffs %% 3 != 0)
})

multi_start <- cds_diversity %>% filter(n_cds_starts >= 2)
cat("\nReading frame diversity (genes with ≥2 CDS starts, n =", nrow(multi_start), "):\n")
cat("  Genes with starts in DIFFERENT frames:",
    sum(multi_start$has_diff_frame),
    sprintf("(%.1f%%)\n", 100 * mean(multi_start$has_diff_frame)))
cat("  Genes with all starts in SAME frame:",
    sum(!multi_start$has_diff_frame),
    sprintf("(%.1f%%)\n", 100 * mean(!multi_start$has_diff_frame)))
```

`r sprintf("%.1f%%", 100 * mean(cds_diversity$n_cds_starts >= 2))` of genes
with multiple coding isoforms use at least two distinct CDS start positions.
Among these, `r sprintf("%.1f%%", 100 * mean(multi_start$has_diff_frame))`
use CDS starts in different reading frames — the start positions differ by a
non-multiple of 3, meaning the downstream protein sequences are in different
frames. These genes encode genuinely distinct reading frames through their
isoform repertoire, not merely N-terminal extensions of the same protein.

## Expression Concentration

```{r sec1-expression-concentration}
# How much of a gene's output comes from its dominant isoform?
# Use DMSO samples for baseline expression
dmso_samples <- sample_meta$sample_id[sample_meta$treatment == "DMSO"]
dmso_mat <- expr_mat[, colnames(expr_mat) %in% dmso_samples]

# CPM per isoform
lib_sizes <- colSums(dmso_mat)
cpm_mat <- sweep(dmso_mat, 2, lib_sizes, "/") * 1e6

# Per-gene dominant fraction
gene_ids <- coding_structures$gene_id[match(rownames(cpm_mat), coding_structures$isoform_id)]
valid <- !is.na(gene_ids) & rownames(cpm_mat) %in% coding_structures$isoform_id

mean_cpm <- rowMeans(cpm_mat[valid, ])
iso_df <- data.frame(
  isoform_id = rownames(cpm_mat)[valid],
  gene_id = gene_ids[valid],
  mean_cpm = mean_cpm,
  stringsAsFactors = FALSE
)

dominant_fraction <- iso_df %>%
  group_by(gene_id) %>%
  filter(n() >= 2, sum(mean_cpm) > 0) %>%
  summarise(
    n_iso = n(),
    total_cpm = sum(mean_cpm),
    max_cpm = max(mean_cpm),
    dominant_pct = 100 * max_cpm / total_cpm,
    .groups = "drop"
  )

cat("Expression concentration (DMSO, multi-isoform coding genes):\n")
cat("  Median dominant isoform share:", round(median(dominant_fraction$dominant_pct), 1), "%\n")
cat("  Genes where dominant >50%:", sum(dominant_fraction$dominant_pct > 50),
    sprintf("(%.1f%%)\n", 100 * mean(dominant_fraction$dominant_pct > 50)))
cat("  Genes where dominant >80%:", sum(dominant_fraction$dominant_pct > 80),
    sprintf("(%.1f%%)\n", 100 * mean(dominant_fraction$dominant_pct > 80)))
```

Transcriptional diversity is the norm. Genes maintain multiple exon paths,
each producing a different reading frame context. Most genes concentrate
expression on a dominant isoform, but devote substantial transcriptional
output to alternatives — including paths that produce NMD substrates.

# 1b. Baseline Splicing and NMD-Enriched Events {#goal1b-baseline}

Characterize baseline alternative splicing (Control C4 pairs) and identify which
splicing event types are enriched in NMD vs Control.

**Methods:** Structural profiles quantify splicing event types for each pair. Sequence
divergence (`quantifyPairDivergence()`) measures exonic sequence shared between isoforms.
NMD vs Control enrichment is tested by Fisher's exact test per event type.

## Figure 1: Sequence Shared — NMD vs Control

```{r goal1-fig1-pct-shared, fig.width=8, fig.height=5}
# ── Data provenance ──
# Starting data: gene-matched profiles → cached divergence
# div_c2 = set_results[["c2_allsamples"]]$divergence
#   Source: quantifyPairDivergence(gene-matched profiles, structures, cds),
#   cached by script 03
#   $pct_shared: percentage of exonic sequence shared between ref and comp
# div_c4: same for Control
# Result: density comparison of sequence similarity between NMD and Control pairs

pct_shared_data <- rbind(
  data.frame(comparison = "NMD",     pct_shared = div_c2$pct_shared),
  data.frame(comparison = "Control", pct_shared = div_c4$pct_shared))
pct_shared_data <- pct_shared_data[!is.na(pct_shared_data$pct_shared), ]

fig1_shared <- ggplot(pct_shared_data, aes(x = pct_shared, fill = comparison)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = pal) +
  labs(title = "Exonic Sequence Shared (%): NMD vs Control",
       subtitle = "NMD pairs shifted right = more similar (targeted splicing changes)",
       x = "Sequence Shared (%)", y = "Density", fill = "")
save_fig(fig1_shared, "fig1_pct_shared_density")

# Summary stats
shared_summary <- pct_shared_data |>
  group_by(comparison) |>
  summarise(n = n(), median = round(median(pct_shared), 1),
            mean = round(mean(pct_shared), 1), .groups = "drop")
render_dt(shared_summary, caption = "Table 1a: Sequence Shared Summary")
```

Columns: **comparison** = NMD or Control; **n** = number of pairs; **median/mean** =
median and mean percentage of exonic sequence shared between reference and
comparator (from `Isopair::quantifyPairDivergence()`).

## Table 1b: Splicing Event Enrichment — NMD vs Control

For each event type, the denominator is all gene-matched pairs (not restricted
to coding) — a pair is counted if it has ≥1 event of that type. Fisher's exact
test compares the proportion of pairs with each event type between NMD (C2) and
Control (C4), both on the same gene-matched set.

```{r goal1-table1b-event-enrichment}
# ── Data provenance ──
# Starting data: gene-matched profiles (set_results[[key]]$profiles)
# compute_event_freq(profiles) counts pairs with each event type using:
#   - Boolean columns (tss_changed, tes_changed, n_a5ss > 0, etc.) from profiles
#   - Partial_IR_5/3 from detailed_events list-column
#   These are computed by buildProfiles → compareSplicingChoices
# Fisher's exact test: event prevalence in NMD vs Control on gene-matched sets

ef_c2 <- compute_event_freq(set_results[["c2_allsamples"]]$profiles)
ef_c4 <- compute_event_freq(set_results[["c4_allsamples"]]$profiles)
n_c2 <- nrow(set_results[["c2_allsamples"]]$profiles)
n_c4 <- nrow(set_results[["c4_allsamples"]]$profiles)

enrichment <- merge(ef_c2, ef_c4, by = "event_type", suffixes = c("_NMD", "_Ctrl"))

# Fisher's exact test per event type
enrichment$fisher_OR <- NA_real_
enrichment$fisher_p <- NA_real_
for (i in seq_len(nrow(enrichment))) {
  a <- enrichment$n_pairs_with_event_NMD[i]
  b <- n_c2 - a
  c <- enrichment$n_pairs_with_event_Ctrl[i]
  d <- n_c4 - c
  ft <- tryCatch(fisher.test(matrix(c(a, b, c, d), nrow = 2)), error = function(e) NULL)
  if (!is.null(ft)) {
    enrichment$fisher_OR[i] <- round(ft$estimate, 2)
    enrichment$fisher_p[i] <- ft$p.value
  }
}

enrichment$direction <- ifelse(enrichment$fisher_p > 0.05, "Non-significant",
  ifelse(enrichment$fisher_OR > 1, "Enriched in NMD", "Enriched in Control"))
enrichment <- enrichment[order(enrichment$fisher_p), ]

enrich_display <- enrichment[, c("event_type", "n_pairs_with_event_NMD", "pct_of_pairs_NMD",
                                 "n_pairs_with_event_Ctrl", "pct_of_pairs_Ctrl",
                                 "fisher_OR", "fisher_p", "direction")]
enrich_display$fisher_p <- signif(enrich_display$fisher_p, 3)
render_dt(enrich_display,
          caption = "Table 1b: Splicing Event Type Enrichment — NMD vs Control")
write.csv(enrichment, file.path(table_dir, "table1b_event_enrichment.csv"), row.names = FALSE)
```

Columns: **event_type** = splicing event type (see Event Type Glossary);
**n_pairs_with_event_NMD** = number of gene-matched NMD (C2) pairs with ≥1
event of this type; **pct_of_pairs_NMD** = percentage of all
`r fmt(n_c2)` NMD pairs; **n_pairs_with_event_Ctrl/pct_of_pairs_Ctrl** =
same for Control (C4, `r fmt(n_c4)` pairs); **fisher_OR** = Fisher's exact
test odds ratio (>1 = NMD-enriched); **fisher_p** = p-value; **direction** =
Enriched in NMD, Enriched in Control, or Non-significant (p > 0.05).

### Figure 1b: Splicing Event Prevalence — NMD vs Control

```{r goal1-fig1b-event-barplot, fig.width=10, fig.height=6}
# ── Data provenance ──
# enrichment: from goal1-table1b-event-enrichment chunk
# Paired barplot of event prevalence (% of pairs) by comparison type

bar_df <- rbind(
  data.frame(event_type = enrichment$event_type,
             comparison = "NMD",
             pct = enrichment$pct_of_pairs_NMD),
  data.frame(event_type = enrichment$event_type,
             comparison = "Control",
             pct = enrichment$pct_of_pairs_Ctrl))

# Order events by NMD prevalence (descending)
evt_order <- enrichment$event_type[order(-enrichment$pct_of_pairs_NMD)]
bar_df$event_type <- factor(bar_df$event_type, levels = evt_order)

# Significance stars
sig_df <- data.frame(
  event_type = factor(enrichment$event_type, levels = evt_order),
  y = pmax(enrichment$pct_of_pairs_NMD, enrichment$pct_of_pairs_Ctrl) + 2,
  label = ifelse(enrichment$fisher_p < 0.001, "***",
           ifelse(enrichment$fisher_p < 0.01, "**",
             ifelse(enrichment$fisher_p < 0.05, "*", "ns"))))

fig1b <- ggplot(bar_df, aes(x = event_type, y = pct, fill = comparison)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(data = sig_df, aes(x = event_type, y = y, label = label),
            inherit.aes = FALSE, size = 4, fontface = "bold") +
  scale_fill_manual(values = pal) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(title = "Splicing Event Prevalence: NMD vs Control",
       subtitle = sprintf("NMD: %s pairs, Control: %s pairs. Significance: Fisher's exact test (* p<0.05, ** p<0.01, *** p<0.001).",
                           fmt(n_c2), fmt(n_c4)),
       x = "", y = "% of Pairs with Event", fill = "Comparison") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

save_fig(fig1b, "fig1b_event_prevalence_barplot", width = 10, height = 6)
```

## Table 1c: Gain/Loss Direction by Event Type

For each event type, the comparator can GAIN sequence (comparator has additional
sequence relative to reference) or LOSE sequence (comparator is missing sequence
present in reference). The table below shows event counts and the percentage of
events in each direction, separately for NMD and Control pairs.

```{r goal1-table1c-gain-loss}
# ── Data provenance ──
# Starting data: gene-matched profiles (set_results[[key]]$profiles)
# $detailed_events: list-column, one data.frame per pair from buildProfiles
#   Each event row has event_type and direction (GAIN or LOSS)
#   Direction is from comparator's perspective (see Event Type Glossary)
# For each comparison (NMD, Control): count events by type × direction,
#   compute % GAIN and % LOSS within each event type

get_gain_loss_pct <- function(profiles, label) {
  all_events <- do.call(rbind, profiles$detailed_events)
  if (is.null(all_events) || nrow(all_events) == 0) return(NULL)
  gl <- all_events |>
    group_by(event_type, direction) |>
    summarise(n = n(), .groups = "drop")
  totals <- gl |> group_by(event_type) |> summarise(total = sum(n), .groups = "drop")
  gl <- merge(gl, totals, by = "event_type")
  gl$pct <- round(100 * gl$n / gl$total, 1)
  gl$comparison <- label
  gl
}

gl_c2 <- get_gain_loss_pct(set_results[["c2_allsamples"]]$profiles, "NMD")
gl_c4 <- get_gain_loss_pct(set_results[["c4_allsamples"]]$profiles, "Control")
gl_all <- rbind(gl_c2, gl_c4)

gl_wide <- gl_all |>
  tidyr::pivot_wider(
    id_cols = event_type,
    names_from = c(comparison, direction),
    values_from = c(n, pct),
    values_fill = 0) |>
  arrange(event_type)

col_order <- c("event_type",
  "n_NMD_GAIN", "n_NMD_LOSS", "pct_NMD_GAIN",
  "n_Control_GAIN", "n_Control_LOSS", "pct_Control_GAIN")
existing_cols <- intersect(col_order, names(gl_wide))
gl_display <- gl_wide[, existing_cols, drop = FALSE]

gl_display$fisher_p <- NA_real_
gl_display$direction_diff <- NA_character_
for (i in seq_len(nrow(gl_display))) {
  ng <- gl_display$n_NMD_GAIN[i]
  nl <- gl_display$n_NMD_LOSS[i]
  cg <- gl_display$n_Control_GAIN[i]
  cl <- gl_display$n_Control_LOSS[i]
  if (ng + nl > 0 && cg + cl > 0) {
    ft <- tryCatch(fisher.test(matrix(c(ng, nl, cg, cl), nrow = 2)), error = function(e) NULL)
    if (!is.null(ft)) {
      gl_display$fisher_p[i] <- signif(ft$p.value, 3)
      pct_nmd <- 100 * ng / (ng + nl)
      pct_ctrl <- 100 * cg / (cg + cl)
      gl_display$direction_diff[i] <- if (ft$p.value > 0.05) "Non-significant"
        else if (pct_nmd > pct_ctrl) "More GAIN in NMD"
        else "More LOSS in NMD"
    }
  }
}

render_dt(gl_display, caption = "Table 1c: Gain/Loss Events by Type — NMD vs Control")
write.csv(gl_display, file.path(table_dir, "table1c_gain_loss_events.csv"), row.names = FALSE)
```

Columns: **event_type** = splicing event type; **n_NMD_GAIN/n_NMD_LOSS** =
count of NMD events with GAIN/LOSS direction (from comparator's perspective);
**pct_NMD_GAIN** = percentage of NMD events of that type with GAIN direction;
**n_Control_GAIN/n_Control_LOSS/pct_Control_GAIN** = same for Control;
**fisher_p** = Fisher's exact test p-value comparing GAIN/LOSS ratio between
NMD and Control; **direction_diff** = "More GAIN in NMD", "More LOSS in NMD",
or "Non-significant" (p > 0.05).

### Figure 1c: GAIN Direction by Event Type — NMD vs Control

```{r goal1-fig1c-gain-loss-barplot, fig.width=10, fig.height=6}
# ── Data provenance ──
# gl_display: from goal1-table1c-gain-loss chunk
# Paired barplot of % GAIN direction by event type, NMD vs Control

bar1c <- rbind(
  data.frame(event_type = gl_display$event_type,
             comparison = "NMD",
             pct_gain = gl_display$pct_NMD_GAIN),
  data.frame(event_type = gl_display$event_type,
             comparison = "Control",
             pct_gain = gl_display$pct_Control_GAIN))

# Order events by NMD % GAIN (descending)
evt_order_1c <- gl_display$event_type[order(-gl_display$pct_NMD_GAIN)]
bar1c$event_type <- factor(bar1c$event_type, levels = evt_order_1c)

# Significance stars
sig1c <- data.frame(
  event_type = factor(gl_display$event_type, levels = evt_order_1c),
  y = pmax(gl_display$pct_NMD_GAIN, gl_display$pct_Control_GAIN) + 3,
  label = ifelse(is.na(gl_display$fisher_p), "",
           ifelse(gl_display$fisher_p < 0.001, "***",
             ifelse(gl_display$fisher_p < 0.01, "**",
               ifelse(gl_display$fisher_p < 0.05, "*", "ns")))))

fig1c <- ggplot(bar1c, aes(x = event_type, y = pct_gain, fill = comparison)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(data = sig1c, aes(x = event_type, y = y, label = label),
            inherit.aes = FALSE, size = 4, fontface = "bold") +
  scale_fill_manual(values = pal) +
  scale_y_continuous(limits = c(0, 105), expand = expansion(mult = c(0, 0))) +
  labs(title = "GAIN Direction by Event Type: NMD vs Control",
       subtitle = "% of events where comparator GAINS sequence relative to reference. Significance: Fisher's exact test.",
       x = "", y = "% Events with GAIN Direction", fill = "Comparison") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

save_fig(fig1c, "fig1c_gain_direction_barplot", width = 10, height = 6)
```

# 2a. Predicting NMD from Isoform Structure — The Unified Model {#sec2a-unified}

Can we predict which isoforms trigger NMD from structural features alone?
We build a model progressively, adding feature categories and measuring
improvement. This overview presents the unified model; subsequent sections
drill into each feature category.

**Data source:** All mashr-classified NMD and non-NMD coding isoforms with
complete structural features (broader than the paired framework). The model
asks "can structural features predict NMD status?" — an isoform-level
question, not a pair-level question.

**Statistical approach:** Elastic net logistic regression (alpha = 0.5) with
10-fold cross-validation on the training set. Holdout chromosomes 1, 3, 5, 7
(~26% of data) are never used for feature selection or model tuning.

```{r sec2a-load-model}
# ── Data provenance ──
# Source: 05l_unified_model.R → unified_model.rds
# Population: ALL mashr-classified NMD/non-NMD coding isoforms with complete
#   structural features (PTC from ptc.rds, 5'UTR from utr5_features_all.rds)
# Train/test: holdout chr 1, 3, 5, 7 (~26%)

unified <- readRDS(file.path(cache_dir, "unified_model.rds"))

cat("Population:", format(unified$metadata$population, big.mark = ","),
    "isoforms with complete features\n")
cat("NMD:", sum(unified$df$is_nmd),
    "| Non-NMD:", sum(!unified$df$is_nmd), "\n")
cat("Holdout chromosomes:", paste(unified$metadata$holdout_chrs, collapse = ", "), "\n\n")

# Feature name mapping for display
feat_labels <- unified$feature_labels
```

## Progressive ROC Curves

```{r sec2a-progressive-roc, fig.width=7, fig.height=7}
# ── Data provenance ──
# Source: unified$elastic_net (elastic net models fit on training data)
# ROC curves from stored roc_test objects (holdout predictions)

en_models <- list(
  list(roc = unified$elastic_net$step1$roc_test,
       name = "Step 1: Stop codon position", color = "#fc8d59"),
  list(roc = unified$elastic_net$step2$roc_test,
       name = "Step 2: + Start codon context", color = "#91bfdb"),
  list(roc = unified$elastic_net$step3$roc_test,
       name = "Step 3: + ORF features (unified)", color = "#1a1a1a")
)

roc_curves <- lapply(en_models, function(spec) {
  auc_val <- as.numeric(auc(spec$roc))
  data.frame(
    fpr = 1 - spec$roc$specificities,
    tpr = spec$roc$sensitivities,
    model = sprintf("%s (AUC = %.3f)", spec$name, auc_val),
    stringsAsFactors = FALSE
  )
})
roc_all <- do.call(rbind, roc_curves)
roc_all$model <- factor(roc_all$model, levels = unique(roc_all$model))

model_colors <- setNames(
  sapply(en_models, `[[`, "color"),
  levels(roc_all$model)
)

fig_roc <- ggplot(roc_all, aes(x = fpr, y = tpr, color = model)) +
  geom_abline(linetype = "dashed", color = "grey70") +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = model_colors) +
  coord_equal() +
  labs(title = "Holdout ROC — Progressive Model Building",
       subtitle = sprintf("Holdout: chr 1, 3, 5, 7 (N = %s)",
                          format(unified$elastic_net$step3$n_test, big.mark = ",")),
       x = "False Positive Rate", y = "True Positive Rate",
       color = "") +
  theme(legend.position = c(0.62, 0.18),
        legend.background = element_rect(fill = "white", color = "grey80"),
        legend.key.width = unit(1.2, "cm"))
save_fig(fig_roc, "fig_progressive_roc", width = 7, height = 7)
```

Each curve represents a progressive elastic net model evaluated on the
holdout chromosomes. Stop codon position alone (the stepped orange curve)
reflects the 4-level categorical predictor — it achieves coarse but
meaningful discrimination (AUC =
`r round(unified$elastic_net$step1$auc_test, 3)`). Adding start codon
features produces a smooth curve with substantially improved sensitivity
across all specificity levels. The final unified model (black) adds ORF
features for further gains, particularly in the high-specificity region.

## Feature Importance

```{r sec2a-coefficients}
# ── Data provenance ──
# Source: unified$elastic_net$step3$cv_fit (elastic net, unified model)
# Non-zero coefficients at lambda.min, sorted by absolute magnitude

en_fit <- unified$elastic_net$step3$cv_fit
coefs <- as.matrix(coef(en_fit, s = "lambda.min"))
nonzero <- coefs[coefs[, 1] != 0 & rownames(coefs) != "(Intercept)", , drop = FALSE]

coef_table <- data.frame(
  r_name = rownames(nonzero),
  Coefficient = nonzero[, 1],
  stringsAsFactors = FALSE
) %>%
  mutate(
    Feature = ifelse(r_name %in% names(feat_labels),
                     feat_labels[r_name], r_name),
    `Abs. Coefficient` = abs(Coefficient)
  ) %>%
  arrange(-`Abs. Coefficient`) %>%
  transmute(Feature, Coefficient = round(Coefficient, 4))

render_dt(coef_table, caption = "Table: Elastic Net Feature Importance (Unified Model)")
```

Column definitions: **Feature** = model variable (display name);
**Coefficient** = elastic net coefficient at lambda.min (standardized
features). Positive coefficients increase predicted NMD probability;
negative coefficients decrease it. Features with zero coefficients are
excluded by the elastic net penalty.

## Complementarity: Stop Codon vs Start Codon Features

```{r sec2a-complementarity}
# Stop-codon-only vs 5'UTR-only vs combined
comp_df <- data.frame(
  Model = c("Stop codon only", "5'UTR only", "Combined"),
  `Train AUC` = round(c(
    unified$elastic_net$step1$auc_train,
    unified$elastic_net$utr5_only$auc_train,
    unified$elastic_net$step3$auc_train
  ), 3),
  `Test AUC` = round(c(
    unified$elastic_net$step1$auc_test,
    unified$elastic_net$utr5_only$auc_test,
    unified$elastic_net$step3$auc_test
  ), 3),
  check.names = FALSE
)
render_dt(comp_df, caption = "Complementarity: Feature Sets Alone vs Combined")
```

Column definitions: **Model** = feature set used; **Train AUC** = AUC on
training chromosomes; **Test AUC** = AUC on holdout chromosomes (1, 3, 5, 7).

Stop codon and start codon features are complementary, not redundant. Each
achieves comparable AUC alone (~0.70–0.72), but combined they reach
`r round(unified$elastic_net$step3$auc_test, 3)`. This means PTC and uORF
mechanisms operate through different structural pathways to trigger NMD.

# 2b. Stop Codons: Where Translation Terminates {#sec2b-stop-codons}

The strongest single feature category is the downstream EJC count. Here we
examine stop codon position in detail using gene-matched C2 (NMD) vs C4
(Control) isoform pairs.

**Methods:** A PTC is defined as a stop codon >50 nt upstream of the last exon-junction
complex (EJC). Fisher's exact test compares PTC rates between NMD (C2) and Control (C4)
comparator isoforms, restricted to pairs where both reference and comparator are coding.
PTC position metrics (distance to EJC, % CDS, number of downstream EJCs) characterize NMD
response strength using mashr meta-analysis posterior means averaged across 5 conditions
(DD, AT, DO_ALI, FB, MV).

## PTC Rates

```{r goal1-ptc-rates-subsetting}
# ── Data provenance ──
# Starting data: gene-matched profiles (profiles_list → set_results$profiles)
#   Pipeline: 01 (mashr NMD classification) → 02 (pair generation + buildProfiles)
#   → gene-matching (this report, STAGE 2) → set_results (STAGE 3)
#
# Subsetting for PTC rate calculation:
#   1. Start with gene-matched C2 (NMD) and C4 (Control) profiles
#   2. Restrict to "coding pairs": both reference AND comparator isoform must have
#      coding_status == "coding" in cds.rds (from Isopair extractCds() on SQANTI CDS)
#   3. Numerator: comparator isoforms with has_ptc == TRUE in ptc.rds
#      PTC definition: stop codon >50 nt upstream of the last exon-junction complex (EJC)
#   4. Denominator: total coding pairs (from step 2)
#   5. Fisher's exact test: 2×2 table of (PTC+/PTC−) × (NMD/Control)

# Show the subsetting explicitly
coding_ids <- cds$isoform_id[cds$coding_status == "coding"]
n_gm_c2 <- nrow(set_results[["c2_allsamples"]]$profiles)
n_gm_c4 <- nrow(set_results[["c4_allsamples"]]$profiles)

p_c2 <- set_results[["c2_allsamples"]]$profiles
cod_c2 <- p_c2[p_c2$reference_isoform_id %in% coding_ids &
                p_c2$comparator_isoform_id %in% coding_ids, ]
p_c4 <- set_results[["c4_allsamples"]]$profiles
cod_c4 <- p_c4[p_c4$reference_isoform_id %in% coding_ids &
                p_c4$comparator_isoform_id %in% coding_ids, ]
```

Starting from `r fmt(n_gm_c2)` gene-matched NMD (C2) pairs and
`r fmt(n_gm_c4)` gene-matched Control (C4) pairs, we restrict to **coding
pairs** where both the reference and comparator isoform are classified as
protein-coding by SQANTI (`coding_status == "coding"` in `cds.rds`). This
yields `r fmt(nrow(cod_c2))` coding NMD pairs and `r fmt(nrow(cod_c4))` coding
Control pairs. The PTC rate is the fraction of these coding comparator isoforms
whose stop codon falls >50 nt upstream of the last exon-junction complex (EJC),
as computed by `Isopair::testPtcAssociation()`.

```{r goal1-ptc-rates}
ptc_rates <- compute_ptc_rates_row("all_samples")
kable(ptc_rates, caption = "Table 2: PTC Rates — all_samples",
      digits = c(0, 0, 0, 0, 1, 0, 1, 2, 2))
write.csv(ptc_rates, file.path(table_dir, "table2_ptc_rates_allsamples.csv"), row.names = FALSE)
```

Columns: **cell_type** = sample set; **n_coding_pairs_NMD/Ctrl** = number of
coding pairs (both ref and comp coding) in NMD (C2) / Control (C4);
**NMD_comp_ptc_n/pct** = count and percentage of NMD comparators with PTC;
**Ctrl_comp_ptc_n/pct** = same for Control; **fisher_OR** = odds ratio from
Fisher's exact test; **p_value** = Fisher p-value.

## Figure 2: Stop Codon Distance to Last EJC

```{r goal1-fig1-stop-dist-pop}
# ── Population derivation ──
# Start from gene-matched comparators, filter to those with non-NA ptc_distance.
# This is broader than Table 2 (not restricted to coding pairs).
fig2_pop <- list()
for (comp in c("c2", "c4")) {
  key <- paste0(comp, "_allsamples")
  comp_ids <- set_results[[key]]$profiles$comparator_isoform_id
  n_total <- length(comp_ids)
  ptc_sub <- ptc[ptc$isoform_id %in% comp_ids & !is.na(ptc$ptc_distance), ]
  fig2_pop[[comp]] <- list(n_total = n_total, n_with_dist = nrow(ptc_sub))
}
```

Starting from `r fmt(fig2_pop$c2$n_total)` gene-matched NMD comparators,
`r fmt(fig2_pop$c2$n_with_dist)` have a non-NA `ptc_distance` in `ptc.rds`
(i.e., have an annotated CDS with a stop codon and ≥1 exon-junction complex).
For Control: `r fmt(fig2_pop$c4$n_with_dist)` of `r fmt(fig2_pop$c4$n_total)`.
This is a broader set than Table 2, which additionally requires **both**
reference and comparator to be coding (pair-level filter). Here only the
comparator needs `ptc_distance` (isoform-level filter).

```{r goal1-fig1-stop-dist, fig.width=8, fig.height=5}
stop_dist_all <- {
  rows <- list()
  for (comp in c("c2", "c4")) {
    key <- paste0(comp, "_allsamples")
    comp_ids <- set_results[[key]]$profiles$comparator_isoform_id
    ptc_sub <- ptc[ptc$isoform_id %in% comp_ids & !is.na(ptc$ptc_distance), ]
    if (nrow(ptc_sub) == 0) next
    rows[[comp]] <- data.frame(
      comparison = comp_label(comp),
      ptc_distance = ptc_sub$ptc_distance,
      has_ptc = ptc_sub$has_ptc)
  }
  do.call(rbind, rows)
}

fig1 <- ggplot(stop_dist_all, aes(x = ptc_distance, fill = comparison)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
  annotate("text", x = 50, y = Inf, label = "PTC threshold (50 nt)",
           hjust = -0.05, vjust = 1.5, size = 3, color = "red") +
  scale_x_continuous(limits = c(
    quantile(stop_dist_all$ptc_distance, 0.01, na.rm = TRUE),
    quantile(stop_dist_all$ptc_distance, 0.99, na.rm = TRUE))) +
  scale_fill_manual(values = pal) +
  labs(title = "Stop Codon Distance to Last EJC: NMD vs Control Comparators",
       subtitle = "All comparator isoforms with annotated stop codon + EJC. Dashed red = 50 nt PTC threshold.",
       x = "Distance from stop codon to last EJC (nt)", y = "Density", fill = "")
save_fig(fig1, "fig2_stop_codon_distance")
```

## Figure 3: NMD Response vs Stop Codon Distance

```{r goal1-fig2-logfc-dist, fig.width=8, fig.height=5}
# ── Data provenance ──
# Starting data: gene-matched C2 all_samples profiles
# Step 1: Extract comparator_isoform_id from gene-matched C2 profiles
# Step 2: Look up ptc features (ptc_distance, has_ptc, orf_length, n_downstream_ejcs)
#   from ptc.rds for those comparators
# Step 3: Merge with meta_logfc — mashr PosteriorMean averaged across 5 conditions
#   (DD, AT, DO_ALI, FB, MV). Source: mashr_isoform_model_2026.3.10.rds
# Result: scatter of NMD response magnitude vs stop codon position

comp_ids_as <- set_results[["c2_allsamples"]]$profiles$comparator_isoform_id
ptc_sub <- ptc[ptc$isoform_id %in% comp_ids_as & !is.na(ptc$ptc_distance),
                c("isoform_id", "ptc_distance", "has_ptc", "orf_length", "n_downstream_ejcs")]
n_with_logfc <- sum(comp_ids_as %in% meta_logfc$txid)
nmd_logfc_dist <- merge(meta_logfc[meta_logfc$txid %in% comp_ids_as, ],
                         ptc_sub, by.x = "txid", by.y = "isoform_id")
```

Of `r fmt(length(comp_ids_as))` gene-matched NMD comparators,
`r fmt(nrow(ptc_sub))` have non-NA `ptc_distance` and
`r fmt(n_with_logfc)` have a mashr posterior mean logFC. After merging
(requiring both), `r fmt(nrow(nmd_logfc_dist))` comparators remain.
`r if (length(comp_ids_as) - n_with_logfc > 0) paste0("The ", fmt(length(comp_ids_as) - n_with_logfc), " comparators absent from the mashr model are novel isoforms not tested for differential expression.") else "All comparators have mashr logFC values."`

```{r goal1-fig2-logfc-dist-plot, fig.width=8, fig.height=5}
fig2 <- ggplot(nmd_logfc_dist, aes(x = ptc_distance, y = mean_logFC)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "loess", se = TRUE, color = "steelblue") +
  geom_vline(xintercept = 50, linetype = "dashed", color = "red", linewidth = 0.5) +
  scale_x_continuous(limits = c(
    quantile(nmd_logfc_dist$ptc_distance, 0.01, na.rm = TRUE),
    quantile(nmd_logfc_dist$ptc_distance, 0.99, na.rm = TRUE))) +
  labs(title = sprintf("NMD Response vs Stop Codon Distance (n = %d)", nrow(nmd_logfc_dist)),
       subtitle = "mashr meta-analysis mean logFC (5 conditions). Blue = loess.",
       x = "Distance from stop codon to last EJC (nt)",
       y = "Mean logFC (mashr)")
save_fig(fig2, "fig3_nmd_response_vs_distance")

# Correlation summary
cor_test <- cor.test(nmd_logfc_dist$ptc_distance, nmd_logfc_dist$mean_logFC,
                      method = "spearman")
kable(data.frame(n = nrow(nmd_logfc_dist),
                 spearman_rho = round(cor_test$estimate, 3),
                 p_value = cor_test$p.value),
      caption = "Spearman correlation: Stop-to-EJC distance vs mean logFC")
```

## Figure 4: NMD Strength by Downstream EJC Count

```{r goal1-fig3-pop}
as_ptc_pos <- nmd_logfc_dist[nmd_logfc_dist$has_ptc == TRUE, ]
as_ptc_pos$logFC <- as_ptc_pos$mean_logFC
```

Restricted to the `r fmt(nrow(as_ptc_pos))` **PTC-positive** comparators from
Figure 3's population of `r fmt(nrow(nmd_logfc_dist))` (`has_ptc == TRUE`),
binned by number of downstream EJCs.

```{r goal1-fig3-ejc-boxplot, fig.width=9, fig.height=5}
# Bin EJC counts >= 7 into "7+"
as_ptc_pos$ejc_bin <- ifelse(as_ptc_pos$n_downstream_ejcs >= 7,
                              "7+", as.character(as_ptc_pos$n_downstream_ejcs))
as_ptc_pos$ejc_bin <- factor(as_ptc_pos$ejc_bin,
                              levels = c(as.character(1:6), "7+"))

ejc_counts <- as_ptc_pos |>
  count(ejc_bin) |>
  mutate(label = paste0("n=", n))

fig3 <- ggplot(as_ptc_pos, aes(x = ejc_bin, y = logFC)) +
  geom_violin(fill = "lightblue", alpha = 0.5) +
  geom_boxplot(width = 0.15, outlier.alpha = 0.2) +
  geom_text(data = ejc_counts,
            aes(x = ejc_bin,
                y = max(as_ptc_pos$logFC, na.rm = TRUE) + 0.3, label = label),
            size = 2.8) +
  labs(title = "NMD Comparators: Mean logFC by Number of Downstream EJCs",
       subtitle = "mashr meta-analysis. logFC rises sharply from 1-4 EJCs, then plateaus.",
       x = "Number of Downstream EJCs", y = "Mean logFC (mashr)")
save_fig(fig3, "fig4_ejc_vs_nmd_strength", width = 9)
```

## Table 3: PTC Rates Across Cell Types

```{r goal1-table2-ct-ptc}
# ── Data provenance ──
# Same pipeline as Table 2 (compute_ptc_rates_row), iterated across main_cts.
# Each cell type uses its own gene-matched profiles from set_results.
ptc_ct <- do.call(rbind, lapply(c("all_samples", main_cts), compute_ptc_rates_row))
render_dt(ptc_ct, caption = "Table 3: PTC Rates Across Cell Types")
write.csv(ptc_ct, file.path(table_dir, "table3_ptc_rates_by_ct.csv"), row.names = FALSE)
```

Same columns as Table 2, one row per cell type. all_samples uses the union
(NMD) and intersection (non-NMD) across AT, DD, DO, FB, MV.


## How PTCs Arise: Splicing Mechanisms {#sec2b-ptc-mechanisms}

**Goal 3:** Identify the splicing mechanisms that produce PTCs in NMD-responsive isoforms.

**Methods:** Frame comparison (`compareIsoformFrames()`) classifies each pair by start codon
and reading frame relationship. Frame walk (`analyzeFrameWalk()`) identifies which splicing
events cause frameshifts. The Sankey diagram traces flows from splice event type through frame
category to PTC outcome for NMD comparators.

## Figure 5: Frame Category Distribution

```{r goal2-fig4-pop}
# Count excluded categories for prose
fc_c2_ps <- set_results[["c2_allsamples"]]$frame_comp$pair_summary
fc_c4_ps <- set_results[["c4_allsamples"]]$frame_comp$pair_summary
n_noncoding_c2 <- sum(fc_c2_ps$frame_category %in% c("non_coding", "no_shared_cds"))
n_noncoding_c4 <- sum(fc_c4_ps$frame_category %in% c("non_coding", "no_shared_cds"))
```

`compareIsoformFrames()` assigns each pair a frame category based on start codon
and reading frame relationship. Of `r fmt(nrow(fc_c2_ps))` NMD pairs,
`r fmt(n_noncoding_c2)` are classified as `non_coding` or `no_shared_cds` and
excluded from the figure (Control: `r fmt(n_noncoding_c4)` of
`r fmt(nrow(fc_c4_ps))` excluded). The denominator (y-axis) is the number of
**coding pairs** (`r fmt(nrow(fc_c2_ps) - n_noncoding_c2)` NMD,
`r fmt(nrow(fc_c4_ps) - n_noncoding_c4)` Control), so the 5 categories sum
to 100%.

```{r goal2-fig4-frame-category, fig.width=10, fig.height=6}
# ── Data provenance ──
# Starting data: gene-matched profiles → cached frame comparison
# set_results[[key]]$frame_comp: compareIsoformFrames() output, cached by script 03
#   Input to compareIsoformFrames(): gene-matched profiles + cds_exons.rds + cds.rds
#   $pair_summary contains frame_category per pair (same_start_in_frame, same_start_frameshift, etc.)
# ptc.rds$has_ptc: PTC status per isoform (isoform-intrinsic, independent of pairing)
# Denominator (total_pairs): nrow(gene-matched profiles) — ensures consistent
#   percentages between profiles and frame_comp (verified in STAGE 4)

coding_cats <- c("same_start_in_frame", "same_start_frameshift",
                 "diff_start_same_frame", "diff_start_diff_frame",
                 "diff_start_diff_frame_with_frameshift")
cat_labels <- c(
  same_start_in_frame = "Same Start\nSame Frame",
  same_start_frameshift = "Same Start\nFrameshift",
  diff_start_same_frame = "Diff Start\nSame Frame",
  diff_start_diff_frame = "Diff Start\nDiff Frame",
  diff_start_diff_frame_with_frameshift = "Diff Start\nDiff Frame + FS")

ptc_pos_ids <- ptc$isoform_id[ptc$has_ptc == TRUE]

# 4-color palette: NMD/Control × PTC+/No PTC
pal4 <- c("NMD PTC+"    = "#d6604d",   # dark orange-red
          "NMD No PTC"  = "#fddbc7",   # light orange
          "Ctrl PTC+"   = "#4393c3",   # dark blue
          "Ctrl No PTC" = "#d1e5f0")   # light blue

stacked_df <- do.call(rbind, lapply(c("c2", "c4"), function(comp) {
  key <- paste0(comp, "_allsamples")
  fc <- set_results[[key]]$frame_comp
  if (is.null(fc)) return(NULL)
  ps <- fc$pair_summary
  # Denominator: coding pairs only (exclude non_coding and no_shared_cds)
  coding_pairs_total <- sum(!ps$frame_category %in% c("non_coding", "no_shared_cds"))
  comp_lbl <- comp_label(comp)
  comp_short <- ifelse(comp_lbl == "NMD", "NMD", "Ctrl")

  do.call(rbind, lapply(coding_cats, function(cat) {
    cat_pairs <- ps[ps$frame_category == cat, ]
    n_cat <- nrow(cat_pairs)
    n_ptc <- sum(cat_pairs$comparator_isoform_id %in% ptc_pos_ids)
    n_no_ptc <- n_cat - n_ptc
    pct_ptc <- 100 * n_ptc / max(coding_pairs_total, 1)
    pct_no_ptc <- 100 * n_no_ptc / max(coding_pairs_total, 1)
    rbind(
      data.frame(frame_category = cat, category_label = cat_labels[cat],
                 comparison = comp_lbl, ptc_status = "PTC+",
                 fill_group = paste(comp_short, "PTC+"),
                 n = n_ptc, pct = pct_ptc),
      data.frame(frame_category = cat, category_label = cat_labels[cat],
                 comparison = comp_lbl, ptc_status = "No PTC",
                 fill_group = paste(comp_short, "No PTC"),
                 n = n_no_ptc, pct = pct_no_ptc))
  }))
}))

stacked_df$category_label <- factor(stacked_df$category_label,
  levels = cat_labels[coding_cats])
stacked_df$fill_group <- factor(stacked_df$fill_group,
  levels = c("NMD PTC+", "NMD No PTC", "Ctrl PTC+", "Ctrl No PTC"))

fig4 <- ggplot(stacked_df, aes(x = category_label, y = pct, fill = fill_group)) +
  geom_col() +
  facet_wrap(~comparison) +
  scale_fill_manual(values = pal4) +
  labs(title = "Frame Category Distribution: NMD vs Control (all_samples)",
       subtitle = "% of total pairs. PTC+ = comparator has premature termination codon.",
       x = "Frame Category", y = "% of Total Pairs", fill = "") +
  theme(axis.text.x = element_text(size = 8))
save_fig(fig4, "fig5_frame_category_distribution", width = 10, height = 6)
```

## Table 4: Frame Category Counts

```{r goal2-table3-frame-cats}
fc_rows <- compute_frame_category_rows("all_samples")
render_dt(fc_rows, caption = "Table 4: Frame Category Distribution — all_samples")
write.csv(fc_rows, file.path(table_dir, "table4_frame_categories.csv"), row.names = FALSE)
```

Columns: **cell_type**; **frame_category** = reading frame relationship between
reference and comparator (from `Isopair::compareIsoformFrames()`);
**NMD_n/NMD_pct** = count and percentage in NMD (C2) pairs;
**Ctrl_n/Ctrl_pct** = same for Control (C4). The denominator for percentages is
**coding pairs** (excluding `non_coding` and `no_shared_cds`), consistent with
Figure 5 — the 5 coding categories sum to 100%. The `non_coding` and
`no_shared_cds` rows show counts only (no percentage).

Note: the coding pair count here differs from Table 2 because `compareIsoformFrames()`
classifies some individually-coding pairs as `no_shared_cds` (both isoforms are
coding but their CDS regions don't overlap). Table 2's "coding pairs" filter
requires only that both isoforms have `coding_status == "coding"` in `cds.rds`.

## How PTCs Arise: Frameshift vs Non-Frameshift Mechanisms

```{r goal2-ptc-mechanisms}
# ── Data provenance ──
# Source: computed inline in this report (not a separate script)
# Methods: METHODS.md § "PTC-Causing Event Attribution"
#
# For each PTC+ NMD comparator, identify the PTC-CAUSING splice event:
#   - Frameshift PTCs: from analyzeFrameWalk() (Isopair::analyzeFrameWalk,
#     cached as fw_c2_allsamples.rds by 03_nmd_analysis_mashr.Rmd)
#   - Non-frameshift PTCs: the event whose genomic coordinates (from
#     buildProfiles() detailed_events) contain the strand-aware PTC position
#     (from Isopair::extractCds → cds.rds)
#
# Strand-aware PTC position:
#   + strand: PTC = cds_stop (larger genomic coordinate = 3' end of CDS)
#   - strand: PTC = cds_start (smaller genomic coordinate = 3' end on minus strand)
# For PTC isoforms, SQANTI annotates the actual (premature) stop, so cds_start/
# cds_stop IS the PTC position (depending on strand).

fc_c2 <- set_results[["c2_allsamples"]]$frame_comp
fw_c2 <- set_results[["c2_allsamples"]]$frame_walk
prof_c2 <- set_results[["c2_allsamples"]]$profiles

fc_c2_ps <- fc_c2$pair_summary
fc_c2_ps$has_ptc <- fc_c2_ps$comparator_isoform_id %in% ptc_pos_ids

frameshift_cats <- c("same_start_frameshift",
                     "diff_start_diff_frame_with_frameshift")
fc_c2_ps$is_frameshift_cat <- fc_c2_ps$frame_category %in% frameshift_cats

# PTC+ pairs by mechanism — restrict to coding pairs (both ref and comp coding)
# to match the coding_pairs_c2 base used in Goals 3-4 and the cumulative flowchart
coding_ids_g2 <- cds$isoform_id[cds$coding_status == "coding"]
ptc_plus <- fc_c2_ps[fc_c2_ps$has_ptc == TRUE &
  fc_c2_ps$comparator_isoform_id %in% coding_ids_g2 &
  fc_c2_ps$reference_isoform_id %in% coding_ids_g2, ]
n_ptc_via_fs    <- sum(ptc_plus$is_frameshift_cat)
n_ptc_via_nonfs <- sum(!ptc_plus$is_frameshift_cat)
n_ptc_total     <- nrow(ptc_plus)

# Also count all coding NMD pairs for context (same coding filter as ptc_plus)
fc_c2_coding <- fc_c2_ps[fc_c2_ps$comparator_isoform_id %in% coding_ids_g2 &
  fc_c2_ps$reference_isoform_id %in% coding_ids_g2, ]
n_coding_nmd <- nrow(fc_c2_coding)

# ── Strand-aware PTC position lookups ──
# On + strand the stop codon is at cds_stop (larger coordinate).
# On - strand the stop codon is at cds_start (smaller coordinate).
cds_start_lookup <- setNames(cds$cds_start, cds$isoform_id)
cds_stop_lookup  <- setNames(cds$cds_stop, cds$isoform_id)
strand_lookup    <- setNames(cds$strand, cds$isoform_id)

# Build lookup: comparator_id → row index in prof_c2
comp_to_idx <- setNames(seq_len(nrow(prof_c2)), prof_c2$comparator_isoform_id)

# Frameshift PTC-causing events (from frame_walk)
fw_fs <- fw_c2$events[fw_c2$events$is_frameshift == TRUE, ]

ptc_event_rows <- list()
for (i in seq_len(nrow(ptc_plus))) {
  comp_id <- ptc_plus$comparator_isoform_id[i]
  is_fs <- ptc_plus$is_frameshift_cat[i]

  if (is_fs) {
    # Frameshift PTC: use the frameshift-causing event from frame_walk
    fs_match <- fw_fs[fw_fs$comparator_isoform_id == comp_id, ]
    if (nrow(fs_match) > 0) {
      ptc_event_rows[[length(ptc_event_rows) + 1]] <- data.frame(
        comparator_id = comp_id,
        mechanism = "Frameshift",
        ptc_causing_event = fs_match$event_type[1],
        attribution = "direct",
        stringsAsFactors = FALSE)
    } else {
      # Frameshift category but no is_frameshift event in frame_walk.
      # This indicates a disagreement between compareIsoformFrames() and
      # analyzeFrameWalk() — do not guess; label as unresolved.
      ptc_event_rows[[length(ptc_event_rows) + 1]] <- data.frame(
        comparator_id = comp_id,
        mechanism = "Frameshift",
        ptc_causing_event = "Unresolved",
        attribution = "unresolved",
        stringsAsFactors = FALSE)
    }
  } else {
    # Non-frameshift PTC: find the event containing the strand-aware PTC position
    prof_idx <- comp_to_idx[comp_id]
    if (is.na(prof_idx)) next
    de <- prof_c2$detailed_events[[prof_idx]]
    if (is.null(de) || nrow(de) == 0) next

    # Strand-aware PTC genomic position
    comp_strand <- strand_lookup[comp_id]
    ptc_pos_genomic <- if (!is.na(comp_strand) && comp_strand == "-") {
      cds_start_lookup[comp_id]
    } else {
      cds_stop_lookup[comp_id]
    }
    if (is.na(ptc_pos_genomic)) next

    # Check which event's coordinates span the PTC position
    de$ev_min <- pmin(de$five_prime, de$three_prime)
    de$ev_max <- pmax(de$five_prime, de$three_prime)
    containing <- de[ptc_pos_genomic >= de$ev_min & ptc_pos_genomic <= de$ev_max, ]

    if (nrow(containing) > 0) {
      ptc_event_rows[[length(ptc_event_rows) + 1]] <- data.frame(
        comparator_id = comp_id,
        mechanism = "In-frame stop",
        ptc_causing_event = containing$event_type[1],
        attribution = "direct",
        stringsAsFactors = FALSE)
    } else {
      ptc_event_rows[[length(ptc_event_rows) + 1]] <- data.frame(
        comparator_id = comp_id,
        mechanism = "In-frame stop",
        ptc_causing_event = "Unresolved",
        attribution = "unresolved",
        stringsAsFactors = FALSE)
    }
  }
}

ptc_events_df <- do.call(rbind, ptc_event_rows)
n_attributed <- sum(ptc_events_df$attribution == "direct")
n_unresolved <- sum(ptc_events_df$attribution == "unresolved")
```

Of `r fmt(n_coding_nmd)` coding NMD pairs, `r fmt(n_ptc_total)` have a PTC in
the comparator. These arise through two mechanisms:

- **Frameshift → PTC**: `r fmt(n_ptc_via_fs)` pairs
  (`r round(100 * n_ptc_via_fs / n_ptc_total, 1)`%) — a splicing event shifts
  the reading frame, and the new frame encounters a premature stop codon. The
  PTC-causing event is identified by `analyzeFrameWalk()`.
- **In-frame stop → PTC**: `r fmt(n_ptc_via_nonfs)` pairs
  (`r round(100 * n_ptc_via_nonfs / n_ptc_total, 1)`%) — the comparator is
  in the same reading frame but a structural change (e.g., a gained exon or
  retained intron) directly introduces an in-frame stop codon. The PTC-causing
  event is identified by finding the event whose genomic coordinates contain the
  comparator's stop codon position.

### PTC position and strand awareness

The PTC genomic position depends on strand: for + strand genes, the stop codon
is at `cds_stop` (the larger genomic coordinate); for - strand genes, it is at
`cds_start` (the smaller genomic coordinate, which is the 3' end in
transcription direction). This mirrors the logic in `Isopair::computePtcStatus`.

### Attribution results

Of `r fmt(nrow(ptc_events_df))` PTC+ pairs, `r fmt(n_attributed)` were
directly attributed (PTC position inside event coordinates or frameshift event
from `analyzeFrameWalk`)`r if (n_unresolved > 0) paste0(", and ", fmt(n_unresolved), " remain unresolved")``r if (n_unresolved == 0) ""`.
`r if (n_unresolved > 0) "Unresolved cases include both in-frame stop pairs where the PTC falls outside all event coordinates and frameshift-category pairs where analyzeFrameWalk() found no frameshift event (indicating a classification disagreement between compareIsoformFrames() and analyzeFrameWalk())."`

```{r goal2-ptc-diagnostic, eval = n_unresolved > 0}
# ── Diagnostic: characterize all unresolved PTC attributions ──
# Two types of unresolved:
#   1. In-frame stop: PTC position outside all event coordinates
#   2. Frameshift: frame_category says frameshift but analyzeFrameWalk() found
#      no is_frameshift event (classification disagreement)
# For each, report mechanism, strand, frame_category, and distance to nearest event.

unresolved <- ptc_events_df[ptc_events_df$attribution == "unresolved", ]

diag_rows <- list()
for (i in seq_len(nrow(unresolved))) {
  uid <- unresolved$comparator_id[i]
  mech <- unresolved$mechanism[i]

  prof_idx <- comp_to_idx[uid]
  fc_row <- fc_c2_ps[fc_c2_ps$comparator_isoform_id == uid, ]
  gene_id <- if (nrow(fc_row) > 0) fc_row$gene_id[1] else NA_character_
  frame_cat <- if (nrow(fc_row) > 0) fc_row$frame_category[1] else NA_character_
  comp_strand <- strand_lookup[uid]

  # PTC position and distance to nearest event (if events exist)
  ptc_pos <- if (!is.na(comp_strand) && comp_strand == "-") {
    cds_start_lookup[uid]
  } else {
    cds_stop_lookup[uid]
  }

  nearest_evt <- NA_character_; min_dist <- NA_real_; n_ev <- 0L
  if (!is.na(prof_idx)) {
    de <- prof_c2$detailed_events[[prof_idx]]
    if (!is.null(de) && nrow(de) > 0) {
      n_ev <- nrow(de)
      if (!is.na(ptc_pos)) {
        de$ev_min <- pmin(de$five_prime, de$three_prime)
        de$ev_max <- pmax(de$five_prime, de$three_prime)
        nearest_idx <- which.min(pmin(abs(ptc_pos - de$ev_min),
                                      abs(ptc_pos - de$ev_max)))
        min_dist <- min(abs(ptc_pos - c(de$ev_min, de$ev_max)))
        nearest_evt <- de$event_type[nearest_idx]
      }
    }
  }

  diag_rows[[length(diag_rows) + 1]] <- data.frame(
    comparator_id = uid, gene_id = gene_id, mechanism = mech,
    strand = comp_strand, frame_category = frame_cat,
    ptc_pos = if (is.na(ptc_pos)) NA_real_ else ptc_pos,
    nearest_event_type = nearest_evt,
    dist_to_nearest_boundary = min_dist,
    n_events = n_ev,
    stringsAsFactors = FALSE)
}
diag_df <- do.call(rbind, diag_rows)
diag_df <- diag_df[order(diag_df$mechanism, diag_df$dist_to_nearest_boundary), ]

render_dt(diag_df,
  caption = paste0("Diagnostic: ", nrow(diag_df),
    " Unresolved PTC Attributions"))
```

`r if (n_unresolved > 0) paste0("The diagnostic table shows ", fmt(n_unresolved), " unresolved cases. **Mechanism** distinguishes the two types: *In-frame stop* pairs have a PTC outside all event coordinates (dist_to_nearest_boundary shows how far); *Frameshift* pairs have a frameshift frame_category but analyzeFrameWalk() found no is_frameshift event, indicating a classification disagreement between the two Isopair functions.")`

```{r goal2-ptc-fs-test}
# ── Data provenance ──
# n_ptc_via_fs, n_ptc_total: from goal2-ptc-mechanisms chunk
# Baseline: frameshift rate among ALL coding NMD pairs (not just PTC+)
# fc_c2_ps$is_frameshift_cat: TRUE for frameshift frame categories
# Binomial test: is the frameshift proportion among PTC+ pairs different
#   from the frameshift proportion among all coding NMD pairs?

n_fs_all_nmd <- sum(fc_c2_coding$is_frameshift_cat)
baseline_fs_rate <- n_fs_all_nmd / n_coding_nmd

bt_fs <- binom.test(n_ptc_via_fs, n_ptc_total, p = baseline_fs_rate)
```

Frameshifts account for **`r fmt(n_ptc_via_fs)`** of **`r fmt(n_ptc_total)`**
PTC events in NMD pairs (**`r round(100 * n_ptc_via_fs / n_ptc_total, 1)`%**),
with in-frame stops accounting for the remaining
**`r round(100 * n_ptc_via_nonfs / n_ptc_total, 1)`%** — the two mechanisms
contribute roughly equally. However, frameshifts are significantly enriched
among PTC+ pairs relative to the overall frameshift rate among all coding NMD
pairs (`r round(100 * baseline_fs_rate, 1)`%; binomial test p =
`r signif(bt_fs$p.value, 3)`).

## Figure 6: PTC-Causing Splice Events (Sankey)

```{r goal2-fig5-sankey}
# ── Data provenance ──
# Combines three sources of PTC attribution:
#   1. Diff-stop frameshift: ptc_events_df (from analyzeFrameWalk)
#   2. Diff-stop in-frame: ptc_events_df (from coordinate containment)
#   3. Same-stop (3'UTR splice): attributed here using nearest-to-stop 3'UTR event
# Same-stop attribution: for each same-stop pair, find splice events downstream
# of the stop codon and attribute to the nearest one (the junction most likely
# to place the critical EJC within the >50 nt window).

# ── Identify same-stop PTC+ pairs (strand-aware) ──
# This duplicates the strand-aware stop comparison from goal3-fig6-pop,
# which has not yet run. Uses the same logic: + strand → cds_stop, - strand → cds_start.
ptc_plus_sankey <- ptc_plus  # from goal2-ptc-mechanisms
ptc_plus_sankey$strand_s <- strand_lookup[ptc_plus_sankey$comparator_isoform_id]
ptc_plus_sankey$comp_stop_s <- ifelse(ptc_plus_sankey$strand_s == "-",
  cds_start_lookup[ptc_plus_sankey$comparator_isoform_id],
  cds_stop_lookup[ptc_plus_sankey$comparator_isoform_id])
ptc_plus_sankey$ref_stop_s <- ifelse(ptc_plus_sankey$strand_s == "-",
  cds_start_lookup[ptc_plus_sankey$reference_isoform_id],
  cds_stop_lookup[ptc_plus_sankey$reference_isoform_id])
ptc_plus_sankey$same_stop_s <- ptc_plus_sankey$comp_stop_s == ptc_plus_sankey$ref_stop_s

same_stop_sankey <- ptc_plus_sankey[which(ptc_plus_sankey$same_stop_s == TRUE), ]

# ── Attribute same-stop pairs to nearest 3'UTR event ──
same_stop_attr <- list()
for (i in seq_len(nrow(same_stop_sankey))) {
  comp_id <- same_stop_sankey$comparator_isoform_id[i]
  stop_pos <- same_stop_sankey$comp_stop_s[i]
  comp_strand <- same_stop_sankey$strand_s[i]

  prof_idx <- comp_to_idx[comp_id]
  if (is.na(prof_idx)) next
  de <- prof_c2$detailed_events[[prof_idx]]
  if (is.null(de) || nrow(de) == 0) next

  # Find events downstream of stop in transcription direction
  if (comp_strand == "+") {
    utr3_ev <- de[de$five_prime >= stop_pos | de$three_prime >= stop_pos, ]
    if (nrow(utr3_ev) == 0) next
    ev_starts <- pmin(utr3_ev$five_prime, utr3_ev$three_prime)
    nearest_idx <- which.min(ev_starts - stop_pos)
  } else {
    utr3_ev <- de[de$five_prime <= stop_pos | de$three_prime <= stop_pos, ]
    if (nrow(utr3_ev) == 0) next
    ev_ends <- pmax(utr3_ev$five_prime, utr3_ev$three_prime)
    nearest_idx <- which.min(stop_pos - ev_ends)
  }

  same_stop_attr[[length(same_stop_attr) + 1]] <- data.frame(
    comparator_id = comp_id,
    mechanism = "3'UTR splice",
    ptc_causing_event = utr3_ev$event_type[nearest_idx],
    attribution = "direct",
    stringsAsFactors = FALSE)
}
same_stop_events <- do.call(rbind, same_stop_attr)
n_same_stop_attributed <- nrow(same_stop_events)

# ── Combine all attributed pairs ──
# Diff-stop attributed (from goal2-ptc-mechanisms), excluding same-stop pairs.
# Same-stop pairs may have been "directly" attributed in goal2 if the stop
# position fell inside an event by coincidence — these belong in 3'UTR splice.
# Pairs with unknown ref stop (coding_status="unknown") are kept in diff-stop.
same_stop_comp_ids <- same_stop_sankey$comparator_isoform_id
diffstop_attr <- ptc_events_df[ptc_events_df$attribution == "direct" &
  !ptc_events_df$comparator_id %in% same_stop_comp_ids, ]
# Same-stop attributed (from above)
all_attr <- rbind(diffstop_attr, same_stop_events)

# Shorten event type labels
all_attr$event_short <- all_attr$ptc_causing_event
all_attr$event_short[all_attr$event_short == "Missing_Internal"] <- "Missing Int."
all_attr$event_short[all_attr$event_short == "Partial_IR_5"] <- "Partial IR 5'"
all_attr$event_short[all_attr$event_short == "Partial_IR_3"] <- "Partial IR 3'"
all_attr$event_short[all_attr$event_short == "Alt_TSS"] <- "Alt TSS"
all_attr$event_short[all_attr$event_short == "Alt_TES"] <- "Alt TES"
all_attr$event_short[all_attr$event_short == "IR_diff"] <- "IR diff"
all_attr$event_short[all_attr$event_short == "IR_diff_3"] <- "IR diff 3'"
all_attr$event_short[all_attr$event_short == "IR_diff_5"] <- "IR diff 5'"

# Collapse rare event types (<3 occurrences) into "Other"
evt_counts <- table(all_attr$event_short)
rare_events <- names(evt_counts[evt_counts < 3])
all_attr$event_short[all_attr$event_short %in% rare_events] <- "Other"

# Also make ptc_attr available for Table 5a/5b (diff-stop attributed only)
ptc_attr <- all_attr[all_attr$mechanism != "3'UTR splice", ]

# Aggregate: event_type → mechanism
sankey_agg <- all_attr |>
  group_by(event_short, mechanism) |>
  summarise(freq = n(), .groups = "drop")

# Build networkD3 data structures
nodes_vec <- unique(c(sankey_agg$event_short, sankey_agg$mechanism))
nodes <- data.frame(name = nodes_vec, stringsAsFactors = FALSE)

links <- data.frame(
  source = match(sankey_agg$event_short, nodes_vec) - 1L,
  target = match(sankey_agg$mechanism, nodes_vec) - 1L,
  value  = sankey_agg$freq
)

# Node colors: mechanism nodes colored, event types grey
mech_colors <- c("Frameshift" = "#d6604d",
                 "In-frame stop" = "#4393c3",
                 "3'UTR splice" = "#66c2a5")
node_colors <- ifelse(nodes$name %in% names(mech_colors),
                      mech_colors[nodes$name], "#bdbdbd")

color_scale <- paste0(
  'd3.scaleOrdinal().domain([',
  paste0('"', nodes$name, '"', collapse = ','), ']).range([',
  paste0('"', node_colors, '"', collapse = ','), '])')

networkD3::sankeyNetwork(
  Links = links, Nodes = nodes,
  Source = "source", Target = "target", Value = "value",
  NodeID = "name", fontSize = 14, nodeWidth = 30, nodePadding = 15,
  colourScale = color_scale, sinksRight = TRUE)
```

This figure maps each PTC-causing splice event to one of three mechanisms:
**Frameshift** (red; `r fmt(sum(all_attr$mechanism == "Frameshift"))` pairs) —
a splice event shifts the reading frame, introducing a new stop codon;
**In-frame stop** (blue; `r fmt(sum(all_attr$mechanism == "In-frame stop"))`) —
a splice event introduces an in-frame stop codon directly;
**3'UTR splice** (green; `r fmt(n_same_stop_attributed)`) —
the comparator shares the reference's stop codon, but a downstream splice event
repositions an EJC, making the normal stop "premature." For 3'UTR splice pairs,
the attributed event is the nearest downstream event to the stop codon.
`r { n_same_stop_unresolved_in_goal2 <- sum(ptc_events_df$attribution == "unresolved" & ptc_events_df$comparator_id %in% same_stop_comp_ids); n_diffstop_unresolved <- n_unresolved - n_same_stop_unresolved_in_goal2; n_same_no_event <- nrow(same_stop_sankey) - n_same_stop_attributed; paste0(fmt(n_diffstop_unresolved), " unresolved diff-stop pairs and ", fmt(n_same_no_event), " same-stop pairs without identifiable 3'UTR events are excluded (", fmt(nrow(all_attr)), " of ", fmt(nrow(ptc_plus)), " PTC+ pairs shown).") }`

## Table 5a: PTC-Causing Events by Mechanism

```{r goal2-fig5-summary}
# Summary table of PTC-causing events by mechanism
# Denominator: all directly attributed pairs (excludes unresolved)
ptc_event_summary <- ptc_attr |>
  group_by(mechanism, ptc_causing_event) |>
  summarise(n = n(), .groups = "drop") |>
  arrange(mechanism, desc(n))
ptc_event_summary$pct <- round(100 * ptc_event_summary$n / nrow(ptc_attr), 1)

render_dt(ptc_event_summary, caption = "Table 5a: PTC-Causing Events by Mechanism (NMD)")
write.csv(ptc_event_summary, file.path(table_dir, "table5a_ptc_causing_events.csv"),
          row.names = FALSE)
```

Columns: **mechanism** = Frameshift or In-frame stop; **ptc_causing_event** =
splice event type that caused the PTC; **n** = number of pairs; **pct** =
percentage of all `r fmt(nrow(ptc_attr))` attributed PTCs.

## Table 5b: PTC-Causing Event Proportions vs Control Baseline

```{r goal2-table4b-baseline}
# ── Data provenance ──
# NMD: PTC-causing events attributed above (both frameshift and in-frame stop)
# Control baseline: all splicing events across all C4 pairs
# Comparison: proportion of each event type among PTC-causing NMD events vs
#   all Control events (the baseline frequency in normal alternative splicing)

# NMD PTC-causing event frequencies (attributed pairs: direct + largest_event)
n_ptc_attr <- nrow(ptc_attr)
ptc_evt_freq <- sort(table(ptc_attr$ptc_causing_event), decreasing = TRUE)
ptc_evt_df <- data.frame(
  event_type = names(ptc_evt_freq),
  n_ptc_events = as.integer(ptc_evt_freq),
  pct_of_ptc = round(100 * as.integer(ptc_evt_freq) / n_ptc_attr, 1),
  stringsAsFactors = FALSE)

# Control baseline: all events across all C4 pairs
ctrl_all_events <- do.call(rbind, set_results[["c4_allsamples"]]$profiles$detailed_events)
ctrl_event_freq <- table(ctrl_all_events$event_type)
ctrl_total <- sum(ctrl_event_freq)

# Merge
ptc_evt_df$n_ctrl_events <- as.integer(ctrl_event_freq[ptc_evt_df$event_type])
ptc_evt_df$n_ctrl_events[is.na(ptc_evt_df$n_ctrl_events)] <- 0L
ptc_evt_df$pct_ctrl <- round(100 * ptc_evt_df$n_ctrl_events / ctrl_total, 1)
ptc_evt_df$enrichment <- round(ptc_evt_df$pct_of_ptc / pmax(ptc_evt_df$pct_ctrl, 0.1), 1)

# Fisher's exact test per event type: is this event over-represented among
# PTC-causing events vs all Control events?
ptc_evt_df$fisher_p <- NA_real_
ptc_evt_df$direction <- NA_character_
for (i in seq_len(nrow(ptc_evt_df))) {
  a <- ptc_evt_df$n_ptc_events[i]      # this event in PTC-causing
  b <- n_ptc_attr - a                    # other events in PTC-causing
  c_val <- ptc_evt_df$n_ctrl_events[i]  # this event in Control
  d_val <- ctrl_total - c_val            # other events in Control
  ft <- tryCatch(fisher.test(matrix(c(a, b, c_val, d_val), nrow = 2)),
                 error = function(e) NULL)
  if (!is.null(ft)) {
    ptc_evt_df$fisher_p[i] <- signif(ft$p.value, 3)
    ptc_evt_df$direction[i] <- if (ft$p.value > 0.05) "Non-significant"
      else if (ptc_evt_df$enrichment[i] > 1) "Enriched in PTC-causing"
      else "Depleted in PTC-causing"
  }
}

render_dt(ptc_evt_df, caption = "Table 5b: PTC-Causing Events (NMD) vs All Events (Control)")
write.csv(ptc_evt_df, file.path(table_dir, "table5b_ptc_event_enrichment.csv"),
          row.names = FALSE)
```

Columns: **event_type** = splice event type; **n_ptc_events** = count of
PTC-causing events of this type; **pct_of_ptc** = % of all
`r fmt(n_ptc_attr)` attributed PTC-causing events; **n_ctrl_events** = count in
gene-matched Control (C4) pairs; **pct_ctrl** = % of `r fmt(ctrl_total)` Control
events; **enrichment** = fold enrichment (% PTC / % Ctrl); **fisher_p** =
Fisher's exact test p-value (2×2: this event vs others × PTC-causing vs
Control); **direction** = Enriched/Depleted in PTC-causing, or Non-significant
(p > 0.05).


## 3'UTR Length and Splicing {#sec2b-3utr}

The unified model (Section 2a) showed that 3'UTR features do not add predictive
power beyond stop codon position. Here we test the long-3'UTR hypothesis directly
and show that 3'UTR splicing operates through the PTC pathway (EJC repositioning),
not independently.

```{r goal3-identify-ptc-neg}
# ── Data provenance ──
# Starting data: gene-matched C2 all_samples profiles → cached PTC association
# ptc_c2_as = set_results[["c2_allsamples"]]$ptc_assoc
#   Source: testPtcAssociation(ptc, gene-matched profiles), cached by script 03
#   $pair_summary: one row per gene-matched pair, with comp_has_ptc flag
# Step 1: Filter to coding pairs (both ref and comp in cds$coding_status == "coding")
# Step 2: Split into PTC+ and PTC- NMD comparators

ptc_c2_as <- set_results[["c2_allsamples"]]$ptc_assoc
pair_summary <- ptc_c2_as$pair_summary

# Restrict to coding pairs (same filter as Table 2)
coding_ids_g3 <- cds$isoform_id[cds$coding_status == "coding"]
coding_pairs_c2 <- pair_summary[pair_summary$comparator_isoform_id %in% coding_ids_g3 &
                                  pair_summary$reference_isoform_id %in% coding_ids_g3, ]

n_ptc_pos <- sum(coding_pairs_c2$comp_has_ptc == TRUE)
n_ptc_neg <- sum(coding_pairs_c2$comp_has_ptc == FALSE)
```

Of the `r fmt(nrow(coding_pairs_c2))` coding NMD pairs,
`r fmt(n_ptc_pos)` (`r round(100 * n_ptc_pos / nrow(coding_pairs_c2), 1)`%)
have a PTC and `r fmt(n_ptc_neg)` (`r round(100 * n_ptc_neg / nrow(coding_pairs_c2), 1)`%)
do not. The table below categorizes these PTC-negative comparators by their stop
codon position relative to exon-exon junctions.

```{r goal3-stop-position}
# ── Data provenance ──
# Starting data: ptc_neg_ids (PTC-negative NMD comparators identified above)
# ptc.rds: per-isoform annotations (n_downstream_ejcs, ptc_distance, etc.)
#   Source: Isopair testPtcAssociation() infrastructure — isoform-intrinsic
# Step 1: Look up ptc features for PTC-neg comparators
# Step 2: Categorize by n_downstream_ejcs (borderline vs stop-in-last-exon)

ptc_neg_ids <- coding_pairs_c2$comparator_isoform_id[coding_pairs_c2$comp_has_ptc == FALSE]
ptc_neg_info <- ptc[ptc$isoform_id %in% ptc_neg_ids, ]

# Categorize
n_borderline <- sum(ptc_neg_info$n_downstream_ejcs > 0, na.rm = TRUE)
n_stop_last_exon <- sum(ptc_neg_info$n_downstream_ejcs == 0, na.rm = TRUE)
n_na <- sum(is.na(ptc_neg_info$n_downstream_ejcs))

stop_cat <- data.frame(
  category = c("Borderline (1+ downstream EJCs, distance <= 50 nt)",
               "Stop in last exon (0 downstream EJCs)",
               "Missing annotation"),
  n = c(n_borderline, n_stop_last_exon, n_na),
  pct = round(100 * c(n_borderline, n_stop_last_exon, n_na) / length(ptc_neg_ids), 1))

kable(stop_cat, caption = "Table 6: Categorization of PTC-Negative NMD Comparators")
```

### 3'UTR Length: Testing the Long-3'UTR Hypothesis

The "long 3'UTR" hypothesis proposes that NMD can be triggered when the distance
between the stop codon and the poly(A) tail is abnormally large, even without a
classical PTC. The idea is that UPF1 — normally displaced by translating ribosomes —
accumulates on long 3'UTRs and triggers degradation (Hogg & Goff 2010; Kurosaki &
Maquat 2016). If this mechanism operates in our data, PTC-negative NMD comparators
should have longer 3'UTRs than Controls.

However, measuring 3'UTR length in PTC+ isoforms requires care. When a frameshift
creates a premature stop codon, the sequence between the PTC and the normal stop
is measured as "3'UTR" even though it was originally coding sequence. This inflates
the apparent 3'UTR length. We address this by computing two measurements for PTC+
isoforms with a different stop codon: the naive measurement (from the comparator's
own stop) and a corrected measurement (from the reference's stop position projected
onto the comparator's exon structure).

## Figure 7: 3'UTR Length by PTC Mechanism

```{r goal3-fig6-pop}
# ── Data provenance ──
# Source: computed inline in this report using cached divergence + cds.rds
# Methods: METHODS.md § "3'UTR Length Analysis with PTC Correction"
# Key functions:
#   Isopair::quantifyPairDivergence() → div_c2/div_c4 (utr3_bp_comp column)
#   Isopair::.computeRegionalBp() → computes 3'UTR as exonic bp after cds_stop
#   Reference-based 3'UTR: computed inline using ref cds_stop on comp exon structure
# Starting data: gene-matched C2 and C4 profiles → cached divergence
# PTC mechanism classification: "same stop" vs "diff stop"
#   same_stop: comparator and reference share the same cds_stop position.
#     The PTC arises because a 3'UTR splicing event repositioned the last EJC
#     downstream, making the existing stop codon >50 nt upstream of the new EJC.
#     The 3'UTR length measurement (from cds_stop to tx end) is correct and
#     unaffected by the PTC classification.
#   diff_stop: comparator has a different cds_stop (= premature stop from
#     frameshift or in-frame stop in the CDS). The measured 3'UTR is inflated
#     because it includes the former CDS region between PTC and normal stop.
#     For these, we also compute a "reference-based 3'UTR": exonic bp downstream
#     of the REFERENCE's cds_stop in the COMPARATOR's exon structure. This is
#     only valid when the reference stop position falls within the comparator's
#     exons (86% of cases).

div_c2 <- set_results[["c2_allsamples"]]$divergence
div_c4 <- set_results[["c4_allsamples"]]$divergence
ptc_pos_comp_ids <- coding_pairs_c2$comparator_isoform_id[coding_pairs_c2$comp_has_ptc == TRUE]

# Classify PTC+ by same vs different stop codon position
# Strand-aware: + strand stop = cds_stop; - strand stop = cds_start
cds_start_lookup_g3 <- setNames(cds$cds_start, cds$isoform_id)
cds_stop_lookup_g3  <- setNames(cds$cds_stop, cds$isoform_id)
strand_lookup_g3    <- setNames(cds$strand, cds$isoform_id)

ptc_pos_ps <- coding_pairs_c2[coding_pairs_c2$comp_has_ptc == TRUE, ]
ptc_pos_ps$strand <- strand_lookup_g3[ptc_pos_ps$comparator_isoform_id]
ptc_pos_ps$comp_stop <- ifelse(ptc_pos_ps$strand == "-",
  cds_start_lookup_g3[ptc_pos_ps$comparator_isoform_id],
  cds_stop_lookup_g3[ptc_pos_ps$comparator_isoform_id])
ptc_pos_ps$ref_stop <- ifelse(ptc_pos_ps$strand == "-",
  cds_start_lookup_g3[ptc_pos_ps$reference_isoform_id],
  cds_stop_lookup_g3[ptc_pos_ps$reference_isoform_id])
ptc_pos_ps$same_stop <- ptc_pos_ps$comp_stop == ptc_pos_ps$ref_stop

same_stop_ids <- ptc_pos_ps$comparator_isoform_id[ptc_pos_ps$same_stop == TRUE]
diff_stop_ids <- ptc_pos_ps$comparator_isoform_id[ptc_pos_ps$same_stop == FALSE]

# For diff-stop pairs, compute reference-based 3'UTR
diff_stop_ps <- ptc_pos_ps[ptc_pos_ps$same_stop == FALSE, ]
ref_based_utr3 <- setNames(rep(NA_integer_, nrow(diff_stop_ps)), diff_stop_ps$comparator_isoform_id)

for (i in seq_len(nrow(diff_stop_ps))) {
  comp_id <- diff_stop_ps$comparator_isoform_id[i]
  ref_stop_pos <- diff_stop_ps$ref_stop[i]
  comp_strand <- strand_lookup_g3[comp_id]
  if (is.na(comp_strand) || is.na(ref_stop_pos)) next

  str_row <- structures[structures$isoform_id == comp_id, ]
  if (nrow(str_row) == 0) next
  exon_s <- str_row$exon_starts[[1]]
  exon_e <- str_row$exon_ends[[1]]

  # Check ref stop is in comparator exons
  in_exon <- any(ref_stop_pos >= exon_s & ref_stop_pos <= exon_e)
  if (!in_exon) next

  utr3 <- 0L
  for (k in seq_along(exon_s)) {
    if (comp_strand == "+") {
      if (exon_e[k] > ref_stop_pos) {
        post_start <- max(exon_s[k], ref_stop_pos + 1L)
        utr3 <- utr3 + as.integer(exon_e[k] - post_start + 1L)
      }
    } else {
      if (exon_s[k] < ref_stop_pos) {
        pre_end <- min(exon_e[k], ref_stop_pos - 1L)
        utr3 <- utr3 + as.integer(pre_end - exon_s[k] + 1L)
      }
    }
  }
  ref_based_utr3[comp_id] <- utr3
}

n_eligible <- sum(!is.na(ref_based_utr3))
n_ineligible <- sum(is.na(ref_based_utr3))

# Lookup for comparator 3'UTR from divergence
utr3_comp_lk <- setNames(div_c2$utr3_bp_comp, div_c2$comparator_isoform_id)

# Build 5-group data
utr3_data <- rbind(
  data.frame(group = "Same-stop PTC+\n(3'UTR splicing)",
    utr3_bp = utr3_comp_lk[same_stop_ids]),
  data.frame(group = "Diff-stop PTC+\n(own stop, inflated)",
    utr3_bp = utr3_comp_lk[diff_stop_ids]),
  data.frame(group = "Diff-stop PTC+\n(ref stop, corrected)",
    utr3_bp = ref_based_utr3[!is.na(ref_based_utr3)]),
  data.frame(group = "NMD PTC-",
    utr3_bp = utr3_comp_lk[ptc_neg_ids]),
  data.frame(group = "Control",
    utr3_bp = div_c4$utr3_bp_comp)
)
utr3_data <- utr3_data[!is.na(utr3_data$utr3_bp) & utr3_data$utr3_bp > 0, ]
utr3_data$group <- factor(utr3_data$group,
  levels = c("Same-stop PTC+\n(3'UTR splicing)",
             "Diff-stop PTC+\n(own stop, inflated)",
             "Diff-stop PTC+\n(ref stop, corrected)",
             "NMD PTC-", "Control"))
utr3_n <- table(utr3_data$group)
```

PTC+ comparators are split by whether the strand-aware stop codon position is
the **same** as the reference or **different** (premature stop from frameshift
or in-frame stop in CDS). Stop codon position: `cds_stop` on + strand,
`cds_start` on - strand.

- **Same-stop PTC+** (`r fmt(utr3_n[[1]])`): The stop codon is the same as the
  reference. A 3'UTR splicing event repositioned the last EJC, making the
  existing stop >50 nt upstream of the new EJC. The 3'UTR length is correctly
  measured.
- **Diff-stop PTC+ (own stop)** (`r fmt(utr3_n[[2]])`): The stop codon moved
  upstream (PTC). The measured 3'UTR is inflated — it includes former CDS.
- **Diff-stop PTC+ (ref stop)** (`r fmt(utr3_n[[3]])`): Same pairs as above, but
  3'UTR is measured from the **reference's** stop codon position projected onto
  the comparator's exon structure. `r fmt(n_eligible)` of `r fmt(n_eligible + n_ineligible)`
  pairs were eligible (reference stop falls within comparator exons).
- **NMD PTC−** (`r fmt(utr3_n[[4]])`): No PTC. 3'UTR is correctly measured.
  *Example: GPAT3 — comp 3'UTR=1,108 bp, ref 3'UTR=1,108 bp.*
- **Control** (`r fmt(utr3_n[[5]])`): C4 comparators.
  *Example: SART1 — comp 3'UTR=1,107 bp, ref 3'UTR=1,107 bp.*

```{r goal3-fig6-utr3-density, fig.width=10, fig.height=6}
pal5_utr3 <- c("Same-stop PTC+\n(3'UTR splicing)" = "#d6604d",
               "Diff-stop PTC+\n(own stop, inflated)" = "#fc8d59",
               "Diff-stop PTC+\n(ref stop, corrected)" = "#fee08b",
               "NMD PTC-" = "#d9ef8b",
               "Control" = "#67a9cf")

fig6 <- ggplot(utr3_data, aes(x = utr3_bp, fill = group)) +
  geom_density(alpha = 0.5) +
  scale_x_log10(labels = scales::comma) +
  scale_fill_manual(values = pal5_utr3) +
  labs(title = "3'UTR Length by PTC Mechanism",
       subtitle = "Correcting for CDS truncation reveals PTC-neg NMD isoforms have normal 3'UTR lengths",
       x = "3'UTR length (bp, log scale)", y = "Density", fill = "")
save_fig(fig6, "fig7_utr3_length_density", width = 10, height = 6)
```

## Table 7: 3'UTR Length Comparisons (Wilcoxon)

```{r goal3-table6-utr3-tests}
grps <- split(utr3_data$utr3_bp, utr3_data$group)

comparisons_list <- list(
  c("Same-stop PTC+\n(3'UTR splicing)", "Control"),
  c("Diff-stop PTC+\n(own stop, inflated)", "Control"),
  c("Diff-stop PTC+\n(ref stop, corrected)", "Control"),
  c("NMD PTC-", "Control"),
  c("Same-stop PTC+\n(3'UTR splicing)", "Diff-stop PTC+\n(ref stop, corrected)")
)

tests <- do.call(rbind, lapply(comparisons_list, function(pair) {
  g1 <- grps[[pair[1]]]
  g2 <- grps[[pair[2]]]
  wt <- wilcox.test(g1, g2)
  U <- as.numeric(wt$statistic)
  rbc <- 1 - (2 * U) / (length(g1) * length(g2))
  data.frame(
    group1 = gsub("\n", " ", pair[1]),
    n_group1 = length(g1), median_group1 = round(median(g1)),
    group2 = gsub("\n", " ", pair[2]),
    n_group2 = length(g2), median_group2 = round(median(g2)),
    median_diff_bp = round(median(g1) - median(g2)),
    rank_biserial_r = round(rbc, 3),
    p_value = signif(wt$p.value, 3),
    stringsAsFactors = FALSE)
}))

render_dt(tests, caption = "Table 7: 3'UTR Length Comparisons (Wilcoxon rank-sum test)")
write.csv(tests, file.path(table_dir, "table7_utr3_wilcoxon.csv"), row.names = FALSE)
```

Columns: **group1/group2** = the two groups being compared (see Figure 7 for
group definitions); **n_group1/n_group2** = number of isoforms in each group;
**median_group1/median_group2** = median 3'UTR length (bp);
**median_diff_bp** = difference in medians (group1 − group2);
**rank_biserial_r** = rank-biserial correlation (effect size; positive =
group1 longer); **p_value** = Wilcoxon rank-sum test p-value.

## Illustrative Example: 3'UTR Measurement Correction (CMAS)

The CMAS gene illustrates why naive 3'UTR measurement is misleading for
PTC-positive isoforms. The NMD comparator (ENST00000534981.5) has a skipped exon
that causes a frameshift, creating a premature stop codon. When we measure 3'UTR
from this PTC, the sequence between the PTC and the normal stop is counted as
"3'UTR" — inflating the measurement by thousands of bases. The corrected
measurement uses the reference's stop codon position instead.

```{r goal3-utr3-example, fig.width=10, fig.height=6}
# CMAS: + strand, ref has 8 exons, comp has 7 (skipped exon 22061281-22061452)
# Same CDS start (22046304), different stop (ref: 22065311, comp: 22062284)

ref_s <- c(22046218, 22055149, 22055455, 22058567, 22060832, 22061281, 22062281, 22065121)
ref_e <- c(22046563, 22055291, 22055610, 22058700, 22060926, 22061452, 22062434, 22065668)
comp_s <- c(22046225, 22055149, 22055455, 22058567, 22060832, 22062281, 22065121)
comp_e <- c(22046563, 22055291, 22055610, 22058700, 22060926, 22062434, 22065674)

# Skipped exon event
se_event <- data.frame(
  five_prime = 22061281, three_prime = 22061452,
  event_type = "SE", direction = "LOSS")

# Annotations on a single panel:
# - PTC marker (red) on comparator at 22062284
# - Reference stop marker (green) on both tracks at 22065311
# - Shaded region: "inflated" 3'UTR (PTC to tx end) on comparator
# - Shaded region: "corrected" 3'UTR (ref stop to tx end) on comparator
ptc_marker <- list(type = "ptc_marker", track = "comp",
  x = 22062284, label = "PTC (comp stop)", color = "#E74C3C")
ref_stop_ref <- list(type = "ptc_marker", track = "ref",
  x = 22065311, label = "Stop", color = "#27AE60")
ref_stop_comp <- list(type = "ptc_marker", track = "comp",
  x = 22065311, label = "Ref stop", color = "#27AE60")

# Region shades to illustrate the two measurements
inflated_region <- list(type = "region_shade", track = "comp",
  xmin = 22062284, xmax = 22065311,
  color = "#E74C3C", alpha = 0.15,
  label = "Former CDS (inflates 3'UTR)")
corrected_region <- list(type = "region_shade", track = "comp",
  xmin = 22065311, xmax = 22065674,
  color = "#27AE60", alpha = 0.20,
  label = "True 3'UTR")

p <- plot_ucsc_pair(
  ref_s, ref_e, comp_s, comp_e, strand = "+",
  ref_cds = list(cds_start = 22046304, cds_stop = 22065311),
  comp_cds = list(cds_start = 22046304, cds_stop = 22062284),
  ref_label = "Reference (ENST00000229329.7)",
  comp_label = "NMD Comparator (ENST00000534981.5)",
  events = se_event,
  annotations = list(ptc_marker, ref_stop_ref, ref_stop_comp,
                     inflated_region, corrected_region),
  title = "CMAS: Naive vs Corrected 3'UTR Measurement",
  subtitle = "Red shading = former CDS counted as 3'UTR when measured from PTC. Green = true 3'UTR from ref stop.")

print(p)
```

The skipped exon (SE LOSS bracket) causes a frameshift that creates a PTC (red
triangle). Measuring 3'UTR from this PTC includes the red-shaded region — former
coding sequence that inflates the measurement. The green triangle marks the
reference stop codon position; measuring from there gives the true 3'UTR
(green-shaded region). This correction is applied to all diff-stop PTC+ pairs
in Figure 7 and Table 7 above.

**Key finding:** PTC-negative NMD comparators have 3'UTR lengths
indistinguishable from Control (p = `r signif(tests$p_value[tests$group1 == "NMD PTC-"], 2)`).
This is strong evidence against the long-3'UTR hypothesis as an explanation for
PTC-negative NMD in this dataset.

The apparently long 3'UTRs in some PTC+ groups are methodological artifacts:
for diff-stop PTC+ isoforms, the "3'UTR" includes former coding sequence
downstream of the premature stop codon. When corrected using the reference stop
position, even these isoforms show only a modest difference from Control. The
CMAS gene example below illustrates this measurement issue.

## 3'UTR Splicing: PTC-Creating, Not PTC-Independent

A separate question is whether 3'UTR splicing contributes to NMD — and if so,
whether it operates through a PTC-dependent or PTC-independent pathway. If 3'UTR
splicing triggers NMD independently of PTCs, we would expect enrichment of 3'UTR
splicing events in PTC-negative NMD comparators. If instead it triggers NMD by
repositioning exon-exon junctions (creating a PTC under the 50-nt rule), we
would expect enrichment only in PTC-positive comparators.

```{r goal3-utr3-splicing}
# ── Data provenance ──
# Source: computed inline using count_utr3_splicing() defined in this chunk
# Methods: METHODS.md § "3'UTR Splicing Enrichment"
# Starting data: gene-matched C2 and C4 profiles (detailed_events from
#   Isopair::buildProfiles() → compareSplicingChoices) + cds.rds (Isopair::extractCds)

# Strand-aware stop codon position for 3'UTR event detection
# + strand: stop = cds_stop; - strand: stop = cds_start
cds_start_lookup_utr3 <- setNames(cds$cds_start, cds$isoform_id)
cds_stop_lookup_utr3  <- setNames(cds$cds_stop, cds$isoform_id)
strand_lookup <- setNames(cds$strand, cds$isoform_id)

count_utr3_splicing <- function(profiles, comp_id_subset = NULL) {
  if (!is.null(comp_id_subset)) {
    idx <- which(profiles$comparator_isoform_id %in% comp_id_subset)
  } else {
    idx <- seq_len(nrow(profiles))
  }
  n_with <- 0L
  n_total <- 0L
  for (i in idx) {
    comp_id <- profiles$comparator_isoform_id[i]
    de <- profiles$detailed_events[[i]]
    comp_strand <- strand_lookup[comp_id]
    comp_stop <- if (!is.na(comp_strand) && comp_strand == "-") {
      cds_start_lookup_utr3[comp_id]
    } else {
      cds_stop_lookup_utr3[comp_id]
    }
    if (is.na(comp_stop) || is.na(comp_strand)) next
    n_total <- n_total + 1L
    if (is.null(de) || nrow(de) == 0) next
    if (comp_strand == "+") {
      utr3_ev <- de[de$five_prime >= comp_stop | de$three_prime >= comp_stop, ]
    } else {
      utr3_ev <- de[de$five_prime <= comp_stop | de$three_prime <= comp_stop, ]
    }
    if (nrow(utr3_ev) > 0) n_with <- n_with + 1L
  }
  list(n_with = n_with, n_total = n_total,
       pct = round(100 * n_with / max(n_total, 1), 1))
}

prof_c4_as <- set_results[["c4_allsamples"]]$profiles

utr3_ptcneg <- count_utr3_splicing(prof_c2, ptc_neg_ids)
utr3_ptcpos <- count_utr3_splicing(prof_c2, ptc_pos_comp_ids)
utr3_ctrl   <- count_utr3_splicing(prof_c4_as)

ft_utr3_neg <- fisher.test(matrix(c(
  utr3_ptcneg$n_with, utr3_ptcneg$n_total - utr3_ptcneg$n_with,
  utr3_ctrl$n_with, utr3_ctrl$n_total - utr3_ctrl$n_with), nrow = 2))
ft_utr3_pos <- fisher.test(matrix(c(
  utr3_ptcpos$n_with, utr3_ptcpos$n_total - utr3_ptcpos$n_with,
  utr3_ctrl$n_with, utr3_ctrl$n_total - utr3_ctrl$n_with), nrow = 2))

utr3_summary <- data.frame(
  group = c("NMD PTC-", "NMD PTC+", "Control"),
  n_with_utr3_event = c(utr3_ptcneg$n_with, utr3_ptcpos$n_with, utr3_ctrl$n_with),
  n_total = c(utr3_ptcneg$n_total, utr3_ptcpos$n_total, utr3_ctrl$n_total),
  pct = c(utr3_ptcneg$pct, utr3_ptcpos$pct, utr3_ctrl$pct),
  vs_control_OR = c(round(as.numeric(ft_utr3_neg$estimate), 2),
                     round(as.numeric(ft_utr3_pos$estimate), 2), NA),
  vs_control_p = c(signif(ft_utr3_neg$p.value, 3),
                    signif(ft_utr3_pos$p.value, 3), NA))

kable(utr3_summary,
      caption = "3'UTR Splicing Prevalence by Group",
      col.names = c("Group", "With 3'UTR Event", "Total", "%",
                     "OR vs Control", "p vs Control"))
```

Columns: **Group**; **With 3'UTR Event** = pairs where ≥1 splice event has
coordinates in the 3'UTR (downstream of the strand-aware stop codon position);
**OR/p vs Control** = Fisher's exact test comparing to Control (C4).

```{r goal3-utr3-same-stop-verification}
# ── Direct verification: 3'UTR splicing creates PTCs via EJC repositioning ──
# For same-stop PTC+ pairs, the stop codon is identical to the reference.
# The PTC status arises because the comparator has more exon-exon junctions
# downstream of the stop codon (from 3'UTR splicing), repositioning the last EJC.
# We can verify by comparing n_downstream_ejcs between ref and comp.

same_stop_ps <- ptc_pos_ps[ptc_pos_ps$same_stop == TRUE, ]
ptc_ejc_lookup <- setNames(ptc$n_downstream_ejcs, ptc$isoform_id)

same_stop_ps$comp_ejcs <- ptc_ejc_lookup[same_stop_ps$comparator_isoform_id]
same_stop_ps$ref_ejcs <- ptc_ejc_lookup[same_stop_ps$reference_isoform_id]
same_stop_ps$ejc_diff <- same_stop_ps$comp_ejcs - same_stop_ps$ref_ejcs

n_comp_more_ejcs <- sum(same_stop_ps$ejc_diff > 0, na.rm = TRUE)
n_same_ejcs <- sum(same_stop_ps$ejc_diff == 0, na.rm = TRUE)
n_comp_fewer_ejcs <- sum(same_stop_ps$ejc_diff < 0, na.rm = TRUE)
n_valid <- sum(!is.na(same_stop_ps$ejc_diff))
```

### 3'UTR Splicing Creates PTCs via EJC Repositioning

Among the `r fmt(nrow(same_stop_ps))` same-stop PTC+ pairs (where the
comparator has the same stop codon as the reference but is classified as PTC+),
we directly verified the mechanism by comparing the number of exon-exon
junctions downstream of the stop codon:

- **`r n_comp_more_ejcs`** pairs (`r round(100 * n_comp_more_ejcs / n_valid, 1)`%):
  comparator has MORE downstream EJCs than reference — confirming that 3'UTR
  splicing added new junctions, repositioning the last EJC and creating the PTC
- **`r n_same_ejcs`** pairs: same number of downstream EJCs (PTC may arise from
  junction repositioning rather than addition)
- **`r n_comp_fewer_ejcs`** pairs: comparator has fewer EJCs (unexpected —
  may reflect annotation differences)

### Conclusion: 3'UTR Mechanisms Operate Through PTC Creation

Three lines of evidence converge on the same conclusion — 3'UTR features do
not independently trigger NMD in this dataset, but 3'UTR splicing does
contribute to NMD by creating PTCs via EJC repositioning:

1. **No NMD subgroup has substantially longer 3'UTRs than Control when measured
   from the appropriate stop codon** (Table 7). PTC-negative NMD comparators
   have the same median 3'UTR as Control (p =
   `r signif(tests$p_value[tests$group1 == "NMD PTC-"], 2)`). Same-stop PTC+
   comparators are also indistinguishable from Control (p =
   `r signif(tests$p_value[tests$group1 == "Same-stop PTC+ (3'UTR splicing)" & tests$group2 == "Control"], 2)`).
   The diff-stop PTC+ group measured at own stop has longer 3'UTRs (p < 10^-54^),
   but this is an artifact of measuring from the PTC rather than the true stop.
   When corrected using the reference stop position, the effect is small (median
   difference `r tests$median_diff_bp[tests$group1 == "Diff-stop PTC+ (ref stop, corrected)" & tests$group2 == "Control"]`
   bp, r = `r tests$rank_biserial_r[tests$group1 == "Diff-stop PTC+ (ref stop, corrected)" & tests$group2 == "Control"]`,
   p = `r signif(tests$p_value[tests$group1 == "Diff-stop PTC+ (ref stop, corrected)" & tests$group2 == "Control"], 2)`).

2. **3'UTR splicing that triggers NMD does so by creating PTCs** (EJC
   repositioning), not through a PTC-independent pathway. The
   `r fmt(nrow(same_stop_ps))` same-stop PTC+ pairs are direct evidence:
   `r round(100 * n_comp_more_ejcs / n_valid, 1)`% have additional downstream
   EJCs in the comparator from 3'UTR splicing events.

3. **3'UTR splicing prevalence** is actually **depleted** in PTC-negative NMD
   vs Control (OR =
   `r round(as.numeric(ft_utr3_neg$estimate), 2)`, p =
   `r signif(ft_utr3_neg$p.value, 3)`). This is consistent with our finding
   that 3'UTR splicing creates PTCs — isoforms that trigger NMD without a PTC
   would be expected to have *less* 3'UTR splicing, not more.

# 2c. Start Codons and ORFs: The 5'UTR Landscape {#sec2c-start-codons}

The unified model's second-largest gain comes from 5'UTR features (Section 2a:
AUC `r round(unified$elastic_net$step2$auc_test, 3)` →
`r round(unified$elastic_net$step3$auc_test, 3)` when adding ORF features).
What about the 5'UTR predicts NMD, and why do overlapping uORFs matter?

**Data source:** Gene-matched C2 vs C4 coding pairs (paired framework). 5'UTR
features from utr5_features.rds. uORF set operations from uorf_comparison.rds.
Splice event co-occurrence from paired profiles.

Upstream open reading frames (uORFs) in the 5'UTR can trigger NMD when
translation of the uORF terminates and the ribosome encounters downstream
exon-exon junctions from the main CDS. Rather than testing a single feature,
we scan 28 5'UTR sequence features and ask which best distinguish
PTC-negative NMD comparators from Controls.

```{r goal4b-utr5-load}
# ── Data provenance ──
# Source script: 05j_utr5_feature_analysis.R → analysis_cache/utr5_features.rds
# Pipeline:
#   1. Isopair::scan5UtrFeatures() computes 28 features per isoform:
#      - ATG/stop codon counts (total and frame-stratified relative to CDS)
#      - ORF counts (in-frame, out-of-frame, overlapping)
#      - Density features (per 100bp), Kozak context, spatial features
#   2. 5'UTR boundaries: strand-aware from genomic CDS + exon structure
#      Validation: 98.6% of isoforms have ATG at computed CDS start position
#   3. 5'UTR sequences extracted from SQANTI corrected transcript FASTA
#   4. Groups: PTC_neg, PTC_pos, Control (same coding pair definitions as §4)
#   5. Univariate Wilcoxon tests (3 comparisons × 28 features)
#   6. Multivariate elastic net (3 models)
#   7. Pair-level delta analysis (comp - ref feature differences)

utr5_data <- readRDS(file.path(cache_dir, "utr5_features.rds"))
utr5_feat <- utr5_data$isoform_features
utr5_feat$group <- factor(utr5_feat$group, levels = c("PTC_neg", "PTC_pos", "Control"))

# Also load the old uorf_analysis.rds for backward-compatible uorf counts
uorf_data <- readRDS(file.path(cache_dir, "uorf_analysis.rds"))
uorf <- uorf_data$per_isoform
uorf$group <- factor(uorf$group, levels = c("PTC_neg", "PTC_pos", "Control"))
```

## Table 8a: 5'UTR Feature Summary by Group

The table below provides an overview of 5'UTR characteristics across the three
groups. If uORFs are a mechanism for PTC-negative NMD, we expect PTC-negative
comparators to have longer 5'UTRs, more uORFs, and specifically more
*overlapping* uORFs (whose stop codon extends into the main CDS, forcing the
ribosome to encounter downstream exon-exon junctions).

```{r goal4b-feature-summary}
# ── Data provenance ──
# Source: utr5_feat (from utr5_features.rds, loaded above)
# Filtered to comparator isoforms (non-NA group) with non-excluded 5'UTRs

comp_feat <- utr5_feat |> filter(!excluded, !is.na(group))

utr5_summary <- comp_feat |>
  group_by(group) |>
  summarise(
    n = n(),
    median_utr5_bp = median(utr5_length),
    median_n_atg = median(n_atg),
    median_n_orfs = median(n_orfs),
    pct_with_uorf = round(100 * mean(n_orfs > 0), 1),
    median_n_orfs_outframe = median(n_orfs_outframe),
    median_n_overlap = median(n_orfs_overlapping),
    pct_overlapping = round(100 * mean(n_orfs_overlapping > 0), 1),
    median_pct_in_orfs = round(median(pct_utr5_in_orfs), 3),
    pct_fully_covered = round(100 * mean(pct_utr5_in_orfs >= 1.0), 1),
    median_longest_orf = median(longest_orf_nt),
    median_strong_kozak = median(n_strong_kozak_atg),
    .groups = "drop")

render_dt(utr5_summary, caption = "Table 8a: 5'UTR Feature Summary by Group")
```

Columns: **group** = PTC_neg (NMD without PTC), PTC_pos (NMD with PTC), or
Control (C4); **n** = isoforms; **median_utr5_bp** = median 5'UTR length (bp);
**median_n_atg** = median ATG count in 5'UTR; **median_n_orfs** = median uORF count
(ATG + in-frame stop, >=9 nt); **pct_with_uorf** = % with >=1 uORF;
**median_n_orfs_outframe** = median out-of-frame uORFs (frame 1 or 2 relative
to CDS); **median_n_overlap** = median overlapping uORFs (stop at/past CDS start);
**pct_overlapping** = % with >=1 overlapping uORF;
**median_pct_in_orfs** = median fraction of 5'UTR occupied by uORFs;
**pct_fully_covered** = % of isoforms where entire 5'UTR is covered by uORFs;
**median_longest_orf** = median length of longest uORF (nt);
**median_strong_kozak** = median count of ATGs with strong Kozak context
(A/G at -3 AND G at +4).

The contrast is striking: PTC-negative NMD comparators have a median of 8 uORFs
compared to 1 in Controls, with nearly all (96.8%) containing at least one uORF.
The `median_pct_in_orfs` column shows that a median of 67.7% of the 5'UTR
sequence is covered by at least one uORF (computed as the union of all
uORF intervals, with no double-counting of overlapping regions), compared to
27.2% in Controls. Overlapping uORFs — those whose stop codon falls at or past
the main CDS start — are found in 61.9% of PTC-negative comparators vs 28.7%
of Controls.

## Table 8a-stop: Stop Codon Distribution in 5'UTR

Stop codons in the 5'UTR determine where uORFs terminate. Higher stop density
means more opportunities for ATG-initiated reading frames to encounter an
in-frame stop, producing complete uORFs. The three stop codons (TAA, TAG, TGA)
may also differ in read-through efficiency, potentially affecting whether a
uORF triggers NMD or is read through by the ribosome.

```{r goal4b-stop-summary}
stop_summary <- comp_feat |>
  group_by(group) |>
  summarise(
    n = n(),
    median_stops = median(n_stop_total),
    mean_stops = round(mean(n_stop_total), 1),
    median_taa = median(n_taa),
    median_tag = median(n_tag),
    median_tga = median(n_tga),
    median_stop_density = round(median(stop_density), 2),
    .groups = "drop")

render_dt(stop_summary, caption = "Table 8a-stop: 5'UTR Stop Codon Distribution by Group")
```

Columns: **median_stops/mean_stops** = total stop codons (TAA+TAG+TGA) in 5'UTR;
**median_taa/tag/tga** = individual stop codon counts; **median_stop_density** =
stops per 100 bp.

Stop codon composition matters for uORF biology because each uORF requires an
in-frame stop to terminate. TGA is the most abundant stop codon in 5'UTRs across
all groups (~55% of all stops), consistent with known nucleotide composition biases
in non-coding sequence. PTC-negative comparators have higher stop density per 100 bp
(2.86 vs 2.48 in Control) — more uORF termination opportunities per unit sequence,
not just more stops from having longer 5'UTRs.

## Table 8b: uORF Enrichment Tests

The features above describe group-level medians. The table below tests whether
the *prevalence* of uORFs — the fraction of isoforms with at least one uORF —
differs between groups. This answers a simpler question than the elastic net:
not which features predict NMD, but whether having any uORF at all is enriched.

```{r goal4b-uorf-tests}
# ── Data provenance ──
# Source: uorf (from uorf_analysis.rds) for backward-compatible Fisher's tests
# "any uORF": >=1 uORF (>=3 codons), "overlapping uORF": stop at/past CDS start

uorf_tests <- list()
test_pairs <- list(
  c("PTC_neg", "Control", "any uORF"),
  c("PTC_pos", "Control", "any uORF"),
  c("PTC_neg", "PTC_pos", "any uORF"),
  c("PTC_neg", "Control", "overlapping uORF")
)

for (tp in test_pairs) {
  sub <- uorf[uorf$group %in% tp[1:2], ]
  sub$group <- droplevels(factor(sub$group))
  if (tp[3] == "any uORF") {
    has_feat <- sub$n_uorfs > 0
  } else {
    has_feat <- sub$n_overlapping_uorfs > 0
  }
  n1 <- sum(sub$group == tp[1])
  n2 <- sum(sub$group == tp[2])
  pct1 <- round(100 * mean(has_feat[sub$group == tp[1]]), 1)
  pct2 <- round(100 * mean(has_feat[sub$group == tp[2]]), 1)

  # Build table with group1 having the feature as the first row/col
  # so OR > 1 means group1 enriched
  tbl <- matrix(c(
    sum(has_feat & sub$group == tp[1]),
    sum(!has_feat & sub$group == tp[1]),
    sum(has_feat & sub$group == tp[2]),
    sum(!has_feat & sub$group == tp[2])
  ), nrow = 2, byrow = TRUE)

  ft <- tryCatch(fisher.test(tbl), error = function(e) NULL)
  if (!is.null(ft)) {
    uorf_tests[[length(uorf_tests) + 1]] <- data.frame(
      comparison = paste(tp[1], "vs", tp[2]),
      test = tp[3],
      n_group1 = n1, pct_group1 = pct1,
      n_group2 = n2, pct_group2 = pct2,
      OR = round(as.numeric(ft$estimate), 2),
      p_value = signif(as.numeric(ft$p.value), 3),
      stringsAsFactors = FALSE)
  }
}
uorf_test_df <- do.call(rbind, uorf_tests)
render_dt(uorf_test_df, caption = "Table 8b: uORF Enrichment Tests (Fisher's exact)")
```

Columns: **comparison** = group pair tested; **test** = "any uORF" (>=1 uORF) or
"overlapping uORF" (>=1 uORF with stop at/past CDS start); **n_group1/n_group2** =
group sizes; **pct_group1/pct_group2** = prevalence (%) in each group;
**OR** = odds ratio for having the feature in group1 vs group2 (OR > 1 = group1
enriched); **p_value** = Fisher's exact test p-value.

## Figure 8a: 5'UTR Length Distribution

```{r goal4b-utr5-density, fig.width=10, fig.height=6}
# ── Data provenance ──
# Source: uorf (from uorf_analysis.rds)
# Filter: utr5_length > 0

uorf_summary <- uorf |>
  group_by(group) |>
  summarise(
    median_utr5_bp = median(utr5_length),
    median_n_uorfs = median(n_uorfs),
    .groups = "drop")

utr5_plot <- uorf[uorf$utr5_length > 0, ]
utr5_plot$group_label <- factor(utr5_plot$group,
  levels = c("PTC_neg", "PTC_pos", "Control"),
  labels = c("NMD PTC-", "NMD PTC+", "Control"))

fig_utr5 <- ggplot(utr5_plot, aes(x = utr5_length, fill = group_label)) +
  geom_density(alpha = 0.5) +
  scale_x_log10(labels = scales::comma) +
  scale_fill_manual(values = c("NMD PTC-" = "#fddbc7", "NMD PTC+" = "#ef8a62",
                                "Control" = "#67a9cf")) +
  labs(title = "5'UTR Length Distribution by Group",
       subtitle = paste0(
         "Median 5'UTR length: NMD PTC- = ",
         uorf_summary$median_utr5_bp[uorf_summary$group == "PTC_neg"], " bp, ",
         "NMD PTC+ = ",
         uorf_summary$median_utr5_bp[uorf_summary$group == "PTC_pos"], " bp, ",
         "Control = ",
         uorf_summary$median_utr5_bp[uorf_summary$group == "Control"], " bp"),
       x = "5'UTR Length (bp, log scale)", y = "Density", fill = "")
save_fig(fig_utr5, "fig8a_utr5_length_distribution", width = 10, height = 6)
```

**Figure 8a** shows that PTC-negative NMD comparators have dramatically longer
5'UTRs than either PTC-positive or Control isoforms (median 724 bp vs 243 and
265 bp). This length difference is a necessary precondition for the uORF
enrichment — longer 5'UTRs provide more sequence in which ATGs and in-frame stops
can occur. However, as we show below, the enrichment goes beyond what length alone
would predict: the *density* of uORF-related features is also elevated.

## Figure 8b: uORF Prevalence Barplot

```{r goal4b-uorf-prevalence-barplot, fig.width=10, fig.height=6}
uorf_prev <- uorf |>
  group_by(group) |>
  summarise(pct = round(100 * mean(n_uorfs > 0), 1), .groups = "drop")
uorf_prev$group_label <- factor(uorf_prev$group,
  levels = c("PTC_neg", "PTC_pos", "Control"),
  labels = c("NMD PTC-", "NMD PTC+", "Control"))

any_uorf_tests <- uorf_test_df[uorf_test_df$test == "any uORF", ]
p_neg_ctrl <- signif(any_uorf_tests$p_value[any_uorf_tests$comparison == "PTC_neg vs Control"], 3)
p_pos_ctrl <- signif(any_uorf_tests$p_value[any_uorf_tests$comparison == "PTC_pos vs Control"], 3)
p_neg_pos  <- signif(any_uorf_tests$p_value[any_uorf_tests$comparison == "PTC_neg vs PTC_pos"], 3)

bar_x <- setNames(1:3, c("NMD PTC-", "NMD PTC+", "Control"))
ymax <- max(uorf_prev$pct) + 15

brackets <- data.frame(
  x1 = c(bar_x["NMD PTC-"], bar_x["NMD PTC+"], bar_x["NMD PTC-"]),
  x2 = c(bar_x["Control"],  bar_x["Control"],  bar_x["NMD PTC+"]),
  y  = c(ymax - 10, ymax - 5, ymax),
  label = c(paste0("p = ", p_neg_ctrl), paste0("p = ", p_pos_ctrl), paste0("p = ", p_neg_pos))
)

fig_prev <- ggplot(uorf_prev, aes(x = group_label, y = pct, fill = group_label)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(pct, "%")), vjust = -0.5, size = 4.5) +
  scale_fill_manual(values = c("NMD PTC-" = "#fddbc7", "NMD PTC+" = "#ef8a62",
                                "Control" = "#67a9cf")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)), limits = c(0, ymax + 3)) +
  geom_segment(data = brackets, aes(x = x1, xend = x2, y = y, yend = y),
               inherit.aes = FALSE, linewidth = 0.4) +
  geom_segment(data = brackets, aes(x = x1, xend = x1, y = y - 1, yend = y),
               inherit.aes = FALSE, linewidth = 0.4) +
  geom_segment(data = brackets, aes(x = x2, xend = x2, y = y - 1, yend = y),
               inherit.aes = FALSE, linewidth = 0.4) +
  geom_text(data = brackets, aes(x = (x1 + x2) / 2, y = y + 1.5, label = label),
            inherit.aes = FALSE, size = 3.5) +
  labs(title = "uORF Prevalence by Group",
       subtitle = "Percentage of isoforms with at least one uORF (>=3 codons)",
       x = "", y = "% with >=1 uORF", fill = "") +
  theme(legend.position = "none")
save_fig(fig_prev, "fig8b_uorf_prevalence_barplot", width = 10, height = 6)
```

**Figure 8b** shows uORF prevalence. Nearly all PTC-negative NMD comparators
(96.2%) have at least one uORF, compared to 63.3% of Controls (OR = 14.6,
p < 10^-180^). Notably, PTC-positive comparators are *less* likely to have uORFs
than Controls (57.2% vs 63.3%, OR = 0.8) — suggesting that the uORF mechanism
is specific to the PTC-negative group, not a general property of NMD-responsive
isoforms.

## Figure 8c: uORF Count Distribution

```{r goal4b-uorf-density, fig.width=10, fig.height=6}
uorf_plot <- uorf[uorf$utr5_length > 0, ]
uorf_plot$group_label <- factor(uorf_plot$group,
  levels = c("PTC_neg", "PTC_pos", "Control"),
  labels = c("NMD PTC-", "NMD PTC+", "Control"))

fig_uorf <- ggplot(uorf_plot, aes(x = n_uorfs, fill = group_label)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 1, alpha = 0.5,
                 position = "identity") +
  scale_x_continuous(breaks = c(0, 5, 10, 15, 20, 25, 30), limits = c(-0.5, 35)) +
  scale_fill_manual(values = c("NMD PTC-" = "#fddbc7", "NMD PTC+" = "#ef8a62",
                                "Control" = "#67a9cf")) +
  labs(title = "uORF Count Distribution by Group",
       subtitle = "PTC-negative NMD isoforms have dramatically more uORFs (median 8 vs 1 in Control)",
       x = "Number of uORFs", y = "Density", fill = "")
save_fig(fig_uorf, "fig8c_uorf_count_distribution", width = 10, height = 6)
```

## Which 5'UTR Features Best Predict NMD Status?

The univariate analyses above show that virtually every 5'UTR feature is elevated
in PTC-negative NMD comparators — but many of these features are correlated with
each other (longer 5'UTR = more ATGs = more ORFs). To identify which features
carry *independent* predictive information, we fit an elastic net logistic
regression (alpha=0.5, 10-fold cross-validation) with 21 features, including
frame-stratified counts and individual stop codon types.
The model achieves a 10-fold cross-validated AUC of `r round(utr5_data$multivariate$cv_auc_a$cvm[which(utr5_data$multivariate$cv_auc_a$lambda == utr5_data$multivariate$cv_auc_a$lambda.min)], 3)` for distinguishing PTC-negative from Control comparators based on 5'UTR features alone — strong predictive power from sequence features. By contrast, the same features achieve AUC = `r round(utr5_data$multivariate$cv_auc_b$cvm[which(utr5_data$multivariate$cv_auc_b$lambda == utr5_data$multivariate$cv_auc_b$lambda.min)], 3)` for PTC-positive vs Control (barely above chance), confirming that 5'UTR features are specific to the PTC-negative NMD mechanism.

```{r goal4b-feature-importance, fig.width=10, fig.height=7}
# ── Data provenance ──
# Source: utr5_data$multivariate$coefs_a (elastic net coefficients)
# Model A: PTC_neg vs Control, elastic net (alpha=0.5), 10-fold CV
# Features: frame-stratified ATG/stop counts, individual stop codon types,
#   ORF counts, density, Kozak, longest/total ORF, orphan ATGs
# Excludes totals (n_atg, n_stop_total) to avoid multicollinearity with components

coefs_a <- utr5_data$multivariate$coefs_a
coefs_a <- coefs_a[order(coefs_a$coefficient), ]
coefs_a$feature <- factor(coefs_a$feature, levels = coefs_a$feature)
coefs_a$direction <- ifelse(coefs_a$coefficient > 0, "Positive", "Negative")

fig_coefs <- ggplot(coefs_a, aes(x = coefficient, y = feature, fill = direction)) +
  geom_col() +
  scale_fill_manual(values = c("Positive" = "#ef8a62", "Negative" = "#67a9cf")) +
  labs(title = "Elastic Net Feature Importance: PTC-neg vs Control",
       subtitle = paste0("Model A: ", nrow(coefs_a), " non-zero features at lambda.min; ",
                          "positive = enriched in PTC-neg"),
       x = "Coefficient", y = "", fill = "") +
  theme(legend.position = "bottom")
save_fig(fig_coefs, "fig8d_feature_importance", width = 10, height = 7)
```

The figure reveals several important patterns:

1. **`n_orfs_overlapping`** (uORFs whose stop codon falls at or past the main
   CDS start) is the strongest predictor. Overlapping uORFs are particularly
   potent NMD triggers because the ribosome translating them necessarily
   encounters downstream exon-exon junctions from the main CDS — exactly the
   configuration that activates the NMD pathway.

2. **`atg_density`** (ATGs per 100 bp) is the second-strongest feature. This
   captures how densely packed the 5'UTR is with potential translation start
   sites, independent of total 5'UTR length.

3. **`pct_utr5_in_orfs`** (fraction of 5'UTR covered by uORFs, computed as
   the union of all uORF intervals without double-counting) ranks third. A
   median 67.7% of the 5'UTR in PTC-negative comparators is covered by at
   least one uORF, vs 27.2% in Controls.

4. **`n_strong_kozak_atg`** (ATGs with strong initiation context: A/G at -3 and
   G at +4) is independently predictive. This suggests that translational
   *efficiency* of uORFs matters — not just their presence but whether ribosomes
   actually initiate at them.

5. **`stop_density`** (stop codons per 100 bp) has a positive coefficient.
   More stops per unit sequence means more opportunities for uORFs to terminate,
   creating more complete ORFs that can trigger NMD.

6. **Frame-specific signals:** In-frame ATGs (`n_atg_frame0`) have a *negative*
   coefficient — consistent with the biology that in-frame uORFs may represent
   N-terminal extensions that do not independently terminate. Out-of-frame
   uORFs must terminate in a different frame from the main CDS, making them
   bona fide NMD triggers.

7. **Stop codon identity:** TAG has a negative coefficient while TGA is positive.
   The biological significance of this is unclear but may relate to differences
   in read-through efficiency between stop codons.

## uORF Detection Validation Against ORFik

```{r goal4b-orfik-validation}
# ── Data provenance ──
# Source: uorf_comparison.rds (05i_uorf_orfik_analysis.R)
# Our method: detectUorfs(method="simple") walks stop codons through full transcript
# ORFik: findORFs(startCodon="ATG", longestORF=FALSE, minimumLength=0) on 5'UTR-only

uorf_comp <- readRDS(file.path(cache_dir, "uorf_comparison.rds"))
ov <- uorf_comp$orfik_validation
```

Our uORF detection method walks the reading frame through the **full transcript
sequence**, not just the 5'UTR. This is necessary to detect **overlapping uORFs**
— those whose ATG is in the 5'UTR but whose stop codon falls in or past the
main CDS. Overlapping uORFs are the top elastic net predictor of PTC-negative
NMD status, so detecting them is essential.

To validate our method, we compared non-overlapping uORF counts (where both
methods should agree) against ORFik's `findORFs()` applied to 5'UTR-only
sequences. Across `r fmt(ov$n_compared)` isoforms, the per-isoform uORF counts
show a Spearman correlation of **`r round(ov$spearman_r, 3)`**, with
`r fmt(ov$n_agree)` isoforms (`r round(100 * ov$n_agree / ov$n_compared, 1)`%)
having identical counts between methods. For the remaining
`r fmt(ov$n_orfik_more)` isoforms, ORFik detected additional short uORFs near
the 3' end of the 5'UTR where our method's stop-codon walk extends past the
5'UTR boundary into the CDS. Our method found zero isoforms with more
non-overlapping uORFs than ORFik.

Beyond this concordance, our method uniquely detected
**`r fmt(ov$n_overlapping_ours)` overlapping uORFs** (ATG in 5'UTR, stop codon
in or past the main CDS) that ORFik cannot find from 5'UTR-only input. These
overlapping uORFs are the most biologically important category: they are the
top elastic net predictor and show the strongest enrichment in PTC-negative
NMD comparators.

## uORF Set Operations: Gained, Lost, and Shared uORFs

The feature analysis above counts uORFs per comparator isoform. Here we go
further by matching individual uORFs between the reference and comparator
within each gene-matched pair, using their shared ATG genomic position. This
identifies which uORFs are *gained* by the comparator (present in comparator
but absent from the reference), *lost*, or *shared*. Critically, we stratify
gained uORFs by overlapping status — the category most associated with NMD.

```{r goal4b-uorf-setops}
# ── Data provenance ──
# Source: uorf_comparison.rds → group_summaries (stratified by overlapping status)
# Gained = present in comparator but not reference (by ATG genomic position)
# Lost = present in reference but not comparator
# Overlapping = uORF stop codon at or past main CDS start

gs <- uorf_comp$group_summaries

render_dt(gs, caption = "Table 8c: uORF Set Operations by Group (with Overlapping Stratification)")
```

For each isoform pair, we compared individual uORFs between reference and
comparator by matching on ATG genomic position. uORFs present only in the
comparator are "gained"; those only in the reference are "lost." Gained uORFs
are further stratified by whether they overlap the main CDS.

Columns: **group** = PTC_neg, PTC_pos, or Control; **pct_any_gained** = % of
pairs where the comparator gains at least one uORF; **pct_gained_overlapping** =
% gaining at least one overlapping uORF; **pct_gained_nonoverlapping** = %
gaining non-overlapping uORFs; **median_gained/median_gained_overlap** = median
counts; **pct_any_lost/pct_shared** = % with lost or shared uORFs.

The key finding: **`r gs$pct_gained_overlapping[gs$group == "PTC_neg"]`% of
PTC-negative NMD pairs gain at least one overlapping uORF**, compared to
`r gs$pct_gained_overlapping[gs$group == "Control"]`% of Control pairs (Fisher's
exact OR = `r round(uorf_comp$event_enrichment$OR[1], 1)` — see below). This is
a 5.4-fold enrichment. PTC-positive pairs are actually *depleted* for gained
overlapping uORFs (9.7%), consistent with their NMD being driven by the PTC
rather than 5'UTR features.

## Splice Event Co-Occurrence

If specific types of splicing changes are responsible for creating overlapping
uORFs, those event types should be enriched in PTC-negative pairs that gain
overlapping uORFs compared to PTC-negative pairs that do not. The table below
tests this by comparing splice event prevalence between the two subgroups within
PTC-negative NMD pairs.

```{r goal4b-event-cooccurrence}
# ── Data provenance ──
# Source: uorf_comparison.rds → event_enrichment
# For PTC_neg pairs: event type frequencies in pairs with gained overlapping uORFs
# vs pairs without. Fisher's exact test for enrichment.

ee <- uorf_comp$event_enrichment
ee$p_adj <- p.adjust(ee$p_value, method = "BH")
ee <- ee[order(ee$p_value), ]

render_dt(ee, caption = "Table 8d: Splice Event Enrichment in Pairs with Gained Overlapping uORFs (PTC_neg)")
```

Columns: **event_type** = splice event type; **n_with_gained_overlap** = PTC_neg
pairs with gained overlapping uORFs that also have this event type;
**pct_gained_overlap** = prevalence in gained-overlap pairs;
**n_without_gained_overlap** = PTC_neg pairs without gained overlapping uORFs;
**pct_without** = prevalence in non-gained pairs; **OR** = Fisher's exact odds
ratio (OR > 1 = enriched in gained-overlap pairs); **p_value** = Fisher p-value.

**Skipped exon events are the most significantly enriched** event type in pairs
that gain overlapping uORFs (59.1% vs 43.7%, OR = 1.87, p = 4.3e-11). This
suggests that exon skipping contributes to overlapping uORF creation — possibly
by shifting the reading frame or altering the stop codon landscape in the 5'UTR.

Notably, alternative TSS events are **not** enriched (83.3% vs 82.8%,
OR = 1.03, p = 0.80) — they are ubiquitous across PTC-negative NMD pairs
regardless of overlapping uORF status. Alt_TSS events are likely important for
PTC-negative NMD more broadly (by extending the 5'UTR and increasing uORF count
overall), but this analysis specifically asks which events distinguish pairs
that gain *overlapping* uORFs from those that do not.

## 5'UTR Features Predict NMD Response Magnitude

```{r goal4b-logfc-correlation, fig.width=10, fig.height=5}
# ── Data provenance ──
# Source: uorf_comparison.rds → logfc_correlation
# Mixed-effects model: logFC ~ pred_nmd_prob + (1 | isoform_id)
# Accounts for 6 repeated cell-type measurements per isoform
# pred_nmd_prob from elastic net Model A (5'UTR features)

lfc <- uorf_comp$logfc_correlation
me <- lfc$mixed_effects
iso_means <- lfc$iso_means

# Scatter plot: isoform-level mean logFC vs predicted NMD probability
plot_data <- iso_means[iso_means$group %in% c("PTC_neg", "PTC_pos"), ]
plot_data$group_label <- factor(plot_data$group,
  levels = c("PTC_neg", "PTC_pos"),
  labels = c("NMD PTC-", "NMD PTC+"))

fig_lfc <- ggplot(plot_data, aes(x = pred_nmd_prob, y = mean_logFC)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "lm", color = "#ef8a62", se = TRUE) +
  facet_wrap(~group_label, scales = "free_y") +
  labs(
    title = "5'UTR Feature Score Predicts NMD Response Magnitude",
    subtitle = paste0(
      "PTC-neg: mixed-effects coef = ",
      round(me$PTC_neg$coefficient, 3), ", p = ",
      signif(me$PTC_neg$p_value, 3),
      "  |  PTC-pos: coef = ",
      round(me$PTC_pos$coefficient, 3), ", p = ",
      signif(me$PTC_pos$p_value, 3)),
    x = "Predicted NMD Probability (from 5'UTR features)",
    y = "Mean mashr logFC (across cell types)")
save_fig(fig_lfc, "fig8e_logfc_correlation", width = 10, height = 5)
```

If uORF-mediated NMD were a binary switch — either triggered or not — we would
expect no relationship between 5'UTR features and the *magnitude* of NMD
response among PTC-negative comparators (they are all NMD-responsive by
definition). Instead, we find a significant positive relationship.

A linear mixed-effects model (`logFC ~ pred_nmd_prob + (1 | isoform_id)`)
accounts for the 6 repeated cell-type measurements per isoform (mashr posterior
means). Within PTC-negative comparators, isoforms with higher predicted NMD
probability (based on 5'UTR features alone) show significantly stronger NMD
response (coefficient = `r round(me$PTC_neg$coefficient, 3)`, t =
`r round(me$PTC_neg$t_value, 1)`, p = `r signif(me$PTC_neg$p_value, 3)`,
n = `r fmt(me$PTC_neg$n_isoforms)` isoforms).

The same model applied to PTC-positive comparators shows **no relationship**
(coefficient = `r round(me$PTC_pos$coefficient, 3)`, p =
`r signif(me$PTC_pos$p_value, 3)`). This confirms that 5'UTR features
specifically modulate NMD in the PTC-negative group — they have no bearing on
NMD response when a PTC is already present.

**This suggests uORF-mediated NMD is a continuous, dose-dependent phenomenon**:
the more the 5'UTR is loaded with overlapping uORFs and strong initiation
contexts, the stronger the NMD response. This is consistent with a model where
multiple uORFs each contribute a small probability of ribosome engagement and
premature termination, with the cumulative effect determining overall NMD
sensitivity.

### Conclusion: 5'UTR uORFs as the Primary Non-PTC NMD Mechanism

The analyses in this section converge on a clear picture:

1. **Overlapping uORFs and ATG density are the strongest predictors** of
   PTC-negative NMD status (elastic net AUC = 0.875). These features are
   specific to PTC-negative comparators (AUC = 0.628 for PTC-positive,
   near chance).

2. **58% of PTC-negative pairs gain overlapping uORFs** compared to 20% of
   Controls (OR = 5.4). Skipped exon events are specifically enriched among
   pairs gaining overlapping uORFs (OR = 1.87, p = 4.3e-11), suggesting exon
   skipping contributes to overlapping uORF creation beyond the general
   structural differences between NMD and Control isoforms.

3. **5'UTR features predict NMD response *magnitude***, not just binary NMD
   status. The mixed-effects coefficient is highly significant in PTC-negative
   comparators (p = `r signif(me$PTC_neg$p_value, 3)`) and null in PTC-positive
   (p = `r signif(me$PTC_pos$p_value, 3)`), suggesting uORF-mediated NMD is a
   continuous, dose-dependent phenomenon.

4. **Our detection method is validated** against ORFik (Spearman r =
   `r round(ov$spearman_r, 3)` for non-overlapping uORFs) while uniquely
   capturing the `r fmt(ov$n_overlapping_ours)` overlapping uORFs that are
   the most biologically important category.

**Limitations:** uORF detection is sequence-based and does not confirm
translation. The correlation with NMD magnitude is observational and does not
establish that specific uORFs cause the observed NMD response. Other non-PTC
NMD mechanisms cannot be ruled out from structural data alone.

# 2d. NMD Response is Continuous {#sec2d-continuous}

Both stop codon and start codon features predict NMD response *magnitude*,
not just binary status. Using raw limma logFC values (independent per-cell-type
estimates, appropriate for mixed-effects modeling), we show that the unified
model's predicted probability correlates with NMD response strength.

**Data source:** Raw limma logFC values from per-cell-type DE files
(6 cell types: AT, DD, DD_ALI, DO, FB, MV). Predicted probabilities from
the unified model (Section 2a).

**Statistical approach:** Spearman correlation on isoform-level mean logFC
(averaged across cell types) for interpretability.

```{r sec2d-load-limma}
# ── Data provenance ──
# Source: limma DE files from /Users/petecastaldi/claude_projects/nmd/isocall_dge/limma/yul/
# Raw unshrunk logFC per cell type — independent estimates (no mashr shrinkage)
# Used as response variable for dose-response analysis

limma_dir <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/limma/yul"
limma_cts <- c(AT = "at", DD = "dd", DD_ALI = "ddali", DO = "do", FB = "fb", MV = "mv")

limma_logfc <- lapply(names(limma_cts), function(ct) {
  f <- file.path(limma_dir, sprintf("nmd_dge_%s_2026.3.3.csv", limma_cts[ct]))
  de <- read.csv(f, stringsAsFactors = FALSE)
  # Use transcript_id column
  id_col <- if ("transcript_id" %in% names(de)) "transcript_id" else "txid"
  data.frame(
    isoform_id = de[[id_col]],
    cell_type = ct,
    logFC = de$logFC,
    stringsAsFactors = FALSE
  )
})
limma_long <- do.call(rbind, limma_logfc)

# Mean logFC per isoform (across cell types)
mean_logfc <- limma_long %>%
  group_by(isoform_id) %>%
  summarise(
    mean_logFC = mean(logFC, na.rm = TRUE),
    n_cts = n(),
    .groups = "drop"
  )

# Merge with unified model scores (NMD isoforms only)
nmd_scored <- unified$scored_isoforms %>%
  filter(is_nmd == 1) %>%
  inner_join(mean_logfc, by = "isoform_id")

cat("NMD isoforms with both model scores and limma logFC:", nrow(nmd_scored), "\n")
cat("Mean logFC (NMD isoforms):", round(mean(nmd_scored$mean_logFC), 3), "\n")
```

## Unified Model Score vs NMD Response Magnitude

```{r sec2d-scatter, fig.width=10, fig.height=5}
# Spearman correlation
rho_all <- cor(nmd_scored$pred_prob, nmd_scored$mean_logFC, method = "spearman")
p_all <- cor.test(nmd_scored$pred_prob, nmd_scored$mean_logFC, method = "spearman")$p.value

# Split by PTC status for side-by-side
ptc_status <- ptc %>%
  filter(isoform_id %in% nmd_scored$isoform_id) %>%
  select(isoform_id, has_ptc)

nmd_scored_ptc <- nmd_scored %>%
  inner_join(ptc_status, by = "isoform_id") %>%
  mutate(PTC = ifelse(has_ptc, "PTC+", "PTC\u2212"))

rho_ptcpos <- with(nmd_scored_ptc[nmd_scored_ptc$has_ptc == TRUE, ],
                    cor(pred_prob, mean_logFC, method = "spearman"))
rho_ptcneg <- with(nmd_scored_ptc[nmd_scored_ptc$has_ptc == FALSE, ],
                    cor(pred_prob, mean_logFC, method = "spearman"))

fig_dose <- ggplot(nmd_scored_ptc,
                   aes(x = pred_prob, y = mean_logFC, color = PTC)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  facet_wrap(~PTC, scales = "free_x") +
  scale_color_manual(values = c("PTC+" = "#ef8a62", "PTC\u2212" = "#67a9cf")) +
  labs(title = "Unified Model Score vs NMD Response Magnitude",
       subtitle = sprintf("All NMD: rho = %.3f | PTC+: %.3f | PTC\u2212: %.3f",
                          rho_all, rho_ptcpos, rho_ptcneg),
       x = "Predicted NMD probability (unified model)",
       y = "Mean logFC (limma, across cell types)",
       color = "") +
  theme(legend.position = "none")
save_fig(fig_dose, "fig_dose_response_unified", width = 10, height = 5)
```

NMD is a continuous response, not a binary switch. The unified model's
predicted probability correlates with actual NMD response magnitude
(Spearman rho = `r round(rho_all, 3)`, p = `r signif(p_all, 3)`). This
holds for both PTC-positive isoforms (where stop codon position determines
response strength) and PTC-negative isoforms (where 5'UTR features predict
response magnitude).

# 3. Shared ATGs — The Same Start Codon, Multiple Fates {#sec3-shared-atgs}

The unified model treats PTC and uORF features as independent predictors.
But biologically, where do the overlapping uORFs come from? We find that the
ATGs creating overlapping uORFs are not random — many are functional CDS starts
in other isoforms from the same gene.

**Data source:** Overlapping uORF ATGs identified from C2 PTC-negative paired
comparators (paired framework). The ATG role analysis then examines ALL coding
isoforms of those genes (broader than pairs) to determine whether the ATG
serves as CDS start in other isoforms.

```{r sec3-shared-atg-analysis}
# ── Data provenance ──
# Source: utr5_features.rds → isoform_features (overlapping uORFs),
#   cds.rds (CDS start positions), structures.rds (strand info)
# Population: Genes with PTC-negative NMD comparators that have overlapping uORFs.
#   For each such gene, we examine ALL expressed coding isoforms.

# PTC-negative NMD comparator isoforms with overlapping uORFs
ptc_c2_as <- set_results[["c2_allsamples"]]$ptc_assoc
coding_ids_set <- cds$isoform_id[cds$coding_status == "coding"]
sec3_coding_c2 <- profiles_list[["c2_allsamples"]] %>%
  filter(reference_isoform_id %in% coding_ids_set,
         comparator_isoform_id %in% coding_ids_set)

ptc_lookup <- ptc_c2_as$pair_summary[, c("comparator_isoform_id", "comp_has_ptc")]
coding_c2_ptc <- merge(sec3_coding_c2, ptc_lookup, by = "comparator_isoform_id", all.x = TRUE)
ptc_neg_comps <- coding_c2_ptc %>%
  filter(comp_has_ptc == FALSE)

# Get UTR5 features for PTC-neg comparators
utr5_paired <- readRDS(file.path(cache_dir, "utr5_features.rds"))
ptc_neg_with_uorfs <- utr5_paired$isoform_features %>%
  filter(isoform_id %in% ptc_neg_comps$comparator_isoform_id,
         n_orfs_overlapping > 0,
         !excluded)

cat("PTC-negative NMD comparators with overlapping uORFs:",
    nrow(ptc_neg_with_uorfs), "\n")

# Get the gene IDs for these isoforms
genes_with_overlap_uorfs <- unique(
  coding_c2_ptc$gene_id[coding_c2_ptc$comparator_isoform_id %in%
                           ptc_neg_with_uorfs$isoform_id]
)
cat("Genes with PTC-neg + overlapping uORFs:", length(genes_with_overlap_uorfs), "\n")

# For each gene, get CDS start positions of ALL coding isoforms
gene_cds_starts <- coding_structures %>%
  filter(gene_id %in% genes_with_overlap_uorfs) %>%
  inner_join(coding_cds %>% select(isoform_id, cds_start, cds_stop),
             by = "isoform_id") %>%
  mutate(
    cds_5prime = ifelse(strand == "+", cds_start, cds_stop)
  ) %>%
  select(isoform_id, gene_id, strand, cds_5prime)

# Count distinct CDS start positions per gene
cds_start_diversity <- gene_cds_starts %>%
  group_by(gene_id) %>%
  summarise(
    n_isoforms = n(),
    n_cds_starts = n_distinct(cds_5prime),
    .groups = "drop"
  )

cat("\nCDS start diversity in these genes:\n")
cat("  Genes with ≥2 CDS starts:",
    sum(cds_start_diversity$n_cds_starts >= 2),
    sprintf("(%.1f%%)\n", 100 * mean(cds_start_diversity$n_cds_starts >= 2)))
cat("  Genes with ≥3 CDS starts:",
    sum(cds_start_diversity$n_cds_starts >= 3),
    sprintf("(%.1f%%)\n", 100 * mean(cds_start_diversity$n_cds_starts >= 3)))

# Check NMD status of isoforms sharing CDS starts
nmd_ids_set <- nmd_class$all_samples$nmd
non_nmd_ids_set <- nmd_class$all_samples$non_nmd

gene_cds_starts$is_nmd <- gene_cds_starts$isoform_id %in% nmd_ids_set
gene_cds_starts$is_non_nmd <- gene_cds_starts$isoform_id %in% non_nmd_ids_set

# For each distinct CDS start in these genes, how many isoforms use it?
cds_start_usage <- gene_cds_starts %>%
  group_by(gene_id, cds_5prime) %>%
  summarise(
    n_isoforms = n(),
    n_nmd = sum(is_nmd),
    n_non_nmd = sum(is_non_nmd),
    .groups = "drop"
  )

# CDS starts used by both NMD and non-NMD isoforms
shared_starts <- cds_start_usage %>%
  filter(n_nmd > 0, n_non_nmd > 0)

cat("\nShared ATG analysis:\n")
cat("  Total distinct CDS starts:", nrow(cds_start_usage), "\n")
cat("  Used by both NMD and non-NMD:", nrow(shared_starts),
    sprintf("(%.1f%%)\n", 100 * nrow(shared_starts) / nrow(cds_start_usage)))
cat("  Non-NMD isoforms using shared starts:", sum(shared_starts$n_non_nmd), "\n")
cat("  NMD isoforms using shared starts:", sum(shared_starts$n_nmd), "\n")
```

The same ATG position can serve as a CDS start in one isoform, an
overlapping uORF in another, or internal CDS in a third — with the role
determined entirely by the exon path. Genes encode overlapping reading
frames through shared ATGs. The exon path — determined by TSS, internal
splicing, and TES — selects which frame is productive. NMD degrades
transcripts where the selected path produces a reading frame that
terminates upstream of EJCs.


# 4. Protein-Level Consequences of Normal Alternative Splicing {#goal5}

This analysis has been moved to a separate report:
**`05g_protein_report_mashr.Rmd`** / **`05g_protein_report_mashr.html`**

That report examines the dominant vs next-best non-NMD isoform for each gene,
including domain overlap analysis, domain enrichment tests, mass-spectrometry
validation, and cell-type-specific patterns.

# 5. Cross-Cell-Type Summary {#summary}

```{r sec6-summary-table}
# ── Data provenance ──
# Starting data: gene-matched profiles and cached analyses across all cell types
# Each row computed via helper functions that access set_results[[key]]:
#   - Pair counts: nrow(gene-matched profiles)
#   - PTC rates: compute_ptc_rates_row() — coding pairs from gene-matched profiles + ptc.rds
#   - Divergence: compute_divergence_row() — from cached quantifyPairDivergence()
#   - Productive frameshifts: from pf_data precompute (all_samples only)
# All helpers operate on gene-matched data (profiles_list was matched in STAGE 2)

summary_rows <- lapply(c("all_samples", main_cts), function(ct) {
  ct_label <- ct_to_label(ct)

  # NMD and Control pair counts
  c2_key <- paste0("c2_", ct_label)
  c4_key <- paste0("c4_", ct_label)
  n_nmd  <- if (c2_key %in% names(set_results)) nrow(set_results[[c2_key]]$profiles) else NA
  n_ctrl <- if (c4_key %in% names(set_results)) nrow(set_results[[c4_key]]$profiles) else NA

  # PTC rates
  ptc_row <- compute_ptc_rates_row(ct)
  ptc_rate_nmd  <- if (!is.null(ptc_row)) ptc_row$NMD_comp_ptc_pct else NA
  ptc_rate_ctrl <- if (!is.null(ptc_row)) ptc_row$Ctrl_comp_ptc_pct else NA
  ptc_or        <- if (!is.null(ptc_row)) ptc_row$fisher_OR else NA

  # PTC-neg %
  ptc_neg_pct <- if (!is.null(ptc_row)) round(100 - ptc_row$NMD_comp_ptc_pct, 1) else NA

  # Median pct_shared for C4
  div_row <- compute_divergence_row(ct)
  med_shared <- if (!is.null(div_row)) div_row$Ctrl_median_pct_shared else NA

  # Productive frameshifts (only available for all_samples)
  n_pf <- if (ct == "all_samples") nrow(pf_data$characterization) else NA

  data.frame(
    cell_type = ct,
    N_NMD_pairs = n_nmd,
    N_Ctrl_pairs = n_ctrl,
    PTC_rate_NMD_pct = ptc_rate_nmd,
    PTC_rate_Ctrl_pct = ptc_rate_ctrl,
    PTC_OR = ptc_or,
    PTC_neg_pct = ptc_neg_pct,
    Median_pct_shared_C4 = med_shared,
    N_productive_FS = n_pf)
})

summary_table <- do.call(rbind, summary_rows)
render_dt(summary_table, caption = "Table 11: Cross-Cell-Type Summary")
write.csv(summary_table, file.path(table_dir, "table11_cross_ct_summary.csv"), row.names = FALSE)
```

Columns: **cell_type** = sample set; **N_NMD/N_Ctrl_pairs** = gene-matched pair
counts; **PTC_rate_NMD/Ctrl_pct** = PTC rate among coding comparators;
**PTC_OR** = Fisher OR for PTC enrichment; **PTC_neg_pct** = percentage of
coding NMD comparators without PTC; **Median_pct_shared_C4** = median exonic
sequence shared in Control pairs (from `quantifyPairDivergence()`);
**N_productive_FS** = productive frameshifts (all_samples only).

# Session Info

```{r session-info}
sessionInfo()
```
