---
title: "[LEGACY 2026-06-13] ORF-Centric NMD Analysis"
subtitle: "Superseded by 05_final_report_gencode_scope_2026-06-15.Rmd as of 2026-06-15. Retained for sensitivity / reproducibility of prior submission."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    fig_width: 8
    fig_height: 5
---

```{r legacy-banner, echo=FALSE, results='asis'}
cat('
<div style="background-color: #fff3cd; border-left: 6px solid #ffc107; padding: 12px; margin: 16px 0; font-size: 14px;">
<strong>⚠ LEGACY REPORT</strong> — This report uses the two-pass TD2 → ref-AUG reclassification
framework, which has been superseded as of 2026-06-15. The current canonical analysis is in
<code>05_final_report_gencode_scope_2026-06-15.Rmd</code>, which uses each comparator\'s own
GENCODE-annotated stop codon (for the all-3-ENST + coding-CDS scope) or ref-AUG projection
on a 1:1 gene-matched re-intersected universe (for the larger Section C scope).
The analyses in this report are retained for reproducibility of the prior submission.
</div>
')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
library(Isopair)
library(dplyr)
library(tidyr)
library(ggplot2)
library(DT)
library(patchwork)
source("visualization_functions.R")

render_dt <- function(df, caption = NULL, pageLength = 15, ...) {
  datatable(df, caption = caption, rownames = FALSE,
            options = list(pageLength = pageLength, scrollX = TRUE), ...)
}

theme_pub <- theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 10, color = "grey40"),
        legend.position = "bottom",
        panel.grid.minor = element_blank())
theme_set(theme_pub)

fmt <- function(x) format(x, big.mark = ",")

save_fig <- function(p, name, width = 8, height = 5) {
  fig_dir <- "figures_mashr"
  if (!dir.exists(fig_dir)) dir.create(fig_dir)
  ggsave(file.path(fig_dir, paste0(name, ".pdf")), p, width = width, height = height)
  ggsave(file.path(fig_dir, paste0(name, ".png")), p, width = width, height = height, dpi = 150)
  p
}
```

# Overview

This report examines NMD from an ORF-centric perspective. The key finding:
**most isoforms classified as "PTC-negative" by computational CDS prediction
actually produce premature stop codons with downstream EJCs when the reading
frame is traced from the reference isoform's CDS start ATG.** The PTC+/PTC-
distinction is largely an artifact of CDS prediction on the NMD comparator.

**ORF definition used throughout:** An ORF is defined as an ATG codon followed
by the first in-frame stop codon (TAA, TAG, or TGA), scanning downstream in
the same reading frame. Minimum length: 9 nucleotides (3 codons).

# 0. Data Loading

```{r load-data}
# ── Data provenance ──
# Primary data: gene-matched C2 (NMD) and C4 (Control) pairs from mashr pipeline
# Infrastructure: cds.rds, structures.rds (from Isopair pipeline)
# Precomputed: ref_atg_analysis.rds (from 05r_ref_atg_analysis.R)
#              orf_landscape.rds (from 05m_orf_landscape.R)

cds <- readRDS("data_mashr/cds.rds")
structures <- readRDS("data_mashr/structures.rds")
nmd_class <- readRDS("data_mashr/nmd_classification.rds")

ref_atg <- readRDS("data_mashr/analysis_cache/ref_atg_analysis.rds")
orf_landscape <- readRDS("data_mashr/analysis_cache/orf_landscape.rds")

c2 <- ref_atg$c2
c4 <- ref_atg$c4

cat("NMD (C2) pairs:", nrow(c2), "\n")
cat("Control (C4) pairs:", nrow(c4), "\n")

# Gene symbol lookup
de_dir <- "/Users/petecastaldi/claude_projects/nmd/isocall_dge/mashr"
gene_symbol_lookup <- character(0)
for (f in list.files(de_dir, pattern = "csv$", full.names = TRUE)) {
  d <- read.csv(f, stringsAsFactors = FALSE)[, c("gene_id", "hgnc_symbol")]
  d <- d[!duplicated(d$gene_id) & !is.na(d$hgnc_symbol) & d$hgnc_symbol != "", ]
  new <- setdiff(d$gene_id, names(gene_symbol_lookup))
  gene_symbol_lookup[d$gene_id[d$gene_id %in% new]] <- d$hgnc_symbol[d$gene_id %in% new]
}
get_gene_sym <- function(gid) {
  sym <- gene_symbol_lookup[gid]
  if (is.na(sym) || sym == "") gid else sym
}
```

# 1. Reference-Anchored PTC Analysis

## 1a. Three Groups

The reference isoform in each pair is the dominant non-NMD isoform — often
GENCODE-annotated with a curated CDS. We trust its CDS start ATG as the
position where the ribosome initiates translation. For each NMD comparator,
we ask: is this ATG still accessible?

```{r groups}
# Group assignment
c2$group <- ifelse(c2$ref_atg_exonic_in_comp, "Ref ATG available", "Ref ATG lost")

group_tab <- c2 %>%
  count(group) %>%
  mutate(pct = round(100 * n / sum(n), 1))

render_dt(group_tab, caption = "NMD Pair Groups by Reference ATG Availability")
```

Column definitions: **group** = whether the reference's CDS start ATG is
exonic in the comparator; **n** = pair count; **pct** = percentage.

`r group_tab$n[group_tab$group == "Ref ATG available"]` pairs
(`r group_tab$pct[group_tab$group == "Ref ATG available"]`%) retain the
reference ATG in their exonic sequence. The ribosome can still initiate at
the same position. `r group_tab$n[group_tab$group == "Ref ATG lost"]` pairs
(`r group_tab$pct[group_tab$group == "Ref ATG lost"]`%) have lost the
reference ATG — the exon containing it is absent from the comparator.

## 1b. Tracing the Reference ATG

For the `r group_tab$n[group_tab$group == "Ref ATG available"]` pairs where
the reference ATG is available, we trace the reading frame from that ATG
through the comparator's transcript sequence and compare to the reference.

```{r orf-tracing, fig.width=10, fig.height=5}
# ORF length comparison: ref ATG traced through ref vs comp
atg_avail <- c2[c2$group == "Ref ATG available" &
                  !is.na(c2$ref_orf_length) & !is.na(c2$comp_orf_length_from_ref_atg), ]

cat("Pairs with valid ORF tracing:", nrow(atg_avail), "\n")
cat("Ref ORF length: median", median(atg_avail$ref_orf_length), "nt\n")
cat("Comp ORF (from ref ATG): median", median(atg_avail$comp_orf_length_from_ref_atg), "nt\n")
cat("Comp ORF shorter:", sum(atg_avail$comp_orf_length_from_ref_atg < atg_avail$ref_orf_length),
    sprintf("(%.1f%%)\n", 100 * mean(atg_avail$comp_orf_length_from_ref_atg < atg_avail$ref_orf_length)))

# Density plot of ORF length ratio
atg_avail$orf_ratio <- atg_avail$comp_orf_length_from_ref_atg / atg_avail$ref_orf_length

fig_ratio <- ggplot(atg_avail, aes(x = orf_ratio)) +
  geom_density(fill = "#ef8a62", alpha = 0.6) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  scale_x_continuous(limits = c(0, 2), breaks = seq(0, 2, 0.25)) +
  labs(title = "ORF Length Ratio: Comparator / Reference (from Reference ATG)",
       subtitle = sprintf("n = %s pairs | Ratio < 1 = ORF truncated in comparator",
                          fmt(nrow(atg_avail))),
       x = "ORF length ratio (comparator / reference)", y = "Density")
save_fig(fig_ratio, "fig_orf_ratio_ref_atg", width = 10, height = 5)
```

## 1c. Reclassification

```{r reclassification}
# Reclassification table
reclass <- c2 %>%
  mutate(original_class = ifelse(original_ptc, "Originally PTC+", "Originally PTC\u2212")) %>%
  count(original_class, category) %>%
  arrange(original_class, -n) %>%
  mutate(pct = round(100 * n / sum(n), 1))

render_dt(reclass, caption = "Reclassification: Original PTC Status vs Reference-ATG-Traced Category")

# Key numbers
orig_ptc_pos <- sum(c2$original_ptc)
orig_ptc_neg <- sum(!c2$original_ptc)
reclass_eff_ptc <- sum(c2$category == "effectively_ptc" & !c2$original_ptc, na.rm = TRUE)
total_eff_ptc <- orig_ptc_pos + reclass_eff_ptc

cat("\nOriginal PTC+:", orig_ptc_pos, "\n")
cat("Originally PTC-, reclassified as effectively PTC:", reclass_eff_ptc,
    sprintf("(%.1f%% of PTC-)\n", 100 * reclass_eff_ptc / orig_ptc_neg))
cat("Total PTC-mediated:", total_eff_ptc, "of", nrow(c2),
    sprintf("(%.1f%%)\n", 100 * total_eff_ptc / nrow(c2)))
```

Column definitions: **original_class** = PTC status from SQANTI CDS annotation;
**category** = reclassification based on tracing the reference ATG through the
comparator; **n** = pair count; **pct** = percentage of all pairs.

`r sprintf("%.1f%%", 100 * reclass_eff_ptc / orig_ptc_neg)` of originally
PTC-negative pairs are reclassified as effectively PTC when the reading frame
is traced from the reference's CDS start. Combined with the original PTC+
pairs, **`r sprintf("%.1f%%", 100 * total_eff_ptc / nrow(c2))`% of all NMD
pairs are PTC-mediated** when viewed from the reference ATG.

## 1d. Splice Events in Reclassified Pairs

```{r reclassified-events, fig.width=10, fig.height=5}
reclass_pairs <- c2[c2$category == "effectively_ptc" & !c2$original_ptc, ]

cat("Reclassified effectively-PTC pairs:", nrow(reclass_pairs), "\n\n")
cat("Attributed splice events:\n")
evt_tab <- sort(table(reclass_pairs$attr_event), decreasing = TRUE)
for (nm in names(evt_tab)) {
  cat(sprintf("  %s: %d (%.1f%%)\n", nm, evt_tab[nm], 100 * evt_tab[nm] / nrow(reclass_pairs)))
}

# Bar chart
evt_df <- data.frame(
  event = names(evt_tab),
  n = as.integer(evt_tab),
  stringsAsFactors = FALSE) %>%
  mutate(pct = round(100 * n / sum(n), 1),
         event = factor(event, levels = event))

fig_events <- ggplot(evt_df, aes(x = event, y = n, fill = event)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = sprintf("%d\n(%.0f%%)", n, pct)), vjust = -0.3, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Splice Events in Reclassified PTC Pairs",
       subtitle = sprintf("Originally PTC\u2212, reclassified as effectively PTC (n = %s)",
                          fmt(nrow(reclass_pairs))),
       x = NULL, y = "Number of pairs") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
save_fig(fig_events, "fig_reclass_splice_events", width = 10, height = 5)
```

# 2. CDS-Agnostic ORF Landscape

Having shown that PTC status depends on which ATG we trace from, we now
drop CDS annotations entirely and compare ORF landscapes across the three
groups using only the transcript sequence and exon structure.

## 2a. ORF Counts by Group

```{r orf-landscape-groups, fig.width=10, fig.height=6}
# Map isoform IDs to groups
# Reference isoforms
ref_ids <- unique(c2$reference_isoform_id)
# NMD comparators by group
comp_atg_avail <- c2$comparator_isoform_id[c2$group == "Ref ATG available"]
comp_atg_lost <- c2$comparator_isoform_id[c2$group == "Ref ATG lost"]
# Control comparators
ctrl_comp_ids <- unique(c4$comparator_isoform_id)

orf_landscape$analysis_group <- case_when(
  orf_landscape$isoform_id %in% ref_ids ~ "Reference",
  orf_landscape$isoform_id %in% comp_atg_avail ~ "NMD / ATG available",
  orf_landscape$isoform_id %in% comp_atg_lost ~ "NMD / ATG lost",
  orf_landscape$isoform_id %in% ctrl_comp_ids ~ "Control",
  TRUE ~ NA_character_
)

orf_grouped <- orf_landscape %>%
  filter(!is.na(analysis_group)) %>%
  mutate(analysis_group = factor(analysis_group,
    levels = c("Reference", "NMD / ATG available", "NMD / ATG lost", "Control")))

# Summary table
orf_summary <- orf_grouped %>%
  group_by(analysis_group) %>%
  summarise(
    n = n(),
    median_orfs = median(n_orfs),
    median_nmd_competent = median(n_nmd_competent),
    median_best_kozak_nmd = round(median(best_kozak_nmd, na.rm = TRUE), 2),
    median_junctions = median(n_junctions),
    pct_any_nmd_comp = round(100 * mean(has_any_nmd_competent), 1),
    .groups = "drop")

render_dt(orf_summary, caption = "ORF Landscape by Analysis Group")
```

Column definitions: **median_orfs** = median plausibly-translated (Kozak-passing)
ORFs per isoform; **median_nmd_competent** = median ORFs whose stop has ≥1
downstream EJC; **median_best_kozak_nmd** = median (across isoforms) of the best
PWM Kozak score among NMD-competent ORFs (continuous log-odds); **median_junctions**
= median exon-exon junctions; **pct_any_nmd_comp** = % of isoforms with at least
one NMD-competent ORF.

```{r orf-boxplots, fig.width=12, fig.height=5}
# Side-by-side boxplots
p1 <- ggplot(orf_grouped, aes(x = analysis_group, y = n_orfs, fill = analysis_group)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
  scale_y_continuous(limits = c(0, 150)) +
  labs(title = "A. Total ORFs", y = "ORFs per isoform", x = NULL) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))

p2 <- ggplot(orf_grouped, aes(x = analysis_group, y = n_nmd_competent, fill = analysis_group)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
  scale_y_continuous(limits = c(0, 80)) +
  labs(title = "B. ORFs with Downstream EJCs", y = "NMD-competent ORFs", x = NULL) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))

fig_orf_box <- p1 | p2
save_fig(fig_orf_box, "fig_orf_landscape_groups", width = 12, height = 5)
```

## 2b. Pair-Level Deltas

```{r pair-deltas, fig.width=10, fig.height=5}
# Compute delta (comp - ref) for NMD and Control pairs
build_deltas <- function(pairs, label) {
  ref_orf <- orf_landscape[match(pairs$reference_isoform_id, orf_landscape$isoform_id),
    c("n_orfs", "n_nmd_competent")]
  comp_orf <- orf_landscape[match(pairs$comparator_isoform_id, orf_landscape$isoform_id),
    c("n_orfs", "n_nmd_competent")]
  data.frame(
    label = label,
    delta_orfs = comp_orf$n_orfs - ref_orf$n_orfs,
    delta_nmd_comp = comp_orf$n_nmd_competent - ref_orf$n_nmd_competent,
    stringsAsFactors = FALSE)
}

deltas_nmd <- build_deltas(c2, "NMD (C2)")
deltas_ctrl <- build_deltas(c4, "Control (C4)")
deltas_all <- rbind(deltas_nmd, deltas_ctrl)

cat("Pair-level deltas (comparator - reference):\n")
for (lb in c("NMD (C2)", "Control (C4)")) {
  sub <- deltas_all[deltas_all$label == lb, ]
  cat(sprintf("  %s: delta ORFs median %+.0f, delta NMD-comp median %+.0f, %% gaining NMD-comp %.1f%%\n",
      lb, median(sub$delta_orfs, na.rm=TRUE), median(sub$delta_nmd_comp, na.rm=TRUE),
      100 * mean(sub$delta_nmd_comp > 0, na.rm=TRUE)))
}

fig_deltas <- ggplot(deltas_all, aes(x = label, y = delta_nmd_comp, fill = label)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = c("NMD (C2)" = "#ef8a62", "Control (C4)" = "#67a9cf")) +
  labs(title = "Change in NMD-Competent ORFs (Comparator \u2212 Reference)",
       subtitle = "NMD comparators gain ORFs with downstream EJCs; Controls do not",
       x = NULL, y = "\u0394 ORFs with downstream EJCs") +
  theme(legend.position = "none")
save_fig(fig_deltas, "fig_delta_nmd_competent", width = 10, height = 5)
```

## 2c. Splice Events: Frameshift Rate

```{r splice-frameshift, fig.width=10, fig.height=5}
# For each pair, check SE events and their frameshift rate
classify_se <- function(profiles, label) {
  n_pairs <- nrow(profiles)
  has_se <- 0L; has_fs_se <- 0L; has_if_se <- 0L
  for (i in seq_len(n_pairs)) {
    de <- profiles$detailed_events[[i]]
    if (is.null(de) || nrow(de) == 0) next
    se <- de[de$event_type == "SE", ]
    if (nrow(se) == 0) next
    has_se <- has_se + 1L
    if (any(!is.na(se$bp_diff) & se$bp_diff %% 3 != 0)) has_fs_se <- has_fs_se + 1L
    if (any(!is.na(se$bp_diff) & se$bp_diff %% 3 == 0)) has_if_se <- has_if_se + 1L
  }
  data.frame(label = label, n_pairs = n_pairs, has_se = has_se,
             has_frameshift_se = has_fs_se, has_inframe_se = has_if_se,
             stringsAsFactors = FALSE)
}

# Need original profiles for detailed_events
profiles_c2 <- readRDS("data_mashr/profiles_c2_allsamples.rds")
profiles_c4 <- readRDS("data_mashr/profiles_c4_allsamples.rds")

shared <- inner_join(
  unique(profiles_c2[, c("gene_id", "reference_isoform_id")]),
  unique(profiles_c4[, c("gene_id", "reference_isoform_id")]),
  by = c("gene_id", "reference_isoform_id"))
gm_c2 <- semi_join(profiles_c2, shared, by = c("gene_id", "reference_isoform_id"))
gm_c4 <- semi_join(profiles_c4, shared, by = c("gene_id", "reference_isoform_id"))
coding_ids <- cds$isoform_id[cds$coding_status == "coding"]
gm_c2_coding <- gm_c2 %>% filter(reference_isoform_id %in% coding_ids,
                                    comparator_isoform_id %in% coding_ids)
gm_c4_coding <- gm_c4 %>% filter(reference_isoform_id %in% coding_ids,
                                    comparator_isoform_id %in% coding_ids)

se_nmd <- classify_se(gm_c2_coding, "NMD")
se_ctrl <- classify_se(gm_c4_coding, "Control")
se_both <- rbind(se_nmd, se_ctrl)

se_both$pct_se <- round(100 * se_both$has_se / se_both$n_pairs, 1)
se_both$pct_fs_among_se <- round(100 * se_both$has_frameshift_se / pmax(se_both$has_se, 1), 1)

render_dt(se_both, caption = "Skipped Exon Events and Frameshift Rate")
```

Column definitions: **has_se** = pairs with at least one SE event;
**has_frameshift_se** = pairs with SE where bp_diff is not divisible by 3;
**pct_fs_among_se** = % of SE-containing pairs where at least one SE causes
a frameshift.

The frameshift rate among SE events is
`r se_both$pct_fs_among_se[se_both$label == "NMD"]`% in NMD pairs vs
`r se_both$pct_fs_among_se[se_both$label == "Control"]`% in Control pairs.
NMD isoforms are enriched for frameshift-causing exon skips — the same
mechanism that creates PTCs when it disrupts the main reading frame.

# 3. Synthesis

```{r synthesis}
# Collect key numbers for inline use
n_total <- nrow(c2)
n_orig_ptc <- orig_ptc_pos
n_reclass <- reclass_eff_ptc
n_total_ptc <- total_eff_ptc
pct_total_ptc <- round(100 * n_total_ptc / n_total, 1)
n_atg_lost <- sum(c2$category == "ref_atg_lost", na.rm = TRUE)
n_truly_nonptc <- sum(c2$category %in% c("no_downstream_ejc", "truncated_no_ejc",
                                           "same_or_longer_with_ejc"), na.rm = TRUE)
```

Of `r fmt(n_total)` gene-matched coding NMD pairs:

- **`r fmt(n_orig_ptc)`** (`r round(100 * n_orig_ptc / n_total, 1)`%) were
  originally classified as PTC+ by SQANTI CDS prediction
- **`r fmt(n_reclass)`** additional pairs are **effectively PTC** when traced
  from the reference ATG — the same reading frame hits a premature stop with
  downstream EJCs
- **Total PTC-mediated: `r fmt(n_total_ptc)` (`r pct_total_ptc`%)**
- **`r fmt(n_atg_lost)`** pairs (`r round(100 * n_atg_lost / n_total, 1)`%)
  lost the reference ATG entirely
- **`r fmt(n_truly_nonptc)`** pairs
  (`r round(100 * n_truly_nonptc / n_total, 1)`%) are truly non-PTC even from
  the reference ATG

The PTC+/PTC- distinction is largely an artifact of computational CDS
prediction. When the CDS predictor (GeneMarkS-T) encounters an NMD comparator
with a disrupted reading frame, it often selects a different ATG as the
"main" CDS start — one whose ORF terminates in the last exon (no downstream
EJCs). This makes the isoform appear PTC-negative. But the reference's
original ATG is still there, and the reading frame from it terminates
prematurely.

The underlying mechanism is consistent: **frameshift-causing exon skips
(`r se_both$pct_fs_among_se[se_both$label == "NMD"]`% of NMD SE events vs
`r se_both$pct_fs_among_se[se_both$label == "Control"]`% in Control)
disrupt reading frames, creating premature stop codons with downstream EJCs.**
Whether we call this a "PTC" or a "non-PTC uORF" depends entirely on which
ATG the CDS predictor selects as the main start codon.

```{r session-info}
sessionInfo()
```
