---
title: "NMD+/PTC- Genes (Reference-AUG) — Hand-off for GO Enrichment"
subtitle: "Standalone per-isoform PTC determination — Isocall / mashr DIE"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    fig_width: 8
    fig_height: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(data.table)
library(knitr)
library(DT)
suppressMessages(library(org.Hs.eg.db))

render_dt <- function(df, caption = NULL, pageLength = 10, ...) {
  datatable(df, caption = caption, rownames = FALSE,
            options = list(pageLength = pageLength, scrollX = TRUE), ...)
}
fmt <- function(x) formatC(x, format = "d", big.mark = ",")
pct <- function(num, den) sprintf("%.1f%%", 100 * num / den)
strip_ver <- function(x) sub("\\..*", "", x)

# ── Paths ──
# This report renders from code/ptcneg_go_handoff/. Repo root is two levels up.
repo_root <- "../.."
out_dir   <- "."   # deliverables land alongside this report (tracked under code/)

# Canonical reference-AUG ORF + PTC + NMD cache, produced by the NMD_orf_model_v5_4ct
# repo (NOT the Isopair pairwise pipeline). die_date inside = 2026.3.10 mashr.
CACHE <- file.path(repo_root, "nmd_orf_model_v5_4ct", "tmp",
                   "ref_orf_ptc_cache_with_nmd_2026.5.12.rds")

.d   <- Sys.Date()
DATE <- sprintf("%d.%d.%d", as.integer(format(.d, "%Y")),
                as.integer(format(.d, "%m")), as.integer(format(.d, "%d")))
```

## 1. Purpose & approach

This report assembles the **NMD+/PTC- gene set** — NMD-responsive isoforms that
lack a premature termination codon (PTC) — and packages it, with a matched gene
background, for downstream **GO enrichment** (topGO or any over-representation
tool, run by Yul).

Two design choices distinguish this from the manuscript's Isopair analysis:

1. **PTC is determined per-isoform from the reference AUG, not from TD2.**
   TD2/SQANTI's predicted CDS is biased against PTC-containing ORFs, so its NMD
   prediction cannot be trusted here (CLAUDE.md §"SQANTI/TD2 CDS bias"). Instead
   we use the **reference-AUG projection**: the gene's annotated reference start
   codon is traced into each isoform, the in-frame stop is located, and downstream
   exon-junction complexes (EJCs) are counted under the 50-nt rule. This needs no
   reference-vs-comparator *pairwise* framework — it is a per-isoform projection
   against the gene's canonical AUG.

2. **The analysis is restricted to reference-AUG-compatible isoforms**, i.e.
   isoforms in which the reference start codon is exonic and traceable
   (`ref_atg_exonic_in_comp == TRUE`). Isoforms that lost the reference AUG are
   out of scope by construction.

**Definitions**

| Term | Operational definition |
|---|---|
| Reference-AUG compatible | `ref_atg_exonic_in_comp == TRUE` |
| PTC- (no PTC) | `n_downstream_ejc == 0` from the ref-AUG stop (categories `no_downstream_ejc`, `truncated_no_ejc`) |
| PTC+ | `n_downstream_ejc > 0` (`effectively_ptc`) — excluded from the target |
| NMD+ (NMD-responsive) | `nmd_responsive_any == TRUE` (mashr-significant up-regulation under SMG1i in ≥1 of AT, DD, FB, MV) |
| **Target (NMD+/PTC-)** | AUG-compatible **and** PTC- **and** NMD+ |

**Provenance.** All calls come from `ref_orf_ptc_cache_with_nmd_2026.5.12.rds`
(`$trace_aug`), produced by the `NMD_orf_model_v5_4ct` repo; NMD status is the
`2026.3.10` isoform-level mashr DIE (4 cell types). This file is a local input
(under a gitignored `tmp/`); regenerate it from the model repo if absent.

## 2. Target set: NMD+/PTC- isoforms (reference-AUG)

```{r load}
# ── Data provenance ──
# $trace_aug: one row per reference-AUG-projected isoform (keyed comparator_isoform_id),
# with ref-AUG ORF/EJC trace + per-cell-type mashr DIE + NMD summary flags.
stopifnot(file.exists(CACHE))
cache <- readRDS(CACHE)
ta <- as.data.table(cache$trace_aug)

# Reference-AUG-compatible isoforms only
augc <- ta[ref_atg_exonic_in_comp == TRUE]
augc[, ptc_neg := !is.na(n_downstream_ejc) & n_downstream_ejc == 0]
augc[, ensembl_gene_id := strip_ver(gene_id)]

# Target: AUG-compatible, PTC-negative, NMD-responsive
target_iso <- augc[ptc_neg == TRUE & nmd_responsive_any == TRUE]

n_aug      <- nrow(augc)
n_tgt_iso  <- nrow(target_iso)
n_tgt_gene <- uniqueN(target_iso$ensembl_gene_id)
```

The reference-AUG-compatible set contains **`r fmt(n_aug)` isoforms**. The target
NMD+/PTC- set is **`r fmt(n_tgt_iso)` isoforms across `r fmt(n_tgt_gene)` genes**.

```{r subgroup}
sub <- target_iso[, .(n_isoforms = .N,
                      n_genes = uniqueN(ensembl_gene_id)), by = category][order(-n_isoforms)]
render_dt(sub, caption = "Table 1: Target composition by ref-AUG PTC- category",
          pageLength = 5)
```

```{r fraction}
# Live PTC-negative fraction among NMD-responsive, AUG-compatible isoforms —
# the standalone analogue of the manuscript's "~15% of NMD isoforms lack a PTC".
nmd_aug   <- augc[nmd_responsive_any == TRUE]
frac_ptcneg <- pct(sum(nmd_aug$ptc_neg), nrow(nmd_aug))
```

Among **`r fmt(nrow(nmd_aug))`** NMD-responsive reference-AUG-compatible isoforms,
**`r fmt(sum(nmd_aug$ptc_neg))` (`r frac_ptcneg`)** are PTC-negative — the
genome-wide, pairwise-free counterpart of the manuscript's PTC-negative NMD
fraction.

```{r per-ct}
# Per-cell-type support: in how many of the 4 CTs is each target isoform
# NMD-significant? (n_ct_sig_005 from the cache.)
ctt <- target_iso[, .N, by = n_ct_sig_005][order(n_ct_sig_005)]
setnames(ctt, "N", "n_isoforms")
render_dt(ctt, caption = "Table 2: Target isoforms by number of cell types with significant NMD response",
          pageLength = 5)
```

## 3. Background universe (AUG-assessed genes)

```{r universe}
# Universe = genes with >=1 reference-AUG-compatible isoform, i.e. genes that
# were actually eligible to be called NMD+/PTC-. This matched background avoids
# diluting the ORA with genes that could never enter the target.
universe <- data.table(ensembl_gene_id = unique(augc$ensembl_gene_id))
n_universe <- nrow(universe)
all_in_uni <- all(target_iso$ensembl_gene_id %in% universe$ensembl_gene_id)
```

The background holds **`r fmt(n_universe)` genes** (≥1 reference-AUG-compatible
isoform). All `r fmt(n_tgt_gene)` target genes are contained in it
(`r all_in_uni`).

## 4. Gene identifiers

```{r ids}
# Symbol + Entrez from org.Hs.eg.db — the same annotation topGO uses via
# annFUN.org, so identifiers stay internally consistent with the enrichment.
all_ens <- union(universe$ensembl_gene_id, target_iso$ensembl_gene_id)
sym <- mapIds(org.Hs.eg.db, keys = all_ens, column = "SYMBOL",
              keytype = "ENSEMBL", multiVals = "first")
ez  <- mapIds(org.Hs.eg.db, keys = all_ens, column = "ENTREZID",
              keytype = "ENSEMBL", multiVals = "first")
id_map <- data.table(ensembl_gene_id = all_ens,
                     hgnc_symbol = sym[all_ens], entrez_id = ez[all_ens])

cov_tab <- data.table(
  set    = c("Target (NMD+/PTC-)", "Background (universe)"),
  n      = c(n_tgt_gene, n_universe),
  symbol = c(pct(sum(!is.na(id_map[ensembl_gene_id %in% target_iso$ensembl_gene_id]$hgnc_symbol)), n_tgt_gene),
             pct(sum(!is.na(id_map[ensembl_gene_id %in% universe$ensembl_gene_id]$hgnc_symbol)), n_universe)),
  entrez = c(pct(sum(!is.na(id_map[ensembl_gene_id %in% target_iso$ensembl_gene_id]$entrez_id)), n_tgt_gene),
             pct(sum(!is.na(id_map[ensembl_gene_id %in% universe$ensembl_gene_id]$entrez_id)), n_universe)))
render_dt(cov_tab, caption = "Table 3: Identifier mapping coverage", pageLength = 5)
```

## 5. Deliverables

```{r exports}
# ── Gene-level target (for GO): one row per gene, with supporting-isoform detail ──
tgt_genes <- target_iso[, .(
  n_ptcneg_isoforms = .N,
  categories        = paste(sort(unique(category)), collapse = ";"),
  max_n_ct_sig      = max(n_ct_sig_005),
  max_abs_logfc     = round(max(abs(max_sig_logfc), na.rm = TRUE), 3)
), by = ensembl_gene_id]
tgt_genes <- merge(tgt_genes, id_map, by = "ensembl_gene_id", all.x = TRUE)
setcolorder(tgt_genes, c("ensembl_gene_id", "hgnc_symbol", "entrez_id"))

# ── Background with target flag (this IS the topGO allGenes source) ──
bg <- merge(universe, id_map, by = "ensembl_gene_id", all.x = TRUE)
bg[, is_target := ensembl_gene_id %in% target_iso$ensembl_gene_id]
setcolorder(bg, c("ensembl_gene_id", "hgnc_symbol", "entrez_id", "is_target"))

# ── Isoform-level detail (traceability / subgroup-stratified GO) ──
iso_detail <- target_iso[, .(
  comparator_isoform_id, reference_isoform_id, ensembl_gene_id,
  category, ref_orf_length, comp_orf_length, n_downstream_ejc,
  n_ct_sig_005, max_sig_logfc, max_sig_direction)]

f_tgt <- file.path(out_dir, sprintf("nmd_ptcneg_refAUG_target_genes_%s.csv", DATE))
f_bg  <- file.path(out_dir, sprintf("nmd_ptcneg_refAUG_background_genes_%s.csv", DATE))
f_sym <- file.path(out_dir, sprintf("nmd_ptcneg_refAUG_target_symbols_%s.txt", DATE))
f_iso <- file.path(out_dir, sprintf("nmd_ptcneg_refAUG_target_isoforms_%s.csv", DATE))

fwrite(tgt_genes, f_tgt)
fwrite(bg, f_bg)
writeLines(sort(unique(na.omit(tgt_genes$hgnc_symbol))), f_sym)
fwrite(iso_detail, f_iso)

exports <- data.table(
  file = c(basename(f_tgt), basename(f_bg), basename(f_sym), basename(f_iso)),
  contents = c(
    sprintf("%s NMD+/PTC- genes — IDs + supporting-isoform detail (GO target)", fmt(nrow(tgt_genes))),
    sprintf("%s AUG-assessed genes — IDs + is_target flag (GO background / topGO allGenes)", fmt(nrow(bg))),
    sprintf("%s HGNC symbols — plain list (Enrichr / g:Profiler paste)", fmt(length(unique(na.omit(tgt_genes$hgnc_symbol))))),
    sprintf("%s target isoforms — per-isoform trace (traceability, subgroup GO)", fmt(nrow(iso_detail)))))
render_dt(exports, caption = "Table 4: Exported deliverables", pageLength = 5)
```

```{r preview}
render_dt(head(tgt_genes[order(-max_n_ct_sig, -max_abs_logfc)], 20),
          caption = "Preview: target gene table (top 20 by cell-type support)")
```

### Running topGO (for Yul)

The background CSV is the topGO universe; column `is_target` flags the
NMD+/PTC- genes. Build the named factor on Entrez IDs and run:

```r
library(topGO); library(org.Hs.eg.db); library(data.table)
bg <- fread("nmd_ptcneg_refAUG_background_genes_2026.6.8.csv")
bg <- bg[!is.na(entrez_id)][!duplicated(entrez_id)]
allGenes <- factor(as.integer(bg$is_target), levels = c(0, 1)); names(allGenes) <- bg$entrez_id
GOdata <- new("topGOdata", ontology = "BP", allGenes = allGenes,
              annot = annFUN.org, mapping = "org.Hs.eg.db", ID = "entrez", nodeSize = 10)
res <- runTest(GOdata, algorithm = "weight01", statistic = "fisher")
GenTable(GOdata, weightFisher = res, topNodes = 30)
```

For Enrichr / g:Profiler, paste `*_target_symbols_*.txt` and supply column
`ensembl_gene_id` of the background CSV as the custom background. Subgroup
GO is available by filtering `category` in the isoform-level file.

## 6. Verification

```{r verify}
checks <- list(
  "Target ⊆ universe"                  = all(target_iso$ensembl_gene_id %in% universe$ensembl_gene_id),
  "Target genes unique"                = !any(duplicated(tgt_genes$ensembl_gene_id)),
  "Universe genes unique"              = !any(duplicated(bg$ensembl_gene_id)),
  "is_target count == target genes"    = sum(bg$is_target) == n_tgt_gene,
  "All target AUG-compatible"          = all(target_iso$ref_atg_exonic_in_comp == TRUE),
  "All target PTC- (0 downstream EJC)" = all(target_iso$n_downstream_ejc == 0),
  "No effectively_ptc in target"       = !any(target_iso$category == "effectively_ptc"),
  "All target NMD-responsive"          = all(target_iso$nmd_responsive_any == TRUE),
  "Category n_isoforms sum to total"   = sum(sub$n_isoforms) == n_tgt_iso
)
chk <- data.table(check = names(checks), pass = unlist(checks))
stopifnot(all(chk$pass))
render_dt(chk, caption = "Table 5: Consistency checks (all must pass)", pageLength = 10)
```

All consistency checks pass.

```{r session}
sessionInfo()
```
