---
title: "Expression distribution of non-NMD-responsive isoforms in NMD+ vs NMD− genes (Smg1i samples)"
author: "Pete Castaldi"
date: "2026-04-17"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
    number_sections: true
---

# Question

For each primary lung cell type (AT, DD, DO, FB, MV), how does the expression
of non-NMD-responsive isoforms in **Smg1i**-treated samples differ between:

- **NMD+ genes**: genes containing at least one NMD-responsive isoform, versus
- **NMD− genes**: genes containing zero NMD-responsive isoforms (but at least
  one non-NMD-responsive isoform, so that they contribute points to the
  comparison).

The analysis is restricted to Smg1i samples and uses library-size-normalized
CPM (**not** TMM-normalized), per project directive.

# Data provenance

- Differential expression: isocall / limma results from
  `isocall_dge/limma/pete/nmd_isocall_dge_{at,dd,do,fb,mv}_2026.3.1.csv`.
  These define each isoform's NMD response in the corresponding cell type.
- Expression matrix: `results/isoform_transitions/Version_6.0/data/isocall/dge_isocall_unfiltered.rds`
  — isocall-derived isoform DGEList (645,273 isoforms × 38 samples),
  unfiltered (no expression filter applied) per user directive. Selected over
  `rds/dge_isoform_nofilter_2026.2.7.rds` because the latter is an
  oarfish-era DGEList whose novel-isoform IDs (`PB.x.y`) do not match the
  isocall DE output's `ENSG....novelN` IDs; the isocall DGEList's row names
  match the DE `transcript_id` column exactly.
- CPM: computed with `edgeR::cpm(counts, lib.size = colSums(counts),
  normalized.lib.sizes = FALSE)` so that column scaling is raw library size
  **only** (no TMM norm factor).
- Cell-type coding: DE file suffix `{at,dd,do,fb,mv}` corresponds to
  `dge$samples$ct` levels `{AT, DD, DO, FB, MV}` (DD_ALI excluded per user
  directive).
- NMD classification thresholds (project standard, memory): NMD-responsive =
  `adj.P.Val < 0.05 & logFC > 0`; non-NMD-responsive = `adj.P.Val > 0.50`;
  isoforms with `0.05 ≤ adj.P.Val ≤ 0.50` are ambiguous and excluded from
  the isoform-level comparison (but they still count toward classifying a
  gene as NMD+ if any sibling isoform is NMD-responsive — see §4).
- Novel isoforms (`ENSG....novelN`): included per user directive.
- All biotypes included per user directive.
- Gene-level expression filter: in each cell type, a gene is retained only
  if its mean DMSO gene CPM (sum of isoform CPMs per sample, averaged
  across that cell type's DMSO samples) is ≥ 1. DMSO (not Smg1i) is used
  so the filter doesn't preferentially retain NMD+ genes whose targets
  are stabilised by Smg1i. See §7 for the before/after counts.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5, dpi = 110
)
options(bitmapType = "cairo")
```

```{r libs}
suppressPackageStartupMessages({
  library(edgeR)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(DT)
  library(knitr)
  library(scales)
})

theme_pub <- theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )

render_dt <- function(df, caption = NULL, digits = 3) {
  num_cols <- names(df)[sapply(df, is.numeric)]
  DT::datatable(
    df,
    caption = caption,
    options = list(pageLength = 10, scrollX = TRUE),
    rownames = FALSE
  ) |> DT::formatSignif(columns = num_cols, digits = digits)
}
```

# Parameters

```{r params}
PROJ_ROOT <- tryCatch(here::here(), error = function(e) normalizePath(".."))
DE_DIR    <- file.path(PROJ_ROOT, "isocall_dge", "limma", "pete")
DGELIST   <- file.path(
  PROJ_ROOT,
  "results", "isoform_transitions", "Version_6.0", "data", "isocall",
  "dge_isocall_unfiltered.rds"
)

CELL_TYPES <- c("AT", "DD", "DO", "FB", "MV")      # DD_ALI excluded per user
DE_FILE_SUFFIX <- c(AT = "at", DD = "dd", DO = "do", FB = "fb", MV = "mv")

NMD_RESP_PADJ  <- 0.05
NMD_RESP_LFC   <- 0
NON_NMD_PADJ   <- 0.50

stopifnot(dir.exists(DE_DIR), file.exists(DGELIST))
```

Parameter definitions: **NMD_RESP_PADJ** = upper bound on adjusted p-value for
NMD-responsive classification; **NMD_RESP_LFC** = lower bound on `logFC` (Smg1i vs
DMSO) for NMD-responsive classification; **NON_NMD_PADJ** = lower bound on
adjusted p-value for non-NMD-responsive classification.

# Load and classify isoforms

```{r classify-isoforms}
# Read the 5 per-cell-type DE files and classify each isoform.
classify_de <- function(de_df) {
  de_df |>
    mutate(
      nmd_class = case_when(
        adj.P.Val < NMD_RESP_PADJ  & logFC > NMD_RESP_LFC ~ "NMD_responsive",
        adj.P.Val > NON_NMD_PADJ                          ~ "non_NMD_responsive",
        TRUE                                              ~ "ambiguous"
      )
    )
}

de_list <- lapply(CELL_TYPES, function(ct) {
  f <- file.path(DE_DIR, paste0("nmd_isocall_dge_", DE_FILE_SUFFIX[[ct]], "_2026.3.1.csv"))
  read.csv(f, stringsAsFactors = FALSE) |>
    classify_de() |>
    mutate(ct = ct)
})
names(de_list) <- CELL_TYPES

# Sanity: all DE files should share the same transcript universe
de_sizes <- sapply(de_list, nrow)
stopifnot(length(unique(de_sizes)) == 1)
```

```{r classify-isoforms-summary}
iso_class_summary <- bind_rows(lapply(de_list, function(d) {
  data.frame(
    cell_type           = d$ct[1],
    n_total             = nrow(d),
    n_NMD_responsive    = sum(d$nmd_class == "NMD_responsive"),
    n_non_NMD_responsive= sum(d$nmd_class == "non_NMD_responsive"),
    n_ambiguous         = sum(d$nmd_class == "ambiguous")
  )
}))
kable(iso_class_summary, caption = "Isoform-level NMD classification per cell type.")
```

Columns: **cell_type** = cell population; **n_total** = number of isoforms in
the DE output; **n_NMD_responsive** = isoforms with `adj.P.Val <
`r NMD_RESP_PADJ`` and `logFC > `r NMD_RESP_LFC``; **n_non_NMD_responsive** =
isoforms with `adj.P.Val > `r NON_NMD_PADJ``; **n_ambiguous** = isoforms whose
`adj.P.Val` falls in the intermediate range (`r NMD_RESP_PADJ` ≤ adj.P.Val ≤
`r NON_NMD_PADJ`), excluded from isoform-level comparisons but still counted
when classifying the parent gene (below).

# Classify genes per cell type

A gene is **NMD+** in cell type *c* if it has ≥1 NMD-responsive isoform in *c*;
otherwise it is **NMD−**. Under this rule, ambiguous isoforms do not change a
gene's classification (they neither promote it to NMD+ nor prevent it from
being NMD−). We further restrict NMD− genes to those with ≥1
**non-NMD-responsive** isoform, since only those contribute points to the
downstream isoform-level comparison.

```{r classify-genes}
gene_class_list <- lapply(de_list, function(d) {
  d |>
    group_by(gene_id) |>
    summarise(
      n_iso              = n(),
      n_nmd_resp         = sum(nmd_class == "NMD_responsive"),
      n_non_nmd_resp     = sum(nmd_class == "non_NMD_responsive"),
      n_ambig            = sum(nmd_class == "ambiguous"),
      .groups = "drop"
    ) |>
    mutate(
      gene_group = case_when(
        n_nmd_resp >= 1                      ~ "NMD+_gene",
        n_nmd_resp == 0 & n_non_nmd_resp >= 1 ~ "NMD-_gene",
        TRUE                                 ~ "excluded"   # only ambiguous isoforms
      )
    )
})
names(gene_class_list) <- CELL_TYPES

gene_class_summary <- bind_rows(lapply(CELL_TYPES, function(ct) {
  g <- gene_class_list[[ct]]
  data.frame(
    cell_type     = ct,
    n_genes_total = nrow(g),
    n_NMDplus     = sum(g$gene_group == "NMD+_gene"),
    n_NMDminus    = sum(g$gene_group == "NMD-_gene"),
    n_excluded    = sum(g$gene_group == "excluded")
  )
}))
kable(gene_class_summary, caption = "Gene-level NMD classification per cell type.")
```

Columns: **cell_type** = cell population; **n_genes_total** = distinct genes
present in the DE output; **n_NMDplus** = genes with ≥1 NMD-responsive isoform
(`NMD+_gene`); **n_NMDminus** = genes with 0 NMD-responsive and ≥1
non-NMD-responsive isoform (`NMD-_gene`); **n_excluded** = genes whose only
isoforms are ambiguous (excluded from the comparison).

# Compute non-TMM CPM per Smg1i sample, average within cell type

```{r load-dge}
dge <- readRDS(DGELIST)
stopifnot(inherits(dge, "DGEList"))

n_smg1i_by_ct <- dge$samples |>
  as.data.frame() |>
  filter(treatment == "Smg1i", ct %in% CELL_TYPES) |>
  count(ct, name = "n_smg1i_samples")
kable(n_smg1i_by_ct, caption = "Number of Smg1i samples contributing to the mean CPM per cell type.")
```

Columns: **ct** = cell population (`dge$samples$ct`); **n_smg1i_samples** =
count of Smg1i-treated samples in the DGEList for that cell type.

```{r compute-cpm}
dge_nontmm <- dge
dge_nontmm$samples$norm.factors <- 1
cpm_mat <- edgeR::cpm(dge_nontmm, normalized.lib.sizes = FALSE)
stopifnot(all(abs(colSums(cpm_mat) - 1e6) < 1))

dge_txids <- rownames(dge_nontmm$counts)
stopifnot(!anyDuplicated(dge_txids))

# Sanity: every DE transcript must be present in the DGEList so the
# inner_join below doesn't silently drop isoforms.
de_txids <- unique(unlist(lapply(de_list, function(d) d$transcript_id)))
missing_in_dge <- setdiff(de_txids, dge_txids)
if (length(missing_in_dge) > 0) {
  stop("DE transcript_ids missing from DGEList rownames: n = ",
       length(missing_in_dge),
       "  e.g., ", paste(head(missing_in_dge, 3), collapse = ", "))
}
```

```{r mean-cpm-per-ct}
# For each cell type, compute mean CPM per isoform across its Smg1i samples.
mean_cpm_by_ct <- lapply(CELL_TYPES, function(ct) {
  smg1i_cols <- rownames(dge_nontmm$samples)[
    dge_nontmm$samples$treatment == "Smg1i" &
      dge_nontmm$samples$ct == ct
  ]
  if (length(smg1i_cols) == 0) {
    stop("No Smg1i samples found for cell type ", ct)
  }
  m <- rowMeans(cpm_mat[, smg1i_cols, drop = FALSE])
  data.frame(transcript_id = names(m), mean_cpm_smg1i = unname(m), ct = ct)
})
names(mean_cpm_by_ct) <- CELL_TYPES
```

Columns in `mean_cpm_by_ct[[ct]]`: **transcript_id** = isoform ID (matches DE
`transcript_id`); **mean_cpm_smg1i** = mean of library-size-normalized (non-TMM)
CPM across the Smg1i samples of cell type `ct`; **ct** = cell type.

# Gene-level expression — distribution and filter calibration

Purpose: the comparison is biased if low-expressed genes are systematically
classified NMD− simply because their isoforms lacked power to reach
`adj.P.Val < `r NMD_RESP_PADJ``. This section summarises gene-level
expression in DMSO and Smg1i per cell type so a minimum-expression filter
can be chosen on the same CPM scale used downstream.

Definition: **gene CPM** = sum of isoform CPMs (non-TMM) over all isoforms
of that gene within a sample; **mean gene CPM by treatment** = mean of
gene CPM across that cell type's samples of that treatment. No filter is
applied yet — this section informs the threshold only.

```{r gene-cpm-by-treatment}
gene_ids_by_tx <- setNames(dge$genes$gene_id, rownames(dge$counts))
stopifnot(!is.null(gene_ids_by_tx), !any(is.na(gene_ids_by_tx)))

gene_cpm_mat <- rowsum(cpm_mat, group = gene_ids_by_tx, reorder = TRUE)

samp <- dge_nontmm$samples |>
  as.data.frame() |>
  mutate(sample_col = rownames(dge_nontmm$samples)) |>
  filter(ct %in% CELL_TYPES, treatment %in% c("DMSO", "Smg1i"))

gene_cpm_summary <- lapply(CELL_TYPES, function(ct_i) {
  lapply(c("DMSO", "Smg1i"), function(tx) {
    cols <- samp$sample_col[samp$ct == ct_i & samp$treatment == tx]
    if (length(cols) == 0) return(NULL)
    mg <- rowMeans(gene_cpm_mat[, cols, drop = FALSE])
    data.frame(ct = ct_i, treatment = tx, gene_id = names(mg), mean_gene_cpm = unname(mg))
  }) |> bind_rows()
}) |> bind_rows()
```

## Quantile table

```{r gene-cpm-quantiles}
q_probs <- c(0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95)

gene_cpm_quantiles <- gene_cpm_summary |>
  group_by(ct, treatment) |>
  summarise(
    n_genes = n(),
    q05 = quantile(mean_gene_cpm, 0.05),
    q10 = quantile(mean_gene_cpm, 0.10),
    q25 = quantile(mean_gene_cpm, 0.25),
    q50 = quantile(mean_gene_cpm, 0.50),
    q75 = quantile(mean_gene_cpm, 0.75),
    q90 = quantile(mean_gene_cpm, 0.90),
    q95 = quantile(mean_gene_cpm, 0.95),
    .groups = "drop"
  ) |>
  mutate(across(starts_with("q"), ~round(.x, 3)))

kable(gene_cpm_quantiles,
      caption = "Quantiles of mean gene-level CPM (non-TMM) by cell type and treatment.")
```

Columns: **ct** = cell type; **treatment** = DMSO or Smg1i; **n_genes** =
distinct genes contributing; **q05–q95** = quantiles of the per-gene mean
CPM within that (cell type, treatment) cell.

## Fraction of genes excluded at candidate thresholds

```{r gene-cpm-threshold-scan}
thresholds <- c(0.1, 0.5, 1, 2, 5, 10)

threshold_scan <- lapply(thresholds, function(thr) {
  gene_cpm_summary |>
    group_by(ct, treatment) |>
    summarise(
      threshold     = thr,
      n_genes       = n(),
      n_kept        = sum(mean_gene_cpm >= thr),
      pct_kept      = round(100 * mean(mean_gene_cpm >= thr), 2),
      pct_excluded  = round(100 * mean(mean_gene_cpm <  thr), 2),
      .groups = "drop"
    )
}) |> bind_rows() |> arrange(treatment, ct, threshold)

kable(threshold_scan,
      caption = "Genes kept/excluded at candidate minimum-mean-CPM thresholds, by cell type and treatment.")
```

Columns: **ct**, **treatment** — as above; **threshold** = candidate
minimum on mean gene CPM; **n_genes** = total distinct genes in that (ct,
treatment) cell; **n_kept** / **pct_kept** = genes with mean gene CPM ≥
threshold; **pct_excluded** = complement.

## Density of log10(mean gene CPM)

```{r gene-cpm-density, fig.width=11, fig.height=6}
ggplot(gene_cpm_summary,
       aes(x = log10(mean_gene_cpm + 0.01),
           colour = treatment, fill = treatment)) +
  geom_density(alpha = 0.25, linewidth = 0.5) +
  facet_wrap(~ ct, ncol = 3, scales = "free_y") +
  scale_colour_manual(values = c("DMSO" = "#2c3e50", "Smg1i" = "#c0392b"),
                      name = "Treatment") +
  scale_fill_manual(values = c("DMSO" = "#2c3e50", "Smg1i" = "#c0392b"),
                    name = "Treatment") +
  labs(
    x = "log10(mean gene CPM + 0.01), non-TMM",
    y = "Density",
    title = "Gene-level expression distribution, DMSO vs Smg1i"
  ) +
  theme_pub
```

All genes in the DGEList are shown (no truncation). The small offset
`+0.01` is only to keep log10 defined at zero.

# Apply DMSO expression filter

Filter motivation: genes with low baseline (DMSO) expression have low power
to reach `adj.P.Val < `r NMD_RESP_PADJ`` for any isoform, so they are
systematically classified NMD− by the "no NMD-responsive isoforms" rule
regardless of their underlying biology. Thresholding on **DMSO** (not
Smg1i) keeps the criterion independent of NMD-target stabilisation, which
preferentially inflates Smg1i expression of exactly the NMD+ genes we
don't want to bias for.

Rule: in each cell type, retain a gene only if its mean DMSO gene CPM
(sum of isoform CPMs summed within each DMSO sample, then averaged across
that cell type's DMSO samples) is ≥ `r 1` CPM. Both NMD+ and NMD− genes
are subject to the same bar.

```{r apply-dmso-filter}
DMSO_CPM_MIN <- 1

dmso_cpm_by_gene <- gene_cpm_summary |>
  filter(treatment == "DMSO") |>
  select(ct, gene_id, dmso_mean_gene_cpm = mean_gene_cpm)

gene_class_list_filt <- lapply(CELL_TYPES, function(ct_i) {
  gcls <- gene_class_list[[ct_i]]
  dm   <- dmso_cpm_by_gene |> filter(ct == ct_i) |> select(-ct)
  gcls |>
    left_join(dm, by = "gene_id") |>
    mutate(
      dmso_mean_gene_cpm = ifelse(is.na(dmso_mean_gene_cpm), 0, dmso_mean_gene_cpm),
      passed_dmso_filter = dmso_mean_gene_cpm >= DMSO_CPM_MIN,
      gene_group_pre  = gene_group,
      gene_group = case_when(
        !passed_dmso_filter                    ~ "excluded_low_expression",
        n_nmd_resp >= 1                        ~ "NMD+_gene",
        n_nmd_resp == 0 & n_non_nmd_resp >= 1  ~ "NMD-_gene",
        TRUE                                   ~ "excluded"
      )
    )
})
names(gene_class_list_filt) <- CELL_TYPES

filt_summary <- bind_rows(lapply(CELL_TYPES, function(ct_i) {
  g <- gene_class_list_filt[[ct_i]]
  n_pre_plus   <- sum(g$gene_group_pre == "NMD+_gene")
  n_pre_minus  <- sum(g$gene_group_pre == "NMD-_gene")
  n_post_plus  <- sum(g$gene_group == "NMD+_gene")
  n_post_minus <- sum(g$gene_group == "NMD-_gene")
  data.frame(
    cell_type              = ct_i,
    n_NMDplus_pre          = n_pre_plus,
    n_NMDminus_pre         = n_pre_minus,
    n_NMDplus_post         = n_post_plus,
    n_NMDminus_post        = n_post_minus,
    pct_NMDplus_dropped    = round(100 * (n_pre_plus  - n_post_plus)  / max(1, n_pre_plus),  2),
    pct_NMDminus_dropped   = round(100 * (n_pre_minus - n_post_minus) / max(1, n_pre_minus), 2),
    n_excluded_lowexpr     = sum(g$gene_group == "excluded_low_expression")
  )
}))

kable(filt_summary,
      caption = paste0("Gene classification before and after the DMSO mean gene CPM \u2265 ",
                       DMSO_CPM_MIN, " filter, per cell type."))
```

Columns: **cell_type** = cell population; **n_NMDplus_pre** /
**n_NMDminus_pre** = NMD+ / NMD− gene counts from the pre-filter
classification (`r format(nrow(gene_class_list_filt[[1]]), big.mark=",")`
genes total per cell type); **n_NMDplus_post** / **n_NMDminus_post** =
counts after requiring mean DMSO gene CPM ≥ `r 1` in the same cell type;
**pct_NMDplus_dropped** / **pct_NMDminus_dropped** = percentage of each
pre-filter set removed by the DMSO filter; **n_excluded_lowexpr** = genes
(of any pre-filter group) that fell below the DMSO CPM threshold.

Interpretation cue: if the NMD− drop % is much larger than the NMD+ drop %,
that confirms the original bias — the NMD− set was disproportionately
made of low-expression genes.

# Assemble the per-isoform comparison dataset

```{r assemble}
# For each cell type, keep only non-NMD-responsive isoforms, attach their
# post-filter parent-gene classification, and attach the Smg1i mean CPM.
iso_cmp_by_ct <- lapply(CELL_TYPES, function(ct) {
  de   <- de_list[[ct]]
  gcls <- gene_class_list_filt[[ct]]
  cpm  <- mean_cpm_by_ct[[ct]]

  de |>
    filter(nmd_class == "non_NMD_responsive") |>
    inner_join(gcls |> select(gene_id, gene_group, dmso_mean_gene_cpm),
               by = "gene_id") |>
    filter(gene_group %in% c("NMD+_gene", "NMD-_gene")) |>
    inner_join(cpm |> select(transcript_id, mean_cpm_smg1i), by = "transcript_id") |>
    mutate(ct = ct, log2_cpm_p1 = log2(mean_cpm_smg1i + 1))
})
names(iso_cmp_by_ct) <- CELL_TYPES

iso_cmp_all <- bind_rows(iso_cmp_by_ct)

cmp_counts <- iso_cmp_all |>
  count(ct, gene_group, name = "n_isoforms") |>
  pivot_wider(names_from = gene_group, values_from = n_isoforms, values_fill = 0)
kable(cmp_counts, caption = "Number of non-NMD-responsive isoforms entering the comparison (after DMSO filter), by cell type and parent-gene class.")
```

Columns: **ct** = cell type; **NMD+_gene** = number of non-NMD-responsive
isoforms whose parent gene has at least one NMD-responsive isoform in that cell
type; **NMD-_gene** = number of non-NMD-responsive isoforms whose parent gene
has no NMD-responsive isoforms in that cell type. Denominator: all
non-NMD-responsive isoforms in that cell type (see `n_non_NMD_responsive` in
the isoform classification table above). The two columns sum to the
denominator less any isoforms whose parent gene was classified as `excluded`
(parent gene has only ambiguous isoforms — not possible here because the
child is by definition non-NMD-responsive, so this sum equals the denominator
within each cell type).

# Distribution comparison

```{r mw-tests}
# Mann–Whitney U test (two-sided) per cell type on log2(CPM+1).
mw_by_ct <- iso_cmp_all |>
  group_by(ct) |>
  group_modify(~ {
    d <- .x
    w <- wilcox.test(
      log2_cpm_p1 ~ gene_group,
      data = d, exact = FALSE, conf.int = FALSE
    )
    tibble(
      n_NMDplus      = sum(d$gene_group == "NMD+_gene"),
      n_NMDminus     = sum(d$gene_group == "NMD-_gene"),
      median_NMDplus = median(d$log2_cpm_p1[d$gene_group == "NMD+_gene"]),
      median_NMDminus= median(d$log2_cpm_p1[d$gene_group == "NMD-_gene"]),
      W              = unname(w$statistic),
      p_value        = w$p.value
    )
  }) |>
  ungroup() |>
  mutate(
    median_diff = median_NMDplus - median_NMDminus,
    direction   = ifelse(median_NMDplus > median_NMDminus,
                         "higher in NMD+ genes",
                         "higher in NMD- genes"),
    p_adj_BH    = p.adjust(p_value, method = "BH")
  )

kable(
  mw_by_ct |>
    select(ct, n_NMDplus, n_NMDminus,
           median_NMDplus, median_NMDminus, median_diff,
           direction, W, p_value, p_adj_BH) |>
    mutate(across(c(median_NMDplus, median_NMDminus, median_diff), ~round(.x, 3)),
           p_value = signif(p_value, 3),
           p_adj_BH= signif(p_adj_BH, 3)),
  caption = "Mann–Whitney U comparison of non-NMD-responsive isoform expression (log2 CPM+1, non-TMM, Smg1i mean) between NMD+ and NMD− genes, per cell type."
)
```

Columns: **ct** = cell type; **n_NMDplus** / **n_NMDminus** = non-NMD-responsive
isoforms whose parent gene is NMD+ / NMD−; **median_NMDplus** /
**median_NMDminus** = median log2(CPM+1) within each parent-gene class;
**median_diff** = `median_NMDplus − median_NMDminus` on the log2 scale;
**direction** = which class has the higher median; **W** = Mann–Whitney U
statistic (two-sided, normal approximation); **p_value** = raw test p-value;
**p_adj_BH** = Benjamini–Hochberg adjusted p-value across the `r length(CELL_TYPES)`
cell-type tests.

## Density plot — pooled across cell types

Pooled view of the same population, ignoring cell-type structure: each
non-NMD-responsive isoform contributes one point per cell type in which it
is classified non-NMD-responsive, so a single isoform may contribute up to
`r length(CELL_TYPES)` observations. This is descriptive, not a valid joint
inference (see the per-cell-type Mann–Whitney table above for the primary
test).

```{r density-pooled, fig.width=9, fig.height=5}
pooled_medians <- iso_cmp_all |>
  group_by(gene_group) |>
  summarise(median_log2cpm_p1 = median(log2_cpm_p1), .groups = "drop")

ggplot(iso_cmp_all,
       aes(x = log2_cpm_p1, colour = gene_group, fill = gene_group)) +
  geom_density(alpha = 0.25, linewidth = 0.7) +
  geom_vline(data = pooled_medians,
             aes(xintercept = median_log2cpm_p1, colour = gene_group),
             linetype = "dashed", linewidth = 0.5, show.legend = FALSE) +
  scale_colour_manual(values = c("NMD+_gene" = "#c0392b", "NMD-_gene" = "#2c3e50"),
                      name = "Parent gene class") +
  scale_fill_manual(values = c("NMD+_gene" = "#c0392b", "NMD-_gene" = "#2c3e50"),
                    name = "Parent gene class") +
  labs(
    x = "log2(mean Smg1i CPM + 1), non-TMM",
    y = "Density",
    title = "Pooled expression distribution of non-NMD-responsive isoforms",
    subtitle = "Dashed lines: group medians"
  ) +
  theme_pub
```

## Density plot — per cell type

Population for this figure: non-NMD-responsive isoforms (adj.P.Val >
`r NON_NMD_PADJ`) in each cell type, expressed as mean log2(CPM+1) across
that cell type's Smg1i samples, colored by whether the parent gene has ≥1
NMD-responsive isoform in the same cell type. Counts per (cell type,
parent-gene class) are shown in the preceding table. Parent genes are
restricted to those passing the DMSO mean gene CPM ≥ `r 1` filter; no
other CPM or gene-filter thresholds are applied.

```{r density, fig.width=11, fig.height=6}
ggplot(iso_cmp_all,
       aes(x = log2_cpm_p1, colour = gene_group, fill = gene_group)) +
  geom_density(alpha = 0.25, linewidth = 0.6) +
  facet_wrap(~ ct, ncol = 3, scales = "free_y") +
  scale_colour_manual(values = c("NMD+_gene" = "#c0392b", "NMD-_gene" = "#2c3e50"),
                      name = "Parent gene class") +
  scale_fill_manual(values = c("NMD+_gene" = "#c0392b", "NMD-_gene" = "#2c3e50"),
                    name = "Parent gene class") +
  labs(
    x = "log2(mean Smg1i CPM + 1), non-TMM",
    y = "Density",
    title = "Expression distribution of non-NMD-responsive isoforms in NMD+ vs NMD− genes"
  ) +
  theme_pub
```

## Empirical cumulative distribution

```{r ecdf, fig.width=11, fig.height=6}
ggplot(iso_cmp_all,
       aes(x = log2_cpm_p1, colour = gene_group)) +
  stat_ecdf(geom = "step", linewidth = 0.6) +
  facet_wrap(~ ct, ncol = 3) +
  scale_colour_manual(values = c("NMD+_gene" = "#c0392b", "NMD-_gene" = "#2c3e50"),
                      name = "Parent gene class") +
  labs(
    x = "log2(mean Smg1i CPM + 1), non-TMM",
    y = "ECDF",
    title = "ECDF of non-NMD-responsive isoform expression, per cell type"
  ) +
  theme_pub
```

## Boxplot with jitter (all isoforms shown)

```{r box, fig.width=10, fig.height=5}
ggplot(iso_cmp_all,
       aes(x = gene_group, y = log2_cpm_p1, fill = gene_group)) +
  geom_boxplot(outlier.shape = NA, width = 0.5, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 0.2, alpha = 0.08) +
  facet_wrap(~ ct, nrow = 1) +
  scale_fill_manual(values = c("NMD+_gene" = "#c0392b", "NMD-_gene" = "#2c3e50"),
                    guide = "none") +
  labs(
    x = NULL,
    y = "log2(mean Smg1i CPM + 1), non-TMM",
    title = "Per-isoform expression by parent-gene class"
  ) +
  theme_pub +
  theme(axis.text.x = element_text(angle = 0))
```

No values are truncated in the three plots above — all non-NMD-responsive
isoforms in the comparison universe are shown (`r nrow(iso_cmp_all)` isoforms
total across the `r length(CELL_TYPES)` cell types).

# Sensitivity: pooled meta-analysis

```{r pooled}
pooled <- wilcox.test(
  log2_cpm_p1 ~ gene_group,
  data = iso_cmp_all, exact = FALSE, conf.int = FALSE
)

pooled_tbl <- tibble(
  n_NMDplus       = sum(iso_cmp_all$gene_group == "NMD+_gene"),
  n_NMDminus      = sum(iso_cmp_all$gene_group == "NMD-_gene"),
  median_NMDplus  = median(iso_cmp_all$log2_cpm_p1[iso_cmp_all$gene_group == "NMD+_gene"]),
  median_NMDminus = median(iso_cmp_all$log2_cpm_p1[iso_cmp_all$gene_group == "NMD-_gene"]),
  median_diff     = median_NMDplus - median_NMDminus,
  W               = unname(pooled$statistic),
  p_value         = pooled$p.value
) |>
  mutate(across(c(median_NMDplus, median_NMDminus, median_diff), ~round(.x, 3)),
         p_value = signif(p_value, 3))

kable(pooled_tbl,
      caption = "Pooled Mann–Whitney U test across all cell types. Note: this
      treats isoforms as independent even though a gene can contribute multiple
      non-NMD isoforms in more than one cell type; interpret as a descriptive
      summary rather than a valid joint inference. See the per-cell-type table
      above for the primary result.")
```

Columns: as defined in the per-cell-type Mann–Whitney table above.

# Full per-isoform results

```{r prepare-full-table}
iso_cmp_full <- iso_cmp_all |>
  transmute(
    cell_type    = ct,
    transcript_id,
    gene_id,
    gene_symbol  = gene_name,
    gene_group,
    adj_P_Val    = signif(adj.P.Val, 3),
    logFC        = round(logFC, 3),
    mean_cpm_smg1i = round(mean_cpm_smg1i, 3),
    log2_cpm_p1  = round(log2_cpm_p1, 3)
  ) |>
  arrange(cell_type, desc(log2_cpm_p1))

FULL_CSV      <- "nonNMD_expression_distribution_2026.4.17_full_table.csv"
FULL_CSV_PATH <- file.path(PROJ_ROOT, "code", FULL_CSV)
write.csv(iso_cmp_full, FULL_CSV_PATH, row.names = FALSE)

PREVIEW_N        <- 2000
iso_cmp_preview  <- head(iso_cmp_full, PREVIEW_N)
n_full           <- nrow(iso_cmp_full)
n_truncated_out  <- n_full - nrow(iso_cmp_preview)
```

Full table: **`r format(n_full, big.mark = ",")` rows** total.
**Truncation disclosed:** the DT widget below shows only the first
`r format(PREVIEW_N, big.mark = ",")` rows after sorting by `cell_type`
(ascending) then `log2_cpm_p1` (descending). The remaining
**`r format(n_truncated_out, big.mark = ",")` rows** are omitted from the
widget but written in full to
[`r FULL_CSV`](`r FULL_CSV`) next to this HTML. Reason: embedding all
`r format(n_full, big.mark = ",")` rows as inline JSON produced a single
~72 MB line that prevented browsers from loading the page. No rows are
omitted from the density, ECDF, or boxplot figures above (those summarise
all `r format(n_full, big.mark = ",")` rows).

```{r dt-preview}
render_dt(
  iso_cmp_preview,
  caption = sprintf(
    "Preview: top %s of %s non-NMD-responsive isoforms (cell_type asc, log2(CPM+1) desc). Full table: %s.",
    format(PREVIEW_N, big.mark = ","),
    format(n_full, big.mark = ","),
    FULL_CSV
  )
)
```

Columns: **cell_type** = cell population; **transcript_id** = isoform ID
(ENST…, or ENSG….novelN for novel isoforms); **gene_id** = parent gene
(versioned ENSG from the DE output); **gene_symbol** = HGNC symbol as
reported in the DE output; **gene_group** = NMD+ / NMD− per-cell-type
classification of the parent gene; **adj_P_Val**, **logFC** = from the
corresponding cell type's limma/isocall DE CSV; **mean_cpm_smg1i** =
library-size-normalized (non-TMM) CPM averaged across Smg1i samples in that
cell type; **log2_cpm_p1** = `log2(mean_cpm_smg1i + 1)`.

# Session info

```{r session}
sessionInfo()
```
